### 2021-03-23 analysis/analysis.Rmd
some models

---
title: "Analysis for Project"
output: 
  html_document: 
    fig_height: 6
    fig_width: 9
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
source(here("analysis","check_packages.R"))
source(here("analysis","useful_functions.R"))
load(here("analysis","output","acs.RData"))
```

# Introduction
<a href="#top">Back to top</a>

Use this R Markdown to perform the main analysis for the project. I use this basically as a lab notebook. It contains the main analysis and a variety of sensitivity analysis. The code in this documents serves as a baseline for the eventual tables and figures that will go into the paper. At the same time it will serve as a record of all supplementary analyses performed.

```{r}
#try a model
model1 <- glm(below_exp_grade~race+as.factor(year), data=acs, family=binomial)
model2 <- update(model1, .~.+degree_mother+degree_father)
model3 <- update(model2, .~.+sqrt(family_income/10000)+own_home+parents_married)
model4 <- update(model3, .~.+foreign_born_mother+foreign_born_father)

coef_map <- list("(Intercept)"="Intercept",
                 "raceBlack"  ="Black",
                 "raceAIAN"   ="AIAN",
                 "raceAPI"    ="API",
                 "raceLatino" ="Latino",
                 "raceAIAN/API"="AIAN/API",
                 "raceAIAN/Black"="AIAN/Black",
                 "raceAIAN/White"="AIAN/White",
                 "raceAPI/Black"="API/Black",
                 "raceAPI/Latino"="API/Latino",
                 "raceAPI/White"="API/White",
                 "raceBlack/Latino"="Black/Latino",
                 "raceBlack/White"="Black/White",
                 "raceLatino/White"="Latino/White",
                 "degree_motherHS diploma"="HS Diploma, Mother",
                 "degree_motherAA degree"="AA Degree, Mother",
                 "degree_motherBA degree"="BA degree, Mother",
                 "degree_motherGrad degree"="Grad degree, Mother",
                 "degree_fatherHS diploma"="HS Diploma, Father",
                 "degree_fatherAA degree"="AA Degree, Father",
                 "degree_fatherBA degree"="BA degree, Father",
                 "degree_fatherGrad degree"="Grad degree, Father",
                 "sqrt(family_income/10000)"="family income (sqrt)",
                 "own_homeTRUE"="Family owns home",
                 "parents_marriedTRUE"="Parents married",
                 "foreign_born_motherTRUE"="Mother foreign-born",
                 "foreign_born_motherTRUE"="Father foreign-born")

knitreg(list(model1, model2, model3, model4))
```

### 2021-03-24 analysis/analysis.Rmd
removed out of data coef_map

@@
 
-coef_map <- list("(Intercept)"="Intercept",
-                 "raceBlack"  ="Black",
-                 "raceAIAN"   ="AIAN",
-                 "raceAPI"    ="API",
-                 "raceLatino" ="Latino",
-                 "raceAIAN/API"="AIAN/API",
-                 "raceAIAN/Black"="AIAN/Black",
-                 "raceAIAN/White"="AIAN/White",
-                 "raceAPI/Black"="API/Black",
-                 "raceAPI/Latino"="API/Latino",
-                 "raceAPI/White"="API/White",
-                 "raceBlack/Latino"="Black/Latino",
-                 "raceBlack/White"="Black/White",
-                 "raceLatino/White"="Latino/White",
-                 "degree_motherHS diploma"="HS Diploma, Mother",
-                 "degree_motherAA degree"="AA Degree, Mother",
-                 "degree_motherBA degree"="BA degree, Mother",
-                 "degree_motherGrad degree"="Grad degree, Mother",
-                 "degree_fatherHS diploma"="HS Diploma, Father",
-                 "degree_fatherAA degree"="AA Degree, Father",
-                 "degree_fatherBA degree"="BA degree, Father",
-                 "degree_fatherGrad degree"="Grad degree, Father",
-                 "sqrt(family_income/10000)"="family income (sqrt)",
-                 "own_homeTRUE"="Family owns home",
-                 "parents_marriedTRUE"="Parents married",
-                 "foreign_born_motherTRUE"="Mother foreign-born",
-                 "foreign_born_motherTRUE"="Father foreign-born")
-
 knitreg(list(model1, model2, model3, model4))

### 2021-03-24 analysis/organize_data.R
changed organization of racial groups

@@ -3,8 +3,9 @@
 #' author: ""
 #' ---
 
-# This script will read in raw data from the input directory, clean it up to produce 
-# the analytical dataset, and then write the analytical data to the output directory. 
+# This script will read in raw data from the input directory, clean it up to
+# produce the analytical dataset, and then write the analytical data to the
+# output directory.
 
 
 # Load libraries and functions --------------------------------------------
@@ -16,17 +17,15 @@ source(here("analysis","useful_functions.R"))
 options(max.print=999999)
 
 code_race <- function(race, hisp) {
-  #TODO: split out pacific islanders
-  #TODO: handle multiracial groups that are single big race (e.g Chinese/Japanese)
   race <- case_when(
     is.na(race) | is.na(hisp) ~ NA_character_,
     hisp != 0 ~ "Latino",
     race == 100 ~ "White",
     race == 200 ~ "Black",
-    race >= 300 & race < 400 ~ "AIAN",
-    race >= 400 & race < 700 ~ "API",
-#    race == 700 ~ "Other",
-#    race > 700 ~ "Multiracial"
+    (race >= 300 & race < 400) | (race>=680 & race<700)  ~ "Indigenous",
+    race >= 400 & race < 680 ~ "Asian",
+    race > 700 ~ "Multiracial",
+    TRUE ~ NA_character_
   )
   return(race)
 }
@@ -74,9 +73,10 @@ acs <- read_fwf(here("analysis","input","usa_00110.dat.gz"),
 #for now just use the level+5 as above grade, but this probably needs refinement
 #to address lots of issues.
 
-#I worry a little bit about specifying grade retention for the older grades where
-#grade retention is only possible at ages when kids are often out of the household
-#and therefore not eligible for the sample. Might create some weird biases.
+#I worry a little bit about specifying grade retention for the older grades
+#where grade retention is only possible at ages when kids are often out of the
+#household and therefore not eligible for the sample. Might create some weird
+#biases.
 
 acs <- acs %>% 
   mutate(race_mother=code_race(raced_mom, hispand_mom),
@@ -115,11 +115,26 @@ acs <- acs %>%
 
 #code the contrast matrix for race to adjust for fractions
 acs$race <- factor(acs$race,
-                   levels=c("White","Black","AIAN","API","Latino",
-                            "AIAN/API","AIAN/Black","AIAN/Latino","AIAN/White",
-                            "API/Black","API/Latino","API/White",
-                            "Black/Latino","Black/White",
[…]

### 2021-03-24 analysis/analysis.Rmd
caterpillar plot of estimates from model 4

@@
 ```
+
+```{r}
+coef_table <- tidy(model4) %>%
+  filter(str_detect(term, "race") & str_detect(term, "/") & !str_detect(term, "Multiracial")) %>%
+  mutate(term=str_remove(term, "race"))
+
+ggplot(coef_table, aes(x=reorder(term, estimate, mean), y=estimate, 
+                       ymin=estimate-1.96*std.error, ymax=estimate+1.96*std.error))+
+  geom_hline(yintercept = 0, linetype=2)+
+  geom_pointrange()+
+  coord_flip()+
+  theme_bw()
+```

### 2021-03-24 analysis/check_packages.R

@@ -5,7 +5,7 @@
 #http://www.vikram-baliga.com/blog/2015/7/19/a-hassle-free-way-to-verify-that-r-packages-are-installed-and-loaded
 
 #add new packages to the chain here
-packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr")
+packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr","broom")
 
 package.check <- lapply(packages, FUN = function(x) {
   if (!require(x, character.only = TRUE)) {

### 2021-03-26 analysis/analysis.Rmd
added lpm models; function for generating difference plots; summary of variables by race

--- 2021-03-24 analysis/analysis.Rmd
@@
 ```
+
+```{r}
+#try a LPM
+model1_lpm <- lm(below_exp_grade~race+as.factor(year), data=acs)
+model2_lpn <- update(model1, .~.+degree_mother+degree_father)
+model3_lpm <- update(model2, .~.+sqrt(family_income/10000)+own_home+parents_married)
+model4_lpm <- update(model3, .~.+foreign_born_mother+foreign_born_father)
+
+knitreg(list(model1_lpm, model2_lpm, model3_lpm, model4_lpm), digits=3)
+```
+
+```{r}
+plot_mr_effects <- function(model) {
+  coef_table <- tidy(model) %>%
+    filter(str_detect(term, "race") & str_detect(term, "/") & !str_detect(term, "Multiracial")) %>%
+    mutate(term=str_remove(term, "race"))
+
+  #add in the placements for each single race group
+  single_race <- tidy(model) %>%
+    filter(str_detect(term, "race") & !str_detect(term, "/") & !str_detect(term, "Multiracial")) %>%
+    select(term, estimate) %>%
+    mutate(term=str_remove(term, "race"))
+  x <- c(0, single_race$estimate)
+  names(x) <- c("White",single_race$term)
+  coef_table$lower <- NA
+  coef_table$higher <- NA
+  for(i in 1:nrow(coef_table)) {
+    groups <- str_split(coef_table$term[i], "/")[[1]]
+    temp <- diff(sort(x[groups]))/2
+    coef_table$lower[i] <- -temp
+    coef_table$higher[i] <- temp
+  }
+
+  # ggplot(coef_table, aes(x=reorder(term, estimate, mean), y=estimate, 
+  #                        ymin=estimate-1.96*std.error, ymax=estimate+1.96*std.error))+
+  #   geom_hline(yintercept = 0, linetype=2)+
+  #   geom_pointrange()+
+  #   geom_point(aes(x=reorder(term, estimate, mean), y=higher, color="red"))+
+  #   geom_point(aes(x=reorder(term, estimate, mean), y=lower, color="red"))+
+  #   coord_flip()+
+  #   theme_bw()
+
+  ggplot(coef_table, aes(x=reorder(term, higher, mean), y=estimate, 
+                         ymin=estimate-1.96*std.error, ymax=estimate+1.96*std.error))+
+    geom_hline(yintercept = 0, linetype=2)+
+    geom_pointrange()+
+    geom_point(aes(x=reorder(term, higher, mean), y=higher, color="red"))+
+    geom_point(aes(x=reorder(term, higher, mean), y=lower, color="red"))+
+    coord_flip()+
+    theme_bw()
+}
+```
+
+```{r}
+plot_mr_effects(model1)
+```
+
+```{r}
+plot_mr_effects(model4)
+```
+
+```{r}
+resources_race <- acs %>% 
+  group_by(race) %>%
+  summarize(mean_age=mean(age),
+            prop_foreign_born=mean(foreign_born),
+            mean_fam_inc=mean(family_income),
+            prop_own_home=mean(own_home),
+            prop_parents_married=mean(parents_married),
+            mean_age_parents=mean((age_birth_mother+age_birth_father)/2),
+            prop_parent_foreign_born=mean(foreign_born_father | foreign_born_mother)) %>%
+  ungroup() %>%
+  filter(!str_detect(race,"Multiracial"))
+
+ggplot(resources_race, aes(x=reorder(race, mean_fam_inc, mean), y=mean_fam_inc))+
+  geom_col()+
+  coord_flip()
+```

### 2021-03-26 analysis/analysis.Rmd
use ordered factor for parent's highest degree

@@
 #try a model
-model1 <- glm(below_exp_grade~race+as.factor(year), data=acs, family=binomial)
+model1 <- glm(below_exp_grade~race+as.factor(year)+foreign_born, data=acs, family=binomial)
 model2 <- update(model1, .~.+degree_mother+degree_father)
@@
   group_by(race) %>%
-  summarize(mean_age=mean(age),
+  summarize(prop_held_back=mean(below_exp_grade),
+            mean_age=mean(age),
             prop_foreign_born=mean(foreign_born),
@@
             prop_own_home=mean(own_home),
+            prop_hs_diploma=mean(degree_father>="HS diploma" | 
+                                   degree_mother>="HS diploma"),
+            prop_ba_degree=mean(degree_father>="BA degree" | 
+                                  degree_mother>="BA degree"),
             prop_parents_married=mean(parents_married),
             mean_age_parents=mean((age_birth_mother+age_birth_father)/2),
-            prop_parent_foreign_born=mean(foreign_born_father | foreign_born_mother)) %>%
+            prop_parent_foreign_born=mean(foreign_born_father | 
+                                            foreign_born_mother)) %>%
   ungroup() %>%

### 2021-03-26 analysis/organize_data.R

@@ -41,8 +41,24 @@ code_degree <- function(educd) {
   )
   degree <- factor(degree,
                  levels=c("Less than HS","HS diploma","AA degree","BA degree",
-                          "Grad degree"))
+                          "Grad degree"),
+                 ordered=TRUE)
+  degree <- ordered_factor(degree)
   return(degree)
+}
+
+#change contrasts for ordered factors to stairstep style
+ordered_factor <- function(fact_var) {
+  ord_fact <- factor(fact_var, ordered=TRUE)
+  categories <- levels(fact_var)
+  n_cat <- length(categories)
+  cont <- matrix(0, n_cat, n_cat-1)
+  cont[col(cont)<row(cont)] <- 1
+  rownames(cont) <- categories
+  colnames(cont) <- paste(categories[2:n_cat], categories[1:(n_cat-1)],
+                          sep=" vs. ")
+  contrasts(ord_fact) <- cont
+  return(ord_fact)
 }
 
 # Read the raw data -------------------------------------------------------

### 2021-10-11 analysis/analysis.Rmd
Maintain 80 character limit in analysis.Rmd
Add plot showing difference from midpoint expectation

@@
 #try a model
-model1 <- glm(below_exp_grade~race+as.factor(year)+foreign_born, data=acs, family=binomial)
+model1 <- glm(below_exp_grade~race+as.factor(year)+foreign_born, data=acs, 
+              family=binomial)
 model2 <- update(model1, .~.+degree_mother+degree_father)
@@
 ```
+
+```{r}
+#make a plot showing difference from middle expectation by model
+coef_table <- bind_rows(tidy(model1) %>%
+                          filter(str_detect(term, "race") & 
+                                   str_detect(term, "/") & 
+                                   !str_detect(term, "Multiracial")) %>%
+                          mutate(model="model1"),
+                        tidy(model4) %>%
+                          filter(str_detect(term, "race") & 
+                                   str_detect(term, "/") & 
+                                   !str_detect(term, "Multiracial")) %>%
+                          mutate(model="model4"))
+ggplot(coef_table, aes(x=reorder(term, estimate, mean), y=estimate, color=model, group=model))+
+  geom_point(position="dodge")+
+  coord_flip()
+```

### 2021-11-02 analysis/analysis.Rmd
Add summary graphs and a better plot for comparing effects
This is all WIP, but I have made some additions to the analysis. In particular, I think I can plot a scatterplot comparing the actual effect to that expected by the midpoint to illustrate the findings pretty well.

--- 2021-03-26 analysis/analysis.Rmd
@@
 
-# Introduction
-<a href="#top">Back to top</a>
+# Grade Retention Profile
 
-Use this R Markdown to perform the main analysis for the project.
-I use this basically as a lab notebook.
-It contains the main analysis and a variety of sensitivity analysis.
-The code in this documents serves as a baseline for the eventual tables and figures that will go into the paper.
-At the same time it will serve as a record of all supplementary analyses performed.
+OK, first lets just confirm that the grade retention variable is working as expected on our sample. I start by looking at the number of kids by age and current grade:
+
+```{r}
+table(acs$age, acs$current_grade)
+```
+
+Ok, this looks more or less like what I would expect.
+We observe two modal ages for each grade.
+Some of the more extreme age values for older grades seem a little strange given that these are only supposed to be currently enrolled students.
+20-year old sixth graders?
+I mean there are relatively few of these cases but still.
+Its also interesting how this really starts expanding past 5th - did the Census Bureau "correct" values below 6th or something?
+
+Now lets calculate the percentage of cases TRUE on the dependent variable across this same table, just to confirm its all working correctly.
+
+```{r}
+tapply(acs$below_exp_grade,acs[,c("age","current_grade")], mean)
+```
+
+Ok, that looks good. NAs indicate cells with no cases. The 1 values start with 7 year old Kindergarteners and then move down in a perfect diagonal from there. So this looks good.
+
+What do the grade retention probabilities look like by grade:
+
+```{r}
+round(tapply(acs$below_exp_grade, acs$current_grade, mean), 3)*100
+```
+
+I guess those numbers look pretty reasonable.
+Its hard to benchmark it off anything.
+Warren et.
+al.
+estimate annual retention rates by grade but that is different than what I am measuring here because this is cumulative (but can also reflect students being caught up).
+In general, these must be underestimates because they do not capture kids who were held back but are still within the modal age.
+
+What about by year?
+
+```{r}
+round(tapply(acs$below_exp_grade, acs[,c("current_grade","year")], mean), 3)*100
+```
+
+Some evidence of declining grade retention prevalence at higher grades.
+
+# Lets look at the multiracial sample
+
+How big is it?
+
+```{r}
+acs %>%
+  filter(str_detect(race, "/") & !str_detect(race, "Multiracial")) %>%
+  ggplot(aes(x=reorder(race, race, length))) +
+  geom_bar()+
+  coord_flip()+
+  labs(x=NULL)+
+  theme_bw()
+```
+
+```{r}
+resources_race <- acs %>% 
+  group_by(race) %>%
+  summarize(prop_held_back=mean(below_exp_grade),
+            mean_age=mean(age),
+            prop_foreign_born=mean(foreign_born),
+            mean_fam_inc=mean(family_income),
+            prop_own_home=mean(own_home),
+            prop_hs_diploma=mean(degree_father>="HS diploma" | 
+                                   degree_mother>="HS diploma"),
+            prop_ba_degree=mean(degree_father>="BA degree" | 
+                                  degree_mother>="BA degree"),
+            prop_parents_married=mean(parents_married),
+            mean_age_parents=mean((age_birth_mother+age_birth_father)/2),
+            prop_parent_foreign_born=mean(foreign_born_father | 
+                                            foreign_born_mother)) %>%
+  ungroup() %>%
+  filter(!str_detect(race,"Multiracial")) %>%
+  mutate(multiracial=str_detect(race, "/"))
+```
+
+```{r}
+ggplot(resources_race, aes(x=reorder(race, mean_fam_inc, mean), y=mean_fam_inc,
+                           fill=multiracial))+
+  geom_col()+
+  coord_flip()+
+  labs(x=NULL, y="mean family income")
+```
+
+```{r}
+ggplot(resources_race, aes(x=reorder(race, prop_own_home, mean), y=prop_own_home,
+                           fill=multiracial))+
+  geom_col()+
+  coord_flip()+
+  labs(x=NULL, y="Proportion who own their own home")
+```
+
+```{r}
+ggplot(resources_race, aes(x=reorder(race, prop_hs_diploma, mean), y=prop_hs_diploma,
+                           fill=multiracial))+
+  geom_col()+
+  coord_flip()+
+  labs(x=NULL, y="Proportion where one parent has HS diploma")
+```
+
+```{r}
+ggplot(resources_race, aes(x=reorder(race, prop_ba_degree, mean), y=prop_ba_degree,
+                           fill=multiracial))+
+  geom_col()+
+  coord_flip()+
+  labs(x=NULL, y="Proportion where one parent has college degree")
+```
+
+## Raw differences in grade retention
+
+```{r}
+ggplot(resources_race, aes(x=reorder(race, prop_held_back, mean), y=prop_held_back,
+                           fill=multiracial))+
+  geom_col()+
+  coord_flip()
+```
+
+## Build some models
+
+Here are the model structure that I will use:
+
+0. Baseline model with dummies for current grade and year
+1. Add race.
+2. Add kids and parents foreign-born status.
+2. Add mother and father highest degree
+3. Add family income (square rooted), own your own home, and whether parents are married.
+
+All models are estimated as logit models. Race is coded using the average between the two monoracial groups as the reference category for all multiracial responses.
 
@@
 #try a model
-model1 <- glm(below_exp_grade~race+as.factor(year)+foreign_born, data=acs, family=binomial)
-model2 <- update(model1, .~.+degree_mother+degree_father)
-model3 <- update(model2, .~.+sqrt(family_income/10000)+own_home+parents_married)
-model4 <- update(model3, .~.+foreign_born_mother+foreign_born_father)
+model0 <- glm(below_exp_grade~as.factor(year)+current_grade, 
+              data=acs, family=binomial)
+model1 <- update(model0, .~.+race)
+model2 <- update(model1, .~.+foreign_born+foreign_born_mother+foreign_born_father)
+model3 <- update(model2, .~.+degree_mother+degree_father)
+model4 <- update(model3, .~.+sqrt(family_income/10000)+own_home+parents_married)
 
@@
 ```{r}
-#try a LPM
-model1_lpm <- lm(below_exp_grade~race+as.factor(year), data=acs)
-model2_lpn <- update(model1, .~.+degree_mother+degree_father)
-model3_lpm <- update(model2, .~.+sqrt(family_income/10000)+own_home+parents_married)
-model4_lpm <- update(model3, .~.+foreign_born_mother+foreign_born_father)
+#run separate models by grade groups
+model4a <- update(model4, data=subset(acs, as.numeric(current_grade)<=7))
+model4b <- update(model4, data=subset(acs, as.numeric(current_grade)>7 &
+                                        as.numeric(current_grade)<10))
+model4c <- update(model4, data=subset(acs, as.numeric(current_grade)>=10))
 
-knitreg(list(model1_lpm, model2_lpm, model3_lpm, model4_lpm), digits=3)
+knitreg(list(model4a, model4b, model4c))
 ```
@@
 
-  # ggplot(coef_table, aes(x=reorder(term, estimate, mean), y=estimate, 
-  #                        ymin=estimate-1.96*std.error, ymax=estimate+1.96*std.error))+
-  #   geom_hline(yintercept = 0, linetype=2)+
-  #   geom_pointrange()+
-  #   geom_point(aes(x=reorder(term, estimate, mean), y=higher, color="red"))+
-  #   geom_point(aes(x=reorder(term, estimate, mean), y=lower, color="red"))+
-  #   coord_flip()+
-  #   theme_bw()
-
   ggplot(coef_table, aes(x=reorder(term, higher, mean), y=estimate, 
@@
 ```{r}
-plot_mr_effects(model1)
+plot_mr_effects(model2)
 ```
@@
 ```
+
+```{r}
+#make a plot showing difference from middle expectation by model
+coef_table <- bind_rows(tidy(model1) %>%
+                          filter(str_detect(term, "race") & 
+                                   str_detect(term, "/") & 
+                                   !str_detect(term, "Multiracial")) %>%
+                          mutate(model="model1"),
+                        tidy(model2) %>%
+                          filter(str_detect(term, "race") & 
+                                   str_detect(term, "/") & 
+                                   !str_detect(term, "Multiracial")) %>%
+                          mutate(model="model2"),
+                        tidy(model4) %>%
+                          filter(str_detect(term, "race") & 
+                                   str_detect(term, "/") & 
+                                   !str_detect(term, "Multiracial")) %>%
+                          mutate(model="model4"))
+ggplot(coef_table, aes(x=reorder(term, estimate, min), y=estimate, color=model, group=model))+
+  geom_point(position="dodge")+
+  coord_flip()
+```
+
+Try a plot that shows the expectation based on being exactly in the middle on the x-axis and the actual value on the y-axis.
+
+```{r}
+plot_effect_scatter <- function(model) {
+  coef_table <- tidy(model) %>%
+    filter(str_detect(term, "race") & !str_detect(term, "Multiracial")) %>%
+    mutate(term=str_remove(term, "race")) %>%
+    select(term, estimate, std.error) %>%
+    bind_rows(tibble(term="White", estimate=0, std.error=NA)) %>%
+    mutate(multiracial=str_detect(term, "/"))
+
+  #calculate midway points
+  temp <- coef_table %>%
+    filter(!multiracial) %>%
+    select(term, estimate)
+  means <- apply(expand.grid(temp$estimate,temp$estimate), 1, mean)
+  coef_table <- expand.grid(temp$term,temp$term) %>%
+    mutate(term=paste(Var1, Var2, sep="/"),
+           estimate_midpoint=means) %>%
+    filter(Var1!=Var2) %>%
+    select(term, estimate_midpoint) %>%
+    right_join(coef_table) %>%
+    mutate(estimate_midpoint=ifelse(is.na(estimate_midpoint), estimate,
+                                    estimate_midpoint),
+           estimate=ifelse(estimate==estimate_midpoint, 0, estimate))
+
+  ggplot(coef_table, aes(x=estimate_midpoint, y=estimate_midpoint+estimate,
+                         color=multiracial,
+                         ymin=estimate_midpoint+estimate-1.96*std.error,
+                         ymax=estimate_midpoint+estimate+1.96*std.error))+
+    geom_abline(intercept = 0, slope=1, linetype=2, color="grey70")+
+   # geom_hline(yintercept = 0, linetype=2, color="grey40")+
+  #  geom_vline(xintercept = 0, linetype=2, color="grey40")+
+    geom_linerange(alpha=0.5)+
+    geom_point()+
+    geom_text_repel(aes(label=term), min.segment.length = 0)+
+    theme_bw()+
+    scale_color_viridis_d(end=0.6)+
+    labs(x="expected log odds ratio based on midway point",
+         y="actual log odds ratio")
+}
+```
+
+## Logit panic sensitivity results
+
+```{r, eval=FALSE}
+#try a LPM
+model1_lpm <- lm(below_exp_grade~race+as.factor(year), data=acs)
+model2_lpn <- update(model1, .~.+degree_mother+degree_father)
+model3_lpm <- update(model2, .~.+sqrt(family_income/10000)+own_home+parents_married)
+model4_lpm <- update(model3, .~.+foreign_born_mother+foreign_born_father)
+
+knitreg(list(model1_lpm, model2_lpm, model3_lpm, model4_lpm), digits=3)
+```

### 2021-11-02 analysis/check_packages.R
Add ggrepel package

@@ -5,7 +5,8 @@
 #http://www.vikram-baliga.com/blog/2015/7/19/a-hassle-free-way-to-verify-that-r-packages-are-installed-and-loaded
 
 #add new packages to the chain here
-packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr","broom")
+packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr","broom",
+             "ggrepel")
 
 package.check <- lapply(packages, FUN = function(x) {
   if (!require(x, character.only = TRUE)) {

### 2021-11-02 analysis/organize_data.R
Remove relic commented out code

@@ -226,16 +226,3 @@ acs <- acs %>%
   filter(!is.na(race) & !is.na(current_grade) & current_grade!="Pre-K")
 
 save(acs, file=here("analysis","output","acs.RData"))
-
-
-# Some models -------------------------------------------------------------
-
-#try a model
-#model <- glm(below_exp_grade~race, 
-#             data=acs, family=binomial)
-
-#model2 <- glm(below_exp_grade~relevel(as.factor(race), "White"), 
-#              data=acs, family=binomial)
-
-  
-  

### 2021-11-03 analysis/organize_data.R
Change code to read in new ipums extract 112

@@ -63,22 +63,24 @@ ordered_factor <- function(fact_var) {
 
 # Read the raw data -------------------------------------------------------
 
-acs <- read_fwf(here("analysis","input","usa_00110.dat.gz"), 
+acs <- read_fwf(here("analysis","input","usa_00112.dat.gz"), 
                 col_positions = 
                   fwf_positions(
-                    start = c(1,11,32,42,55,56,61,74,77,88,98, 99,104,108,111,
-                              126,128,135,138,147,148,151,154,163,166,175,178,
-                              187,190),
-                    end   = c(4,18,41,54,55,60,72,74,83,97,98,101,106,110,113,
-                              127,134,137,140,147,148,153,156,165,168,177,180,
-                              189,192),
-                    col_names = c("year","serial","hhwt","cluster","metro",
-                                  "puma","strata","ownershp","hhincome","perwt",
+                    start = c(1,11,32,42,55,57,58,71,74,85,95,97,101,105,106,
+                              111,115,118,133,135,142,146,158,161,170,171,174,
+                              177,186,189,198,201,210,213),
+                    end   = c(4,18,41,54,56,57,69,71,80,94,96,98,104,105,108,
+                              113,117,120,134,141,145,149,158,163,170,171,176,
+                              179,188,191,200,203,212,215),
+                    col_names = c("year","serial","hhwt","cluster","state",
+                                  "metro","strata","ownership","hhincome",
+                                  "perwt","momrule","poprule","related",
                                   "sex","age","raced","hispand","bpl",
-                                  "gradeattd","ftotinc","age_mom","age_pop",
-                                  "marst_mom","marst_pop","raced_mom",
-                                  "raced_pop","hispand_mom","hispand_pop",
-                                  "bpl_mom","bpl_pop","educd_mom","educd_pop")),
+                                  "gradattd","ftotinc","related_mom",
+                                  "related_pop","age_mom","age_pop","marst_mom",
+                                  "marst_pop","raced_mom","raced_pop",
+                                  "hispand_mom","hispand_pop","bpl_mom",
+                                  "bpl_dad","educd_mom","educd_pop")),
                 col_types = cols(.default = "i",
                                  cluster = col_double()),
                 progress = FALSE)
@@ -95,6 +97,8 @@ acs <- read_fwf(here("analysis","input","usa_00110.dat.gz"),
 #biases.
 
 acs <- acs %>% 
[…]

### 2021-11-03 analysis/organize_data.R
Code new ACS data
Fixed a few screwed up names in my read_fwf commands and added variables for metro area and sex. I also removed pre-K individuals and filtered to only include cases with both parents identified.

@@ -73,32 +73,24 @@ acs <- read_fwf(here("analysis","input","usa_00112.dat.gz"),
                               113,117,120,134,141,145,149,158,163,170,171,176,
                               179,188,191,200,203,212,215),
                     col_names = c("year","serial","hhwt","cluster","state",
-                                  "metro","strata","ownership","hhincome",
+                                  "metro","strata","ownershp","hhincome",
                                   "perwt","momrule","poprule","related",
                                   "sex","age","raced","hispand","bpl",
-                                  "gradattd","ftotinc","related_mom",
+                                  "gradeattd","ftotinc","related_mom",
                                   "related_pop","age_mom","age_pop","marst_mom",
                                   "marst_pop","raced_mom","raced_pop",
                                   "hispand_mom","hispand_pop","bpl_mom",
-                                  "bpl_dad","educd_mom","educd_pop")),
+                                  "bpl_pop","educd_mom","educd_pop")),
                 col_types = cols(.default = "i",
                                  cluster = col_double()),
                 progress = FALSE)
 
 # Recode variables ------------------------------------------------------------
 
-#Code grade retention
-#for now just use the level+5 as above grade, but this probably needs refinement
-#to address lots of issues.
-
-#I worry a little bit about specifying grade retention for the older grades
-#where grade retention is only possible at ages when kids are often out of the
-#household and therefore not eligible for the sample. Might create some weird
-#biases.
-
 acs <- acs %>% 
-  #remove cases without both parents
-  filter(momrule!=0 & poprule!=0) %>%
+  #remove cases without both parents and pre-K
+  filter(momrule!=0 & poprule!=0 & gradeattd>10) %>%
+  #recode variables
   mutate(race_mother=code_race(raced_mom, hispand_mom),
          race_father=code_race(raced_pop, hispand_pop),
          age_birth_mother=age_mom-age,
@@ -112,7 +104,6 @@ acs <- acs %>%
                      pmap_chr(list(race_mother, race_father),
                               ~paste(sort(c(...)), collapse = "/"))),
          current_grade=factor(case_when(
[…]

### 2021-11-03 analysis/organize_data.R
Made several changes to organize data
* remove age-discrepant parents (outside 15-44 at birth for moms and outside 15-60 for dads)
* move functions to useful_functions.R
* Add new variables to final saved dataset

@@ -16,51 +16,6 @@ source(here("analysis","check_packages.R"))
 source(here("analysis","useful_functions.R"))
 options(max.print=999999)
 
-code_race <- function(race, hisp) {
-  race <- case_when(
-    is.na(race) | is.na(hisp) ~ NA_character_,
-    hisp != 0 ~ "Latino",
-    race == 100 ~ "White",
-    race == 200 ~ "Black",
-    (race >= 300 & race < 400) | (race>=680 & race<700)  ~ "Indigenous",
-    race >= 400 & race < 680 ~ "Asian",
-    race > 700 ~ "Multiracial",
-    TRUE ~ NA_character_
-  )
-  return(race)
-}
-
-code_degree <- function(educd) {
-  degree <- case_when(
-    is.na(educd) | educd==1 ~  NA_character_,
-    educd<62 ~ "Less than HS",
-    educd<81 ~ "HS diploma",
-    educd<101 ~ "AA degree",
-    educd==101 ~ "BA degree",
-    TRUE ~ "Grad degree"
-  )
-  degree <- factor(degree,
-                 levels=c("Less than HS","HS diploma","AA degree","BA degree",
-                          "Grad degree"),
-                 ordered=TRUE)
-  degree <- ordered_factor(degree)
-  return(degree)
-}
-
-#change contrasts for ordered factors to stairstep style
-ordered_factor <- function(fact_var) {
-  ord_fact <- factor(fact_var, ordered=TRUE)
-  categories <- levels(fact_var)
-  n_cat <- length(categories)
-  cont <- matrix(0, n_cat, n_cat-1)
-  cont[col(cont)<row(cont)] <- 1
-  rownames(cont) <- categories
-  colnames(cont) <- paste(categories[2:n_cat], categories[1:(n_cat-1)],
-                          sep=" vs. ")
-  contrasts(ord_fact) <- cont
-  return(ord_fact)
-}
-
 # Read the raw data -------------------------------------------------------
 
 acs <- read_fwf(here("analysis","input","usa_00112.dat.gz"), 
@@ -70,7 +25,7 @@ acs <- read_fwf(here("analysis","input","usa_00112.dat.gz"),
                               111,115,118,133,135,142,146,158,161,170,171,174,
                               177,186,189,198,201,210,213),
                     end   = c(4,18,41,54,56,57,69,71,80,94,96,98,104,105,108,
-                              113,117,120,134,141,145,149,158,163,170,171,176,
[…]

### 2021-11-03 analysis/useful_functions.R

@@ -1,3 +1,62 @@
 ## useful_functions.R
 
 #Put any functions that might be used across multiple scripts here so that they can be properly sourced in.
+
+code_race <- function(race, hisp) {
+  race <- case_when(
+    is.na(race) | is.na(hisp) ~ NA_character_,
+    hisp != 0 ~ "Latino",
+    race == 100 ~ "White",
+    race == 200 ~ "Black",
+    (race >= 300 & race < 400) | (race>=680 & race<700)  ~ "Indigenous",
+    race >= 400 & race < 680 ~ "Asian",
+    race > 700 ~ "Multiracial",
+    TRUE ~ NA_character_
+  )
+  return(race)
+}
+
+code_degree <- function(educd) {
+  degree <- case_when(
+    is.na(educd) | educd==1 ~  NA_character_,
+    educd<62 ~ "Less than HS",
+    educd<81 ~ "HS diploma",
+    educd<101 ~ "AA degree",
+    educd==101 ~ "BA degree",
+    TRUE ~ "Grad degree"
+  )
+  degree <- factor(degree,
+                   levels=c("Less than HS","HS diploma","AA degree","BA degree",
+                            "Grad degree"),
+                   ordered=TRUE)
+  degree <- ordered_factor(degree)
+  return(degree)
+}
+
+#change contrasts for ordered factors to stairstep style
+ordered_factor <- function(fact_var) {
+  ord_fact <- factor(fact_var, ordered=TRUE)
+  categories <- levels(fact_var)
+  n_cat <- length(categories)
+  cont <- matrix(0, n_cat, n_cat-1)
+  cont[col(cont)<row(cont)] <- 1
+  rownames(cont) <- categories
+  colnames(cont) <- paste(categories[2:n_cat], categories[1:(n_cat-1)],
+                          sep=" vs. ")
+  contrasts(ord_fact) <- cont
+  return(ord_fact)
+}
+
+## Determine if race coding is consistent with reported race
+
+#first create vectors of values from raced that include each of the major
+#race groups (with the exception of Latinos who are defined differently)
+# other should be considered consistent with all cases 
+
+#include_white <- c(100,700,801:827,901:922,950:964,990)
+#include_black <- c(200,700,801,830:845,901:904,917,930:935,950:955,970:973,
+#                   980:983, 985:986,990)
+#include_indig <- c(300:399,680:699,700,802,820:825,840:842,850:856,860,
+#                   865:868,890:893,901,903,905:907,
+#                   890)
+#include_asian <- c()

### 2021-11-03 analysis/organize_data.R
Add first attempt at consistent race function
This function compares child's race report to parent's race report in an attempt to remove cases where they are completely inconsistent

@@ -185,6 +185,25 @@ acs <- acs %>%
 # groups but not something completely unrelated. That is a pretty complicated
 # function though given the number of categories available.
 
+mean(acs$raced==acs$raced_mom & acs$hispand==acs$hispand_mom)
+mean(acs$raced==acs$raced_pop & acs$hispand==acs$hispand_pop)
+
+mean((acs$raced==acs$raced_mom & acs$hispand==acs$hispand_mom)  | 
+       (acs$raced==acs$raced_pop & acs$hispand==acs$hispand_pop))
+
+#About 95.2% of cases are exactly consistent with either mother or father
+
+# Based on this, I created a function that tries to identify consistent cases.
+# TODO: It allows multiracial kids to be considered ok in any multiracial category. I 
+# should probably refine that a bit in the future.
+
+mean(is_race_consistent(acs$raced, acs$hispand, acs$raced_mom, acs$hispand_mom,
+                        acs$raced_pop, acs$hispand_pop, acs$race))
+
+acs <- acs %>%
+  filter(is_race_consistent(raced, hispand, raced_mom, hispand_mom, raced_pop, 
+                            hispand_pop, race))
+
 # Trim to analytical data -------------------------------------------------
 
 acs <- acs %>%

### 2021-11-03 analysis/useful_functions.R

@@ -48,6 +48,39 @@ ordered_factor <- function(fact_var) {
 }
 
 ## Determine if race coding is consistent with reported race
+is_race_consistent <- function(race_child, hispan_child, race_mom, hispan_mom,
+                               race_dad, hispan_dad, race_constructed) {
+  ## CASE 1: child's race is exactly consistent with one of parent's races
+  ## This case should take care of about 95% of the data
+  TF <- (race_child==race_mom & hispan_child==hispan_mom)  | 
+    (race_child==race_dad & hispan_child==hispan_dad)
+
+  #if both parents report indigenous allow a couple of residual indigenous
+  #categories
+  TF <- TF | 
+    (race_constructed=="Indigenous" & (race_child==361 | race_child==362))
+  
+  #if both parents report Asian allow residual and some combined categories
+  TF <- TF |
+    (race_constructed=="Asian" & race_child %in% c(671:679))
+
+  #if both parents are Latino, allow residual category
+  TF <- TF |
+    (race_constructed=="Latino" & hispan_child==498)
+  
+  #if neither parent is Latino, but child is classified as Latino then reject
+  TF <- TF & 
+    (str_detect(race_constructed, "Latino") | 
+       (!str_detect(race_constructed, "Latino") & hispan_child==0))
+  
+  #if multiracial category, we will consider other or any multiracial category
+  #as ok
+  TF <- TF | 
+    (str_detect(race_constructed, "/") & race_child>=700)
+  
+  return(TF)
+}
+
 
 #first create vectors of values from raced that include each of the major
 #race groups (with the exception of Latinos who are defined differently)

### 2021-11-03 analysis/useful_functions.R
Allow capture of residual categories for multiracial respondents

@@ -55,18 +55,19 @@ is_race_consistent <- function(race_child, hispan_child, race_mom, hispan_mom,
   TF <- (race_child==race_mom & hispan_child==hispan_mom)  | 
     (race_child==race_dad & hispan_child==hispan_dad)
 
-  #if both parents report indigenous allow a couple of residual indigenous
-  #categories
+  #if at least one parent reports indigenous allow a couple of residual 
+  #indigenous categories
   TF <- TF | 
-    (race_constructed=="Indigenous" & (race_child==361 | race_child==362))
+    (str_detect(race_constructed, "Indigenous") & race_child %in% 361:362)
   
-  #if both parents report Asian allow residual and some combined categories
+  #if at least one parent reports Asian allow residual and some combined 
+  #categories
   TF <- TF |
-    (race_constructed=="Asian" & race_child %in% c(671:679))
+    (str_detect(race_constructed, "Asian") & race_child %in% 671:679)
 
-  #if both parents are Latino, allow residual category
+  #if at least one parent is Latino, allow residual category
   TF <- TF |
-    (race_constructed=="Latino" & hispan_child==498)
+    (str_detect(race_constructed, "Latino") & hispan_child==498)
   
   #if neither parent is Latino, but child is classified as Latino then reject
   TF <- TF & 

### 2021-11-03 analysis/organize_data.R
Add race consistency function
Function checks for whether the race reporting of child is broadly consistent with the race of parents. This is complex given the large number of mulitracial categories. I generally err on the side of liberal inclusion here. The goal is simply to eliminate cases where there is a very clear disrepancy.

@@ -201,8 +201,18 @@ mean(is_race_consistent(acs$raced, acs$hispand, acs$raced_mom, acs$hispand_mom,
                         acs$raced_pop, acs$hispand_pop, acs$race))
 
 acs <- acs %>%
-  filter(is_race_consistent(raced, hispand, raced_mom, hispand_mom, raced_pop, 
-                            hispand_pop, race))
+  mutate(race_consistent=is_race_consistent(raced, hispand, raced_mom, 
+                                            hispand_mom, raced_pop, 
+                                            hispand_pop, race))
+
+mean(acs$race_consistent)
+
+tapply(acs$race_consistent, acs$race, mean)
+
+#its a bit low for Latino/Asian which may I think reflect some issues for 
+#Filipinos, but in general pretty good.
+acs <- acs %>%
+  filter(race_consistent)
 
 # Trim to analytical data -------------------------------------------------
 

### 2021-11-03 analysis/useful_functions.R

@@ -50,34 +50,79 @@ ordered_factor <- function(fact_var) {
 ## Determine if race coding is consistent with reported race
 is_race_consistent <- function(race_child, hispan_child, race_mom, hispan_mom,
                                race_dad, hispan_dad, race_constructed) {
-  ## CASE 1: child's race is exactly consistent with one of parent's races
-  ## This case should take care of about 95% of the data
+  # start with the case wherechild's race is exactly consistent with one of 
+  # parent's races. This case should take care of about 95% of the data and 
+  # should take care of most single race respondents and biracial respondents
+  #  that identified with one parent
   TF <- (race_child==race_mom & hispan_child==hispan_mom)  | 
     (race_child==race_dad & hispan_child==hispan_dad)
 
+  #in cases where either parent is hispanic allow this to match even if
+  #race does not, since there may be discrepancy here in terms of reporting
+  #other
+  TF <- TF | 
+    (hispan_dad!=0 & hispan_child==hispan_dad) | 
+    (hispan_mom!=0 & hispan_child==hispan_mom)
+  
   #if at least one parent reports indigenous allow a couple of residual 
   #indigenous categories
   TF <- TF | 
-    (str_detect(race_constructed, "Indigenous") & race_child %in% 361:362)
+    (str_detect(race_constructed, "Indigenous") & 
+       race_child %in% c(361:362,855,942))
   
   #if at least one parent reports Asian allow residual and some combined 
   #categories
   TF <- TF |
-    (str_detect(race_constructed, "Asian") & race_child %in% 671:679)
+    (str_detect(race_constructed, "Asian") & race_child %in% c(671:679,887))
 
   #if at least one parent is Latino, allow residual category
   TF <- TF |
     (str_detect(race_constructed, "Latino") & hispan_child==498)
   
+  #if multiracial category consider "other" to be ok
+  TF <- TF | 
+    (str_detect(race_constructed, "/") & race_child==700)
+  
+  #now go through specific multiracial possibilities
+  #non-Latino ones are a pain but straightforward
+  #we allow other to operate as a wildcard
+  TF <- TF | 
+    (str_detect(race_constructed, "Black/White") & 
+       race_child %in% c(801,826,827,845,901:904,907,917,921:923,925,932,934,
+                         935,950:955,961:963,971:973,980:986,990))
+  TF <- TF | 
[…]

### 2021-11-03 analysis/useful_functions.R
Forgot some comments in prior commit

@@ -116,6 +116,9 @@ is_race_consistent <- function(race_child, hispan_child, race_mom, hispan_mom,
        race_child %in% c(850:854,856:868,880:886,890:893,905,910:925,930,
                          932:944,950,952:976,980:990))
   
+  #Part-Latino respondents should have already been picked up by other
+  #conditions.
+  
   #if neither parent is Latino, but child is classified as Latino then reject
   #this should have been dealt with by first two conditions, but may have been
   #re-introduced in later cases
@@ -123,19 +126,5 @@ is_race_consistent <- function(race_child, hispan_child, race_mom, hispan_mom,
     (str_detect(race_constructed, "Latino") | 
        (!str_detect(race_constructed, "Latino") & hispan_child==0))
 
-  
   return(TF)
 }
-
-
-#first create vectors of values from raced that include each of the major
-#race groups (with the exception of Latinos who are defined differently)
-# other should be considered consistent with all cases 
-
-#include_white <- c(100,700,801:827,901:922,950:964,990)
-#include_black <- c(200,700,801,830:845,901:904,917,930:935,950:955,970:973,
-#                   980:983, 985:986,990)
-#include_indig <- c(300:399,680:699,700,802,820:825,840:842,850:856,860,
-#                   865:868,890:893,901,903,905:907,
-#                   890)
-#include_asian <- c()

### 2021-11-04 analysis/analysis.Rmd
Add state and metro fixed effects to the models

@@
 #try a model
-model0 <- glm(below_exp_grade~as.factor(year)+current_grade, 
+model0 <- glm(below_exp_grade~as.factor(year)+current_grade+as.factor(state)+metro, 
               data=acs, family=binomial)
@@
 #run separate models by grade groups
-model4a <- update(model4, data=subset(acs, as.numeric(current_grade)<=7))
-model4b <- update(model4, data=subset(acs, as.numeric(current_grade)>7 &
-                                        as.numeric(current_grade)<10))
+model4a <- update(model4, data=subset(acs, as.numeric(current_grade)<=6))
+model4b <- update(model4, data=subset(acs, as.numeric(current_grade)>6 &
+                                        as.numeric(current_grade)<=9))
 model4c <- update(model4, data=subset(acs, as.numeric(current_grade)>=10))
@@
 plot_mr_effects(model4)
-```
-
-```{r}
-resources_race <- acs %>% 
-  group_by(race) %>%
-  summarize(prop_held_back=mean(below_exp_grade),
-            mean_age=mean(age),
-            prop_foreign_born=mean(foreign_born),
-            mean_fam_inc=mean(family_income),
-            prop_own_home=mean(own_home),
-            prop_hs_diploma=mean(degree_father>="HS diploma" | 
-                                   degree_mother>="HS diploma"),
-            prop_ba_degree=mean(degree_father>="BA degree" | 
-                                  degree_mother>="BA degree"),
-            prop_parents_married=mean(parents_married),
-            mean_age_parents=mean((age_birth_mother+age_birth_father)/2),
-            prop_parent_foreign_born=mean(foreign_born_father | 
-                                            foreign_born_mother)) %>%
-  ungroup() %>%
-  filter(!str_detect(race,"Multiracial"))
-
-ggplot(resources_race, aes(x=reorder(race, mean_fam_inc, mean), y=mean_fam_inc))+
-  geom_col()+
-  coord_flip()
-```
-
-```{r}
-#make a plot showing difference from middle expectation by model
-coef_table <- bind_rows(tidy(model1) %>%
-                          filter(str_detect(term, "race") & 
-                                   str_detect(term, "/") & 
-                                   !str_detect(term, "Multiracial")) %>%
-                          mutate(model="model1"),
-                        tidy(model2) %>%
-                          filter(str_detect(term, "race") & 
-                                   str_detect(term, "/") & 
-                                   !str_detect(term, "Multiracial")) %>%
-                          mutate(model="model2"),
-                        tidy(model4) %>%
-                          filter(str_detect(term, "race") & 
-                                   str_detect(term, "/") & 
-                                   !str_detect(term, "Multiracial")) %>%
-                          mutate(model="model4"))
-ggplot(coef_table, aes(x=reorder(term, estimate, min), y=estimate, color=model, group=model))+
-  geom_point(position="dodge")+
-  coord_flip()
 ```

### 2021-11-11 analysis/analysis.Rmd
Panelize scatterplot
The function for making the scatterplot now takes a named list of models and then panelizes the resulting figures. It works, but things are a bit crunched and busy so I am going to keep working on alternative.

@@
 ```{r}
-plot_effect_scatter <- function(model) {
-  coef_table <- tidy(model) %>%
+plot_effect_scatter <- function(models) {
+  coef_table <- models %>%
+    map(function(x) {
+      coefs <- tidy(x) %>%
         filter(str_detect(term, "race") & !str_detect(term, "Multiracial")) %>%
@@
 
-  #calculate midway points
-  temp <- coef_table %>%
+      #now get midpoint values
+      temp <- coefs %>%
         filter(!multiracial) %>%
@@
       means <- apply(expand.grid(temp$estimate,temp$estimate), 1, mean)
-  coef_table <- expand.grid(temp$term,temp$term) %>%
+      coefs <- expand.grid(temp$term,temp$term) %>%
         mutate(term=paste(Var1, Var2, sep="/"),
@@
         select(term, estimate_midpoint) %>%
-    right_join(coef_table) %>%
+        right_join(coefs) %>%
         mutate(estimate_midpoint=ifelse(is.na(estimate_midpoint), estimate,
@@
            estimate=ifelse(estimate==estimate_midpoint, 0, estimate))
+    }) %>%
+    bind_rows(.id="model") %>%
+    mutate(model=factor(model, levels=names(models)))
 
@@
     geom_abline(intercept = 0, slope=1, linetype=2, color="grey70")+
-   # geom_hline(yintercept = 0, linetype=2, color="grey40")+
-  #  geom_vline(xintercept = 0, linetype=2, color="grey40")+
     geom_linerange(alpha=0.5)+
@@
     geom_text_repel(aes(label=term), min.segment.length = 0)+
+    facet_wrap(~model)+
     theme_bw()+
-    scale_color_viridis_d(end=0.6)+
+    theme(legend.position = "none")+
+    scale_color_viridis_d(end=0.6, direction = -1)+
     labs(x="expected log odds ratio based on midway point",
@@
 }
+```
+
+```{r}
+plot_effect_scatter(list("controls for birthplace"=model2, "+controls for parent resources"=model4))
 ```

### 2021-11-11 analysis/check_packages.R

@@ -6,7 +6,7 @@
 
 #add new packages to the chain here
 packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr","broom",
-             "ggrepel")
+             "ggrepel","purrr")
 
 package.check <- lapply(packages, FUN = function(x) {
   if (!require(x, character.only = TRUE)) {

### 2021-11-11 analysis/analysis.Rmd
Panelize figure by group
To keep things a little cleaner,  I redesigned the panelized figure to panelize by each multracial group which allows readers to much more easily see where the multiracial group falls relative to the monoracial constituent groups. This new function only takes one model. I think trying to compare models in the same figure would be too difficult, so think about other ways of showing change across controls.

--- 2021-11-04 analysis/analysis.Rmd
@@
 plot_effect_scatter <- function(model) {
-  coef_table <- tidy(model) %>%
+
+  coefs <- tidy(model) %>%
     filter(str_detect(term, "race") & !str_detect(term, "Multiracial")) %>%
@@
 
-  #calculate midway points
-  temp <- coef_table %>%
-    filter(!multiracial) %>%
-    select(term, estimate)
-  means <- apply(expand.grid(temp$estimate,temp$estimate), 1, mean)
-  coef_table <- expand.grid(temp$term,temp$term) %>%
-    mutate(term=paste(Var1, Var2, sep="/"),
-           estimate_midpoint=means) %>%
-    filter(Var1!=Var2) %>%
-    select(term, estimate_midpoint) %>%
-    right_join(coef_table) %>%
-    mutate(estimate_midpoint=ifelse(is.na(estimate_midpoint), estimate,
-                                    estimate_midpoint),
-           estimate=ifelse(estimate==estimate_midpoint, 0, estimate))
+  #ok convert this to have single race groups for each multiracial group
+  #panel and get the assumed midpoint value
+  coef_table <- map(as.list(str_subset(coefs$term, "/")), function(x) {
+    coefs %>%
+      filter(term %in% c(x, str_split(x, "/")[[1]])) %>%
+      mutate(mrace=factor(x,
+                          levels=c("Black/White","White/Indigenous",
+                                   "Black/Indigenous",
+                                   "White/Latino","Black/Latino",
+                                   "Indigenous/Latino",
+                                   "White/Asian","Black/Asian",
+                                   "Indigenous/Asian","Latino/Asian")),
+             estimate_mid=ifelse(multiracial, 
+                                 mean(estimate[!multiracial]),
+                                 estimate),
+             estimate=ifelse(multiracial, estimate_mid+estimate,
+                             estimate))
+  }) %>%
+    bind_rows()
 
-  ggplot(coef_table, aes(x=estimate_midpoint, y=estimate_midpoint+estimate,
+  ggplot(coef_table, aes(x=estimate_mid, y=estimate,
                          color=multiracial,
-                         ymin=estimate_midpoint+estimate-1.96*std.error,
-                         ymax=estimate_midpoint+estimate+1.96*std.error))+
-    geom_abline(intercept = 0, slope=1, linetype=2, color="grey70")+
-   # geom_hline(yintercept = 0, linetype=2, color="grey40")+
-  #  geom_vline(xintercept = 0, linetype=2, color="grey40")+
+                         ymin=estimate-1.96*std.error,
+                         ymax=estimate+1.96*std.error))+
+    geom_abline(intercept = 0, slope=1, linetype=1, color="grey80")+
+    geom_hline(data=subset(coef_table, !multiracial),
+               aes(yintercept=estimate), linetype=3, color="grey80")+
     geom_linerange(alpha=0.5)+
     geom_point()+
-    geom_text_repel(aes(label=term), min.segment.length = 0)+
+    geom_text_repel(aes(label=term), size=3)+
+    facet_wrap(~mrace, ncol=3)+
     theme_bw()+
-    scale_color_viridis_d(end=0.6)+
-    labs(x="expected log odds ratio based on midway point",
-         y="actual log odds ratio")
+    theme(legend.position = "none", panel.grid = element_blank())+
+    scale_color_manual(values=c("grey40","black"))+
+    labs(x="expected log odds ratio of grade retention based on midpoint",
+         y="actual log odds ratio of grade retenion")
 }
+```
+
+```{r}
+temp <- plot_effect_scatter(model4)
 ```

### 2021-11-11 analysis/analysis.Rmd
Remove old plot function
The original plot I had is not nearly as useful so I am getting rid of the code

@@
 
-```{r}
-plot_mr_effects <- function(model) {
-  coef_table <- tidy(model) %>%
-    filter(str_detect(term, "race") & str_detect(term, "/") & !str_detect(term, "Multiracial")) %>%
-    mutate(term=str_remove(term, "race"))
-
-  #add in the placements for each single race group
-  single_race <- tidy(model) %>%
-    filter(str_detect(term, "race") & !str_detect(term, "/") & !str_detect(term, "Multiracial")) %>%
-    select(term, estimate) %>%
-    mutate(term=str_remove(term, "race"))
-  x <- c(0, single_race$estimate)
-  names(x) <- c("White",single_race$term)
-  coef_table$lower <- NA
-  coef_table$higher <- NA
-  for(i in 1:nrow(coef_table)) {
-    groups <- str_split(coef_table$term[i], "/")[[1]]
-    temp <- diff(sort(x[groups]))/2
-    coef_table$lower[i] <- -temp
-    coef_table$higher[i] <- temp
-  }
-
-  ggplot(coef_table, aes(x=reorder(term, higher, mean), y=estimate, 
-                         ymin=estimate-1.96*std.error, ymax=estimate+1.96*std.error))+
-    geom_hline(yintercept = 0, linetype=2)+
-    geom_pointrange()+
-    geom_point(aes(x=reorder(term, higher, mean), y=higher, color="red"))+
-    geom_point(aes(x=reorder(term, higher, mean), y=lower, color="red"))+
-    coord_flip()+
-    theme_bw()
-}
-```
-
-```{r}
-plot_mr_effects(model2)
-```
-
-```{r}
-plot_mr_effects(model4)
-```
-
 Try a plot that shows the expectation based on being exactly in the middle on the x-axis and the actual value on the y-axis.
@@
 ```{r}
-temp <- plot_effect_scatter(model4)
+plot_effect_scatter(model4)
 ```

### 2021-11-11 analysis/analysis.Rmd
Add plot_effect_point function
Instead of using a scatterplot, this function creates a dot plot panelized by multiracial groups. I think it is far easier to understand than the scatterplots.

@@
 
+```{r}
+plot_effect_point <- function(model) {
+
+  coefs <- tidy(model) %>%
+    filter(str_detect(term, "race") & !str_detect(term, "Multiracial")) %>%
+    mutate(term=str_remove(term, "race")) %>%
+    select(term, estimate, std.error) %>%
+    bind_rows(tibble(term="White", estimate=0, std.error=NA)) %>%
+    mutate(multiracial=str_detect(term, "/"))
+
+  #ok convert this to have single race groups for each multiracial group
+  #panel and get the assumed midpoint value
+  coef_table <- map(as.list(str_subset(coefs$term, "/")), function(x) {
+    coefs %>%
+      filter(term %in% c(x, str_split(x, "/")[[1]])) %>%
+      mutate(mrace=factor(x,
+                          levels=c("Black/White",
+                                   "White/Indigenous",
+                                   "Latino/Asian",
+                                   "Black/Latino",
+                                   "White/Latino",
+                                   "Indigenous/Latino",
+                                   "Black/Asian",
+                                   "White/Asian",
+                                   "Indigenous/Asian",
+                                   "Black/Indigenous")),
+             estimate_mid=ifelse(multiracial, 
+                                 mean(estimate[!multiracial]),
+                                 estimate),
+             estimate=ifelse(multiracial, estimate_mid+estimate,
+                             estimate))
+  }) %>%
+    bind_rows()
+
+  # I want term to be ordered so that multiracial groups are always between two
+  # constituent monoracial groups and monoracial groups to be sorted by the 
+  # magnitude of the estimate option. This is complicated but I should be able
+  # to do this by inserting multiracial groups immediately below the higher 
+  # of the two groups
+
+  #get terms ordered
+  race_names <- levels(reorder(factor(coef_table$term), coef_table$estimate))
+  #pull out single race and multiple race as separate vectors
+  race_single <- str_subset(race_names, "/", negate=TRUE)
+  race_multi <- str_subset(race_names, "/", negate=FALSE)
+  #start with single races
+  race_full <- race_single
+  #now loop through multiple race names and append based on the higher group
+  for(race in race_multi) {
+    temp <- str_split(race, "/")[[1]]
+    idx <- max(which(race_full==temp[1]), which(race_full==temp[2]))
+    race_full <- append(race_full, race, idx-1)
+  }
+
+  #now use this as the levels of term as a factor
+  coef_table <- coef_table %>%
+    mutate(term=factor(term, levels=race_full))
+
+  ggplot(coef_table, aes(x=term, y=estimate,
+                         color=multiracial,
+                         ymin=estimate-1.96*std.error,
+                         ymax=estimate+1.96*std.error))+
+    geom_hline(aes(yintercept=estimate_mid, linetype=multiracial), 
+               color="grey80")+
+    geom_linerange(alpha=0.5)+
+    geom_point()+
+    coord_flip()+
+    facet_wrap(~mrace, ncol=3, scales="free_y")+
+    theme_bw()+
+    theme(legend.position = "none", panel.grid = element_blank())+
+    scale_color_manual(values=c("grey40","black"))+
+    scale_linetype_manual(values=c(3,1))+
+    labs(x=NULL,
+         y="actual log odds ratio of grade retenion")
+}
+```
+
 ## Logit panic sensitivity results

### 2021-11-11 analysis/analysis.Rmd
Move plotting functions to useful functions

--- 2021-11-11 analysis/analysis.Rmd
@@
 ```{r}
-plot_effect_scatter <- function(model) {
 
-  coefs <- tidy(model) %>%
-    filter(str_detect(term, "race") & !str_detect(term, "Multiracial")) %>%
-    mutate(term=str_remove(term, "race")) %>%
-    select(term, estimate, std.error) %>%
-    bind_rows(tibble(term="White", estimate=0, std.error=NA)) %>%
-    mutate(multiracial=str_detect(term, "/"))
-
-  #ok convert this to have single race groups for each multiracial group
-  #panel and get the assumed midpoint value
-  coef_table <- map(as.list(str_subset(coefs$term, "/")), function(x) {
-    coefs %>%
-      filter(term %in% c(x, str_split(x, "/")[[1]])) %>%
-      mutate(mrace=factor(x,
-                          levels=c("Black/White","White/Indigenous",
-                                   "Black/Indigenous",
-                                   "White/Latino","Black/Latino",
-                                   "Indigenous/Latino",
-                                   "White/Asian","Black/Asian",
-                                   "Indigenous/Asian","Latino/Asian")),
-             estimate_mid=ifelse(multiracial, 
-                                 mean(estimate[!multiracial]),
-                                 estimate),
-             estimate=ifelse(multiracial, estimate_mid+estimate,
-                             estimate))
-  }) %>%
-    bind_rows()
-
-  ggplot(coef_table, aes(x=estimate_mid, y=estimate,
-                         color=multiracial,
-                         ymin=estimate-1.96*std.error,
-                         ymax=estimate+1.96*std.error))+
-    geom_abline(intercept = 0, slope=1, linetype=1, color="grey80")+
-    geom_hline(data=subset(coef_table, !multiracial),
-               aes(yintercept=estimate), linetype=3, color="grey80")+
-    geom_linerange(alpha=0.5)+
-    geom_point()+
-    geom_text_repel(aes(label=term), size=3)+
-    facet_wrap(~mrace, ncol=3)+
-    theme_bw()+
-    theme(legend.position = "none", panel.grid = element_blank())+
-    scale_color_manual(values=c("grey40","black"))+
-    labs(x="expected log odds ratio of grade retention based on midpoint",
-         y="actual log odds ratio of grade retenion")
-}
 ```
@@
 plot_effect_scatter(model4)
+```
+
+```{r}
+
+```
+
+```{r}
+plot_effect_point(model4)
 ```

### 2021-11-11 analysis/useful_functions.R

@@ -2,6 +2,9 @@
 
 #Put any functions that might be used across multiple scripts here so that they can be properly sourced in.
 
+
+# Functions for coding and cleaning ---------------------------------------
+
 code_race <- function(race, hisp) {
   race <- case_when(
     is.na(race) | is.na(hisp) ~ NA_character_,
@@ -127,4 +130,129 @@ is_race_consistent <- function(race_child, hispan_child, race_mom, hispan_mom,
        (!str_detect(race_constructed, "Latino") & hispan_child==0))
 
   return(TF)
+}
+
+
+# Functions for plotting effects ------------------------------------------
+
+plot_effect_scatter <- function(model) {
+  
+  coefs <- tidy(model) %>%
+    filter(str_detect(term, "race") & !str_detect(term, "Multiracial")) %>%
+    mutate(term=str_remove(term, "race")) %>%
+    select(term, estimate, std.error) %>%
+    bind_rows(tibble(term="White", estimate=0, std.error=NA)) %>%
+    mutate(multiracial=str_detect(term, "/"))
+  
+  #ok convert this to have single race groups for each multiracial group
+  #panel and get the assumed midpoint value
+  coef_table <- map(as.list(str_subset(coefs$term, "/")), function(x) {
+    coefs %>%
+      filter(term %in% c(x, str_split(x, "/")[[1]])) %>%
+      mutate(mrace=factor(x,
+                          levels=c("Black/White","White/Indigenous",
+                                   "Black/Indigenous",
+                                   "White/Latino","Black/Latino",
+                                   "Indigenous/Latino",
+                                   "White/Asian","Black/Asian",
+                                   "Indigenous/Asian","Latino/Asian")),
+             estimate_mid=ifelse(multiracial, 
+                                 mean(estimate[!multiracial]),
+                                 estimate),
+             estimate=ifelse(multiracial, estimate_mid+estimate,
+                             estimate))
+  }) %>%
+    bind_rows()
+  
+  ggplot(coef_table, aes(x=estimate_mid, y=estimate,
+                         color=multiracial,
+                         ymin=estimate-1.96*std.error,
+                         ymax=estimate+1.96*std.error))+
+    geom_abline(intercept = 0, slope=1, linetype=1, color="grey80")+
+    geom_hline(data=subset(coef_table, !multiracial),
+               aes(yintercept=estimate), linetype=3, color="grey80")+
+    geom_linerange(alpha=0.5)+
+    geom_point()+
[…]

### 2021-11-13 analysis/useful_functions.R
Add calculate_cond_means function
This function will create the coefficient table that I use to make the dotplot of effects. I am separating the creation of the table from the actual ggplot. Right now the function just uses some reasonable assumptions as baselines for the non-race variables. The function returns probabilities by default but there is an argument to return log-odds instead.

@@ -182,14 +182,30 @@ plot_effect_scatter <- function(model) {
          y="actual log odds ratio of grade retenion")
 }
 
-plot_effect_point <- function(model) {
+calculate_cond_means <- function(model, type="response") {
   
-  coefs <- tidy(model) %>%
-    filter(str_detect(term, "race") & !str_detect(term, "Multiracial")) %>%
-    mutate(term=str_remove(term, "race")) %>%
-    select(term, estimate, std.error) %>%
-    bind_rows(tibble(term="White", estimate=0, std.error=NA)) %>%
+  #TODO: Estimate means for each variable, currently just supplying some
+  #reasonable baseline
+  fake_data <- data.frame(race=levels(acs$race),
+                          year=2015,
+                          current_grade="6th",
+                          state=13,
+                          metro="Metro non-central",
+                          foreign_born=FALSE,
+                          foreign_born_mother=FALSE,
+                          foreign_born_father=FALSE,
+                          degree_mother="HS",
+                          degree_father="HS",
+                          family_income=mean(acs$family_income),
+                          own_home=TRUE,
+                          parents_married=TRUE)
+  
+  predicted <- predict(model, fake_data, se=TRUE, type=type)
+  
+  coefs <- tibble(term=fake_data$race, estimate=predicted$fit, 
+                  se=predicted$se.fit) %>%
     mutate(multiracial=str_detect(term, "/"))
+  
   
   #ok convert this to have single race groups for each multiracial group
   #panel and get the assumed midpoint value
@@ -210,16 +226,11 @@ plot_effect_point <- function(model) {
              estimate_mid=ifelse(multiracial, 
                                  mean(estimate[!multiracial]),
                                  estimate),
-             estimate=ifelse(multiracial, estimate_mid+estimate,
-                             estimate))
+             se_mid=ifelse(multiracial, 
+                                  sqrt(sum(se[!multiracial]^2)/4),
+                                  se))
   }) %>%
     bind_rows()
-  
-  # I want term to be ordered so that multiracial groups are always between two
-  # constituent monoracial groups and monoracial groups to be sorted by the 
-  # magnitude of the estimate option. This is complicated but I should be able
-  # to do this by inserting multiracial groups immediately below the higher 
[…]

### 2021-11-14 analysis/analysis.Rmd
Remove oout of date code chunks
Move conditional prob plot to analysis.Rmd

@@
 ```{r}
+coef_table <- calculate_cond_means(model4)
 
-```
-
-```{r}
-plot_effect_scatter(model4)
-```
-
-```{r}
-
-```
-
-```{r}
-plot_effect_point(model4)
+ggplot(coef_table, aes(x=term, y=estimate,
+                       color=multiracial,
+                       ymin=estimate-1.39*se,
+                       ymax=estimate+1.39*se))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.39*se_mid,
+                ymax=estimate_mid+1.39*se_mid), 
+            fill="grey90", color="grey90", alpha=0.7)+
+  geom_linerange(data=subset(coef_table, multiracial))+
+  geom_point()+
+  coord_flip()+
+  facet_wrap(~mrace, ncol=3, scales="free_y")+
+  theme_bw()+
+  theme(legend.position = "none", panel.grid = element_blank())+
+  scale_color_manual(values=c("grey40","red"))+
+  scale_linetype_manual(values=c(3,1))+
+  scale_y_continuous(labels = scales::percent)+
+  labs(x=NULL,
+       y="predicted probability of being behind expected grade")
 ```

### 2021-11-14 analysis/useful_functions.R

@@ -253,20 +253,3 @@ calculate_cond_means <- function(model, type="response") {
   return(coef_table)
 }
 
-#test ggplot code, will need to move eventually
-ggplot(coef_table, aes(x=term, y=estimate,
-                       color=multiracial,
-                       ymin=estimate-1.39*se,
-                       ymax=estimate+1.39*se))+
-  geom_hline(aes(yintercept=estimate_mid, linetype=multiracial), 
-             color="grey80")+
-  geom_linerange(alpha=0.5)+
-  geom_point()+
-  coord_flip()+
-  facet_wrap(~mrace, ncol=3, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","black"))+
-  scale_linetype_manual(values=c(3,1))+
-  labs(x=NULL,
-       y="actual log odds ratio of grade retenion")

### 2021-11-14 analysis/useful_functions.R
remove out of data plotting function

@@ -135,53 +135,6 @@ is_race_consistent <- function(race_child, hispan_child, race_mom, hispan_mom,
 
 # Functions for plotting effects ------------------------------------------
 
-plot_effect_scatter <- function(model) {
-  
-  coefs <- tidy(model) %>%
-    filter(str_detect(term, "race") & !str_detect(term, "Multiracial")) %>%
-    mutate(term=str_remove(term, "race")) %>%
-    select(term, estimate, std.error) %>%
-    bind_rows(tibble(term="White", estimate=0, std.error=NA)) %>%
-    mutate(multiracial=str_detect(term, "/"))
-  
-  #ok convert this to have single race groups for each multiracial group
-  #panel and get the assumed midpoint value
-  coef_table <- map(as.list(str_subset(coefs$term, "/")), function(x) {
-    coefs %>%
-      filter(term %in% c(x, str_split(x, "/")[[1]])) %>%
-      mutate(mrace=factor(x,
-                          levels=c("Black/White","White/Indigenous",
-                                   "Black/Indigenous",
-                                   "White/Latino","Black/Latino",
-                                   "Indigenous/Latino",
-                                   "White/Asian","Black/Asian",
-                                   "Indigenous/Asian","Latino/Asian")),
-             estimate_mid=ifelse(multiracial, 
-                                 mean(estimate[!multiracial]),
-                                 estimate),
-             estimate=ifelse(multiracial, estimate_mid+estimate,
-                             estimate))
-  }) %>%
-    bind_rows()
-  
-  ggplot(coef_table, aes(x=estimate_mid, y=estimate,
-                         color=multiracial,
-                         ymin=estimate-1.96*std.error,
-                         ymax=estimate+1.96*std.error))+
-    geom_abline(intercept = 0, slope=1, linetype=1, color="grey80")+
-    geom_hline(data=subset(coef_table, !multiracial),
-               aes(yintercept=estimate), linetype=3, color="grey80")+
-    geom_linerange(alpha=0.5)+
-    geom_point()+
-    geom_text_repel(aes(label=term), size=3)+
-    facet_wrap(~mrace, ncol=3)+
-    theme_bw()+
-    theme(legend.position = "none", panel.grid = element_blank())+
-    scale_color_manual(values=c("grey40","black"))+
-    labs(x="expected log odds ratio of grade retention based on midpoint",
-         y="actual log odds ratio of grade retenion")
-}
-
[…]

### 2021-11-14 analysis/useful_functions.R
Remove stairstep coding for parent degree
Now that I am needing to mean center all variables for plots, the stairstep coding is not helpful

@@ -22,17 +22,14 @@ code_race <- function(race, hisp) {
 code_degree <- function(educd) {
   degree <- case_when(
     is.na(educd) | educd==1 ~  NA_character_,
-    educd<62 ~ "Less than HS",
-    educd<81 ~ "HS diploma",
-    educd<101 ~ "AA degree",
-    educd==101 ~ "BA degree",
-    TRUE ~ "Grad degree"
+    educd<62 ~ "LHS",
+    educd<81 ~ "HS",
+    educd<101 ~ "AA",
+    educd==101 ~ "BA",
+    TRUE ~ "G"
   )
   degree <- factor(degree,
-                   levels=c("Less than HS","HS diploma","AA degree","BA degree",
-                            "Grad degree"),
-                   ordered=TRUE)
-  degree <- ordered_factor(degree)
+                   levels=c("LHS","HS","AA","BA","G"))
   return(degree)
 }
 

### 2021-11-14 analysis/useful_functions.R
Remove fractional race contrasts
These are not really needed now that I am using conditional means to display the results. I have the moved the code for creating this contrast to a function in the useful functions script in case I want to use it later.

@@ -47,6 +47,24 @@ ordered_factor <- function(fact_var) {
   return(ord_fact)
 }
 
+#This function will return contrasts using fractions of single race
+#responses for multiracial respondents
+race_half_contrast <- function(race) {
+  #the default treatment contrast will do most of the work, but I need to 
+  #put in the 0.5 cases
+  contr_race <- contrasts(race)
+  #lets loop through rows for multiracials and decide where the 0.5 should go
+  for(i in str_which(rownames(contr_race),"/")) {
+    race_name <- rownames(contr_race)[i]
+    comp_races <- str_split(race_name, "/")[[1]]
+    fraction <- 1/length(comp_races)
+    #one category is missing because it is the reference so remove if so
+    comp_races <- comp_races[comp_races %in% colnames(contr_race)]
+    contr_race[race_name, comp_races] <- fraction
+  }
+  return(contr_race)
+}
+
 ## Determine if race coding is consistent with reported race
 is_race_consistent <- function(race_child, hispan_child, race_mom, hispan_mom,
                                race_dad, hispan_dad, race_constructed) {

### 2021-11-14 analysis/analysis.Rmd
Code design matrix
I have created a design matrix from the analytical dataset that is used for the actual models. It allows me to easily mean center everything and then set defaults for the predict command.

@@
 
-```{r}
-#try a model
-model0 <- glm(below_exp_grade~as.factor(year)+current_grade+as.factor(state)+metro, 
-              data=acs, family=binomial)
-model1 <- update(model0, .~.+race)
-model2 <- update(model1, .~.+foreign_born+foreign_born_mother+foreign_born_father)
-model3 <- update(model2, .~.+degree_mother+degree_father)
-model4 <- update(model3, .~.+sqrt(family_income/10000)+own_home+parents_married)
+```{r design-matrix}
 
-knitreg(list(model1, model2, model3, model4))
+metro <- model.matrix(~metro, acs)[,-1]
+colnames(metro) <- c("met_non_metro","met_central_city","met_non_central_city",
+                     "met_city_indeter")
+#mean center
+metro <- t(t(metro)-apply(metro, 2, mean))
+metro <- as.data.frame(metro)
+
+years <- model.matrix(~as.factor(year), acs)[,-1]
+colnames(years) <- paste("y", 2011:2019, sep="")
+#mean center
+years <- t(t(years)-apply(years, 2, mean))
+years <- as.data.frame(years)
+
+states <- model.matrix(~as.factor(state), acs)[,-1]
+colnames(states) <- paste("s", 1:50, sep="")
+#mean center
+states <- t(t(states)-apply(states, 2, mean))
+states <- as.data.frame(states)
+
+grades <- model.matrix(~current_grade, acs)[,-1]
+#mean center
+grades <- t(t(grades)-apply(grades, 2, mean))
+grades <- as.data.frame(grades)
+
+degree <- model.matrix(~degree_mother+degree_father, acs)[,-1]
+#mean center
+degree <- t(t(degree)-apply(degree, 2, mean))
+degree <- as.data.frame(degree)
+
+design_matrix <- cbind(states, years, grades, metro, degree)
+
+design_matrix$foreign_born <- acs$foreign_born-mean(acs$foreign_born)
+design_matrix$foreign_born_mother <- acs$foreign_born_mother-mean(acs$foreign_born_mother)
+design_matrix$foreign_born_father <- acs$foreign_born_father-mean(acs$foreign_born_father)
+
+design_matrix$income_trans <- sqrt(acs$family_income/10000)
+design_matrix$income_trans <- design_matrix$income_trans-mean(design_matrix$income_trans)
+design_matrix$own_home <- acs$own_home-mean(acs$own_home)
+design_matrix$parents_married <- acs$parents_married-mean(acs$parents_married)
+
+design_matrix$race <- acs$race
+design_matrix$below_exp_grade <- acs$below_exp_grade
 ```
@@
 ```{r}
+#start with raw race differences
+model0 <- glm(below_exp_grade~race, data=design_matrix, family=binomial)
+model1 <- update(model0, .~.+y2011+y2012+y2013+y2014+y2015+y2016+y2017+y2018+y2019+
+                current_grade1st+current_grade2nd+current_grade3rd+current_grade4th+
+                current_grade5th+current_grade6th+current_grade7th+current_grade8th+
+                current_grade9th+current_grade10th+current_grade11th+current_grade10th+
+                current_grade11th+current_grade12th+
+                met_non_metro+met_central_city+met_non_central_city+met_city_indeter+
+                s1+s2+s3+s4+s5+s6+s7+s8+s9+s10+s11+s12+s13+s14+s15+s16+s17+s18+s19+
+                s20+s21+s22+s23+s24+s25+s26+s27+s28+s29+s30+s31+s32+s33+s34+s35+s36+
+                s37+s38+s39+s40+s41+s42+s43+s44+s45+s46+s47+s48+s49+s50)
+model2 <- update(model1, .~.+foreign_born+foreign_born_mother+foreign_born_father)
+model3 <- update(model2, .~.+degree_motherHS+degree_motherAA+degree_motherBA+
+                   degree_motherG+degree_fatherHS+degree_fatherAA+
+                   degree_fatherBA+degree_fatherG)
+model4 <- update(model3, .~.+income_trans+own_home+parents_married)
+
+knitreg(list(model0, model1, model2, model3, model4))
+```
+
+```{r eval=FALSE}
 #run separate models by grade groups
@@
 Try a plot that shows the expectation based on being exactly in the middle on the x-axis and the actual value on the y-axis.
+
+```{r}
+#dateset for predict
+df_pred <- data.frame(race=levels(acs$race),
+                      y2011=0, y2012=0, y2013=0, y2014=0, y2015=0, y2016=0,
+                      y2017=0, y2018=0, y2019=0,
+                      current_grade1st=0, current_grade2nd=0, current_grade3rd=0,
+                      current_grade4th=0, current_grade5th=0, current_grade6th=0,
+                      current_grade7th=0, current_grade8th=0, current_grade9th=0,
+                      current_grade10th=0, current_grade11th=0, current_grade12th=0,
+                      met_non_metro=0, met_central_city=0, met_non_central_city=0,
+                      met_city_indeter=0,
+                      s1=0, s2=0, s3=0, s4=0, s5=0, s6=0, s7=0, s8=0, s9=0, s10=0,
+                      s11=0, s12=0, s13=0, s14=0, s15=0, s16=0, s17=0, s18=0, s19=0,
+                      s20=0, s21=0, s22=0, s23=0, s24=0, s25=0, s26=0, s27=0, s28=0,
+                      s29=0, s30=0, s31=0, s32=0, s33=0, s34=0, s35=0, s36=0, s37=0,
+                      s38=0, s39=0, s40=0, s41=0, s42=0, s43=0, s44=0, s45=0, s46=0,
+                      s47=0, s48=0, s49=0, s50=0,
+                      degree_motherHS=0, degree_motherAA=0, degree_motherBA=0,
+                      degree_motherG=0, degree_fatherHS=0, degree_fatherAA=0, 
+                      degree_fatherBA=0, degree_fatherG=0,
+                      foreign_born=0, foreign_born_father=0, foreign_born_mother=0,
+                      income_trans=0, own_home=0, parents_married=0)
+```
 

### 2021-11-14 analysis/useful_functions.R
Modify calculate_cond_means function to take a prediction dataset

@@ -150,27 +150,11 @@ is_race_consistent <- function(race_child, hispan_child, race_mom, hispan_mom,
 
 # Functions for plotting effects ------------------------------------------
 
-calculate_cond_means <- function(model, type="response") {
+calculate_cond_means <- function(model, at=df_pred, type="response") {
   
-  #TODO: Estimate means for each variable, currently just supplying some
-  #reasonable baseline
-  fake_data <- data.frame(race=levels(acs$race),
-                          year=2015,
-                          current_grade="6th",
-                          state=13,
-                          metro="Metro non-central",
-                          foreign_born=FALSE,
-                          foreign_born_mother=FALSE,
-                          foreign_born_father=FALSE,
-                          degree_mother="HS",
-                          degree_father="HS",
-                          family_income=mean(acs$family_income),
-                          own_home=TRUE,
-                          parents_married=TRUE)
+  predicted <- predict(model, df_pred, se=TRUE, type=type)
   
-  predicted <- predict(model, fake_data, se=TRUE, type=type)
-  
-  coefs <- tibble(term=fake_data$race, estimate=predicted$fit, 
+  coefs <- tibble(term=df_pred$race, estimate=predicted$fit, 
                   se=predicted$se.fit) %>%
     mutate(multiracial=str_detect(term, "/"))
   

### 2021-11-15 analysis/analysis.Rmd
Update models and plots

@@
 
-```{r}
+```{r tab-grade-by-age}
 table(acs$age, acs$current_grade)
@@
 20-year old sixth graders?
-I mean there are relatively few of these cases but still.
+I mean there are relatively few of these cases but still seems a bit odd.
 Its also interesting how this really starts expanding past 5th - did the Census Bureau "correct" values below 6th or something?
@@
 
-0. Baseline model with dummies for current grade and year
-1. Add race.
+0. Raw race differences
+1. Add fixed effects for state, current grade, year, and metro
 2. Add kids and parents foreign-born status.
-2. Add mother and father highest degree
-3. Add family income (square rooted), own your own home, and whether parents are married.
+3. Add mother and father highest degree
+4. Add family income (square rooted), own your own home, and whether parents are married.
 
-All models are estimated as logit models. Race is coded using the average between the two monoracial groups as the reference category for all multiracial responses.
+All models are estimated as logit models.
+In order to calculate conditional means for the plots below, I need to be able to feed back in data to the `predict` command, including fractional values for 1/0 dummy variables.
+Therefore, rather than use the `acs` data itself, I need to create a purely numeric design matrix where each factor variable is expanded out into a set of 1/0 dummies.
 
@@
 design_matrix$foreign_born <- acs$foreign_born-mean(acs$foreign_born)
-design_matrix$foreign_born_mother <- acs$foreign_born_mother-mean(acs$foreign_born_mother)
-design_matrix$foreign_born_father <- acs$foreign_born_father-mean(acs$foreign_born_father)
+design_matrix$foreign_born_mother <- acs$foreign_born_mother-
+  mean(acs$foreign_born_mother)
+design_matrix$foreign_born_father <- acs$foreign_born_father-
+  mean(acs$foreign_born_father)
 
 design_matrix$income_trans <- sqrt(acs$family_income/10000)
-design_matrix$income_trans <- design_matrix$income_trans-mean(design_matrix$income_trans)
+design_matrix$income_trans <- design_matrix$income_trans-
+  mean(design_matrix$income_trans)
 design_matrix$own_home <- acs$own_home-mean(acs$own_home)
-design_matrix$parents_married <- acs$parents_married-mean(acs$parents_married)
+design_matrix$parents_married <- acs$parents_married-
+  mean(acs$parents_married)
 
@@
 
+Now, I can run the models. Warning, the code chunks below take a bit of time to run!
+
 ```{r}
-#start with raw race differences
 model0 <- glm(below_exp_grade~race, data=design_matrix, family=binomial)
-model1 <- update(model0, .~.+y2011+y2012+y2013+y2014+y2015+y2016+y2017+y2018+y2019+
-                current_grade1st+current_grade2nd+current_grade3rd+current_grade4th+
-                current_grade5th+current_grade6th+current_grade7th+current_grade8th+
-                current_grade9th+current_grade10th+current_grade11th+current_grade10th+
+model1 <- update(model0, 
+                 .~.+y2011+y2012+y2013+y2014+y2015+y2016+y2017+y2018+y2019+
+                   current_grade1st+current_grade2nd+current_grade3rd+
+                   current_grade4th+current_grade5th+current_grade6th+
+                   current_grade7th+current_grade8th+current_grade9th+
+                   current_grade10th+current_grade11th+current_grade10th+
                    current_grade11th+current_grade12th+
-                met_non_metro+met_central_city+met_non_central_city+met_city_indeter+
-                s1+s2+s3+s4+s5+s6+s7+s8+s9+s10+s11+s12+s13+s14+s15+s16+s17+s18+s19+
-                s20+s21+s22+s23+s24+s25+s26+s27+s28+s29+s30+s31+s32+s33+s34+s35+s36+
-                s37+s38+s39+s40+s41+s42+s43+s44+s45+s46+s47+s48+s49+s50)
-model2 <- update(model1, .~.+foreign_born+foreign_born_mother+foreign_born_father)
+                   met_non_metro+met_central_city+met_non_central_city+
+                   met_city_indeter+
+                   s1+s2+s3+s4+s5+s6+s7+s8+s9+s10+s11+s12+s13+s14+s15+s16+s17+
+                   s18+s19+s20+s21+s22+s23+s24+s25+s26+s27+s28+s29+s30+s31+s32+
+                   s33+s34+s35+s36+s37+s38+s39+s40+s41+s42+s43+s44+s45+s46+s47+
+                   s48+s49+s50)
+model2 <- update(model1, .~.+foreign_born+foreign_born_mother+
+                   foreign_born_father)
 model3 <- update(model2, .~.+degree_motherHS+degree_motherAA+degree_motherBA+
@@
 
-```{r eval=FALSE}
-#run separate models by grade groups
-model4a <- update(model4, data=subset(acs, as.numeric(current_grade)<=6))
-model4b <- update(model4, data=subset(acs, as.numeric(current_grade)>6 &
-                                        as.numeric(current_grade)<=9))
-model4c <- update(model4, data=subset(acs, as.numeric(current_grade)>=10))
+The model results themselves are not that useful. I want to compare each multiracial group to:
 
-knitreg(list(model4a, model4b, model4c))
-```
+1. Its two constituent monoracial groups (e.g. Compare Black/White to White and Black)
+2. The assumption that the biracial group will be halfway between the two constituent monoracial groups.
 
-Try a plot that shows the expectation based on being exactly in the middle on the x-axis and the actual value on the y-axis.
+I can do this by calculating conditional means (or proportions to be more accurate) of the dependent variable across all of the possible race groups.
+For each conditional mean, I can also calculate standard errors for these estimates by the delta method.
+This can all be done via the `predict` command.
+
+To compare the results for each biracial gropu to the halfway assumption, I need to calculate the value midway between the two consituent groups and the standard error of that value.
+Calculating the midway value is just a straightforward mean calculation.
+To calculate the standard error, I take:
+
+$$\frac{\sqrt{s_1^2+s_2^2}}{2}$$
+where $s_1$ is the standard error for monoracial group 1 and $s_2$ is the standard error for monoracial group 2.
+This formula follows from the standard formulas for the variance of the sum of random variables.
+
+There are two additional issues that have to be addressed though.
+First, I need to decide whether I want to calculate conditional means at the level of the response (proportions) or the linear transformation (log-odds).
+For ease of interpretability, I think the best option here is the former for the main presentation of results, but I also want to test the sensitivity of these results if I do it the other way.
+
+Second, the conditional means only apply at specific values of the independent variables.
+So I need to hold all of the other variables at some value.
+Now the relative difference between racial groups wouldn't change if this was a linear model (or if I used log-odds rather than proportions) so those values wouldn't matter much in that case.
+However, if I am estimating the actual proportions, then I no longer have a linear model and I am now caught up indirectly in the *logit panic*.
+I don't think it will matter that much because the range of predicted values of grade retention is not going to be that high (like between 2% and 10%).
+However, I should test the sensitivity of the results to different specifications.
+As a default, it makes sense to use the mean values for all other variables.
+Thankfully, because I already mean-centered the design matrix, this means that all other variables should be at zero.
 
@@
 
+Now in order to calculate all the values that I need, I have created a function that will put this into a nice table for later plotting.
+
+```{r}
+coef_table <- calculate_cond_means(model0)
+```
+
+Now I can use this nice table to plot out the results. I illustrate what this basic plot looks like below for the case of black/white individuals.
+
+```{r}
+g <- coef_table %>%
+  filter(mrace=="Black/White") %>%
+  ggplot(aes(x=term, y=estimate, color=multiracial,
+             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
+                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
+            color=NA, alpha=0.2)+
+  geom_linerange(data=subset(coef_table, multiracial & mrace=="Black/White"))+
+  geom_point()+
+  coord_flip()+
+  theme_bw()+
+  theme(legend.position = "none", panel.grid = element_blank())+
+  scale_color_manual(values=c("grey40","red"))+
+  scale_fill_manual(values=c("grey40","red"))+
+  scale_linetype_manual(values=c(3,1))+
+  scale_y_continuous(labels = scales::percent)+
+  labs(x=NULL,
+       y="predicted probability of being behind expected grade",
+       title="Probability of being behind grade for biracial black/white respondents and\nmonoracial comparison groups",
+       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
+
+g+annotate("text", 1, 0.037, hjust="left",
+             label="Grey bands give 83%\nconfidence interval around\nmonoracial estimates")+
+  annotate("text", 2, 0.0475, hjust="left",
+             label="Red band gives 83%\nconfidence interval around\nmidpoint estimate")+
+  annotate("text", 2.5, 0.038, hjust="left",
+             label="Actual biracial estimate\nshown with 83% confidence\ninterval")
+```
+This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions and their binomial standard errors.
+The grey bands for the monoracial groups bookmark the range of results we might expect and the red band gives what would be the midpoint value between the groups, adjusting for statistical uncertainty.
+I can then see where the actual biracial estimate falls within this range.
+The results here indicate that biracial black/white students have probabilities of grade retention much closer to the lower values of white than black students.
+However, their probability is still a bit higher than that for whites.
+
+Why I am using 83% (technically 83.4%) confidence intervals?
+This is a nice visual trick.
+People often incorrectly believe that overlapping 95% confidence intervals on two means indicates that the difference in the means is not statistically significant at $p<.05$.
+This is not always true and leads in practice to a much lower $\alpha$ for rejection.
+For that to be true, one needs to use 1.386 rather than 1.96 in estimating the confidence interval which produces an 83.4% confidence interval.
+It turns out that 1.386 is equal to $1.96/\sqrt{2}$.
+The intuition behind this technique is given nicely [here](https://chris-said.io/2014/12/01/independent-t-tests-and-the-83-confidence-interval-a-useful-trick-for-eyeballing-your-data/).
+Because I use this interval, we can know that if intervals (or bands) do not overlap, then the difference in means between the two groups is statistically significant at $p<.05$.
+In this case, the biracial estimate is statistically significantly different from both monoracial estimates as well as the "halfway" estimate.
+In short, biracial black/white students are held back at rates much closer but somewhat higher than whites, rather than blacks.
+
+The results here are the same as the raw differences, but I can also fit this graph to any of the models, so I can effectively give conditional probabilities.
+The plot below does this for model 4, by panelizing each multiracial group.
+
 ```{r}
@@
 
-ggplot(coef_table, aes(x=term, y=estimate,
-                       color=multiracial,
-                       ymin=estimate-1.39*se,
-                       ymax=estimate+1.39*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.39*se_mid,
-                ymax=estimate_mid+1.39*se_mid), 
-            fill="grey90", color="grey90", alpha=0.7)+
+ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
+                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
+                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
+            color=NA, alpha=0.2)+
   geom_linerange(data=subset(coef_table, multiracial))+
@@
   scale_color_manual(values=c("grey40","red"))+
+  scale_fill_manual(values=c("grey40","red"))+
   scale_linetype_manual(values=c(3,1))+
@@
   labs(x=NULL,
-       y="predicted probability of being behind expected grade")
+       y="predicted probability of being behind expected grade",
+       title="Probability of being behind grade for biracial respondents and\nmonoracial comparison groups",
+       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
 ```
 
-## Logit panic sensitivity results
+As the table here shows, after controlling for other stuff, most multiracial groups are quite similar to the monoracial constituent group with the lower grade retention rate.
+The only case where the biracial group is more like the higher monoracial group is for Latino/Asians.
+The results for indigenous groups are also quite noisy due to small size, but are generally consistent with the "below the halfway point."
 
-```{r, eval=FALSE}
-#try a LPM
-model1_lpm <- lm(below_exp_grade~race+as.factor(year), data=acs)
-model2_lpn <- update(model1, .~.+degree_mother+degree_father)
-model3_lpm <- update(model2, .~.+sqrt(family_income/10000)+own_home+parents_married)
-model4_lpm <- update(model3, .~.+foreign_born_mother+foreign_born_father)
+To compare these results across models, I can try a very large figure that panelizes by model on the column and group on the row.
 
-knitreg(list(model1_lpm, model2_lpm, model3_lpm, model4_lpm), digits=3)
+```{r, fig.height=12}
+coef_table <- calculate_cond_means(model0) %>%
+  mutate(model="crude")
+coef_table <- calculate_cond_means(model1) %>%
+  mutate(model="fixed effects") %>%
+  bind_rows(coef_table)
+coef_table <- calculate_cond_means(model2) %>%
+  mutate(model="cultural resources") %>%
+  bind_rows(coef_table)
+coef_table <- calculate_cond_means(model4) %>%
+  mutate(model="parent resources") %>%
+  bind_rows(coef_table) %>%
+  mutate(model=factor(model, levels=c("crude","fixed effects",
+                                      "cultural resources",
+                                      "parent resources")),
+         mrace=factor(mrace, levels=c("Black/White",
+                                      "Black/Latino",
+                                      "Black/Asian",
+                                      "Black/Indigenous",
+                                      "White/Indigenous",
+                                      "White/Latino",
+                                      "White/Asian",
+                                      "Latino/Asian",
+                                      "Indigenous/Latino",
+                                      "Indigenous/Asian")))
+
+ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
+                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
+                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
+            color=NA, alpha=0.2)+
+  geom_linerange(data=subset(coef_table, multiracial))+
+  geom_point()+
+  coord_flip()+
+  facet_grid(mrace~model, scales="free_y")+
+  theme_bw()+
+  theme(legend.position = "none", panel.grid = element_blank())+
+  scale_color_manual(values=c("grey40","red"))+
+  scale_fill_manual(values=c("grey40","red"))+
+  scale_linetype_manual(values=c(3,1))+
+  scale_y_continuous(labels = scales::percent)+
+  labs(x=NULL,
+       y="predicted probability of being behind expected grade",
+       title="Probability of being behind grade for biracial respondents and\nmonoracial comparison groups",
+       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
 ```
+
+Summarize these results.
+
+## Sensitivity tests
+
+```{r eval=FALSE}
+#run separate models by grade groups
+model4a <- update(model4, data=subset(acs, as.numeric(current_grade)<=6))
+model4b <- update(model4, data=subset(acs, as.numeric(current_grade)>6 &
+                                        as.numeric(current_grade)<=9))
+model4c <- update(model4, data=subset(acs, as.numeric(current_grade)>=10))
+
+knitreg(list(model4a, model4b, model4c))
+```

### 2021-11-15 analysis/analysis.Rmd
Clean up analysis.Rmd
I put names on all code chunks, placed tabular output into proper kable tables and  wrote some text for the front end.

@@
 
-# Grade Retention Profile
+# Grade retention summary
 
@@
 ```{r tab-grade-by-age}
-table(acs$age, acs$current_grade)
+tbl <- table(acs$age, acs$current_grade)
+
+tbl %>%
+  kbl(caption="Cross-tabulation of age and current grade of student") %>%
+  kable_minimal()
 ```
@@
 
-```{r}
+```{r check-retention-code}
 tapply(acs$below_exp_grade,acs[,c("age","current_grade")], mean)
@@
 
-```{r}
-round(tapply(acs$below_exp_grade, acs$current_grade, mean), 3)*100
+```{r retention-by-grade}
+acs %>%
+  group_by(current_grade) %>%
+  summarize(below_grade=mean(below_exp_grade)) %>%
+  ggplot(aes(x=current_grade, y=below_grade))+
+  geom_col()+
+  scale_y_continuous(labels=scales::percent)+
+  labs(x="current grade", y="percent clearly behind grade level for age")+
+  theme_bw()
 ```
@@
 
-```{r}
-round(tapply(acs$below_exp_grade, acs[,c("current_grade","year")], mean), 3)*100
-```
-
-Some evidence of declining grade retention prevalence at higher grades.
-
-# Lets look at the multiracial sample
-
-How big is it?
-
-```{r}
+```{r retention-by-year}
 acs %>%
-  filter(str_detect(race, "/") & !str_detect(race, "Multiracial")) %>%
-  ggplot(aes(x=reorder(race, race, length))) +
-  geom_bar()+
-  coord_flip()+
-  labs(x=NULL)+
+  group_by(year) %>%
+  summarize(below_grade=mean(below_exp_grade)) %>%
+  ggplot(aes(x=year, y=below_grade))+
+  geom_line()+
+  geom_point()+
+  scale_x_continuous(breaks=c(2010,2014,2018))+
+  scale_y_continuous(labels=scales::percent)+
+  labs(x="current grade", y="percent clearly behind grade level for age")+
   theme_bw()
@@
 
-```{r}
-resources_race <- acs %>% 
-  group_by(race) %>%
-  summarize(prop_held_back=mean(below_exp_grade),
-            mean_age=mean(age),
-            prop_foreign_born=mean(foreign_born),
-            mean_fam_inc=mean(family_income),
-            prop_own_home=mean(own_home),
-            prop_hs_diploma=mean(degree_father>="HS diploma" | 
-                                   degree_mother>="HS diploma"),
-            prop_ba_degree=mean(degree_father>="BA degree" | 
-                                  degree_mother>="BA degree"),
-            prop_parents_married=mean(parents_married),
-            mean_age_parents=mean((age_birth_mother+age_birth_father)/2),
-            prop_parent_foreign_born=mean(foreign_born_father | 
-                                            foreign_born_mother)) %>%
-  ungroup() %>%
-  filter(!str_detect(race,"Multiracial")) %>%
-  mutate(multiracial=str_detect(race, "/"))
+Overall, there has been some decline over time in the prevalence of grade retention.
+
+Lets look at this by grade and year
+
+```{r retention-by-year-grade}
+acs %>%
+  group_by(current_grade, year) %>%
+  summarize(below_grade=mean(below_exp_grade)) %>%
+  ggplot(aes(x=year, y=below_grade, group=current_grade, color=current_grade))+
+  geom_line()+
+  geom_point()+
+  scale_x_continuous(breaks=c(2010,2014,2018))+
+  scale_y_continuous(labels=scales::percent)+
+  scale_color_viridis_d()+
+  labs(x="current grade", y="percent clearly behind grade level for age")+
+  theme_bw()
 ```
 
-```{r}
-ggplot(resources_race, aes(x=reorder(race, mean_fam_inc, mean), y=mean_fam_inc,
-                           fill=multiracial))+
-  geom_col()+
-  coord_flip()+
-  labs(x=NULL, y="mean family income")
+The evidence here suggests that the decline is driven by the higher grades where grade retention was higher overall.
+
+# Descriptive statistics
+
+Lets looks at sample characteristics by race.
+
+```{r desc-stats}
+tbl_summary <- acs %>%
+  group_by(race) %>%
+  summarize(n=n(), 
+            held_back=round(mean(below_exp_grade), 3)*100,
+            mean_age=round(mean(age), 1),
+            f_born=round(mean(foreign_born), 3)*100,
+            par_fb=round(mean(foreign_born_father | 
+                                       foreign_born_mother), 3)*100,
+            fam_inc=round(mean(family_income), 0),
+            own_home=round(mean(own_home), 3)*100,
+            par_hs=round(mean(degree_father!="LHS" | 
+                                degree_mother!="LHS"), 3)*100,
+            par_ba=round(mean(degree_father=="BA" | 
+                                degree_father=="G" | 
+                                degree_mother=="BA" | 
+                                degree_mother=="G"), 3)*100,
+            par_mar=round(mean(parents_married), 3)*100) %>%
+  arrange(desc(n))
+
+index_biracial <- which(str_detect(tbl_summary$race, "/"))
+
+tbl_summary %>%
+  mutate_if(is.numeric, format, big.mark=",") %>%
+  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.", align = c("l",rep("r", ncol(tbl_summary)-1))) %>%
+  kable_paper() %>%
+  row_spec(index_biracial, background = "#D3D3D3")
 ```
 
-```{r}
-ggplot(resources_race, aes(x=reorder(race, prop_own_home, mean), y=prop_own_home,
-                           fill=multiracial))+
-  geom_col()+
-  coord_flip()+
-  labs(x=NULL, y="Proportion who own their own home")
-```
-
-```{r}
-ggplot(resources_race, aes(x=reorder(race, prop_hs_diploma, mean), y=prop_hs_diploma,
-                           fill=multiracial))+
-  geom_col()+
-  coord_flip()+
-  labs(x=NULL, y="Proportion where one parent has HS diploma")
-```
-
-```{r}
-ggplot(resources_race, aes(x=reorder(race, prop_ba_degree, mean), y=prop_ba_degree,
-                           fill=multiracial))+
-  geom_col()+
-  coord_flip()+
-  labs(x=NULL, y="Proportion where one parent has college degree")
-```
-
-## Raw differences in grade retention
-
-```{r}
-ggplot(resources_race, aes(x=reorder(race, prop_held_back, mean), y=prop_held_back,
-                           fill=multiracial))+
-  geom_col()+
-  coord_flip()
-```
-
-## Build some models
+# Model building time
 
@@
 
-```{r}
+```{r build-models, results='asis'}
 model0 <- glm(below_exp_grade~race, data=design_matrix, family=binomial)
@@
 
-$$\frac{\sqrt{s_1^2+s_2^2}}{2}$$
-where $s_1$ is the standard error for monoracial group 1 and $s_2$ is the standard error for monoracial group 2.
+$$\frac{\sqrt{s_1^2+s_2^2}}{2}$$ where $s_1$ is the standard error for monoracial group 1 and $s_2$ is the standard error for monoracial group 2.
 This formula follows from the standard formulas for the variance of the sum of random variables.
@@
 
-```{r}
+```{r create-predictor-data}
 #dateset for predict
@@
 
-```{r}
+```{r calc-cond-means}
 coef_table <- calculate_cond_means(model0)
@@
 
-```{r}
+```{r example-plot}
 g <- coef_table %>%
@@
        y="predicted probability of being behind expected grade",
-       title="Probability of being behind grade for biracial black/white respondents and\nmonoracial comparison groups",
+       title="Probability of being behind grade for biracial black/white respondents and monoracial\ncomparison groups",
        caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
@@
 ```
+
 This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions and their binomial standard errors.
@@
 
-```{r}
+```{r cond-means-model4}
 coef_table <- calculate_cond_means(model4)
@@
        y="predicted probability of being behind expected grade",
-       title="Probability of being behind grade for biracial respondents and\nmonoracial comparison groups",
+       title="Probability of being behind grade for biracial respondents and monoracial comparison groups",
        caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
@@
 
-```{r, fig.height=12}
+```{r cond-means-model-compare, fig.height=12}
 coef_table <- calculate_cond_means(model0) %>%
@@
 
-## Sensitivity tests
+# Sensitivity tests
 
-```{r eval=FALSE}
+```{r models-by-age, eval=FALSE}
 #run separate models by grade groups

### 2021-11-15 analysis/check_packages.R

@@ -6,7 +6,7 @@
 
 #add new packages to the chain here
 packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr","broom",
-             "ggrepel","purrr")
+             "ggrepel","purrr", "kableExtra", "tibble","scales")
 
 package.check <- lapply(packages, FUN = function(x) {
   if (!require(x, character.only = TRUE)) {

### 2021-11-15 analysis/check_packages.R
Change formatting of grade by age table
I added shading to indicate ages that were grade consistent.

@@ -6,7 +6,7 @@
 
 #add new packages to the chain here
 packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr","broom",
-             "ggrepel","purrr", "kableExtra", "tibble","scales")
+             "ggrepel","purrr", "kableExtra", "tibble","scales","janitor")
 
 package.check <- lapply(packages, FUN = function(x) {
   if (!require(x, character.only = TRUE)) {

### 2021-11-15 analysis/analysis.Rmd

@@
 
-tbl %>%
-  kbl(caption="Cross-tabulation of age and current grade of student") %>%
-  kable_minimal()
+acs %>% 
+  tabyl(age, current_grade) %>%
+  mutate_if(is.numeric, format, big.mark=",") %>%
+  kbl(caption="Cross-tabulation of age and current grade of student",
+      align=c("l",rep("r",13))) %>%
+  kable_paper() %>%
+  column_spec(2, background=c(rep("#D3D3D3",2), rep("white", 14))) %>%
+  column_spec(3, background=c("white", 
+                              rep("#D3D3D3",2), 
+                              rep("white", 13))) %>%
+  column_spec(4, background=c(rep("white", 2), 
+                              rep("#D3D3D3",2), 
+                              rep("white", 12))) %>%
+  column_spec(5, background=c(rep("white", 3), 
+                              rep("#D3D3D3",2), 
+                              rep("white", 11))) %>%
+  column_spec(6, background=c(rep("white", 4), 
+                              rep("#D3D3D3",2), 
+                              rep("white", 10))) %>%
+  column_spec(7, background=c(rep("white", 5), 
+                              rep("#D3D3D3",2), 
+                              rep("white", 9))) %>%
+  column_spec(8, background=c(rep("white", 6), 
+                              rep("#D3D3D3",2), 
+                              rep("white", 8))) %>%
+  column_spec(9, background=c(rep("white", 7), 
+                              rep("#D3D3D3",2), 
+                              rep("white", 7))) %>%
+  column_spec(10, background=c(rep("white", 8), 
+                               rep("#D3D3D3",2), 
+                               rep("white", 6))) %>%
+  column_spec(11, background=c(rep("white", 9), 
+                               rep("#D3D3D3",2), 
+                               rep("white", 5))) %>%
+  column_spec(12, background=c(rep("white", 10), 
+                               rep("#D3D3D3",2), 
+                               rep("white", 4))) %>%
+  column_spec(13, background=c(rep("white", 11), 
+                               rep("#D3D3D3",2), 
+                               rep("white", 3))) %>%
+  column_spec(14, background=c(rep("white", 12), 
+                               rep("#D3D3D3",2), 
+                               rep("white", 2)))
 ```

### 2021-11-16 analysis/organize_data.R
Add sample design variables to analytical data

@@ -203,7 +203,8 @@ acs <- acs %>%
 # Trim to analytical data -------------------------------------------------
 
 acs <- acs %>%
-  select(hhid, perwt, year, state, 
+  select(hhid, perwt, cluster, strata,
+         year, state, 
          age, current_grade, below_exp_grade, 
          race, sex, metro,
          age_birth_mother, age_birth_father, degree_mother, degree_father, 

### 2021-11-16 analysis/check_packages.R
Add survey package

@@ -6,7 +6,8 @@
 
 #add new packages to the chain here
 packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr","broom",
-             "ggrepel","purrr", "kableExtra", "tibble","scales","janitor")
+             "ggrepel","purrr", "kableExtra", "tibble","scales","janitor",
+             "survey")
 
 package.check <- lapply(packages, FUN = function(x) {
   if (!require(x, character.only = TRUE)) {

### 2021-11-16 analysis/organize_data.R
Remove hhid variable
The cluster variable already uniquely identifies households by year so no need for the constructed hhid variable

@@ -81,8 +81,7 @@ acs <- acs %>%
          sex=factor(sex, levels=1:2, labels=c("Male","Female")),
          metro=factor(metro, levels=0:4, 
                       labels=c("Indeterminable","Non-metro","Central city",
-                               "Metro non-central","Metro indeterminable")),
-         hhid=serial*10000+year) %>%
+                               "Metro non-central","Metro indeterminable"))) %>%
   #remove respondents with multiracial parents
   filter(!str_detect(race, "Multiracial"))
 
@@ -203,7 +202,7 @@ acs <- acs %>%
 # Trim to analytical data -------------------------------------------------
 
 acs <- acs %>%
-  select(hhid, perwt, cluster, strata,
+  select(pcluster, strata, perwt,
          year, state, 
          age, current_grade, below_exp_grade, 
          race, sex, metro,

### 2021-11-16 analysis/organize_data.R
Fix typo in previous commit

@@ -202,7 +202,7 @@ acs <- acs %>%
 # Trim to analytical data -------------------------------------------------
 
 acs <- acs %>%
-  select(pcluster, strata, perwt,
+  select(cluster, strata, perwt,
          year, state, 
          age, current_grade, below_exp_grade, 
          race, sex, metro,

### 2021-11-16 analysis/analysis.Rmd
Adjust models for sampling design
I use the survey library to adjust for the clustering of observations in households and to apply person weights. I also would like to apply the strata effects (because these can reduce SEs!) but currently svydesign is throwing up errors when I do.

@@
 design_matrix$below_exp_grade <- acs$below_exp_grade
+
+# add design effects
+design_matrix$cluster <- acs$cluster
+design_matrix$strata <- acs$strata
+design_matrix$perwt <- acs$perwt
+
+#TODO: It won't let me add strata - gives some error. This sucks because 
+#strata may help to reduce SEs
+design_matrix <- svydesign(ids=~cluster, weights=~perwt, data=design_matrix)
+
 ```
 
-Now, I can run the models. Warning, the code chunks below take a bit of time to run!
+Now, I can run the models.
+I am using the `svyglm` command which (given the design set up above) adjusts for clustering and weights observations based on sample weights.
+I also want to figure out how to adjust for strata, but `svydesign` is throwing up errors when I attemp to do that.
+Warning, the code chunks below take a bit of time to run!
 
 ```{r build-models, results='asis'}
-model0 <- glm(below_exp_grade~race, data=design_matrix, family=binomial)
+model0 <- svyglm(below_exp_grade~race, design=design_matrix, family=binomial)
 model1 <- update(model0, 
@@
 
-g+annotate("text", 1, 0.037, hjust="left",
+g+annotate("text", 1, 0.0375, hjust="left",
              label="Grey bands give 83%\nconfidence interval around\nmonoracial estimates")+
-  annotate("text", 2, 0.0475, hjust="left",
+  annotate("text", 2, 0.048, hjust="left",
              label="Red band gives 83%\nconfidence interval around\nmidpoint estimate")+
-  annotate("text", 2.5, 0.038, hjust="left",
+  annotate("text", 2.5, 0.039, hjust="left",
              label="Actual biracial estimate\nshown with 83% confidence\ninterval")

### 2021-11-16 analysis/organize_data.R
Adjust family income for inflation
This adjustment should resolves issue #13

@@ -76,6 +76,18 @@ acs <- acs %>%
          below_exp_grade=age>=as.numeric(current_grade)+6,
          family_income=ifelse(ftotinc==9999999, NA, 
                               ifelse(ftotinc<0, 0, ftotinc)),
+         #now adjust for inflation
+         family_income=case_when(
+           year==2010 ~ family_income*1.18,
+           year==2011 ~ family_income*1.14,
+           year==2012 ~ family_income*1.12,
+           year==2013 ~ family_income*1.10,
+           year==2014 ~ family_income*1.08,
+           year==2015 ~ family_income*1.08,
+           year==2016 ~ family_income*1.07,
+           year==2017 ~ family_income*1.05,
+           year==2018 ~ family_income*1.02,
+           year==2019 ~ family_income*1),
          own_home=ifelse(ownershp==0, NA, ownershp==1),
          foreign_born=bpl>=100,
          sex=factor(sex, levels=1:2, labels=c("Male","Female")),

### 2021-11-17 analysis/organize_data.R
Read in new data
Change index positions for new fixed width data, code new variables, and filter analytical data by data quality flags

@@ -18,24 +18,29 @@ options(max.print=999999)
 
 # Read the raw data -------------------------------------------------------
 
-acs <- read_fwf(here("analysis","input","usa_00112.dat.gz"), 
+acs <- read_fwf(here("analysis","input","usa_00114.dat.gz"), 
                 col_positions = 
                   fwf_positions(
-                    start = c(1,11,32,42,55,57,58,71,74,85,95,97,101,105,106,
-                              111,115,118,133,135,142,146,158,161,170,171,174,
-                              177,186,189,198,201,210,213),
-                    end   = c(4,18,41,54,56,57,69,71,80,94,96,98,104,105,108,
-                              113,117,120,134,141,145,149,160,163,170,171,176,
-                              179,188,191,200,203,212,215),
-                    col_names = c("year","serial","hhwt","cluster","state",
-                                  "metro","strata","ownershp","hhincome",
-                                  "perwt","momrule","poprule","related",
-                                  "sex","age","raced","hispand","bpl",
-                                  "gradeattd","ftotinc","related_mom",
-                                  "related_pop","age_mom","age_pop","marst_mom",
-                                  "marst_pop","raced_mom","raced_pop",
-                                  "hispand_mom","hispand_pop","bpl_mom",
-                                  "bpl_pop","educd_mom","educd_pop")),
+                    start = c(1,42,55,57,58,71,85,95,97,101,105,106,111,115,118,
+                              126,134,136,143,144,145,146,147,151,163,166,175,
+                              176,179,182,191,194,203,206,215,216,219,222,231,
+                              232,235,236),
+                    end   = c(4,54,56,57,69,71,94,96,98,104,105,108,113,117,120,
+                              126,135,142,143,144,145,146,150,154,165,168,175,
+                              176,181,184,193,196,205,208,215,216,221,224,231,
+                              232,235,236),
+                    col_names = c("year","cluster","state","metro","strata",
+                                  "ownershp","perwt","momrule","poprule",
+                                  "related","sex","age","raced","hispand",
+                                  "bpl","speakeng","gradeattd","ftotinc",
+                                  "qage","qhispan","qrace","qgradeat",
[…]

### 2021-11-17 analysis/check_packages.R
Add modmarg and sandwich packages

@@ -7,7 +7,7 @@
 #add new packages to the chain here
 packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr","broom",
              "ggrepel","purrr", "kableExtra", "tibble","scales","janitor",
-             "survey")
+             "survey","modmarg","sandwich")
 
 package.check <- lapply(packages, FUN = function(x) {
   if (!require(x, character.only = TRUE)) {

### 2021-11-17 analysis/analysis.Rmd
Don't use weights in estimation

@@
             f_born=round(mean(foreign_born), 3)*100,
+            spk_eng_well=round(mean(spk_eng=="Yes"), 3)*100,
+            spk_eng_par=round(mean(spk_eng_father=="Yes" & 
+                                     spk_eng_mother=="Yes"), 3)*100,
             par_fb=round(mean(foreign_born_father | 
@@
 
-design_matrix <- cbind(states, years, grades, metro, degree)
+spk_eng <- model.matrix(~spk_eng+spk_eng_father+spk_eng_mother, acs)[,-1]
+#mean center
+spk_eng <- t(t(spk_eng)-apply(spk_eng, 2, mean))
+spk_eng <- as.data.frame(spk_eng)
+
+design_matrix <- cbind(states, years, grades, metro, degree, spk_eng)
 
@@
 
-#TODO: It won't let me add strata - gives some error. This sucks because 
-#strata may help to reduce SEs
-design_matrix <- svydesign(ids=~cluster, weights=~perwt, data=design_matrix)
-
+design_matrix <- svydesign(ids=~cluster, data=design_matrix)
 ```
@@
 model2 <- update(model1, .~.+foreign_born+foreign_born_mother+
-                   foreign_born_father)
+                   foreign_born_father+spk_engSomewhat+spk_engYes+
+                   spk_eng_motherSomewhat+spk_eng_motherYes+
+                   spk_eng_fatherSomewhat+spk_eng_fatherYes)
 model3 <- update(model2, .~.+degree_motherHS+degree_motherAA+degree_motherBA+
@@
                       foreign_born=0, foreign_born_father=0, foreign_born_mother=0,
+                      spk_engSomewhat=0, spk_engYes=0, spk_eng_fatherSomewhat=0, 
+                      spk_eng_fatherYes=0, spk_eng_motherSomewhat=0, spk_engYes=0,
                       income_trans=0, own_home=0, parents_married=0)

### 2021-11-17 analysis/analysis.Rmd
Switch from predict to modmarg to calculate marginal means
The results will change slightly because I am now estimating means averaged across all respondents rather than at the mean respondent.

@@
 
-All models are estimated as logit models.
-In order to calculate conditional means for the plots below, I need to be able to feed back in data to the `predict` command, including fractional values for 1/0 dummy variables.
-Therefore, rather than use the `acs` data itself, I need to create a purely numeric design matrix where each factor variable is expanded out into a set of 1/0 dummies.
-
-```{r design-matrix}
-
-metro <- model.matrix(~metro, acs)[,-1]
-colnames(metro) <- c("met_non_metro","met_central_city","met_non_central_city",
-                     "met_city_indeter")
-#mean center
-metro <- t(t(metro)-apply(metro, 2, mean))
-metro <- as.data.frame(metro)
-
-years <- model.matrix(~as.factor(year), acs)[,-1]
-colnames(years) <- paste("y", 2011:2019, sep="")
-#mean center
-years <- t(t(years)-apply(years, 2, mean))
-years <- as.data.frame(years)
-
-states <- model.matrix(~as.factor(state), acs)[,-1]
-colnames(states) <- paste("s", 1:50, sep="")
-#mean center
-states <- t(t(states)-apply(states, 2, mean))
-states <- as.data.frame(states)
-
-grades <- model.matrix(~current_grade, acs)[,-1]
-#mean center
-grades <- t(t(grades)-apply(grades, 2, mean))
-grades <- as.data.frame(grades)
-
-degree <- model.matrix(~degree_mother+degree_father, acs)[,-1]
-#mean center
-degree <- t(t(degree)-apply(degree, 2, mean))
-degree <- as.data.frame(degree)
-
-spk_eng <- model.matrix(~spk_eng+spk_eng_father+spk_eng_mother, acs)[,-1]
-#mean center
-spk_eng <- t(t(spk_eng)-apply(spk_eng, 2, mean))
-spk_eng <- as.data.frame(spk_eng)
-
-design_matrix <- cbind(states, years, grades, metro, degree, spk_eng)
-
-design_matrix$foreign_born <- acs$foreign_born-mean(acs$foreign_born)
-design_matrix$foreign_born_mother <- acs$foreign_born_mother-
-  mean(acs$foreign_born_mother)
-design_matrix$foreign_born_father <- acs$foreign_born_father-
-  mean(acs$foreign_born_father)
-
-design_matrix$income_trans <- sqrt(acs$family_income/10000)
-design_matrix$income_trans <- design_matrix$income_trans-
-  mean(design_matrix$income_trans)
-design_matrix$own_home <- acs$own_home-mean(acs$own_home)
-design_matrix$parents_married <- acs$parents_married-
-  mean(acs$parents_married)
-
-design_matrix$race <- acs$race
-design_matrix$below_exp_grade <- acs$below_exp_grade
-
-# add design effects
-design_matrix$cluster <- acs$cluster
-design_matrix$strata <- acs$strata
-design_matrix$perwt <- acs$perwt
-
-design_matrix <- svydesign(ids=~cluster, data=design_matrix)
-```
-
-Now, I can run the models.
-I am using the `svyglm` command which (given the design set up above) adjusts for clustering and weights observations based on sample weights.
-I also want to figure out how to adjust for strata, but `svydesign` is throwing up errors when I attemp to do that.
-Warning, the code chunks below take a bit of time to run!
+All models are estimated as logit models. Warning, the code chunks below take a bit of time to run!
 
 ```{r build-models, results='asis'}
-model0 <- svyglm(below_exp_grade~race, design=design_matrix, family=binomial)
-model1 <- update(model0, 
-                 .~.+y2011+y2012+y2013+y2014+y2015+y2016+y2017+y2018+y2019+
-                   current_grade1st+current_grade2nd+current_grade3rd+
-                   current_grade4th+current_grade5th+current_grade6th+
-                   current_grade7th+current_grade8th+current_grade9th+
-                   current_grade10th+current_grade11th+current_grade10th+
-                   current_grade11th+current_grade12th+
-                   met_non_metro+met_central_city+met_non_central_city+
-                   met_city_indeter+
-                   s1+s2+s3+s4+s5+s6+s7+s8+s9+s10+s11+s12+s13+s14+s15+s16+s17+
-                   s18+s19+s20+s21+s22+s23+s24+s25+s26+s27+s28+s29+s30+s31+s32+
-                   s33+s34+s35+s36+s37+s38+s39+s40+s41+s42+s43+s44+s45+s46+s47+
-                   s48+s49+s50)
-model2 <- update(model1, .~.+foreign_born+foreign_born_mother+
-                   foreign_born_father+spk_engSomewhat+spk_engYes+
-                   spk_eng_motherSomewhat+spk_eng_motherYes+
-                   spk_eng_fatherSomewhat+spk_eng_fatherYes)
-model3 <- update(model2, .~.+degree_motherHS+degree_motherAA+degree_motherBA+
-                   degree_motherG+degree_fatherHS+degree_fatherAA+
-                   degree_fatherBA+degree_fatherG)
-model4 <- update(model3, .~.+income_trans+own_home+parents_married)
+model0 <- glm(below_exp_grade~race, data=acs, family=binomial)
+model1 <- update(model0, as.factor(year)+current_grade+as.factor(state)+metro)
+model2 <- update(model1, 
+                 .~.+foreign_born+foreign_born_mother+foreign_born_father+
+                   spk_eng+spk_eng_mother+spk_eng_father)
+model3 <- update(model2, .~.+degree_mother+degree_father)
+model4 <- update(model3, .~.+sqrt(family_income/10000)+own_home+parents_married)
 
@@
 
-I can do this by calculating conditional means (or proportions to be more accurate) of the dependent variable across all of the possible race groups.
-For each conditional mean, I can also calculate standard errors for these estimates by the delta method.
-This can all be done via the `predict` command.
+I can do this by calculating marginal means (or proportions to be more accurate) of the dependent variable across all of the possible race groups.
+In order to calculate these marginal means, I use the `marg` command in the `modmargs` library.
+This gives me results equivalent to average marginal effects but on the level (e.g.
+mean) rather than the effect (different in mean).
+Standard errors for these marginal means are calculated by the [delta method](https://cran.r-project.org/web/packages/modmarg/vignettes/delta-method.html).
+I adjust all SEs for clustering of respondents within households.
 
-To compare the results for each biracial gropu to the halfway assumption, I need to calculate the value midway between the two consituent groups and the standard error of that value.
+To compare the results for each biracial group to the halfway assumption, I need to calculate the value midway between the two consituent groups and the standard error of that value.
 Calculating the midway value is just a straightforward mean calculation.
@@
 
-There are two additional issues that have to be addressed though.
-First, I need to decide whether I want to calculate conditional means at the level of the response (proportions) or the linear transformation (log-odds).
-For ease of interpretability, I think the best option here is the former for the main presentation of results, but I also want to test the sensitivity of these results if I do it the other way.
+In order to calculate all the values that I need, I have created a function that will put this into a nice table for later plotting.
 
-Second, the conditional means only apply at specific values of the independent variables.
-So I need to hold all of the other variables at some value.
-Now the relative difference between racial groups wouldn't change if this was a linear model (or if I used log-odds rather than proportions) so those values wouldn't matter much in that case.
-However, if I am estimating the actual proportions, then I no longer have a linear model and I am now caught up indirectly in the *logit panic*.
-I don't think it will matter that much because the range of predicted values of grade retention is not going to be that high (like between 2% and 10%).
-However, I should test the sensitivity of the results to different specifications.
-As a default, it makes sense to use the mean values for all other variables.
-Thankfully, because I already mean-centered the design matrix, this means that all other variables should be at zero.
-
-```{r create-predictor-data}
-#dateset for predict
-df_pred <- data.frame(race=levels(acs$race),
-                      y2011=0, y2012=0, y2013=0, y2014=0, y2015=0, y2016=0,
-                      y2017=0, y2018=0, y2019=0,
-                      current_grade1st=0, current_grade2nd=0, current_grade3rd=0,
-                      current_grade4th=0, current_grade5th=0, current_grade6th=0,
-                      current_grade7th=0, current_grade8th=0, current_grade9th=0,
-                      current_grade10th=0, current_grade11th=0, current_grade12th=0,
-                      met_non_metro=0, met_central_city=0, met_non_central_city=0,
-                      met_city_indeter=0,
-                      s1=0, s2=0, s3=0, s4=0, s5=0, s6=0, s7=0, s8=0, s9=0, s10=0,
-                      s11=0, s12=0, s13=0, s14=0, s15=0, s16=0, s17=0, s18=0, s19=0,
-                      s20=0, s21=0, s22=0, s23=0, s24=0, s25=0, s26=0, s27=0, s28=0,
-                      s29=0, s30=0, s31=0, s32=0, s33=0, s34=0, s35=0, s36=0, s37=0,
-                      s38=0, s39=0, s40=0, s41=0, s42=0, s43=0, s44=0, s45=0, s46=0,
-                      s47=0, s48=0, s49=0, s50=0,
-                      degree_motherHS=0, degree_motherAA=0, degree_motherBA=0,
-                      degree_motherG=0, degree_fatherHS=0, degree_fatherAA=0, 
-                      degree_fatherBA=0, degree_fatherG=0,
-                      foreign_born=0, foreign_born_father=0, foreign_born_mother=0,
-                      spk_engSomewhat=0, spk_engYes=0, spk_eng_fatherSomewhat=0, 
-                      spk_eng_fatherYes=0, spk_eng_motherSomewhat=0, spk_engYes=0,
-                      income_trans=0, own_home=0, parents_married=0)
-```
-
-Now in order to calculate all the values that I need, I have created a function that will put this into a nice table for later plotting.
-
-```{r calc-cond-means}
-coef_table <- calculate_cond_means(model0)
+```{r calc-cond-means, warning=FALSE}
+coef_table_mod0 <- calculate_cond_means(model0)
 ```
@@
 ```{r example-plot}
-g <- coef_table %>%
+g <- coef_table_mod0 %>%
   filter(mrace=="Black/White") %>%
@@
             color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial & mrace=="Black/White"))+
+  geom_linerange(data=subset(coef_table_mod0, multiracial & mrace=="Black/White"))+
   geom_point()+
@@
 
-g+annotate("text", 1, 0.0375, hjust="left",
-             label="Grey bands give 83%\nconfidence interval around\nmonoracial estimates")+
-  annotate("text", 2, 0.048, hjust="left",
-             label="Red band gives 83%\nconfidence interval around\nmidpoint estimate")+
-  annotate("text", 2.5, 0.039, hjust="left",
-             label="Actual biracial estimate\nshown with 83% confidence\ninterval")
+g+annotate("text", 1, 0.035, hjust="left",
+             label="Grey bands give 83.4%\nconfidence interval around\nmonoracial estimates")+
+  annotate("text", 2, 0.045, hjust="left",
+             label="Red band gives 83.4%\nconfidence interval around\nmidpoint estimate")+
+  annotate("text", 2.5, 0.036, hjust="left",
+             label="Actual biracial estimate\nshown with 83.4% confidence\ninterval")
 ```
 
-This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions and their binomial standard errors.
+This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions and their binomial standard errors (although the SEs are slightly different because this model accounts for clustering).
 The grey bands for the monoracial groups bookmark the range of results we might expect and the red band gives what would be the midpoint value between the groups, adjusting for statistical uncertainty.
@@
 ```{r cond-means-model4}
-coef_table <- calculate_cond_means(model4)
+coef_table_mod4 <- calculate_cond_means(model4)
 
-ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
+ggplot(coef_table_mod4, aes(x=term, y=estimate, color=multiracial,
                        ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
@@
             color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial))+
+  geom_linerange(data=subset(coef_table_mod4, multiracial))+
   geom_point()+
@@
 ```{r cond-means-model-compare, fig.height=12}
-coef_table <- calculate_cond_means(model0) %>%
+#calculate remaining marginal means
+coef_table_mod1 <- calculate_cond_means(model1)
+coef_table_mod2 <- calculate_cond_means(model2)
+
+coef_table <- coef_table_mod0 %>%
   mutate(model="crude")
-coef_table <- calculate_cond_means(model1) %>%
+coef_table <- coef_table_mod1 %>%
   mutate(model="fixed effects") %>%
   bind_rows(coef_table)
-coef_table <- calculate_cond_means(model2) %>%
+coef_table <- coef_table_mod2 %>%
   mutate(model="cultural resources") %>%
   bind_rows(coef_table)
-coef_table <- calculate_cond_means(model4) %>%
+coef_table <- coef_table_mod4 %>%
   mutate(model="parent resources") %>%

### 2021-11-17 analysis/useful_functions.R

@@ -162,14 +162,16 @@ is_race_consistent <- function(race_child, hispan_child, race_mom, hispan_mom,
 
 # Functions for plotting effects ------------------------------------------
 
-calculate_cond_means <- function(model, at=df_pred, type="response") {
+calculate_cond_means <- function(model, cluster_var=acs$cluster) {
   
-  predicted <- predict(model, df_pred, se.fit=TRUE, type=type)
+  means_marginal <- marg(model, "race", type="levels", 
+                         vcov_mat=vcovCL(model, cluster=cluster_var))[[1]]
   
   #figuring out how to pull things out of this object is really 
   #annoying
-  coefs <- tibble(term=df_pred$race, estimate=coef(predicted), 
-                  se=as.vector(sqrt(vcov(predicted)))) %>%
+  coefs <- tibble(term=str_remove(means_marginal$Label, "race = "), 
+                  estimate=means_marginal$Margin, 
+                  se=means_marginal$Standard.Error) %>%
     mutate(multiracial=str_detect(term, "/"))
   
   

### 2021-11-17 analysis/analysis.Rmd
Fix a few things from previous commit
Fixed a few typos and supressed warnings in  code chunks that call marg

@@
 model0 <- glm(below_exp_grade~race, data=acs, family=binomial)
-model1 <- update(model0, as.factor(year)+current_grade+as.factor(state)+metro)
+model1 <- update(model0, .~.+as.factor(year)+current_grade+as.factor(state)+metro)
 model2 <- update(model1, 
@@
 
-```{r cond-means-model4}
+```{r cond-means-model4, warning=FALSE}
 coef_table_mod4 <- calculate_cond_means(model4)
@@
 
-```{r cond-means-model-compare, fig.height=12}
+```{r cond-means-model-compare, fig.height=12, warning=FALSE}
 #calculate remaining marginal means

### 2021-11-17 analysis/useful_functions.R
Change function name to calculate_marg_means

@@ -162,7 +162,7 @@ is_race_consistent <- function(race_child, hispan_child, race_mom, hispan_mom,
 
 # Functions for plotting effects ------------------------------------------
 
-calculate_cond_means <- function(model, cluster_var=acs$cluster) {
+calculate_marg_means <- function(model, cluster_var=acs$cluster) {
   
   means_marginal <- marg(model, "race", type="levels", 
                          vcov_mat=vcovCL(model, cluster=cluster_var))[[1]]

### 2021-11-17 analysis/analysis.Rmd

@@
 ```{r calc-cond-means, warning=FALSE}
-coef_table_mod0 <- calculate_cond_means(model0)
+coef_table_mod0 <- calculate_marg_means(model0)
 ```
@@
 ```{r cond-means-model4, warning=FALSE}
-coef_table_mod4 <- calculate_cond_means(model4)
+coef_table_mod4 <- calculate_marg_means(model4)
 
@@
 #calculate remaining marginal means
-coef_table_mod1 <- calculate_cond_means(model1)
-coef_table_mod2 <- calculate_cond_means(model2)
+coef_table_mod1 <- calculate_marg_means(model1)
+coef_table_mod2 <- calculate_marg_means(model2)
 

### 2021-11-18 analysis/analysis.Rmd
Add description of results for Black/White and Black/Latino respondents

@@
 
-Summarize these results.
+Ok, let me try to summarize these results, by group.
+
+## Black/White results
+
+For the first three models, the results are pretty consistent.
+Black/White biracials are much closer to the lower grade retention rates of whites, although they have rates that are clearly still higher than whites.
+Controlling for parental resources changes things considerably.
+First, the gap between Blacks and Whites shrinks considerably (visually it looks like its cut about in half).
+Second, Black/White biracials actually have lower rates than either group.
+They aren't that different than whites in effect size but the difference is statistically significant.
+
+Why does the big change occur in the parental resources model.
+The descriptive stat table shows that the family resources of Black/White biracials are closer to those of black students than white students.
+So, Black/White biracials are at a disadvantage relative to whites, which is controlled for in the last model.
+
+## Black/Latino results
+
+In the crude model, its clear that there isn't much difference in the grade retention rates of Blacks and Latinos, with Blacks having a slightly higher rate than Latinos.
+The Black/Latino group has rates well below either of these groups.
+In fact, their grade retention rates in the crude model are closer to Whites.
+
+Interestingly, the fixed effects model reverses the ordering of Blacks and Latinos, but this is then reversed again in the cultural resources model.
+In addition the gap between Latinos and Blacks grows a little in the cultural resources model, which is expected.
+In all of these models, the Black/Latino group remains below either of those two groups, although the gap shrinks considerably across models.
+
+Things change quite a bit in the parent resources model.
+The gap between blacks and Latinos grows and the Black/Latino group now looks almost identical to the Latino group.
+So, some of their lower rate was due to better family resources, particularly relative to Latinos, as can be seen on the table of descriptive stats.
+
+## Black/Asian results
 

### 2021-11-18 analysis/analysis.Rmd
Add table of AME for monoracial group differences by model

@@
 
+## AMEs for single race groups
+
+For the paper, I want a table showing the changes across single race groups. Lets go ahead and build that here based on the AMEs.
+
+```{r monoracial-results}
+results_monoracial <- map(list(model0, model1, model2, model3, model4),
+                          function(x) {
+                            marg(x, "race", type="effects", 
+                                 vcov_mat=vcovCL(model, cluster=acs$cluster),
+                                 at_var_interest=c("White","Black","Indigenous",
+                                                   "Asian","Latino"))[[1]][-1,]
+                          })
+
+knitreg(map(results_monoracial, convert_marg),
+        digits=3, caption="Average marginal differences in probability of being behind grade relative to whites",
+        custom.gof.rows=list("State fixed effects" =  c("No","Yes","Yes","Yes","Yes"),
+                             "Year fixed effects" =   c("No","Yes","Yes","Yes","Yes"),
+                             "Grade fixed effects" =  c("No","Yes","Yes","Yes","Yes"),
+                             "Metro fixed effects" =  c("No","Yes","Yes","Yes","Yes"),
+                             "Nativity and language" = c("No","No","Yes","Yes","Yes"),
+                             "Parent's education" =    c("No","No","No","Yes","Yes"),
+                             "Family resources" =      c("No","No","No","No","Yes")))
+
+```
+
+The results indicate that controlling for everything in model 5 reduces the racial disparities substantially.
+Comparing Model 5 to Model 2, the controls cut the effect in half for the Black and Indigenous groups.
+It has less effect and slightly increases the gap for Asians because Asians already had lower rates than Whites.
+The biggest change is for Latinos.
+In model 2, Latinos had the second highest probabilities and were 2% higher than whites.
+After controlling for everything, that actually reverse direction, such that Latinos have slightly lower expected probabilities than whites (.3% less).
+
+The gap is cut in half for Latinos by the nativity and language variables.
+These variables also increase the advantage of Asians as expected.
+they have no effect on the Black and Indigenous effects as expected.
+
+So, I have to consider the shrinking (and in the case of Latinos, flipping direction) of these effects with controls when evaluating the placement of biracials.
+
 # Sensitivity tests

### 2021-11-18 analysis/useful_functions.R

@@ -221,3 +221,15 @@ calculate_marg_means <- function(model, cluster_var=acs$cluster) {
   return(coef_table)
 }
 
+convert_marg <- function(model) {
+  tr <- createTexreg(
+    coef.names = str_remove(model$Label, "race = "), 
+    coef = model$Margin, 
+    se = model$Standard.Error, 
+    pvalues = model$P.Value,
+    gof.names = c("N"), 
+    gof = c(nrow(acs)), 
+    gof.decimal = c(F)
+  )
+}
+

### 2021-11-22 analysis/organize_data.R
Order the current grade factor
I changed the current_grade varable to be an ordered factor to make it easier to separate analysis by grade level. I also changed it back to a simple treatment contrast so it goes into the models in the same way.

@@ -83,7 +83,8 @@ acs <- acs %>%
            gradeattd == 52 ~ "10th",
            gradeattd == 53 ~ "11th",
            gradeattd == 54 ~ "12th"),
-           levels=c("K","1st","2nd","3rd",paste(4:12,"th",sep=""))),
+           levels=c("K","1st","2nd","3rd",paste(4:12,"th",sep="")),
+           ordered=TRUE),
          below_exp_grade=age>=as.numeric(current_grade)+6,
          family_income=ifelse(ftotinc==9999999, NA, 
                               ifelse(ftotinc<0, 0, ftotinc)),
@@ -109,7 +110,7 @@ acs <- acs %>%
   #remove respondents with multiracial parents
   filter(!str_detect(race, "Multiracial"))
 
-#code the contrast matrix for race to adjust for fractions
+#factorize the race variable
 acs$race <- factor(acs$race,
                    levels=c("White","Black","Indigenous","Asian","Latino",
                             "Black/White","Black/Indigenous","Black/Latino",
@@ -123,6 +124,11 @@ acs$race <- factor(acs$race,
                             "White/Indigenous","White/Latino","White/Asian",
                             "Indigenous/Latino","Indigenous/Asian",
                             "Latino/Asian"))
+
+#code the current_grade contrast back to treatment dummies
+x <- contr.treatment(length(levels(acs$current_grade)))
+colnames(x) <- levels(acs$current_grade)[-1]
+contrasts(acs$current_grade) <- x
 
 # Check Yourself Before You Wreck Yourself --------------------------------
 
@@ -147,7 +153,6 @@ table(acs$bpl_pop, acs$foreign_born_father, exclude=NULL)
 table(acs$speakeng, acs$spk_eng, exclude=NULL)
 table(acs$speakeng_mom, acs$spk_eng_mother, exclude=NULL)
 table(acs$speakeng_pop, acs$spk_eng_father, exclude=NULL)
-
 
 ##parents married
 table(acs$marst_mom, acs$marst_pop, acs$parents_married, exclude=NULL)

### 2021-11-22 analysis/analysis.Rmd
Add sensitivity analysis separating models by current grade groups
I group the results by elementary,  middle, and high school

@@
 
-```{r monoracial-results}
+```{r monoracial-results, warning=FALSE}
 results_monoracial <- map(list(model0, model1, model2, model3, model4),
@@
                             marg(x, "race", type="effects", 
-                                 vcov_mat=vcovCL(model, cluster=acs$cluster),
+                                 vcov_mat=vcovCL(x, cluster=acs$cluster),
                                  at_var_interest=c("White","Black","Indigenous",
@@
 
-```{r models-by-age, eval=FALSE}
+## Models by grade
+
+I want to separate the results by age more or less, but I cannot take age directly because then some kids cannot logically be behind grade at certain grade levels.
+For example, if I restrict to 11 or younger, then none of the 5th graders will be behind grade, by design.
+So, I will restrict by current grade level which indirectly gets at age.
+I am going to split it by elementary, middle school, and high school.
+
+```{r models-by-grade, warning=FALSE}
 #run separate models by grade groups
-model4a <- update(model4, data=subset(acs, as.numeric(current_grade)<=6))
-model4b <- update(model4, data=subset(acs, as.numeric(current_grade)>6 &
-                                        as.numeric(current_grade)<=9))
-model4c <- update(model4, data=subset(acs, as.numeric(current_grade)>=10))
+model4_elem <- update(model4, data=subset(acs, current_grade<="5th"))
+model4_middle <- update(model4, data=subset(acs, current_grade>"5th" & 
+                                        current_grade<"9th"))
+model4_hs <- update(model4, data=subset(acs, current_grade>="9th"))
 
-knitreg(list(model4a, model4b, model4c))
+coef_table_elem <- calculate_marg_means(model4_elem, 
+                                        cluster_var=subset(acs, current_grade<="5th")$cluster)
+coef_table_middle <- calculate_marg_means(model4_middle,
+                                          cluster_var=subset(acs, current_grade>"5th" & 
+                                        current_grade<"9th")$cluster)
+coef_table_hs <- calculate_marg_means(model4_hs,
+                                      cluster_var=subset(acs, current_grade>="9th")$cluster)
+
+coef_table <- coef_table_elem %>%
+  mutate(model="elementary")
+coef_table <- coef_table_middle %>%
+  mutate(model="middle school") %>%
+  bind_rows(coef_table)
+coef_table <- coef_table_hs %>%
+  mutate(model="high school") %>%
+  bind_rows(coef_table) %>%
+  mutate(model=factor(model, levels=c("elementary","middle school",
+                                      "high school")),
+         mrace=factor(mrace, levels=c("Black/White",
+                                      "Black/Latino",
+                                      "Black/Asian",
+                                      "Black/Indigenous",
+                                      "White/Indigenous",
+                                      "White/Latino",
+                                      "White/Asian",
+                                      "Latino/Asian",
+                                      "Indigenous/Latino",
+                                      "Indigenous/Asian")))
+
+ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
+                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
+                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
+            color=NA, alpha=0.2)+
+  geom_linerange(data=subset(coef_table, multiracial))+
+  geom_point()+
+  coord_flip()+
+  facet_grid(mrace~model, scales="free_y")+
+  theme_bw()+
+  theme(legend.position = "none", panel.grid = element_blank())+
+  scale_color_manual(values=c("grey40","red"))+
+  scale_fill_manual(values=c("grey40","red"))+
+  scale_linetype_manual(values=c(3,1))+
+  scale_y_continuous(labels = scales::percent)+
+  labs(x=NULL,
+       y="predicted probability of being behind expected grade",
+       title="Probability of being behind grade for biracial respondents and monoracial comparison groups",
+       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
 ```

### 2021-11-22 analysis/analysis.Rmd
Better figure to test coding of below expected grade

@@
 ```{r tab-grade-by-age}
-tbl <- table(acs$age, acs$current_grade)
-
 acs %>% 
@@
 ```{r check-retention-code}
-tapply(acs$below_exp_grade,acs[,c("age","current_grade")], mean)
+acs %>%
+  group_by(age, current_grade) %>%
+  summarize(below=mean(below_exp_grade)) %>%
+  ggplot(aes(x=current_grade, y=age, fill=below))+
+  geom_tile(color="grey20")+
+  geom_text(aes(label=below))+
+  theme_bw()+
+  scale_fill_viridis_c()+
+  scale_y_continuous(breaks=5:20, minor_breaks = NULL)+
+  labs(x="current grade", y="age", fill="proportion below grade level")
 ```
 
-Ok, that looks good. NAs indicate cells with no cases. The 1 values start with 7 year old Kindergarteners and then move down in a perfect diagonal from there. So this looks good.
+Ok, that looks good.
 

### 2021-11-22 analysis/analysis.Rmd
Improve graph showing below expected grade by current grade
Rather than a barplot, doing a line and dot plot with a smooth function and error bars.

@@
   group_by(current_grade) %>%
-  summarize(below_grade=mean(below_exp_grade)) %>%
-  ggplot(aes(x=current_grade, y=below_grade))+
-  geom_col()+
+  summarize(below_grade=mean(below_exp_grade),
+            n=n(),
+            se=sqrt((below_grade*(1-below_grade))/n)) %>%
+  ggplot(aes(x=current_grade, y=below_grade, group=1,
+             ymin=below_grade-1.96*se, ymax=below_grade+1.96*se))+
+  geom_smooth(se=FALSE)+
+  geom_line(linetype=2, color="grey30")+
+  geom_point()+
+  geom_linerange()+
   scale_y_continuous(labels=scales::percent)+
@@
   scale_y_continuous(labels=scales::percent)+
-  scale_color_viridis_d()+
+  scale_color_manual(values=rainbow(14))+
   labs(x="current grade", y="percent clearly behind grade level for age")+

### 2021-11-22 analysis/analysis.Rmd
Add smooth trend line to grade reetention by year

@@
   geom_point()+
+  geom_smooth()+
   scale_x_continuous(breaks=c(2010,2014,2018))+

### 2021-11-22 analysis/analysis.Rmd
Add better variable headers in descriptive stats table

--- 2021-11-22 analysis/analysis.Rmd
@@
   geom_point()+
+  geom_smooth(se=FALSE)+
   scale_x_continuous(breaks=c(2010,2014,2018))+
@@
   mutate_if(is.numeric, format, big.mark=",") %>%
-  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.", align = c("l",rep("r", ncol(tbl_summary)-1))) %>%
+  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.",
+      align = c("l",rep("r", ncol(tbl_summary)-1)),
+      col.names=c("race","n","% below expected grade", "mean age",
+                  "% foreign-born","% speak English well",
+                  "% Both parents speak english well","% either parent foreign-born",
+                  "mean family income","% own home","% either parent, HS diploma",
+                  "% Either parent, BA degree","% parents married")) %>%
   kable_paper() %>%

### 2021-11-22 analysis/analysis.Rmd
Add sex to model to address issue #21

@@
 0.  Raw race differences
-1.  Add fixed effects for state, current grade, year, and metro
+1.  Add fixed effects for state, current grade, year, metro, and sex
 2.  Add kids and parents foreign-born status.
@@
 model0 <- glm(below_exp_grade~race, data=acs, family=binomial)
-model1 <- update(model0, .~.+as.factor(year)+current_grade+as.factor(state)+metro)
+model1 <- update(model0, 
+                 .~.+as.factor(year)+current_grade+as.factor(state)+metro+sex)
 model2 <- update(model1, 

### 2021-11-22 analysis/analysis.Rmd
Add analysis of best way to fit grade-year interaction
I tested multiple models and the best fit was given by a linear year trend interacted with grade. This addresses issue #20.

@@
 
-All models are estimated as logit models. Warning, the code chunks below take a bit of time to run!
+All models are estimated as logit models.
+
+Before running the full models, I want to test out how I can parsimoniously address differences in the time trends by grade.
+
+```{r}
+model_add_cat <- glm(below_exp_grade~as.factor(year)+current_grade, 
+                     data=acs, family=binomial)
+model_add_cont <- glm(below_exp_grade~I(year-2010)+current_grade, 
+                     data=acs, family=binomial)
+model_inter_cont <- glm(below_exp_grade~I(year-2010)*current_grade, 
+                     data=acs, family=binomial)
+model_inter_cat <- glm(below_exp_grade~as.factor(year)*current_grade, 
+                     data=acs, family=binomial)
+
+sapply(list("categorical, additive"=model_add_cat,
+            "categorical, interaction"=model_inter_cat, 
+            "continuous, additive"=model_add_cont, 
+            "continuous, interaction"=model_inter_cont), 
+       BIC) %>%
+    enframe("model","BIC") %>%
+  kbl()
+
+# free some memory
+rm(model_add_cat, model_add_cont, model_inter_cat, model_inter_cont)
+```
+
+The preferred model by BIC is clearly the model that fits year as a continuous measure and interacts it with current grade, so that is what I will do.
+
+Now lets run the full models. Warning, the code chunks below take a bit of time to run!
 
@@
 model1 <- update(model0, 
-                 .~.+as.factor(year)+current_grade+as.factor(state)+metro+sex)
+                 .~.+I(year-2010)*current_grade+as.factor(state)+metro+sex)
 model2 <- update(model1, 

### 2021-11-23 analysis/analysis.Rmd
Add state names
I have replaced the numeric values of statefip with actual state names.

@@
 model1 <- update(model0, 
-                 .~.+I(year-2010)*current_grade+as.factor(state)+metro+sex)
+                 .~.+I(year-2010)*current_grade+state+metro+sex)
 model2 <- update(model1, 

### 2021-11-23 analysis/organize_data.R

@@ -29,7 +29,7 @@ acs <- read_fwf(here("analysis","input","usa_00114.dat.gz"),
                               126,135,142,143,144,145,146,150,154,165,168,175,
                               176,181,184,193,196,205,208,215,216,221,224,231,
                               232,235,236),
-                    col_names = c("year","cluster","state","metro","strata",
+                    col_names = c("year","cluster","statefip","metro","strata",
                                   "ownershp","perwt","momrule","poprule",
                                   "related","sex","age","raced","hispand",
                                   "bpl","speakeng","gradeattd","ftotinc",
@@ -106,7 +106,8 @@ acs <- acs %>%
          sex=factor(sex, levels=1:2, labels=c("Male","Female")),
          metro=factor(metro, levels=0:4, 
                       labels=c("Indeterminable","Non-metro","Central city",
-                               "Metro non-central","Metro indeterminable"))) %>%
+                               "Metro non-central","Metro indeterminable")),
+         state=code_state_names(statefip)) %>%
   #remove respondents with multiracial parents
   filter(!str_detect(race, "Multiracial"))
 
@@ -131,6 +132,9 @@ colnames(x) <- levels(acs$current_grade)[-1]
 contrasts(acs$current_grade) <- x
 
 # Check Yourself Before You Wreck Yourself --------------------------------
+
+## state names
+table(acs$state, acs$statefip, exclude=NULL)
 
 ## race variable
 table(acs$raced_mom, acs$race_mother, exclude=NULL)

### 2021-11-23 analysis/useful_functions.R

@@ -45,6 +45,63 @@ code_speakeng <- function(speakeng) {
   return(spk_eng)
 }
 
+code_state_names <- function(state) {
+  state_name <- case_when(
+    state==1  ~ "Alabama",
+    state==2  ~ "Alaska",
+    state==4  ~ "Arizona",
+    state==5  ~ "Arkansas",
+    state==6  ~ "California",
+    state==8  ~ "Colorado",
+    state==9  ~ "Connecticut",
+    state==10 ~ "Delaware",
+    state==11 ~ "DC",
+    state==12 ~ "Florida",
+    state==13 ~ "Georgia",
+    state==15 ~ "Hawaii",
+    state==16 ~ "Idaho",
+    state==17 ~ "Illinois",
+    state==18 ~ "Indiana",
+    state==19 ~ "Iowa",
+    state==20 ~ "Kansas",
+    state==21 ~ "Kentucky",
+    state==22 ~ "Louisiana",
+    state==23 ~ "Maine",
+    state==24 ~ "Maryland",
+    state==25 ~ "Massachusetts",
+    state==26 ~ "Michigan",
+    state==27 ~ "Minnesota",
+    state==28 ~ "Mississippi",
+    state==29 ~ "Missouri",
+    state==30 ~ "Montana",
+    state==31 ~ "Nebraska",
+    state==32 ~ "Nevada",
+    state==33 ~ "New Hampshire",
+    state==34 ~ "New Jersey",
+    state==35 ~ "New Mexico",
+    state==36 ~ "New York",
+    state==37 ~ "North Carolina",
+    state==38 ~ "North Dakota",
+    state==39 ~ "Ohio",
+    state==40 ~ "Oklahoma",
+    state==41 ~ "Oregon",
+    state==42 ~ "Pennsylvania",
+    state==44 ~ "Rhode Island",
+    state==45 ~ "South Carolina",
+    state==46 ~ "South Dakota",
+    state==47 ~ "Tennessee",
+    state==48 ~ "Texas",
+    state==49 ~ "Utah",
+    state==50 ~ "Vermont",
+    state==51 ~ "Virginia",
+    state==53 ~ "Washington",
+    state==54 ~ "West Virginia",
+    state==55 ~ "Wisconsin",
+    state==56 ~ "Wyoming"
+  )
+  
+}
+
 #change contrasts for ordered factors to stairstep style
 ordered_factor <- function(fact_var) {
   ord_fact <- factor(fact_var, ordered=TRUE)

### 2021-11-23 analysis/organize_data.R
Leave momrule and poprule in analytical data
This will make it possible for me to restrict some models to more accurate links to address issue #5

@@ -243,6 +243,7 @@ acs <- acs %>%
 acs <- acs %>%
   select(cluster, strata, perwt,
          year, state, 
+         momrule, poprule,
          age, current_grade, below_exp_grade, 
          race, sex, metro,
          age_birth_mother, age_birth_father, degree_mother, degree_father, 

### 2021-11-23 analysis/analysis.Rmd
Add code chunk name

@@
 
-```{r}
+```{r models-year-grade}
 model_add_cat <- glm(below_exp_grade~as.factor(year)+current_grade, 

### 2021-11-23 analysis/analysis.Rmd
Add asis results for monoracial outcome table

@@
 
-```{r monoracial-results, warning=FALSE}
+```{r monoracial-results, warning=FALSE, results='asis'}
 results_monoracial <- map(list(model0, model1, model2, model3, model4),

### 2021-11-23 analysis/analysis.Rmd
Add model with strict parent linking

@@
 
-```{r models-by-grade, warning=FALSE}
+```{r models-by-grade, warning=FALSE, fig.height=12}
 #run separate models by grade groups
@@
 ```
+
+## Models with stricter parent linking
+
+```{r models-parent-link, warning=FALSE, fig.height=12}
+#run separate models by grade groups
+model4_best_link <- update(model4, data=subset(acs, momrule==11 & poprule==11))
+
+coef_table_best_link <- calculate_marg_means(model4_best_link, 
+                                        cluster_var=subset(acs, momrule==11 &
+                                                             poprule==11)$cluster)
+
+coef_table <- coef_table_best_link %>%
+  mutate(model="only direct links")
+coef_table <- coef_table_mod4 %>%
+  mutate(model="all links") %>%
+  bind_rows(coef_table) %>%
+  mutate(model=factor(model, levels=c("all links","only direct links")),
+         mrace=factor(mrace, levels=c("Black/White",
+                                      "Black/Latino",
+                                      "Black/Asian",
+                                      "Black/Indigenous",
+                                      "White/Indigenous",
+                                      "White/Latino",
+                                      "White/Asian",
+                                      "Latino/Asian",
+                                      "Indigenous/Latino",
+                                      "Indigenous/Asian")))
+
+ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
+                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
+                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
+            color=NA, alpha=0.2)+
+  geom_linerange(data=subset(coef_table, multiracial))+
+  geom_point()+
+  coord_flip()+
+  facet_grid(mrace~model, scales="free_y")+
+  theme_bw()+
+  theme(legend.position = "none", panel.grid = element_blank())+
+  scale_color_manual(values=c("grey40","red"))+
+  scale_fill_manual(values=c("grey40","red"))+
+  scale_linetype_manual(values=c(3,1))+
+  scale_y_continuous(labels = scales::percent)+
+  labs(x=NULL,
+       y="predicted probability of being behind expected grade",
+       title="Probability of being behind grade for biracial respondents and monoracial comparison groups",
+       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
+```

### 2021-11-23 analysis/analysis.Rmd
Add timestamp to analysis.Rmd
Add summary of models

@@
 ---
+
+```{r time-start}
+timestamp()
+```
 
@@
 
-Ok, let me try to summarize these results, by group.
+## Summarizing the model results
 
-## Black/White results
+Ok, let me first try to summarize these results, by group.
+
+### Black/White results
 
@@
 
-Why does the big change occur in the parental resources model.
-The descriptive stat table shows that the family resources of Black/White biracials are closer to those of black students than white students.
+Why does the big change occur in the parental resources model?
+The descriptive stat table shows that the family resources of Black/White biracials are only slightly better than those of Black students and substantially worse than those of White students.
 So, Black/White biracials are at a disadvantage relative to whites, which is controlled for in the last model.
 
-## Black/Latino results
+### Black/Latino results
 
@@
 The gap between blacks and Latinos grows and the Black/Latino group now looks almost identical to the Latino group.
-So, some of their lower rate was due to better family resources, particularly relative to Latinos, as can be seen on the table of descriptive stats.
+So, some of their lower probability was due to better family resources, particularly relative to Latinos, as can be seen on the table of descriptive stats.
+Once that is controlled for, their outcomes are indistinguishable from Latino students.
 
-## Black/Asian results
+### Black/Asian results
+
+The crude and fixed effects models indicate that Black/Asian students have grade retention rates similar to the lower grade retention rates of Asian students.
+Controlling for cultural resources changes this somewhat with a point estimate that is higher than that of Asian students.
+This makes sense because Black/Asian students are much more likely than Asian students to be native born and speak English well with native-born parents who also speak English well.
+So this gave them and advantage which is removed in that model.
+The family resources model does not change the relative position of Black/Asian students, but does shrink the overall racial gap between Black and Asian students.
+
+In the last two models, the confidence intervals overlap a bit, leading to greater uncertainty in the relatve placement of Black/Asian biracials.
+The point estimate suggests that their grade retention probability is between that of Asians and the half-way point, but the edge of the confidence intervals touch both those bands, so I cannot statistically distinguish their placement relative to those two possibilities.
+In any case, the results strongly rule out them being close to the outcomes of Black students, contra the hypodescent expectation.
+
+### Black/Indigenous results
+
+These results are quite noisy because of the smaller sample sizes involved.
+These are the two monoracial groups with the highest dropout rates, with or without controls.
+However, the difference between them is also quite large due to the very high grade retention probability for Indigenous students.
+This gap shrinks significantly in the model that adjusts for parental resources.
+
+Because of the noise its quite difficult to determine the relative placement of Black/Indigenous students at all, but we can rule out across all models that they have similar grade retentions rates to the higher grade retention of Indigenous students.
+The point estimate suggests a placement somewhere between the lower Black rate and the halfway point, but the wide CIs mean we cannot rule out either of those possibilities.
+
+The point estimate also gets much closer to the lower black rate once we control for family resources.
+
+### White/Indigenous results
+
+The results here consistently show that White/Indigenous students have grade retention rates about midway between the rates of whites and the halfway point.
+So they are clearly in the "lower half" between the groups and closer to White than Indigenous students.
+The overall gap between White and Indigenous students gets smaller with controls, but the relative placement of White/Indigenous students does not change.
+
+###  White/Latino results
+
+Importantly, the ordering of the monoracial groups shifts as controls are added to the model.
+The crude and fixed effets models show that White students are much less likely to be held back relative to Latino students.
+Controlling for cultural resources reduces this gap substantially and the inclusion of family resources flips its so that Latino students have slightly lower grade retention rates than Whites.
+So in the final model, there is not much room to find a midway point for White/Latino biracials.
+The final model shows outcomes that are similar to the lower rates of Latinos, although I need to check this more carefully because the CIs are not visible here relative to the dot (TODO).
+
+In the models prior to the inclusion of family resources, White/Latino biracials had grade retention rates similar to those of Whites, who had the lower rate.
+So despite the switch in ordering between Whites and Latinos, White/Latino biracials tend to look like the group with lower rates.
+
+### White/Asian results
+
+In all of the models prior to the inclusion of family resources, White/Asian students had grade retention rates slightly lower than the rates of Asians - the monoracial group with the lower rates.
+However, once controls are added for family resources, Asian/White students look just like Asian students (the lower group).
+This is because White/Asian students have the greatest family resources of any group in the dataset, even higher than those of White and Asian students.
+
+### Latino/Asian results
+
+This is the group with the biggest outlier in terms of the effect.
+The initial results in the crude model suggested that Latino/Asian students had rates similar to the lower rates of Asian students.
+Controlling for cultural resources shifted this to rates that were exactly midway between the two monoracial groups.
+Controlling for family resources produced a point estimate that was *higher* than either monoracial group although not statistically distinguishable from the higher rates of Latinos.
+Notably, the gap between Asians and Latinos is also quite small in the final model.
+
+The change across models is a result of the high cultural and family resources of Latino/Asian students.
+However, their family resources do not seem particularly advantaged relative to Asian students, so something a bit complex is happening in the final model there.
+They look relatively worse compared to Latinos, but the relative difference doesn't change as much with respect to Asians.
+Essentially, once I control for family resources, they are "leapfrogged" by the Latino group.
+
+### Indigenous/Latino results
+
+Controlling for cultural resources makes a big difference here.
+Initially Indigenous/Latino students look the same as the lower Latino students.
+Controlling for cultural resources makes their outcomes indistinguishably from the midway point, althoug the point estimate is still in the lower half.
+Controlling for family resources has little effect.
+
+Notably, because of sample size the error bars are substantially here.
+
+### Indigenous/Asian results
+
+This is a very small group ($n=521$) so the results are pretty noisy with wide CIs.
+However, the results are consistent across estimates.
+The grade retention rates of Indigenous/Asian students are similar to those of the lower Asian students.
+In the final model, the point estimate is almost identical to those for Asian students.
+
+Overall, I can put these results into a table where I specify the potential placement of each biracial group relative to their monoracial peers. There are seven possible outcomes:
+
+* Lower than both groups
+* Same as lower group
+* Lower half
+* Halfway between
+* Upper half
+* Same as higher
+* Higher than both groups
+
+For the two ends of the distribution and the lower half/upper half, I only consider these cases if it is statistically distinguishable from being in the other categories, respectively.
+For all other cases, I place an X in a cell if the confidence interval indicates this as a possibility.
+Asterisks indicate the placement of the point estimate for cases with uncertainty.
+
+|                   | Lower than both | Same as lower | Lower half | Halfway | Upper half | Same as higher | Higher than both |
+|-------------------|:---------------:|:-------------:|:----------:|:-------:|:----------:|:--------------:|:----------------:|
+| Black/White       |        X        |               |            |         |            |                |                  |
+| Black/Latino      |                 |       X       |            |         |            |                |                  |
+| White/Asian       |                 |       X       |            |         |            |                |                  |
+| White/Latino      |                 |       X       |            |         |            |                |                  |
+| Indigenous/Asian  |                 |       X       |            |         |            |                |                  |
+| Black Asian       |                 |       X       |     X*     |    X    |            |                |                  |
+| Black/Indigenous  |                 |       X       |     X*     |    X    |            |                |                  |
+| Indigenous/White  |                 |               |      X     |         |            |                |                  |
+| Indigenous/Latino |                 |               |            |    X    |            |                |                  |
+| Latino/Asian      |                 |               |            |         |            |        X       |                  |
+
+The results I show here are based on model 4.
+They clearly show that most biracial groups are closer in outcomes to the monoracial group with the lower rates.
+In fact, half the groups have rates that are either indistinguishable from the lower group or lower than *both* groups.
+
+Being a part-black group in no way makes one more likely to suffer from hypodescent.
+In fact, the results for Black/White biracials where we would expect hypodescent to play the largest role are actually the farthest from hypodescent as possible.
+If any group seems to suffer from a pattern of hypodescent to some degree it is part-Indigenous groups and the Latino/Asian group.
+So something else is going on here.
+
+In general, the results show that biracial groups are able to "slip through" the racialized structures that affect their more disadvantaged monoracial groups and this is not just a function of observable differences in class and nativity.
 
@@
 ```
+
+```{r time-end}
+timestamp()
+```

### 2021-11-23 analysis/analysis.Rmd
Add discussion of average adjusted prediction

@@
 
-The model results themselves are not that useful. I want to compare each multiracial group to:
+The model results themselves are not that useful directly. I want to compare each multiracial group to:
 
@@
 
-I can do this by calculating marginal means (or proportions to be more accurate) of the dependent variable across all of the possible race groups.
-In order to calculate these marginal means, I use the `marg` command in the `modmargs` library.
-This gives me results equivalent to average marginal effects but on the level (e.g.
-mean) rather than the effect (different in mean).
-Standard errors for these marginal means are calculated by the [delta method](https://cran.r-project.org/web/packages/modmarg/vignettes/delta-method.html).
-I adjust all SEs for clustering of respondents within households.
+I can do this by calculating the *average adjusted prediction* of the dependent variable across all of the possible race groups.
+The average adjusted prediction (this is the stata-ese language - it is also called variously a predictive margin among other things).
+The adjusted predition in general is just the predicted value of the dependent variable for some set of values on the covariates.
+In this case, I want to predict on the level of the response to get predicted probabilities of being held back.
+The average adjusted prediction essentially gets the adjusted prediction for every observation in the data based on their covariates and then takes the average.
+Essentially, this is equivalent to treating every respondent as if they belonged to group A when calculating the average adjusted prediction for group A.
+The average adjusted prediction is directly related to the *average marginal effect* (AME) in that the AME for a group difference is simply the difference in the average adjusted prediction across groups.
 
-To compare the results for each biracial group to the halfway assumption, I need to calculate the value midway between the two consituent groups and the standard error of that value.
+Its easy to get a simple adjusted prediction at specific values using the `predict` command, but more complicated for the average adjusted prediction.
+In order to calculate these average adjusted predictions, I use the `marg` command in the `modmargs` library.
+This function has a nice `type="levels"` argument that directly gives you the average adjusted prediction.
+Standard errors for these average adjusted predictions are calculated by the [delta method](https://cran.r-project.org/web/packages/modmarg/vignettes/delta-method.html).
+I also adjust all SEs for clustering of respondents within households.
+
+To compare the results for each biracial group to the halfway assumption, I need to calculate the value midway between the two constituent groups and the standard error of that value.
 Calculating the midway value is just a straightforward mean calculation.
@@
 This is not always true and leads in practice to a much lower $\alpha$ for rejection.
-For that to be true, one needs to use 1.386 rather than 1.96 in estimating the confidence interval which produces an 83.4% confidence interval.
-It turns out that 1.386 is equal to $1.96/\sqrt{2}$.
+A much better approximation is given when one uses1.386 rather than 1.96 in estimating the confidence interval, which produces an 83.4% confidence interval.
 The intuition behind this technique is given nicely [here](https://chris-said.io/2014/12/01/independent-t-tests-and-the-83-confidence-interval-a-useful-trick-for-eyeballing-your-data/).
@@
 
-The results here are the same as the raw differences, but I can also fit this graph to any of the models, so I can effectively give conditional probabilities.
+Its important to note that this technique is just an approximation because it depends on the assumption that the standard errors for the two means are the same, which is not going to be true here.
+However, it should give results that are quite close.
+Correcting for unequal SEs is far more complicated because the interval used will vary for any pairwise comparison which I am not doing here.
+
+The results here are the same as the raw differences, but I can also fit this graph to any of the models, using the average adjusted predictions.
 The plot below does this for model 4, by panelizing each multiracial group.
@@
   labs(x=NULL,
-       y="predicted probability of being behind expected grade",
+       y="average adjusted predicted probability of being behind expected grade",
        title="Probability of being behind grade for biracial respondents and monoracial comparison groups",
@@
                              "Family resources" =      c("No","No","No","No","Yes")))
-
 ```

### 2021-11-23 analysis/analysis.Rmd
Restructure headers in analysis.Rmd

@@
 
-# Model building time
+# Model building
 
-Here are the model structure that I will use:
-
-0.  Raw race differences
-1.  Add fixed effects for state, current grade, year, metro, and sex
-2.  Add kids and parents foreign-born status.
-3.  Add mother and father highest degree
-4.  Add family income (square rooted), own your own home, and whether parents are married.
-
-All models are estimated as logit models.
+## Functional form of year-grade interactions
 
@@
 
-Now lets run the full models. Warning, the code chunks below take a bit of time to run!
+## Running the models
+
+Here are the model structure that I will use:
+
+0.  Raw race differences
+1.  Add fixed effects for state, current grade, year, metro, and sex
+2.  Add kids and parents foreign-born status.
+3.  Add mother and father highest degree
+4.  Add family income (square rooted), own your own home, and whether parents are married.
+
+All models are estimated as logit models. Warning, the code chunks below take a bit of time to run!
 
@@
 
-The model results themselves are not that useful directly. I want to compare each multiracial group to:
+## Differences for monoracial groups
+
+The model results here are not very interesting directly because everything is relative to a single reference group.
+However, before proceeding with a technique for understanding the outcomes for biracial respondents, its worthwhile to look more closely at the differences between monoracial groups.
+The table below shows these differences using average marginal effects (AME) with whites as the reference.
+
+```{r monoracial-results, warning=FALSE, results='asis'}
+results_monoracial <- map(list(model0, model1, model2, model3, model4),
+                          function(x) {
+                            marg(x, "race", type="effects", 
+                                 vcov_mat=vcovCL(x, cluster=acs$cluster),
+                                 at_var_interest=c("White","Black","Indigenous",
+                                                   "Asian","Latino"))[[1]][-1,]
+                          })
+
+knitreg(map(results_monoracial, convert_marg),
+        digits=3, caption="Average marginal differences in probability of being behind grade relative to whites",
+        custom.gof.rows=list("State fixed effects" =  c("No","Yes","Yes","Yes","Yes"),
+                             "Year fixed effects" =   c("No","Yes","Yes","Yes","Yes"),
+                             "Grade fixed effects" =  c("No","Yes","Yes","Yes","Yes"),
+                             "Metro fixed effects" =  c("No","Yes","Yes","Yes","Yes"),
+                             "Nativity and language" = c("No","No","Yes","Yes","Yes"),
+                             "Parent's education" =    c("No","No","No","Yes","Yes"),
+                             "Family resources" =      c("No","No","No","No","Yes")))
+```
+
+The results indicate that controlling for everything in Model 5 reduces the racial disparities substantially.
+Comparing Model 5 to Model 2, the controls cut the effect in half for the Black and Indigenous groups.
+It has less effect and slightly increases the gap for Asians because Asians already had lower rates than Whites.
+The biggest change is for Latinos.
+In model 2, Latinos had the second highest probabilities and were 2% higher than whites.
+After controlling for everything, that actually reverse direction, such that Latinos have slightly lower expected probabilities than whites (.3% less).
+
+The gap is cut in half for Latinos by the nativity and language variables.
+These variables also increase the advantage of Asians as expected.
+they have no effect on the Black and Indigenous effects as expected.
+
+So, I have to consider the shrinking (and in the case of Latinos, flipping direction) of these effects with controls when evaluating the placement of biracials.
+
+## Average adjusted preditions for biracials
+
+I want to compare each multiracial group to:
 
@@
 
-## AMEs for single race groups
+## Sensitivity tests
 
-For the paper, I want a table showing the changes across single race groups. Lets go ahead and build that here based on the AMEs.
-
-```{r monoracial-results, warning=FALSE, results='asis'}
-results_monoracial <- map(list(model0, model1, model2, model3, model4),
-                          function(x) {
-                            marg(x, "race", type="effects", 
-                                 vcov_mat=vcovCL(x, cluster=acs$cluster),
-                                 at_var_interest=c("White","Black","Indigenous",
-                                                   "Asian","Latino"))[[1]][-1,]
-                          })
-
-knitreg(map(results_monoracial, convert_marg),
-        digits=3, caption="Average marginal differences in probability of being behind grade relative to whites",
-        custom.gof.rows=list("State fixed effects" =  c("No","Yes","Yes","Yes","Yes"),
-                             "Year fixed effects" =   c("No","Yes","Yes","Yes","Yes"),
-                             "Grade fixed effects" =  c("No","Yes","Yes","Yes","Yes"),
-                             "Metro fixed effects" =  c("No","Yes","Yes","Yes","Yes"),
-                             "Nativity and language" = c("No","No","Yes","Yes","Yes"),
-                             "Parent's education" =    c("No","No","No","Yes","Yes"),
-                             "Family resources" =      c("No","No","No","No","Yes")))
-```
-
-The results indicate that controlling for everything in model 5 reduces the racial disparities substantially.
-Comparing Model 5 to Model 2, the controls cut the effect in half for the Black and Indigenous groups.
-It has less effect and slightly increases the gap for Asians because Asians already had lower rates than Whites.
-The biggest change is for Latinos.
-In model 2, Latinos had the second highest probabilities and were 2% higher than whites.
-After controlling for everything, that actually reverse direction, such that Latinos have slightly lower expected probabilities than whites (.3% less).
-
-The gap is cut in half for Latinos by the nativity and language variables.
-These variables also increase the advantage of Asians as expected.
-they have no effect on the Black and Indigenous effects as expected.
-
-So, I have to consider the shrinking (and in the case of Latinos, flipping direction) of these effects with controls when evaluating the placement of biracials.
-
-# Sensitivity tests
-
-## Models by grade
+### Models by grade
 
@@
 
-## Models with stricter parent linking
+### Models with stricter parent linking
 

### 2021-11-23 analysis/analysis.Rmd
Add description of grade specific model
This closes issue #23

@@
 
+The results get quite a bit noisier here because we are cutting the sample size substantially in each panel.
+The general pattern is that the results tend to look pretty similar across grade levels.
+What variations we do see could easily be the result of statistical noise as they tend to move around the most in cases with high error bands.
+The two notable cases where something might be going on:
+
+* For White/Black students, grade retention is clearly in the lower-half between Whites and the midway point rather than below whites for high school students.
+Notably, this is also where the white/black grows the largest.
+So the advantage that these students have relative to everyone else only holds for elementary and middle school.
+Nonetheless, their outcomes are still much closer to White students than Black students.
+* For White/Indigenous students, the placement varies across panels.
+In elementary school, these students are indistinguishable from White students.
+They are perfectly in the middle in middle school, and then in the lower half in high school.
+So it seems that their lowest outcomes happen earlier and on average you end up with a "lower half" across all grades.
+
+In no case do the results here suggest a biasing effect of non-biological kids.
+We would expect that to be greater in higher grades, but the results are generally consistent across grades.
+In cases where they are not consistent, you would expect the results to move toward the larger population group with age, but I do not see that.
+
 ### Models with stricter parent linking

### 2021-11-24 analysis/analysis.Rmd
Add discussion of results using only parental links
Closes #5
Add models predicting family and cultural resources
Addresses issue #23, but I need to think a bit more about parents' foreign-born status before I close it.

@@
 
+The results here are extraordinarily stable across both specifications.
+
+# Differences in Cultural and Family Resources
+
+I want a parsimonious way to visualize differences in cultural and family resources.
+I can sort of do it by looking at the big descriptive statistics table, but it can be a little hard to summarize.
+Doing each possible variable separately would be extremely tedious here, so I am going to create scores across variables for these two measures and then model them as outcomes.
+I can then plot the results just like I do for my actual outcome measure.
+
+I will start by just looking at the $\alpha$ score for the correlation matrix of each case.
+I then create scores scaled to have a mean of zero and a standard deviation of one and fit these with a linear model using the same structure as for the models above.
+
+## Cultural Resources
+
+```{r cult-resource-alpha}
+cult_resources <- cbind(as.numeric(!acs$foreign_born), 
+                        as.numeric(!acs$foreign_born_father), 
+                        as.numeric(!acs$foreign_born_mother),
+                        as.numeric(factor(acs$spk_eng, 
+                                          levels=c("No","Somewhat","Yes"))),
+                        as.numeric(factor(acs$spk_eng_father, 
+                                          levels=c("No","Somewhat","Yes"))),
+                        as.numeric(factor(acs$spk_eng_mother, 
+                                          levels=c("No","Somewhat","Yes"))))
+
+alpha(cor(cult_resources))
+```
+
+That all holds together well.
+However, I run into a problem when I think about making a score out of this because the foreign-born status of parents actually helps to reduce the likelihood of grade retention while all of the other variables increase it.
+For now, I am just reverse coding it, but this leads to results that are far from intuitive.
+I think the simple score approach might not work well in this case.
+
+```{r cult-resource-score}
+#scale them all, add them up, and then scale again
+#one catch though, I want to reverse score foreign born mother and father
+#because these reduce the likelihood of grade retention, other things
+#controlled
+acs$cult_resources <- scale(scale(as.numeric(!acs$foreign_born))+
+                              scale(as.numeric(acs$foreign_born_father))+
+                              scale(as.numeric(acs$foreign_born_mother))+
+                              scale(as.numeric(factor(acs$spk_eng, 
+                                          levels=c("No","Somewhat","Yes"))))+
+                              scale(as.numeric(factor(acs$spk_eng_father, 
+                                          levels=c("No","Somewhat","Yes"))))+
+                              scale(as.numeric(factor(acs$spk_eng_mother, 
+                                          levels=c("No","Somewhat","Yes")))))
+```
+
+```{r cult-resource-model, warning=FALSE}
+#If I use the glm framework then I should be able to feed this back into
+#the marg command
+model_cultr <- glm(cult_resources~race+I(year-2010)*current_grade+state+metro+sex+
+                    +degree_mother+degree_father+sqrt(family_income/10000)+
+                     own_home+parents_married, 
+                  data=acs, family="gaussian")
+coef_table_cultr <- calculate_marg_means(model_cultr)
+
+ggplot(coef_table_cultr, aes(x=term, y=estimate, color=multiracial,
+                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
+                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
+            color=NA, alpha=0.2)+
+  geom_linerange(data=subset(coef_table_cultr, multiracial))+
+  geom_point()+
+  coord_flip()+
+  facet_wrap(~mrace, scales="free_y")+
+  theme_bw()+
+  theme(legend.position = "none", panel.grid = element_blank())+
+  scale_color_manual(values=c("grey40","red"))+
+  scale_fill_manual(values=c("grey40","red"))+
+  scale_linetype_manual(values=c(3,1))+
+  labs(x=NULL,
+       y="average adjusted cultural resource score (standard deviation units)",
+       title="Average cultural resources for biracial respondents and monoracial comparison groups",
+       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
+```
+
+Because of the issue with foreign-born parents, these results are not particularly intuitive.
+It tends to show that part-Asian and part-Latino respondents have high cultural resources, but generally not higher than the highest of their constituent groups.
+Other groups have results half-way in between.
+
+## Family Resources
+
+```{r fam-resource-score, warning=FALSE}
+fam_resources <- cbind(acs$family_income, as.numeric(acs$degree_father), 
+                       as.numeric(acs$degree_mother), as.numeric(acs$own_home))
+
+alpha(cor(fam_resources))
+
+#scale them all, add them up, and then scale again
+acs$fam_resources <- scale(scale(acs$family_income)+scale(acs$own_home)+
+  scale(as.numeric(acs$degree_father)+scale(as.numeric(acs$degree_mother))))
+```
+
+```{r fam-resource-model}
+#If I use the glm framework then I should be able to feed this back into
+#the marg command
+model_famr <- glm(fam_resources~race+I(year-2010)*current_grade+state+metro+sex+
+                    foreign_born+foreign_born_mother+foreign_born_father+
+                    spk_eng+spk_eng_mother+spk_eng_father, 
+                  data=acs, family="gaussian")
+coef_table_famr <- calculate_marg_means(model_famr)
+
+ggplot(coef_table_famr, aes(x=term, y=estimate, color=multiracial,
+                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
+                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
+            color=NA, alpha=0.2)+
+  geom_linerange(data=subset(coef_table_famr, multiracial))+
+  geom_point()+
+  coord_flip()+
+  facet_wrap(~mrace, scales="free_y")+
+  theme_bw()+
+  theme(legend.position = "none", panel.grid = element_blank())+
+  scale_color_manual(values=c("grey40","red"))+
+  scale_fill_manual(values=c("grey40","red"))+
+  scale_linetype_manual(values=c(3,1))+
+  labs(x=NULL,
+       y="average adjusted family resource score (standard deviation units)",
+       title="Average family resources for biracial respondents and monoracial comparison groups",
+       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
+```
+
+In most cases, family resources are roughly midway between those of the two monoracial groups. However, there are a couple of notable exceptions:
+
+* The average family resources of Black/White students are only marginally better than those of Black students which are far below those of the average white students.
+* The White/Asian group has family resources higher then either of the two constituent groups which already have the highest resources of any monoracial group.
+The White/Asian group has the highest family resources of any group here.
+* The Indigenous/Latino and Black/Latino groups have family resources slightly lower than either of hte constituent groups, but in both cases there are hardly any differences between the monoracial groups themselves.
+So you would not expect controlling for these variables to change much for these groups.
+
 ```{r time-end}

### 2021-11-24 analysis/check_packages.R

@@ -7,7 +7,7 @@
 #add new packages to the chain here
 packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr","broom",
              "ggrepel","purrr", "kableExtra", "tibble","scales","janitor",
-             "survey","modmarg","sandwich")
+             "survey","modmarg","sandwich","psych")
 
 package.check <- lapply(packages, FUN = function(x) {
   if (!require(x, character.only = TRUE)) {

### 2021-11-24 analysis/analysis.Rmd
Switch analysis.Rmd to robobook html output
I am using the rmdformats library to output results using the robobook format. This gives me nicer formatting and a dynamic table of contents to get around

@@
 output: 
-  html_document: 
+  rmdformats::robobook:
+    highlight: tango
+    toc_depth: 2
     fig_height: 6
     fig_width: 9
-    toc: yes
-    toc_depth: 4
 ---
@@
     enframe("model","BIC") %>%
-  kbl()
+  kbl() %>%
+  kable_styling(full_width=FALSE)
 

### 2021-11-24 analysis/check_packages.R

@@ -7,7 +7,7 @@
 #add new packages to the chain here
 packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr","broom",
              "ggrepel","purrr", "kableExtra", "tibble","scales","janitor",
-             "survey","modmarg","sandwich","psych")
+             "survey","modmarg","sandwich","psych","rmdformats")
 
 package.check <- lapply(packages, FUN = function(x) {
   if (!require(x, character.only = TRUE)) {

### 2021-11-24 analysis/analysis.Rmd
Add captions
Now that I am using the robobook format, I should be able to add nice captions to my output.

@@
 
-```{r check-retention-code}
+```{r check-retention-code, caption="The proportion of students measured at below grade level by age and current grade.
+This graph is used to ensure that the dependent variable is calculated correctly."}
 acs %>%
@@
 
-```{r retention-by-grade}
+```{r retention-by-grade, caption="Percent of students clearly behind grade level by current grade. Blue lines fit a smoothed curve to the data."}
 acs %>%
@@
 
-```{r retention-by-year}
+```{r retention-by-year, caption="Trends over time in the percent of students clearly behind grade level. Blue lines fit a smoothed curve to the data."}
 acs %>%
@@
 
-```{r retention-by-year-grade}
+```{r retention-by-year-grade, caption="Trends over time and grade in the percent of students clearly behind grade level. Lines are calculated by smoothing."}
 acs %>%
@@
   ggplot(aes(x=year, y=below_grade, group=current_grade, color=current_grade))+
-  geom_line()+
+  #geom_line()+
+  geom_smooth(se=FALSE)+
   geom_point()+
@@
   scale_y_continuous(labels=scales::percent)+
-  scale_color_manual(values=rainbow(14))+
   labs(x="current grade", y="percent clearly behind grade level for age")+
@@
 knitreg(map(results_monoracial, convert_marg),
+        caption="Differences in the probability of being behind grade by race based on various statistical models. All estimates are average marginal effects.",
         digits=3, caption="Average marginal differences in probability of being behind grade relative to whites",
@@
 
-```{r example-plot}
+```{r example-plot, caption="Probability of being behind grade for biracial black/white respondents in comparison to their monoracial comparison groups.
+83.4% confidence bands shown around each estimate.
+Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level"}
 g <- coef_table_mod0 %>%
@@
   labs(x=NULL,
-       y="predicted probability of being behind expected grade",
-       title="Probability of being behind grade for biracial black/white respondents and monoracial\ncomparison groups",
-       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
+       y="predicted probability of being behind expected grade")
 
@@
 
-```{r cond-means-model4, warning=FALSE}
+```{r cond-means-model4, warning=FALSE, caption="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
+83.4% confidence bands shown around each estimate.
+Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
+Estimates are based on models that adjust for state, year, current grade, metropolitan status, gender, foreign born status, parent's foreign born status, English proficieny, parent's English proficiency, parent's education, family income, home ownership, and family structure."}
 coef_table_mod4 <- calculate_marg_means(model4)
@@
   labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade",
-       title="Probability of being behind grade for biracial respondents and monoracial comparison groups",
-       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
+       y="average adjusted predicted probability of being behind expected grade")
 ```
@@
 
-```{r cond-means-model-compare, fig.height=12, warning=FALSE}
+```{r cond-means-model-compare, fig.height=12, warning=FALSE, caption="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
+83.4% confidence bands shown around each estimate.
+Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
+Each panel is based on a different model that accounts for certain variables.
+Subsequent panels cumulatively include all prior terms.}
 #calculate remaining marginal means
@@
 
-```{r models-by-grade, warning=FALSE, fig.height=12}
+```{r models-by-grade, warning=FALSE, fig.height=12, caption="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
+83.4% confidence bands shown around each estimate.
+Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
+Panels compare results when data are restricted to certain grade ranges."}
 #run separate models by grade groups
@@
 
-```{r models-parent-link, warning=FALSE, fig.height=12}
+```{r models-parent-link, warning=FALSE, fig.height=12, caption="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
+83.4% confidence bands shown around each estimate.
+Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
+Panels compare results depending on whether results are restricted only to children linked to a head of household or not."}
 #run separate models by grade groups
@@
 
-```{r cult-resource-model, warning=FALSE}
+```{r cult-resource-model, warning=FALSE, caption="Average cultural resources for biracial respondents in comparison to their monoracial comparison groups.
+83.4% confidence bands shown around each estimate.
+Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
+Cultural resources are determined by a summated score of standardized native-born status, parents' native-born status, English proficieny, and parents' English proficiency.
+Parents' native-born status is reverse coded based on model results which show it increases grade retention risk."}
 #If I use the glm framework then I should be able to feed this back into
@@
   labs(x=NULL,
-       y="average adjusted cultural resource score (standard deviation units)",
-       title="Average cultural resources for biracial respondents and monoracial comparison groups",
-       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
+       y="average adjusted cultural resource score (standard deviation units)")
 ```
@@
 
-```{r fam-resource-model}
+```{r fam-resource-model, warning=FALSE, caption="Average family resources for biracial respondents in comparison to their monoracial comparison groups.
+83.4% confidence bands shown around each estimate.
+Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
+Family resources are determined by a summated score of standardized family income, parental education, and home ownership."}
 #If I use the glm framework then I should be able to feed this back into
@@
   labs(x=NULL,
-       y="average adjusted family resource score (standard deviation units)",
-       title="Average family resources for biracial respondents and monoracial comparison groups",
-       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
+       y="average adjusted family resource score (standard deviation units)")
 ```

### 2021-11-24 analysis/analysis.Rmd
Change title of analylsis.Rmd
Insert missing quotation mark

@@
 ---
-title: "Analysis for Project"
+title: "Analysis for Biracial Grade Retention Project"
 output: 
@@
 Each panel is based on a different model that accounts for certain variables.
-Subsequent panels cumulatively include all prior terms.}
+Subsequent panels cumulatively include all prior terms."}
 #calculate remaining marginal means

### 2021-11-24 analysis/analysis.Rmd
Fix double caption

@@
 knitreg(map(results_monoracial, convert_marg),
-        caption="Differences in the probability of being behind grade by race based on various statistical models. All estimates are average marginal effects.",
         digits=3, caption="Average marginal differences in probability of being behind grade relative to whites",

### 2021-11-24 analysis/analysis.Rmd
Change caption to fig.cap

@@
 
-```{r check-retention-code, caption="The proportion of students measured at below grade level by age and current grade.
+```{r check-retention-code, fig.cap="The proportion of students measured at below grade level by age and current grade.
 This graph is used to ensure that the dependent variable is calculated correctly."}
@@
 
-```{r retention-by-grade, caption="Percent of students clearly behind grade level by current grade. Blue lines fit a smoothed curve to the data."}
+```{r retention-by-grade, fig.cap="Percent of students clearly behind grade level by current grade. Blue lines fit a smoothed curve to the data."}
 acs %>%
@@
 
-```{r retention-by-year, caption="Trends over time in the percent of students clearly behind grade level. Blue lines fit a smoothed curve to the data."}
+```{r retention-by-year, fig.cap="Trends over time in the percent of students clearly behind grade level. Blue lines fit a smoothed curve to the data."}
 acs %>%
@@
 
-```{r retention-by-year-grade, caption="Trends over time and grade in the percent of students clearly behind grade level. Lines are calculated by smoothing."}
+```{r retention-by-year-grade, fig.cap="Trends over time and grade in the percent of students clearly behind grade level. Lines are calculated by smoothing."}
 acs %>%
@@
 
-```{r example-plot, caption="Probability of being behind grade for biracial black/white respondents in comparison to their monoracial comparison groups.
+```{r example-plot, fig.cap="Probability of being behind grade for biracial black/white respondents in comparison to their monoracial comparison groups.
 83.4% confidence bands shown around each estimate.
@@
 
-```{r cond-means-model4, warning=FALSE, caption="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
+```{r cond-means-model4, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
 83.4% confidence bands shown around each estimate.
@@
 
-```{r cond-means-model-compare, fig.height=12, warning=FALSE, caption="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
+```{r cond-means-model-compare, fig.height=12, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
 83.4% confidence bands shown around each estimate.
@@
 
-```{r models-by-grade, warning=FALSE, fig.height=12, caption="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
+```{r models-by-grade, warning=FALSE, fig.height=12, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
 83.4% confidence bands shown around each estimate.
@@
 
-```{r models-parent-link, warning=FALSE, fig.height=12, caption="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
+```{r models-parent-link, warning=FALSE, fig.height=12, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
 83.4% confidence bands shown around each estimate.
@@
 
-```{r cult-resource-model, warning=FALSE, caption="Average cultural resources for biracial respondents in comparison to their monoracial comparison groups.
+```{r cult-resource-model, warning=FALSE, fig.cap="Average cultural resources for biracial respondents in comparison to their monoracial comparison groups.
 83.4% confidence bands shown around each estimate.
@@
 
-```{r fam-resource-model, warning=FALSE, caption="Average family resources for biracial respondents in comparison to their monoracial comparison groups.
+```{r fam-resource-model, warning=FALSE, fig.cap="Average family resources for biracial respondents in comparison to their monoracial comparison groups.
 83.4% confidence bands shown around each estimate.

### 2021-11-24 analysis/analysis.Rmd
Add code folding to analysis.Rmd output

@@
     fig_width: 9
+    code_folding: hide
 ---

### 2021-11-24 analysis/analysis.Rmd
Add lightbox support to analysis.Rmd output

--- 2021-11-24 analysis/analysis.Rmd
@@
     fig_width: 9
+    code_folding: hide
+    lightbox: TRUE
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, eval=TRUE)
 library(here)
@@
 
-## Average adjusted preditions for biracials
+## Average adjusted predictions for biracials
 

### 2022-03-30 analysis/analysis.Rmd
Save some output for use in paper

@@
                              "Family resources" =      c("No","No","No","No","Yes")))
+
+save(results_monoracial, file=here("analysis","output","monoracial_results.RData"))
 ```
@@
        y="average adjusted predicted probability of being behind expected grade")
+
+save(coef_table, file=here("analysis","output","coef_table.RData"))
 ```

### 2022-03-31 products/paper/main.Rmd
Start to paper
Just a few outlines and tables and figures for now, but its a start!

---
title: "The Title"
author:
- affiliation: University of Oregon, Sociology
  name: Aaron Gullickson
keywords: keywords
#FYI: the thanks and abstract are easier to view and edit in Atom which does proper linewrapping
thanks: Thanks to people and stuff
abstract: This is a test abstract
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    template: ./resources/aog-latex-ms.tex
  word_document:
    reference_docx: ./resources/aog_word_style.docx
fontfamily: mathpazo
fontsize: 11pt
anonymous: false #anonymize, double-space, and ragged right
endnotes: false #make footnotes endnotes
endfloat: false #move table and figures to the end
bibliography: ../project.bib
biblio-style: ./resources/ajs.bst
---

```{r setup, echo=FALSE, warning=FALSE, error=FALSE, include=FALSE}
# If you want to do things in one R Markdown document then use the 
# chunk option eval to toggle between running with and without 
# compiling R code chunks
knitr::opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE, 
                      warning = FALSE, eval = TRUE)
library(here)
source(here("analysis","check_packages.R"))
source(here("analysis","useful_functions.R"))
```

```{r load-data}
load(here("analysis","output","acs.RData"))
load(here("analysis","output","coef_table.RData"))
load(here("analysis","output","monoracial_results.RData"))
```

# Introduction

# Background

# Data and Methods

-   Discuss ACS data source. Only includes children living in a household with two parents who are likely biological parents.
-   How are multiracial children identified?
-   How is grade retention determined?
-   Model building - basic model is a logit model of grade retention. All models include:
    -   state fixed effects
    -   linear year effect
    -   grade fixed effect
    -   grade * linear year fixed effect
    -   location of student (metro, suburb, rural) fixed effects
-   All models also account for occassional clustering of multiple respondents in the same household.
-   I then test nested models that include other family resources that may affect grade retention:
    -   nativity (own and each parent) and english proficiency (own and each parent)
    -   parents' education, measured by highest degree of each parent
    -   other family resources: family income (square rooted), whether they own their own home, and whether parents are married.
-   Because the research question is about where biracials fit relative to their monoracial comparison groups, no coding of reference categories will produce a complete results from model output. Instead, I use the model to calculate average adjusted predicted probabilities of grade retention for each racial (mono and bi) group. Explain what an average adjusted predicted probability is (like an AME but on the level rather than the difference).
-   Given these predicted probabilities, I then use a graphical technique to display most results. Go through example using the raw probabilities for white, black, and white/black respondents.
-   Show the figure for raw model 0 data for white/black

```{r example-plot, fig.cap="Probability of being behind grade for biracial black/white respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level"}
g <-  coef_table %>%
  filter(mrace=="Black/White" & model=="crude") %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial & mrace=="Black/White" &
                               model=="crude"))+
  geom_point()+
  coord_flip()+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade")

g+annotate("text", 1, 0.035, hjust="left",
             label="Grey bands give 83.4%\nconfidence interval around\nmonoracial estimates")+
  annotate("text", 2, 0.045, hjust="left",
             label="Red band gives 83.4%\nconfidence interval around\nmidpoint estimate")+
  annotate("text", 2.5, 0.036, hjust="left",
             label="Actual biracial estimate\nshown with 83.4% confidence\ninterval")
```

# Results

-   Show table of results for monoracial respondents across different models

```{r monoracial-results, results='asis'}
knitreg(map(results_monoracial, convert_marg),
        digits=3, caption="Average marginal differences in the probability of being behind grade by monoracial group, relative to a monoracial white student",
        caption.above=TRUE,
        custom.gof.rows=
          list("State fixed effects" = c("Yes","Yes","Yes","Yes"),
               "Grade fixed effects" = c("Yes","Yes","Yes","Yes"),
               "Year linear effects" = c("Yes","Yes","Yes","Yes"),
               "Year x Grade effects"  = c("Yes","Yes","Yes","Yes"),
               "Location fixed effects" = c("Yes","Yes","Yes","Yes"),
               "Nativity and language" = c("No","Yes","Yes","Yes"),
               "Parent's education" = c("No","No","Yes","Yes"),
               "Family resources" = c("No","No","No","Yes")))
```

-   Show panelized figures for each group from fixed effect model to full model.

```{r cond-means-model-compare, fig.height=9, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Each panel is based on a different model that accounts for certain variables. Subsequent panels cumulatively include all prior terms."}
coef_table %>%
  filter(model=="fixed effects" | model=="parent resources") %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial 
                             & (model=="fixed effects" | 
                                  model=="parent resources")))+
  geom_point()+
  coord_flip()+
  facet_grid(mrace~model, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="average adjusted predicted probability of being behind expected grade")
```

# Conclusions

### 2022-04-04 products/paper/main.qmd
Move paper to quarto document and write rought draft of data and methods section

---
title: Grade Retention and Mixed Race
author: Aaron Gullickson
abstract: This is a test abstract
thanks: Thanks to people and stuff
format:
  pdf:
    documentclass: scrartcl
    papersize: letter
    fig-height: 4
    fig-pos: htb
    #template: resources/aog-latex-ms.tex
bibliography: ../project.bib
execute: 
  eval: true
  echo: false
  warning: false
  error: false
editor: source
---

```{r setup}
#| include: false
library(here)
source(here("analysis","check_packages.R"))
source(here("analysis","useful_functions.R"))
```

```{r load-data}
load(here("analysis","output","acs.RData"))
load(here("analysis","output","coef_table.RData"))
load(here("analysis","output","monoracial_results.RData"))
```

# Introduction

# Background

# Data and Methods

Data for this analysis come from the American Community Survey (ACS), an annual 1-in-100 survey of the US population, produced by the Census Bureau. To increase sample size for smaller populations of biracial respondents, I pool ACS data for a full decade from 2010-2019. While pooling data helps to increase sample size, it may generate biases if the likelihood of grade retention is changing over time in complex way. I address this issue in the modeling techniques discussed below. All data were extracted from the Minnesota Population Center's Integrated Public Use Microdata Series (CITATION).

Ideally, I would restrict the sample only to children living in a household with two biological parents. However, the ACS data only records the family relationship between the head of household and all other members of the family. Since 2008, the ACS also distinguishes between biological children, adopted children, and stepchildren. Respondents can therefore be identified as biological children of one member of the household if that member is the head of household or the biological child of the head of household. However, identifying whether the partner/spouse of that biological parent is also the biological parent rather than a stepparent is more difficult and cannot be determined perfectly.

Prior work has used a variety of additional restrictions to limit the analysis to children who are more likely to be the biological children of both parents (CITE Saenz and Xie & Goyette). Some of those procedures rely upon an assumption that most parents would be married that was more tenable in historical but less tenable in contemporary data. Therefore, I do not explicitly restrict by marital status, I do restrict cases in two other ways. First, I restrict the sample to cases where both parents were of a reasonable age at the birth of the child (aged 15-44 for mothers and aged 15-60 for fathers). Second, I restrict the sample to only those cases where the reported race and Hispanicity of the child is consistent with at least one of the parents.

The final analytical sample consists of children aged 5-20 who are currently enrolled in the K-12 school system and live in a household with two parents who are both likely to be biological parents. This necessary sample restriction will bias the family structure of the analytical sample relative to the total population of K-12 schoolchildren, limiting it exclusively to households with two parents present and overrepresenting the proportion of cases in which those parents are married. The result of this bias is well-known. Children living in two-parent nuclear households have greater educational success than other children as a result of material, social, and cultural resources (CITATIONS). Therefore, the overall estimate of grade retention will be biased upward. However, the goal of this study is to estimate differences between groups which will not necessarily be affected by this bias, unless the differences across racial groups in other family forms are different.

The race of children is calculated by a cross-tabulation of parent's races. A full cross-tabulation of the 17 (CHECK) categories of race in the ACS combined with the 4 (CHECK) categories of Hispanicity would produce far too many combined categories to be analytically useful. Therefore, I collapse each parent's race into the categories of White, Black, Asian, Latino, and Indigenous. These categories form the "ethnoracial pentagon" that is commonly used in popular practice and government tabulation to identify race in the US (CITE Hollinger). Respondents are identified as Latino if they responded with a Latino identitification to the Hispanicity question, regardless of their response to the race question. Respondents are identified as Indigenous if they responded as either American Indian/Alaska Native or as Pacific Islander. Because the goal of my analysis is to examine specifically the outcomes of "first-generation" biracial children, I exclude cases where at least one of the parents identified with multiple races.

the cross-tabulation of parent's races leads to ten distinct biracial categories that respondents may fall into. The total sample size for each of these biracial categories is shown in @fig-sample-size. The sample sizes are quite large for many of the groups, suggesting substantial statistical power in determining the relative outcomes for these groups. Even among the smallest group of Indigenous/Asian, the sample size is substantially greater than one would see in typical survey data ($n=519$).

```{r}
#| label: fig-sample-size
#| fig-cap: Sample size of biracial respondents combined across American Community Survey data.
acs %>%
  filter(str_detect(race, "/")) %>%
  ggplot(aes(x=reorder(race, race, length)))+
  geom_bar()+
  coord_flip()+
  labs(x=NULL, y="number of respondents")+
  scale_y_continuous(labels = scales::comma)+
  theme_bw()
```

To measure grade retention, I compare the age of respondents to their current grade in school. Because the ACS lacks data on the exact birthday of respondents, it is not possible to exactly identify students who are below the expected grade level. All grades include two expected age groups and a student of a given age could always reasonable be expected to be in two different grades. For example, a first grader is expected to be either six or seven years of age, while a second grader is expected to be either seven or eight years of age. Therefore, with information only on current grade and age in years, it is impossible to determine whether a seven year old in first grade has been held back or is at the expected grade level.

Instead of trying to identify exact grade retention, I instead identify cases in which students are clearly behind grade expectations. This is identified when the age of the respondent is greater than either of the possible ages for a student of that grade (e.g. a first grader who is eight years old). This measure will underestimate the total proportion of students that have been held back. However, the goal of the analysis is to examine differences across groups rather than total levels. The only factor determining this bias is whether the birthday of the student takes place earlier or later in the year, and is unlikely to vary substantially across groups.[^1]

[^1]: Biracial respondents are slightly younger than monoracial respondents in the sample as a whole because interracial unions are becoming more common over time HOW STRONG COULD THE BIAS BE HERE AND WHAT CAN I DO ABOUT IT?

The analysis is driven by a set of logit models that predict the likelihood of a student being clearly behind grade. For the baseline model, I consider several characteristics that may vary across groups and thus need to be controlled in all models. First, states dictates educational policy and thus can differ substantially in the likelihood of grade retention so I include fixed effects for state of residence in all analysis. I cannot identify specific school districts, but I also include dummy variables indicating whether the student lived in a central city, suburban, or rural area.[^2]

[^2]: The location cannot be determined for all students and I therefore also include a fourth category of unknown in all models.

The probability of being clearly behind grade also increases with grade and so I include fixed effects for the current grade of the respondent. Finally, there may be changes over time in the probability of being clearly behind grade. In general, educational reforms have recently decreased the frequency of grade retention (CITE). As @fig-year-grade shows, the percent of students clearly behind grade level has declined over time, but this decline has been more substantial at higher grade levels. For the elementary grades, there is no evidence of a decline at all.

```{r}
#| label: fig-year-grade
#| fig-cap: Trends over time and grade in the percent of students clearly behind grade level. Lines are fit to each set of points by grade via lowess smoothing.

acs %>%
  group_by(current_grade, year) %>%
  summarize(below_grade=mean(below_exp_grade)) %>%
  ggplot(aes(x=year, y=below_grade, group=current_grade, color=current_grade))+
  #geom_line()+
  geom_smooth(se=FALSE)+
  geom_point()+
  scale_x_continuous(breaks=c(2010,2014,2018))+
  scale_y_continuous(labels=scales::percent)+
  labs(x="year", y="percent clearly behind grade level for age",
       color="current grade")+
  theme_bw()
```

To account for this grade-specific decline in the models I include an interaction between a linear year term and current grade. Sensitivity analysis showed that this functional form was preferred by BIC to models with no interaction and a model with interaction terms that treated year as a categorical variable. (SHOULD I BE CONSIDERING THE SAME FOR STATE?)

The baseline model also includes the race of the respondent. I then add additional terms measuring material and cultural resources that may account for racial differences in a set of subsequent models. First, I include measures of nativity and English proficiency for both the respondent and each of their parents. Second, I include a categorical measure of highest degree earned for each parent. Finally, I include other measures of family income (square rooted), home ownership, and whether the parents are married. All models account for the clustering of multiple respondents within the same household.

Output from the full models is available in the supplementary materials provided with this article. However, this output is not particularly useful for directly evaluating the relative placement of biracial students compared to their monoracial peers (e.g. a White/Black student compared to White and Black students). Throughout this article, I instead use a visual approach to illustrate this placement.

To illustrate via an example, I take the raw percent of students clearly behind grade for White, Black, and Black/White students. These percents are 3.42%, 5.30%, and 3.64%. I then display these probabilities in @fig-example. This figure clearly shows the placement of Black/White students (shown in red) relative to White and Black students. I show uncertainty for all three cases with 83.4% confidence intervals. I use 83.4% confidence intervals because overlap on such intervals indicates that the difference between the two measures is statistically distinguishable from zero at $p<.05$ (CITE). This is not true of overlap with 95% confidence intervals. The lack of overlap of the Black/White probability with either the White or Black bands indicates that it is statistically distinguishable from both monoracial probabilities.

```{r example-plot}
#| label: fig-example
#| fig-cap: Probability of being behind grade for biracial black/white respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
g <-  coef_table %>%
  filter(mrace=="Black/White" & model=="crude") %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial & mrace=="Black/White" &
                               model=="crude"))+
  geom_point()+
  coord_flip()+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade")

g+annotate("text", 1, 0.035, hjust="left",
             label="Grey bands give 83.4%\nconfidence interval around\nmonoracial estimates")+
  annotate("text", 2, 0.045, hjust="left",
             label="Red band gives 83.4%\nconfidence interval around\nmidpoint estimate")+
  annotate("text", 2.5, 0.036, hjust="left",
             label="Actual biracial estimate\nshown with 83.4% confidence\ninterval")
```

Additionally, I include a measure of the "halfway" point between the two racial groups.^[The standard error of this halfway estimate used to construct the confidence interval is given by $\sqrt{(s_1^2+s_2^2)/2}$.] This allows me to additionally determine where the biracial group falls relative to the expecation of being halfway between the two constituent monoracial groups. In this case, we can see that Black/White students are much closer to White students in their risk of grade retention, and have grade retention rates much lower than bpth Black students and the halfway expectation.

@fig-example shows the raw differences between groups. To create this same figure from model results that account for potential confounders, I calculate average predicted probabilities (APP). Given a logit model, APPs account for control variables within a model to estimate the average predicted probability of the outcome for a given category. APPs are akin to average marginal effects (AME) and estimated in the same way (the delta method). Whereas AMEs estimate differences or slopes, APPs estimate the level for a given category.

# Results

-   Show table of results for monoracial respondents across different models

```{r monoracial-results}
#| results: asis
#| label: tbl-monoracial
#| tbl-cap: Average marginal differences in the probability of being behind grade by monoracial group, relative to a monoracial white student
texreg(map(results_monoracial, convert_marg),
       digits=3,
       caption.above=TRUE,
       custom.gof.rows=
         list("State fixed effects" = c("Yes","Yes","Yes","Yes"),
              "Grade fixed effects" = c("Yes","Yes","Yes","Yes"),
              "Year linear effects" = c("Yes","Yes","Yes","Yes"),
              "Year x Grade effects"  = c("Yes","Yes","Yes","Yes"),
              "Location fixed effects" = c("Yes","Yes","Yes","Yes"),
              "Nativity and language" = c("No","Yes","Yes","Yes"),
              "Parent's education" = c("No","No","Yes","Yes"),
              "Family resources" = c("No","No","No","Yes")))
```

-   Show panelized figures for each group from fixed effect model to full model.

```{r cond-means-model-compare}
#| label: fig-full-models
#| fig-cap: Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Each panel is based on a different model that accounts for certain variables. Subsequent panels cumulatively include all prior terms.
#| fig-height: 9
coef_table %>%
  filter(model=="fixed effects" | model=="parent resources") %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial 
                             & (model=="fixed effects" | 
                                  model=="parent resources")))+
  geom_point()+
  coord_flip()+
  facet_grid(mrace~model, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="average adjusted predicted probability of being behind expected grade")
```

### 2022-04-04 products/paper/main.qmd
Replace figure of sample sizes with descriptive stats table

@@
 the cross-tabulation of parent's races leads to ten distinct biracial categories that respondents may fall into.
-The total sample size for each of these biracial categories is shown in @fig-sample-size.
+The total sample size for each of these biracial categories is shown in @tbl-descriptive, along with several descriptive statistics.
 The sample sizes are quite large for many of the groups, suggesting substantial statistical power in determining the relative outcomes for these groups.
-Even among the smallest group of Indigenous/Asian, the sample size is substantially greater than one would see in typical survey data ($n=519$).
+Even among the smallest group of Indigenous/Asian, the sample size is substantially greater than one would see in typical survey data.
 
 ```{r}
+#| label: tbl-descriptive
+#| tbl-cap: Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.
+tbl_summary <- acs %>%
+  group_by(race) %>%
+  summarize(n=n(), 
+            held_back=percent(mean(below_exp_grade), accuracy=0.01),
+            mean_age=round(mean(age), 1),
+            f_born=percent(mean(foreign_born),  accuracy=0.1),
+            par_fb=percent(mean(foreign_born_father | 
+                                  foreign_born_mother), accuracy=0.1),
+            fam_inc=dollar(round(mean(family_income), 0)),
+            own_home=percent(mean(own_home), accuracy=0.1),
+            par_ba=percent(mean(degree_father=="BA" | 
+                                  degree_father=="G" | 
+                                  degree_mother=="BA" | 
+                                  degree_mother=="G"), accuracy=0.1)
+  ) %>%
+  arrange(desc(n)) %>%
+  mutate(n=comma(n))
+
+index_biracial <- which(str_detect(tbl_summary$race, "/"))
+
+tbl_summary %>%
+  kbl(booktabs = TRUE,
+      linesep = NULL,
+      align = c("l",rep("r", ncol(tbl_summary)-1)),
+      col.names=c("race",
+                  "n",
+                  "clearly below expected grade", 
+                  "mean age",
+                  "foreign-born",
+                  "either parent foreign-born",
+                  "mean family income",
+                  "own home",
+                  "either parent, college degree"
+      )) %>%
+  column_spec(3:10, width = "1.6cm") %>%
+  row_spec(index_biracial, background = "#D3D3D3") %>%
+  landscape()
+```
+
+```{r}
+#| eval: false
 #| label: fig-sample-size
@@
 Sensitivity analysis showed that this functional form was preferred by BIC to models with no interaction and a model with interaction terms that treated year as a categorical variable.
-(SHOULD I BE CONSIDERING THE SAME FOR STATE?)
+(==SHOULD I BE CONSIDERING THE SAME FOR STATE?==)
 

### 2022-04-04 analysis/check_packages.R

@@ -7,7 +7,7 @@
 #add new packages to the chain here
 packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr","broom",
              "ggrepel","purrr", "kableExtra", "tibble","scales","janitor",
-             "survey","modmarg","sandwich","psych","rmdformats")
+             "survey","modmarg","sandwich","psych","rmdformats","scales")
 
 package.check <- lapply(packages, FUN = function(x) {
   if (!require(x, character.only = TRUE)) {

### 2022-04-04 products/paper/main.qmd
Fix caption on texreg

@@
        digits=3,
+       caption = "Average marginal differences in the probability of being clearly behind grade by monoracial group, relative to a monoracial white student",
        caption.above=TRUE,
@@
          list("State fixed effects" = c("Yes","Yes","Yes","Yes"),
+              "Location fixed effects" = c("Yes","Yes","Yes","Yes"),
               "Grade fixed effects" = c("Yes","Yes","Yes","Yes"),
               "Year linear effects" = c("Yes","Yes","Yes","Yes"),
-              "Year x Grade effects"  = c("Yes","Yes","Yes","Yes"),
-              "Location fixed effects" = c("Yes","Yes","Yes","Yes"),
+              "Year x grade effects"  = c("Yes","Yes","Yes","Yes"),
               "Nativity and language" = c("No","Yes","Yes","Yes"),

### 2022-04-04 products/paper/main.qmd
Split model of results into two separate models

@@
 load(here("analysis","output","monoracial_results.RData"))
+# change names of models for display
+coef_table <- coef_table %>%
+  mutate(model=factor(model,
+                      levels=c("crude","fixed effects","cultural resources",
+                               "parent resources"),
+                      labels=c("Raw","Baseline model","Cultural model",
+                               "Full model")))
 ```
@@
 g <-  coef_table %>%
-  filter(mrace=="Black/White" & model=="crude") %>%
+  filter(mrace=="Black/White" & model=="Raw") %>%
   ggplot(aes(x=term, y=estimate, color=multiracial,
@@
   geom_linerange(data=subset(coef_table, multiracial & mrace=="Black/White" &
-                               model=="crude"))+
+                               model=="Raw"))+
   geom_point()+
@@
 
-```{r cond-means-model-compare}
-#| label: fig-full-models
-#| fig-cap: Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
+```{r fig-black-models}
+#| label: fig-black-models
+#| fig-cap: Probability of being behind grade for biracial respondents with one black parent, in comparison to their monoracial comparison groups.
 83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
-Each panel is based on a different model that accounts for certain variables.
-Subsequent panels cumulatively include all prior terms.
-#| fig-height: 9
+Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level, Baseline model includes year, grade, location, and state fixed effects.
+Full models include control variables for nativity, English proficiency, income, education, home ownership, and marital status of parents.
+#| fig-height: 7.5
 coef_table %>%
-  filter(model=="fixed effects" | model=="parent resources") %>%
+  filter(model=="Baseline model" | model=="Full model") %>%
+  filter(str_detect(mrace, "Black")) %>%
   ggplot(aes(x=term, y=estimate, color=multiracial,
@@
   geom_linerange(data=subset(coef_table, multiracial 
-                             & (model=="fixed effects" | 
-                                  model=="parent resources")))+
+                             & (model=="Baseline model" | 
+                                  model=="Full model") & 
+                               str_detect(mrace, "Black")))+
   geom_point()+
@@
   labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
+       y="average predicted probability of being clearly behind expected grade")
 ```
+
+```{r fig-other-models}
+#| label: fig-other-models
+#| fig-cap: Probability of being behind grade for non-black biracial respondents, in comparison to their monoracial comparison groups.
+83.4% confidence bands shown around each estimate.
+Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level, Baseline model includes year, grade, location, and state fixed effects.
+Full models include control variables for nativity, English proficiency, income, education, home ownership, and marital status of parents.
+#| fig-height: 7.5
+coef_table %>%
+  filter(model=="Baseline model" | model=="Full model") %>%
+  filter(!str_detect(mrace, "Black")) %>%
+  ggplot(aes(x=term, y=estimate, color=multiracial,
+             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
+                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
+            color=NA, alpha=0.2)+
+  geom_linerange(data=subset(coef_table, multiracial 
+                             & (model=="Baseline model" | 
+                                  model=="Full model") & 
+                               !str_detect(mrace, "Black")))+
+  geom_point()+
+  coord_flip()+
+  facet_grid(mrace~model, scales="free_y")+
+  theme_bw()+
+  theme(legend.position = "none", panel.grid = element_blank())+
+  scale_color_manual(values=c("grey40","red"))+
+  scale_fill_manual(values=c("grey40","red"))+
+  scale_linetype_manual(values=c(3,1))+
+  scale_y_continuous(labels = scales::percent)+
+  labs(x=NULL,
+       y="average predicted probability of being clearly behind expected grade")
+```

### 2022-04-06 analysis/analysis.Rmd
Add model testing state*year interaction
It was not preferred to my existing set up.

@@
 ```{r models-year-grade}
-model_add_cat <- glm(below_exp_grade~as.factor(year)+current_grade, 
+model_add_cat <- glm(below_exp_grade~as.factor(year)+current_grade+state, 
                      data=acs, family=binomial)
-model_add_cont <- glm(below_exp_grade~I(year-2010)+current_grade, 
+model_add_cont <- glm(below_exp_grade~I(year-2010)+current_grade+state, 
                      data=acs, family=binomial)
-model_inter_cont <- glm(below_exp_grade~I(year-2010)*current_grade, 
+model_inter_cont <- glm(below_exp_grade~I(year-2010)*current_grade+state, 
                      data=acs, family=binomial)
-model_inter_cat <- glm(below_exp_grade~as.factor(year)*current_grade, 
+model_inter_cat <- glm(below_exp_grade~as.factor(year)*current_grade+state, 
+                     data=acs, family=binomial)
+model_inter_state <- glm(below_exp_grade~I(year-2010)*current_grade+
+                            I(year-2010)*state, 
                      data=acs, family=binomial)
@@
             "continuous, additive"=model_add_cont, 
-            "continuous, interaction"=model_inter_cont), 
+            "continuous, interaction"=model_inter_cont,
+            "add state*year continuous interaction"=model_inter_state),
        BIC) %>%
@@
 # free some memory
-rm(model_add_cat, model_add_cont, model_inter_cat, model_inter_cont)
+rm(model_add_cat, model_add_cont, model_inter_cat, model_inter_cont,
+   model_inter_state)
 ```

### 2022-04-06 products/paper/main.qmd
Add rough results section

@@
 title: Grade Retention and Mixed Race
-author: Aaron Gullickson
+author: Aaron Gullickson, *University of Oregon*
 abstract: This is a test abstract
@@
     fig-pos: htb
+    keep-tex: false
     #template: resources/aog-latex-ms.tex
@@
 
+==ACKNOWLEDGE POTENTIAL ISSUE WITH DROPOUT STUDENTS==
+
 The analysis is driven by a set of logit models that predict the likelihood of a student being clearly behind grade.
@@
             color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial & mrace=="Black/White" &
+  geom_linerange(data=subset(coef_table, multiracial & 
+                               mrace=="Black/White" &
                                model=="Raw"))+
@@
 
-Additionally, I include a measure of the "halfway" point between the two racial groups.^[The standard error of this halfway estimate used to construct the confidence interval is given by $\sqrt{(s_1^2+s_2^2)/2}$.] This allows me to additionally determine where the biracial group falls relative to the expecation of being halfway between the two constituent monoracial groups.
+Additionally, I include a measure of the "halfway" point between the two racial groups.[^3] This allows me to additionally determine where the biracial group falls relative to the expecation of being halfway between the two constituent monoracial groups.
 In this case, we can see that Black/White students are much closer to White students in their risk of grade retention, and have grade retention rates much lower than bpth Black students and the halfway expectation.
+
+[^3]: The standard error of this halfway estimate used to construct the confidence interval is given by $\sqrt{(s_1^2+s_2^2)/2}$.
 
@@
 
--   Show table of results for monoracial respondents across different models
+I begin by showing the relative risk of grade retention among monoracial respondents across models.
+@tbl-monoracial reports average marginal effects on the probability of grade retention for each monoracial minority group in comparison to White students.
+Understanding these differences and how they change across models helps me to understand the potential for how biracial respondents might be positioned.
+
+Model 1 only controls for structural factors such as state of residence, urbanicity, grade, and year.
+Thus, these results provide a good estimate of the differences across monoracial groups without controlling for differences in cultural and material resources across groups.
+The results show that Black, Indigenous, and Latino students all have substantially higher probability of grade retention than White students.
+Indigenous students have substantially higher risk than all other students, with a probability of grade retention that is 4.1% percentage points higher than White students.
+Asian students have the lowest probability of grade retention, and their risk is slightly lower than Whites.
+
+The subsequent models control for a variety of additional variables.
+Model 2 includes controls for whether the student and each of their parents are native-born and speak English well.
+Model 3 controls for the highest degree received for each parent, and Model 4 controls for family income (square rooted), home ownership, and whether the student's parents are married.
+
+In general Model 2 and Model 3 have substantial effects on the observed inequality across racial groups, while the additional variables from Model 4 have less impact.
+Controlling for the characteristics related to immigration in Model 2 cuts the gap between White and Latino students in half and increases the gap between Asian and White students.
+It has no impact on the gap between White and Black or White and Indigenous students because most of these students are not first or second generation.
+
+Controlling for education in Model 3 reduces the gap substantially for Black and Indigenous students and removes the gap entirely for Latino students.
+In fact, the results of both Models 3 and 4, suggest that, when holding constant cultural and family resources, Latino students are slightly less likely than White students to be held back.
+
+SUMMARIZE THIS?
 
@@
        caption.above=TRUE,
+       label="tbl-monoracial",  #TODO: This does not work
+       float.pos="htb",
        custom.gof.rows=
@@
 
--   Show panelized figures for each group from fixed effect model to full model.
+I now turn to the results for biracial respondents.
+Partially to present results in a more easily digestible way, I split these results into two separate figures.
+In @fig-black-models, I show the placement of each biracial group involving one black parent in comparison to their two monoracial constituent groups.
+To allow comparison across models, I compare the results from the baseline model (Model 1 in @tbl-monoracial) and the full model that accounts for all cultural and family resources (Model 4 in @tbl-monoracial).
 
@@
 #| label: fig-black-models
+#| fig-pos: p
 #| fig-cap: Probability of being behind grade for biracial respondents with one black parent, in comparison to their monoracial comparison groups.
@@
 
+I focus first on biracial groups with one Black parent because of a stronger historical expectation that these students will be identified as Black by the "one-drop rule" and as a result will have similar outcomes to Black students.
+In three cases, this hypothesis can be soundly rejected.
+For Black/White, Black/Latino, and Black/Asian students, The probability of being clearly behind expected grade level is substantially lower than that for monoracial black students and the difference is statistically distinguishable from zero.
+The only case where this is not true is for Black/Indigenous students.
+However this case is also notable for the fact that monoracial Black students have lower risk of grade retention than monoracial Indigenous students.
+In all four cases, these biracial students have a risk of grade retention that is closer to the lower risk monoracial group.
+
+The results also suggest that each of these biracial groups has a risk of grade retention lower than the halfway point, although this hypothesis cannot be rejected completely for the Black/Asian and Black/Indigenous groups, due to the greater degree of uncertainty in the point estimates for these groups.
+
+Controlling for resource variables has some effect on the placement of Black/White and Black/Latino biracial respondents.
+In the case of Black/White biracial respondents, Black/White students initially have a risk of grade retention slightly higher than and statistically distinguishable from White student.
+After controlling for cultural and family resources, however, Black/White respondents have a risk of grade retention that is slightly lower and statistically distinguishable from White students.
+This change reflects the relatively low level of parental education and income among Black/White students, which is much closer to Black students than White students, as shown on @tbl-descriptive.
+
+An opposite pattern is observed for Black/Latino respondents.
+Black/Latino respondents initially have a risk of grade retention substantially lower than both Black and Latino students, who have similar risks.
+However, after controlling for cultural and family resources, Black/Latino students have a risk of grade retention that is indistinguishable from the grade retention risk of Latino students and substantially lower than Black students.
+The change largely reflects the fact that Black/Latino students have greater cultural resources than Latino students and once this advantage is accounted for, they have a similar risk.
+
+Altogether, the results here strongly suggest that the risk of grade retention faced by these biracial students with one black parent is generally similar to that of their constituent monoracial group with the lower risk of grade retention.
+This finding is inconsistent both with the expectation of the one-drop rule and the "in-between" hypothesis.
+I now turn to the experiences of the remaining biracial groups.
+
 ```{r fig-other-models}
 #| label: fig-other-models
+#| fig-pos: p
 #| fig-cap: Probability of being behind grade for non-black biracial respondents, in comparison to their monoracial comparison groups.
@@
 ```
+
+@fig-other-models shows the results for these remaining biracial groups.
+In general, the results here are consistent with those for part-Black biracial respondents, with some important caveats discussed below.
+In general, the point estimates suggest that the risk of each biracial group is either quite close to the monoracial group with the lower risk or in-between this group's risk and the halfway point.
+
+Any results for part-Latino students are complicated by the substantial change in the monoracial Latino risk across models.
+For example, while Latinos have substantially higher risk of grade retention than Whites in the baseline model, there is only a very small difference favoring Latinos in the full model.
+Nonetheless, even in this case, we see that White/Latinos have risk closer to those of the lower Latino group, but distinguishable from both groups.
+
+For all groups, the full models generally indicate that each biracial group has a risk of grade retention similar to the risk of the lower monoracial constituent group or has risk higher than this group but less than the halfway point.
+There are two exceptions to this general pattern.
+First, for the Indigenous/Latino group, I cannot rule the possibility that this group has risk roughly halfway between the two constituent groups, although the point estimate favors a result closer to the lower Latino group.
+Second, for Latino/Asian students, I observe the only case where the risk for the biracial group is higher than either monoracial group.
+However, the confidence bands suggest that the risk for the Latino/Asian group cannot be differentiated from the rates for Latinos, the higher of the two monoracial constituent groups.
+This result marks the one substantial deviation from the general pattern.
+It only emerges in the models that control for cultural and family resources because Latino/Asian students come from higher income and more educated households than Latino students, on average.
+
+Aside from the effect on Latino/Asian students, controlling for cultural and family resources has moderate effects on the placement of the remaining biracial groups.
+The only case with a substantial shift is that for White/Asian students who initially appear to have lower risk than either White or Asian students.
+After controlling for these resources, the risks of White/Asian students are indistinguishable from the lower risk of Asian students.
+This is due to the fact that White/Asian students come from the most advantaged households of any group in the dataset.
+
+## Conclusions

### 2022-04-18 products/paper/main.qmd
Re-organize discussion of example figure in methods section
I now illustrate the figure using the raw percents from Table 1 prior to discussing the models which I think is a more natural flow

@@
 For example, a first grader is expected to be either six or seven years of age, while a second grader is expected to be either seven or eight years of age.
-Therefore, with information only on current grade and age in years, it is impossible to determine whether a seven year old in first grade has been held back or is at the expected grade level.
+Therefore, with information only on current grade and age in years, it is impossible to determine whether a seven year old in first grade is behind the expected grade level.
 
@@
 
+@tbl-descriptive shows the percent of each racial group that is clearly behind the expected grade level.
+My goal is understand the level of grade retention for each biracial group relative to its constituent monoracial groups (e.g.
+a Black/White student compared to White and Black students).
+Throughout this article, I use a visual approach to illustrate this placement.
+
+As an example, I display the percent of students clearly behind grade for White, Black, and Black/White students in @fig-example.
+The point estimate and confidence interval for the Black/White group is shown in red, while the two monoracial groups are shown via grey dots and confidence bands.
+For all three cases, I use 83.4% confidence intervals, because overlap on such intervals indicates that the observed difference between the two measures in the sample is not statistically distinguishable from zero in the population at $p<.05$ (CITE).[^2] In this case, the lack of overlap between the Black/White probability and either the White or Black band indicates that the Black/White probability of clearly being behind grade level is statistically distinguishable from both monoracial probabilities.
+
+[^2]: While 95% confidence intervals are more common in practice, the overlap of such intervals does not necessarily indicate a lack of statistical significance
+
+```{r example-plot}
+#| label: fig-example
+#| fig-cap: Probability of being behind grade for biracial black/white respondents in comparison to their monoracial comparison groups.
+83.4% confidence bands shown around each estimate.
+Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
+g <-  coef_table %>%
+  filter(mrace=="Black/White" & model=="Raw") %>%
+  ggplot(aes(x=term, y=estimate, color=multiracial,
+             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
+                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
+            color=NA, alpha=0.2)+
+  geom_linerange(data=subset(coef_table, multiracial & 
+                               mrace=="Black/White" &
+                               model=="Raw"))+
+  geom_point()+
+  coord_flip()+
+  theme_bw()+
+  theme(legend.position = "none", panel.grid = element_blank())+
+  scale_color_manual(values=c("grey40","red"))+
+  scale_fill_manual(values=c("grey40","red"))+
+  scale_linetype_manual(values=c(3,1))+
+  scale_y_continuous(labels = scales::percent)+
+  labs(x=NULL,
+       y="predicted probability of being behind expected grade")
+
+g+annotate("text", 1, 0.035, hjust="left",
+             label="Grey bands give 83.4%\nconfidence interval around\nmonoracial estimates")+
+  annotate("text", 2, 0.045, hjust="left",
+             label="Red band gives 83.4%\nconfidence interval around\nmidpoint estimate")+
+  annotate("text", 2.5, 0.036, hjust="left",
+             label="Actual biracial estimate\nshown with 83.4% confidence\ninterval")
+```
+
+Additionally, I include a measure of the "halfway" point between the two racial groups.[^3] This halfway point allows me to determine where the biracial group falls relative to the expecation of being halfway between the two constituent monoracial groups.
+In this case, Black/White students are much closer to White students in their risk of grade retention, and have grade retention rates much lower than both Black students and the halfway expectation.
+
+[^3]: The standard error of this halfway estimate used to construct the confidence interval is given by $\sqrt{(s_1^2+s_2^2)/2}$ where $s_1$ and $s_2$ are the standard errors of the estimates for the two monoracial groups.
+
+@fig-example
+
 The analysis is driven by a set of logit models that predict the likelihood of a student being clearly behind grade.
@@
 First, states dictates educational policy and thus can differ substantially in the likelihood of grade retention so I include fixed effects for state of residence in all analysis.
-I cannot identify specific school districts, but I also include dummy variables indicating whether the student lived in a central city, suburban, or rural area.[^2]
+I cannot identify specific school districts, but I also include dummy variables indicating whether the student lived in a central city, suburban, or rural area.[^4]
 
-[^2]: The location cannot be determined for all students and I therefore also include a fourth category of unknown in all models.
+[^4]: The location cannot be determined for all students and I therefore also include a fourth category of unknown in all models.
 
@@
 Throughout this article, I instead use a visual approach to illustrate this placement.
-
-To illustrate via an example, I take the raw percent of students clearly behind grade for White, Black, and Black/White students.
-These percents are 3.42%, 5.30%, and 3.64%.
-I then display these probabilities in @fig-example.
-This figure clearly shows the placement of Black/White students (shown in red) relative to White and Black students.
-I show uncertainty for all three cases with 83.4% confidence intervals.
-I use 83.4% confidence intervals because overlap on such intervals indicates that the difference between the two measures is statistically distinguishable from zero at $p<.05$ (CITE).
-This is not true of overlap with 95% confidence intervals.
-The lack of overlap of the Black/White probability with either the White or Black bands indicates that it is statistically distinguishable from both monoracial probabilities.
-
-```{r example-plot}
-#| label: fig-example
-#| fig-cap: Probability of being behind grade for biracial black/white respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
-g <-  coef_table %>%
-  filter(mrace=="Black/White" & model=="Raw") %>%
-  ggplot(aes(x=term, y=estimate, color=multiracial,
-             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial & 
-                               mrace=="Black/White" &
-                               model=="Raw"))+
-  geom_point()+
-  coord_flip()+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="predicted probability of being behind expected grade")
-
-g+annotate("text", 1, 0.035, hjust="left",
-             label="Grey bands give 83.4%\nconfidence interval around\nmonoracial estimates")+
-  annotate("text", 2, 0.045, hjust="left",
-             label="Red band gives 83.4%\nconfidence interval around\nmidpoint estimate")+
-  annotate("text", 2.5, 0.036, hjust="left",
-             label="Actual biracial estimate\nshown with 83.4% confidence\ninterval")
-```
-
-Additionally, I include a measure of the "halfway" point between the two racial groups.[^3] This allows me to additionally determine where the biracial group falls relative to the expecation of being halfway between the two constituent monoracial groups.
-In this case, we can see that Black/White students are much closer to White students in their risk of grade retention, and have grade retention rates much lower than bpth Black students and the halfway expectation.
-
-[^3]: The standard error of this halfway estimate used to construct the confidence interval is given by $\sqrt{(s_1^2+s_2^2)/2}$.
 

### 2022-04-18 products/paper/main.qmd
Edit of methods section

@@
 To increase sample size for smaller populations of biracial respondents, I pool ACS data for a full decade from 2010-2019.
-While pooling data helps to increase sample size, it may generate biases if the likelihood of grade retention is changing over time in complex way.
+While pooling data helps to increase sample size, it may generate biases if the likelihood of grade retention is changing over time in complex ways.
 I address this issue in the modeling techniques discussed below.
-All data were extracted from the Minnesota Population Center's Integrated Public Use Microdata Series (CITATION).
+All data were extracted from the Minnesota Population Center's Integrated Public Use Microdata Series [@ruggles_ipums_2020].
 
 Ideally, I would restrict the sample only to children living in a household with two biological parents.
-However, the ACS data only records the family relationship between the head of household and all other members of the family.
+However, the ACS only records the direct family relationship between the head of household andother members of the family.
 Since 2008, the ACS also distinguishes between biological children, adopted children, and stepchildren.
-Respondents can therefore be identified as biological children of one member of the household if that member is the head of household or the biological child of the head of household.
-However, identifying whether the partner/spouse of that biological parent is also the biological parent rather than a stepparent is more difficult and cannot be determined perfectly.
+Respondents can therefore be identified as biological children of another member of the household only if that member is the head of household or the biological child of the head of household.
+Determining whether the partner/spouse of that biological parent is the other biological parent is more difficult.
 
-Prior work has used a variety of additional restrictions to limit the analysis to children who are more likely to be the biological children of both parents (CITE Saenz and Xie & Goyette).
-Some of those procedures rely upon an assumption that most parents would be married that was more tenable in historical but less tenable in contemporary data.
-Therefore, I do not explicitly restrict by marital status, I do restrict cases in two other ways.
+Prior work has used a variety of additional restrictions to limit the analysis to children who are more likely to be the biological children of both parents [@saenz_persistence_1995; @xie_racial_1997].
+Some of those procedures rely upon an assumption that most parents would be married that was more tenable in historical data than contemporary data.
+Therefore, I do not explicitly restrict by marital status, I do restrict cases in two other ways that are consistent with prior work.
 First, I restrict the sample to cases where both parents were of a reasonable age at the birth of the child (aged 15-44 for mothers and aged 15-60 for fathers).
@@
 The final analytical sample consists of children aged 5-20 who are currently enrolled in the K-12 school system and live in a household with two parents who are both likely to be biological parents.
-This necessary sample restriction will bias the family structure of the analytical sample relative to the total population of K-12 schoolchildren, limiting it exclusively to households with two parents present and overrepresenting the proportion of cases in which those parents are married.
+This necessary sample restriction will bias the family structure of the analytical sample relative to the total population of K-12 schoolchildren, limiting it exclusively to households with two parents present and over-representing the proportion of cases in which those parents are married.
 The result of this bias is well-known.
 Children living in two-parent nuclear households have greater educational success than other children as a result of material, social, and cultural resources (CITATIONS).
-Therefore, the overall estimate of grade retention will be biased upward.
-However, the goal of this study is to estimate differences between groups which will not necessarily be affected by this bias, unless the differences across racial groups in other family forms are different.
+Therefore, this sample restriction will upwardly bias estimates of grade retention.
+However, this bias is less problematic for the goal of this study, which is to estimate differences between groups.
+The sample restriction effectively eliminates differences in family structure across racial groups which may account for some observed differences in grade retention.
+However, the remaining differences observed here are likely to hold for the full population in the absence of strong racial differences in the effect of family structure on grade retention.
 
 The race of children is calculated by a cross-tabulation of parent's races.
-A full cross-tabulation of the 17 (CHECK) categories of race in the ACS combined with the 4 (CHECK) categories of Hispanicity would produce far too many combined categories to be analytically useful.
+A full cross-tabulation of the race/Hispanicity of each parent would produce far too many combined categories to be analytically useful.
 Therefore, I collapse each parent's race into the categories of White, Black, Asian, Latino, and Indigenous.
-These categories form the "ethnoracial pentagon" that is commonly used in popular practice and government tabulation to identify race in the US (CITE Hollinger).
-Respondents are identified as Latino if they responded with a Latino identitification to the Hispanicity question, regardless of their response to the race question.
+These categories form the "ethnoracial pentagon" that is commonly used in popular practice and government tabulation to identify race in the US [@hollinger_postethnic_1995].
+Respondents are identified as Latino base on their response to the Hispanicity question, regardless of their response to the race question.
 Respondents are identified as Indigenous if they responded as either American Indian/Alaska Native or as Pacific Islander.
-Because the goal of my analysis is to examine specifically the outcomes of "first-generation" biracial children, I exclude cases where at least one of the parents identified with multiple races.
+Because the goal of my analysis is to examine specifically the outcomes of "first-generation" biracial children, I exclude cases where at least one parent identified with multiple races.
 
-the cross-tabulation of parent's races leads to ten distinct biracial categories that respondents may fall into.
+The cross-tabulation of parent's races leads to ten distinct biracial categories.
 The total sample size for each of these biracial categories is shown in @tbl-descriptive, along with several descriptive statistics.
@@
       )) %>%
-  column_spec(3:10, width = "1.6cm") %>%
-  row_spec(index_biracial, background = "#D3D3D3") %>%
-  landscape()
-```
-
-```{r}
-#| eval: false
-#| label: fig-sample-size
-#| fig-cap: Sample size of biracial respondents combined across American Community Survey data.
-acs %>%
-  filter(str_detect(race, "/")) %>%
-  ggplot(aes(x=reorder(race, race, length)))+
-  geom_bar()+
-  coord_flip()+
-  labs(x=NULL, y="number of respondents")+
-  scale_y_continuous(labels = scales::comma)+
-  theme_bw()
+  kable_styling(font_size=8) %>%
+  column_spec(3:10, width = "1.2cm") %>%
+  row_spec(index_biracial, background = "#D3D3D3") 
+#%>% landscape()
 ```
@@
 Because the ACS lacks data on the exact birthday of respondents, it is not possible to exactly identify students who are below the expected grade level.
-All grades include two expected age groups and a student of a given age could always reasonable be expected to be in two different grades.
+A student of any given age in years can always be expected in two different grades.
 For example, a first grader is expected to be either six or seven years of age, while a second grader is expected to be either seven or eight years of age.
@@
 
-Instead of trying to identify exact grade retention, I instead identify cases in which students are clearly behind grade expectations.
-This is identified when the age of the respondent is greater than either of the possible ages for a student of that grade (e.g.
+Instead of trying to identify exact grade retention, I instead identify cases in which students are clearly below expected grade.
+A student is clearly behind expected grade when the age of the student is greater than either of the possible ages for a student of that grade (e.g.
 a first grader who is eight years old).
 This measure will underestimate the total proportion of students that have been held back.
-However, the goal of the analysis is to examine differences across groups rather than total levels.
+However, the goal of this study is to examine differences across groups rather than total levels.
 The only factor determining this bias is whether the birthday of the student takes place earlier or later in the year, and is unlikely to vary substantially across groups.[^1]
@@
 
-@tbl-descriptive shows the percent of each racial group that is clearly behind the expected grade level.
-My goal is understand the level of grade retention for each biracial group relative to its constituent monoracial groups (e.g.
+@tbl-descriptive shows the percent of each monoracial and biracial group that is clearly below the expected grade.
+My goal is to understand the degree of grade retention for each biracial group relative to its constituent monoracial groups (e.g.
 a Black/White student compared to White and Black students).
@@
 As an example, I display the percent of students clearly behind grade for White, Black, and Black/White students in @fig-example.
-The point estimate and confidence interval for the Black/White group is shown in red, while the two monoracial groups are shown via grey dots and confidence bands.
-For all three cases, I use 83.4% confidence intervals, because overlap on such intervals indicates that the observed difference between the two measures in the sample is not statistically distinguishable from zero in the population at $p<.05$ (CITE).[^2] In this case, the lack of overlap between the Black/White probability and either the White or Black band indicates that the Black/White probability of clearly being behind grade level is statistically distinguishable from both monoracial probabilities.
+The point estimate and confidence interval for the Black/White group are shown in red, while the two monoracial groups are shown via grey dots and confidence bands.
+For all three cases, I use 83.4% confidence intervals, because overlap on such intervals indicates that the observed difference between the two measures in the sample is not statistically distinguishable from zero in the population at $p<.05$ (CITE).[^2] In this case, the lack of overlap between any of the confidence bands indicates that the Black/White probability of being clearly below expected grade is statistically distinguishable from both monoracial probabilities.
 
-[^2]: While 95% confidence intervals are more common in practice, the overlap of such intervals does not necessarily indicate a lack of statistical significance
+[^2]: The correspondence to $p<.05$ is only approximate due to differences in the two standard errors between estimates. However, this approximation will be extremely close in almost all cases.
 
@@
 #| label: fig-example
-#| fig-cap: Probability of being behind grade for biracial black/white respondents in comparison to their monoracial comparison groups.
+#| fig-cap: Probability of being clearly below expected grade for biracial black/white respondents in comparison to their monoracial comparison groups.
 83.4% confidence bands shown around each estimate.
@@
 
-Additionally, I include a measure of the "halfway" point between the two racial groups.[^3] This halfway point allows me to determine where the biracial group falls relative to the expecation of being halfway between the two constituent monoracial groups.
-In this case, Black/White students are much closer to White students in their risk of grade retention, and have grade retention rates much lower than both Black students and the halfway expectation.
+Additionally, I include a measure of the "halfway" point between the two racial groups.[^3] This halfway point allows me to determine where the biracial group falls relative to the expectation of being halfway between the two constituent monoracial groups.
+In this case, Black/White students are much closer to White students in their risk of grade retention, and have probabilities much lower than both Black students and the halfway expectation.
 
@@
 
-@fig-example
+@fig-example shows the raw differences between the three racial groups of interest.
+However, in practice, I want to estimate differences across groups while holding constant a variety of potentially confounding variables.
+To do this, I estimate a set of logit models that predict the likelihood of a student being clearly behind expected grade by race and a variety of other variables.
+I then construct figures similar to @fig-example by calculating from the model the average predicted probabilities (APP) for each racial group.
+APPs estimate the average probability of the outcome by some covariate across all cases, while holding constant all other variables.
+APPs are akin to average marginal effects (AME) and estimated in the same way via the delta method.
+Whereas AMEs estimate differences or slopes, APPs estimate the level for a given category.[^4]
 
-The analysis is driven by a set of logit models that predict the likelihood of a student being clearly behind grade.
-For the baseline model, I consider several characteristics that may vary across groups and thus need to be controlled in all models.
-First, states dictates educational policy and thus can differ substantially in the likelihood of grade retention so I include fixed effects for state of residence in all analysis.
-I cannot identify specific school districts, but I also include dummy variables indicating whether the student lived in a central city, suburban, or rural area.[^4]
+[^4]: Output from the full models used to construct these APPs is available in the supplementary materials provided with this article.
 
-[^4]: The location cannot be determined for all students and I therefore also include a fourth category of unknown in all models.
+I begin with a baseline model that adjusts for a variety of nuisance characteristics that may vary across groups and thus need to be controlled in all models.
+First, I include fixed effects for state of residence because states dictate educational policy and thus can differ substantially in the likelihood of grade retention.
+I cannot identify specific school districts, but I also include dummy variables indicating whether the student lived in a central city, suburban, or rural area.[^5]
 
-The probability of being clearly behind grade also increases with grade and so I include fixed effects for the current grade of the respondent.
-Finally, there may be changes over time in the probability of being clearly behind grade.
-In general, educational reforms have recently decreased the frequency of grade retention (CITE).
-As @fig-year-grade shows, the percent of students clearly behind grade level has declined over time, but this decline has been more substantial at higher grade levels.
+[^5]: The location cannot be determined for all students and I therefore also include a fourth category of unknown in all models.
+
+The probability of being clearly below expected grade also increases with the student's current grade, so I include fixed effects for the current grade of the student.
+Additionally, @fig-year-grade shows that the percent of students clearly behind grade level has declined over time, but this decline has been much more substantial at higher grade levels.
 For the elementary grades, there is no evidence of a decline at all.
+To account for this grade-specific decline in the models I include an interaction between a linear year term and current grade.[^6]
+
+[^6]: Sensitivity analysis showed that this functional form was preferred by the Bayesian information criterion to models with no interaction and a model with interaction terms that treated year as a categorical variable.
+I also considered similar interactions between state and year, but none of the models with these interaction terms were preferred.
 
@@
 
-To account for this grade-specific decline in the models I include an interaction between a linear year term and current grade.
-Sensitivity analysis showed that this functional form was preferred by BIC to models with no interaction and a model with interaction terms that treated year as a categorical variable.
-(==SHOULD I BE CONSIDERING THE SAME FOR STATE?==)
-
 The baseline model also includes the race of the respondent.
@@
 Second, I include a categorical measure of highest degree earned for each parent.
-Finally, I include other measures of family income (square rooted), home ownership, and whether the parents are married.
-All models account for the clustering of multiple respondents within the same household.
-
-Output from the full models is available in the supplementary materials provided with this article.
-However, this output is not particularly useful for directly evaluating the relative placement of biracial students compared to their monoracial peers (e.g.
-a White/Black student compared to White and Black students).
-Throughout this article, I instead use a visual approach to illustrate this placement.
-
-@fig-example shows the raw differences between groups.
-To create this same figure from model results that account for potential confounders, I calculate average predicted probabilities (APP).
-Given a logit model, APPs account for control variables within a model to estimate the average predicted probability of the outcome for a given category.
-APPs are akin to average marginal effects (AME) and estimated in the same way (the delta method).
-Whereas AMEs estimate differences or slopes, APPs estimate the level for a given category.
+Finally, I include measures of family income (square rooted), home ownership, and whether the parents are married that capture other family resources.
+All standard errors reported here also account for the occassional clustering of multiple respondents within the same household.
 

### 2022-04-18 products/paper/main.qmd
Edit the results section
Still needs quite a bit of work, though

@@
     fig-height: 4
-    fig-pos: htb
+    fig-pos: t
     keep-tex: false
@@
 Finally, I include measures of family income (square rooted), home ownership, and whether the parents are married that capture other family resources.
-All standard errors reported here also account for the occassional clustering of multiple respondents within the same household.
+All standard errors reported here also account for the clustering of multiple respondents within the same household.
 
@@
 
-I begin by showing the relative risk of grade retention among monoracial respondents across models.
+I begin by showing the relative risk of being below expected grade among monoracial respondents across models.
 @tbl-monoracial reports average marginal effects on the probability of grade retention for each monoracial minority group in comparison to White students.
@@
 
-Model 1 only controls for structural factors such as state of residence, urbanicity, grade, and year.
-Thus, these results provide a good estimate of the differences across monoracial groups without controlling for differences in cultural and material resources across groups.
+Model 1 only controls for structural factors such as state of residence, student location, grade, and year.
+These results provide a baseline estimate of the differences across monoracial groups without controlling for differences in cultural and material resources across groups.
 The results show that Black, Indigenous, and Latino students all have substantially higher probability of grade retention than White students.
@@
 Controlling for the characteristics related to immigration in Model 2 cuts the gap between White and Latino students in half and increases the gap between Asian and White students.
-It has no impact on the gap between White and Black or White and Indigenous students because most of these students are not first or second generation.
+It has no impact on the gap between White and Black or White and Indigenous students because most of these students are native-born with native-born parents.
 
 Controlling for education in Model 3 reduces the gap substantially for Black and Indigenous students and removes the gap entirely for Latino students.
-In fact, the results of both Models 3 and 4, suggest that, when holding constant cultural and family resources, Latino students are slightly less likely than White students to be held back.
+In fact, the results of both Models 3 and 4 suggest that, when holding constant cultural and family resources, Latino students are slightly less likely than White students to be held back.
 
@@
        label="tbl-monoracial",  #TODO: This does not work
-       float.pos="htb",
+       float.pos="t",
        custom.gof.rows=
@@
 I now turn to the results for biracial respondents.
-Partially to present results in a more easily digestible way, I split these results into two separate figures.
+Due to the number of comparisons being made, I split these results into two separate figures.
 In @fig-black-models, I show the placement of each biracial group involving one black parent in comparison to their two monoracial constituent groups.
@@
 In three cases, this hypothesis can be soundly rejected.
-For Black/White, Black/Latino, and Black/Asian students, The probability of being clearly behind expected grade level is substantially lower than that for monoracial black students and the difference is statistically distinguishable from zero.
-The only case where this is not true is for Black/Indigenous students.
-However this case is also notable for the fact that monoracial Black students have lower risk of grade retention than monoracial Indigenous students.
-In all four cases, these biracial students have a risk of grade retention that is closer to the lower risk monoracial group.
+For Black/White, Black/Latino, and Black/Asian students, The probability of being clearly below expected grade is substantially lower than that for monoracial black students and the difference is statistically distinguishable from zero.
+Only in the case of Black/Indigenous students, do I observe outcomes that are indistinguishable from those of Black students.
+However, in this case,monoracial Black students have lower risk of grade retention than monoracial Indigenous students.
+Thus, In all four cases, part-Black biracial students have a risk of grade retention that is closer to the lower risk monoracial group.
 
-The results also suggest that each of these biracial groups has a risk of grade retention lower than the halfway point, although this hypothesis cannot be rejected completely for the Black/Asian and Black/Indigenous groups, due to the greater degree of uncertainty in the point estimates for these groups.
+The results also suggest that each of these biracial groups has a risk of grade retention lower than the halfway point, although this hypothesis cannot be rejected completely for the Black/Asian and Black/Indigenous groups, due to a higher degree of uncertainty for these smaller groups.
 
 Controlling for resource variables has some effect on the placement of Black/White and Black/Latino biracial respondents.
-In the case of Black/White biracial respondents, Black/White students initially have a risk of grade retention slightly higher than and statistically distinguishable from White student.
+In the case of Black/White biracial respondents, Black/White students initially have a risk of grade retention slightly higher than and statistically distinguishable from White students.
 After controlling for cultural and family resources, however, Black/White respondents have a risk of grade retention that is slightly lower and statistically distinguishable from White students.
@@
 This finding is inconsistent both with the expectation of the one-drop rule and the "in-between" hypothesis.
-I now turn to the experiences of the remaining biracial groups.
+
+@fig-other-models shows the results for the remaining biracial groups.
+In general, the results here are consistent with those for part-Black biracial respondents, with some important caveats discussed below.
+The point estimates indicate that the risk of each biracial group is either quite close to the monoracial group with the lower risk or in-between this group's risk and the halfway point.
 
@@
 
-@fig-other-models shows the results for these remaining biracial groups.
-In general, the results here are consistent with those for part-Black biracial respondents, with some important caveats discussed below.
-In general, the point estimates suggest that the risk of each biracial group is either quite close to the monoracial group with the lower risk or in-between this group's risk and the halfway point.
-
-Any results for part-Latino students are complicated by the substantial change in the monoracial Latino risk across models.
-For example, while Latinos have substantially higher risk of grade retention than Whites in the baseline model, there is only a very small difference favoring Latinos in the full model.
-Nonetheless, even in this case, we see that White/Latinos have risk closer to those of the lower Latino group, but distinguishable from both groups.
-
 For all groups, the full models generally indicate that each biracial group has a risk of grade retention similar to the risk of the lower monoracial constituent group or has risk higher than this group but less than the halfway point.
 There are two exceptions to this general pattern.
-First, for the Indigenous/Latino group, I cannot rule the possibility that this group has risk roughly halfway between the two constituent groups, although the point estimate favors a result closer to the lower Latino group.
-Second, for Latino/Asian students, I observe the only case where the risk for the biracial group is higher than either monoracial group.
+First, for the Indigenous/Latino group, I cannot rule out the possibility that this group has risk roughly halfway between the two constituent groups, although the point estimate favors a result closer to the lower Latino group.
+Second, Latino/Asian students are the only case where the risk for the biracial group is higher than either monoracial group.
 However, the confidence bands suggest that the risk for the Latino/Asian group cannot be differentiated from the rates for Latinos, the higher of the two monoracial constituent groups.
@@
 
+Any results for part-Latino students are complicated by the substantial change in the monoracial Latino risk across models.
+For example, while Latinos have substantially higher risk of grade retention than Whites in the baseline model, there is only a very small difference favoring Latinos in the full model.
+Nonetheless, even in this case, we see that White/Latinos have risk closer to those of the lower Latino group, but distinguishable from both groups.
+
 ## Conclusions
+
+## References

### 2022-04-20 products/paper/main.qmd
Add outline of introduction and background section

@@
 title: Grade Retention and Mixed Race
-author: Aaron Gullickson, *University of Oregon*
+author: 
+  - Aaron Gullickson, *University of Oregon*
 abstract: This is a test abstract
@@
     keep-tex: false
-    #template: resources/aog-latex-ms.tex
 bibliography: ../project.bib
@@
 
+* Research on mixed-race individuals in the US has exploded over the last two decades, driven by both the demographic phenomenon of a biracial baby boom and changes to statistical reporting that allow multiple race reporting.
+* Although, multiraciality in the US is a phenomenon that is in fact older than the country itself (CITATIONS), much of the research has focused particularly on the experiences of first-generation "biracial" individuals whose parents belong to different single race groups.
+The reasons for this are:
+  * These individuals are coming of age in a post-Civil Rights era in which historical norms of racial identification may be dissolving and/or shifting.
+  * The demographic weight of the biracial baby boom suggests that these individuals will play an outsized role in the future of race in the US.
+* Research has focused on a variety of areas with regard to this group.
+In particular, studies of how such individuals identify and are classified by others has been extremely active.
+Significantly less work has been done on the life chances of first-generation biracials relative to their monoracial peers.
+* Such work is critically important because it helps us understand how contemporary biracials are sorted into a pre-existing system of structural racial inequality. 
+* Researchers are well aware of the importance of this question, the relative lack of work on the topic reflects two important limitations in our data sources.
+  * First, simply using self-identification in large scale sources like the Census, ACS, or CPS data is generally seen as problematic.
+We know that many biracial individuals will not identify as multiracial and some individuals who identify as multiracial will not be first-generation biracials.
+Thus, any observed differences across groups may be the result of selection biases in who decides to identify as multiracial.
+  * Second, other data sources that allow identification of biracial respondents by parental race often contain samples of biracial respondents that are too small to facilitate analysis.
+This is particularly true when researchers want to consider outcomes separately by the specific parental racial groups involved, as they should.
+* One way that researchers have overcome this limitation in prior work is by the identification of biracial children living with both biological parents in large-scale Census and ACS data.
+This approach has been primarily used to look at how such children are racially classified (citations), but has also been used to look at residential settlement patterns (citations).
+Its more difficult to look at traditional educational and labor market outcomes, however, precisely because the biracial respondents identified will be minors and thus will not have accumulated much in the way of life outcomes, for good or bad.
+* One important outcome is available, however, for children.
+Grade retention (e.g.
+being held back in school).
+We know that grade retention is relatively common in the US and that the likelihood of grade retention is not equally distributed for children but is rather related to socioeconomic status, family structure, and race (citations).
+We also know that grade retention is related to later negative educational and labor market outcomes (citations).
+Thus, grade retention provides one of the earliest measures of the way systems of stratification sort individuals into a system of structural inequality.
+* In this study, I analyze the grade retention outcomes of biracial children relative to their monoracial peers using ACS data from 2010-2019.
+The results provide a window into the placement of biracial children within an existing system of racial inequality in our educational systems.
+==TAKING TOO LONG TO GET HERE, MIGHT WANT TO RE-ORDER SOME THINGS==
+
 # Background
+
+* The biracial baby boom - why we should care about outcomes for this particular group.
+
+* What should we expect about the outcomes for biracials relative to their monoracial peer groups?
+  * In each case, our interest is in the outcomes for the biracial group relative to the two constituent groups that their parents identify with.
+So for example, how to the outcomes of a Black/White individual compare to monoracial Black and monoracial White individuals?
+  * The most direct pathway for understanding these outcomes is via how such individuals are classified by those in positions of power and control, such as teachers, administrators, employers, etc.
+To the extent that biracial individuals are classified as more like one of their monoracial constituent groups, they will receive the same treatment by individuals and either reap similar benefits or suffer similar discrimination and penalties.
+  * Specifically, I consider two long-standing norms of *hypodescent* and *hyperdescent* in understanding possible placement of biracial individuals (cite F James Davis).
+Hypodescent indicates a norm in which individuals of mixed racial ancestry are classified as members of the lower status group within a racial hierarchy.
+In the US, hypodescent is canonically tied to the "one-drop rule", a norm in which those of part-Black ancestry were simply viewed as Black, irrespective of other ancestry (CITATIONS).
+More generally, if hypodescent is the common practice, we would expect biracial individual's outcomes to be closest those of the monoracial constituent group with the lower outcomes.
+  * Hyperdescent refers to the opposite situation in which a mixed race individual is classified as a member of the higher status group.
+Because hyperdescent blends individuals into the often unmarked dominant group, it is often more difficult to see.
+Indigenous ancestry (American Indian, Pacific Islander, and Alaska Native) has historically followed a logic of hyperdescent in which a mixing of indigenous ancestry does not disqualify individuals from claiming membership with another group (most notably Whites).
+In the case of outcomes, a pattern of hyperdescent would be indicated if biracial individual's outcomes are closest to those of the monoracial constituent group with the higher outcomes.
+  * Discuss work on the contemporary strength of these norms? 
+  * The expectations of hypodescent/hyperdescent are that the biracial respondent will have outcomes close to one of the monoracial constituent groups.
+Another possibility is that biracial respondents will have outcomes intermediate between their monoracial constituent groups.
+There could be several reasons for this:
+    * The racial ambiguity of biracial respondents may protect them from the worst stigmatization and discrimination in social settings, but by the same coin it may allow them to benefit less from membership in advantaged groups.
+Thus, a greater recognition of their "blended" ancestry may place them in a halfway position.
+    * Even in cases where hypodescent and hyperdescent are still norms, physical differentation such as skin tone may provide advantages and disadvantages among members who are classified in the same way.
+We know that skin tone makes a difference even for monoracial Blacks, so the one-drop rule while providing a single classification does not create uniform experiences for all Blacks.
+(==CITE MONKS WORK HERE I THINK==).
+    * Successful outcomes within social systems depend on a variety of non-material, symbolic resources (e.g.
+social and cultural capital) that we rarely completely capture in our measurements.
+Some of the differences in outcomes across racial groups may reflect differences in those resources (CITATIONS?).
+Given that biracial children have parents belonging to two different races, they may possess such resources that are roughly half-way between the two constituent groups on average, and thus will place them roughly in between the groups.
+In effect, the in-between outcomes for biracial children may loosely reflect the "black box" package of differences across racial groups that a racial dummy variables stands in for in our research practices.
+    * Two other possibilities don't suggest the actual placement, but rather may modify it in the face of other prevalent patterns.
+Hybridity itself may be a boon or a burden on biracial individuals that affects their outcomes in, respectively, positive or negative ways.
+      * On the positive side of the ledger, hybridity may enabled a larger cultural "toolkit" that allows mixed race individuals to more easily navigate a variety of social situations.
+On the negative side of the ledger, the longstanding "marginal man" hypothesis suggests that the dissonance of living in two different worlds may lead to negative mental health outcomes for mixed race individuals, which could spill over into other types of outcomes.
+
+* In all of this, its crucially important to understand how other resources affect our understanding of the placement of biracial individuals.
+Biracial individuals are the products of interracial unions and individuals are not randomly selected into such unions.
+This can lead to underlying differences in the socioeconomic resources that biracial individuals possess relative to their monoracial peers.
+However, this selection can be quite complex.
+(DISCUSS CHENG AND LIVELY HERE?)
+  * Status exchange suggests that members of lower status racial groups will be positively selected into interracial unions, while higher status racial groups will be negatively selected.
+How this affects the relative resources of biracial individuals will depend then in some part on the relative strength of these phenomena, the gender balance in a given form of interracial unions, and how the gender of parents affects outcomes.
+Status exchange may also not be present in all forms of interracial unions.
+
+* Review previous studies on outcomes
+  * Campbell
+  * Herman
+  * Holloway/Ellis stuff
+  * Udry? - also related stuff by Cheng? How does Cheng and Powell fit in here?
+  * Do some more bibliographic work, particularly forward searching to see if I am missing something big here.
+  * Limitations of this work
+    * Small sample sizes lead to inconsistent estimates in most of the survey data.
+    * In many cases, impossible to understand outcomes for part-Latino biracials.
+
+* Why grade retention as an outcome?
+  * Grade retention is a relatively common practice. Discuss overall trends in grade retention and connection to educational reforms.
+  * Grade retention has negative effect on later outcomes, even controlling for cognitive differences (CITE WORK HERE).
+  * Clear inequalities by race in practices of grade retention as well as by SES (CITE WORK).
+  * In effect, grade retention may be for many individuals the first encounter with the wider stratification system that sorts individuals and helps to perpetuate inequality.
+
+* What I do  and how it improves our understanding.
 
@@
 
+* Big claim: The results are consistent with a new "regime of hyperdescent" operating across all racial groups.
+This contributes to a muddying of racial distinctiveness via hybridity, while at the same time isolating monoracial members of the most stigmatized racial groups, consistent with arguments like Bonilla-Silva's Latin Americanization.
+
 ## References

### 2022-08-15 products/paper/main.qmd
Add first draft of introduction

--- 2022-04-18 products/paper/main.qmd
@@
 title: Grade Retention and Mixed Race
-author: Aaron Gullickson, *University of Oregon*
+author: 
+  - Aaron Gullickson, *University of Oregon*
 abstract: This is a test abstract
-thanks: Thanks to people and stuff
+thanks: The corresponding author can be reached at aarong@uoregon.edu. All code and data for this project are available at ...
 format:
@@
     keep-tex: false
-    #template: resources/aog-latex-ms.tex
+  docx:
+    toc: false
 bibliography: ../project.bib
@@
 
+The experiences of mixed race individuals will play an important role in the future of race in the US.
+A rise in interracial marriage and coupling since the 1960s has resulted in a "biracial baby boom" of individuals whose parents cross racial lines (==CITE==).
+This demographic phenomenon may have a potentially transformative effect on racial boundaries.
+Although race mixing has always been a feature of US history (and the history of any racialized social system) (==CITE==), this biracial boom is notable for its demographic size and for occurring in  post-Civil Rights ear in which historical norms govern how to handle racial mixture may be dissolving or shifting.
+Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of how these biracial individuals identify their own race and how they are sorted into the prevailing racial hierarchy that reinforces a system of racial inequality.
+
+Extensive scholarship documents the identification and classification decisions of multiracial individuals (==CITE==).
+==SUMMARIZE BRIEFLY==.
+On the other hand, only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers.
+The lack of work on this topic, despite its acknowledged relevance, results from two important limitations in most available data sources.
+
+First, using direct identification of biracial respondents on data sources such as the Census or large scale surveys is problematic.
+Research on the identification and classification of biracial individuals indicates that many individuals whose parents belong to different races will not always or consistently identify with both races (==CITE==).
+Social status and social mobility has been shown to predict racial identification (==CITE==) and this process of reverse causation is likely to be strongest for more ambiguous groups such as biracials.
+Thus, differences in outcome may in some part produce differences in identification rather than the other way around when examining biracial individuals identified by self-classification.
+
+This problem of reverse causation can be largely addressed when researchers have access to the racial identification of a respondent's biological parents.
+While such information is available on some large scale longitudinal surveys, researchers then run into a second problem of sample size.
+The samples of biracial respondents in such cases are generally far too small to make reliable statistical inferences.
+This limitation is particularly severe for biracial individuals with two non-white parents.
+
+In this article, I exploit features of the American Community Survey to identify one life outcome that is available in large scale Census data - grade retention.
+Being held back in school is a relatively common occurrence that has consequences for later educational transitions and is disproportionately applied by race (==CITATIONS==).
+Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience in their life and within it we can clearly see the imprint of the racial hierarchy.
+Because, this outcome occurs to children living in an often natal household, I can identify the race of both biological parents for many cases.
+This identification allows me to compare the likelihood of grade retention for biracial children in comparison to their monoracial peers.
+Pooling data from 2010 to 2019, I am able to draw a large enough sample to make reliable statistical inferences for most biracial groups, including those with two non-white parents.
+
+The results provide a window into the placement of biracial children within an existing system of racial inequality in our educational systems. ==PITHY SUMMARY==
+
 # Background
+
+-   The biracial baby boom - why we should care about outcomes for this particular group.
+
+-   What should we expect about the outcomes for biracials relative to their monoracial peer groups?
+
+    -   In each case, our interest is in the outcomes for the biracial group relative to the two constituent groups that their parents identify with.
+So for example, how to the outcomes of a Black/White individual compare to monoracial Black and monoracial White individuals?
+    -   The most direct pathway for understanding these outcomes is via how such individuals are classified by those in positions of power and control, such as teachers, administrators, employers, etc.
+To the extent that biracial individuals are classified as more like one of their monoracial constituent groups, they will receive the same treatment by individuals and either reap similar benefits or suffer similar discrimination and penalties.
+    -   Specifically, I consider two long-standing norms of *hypodescent* and *hyperdescent* in understanding possible placement of biracial individuals (cite F James Davis).
+Hypodescent indicates a norm in which individuals of mixed racial ancestry are classified as members of the lower status group within a racial hierarchy.
+In the US, hypodescent is canonically tied to the "one-drop rule", a norm in which those of part-Black ancestry were simply viewed as Black, irrespective of other ancestry (CITATIONS).
+More generally, if hypodescent is the common practice, we would expect biracial individual's outcomes to be closest those of the monoracial constituent group with the lower outcomes.
+    -   Hyperdescent refers to the opposite situation in which a mixed race individual is classified as a member of the higher status group.
+Because hyperdescent blends individuals into the often unmarked dominant group, it is often more difficult to see.
+Indigenous ancestry (American Indian, Pacific Islander, and Alaska Native) has historically followed a logic of hyperdescent in which a mixing of indigenous ancestry does not disqualify individuals from claiming membership with another group (most notably Whites).
+In the case of outcomes, a pattern of hyperdescent would be indicated if biracial individual's outcomes are closest to those of the monoracial constituent group with the higher outcomes.
+    -   Discuss work on the contemporary strength of these norms?
+    -   The expectations of hypodescent/hyperdescent are that the biracial respondent will have outcomes close to one of the monoracial constituent groups.
+Another possibility is that biracial respondents will have outcomes intermediate between their monoracial constituent groups.
+There could be several reasons for this:
+        -   The racial ambiguity of biracial respondents may protect them from the worst stigmatization and discrimination in social settings, but by the same coin it may allow them to benefit less from membership in advantaged groups.
+Thus, a greater recognition of their "blended" ancestry may place them in a halfway position.
+        -   Even in cases where hypodescent and hyperdescent are still norms, physical differentation such as skin tone may provide advantages and disadvantages among members who are classified in the same way.
+We know that skin tone makes a difference even for monoracial Blacks, so the one-drop rule while providing a single classification does not create uniform experiences for all Blacks.
+(==CITE MONKS WORK HERE I THINK==).
+        -   Successful outcomes within social systems depend on a variety of non-material, symbolic resources (e.g.
+social and cultural capital) that we rarely completely capture in our measurements.
+Some of the differences in outcomes across racial groups may reflect differences in those resources (CITATIONS?).
+Given that biracial children have parents belonging to two different races, they may possess such resources that are roughly half-way between the two constituent groups on average, and thus will place them roughly in between the groups.
+In effect, the in-between outcomes for biracial children may loosely reflect the "black box" package of differences across racial groups that a racial dummy variables stands in for in our research practices.
+        -   Two other possibilities don't suggest the actual placement, but rather may modify it in the face of other prevalent patterns.
+Hybridity itself may be a boon or a burden on biracial individuals that affects their outcomes in, respectively, positive or negative ways.
+            -   On the positive side of the ledger, hybridity may enabled a larger cultural "toolkit" that allows mixed race individuals to more easily navigate a variety of social situations.
+On the negative side of the ledger, the longstanding "marginal man" hypothesis suggests that the dissonance of living in two different worlds may lead to negative mental health outcomes for mixed race individuals, which could spill over into other types of outcomes.
+
+-   In all of this, its crucially important to understand how other resources affect our understanding of the placement of biracial individuals.
+Biracial individuals are the products of interracial unions and individuals are not randomly selected into such unions.
+This can lead to underlying differences in the socioeconomic resources that biracial individuals possess relative to their monoracial peers.
+However, this selection can be quite complex.
+(DISCUSS CHENG AND LIVELY HERE?)
+
+    -   Status exchange suggests that members of lower status racial groups will be positively selected into interracial unions, while higher status racial groups will be negatively selected.
+How this affects the relative resources of biracial individuals will depend then in some part on the relative strength of these phenomena, the gender balance in a given form of interracial unions, and how the gender of parents affects outcomes.
+Status exchange may also not be present in all forms of interracial unions.
+
+-   Review previous studies on outcomes
+
+    -   Campbell
+    -   Herman
+    -   Holloway/Ellis stuff
+    -   Udry? - also related stuff by Cheng? How does Cheng and Powell fit in here?
+    -   Do some more bibliographic work, particularly forward searching to see if I am missing something big here.
+    -   Limitations of this work
+        -   Small sample sizes lead to inconsistent estimates in most of the survey data.
+        -   In many cases, impossible to understand outcomes for part-Latino biracials.
+
+-   Why grade retention as an outcome?
+
+    -   Grade retention is a relatively common practice. Discuss overall trends in grade retention and connection to educational reforms.
+    -   Grade retention has negative effect on later outcomes, even controlling for cognitive differences (CITE WORK HERE).
+    -   Clear inequalities by race in practices of grade retention as well as by SES (CITE WORK).
+    -   In effect, grade retention may be for many individuals the first encounter with the wider stratification system that sorts individuals and helps to perpetuate inequality.
+
+-   What I do and how it improves our understanding.
 
@@
 
-## Conclusions
+# Conclusions
 
-## References
+-   Big claim: The results are consistent with a new "regime of hyperdescent" operating across all racial groups.
+This contributes to a muddying of racial distinctiveness via hybridity, while at the same time isolating monoracial members of the most stigmatized racial groups, consistent with arguments like Bonilla-Silva's Latin Americanization.
+
+# Acknowledgments {.appendix}
+
+I am grateful to...
+
+# References {.appendix}

### 2022-08-15 products/paper/main.qmd
Add rough draft of first part of background section

@@
 
--   The biracial baby boom - why we should care about outcomes for this particular group.
+How might we expect the outcomes for biracial individuals do differ from their monoracial peers?
+In every case, the relevant comparison is how a biracial person fares relative to the two constituent groups with which their parents identify.
+For example, in the case of an individual with one Black and one White parent, we would compare the average outcomes for such individuals to the average outcomes for individuals with two White and two Black parents.
+The outcome for the Black/White group may be  similar to one of the monoracial groups, it may me somewhere in between the outcomes for the two monoracial groups, or it may be higher or lower than either of the monoracial groups.
 
--   What should we expect about the outcomes for biracials relative to their monoracial peer groups?
+The most direct pathway for understanding the placement of this outcome is via how biracial individuals are classified by those in positions of power and control, such as teachers, administrators, and employers.
+To the extent that biracial individuals are classified as more like one of their monoracial constituent groups.
+they will receive the same treatment and either reap similar rewards or suffer similar discrimination and penalties.
 
-    -   In each case, our interest is in the outcomes for the biracial group relative to the two constituent groups that their parents identify with.
-So for example, how to the outcomes of a Black/White individual compare to monoracial Black and monoracial White individuals?
-    -   The most direct pathway for understanding these outcomes is via how such individuals are classified by those in positions of power and control, such as teachers, administrators, employers, etc.
-To the extent that biracial individuals are classified as more like one of their monoracial constituent groups, they will receive the same treatment by individuals and either reap similar benefits or suffer similar discrimination and penalties.
-    -   Specifically, I consider two long-standing norms of *hypodescent* and *hyperdescent* in understanding possible placement of biracial individuals (cite F James Davis).
-Hypodescent indicates a norm in which individuals of mixed racial ancestry are classified as members of the lower status group within a racial hierarchy.
-In the US, hypodescent is canonically tied to the "one-drop rule", a norm in which those of part-Black ancestry were simply viewed as Black, irrespective of other ancestry (CITATIONS).
-More generally, if hypodescent is the common practice, we would expect biracial individual's outcomes to be closest those of the monoracial constituent group with the lower outcomes.
-    -   Hyperdescent refers to the opposite situation in which a mixed race individual is classified as a member of the higher status group.
-Because hyperdescent blends individuals into the often unmarked dominant group, it is often more difficult to see.
-Indigenous ancestry (American Indian, Pacific Islander, and Alaska Native) has historically followed a logic of hyperdescent in which a mixing of indigenous ancestry does not disqualify individuals from claiming membership with another group (most notably Whites).
-In the case of outcomes, a pattern of hyperdescent would be indicated if biracial individual's outcomes are closest to those of the monoracial constituent group with the higher outcomes.
-    -   Discuss work on the contemporary strength of these norms?
-    -   The expectations of hypodescent/hyperdescent are that the biracial respondent will have outcomes close to one of the monoracial constituent groups.
-Another possibility is that biracial respondents will have outcomes intermediate between their monoracial constituent groups.
-There could be several reasons for this:
-        -   The racial ambiguity of biracial respondents may protect them from the worst stigmatization and discrimination in social settings, but by the same coin it may allow them to benefit less from membership in advantaged groups.
-Thus, a greater recognition of their "blended" ancestry may place them in a halfway position.
-        -   Even in cases where hypodescent and hyperdescent are still norms, physical differentation such as skin tone may provide advantages and disadvantages among members who are classified in the same way.
-We know that skin tone makes a difference even for monoracial Blacks, so the one-drop rule while providing a single classification does not create uniform experiences for all Blacks.
-(==CITE MONKS WORK HERE I THINK==).
-        -   Successful outcomes within social systems depend on a variety of non-material, symbolic resources (e.g.
-social and cultural capital) that we rarely completely capture in our measurements.
-Some of the differences in outcomes across racial groups may reflect differences in those resources (CITATIONS?).
-Given that biracial children have parents belonging to two different races, they may possess such resources that are roughly half-way between the two constituent groups on average, and thus will place them roughly in between the groups.
-In effect, the in-between outcomes for biracial children may loosely reflect the "black box" package of differences across racial groups that a racial dummy variables stands in for in our research practices.
-        -   Two other possibilities don't suggest the actual placement, but rather may modify it in the face of other prevalent patterns.
-Hybridity itself may be a boon or a burden on biracial individuals that affects their outcomes in, respectively, positive or negative ways.
-            -   On the positive side of the ledger, hybridity may enabled a larger cultural "toolkit" that allows mixed race individuals to more easily navigate a variety of social situations.
-On the negative side of the ledger, the longstanding "marginal man" hypothesis suggests that the dissonance of living in two different worlds may lead to negative mental health outcomes for mixed race individuals, which could spill over into other types of outcomes.
+Two well known norms have long governed such patterns of identification.
+The pattern of *hypodescent* dictates that individuals with mixed racial ancestry will be classified as members of the lower status group while the pattern of *hyperdescent* dictates that such individuals will be classified as members of the higher status  (==CITATIONS==) Hypodescent, in the form of the "one drop rule" has long characterized the experiences of mixed part-Black individuals who are treated as exclusively Black (==CITE F JAMES DAVIS==).
+The pattern of hyperdescent blends mixed race individuals into the often unmarked higher status group and is thus more difficult to see in practice.
+Indigenous ancestry (American Indian, Pacific Islander, and Alaska Native) has historically followed a logic of hyperdescent in which a mixing of indigenous ancestry does not disqualify individuals from claiming to be White (==CITATIONS==).
+==NEED TO MIX IN SOME OF THE COMMENTS IN ANN AND ALIYAS RECENT PAPER==
+
+Norms of hyperdescent and hypodescent may account for cases where the biracial group has outcomes similar to one of the monoracial constituent groups.
+However, two issues may diminish the relevance of these norms for understanding the placement of contemporary biracial populations.
+First, for individuals with Asian and Latino ancestry, no historical norm governs the placement of individuals with mixed race.
+Second, the historical norms for those of part-Black and part-Indigenous ancestry may also be breaking down (CITATION).
+MORE TO DO HERE In both of those cases, it is not entirely clear if such norms exist outside of cases in which the other component in the mixing is Whiteness.
+Thus, while these norms may give us some historical expectations for certain groups, they are of limited predictive value more generally.
+
+Patterns of hyperdescent and hypodescent expects biracial individuals to have outcomes more similar to one monoracial group.
+Another possibility is that biracial individuals will have outcomes that are somewhere between their two constituent monoracial groups.
+Why might such an in-between status exist?
+There are several possibilities.
+
+Racial ambiguity in physical appearance may help to produce an in-between status.
+This racial ambiguity may protect biracial respondents from the worst stigmatization and discrimination in social settings.
+However, by the same coin, it may allow them to benefit less from membership in advantaged groups.
+The mechanisms for this process may be two-fold.
+Racial ambiguity may "soften" any particalar experience of discrimination because the person in question does not conform to an ideal type and therefore may be less likely to trigger a sense of racial threat or antipathy.
+Additionally, racial ambiguity may contribute to more probabilistic and uncertain assessments of racial identification by various social actors that contribute to a lower sum total of disadvantage.
+
+Notably, both of these mechanisms could co-exist with norms of hypo/hyperdescent.
+We know, for example, that lighter skin tone among monoracial Black individuals is associated with better outcomes (==CITATION==).
+So, while the one-drop rule provides a consistent classification system, it does not create uniform experiences for all Blacks and skin tone may help to "soften" the experience of discrimination.
+Similarly, if an in-between status results for biracial individuals becaise they are less consistently classified, that does not mean that the classification system and the norms that underlie it are weakening.
+(==CITE MONK HERE==)
+
+Aside from ambiguity in physical appearance, a variety of non-material symbolic resources (e.g.
+social and cultural capital) may also play a role in producing an in-between status.
+We know that racial disparities persist across a variety of outcomes even after controlling for a wide variety of socioeconomic variables (==CITATIONS==).
+The residual effect of race here is a "black box" that we often think of as compromising some combination of (a) discrimination and (b) other forms of social and cultural capital not well captured in our data that are differently distributed by race.
+Biracial individuals may inherit this black box of unobserved resources in weights related to their parent's races, thus helping to produce results that are in-between the two constituent groups.
+(==I SHOULD GIVE SOME EXAMPLES AND CITATIONS OF WHAT I AM TALKING ABOUT HERE==)
+
+All of these arguments relate to where we might expect the outcomes of biracial individuals to fall relative to their monoracial peers.
+However, the experience of hybridity itself may have positive and negative effects that do not tell us the location itself but rather may modify it in the context of other patterns.
+On the positive side, hybridity may enlarge the cultural "toolkit" (==CITATION==) of biracial individuals.
+This enlarged toolkit may allow such individuals to more easily navigate a variety of social situations (e.g.
+code switch).
+On the negative side, Stonequist's (==CITATION==) "marginal man" hypothesis suggests that the dissonance of an identity that crosses strong and salient racial boundaries will be experienced by biracial individuals as not fitting in anywhere.
+This may result in negative mental health experiences for biracial individuals that can spill over into other types of outcomes.
+
+==NO FORMAL HYPOTHESIS==
 

### 2022-08-18 products/paper/main.qmd
Add rought draft of role of class subsection

@@
 
+## Where do Biracials Fit?
+
 How might we expect the outcomes for biracial individuals do differ from their monoracial peers?
@@
 On the negative side, Stonequist's (==CITATION==) "marginal man" hypothesis suggests that the dissonance of an identity that crosses strong and salient racial boundaries will be experienced by biracial individuals as not fitting in anywhere.
-This may result in negative mental health experiences for biracial individuals that can spill over into other types of outcomes.
+This liminality may result in negative mental health experiences for biracial individuals that may affect other types of outcomes as well.
 
@@
 
--   In all of this, its crucially important to understand how other resources affect our understanding of the placement of biracial individuals.
-Biracial individuals are the products of interracial unions and individuals are not randomly selected into such unions.
-This can lead to underlying differences in the socioeconomic resources that biracial individuals possess relative to their monoracial peers.
-However, this selection can be quite complex.
-(DISCUSS CHENG AND LIVELY HERE?)
+## The Role of Class
 
-    -   Status exchange suggests that members of lower status racial groups will be positively selected into interracial unions, while higher status racial groups will be negatively selected.
-How this affects the relative resources of biracial individuals will depend then in some part on the relative strength of these phenomena, the gender balance in a given form of interracial unions, and how the gender of parents affects outcomes.
-Status exchange may also not be present in all forms of interracial unions.
+In addition to these specifically racial mechanisms that may drive the placement of biracial individuals, researchers must also pay careful attention to how socioeconomic background may affect the placement of biracial individuals within a racial hierarchy.
+Socioeconomic background plays a role in racial inequality more generally as we know that accounting for socioeconomic difference often reduces the magnitude of observed racial inequality, suggesting that some of racial inequality we observe operates through class differences (CITATION).
+This issue is particularly important for biracial individuals because of potential class selection into interracial unions.
+The likelihood of being an interracial union varies by class background for at least some groups (CITATION), although it varies in complex ways and we know more about interracial marriages than other types of unions that may produce children.
+For example, black partners tend to be positively selected with regard to education into Black/White marriages while white partners are negatively selected (CITATION).
+According to status exchange theory (CITATIONS), this pattern may be common to all types of interracial marriages within an unequal status hierarchy, although the evidence of status exchange for other forms of marriages is weaker (CITATION).
+How this affects the relative resources of biracial individuals will depend then in some part on the relative strength of these phenomena for a given pairing, the gender balance in that pairing, and how the gender of parents affects outcomes.
+
+Regardless of the underlying complexity, delineating the role of this selection process in the prior generation as opposed to other mechanisms is key to understanding the overall process that sorts biracial individuals into a racialized system.
+Although this approach is sometimes characterized as pitting race vs class and minimizing the importance of race, that is not case here.
+Instead, what this delineation allows us to do is understanding how race affects the outcomes of biracial individuals in a multigenerational process and allow us to better pinpoint the generational locus of certain mechanisms.
+
+## Prior Work
 
@@
         -   In many cases, impossible to understand outcomes for part-Latino biracials.
+
+## Grade Retention as an Outcome
 

### 2022-08-18 products/paper/main.qmd
Add changes from first editing of into and part of background

@@
 This demographic phenomenon may have a potentially transformative effect on racial boundaries.
-Although race mixing has always been a feature of US history (and the history of any racialized social system) (==CITE==), this biracial boom is notable for its demographic size and for occurring in  post-Civil Rights ear in which historical norms govern how to handle racial mixture may be dissolving or shifting.
-Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of how these biracial individuals identify their own race and how they are sorted into the prevailing racial hierarchy that reinforces a system of racial inequality.
+Although race mixing has always been a feature of US history (==CITE==), this biracial boom is notable for its demographic size and for occurring in a post-Civil Rights era in which historical norms governing how to handle racial mixture may be shifting.
+Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of how these biracial individuals identify their own race and how they are sorted into a prevailing racial hierarchy that reinforces a system of racial inequality.
 
@@
 On the other hand, only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers.
-The lack of work on this topic, despite its acknowledged relevance, results from two important limitations in most available data sources.
+(==CITATION==) The lack of work on this topic, despite its acknowledged relevance, results from two important limitations in most available data sources.
 
 First, using direct identification of biracial respondents on data sources such as the Census or large scale surveys is problematic.
-Research on the identification and classification of biracial individuals indicates that many individuals whose parents belong to different races will not always or consistently identify with both races (==CITE==).
-Social status and social mobility has been shown to predict racial identification (==CITE==) and this process of reverse causation is likely to be strongest for more ambiguous groups such as biracials.
-Thus, differences in outcome may in some part produce differences in identification rather than the other way around when examining biracial individuals identified by self-classification.
+Research on the identification and classification of biracial individuals indicates that individuals whose parents belong to different races will not always or consistently identify with both races (==CITE==).
+Furthermore, social status and social mobility has been shown to predict racial identification (==CITE==) and this process of reverse causation is likely to be strongest for more ambiguous groups such as biracials.
+Thus, differences in outcome may in some part produce differences in identification rather than the other way around when examining biracial individuals identified by self-identification.
 
@@
 
-In this article, I exploit features of the American Community Survey to identify one life outcome that is available in large scale Census data - grade retention.
+When children reside with both biological parents, biracial individuals can be identified in large scale data sources such as the Census.
+Researchers have long utilized the ability to identify such biracial children to examine the classification choices made on behalf of these children (==CITATIONS==).
+Because of their age, it is more difficult to analyze the outcomes of such children due to the simple fact that they haven't had many measurable life outcomes yet.
+
+However, one important outcome, grade retention, is measurable and important even for young children.
 Being held back in school is a relatively common occurrence that has consequences for later educational transitions and is disproportionately applied by race (==CITATIONS==).
-Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience in their life and within it we can clearly see the imprint of the racial hierarchy.
-Because, this outcome occurs to children living in an often natal household, I can identify the race of both biological parents for many cases.
-This identification allows me to compare the likelihood of grade retention for biracial children in comparison to their monoracial peers.
+Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience and within it we can clearly see the imprint of the racial hierarchy.
+In this article, I use data from the American Community Survey to examine the likelihood of being held back in school for biracial children in comparison to their monoracial peers.
 Pooling data from 2010 to 2019, I am able to draw a large enough sample to make reliable statistical inferences for most biracial groups, including those with two non-white parents.
@@
 
-The most direct pathway for understanding the placement of this outcome is via how biracial individuals are classified by those in positions of power and control, such as teachers, administrators, and employers.
+The most direct pathway for such placement is via how biracial individuals are classified by those in positions of power and control, such as teachers, administrators, and employers.
 To the extent that biracial individuals are classified as more like one of their monoracial constituent groups.
@@
 
-Two well known norms have long governed such patterns of identification.
-The pattern of *hypodescent* dictates that individuals with mixed racial ancestry will be classified as members of the lower status group while the pattern of *hyperdescent* dictates that such individuals will be classified as members of the higher status  (==CITATIONS==) Hypodescent, in the form of the "one drop rule" has long characterized the experiences of mixed part-Black individuals who are treated as exclusively Black (==CITE F JAMES DAVIS==).
+Two well known norms have long governed such patterns of classification.
+The pattern of *hypodescent* dictates that individuals with mixed racial ancestry will be classified as members of the lower status group while the pattern of *hyperdescent* dictates that such individuals will be classified as members of the higher status group (==CITATIONS==).
+Hypodescent, in the form of the "one drop rule" has long characterized the experiences of mixed part-Black individuals who are treated as exclusively Black (==CITE F JAMES DAVIS==).
 The pattern of hyperdescent blends mixed race individuals into the often unmarked higher status group and is thus more difficult to see in practice.
@@
 So, while the one-drop rule provides a consistent classification system, it does not create uniform experiences for all Blacks and skin tone may help to "soften" the experience of discrimination.
-Similarly, if an in-between status results for biracial individuals becaise they are less consistently classified, that does not mean that the classification system and the norms that underlie it are weakening.
 (==CITE MONK HERE==)
@@
 
-All of these arguments relate to where we might expect the outcomes of biracial individuals to fall relative to their monoracial peers.
-However, the experience of hybridity itself may have positive and negative effects that do not tell us the location itself but rather may modify it in the context of other patterns.
+The experience of hybridity itself may also positively and/or negative affect the outcomes for biracial individuals.
 On the positive side, hybridity may enlarge the cultural "toolkit" (==CITATION==) of biracial individuals.
@@
 On the negative side, Stonequist's (==CITATION==) "marginal man" hypothesis suggests that the dissonance of an identity that crosses strong and salient racial boundaries will be experienced by biracial individuals as not fitting in anywhere.
-This liminality may result in negative mental health experiences for biracial individuals that may affect other types of outcomes as well.
+This liminality may result in negative mental health experiences for biracial individuals that can then affect other types of outcomes as well.
 
@@
 
-In addition to these specifically racial mechanisms that may drive the placement of biracial individuals, researchers must also pay careful attention to how socioeconomic background may affect the placement of biracial individuals within a racial hierarchy.
-Socioeconomic background plays a role in racial inequality more generally as we know that accounting for socioeconomic difference often reduces the magnitude of observed racial inequality, suggesting that some of racial inequality we observe operates through class differences (CITATION).
+In addition to the aforementioned racial mechanisms that may drive the placement of biracial individuals, researchers must also pay careful attention to how socioeconomic background may affect the placement of biracial individuals within a racial hierarchy.
+Socioeconomic background plays a role in racial inequality more generally as we know that accounting for socioeconomic difference often reduces the magnitude of observed racial inequality, suggesting that some of the racial inequality we observe operates through class differences (CITATION).
 This issue is particularly important for biracial individuals because of potential class selection into interracial unions.

### 2022-08-18 products/paper/main.qmd
Add a different start to intermarriage section

@@
 
-## The Role of Class
+## Selection into Intermarriage
+
+The placement of biracial individuals may also be affected by mechanisms that drive selection into interracial unions in the previous generation.
+Such selection may lead to differences on a variety of parental background characteristics between biracial individuals and their monoracial peers.
+How this selection will affect a biracial child's outcomes for a given type of pairing are complex because it depends on the strength and direction of that selection on each partner, its consistency across different types of unions, and how strongly the selected characteristic predicts outcomes in the next generation.
+
+Most research on such selection has focused on socioeconomic characteristics.
+Status exchange theory predicts that given an unequal racial hierarchy, individuals from a lower status racial group will be positively selected on socioeconomic characteristics into intermarriage with a higher status group, while individuals from a higher status racial group will be similarly negatively selected into the same marriage (CITATION).
+Using education as the socioeconomic parameter, prior research has shown strong evidence of this pattern in the case of Black/White marriages in the US (CITATION).
+Evidence of this pattern for other forms of intermarriage is less clear (CITATION).
+
+Alternatively,
+
+---
+
+The aforementioned mechanisms all relate to
 
 In addition to the aforementioned racial mechanisms that may drive the placement of biracial individuals, researchers must also pay careful attention to how socioeconomic background may affect the placement of biracial individuals within a racial hierarchy.
-Socioeconomic background plays a role in racial inequality more generally as we know that accounting for socioeconomic difference often reduces the magnitude of observed racial inequality, suggesting that some of the racial inequality we observe operates through class differences (CITATION).
+Accounting for socioeconomic difference often reduces the magnitude of observed racial inequality on a given outcome, suggesting that some of the racial inequality we observe operates through class differences (CITATION).
 This issue is particularly important for biracial individuals because of potential class selection into interracial unions.

### 2022-08-19 products/paper/main.qmd
Add outline for rest of selection section

@@
 Alternatively,
+
+1. its complicated
+2.
+Rather than sort out a priori predictions, its important to control for most likely selected characteristics.
+In the case I explore here socioeconomic status and traits measuring assimilation such as foreign-born status and english language proficiency.
+3. This is not a "race vs. class" argument that diminishes the importance of race. Rather it delineates where generationally and by what mechanisms race might matter for biracial individuals.
 

### 2022-08-19 products/paper/main.qmd
Update biblio

@@
 First, using direct identification of biracial respondents on data sources such as the Census or large scale surveys is problematic.
-Research on the identification and classification of biracial individuals indicates that individuals whose parents belong to different races will not always or consistently identify with both races (==CITE==).
+Research on the identification and classification of biracial individuals indicates that individuals whose parents belong to different races will not always or consistently identify with both races [@liebler_america_2017] (==OTHER CITE?==).
 Furthermore, social status and social mobility has been shown to predict racial identification (==CITE==) and this process of reverse causation is likely to be strongest for more ambiguous groups such as biracials.
@@
 
----
+------------------------------------------------------------------------
 

### 2022-08-24 products/paper/main.qmd
Add outline for a new approach to first part of background section

@@
 The outcome for the Black/White group may be similar to one of the monoracial groups, it may me somewhere in between the outcomes for the two monoracial groups, or it may be higher or lower than either of the monoracial groups.
+
+Possible restructuring of this whole section:
+
+* To understand where biracials fit, it is helpful to think about what mechanisms drive the racial hierarchy more broadly.
+  * First, racial inequality in outcomes may be produced by the cumulative sum of individual acts of discrimination as well as systematic forms of bias and discrimination.
+Individual discrimination need not be consciously performed and may instead reflect cognitive biases and schema that individuals unknowingly rely upon that produce differential outcomes by race.
+This discrimination will be more consequential when performed by important gatekeepers of opportunity such as teachers, health care workers, and employers.
+  * Systematic discrimination, on the other hand, reflect biases built into institutional systems that channel rewards to certain racial groups.
+This can operate through such things as norms, laws, and organizational practices.
+In our current "colorblind" era, such mechanisms may be formally race neutral, but produce disparate impacts by race.
+  * In either case, the experiences of biracial individuals within such a system of discrimination depend on how they are classified by others.
+To the extent that biracial individuals are classified as more like one of their monoracial constituent groups.
+they will receive the same treatment and either reap similar rewards or suffer similar discrimination and penalties.
+  * discussion of hypodescent/hyperdescent as governing classification into single-race groups. 
+  * Iversen et. al.'s critique of hypodescent/hyperdescent, higlighting:
+    * historical specificy of hypodescent and hyperdescent pattern
+    * equally consistent with a "succession/recession" perspective
+    * regimes governing certain groups may differ due to historical circumstances.
+  * How does this questions of classification address the case where biracial individuals might be in-between? A couple of possibilities.
+    * Racial ambiguity could lead to more inconsistent classification into single race categories across observers submitting biracials to differing exposure to advantages/distadvantages somewhere between  their two monoracial constituent groups.
+    * Greater physical ambiguity may allow biracials to avoid worst aspects of discrimination because they don't match the "ideal type" for most discriminated group.
+We already know that this process exists among monoracial blacks in terms of skin tone, so this is not unique to multiracials.
+    * It is also possible that biracials are "correctly" classified by observers into a "middleman" minority position.
+There is evidence of this kind of practice in Latin America, and it may become more common in the US (Bonilla-Silva).
+It is unlikely that observers are capable of differentiating fine divisions of ancestry but rather can put people into a nebulous middleman position.
+* Some racial inequality may also be explained by differential distributions of resources across racial groups.
+On average, non-whites come from a lower SES background than whites and this may indirectly lead to observed racial differences.
+Although much of the focus of the "race vs.
+class" debate is on SES characteristics, less concretely defined social and cultural capital differences across race may exacerbate racial inequality.
+  * The role of these resource differences in biracial's position depends on what resources the typical parent in an interracial union brings with them.
+This is a complex and not-well understood process even for easily measurable characteristics like education.
+...
+  * the black box of symbolic resources.
+* Aside from these major mechanisms for producing racial inequality, the experience of hybridity itself may affect the outcomes of biracial individuals relative to their monoracial peers.
+  * Marginal man theory
+  * Enlarged toolkits
 

### 2022-08-30 products/paper/main.qmd
Rework first section of Where do Biracials fit
This is the part focused on "pure" race explanations for the hierarchy and the classification of biracials

@@
 The outcome for the Black/White group may be similar to one of the monoracial groups, it may me somewhere in between the outcomes for the two monoracial groups, or it may be higher or lower than either of the monoracial groups.
+
+To better understand where biracial individuals might fit, it is necessary to understand what is producing a contemporary system of racial inequality overall.
+Here, I distinguish three mechanisms that may help us understand the biracial position within this racial hierarchy.
+First, I consider arguments for the racial hierarchy that focus on the effects of racial discrimination at both an individual and systemic level.
+Second, I consider arguments regarding different distributions of resources across racial groups that produce different average arguments.
+Thirs, I consider arguments regarding how the nature of hybridity itself may affect outcomes.
+
+Discrimination and bias by race play an important role in overall levels of racial inequality.
+Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unkwnowningly rely upon and reinforce in social interactions.
+Such discrimination will be most consequential when performed by important gatekeepers of opportunity such as teachers, health care workers, and employers.
+
+Discrimination need not just occur on an individual basis, but can also be systemic in nature when biases are built into institutional systems.
+Such systemic racism has two effects.
+First, it provides norms and ideologies that create, support, and reinforce the cognitive schemas that individuals use in everyday practice.
+Second, norms, laws, and organizational practices can have built in biases that channel rewards to certain racial groups.
+In our current "colorblind" regime, such systems may be formally race neutral but produce disparate impacts by race.
+
+The experiences of biracial individuals within such a system of discrimination depend almost entirely on how they are classified by others.
+To the extent that biracial individuals are classified as more like one of their monoracial constituent groups than the other, they will receive the same treatment and reap similar rewards or suffer similar penalties.
+
+The strongest historical precedent for understanding such classification in the US is the "one drop rule" norm concering Black ancestry.
+According to the one-drop rule, individuals of known Black ancestry are considered exclusively Black, regardless of other ancestries.
+Davis (==CITE==) generalizes this practice to the idea of *hypodescent* in which individuals with mixed racial ancestry will be classified as members of the lower status group.
+A symmetric, but opposite, pattern of *hyperdescent* has governed the experiences of individuals with American Indian ancestry (==CITE==).
+
+Although the hypodescent/hyperdescent paradigm is often used to theoretically frame the possible contemporary experiences of multiracial individuals, there are reasons to be cautious about the usefulness of this paradigm to both the current and historical situations.
+Although often couched in generalizable term, hypodescent and hyperdescent are actually only observed in two specific historical cases involving Black ancestry and American Indian ancestry.
+As CITE note, these two cases could just as parsimoniously be explained by a "dominance" model which ranks ancestries by their tendencies to be supercessive or recessive in determining identification and classification.
+In this framework, among the historical "racial triangle" of American colonialism, Black ancestry is supercessive, Indigenous ancestry is recessive, and White ancestry falls somewhere in between.
+
+The development of these particular norms may also reflect the historical regime in which they were developed (CITE).
+In this case, neither the hypodescent/hyperdescent or dominance paradigm may help us completely to understand the experiences of racialized groups such as Latinos and Asians.
+Second, the historical norms for all groups may also be breaking up or transforming in the post-Civil Rights era (CITE).
+
+If outcomes for biracial individuals all closer to one of their constituent groups, then we may have stronger evidence about such norms of classification.
+However, what if those outcomes fall in-between the two groups?
+How might classification and discrimination produce such an in-between status?
+There are several possibilities here.
+
+First, racial ambiguity in appearance may lead to more inconsistent classification into single race categories across observers.
+Even if the norms consistently indicate a certain type of classification, correct identification by observers is still a probabilistic phenomenon in all social interactions and likely to be more heterogenous for biracial individuals.
+As a result, biracial individuals will experience a lower sum total of discrimination and bias affecting the lower status group but will also not reap the same rewards as a member of the higher status group.
+
+Second, even when "correctly" classified, greater physical ambiguity may allow biracial individuals to avoid the worst instances of bias and discrimination because they do not conform to the ideal type for the group facing the most discrimination.
+In other words, the experience of discrimination may be "softened" because their ambigious physical appearance provokes less racial antipathy.
+The US in fact already has a long history of this process.
+Because of the one-drop rule, Black individuals vary substantially in skin tone and those individuals with a lighter skin tone have better outcomes.
+Thus light skin can moderate (but not eliminate) the stigma of Blackness even among monoracial individuals.
+
+Third, it is also possible that biracial individuals may be classified into a distinct "middleman" minority position.
+This kind of middleman status is common in Latin America (CITATIONS) and Bonilla-Silva (CITE) has argued that this may be a possible future for the US as well.
+In this case, many biracial groups may occupy a nebulous and ill-defined middleman or buffer class position in-between whites and darker-skinned non-whites, helping to cement a new kind of racial hierarchy.
 

### 2022-08-31 products/paper/main.qmd
Restructure headings; correct description of calculating child race

@@
 
-# Background
-
-## Where do Biracials Fit?
+# Where do Biracials Fit?
 
@@
 Second, even when "correctly" classified, greater physical ambiguity may allow biracial individuals to avoid the worst instances of bias and discrimination because they do not conform to the ideal type for the group facing the most discrimination.
-In other words, the experience of discrimination may be "softened" because their ambigious physical appearance provokes less racial antipathy.
+In other words, the experience of discrimination may be "softened" because their ambiguous physical appearance provokes less racial antipathy.
 The US in fact already has a long history of this process.
@@
 
-## Prior Work
+# Prior Work
 
@@
 
-## Grade Retention as an Outcome
+# Grade Retention as an Outcome
 
@@
 These categories form the "ethnoracial pentagon" that is commonly used in popular practice and government tabulation to identify race in the US [@hollinger_postethnic_1995].
-Respondents are identified as Latino base on their response to the Hispanicity question, regardless of their response to the race question.
-Respondents are identified as Indigenous if they responded as either American Indian/Alaska Native or as Pacific Islander.
+Parents are identified as Latino based on their response to the Hispanicity question, regardless of their response to the race question.
+Parents are identified as Indigenous if they responded as either American Indian/Alaska Native or as Pacific Islander.
 Because the goal of my analysis is to examine specifically the outcomes of "first-generation" biracial children, I exclude cases where at least one parent identified with multiple races.

### 2022-08-31 products/paper/main.qmd
Full rough draft of biracial fit section

@@
 First, I consider arguments for the racial hierarchy that focus on the effects of racial discrimination at both an individual and systemic level.
-Second, I consider arguments regarding different distributions of resources across racial groups that produce different average arguments.
-Thirs, I consider arguments regarding how the nature of hybridity itself may affect outcomes.
+Second, I consider arguments regarding different distributions of resources across racial groups that produce different average outcomes.
+Third, I consider arguments regarding how the nature of hybridity itself may affect outcomes.
+
+## Discrimination and Classification
 
@@
 
-Possible restructuring of this whole section:
+## Differential Resources
 
-* To understand where biracials fit, it is helpful to think about what mechanisms drive the racial hierarchy more broadly.
-  * First, racial inequality in outcomes may be produced by the cumulative sum of individual acts of discrimination as well as systematic forms of bias and discrimination.
-Individual discrimination need not be consciously performed and may instead reflect cognitive biases and schema that individuals unknowingly rely upon that produce differential outcomes by race.
-This discrimination will be more consequential when performed by important gatekeepers of opportunity such as teachers, health care workers, and employers.
-  * Systematic discrimination, on the other hand, reflect biases built into institutional systems that channel rewards to certain racial groups.
-This can operate through such things as norms, laws, and organizational practices.
-In our current "colorblind" era, such mechanisms may be formally race neutral, but produce disparate impacts by race.
-  * In either case, the experiences of biracial individuals within such a system of discrimination depend on how they are classified by others.
-To the extent that biracial individuals are classified as more like one of their monoracial constituent groups.
-they will receive the same treatment and either reap similar rewards or suffer similar discrimination and penalties.
-  * discussion of hypodescent/hyperdescent as governing classification into single-race groups. 
-  * Iversen et. al.'s critique of hypodescent/hyperdescent, higlighting:
-    * historical specificy of hypodescent and hyperdescent pattern
-    * equally consistent with a "succession/recession" perspective
-    * regimes governing certain groups may differ due to historical circumstances.
-  * How does this questions of classification address the case where biracial individuals might be in-between? A couple of possibilities.
-    * Racial ambiguity could lead to more inconsistent classification into single race categories across observers submitting biracials to differing exposure to advantages/distadvantages somewhere between  their two monoracial constituent groups.
-    * Greater physical ambiguity may allow biracials to avoid worst aspects of discrimination because they don't match the "ideal type" for most discriminated group.
-We already know that this process exists among monoracial blacks in terms of skin tone, so this is not unique to multiracials.
-    * It is also possible that biracials are "correctly" classified by observers into a "middleman" minority position.
-There is evidence of this kind of practice in Latin America, and it may become more common in the US (Bonilla-Silva).
-It is unlikely that observers are capable of differentiating fine divisions of ancestry but rather can put people into a nebulous middleman position.
-* Some racial inequality may also be explained by differential distributions of resources across racial groups.
-On average, non-whites come from a lower SES background than whites and this may indirectly lead to observed racial differences.
-Although much of the focus of the "race vs.
-class" debate is on SES characteristics, less concretely defined social and cultural capital differences across race may exacerbate racial inequality.
-  * The role of these resource differences in biracial's position depends on what resources the typical parent in an interracial union brings with them.
-This is a complex and not-well understood process even for easily measurable characteristics like education.
-...
-  * the black box of symbolic resources.
-* Aside from these major mechanisms for producing racial inequality, the experience of hybridity itself may affect the outcomes of biracial individuals relative to their monoracial peers.
-  * Marginal man theory
-  * Enlarged toolkits
+Racial inequality may also be at least partly explained by different distributions of material, cultural, and social resources across racial groups.
+On average, non-whites come from a lower socioeconomic background (e.g.
+parental income, wealth, and education) than whites and such socioeconomic background is consequential for outcomes.
+Other less materially defined resources also vary across groups such as cultural styles, social networks, and neighborhood quality.
+When such characteristics are held constant or "controlled", the remaining differences between racial groups tend to be smaller, although such differences rarely disappear.
+These differences in background resources reflect the accumulation of historical processes of discrimination.
+Thus, observed racial inequality reflects both contemporary processes of discrimination and bias and the intergenerational accumulation of disadvantage.
 
-The most direct pathway for such placement is via how biracial individuals are classified by those in positions of power and control, such as teachers, administrators, and employers.
-To the extent that biracial individuals are classified as more like one of their monoracial constituent groups.
-they will receive the same treatment and either reap similar rewards or suffer similar discrimination and penalties.
+How do these differential distributions of resources affect biracials?
+A naive argument would be that because biracial have parents from two different racial groups, they will on average inherit parental resources roughly halfway in-between their two constituent groups (CITE CHEW ET.
+AL.).
+Prior research looking at biracial children in the Census data has found some evidence to support this argument (CITE CHEW ET.
+AL.).
+If such were the case, then the resources of biracial individuals would predict outcomes in-between the two constituent groups, barring other mechanisms.
 
-Two well known norms have long governed such patterns of classification.
-The pattern of *hypodescent* dictates that individuals with mixed racial ancestry will be classified as members of the lower status group while the pattern of *hyperdescent* dictates that such individuals will be classified as members of the higher status group (==CITATIONS==).
-Hypodescent, in the form of the "one drop rule" has long characterized the experiences of mixed part-Black individuals who are treated as exclusively Black (==CITE F JAMES DAVIS==).
-The pattern of hyperdescent blends mixed race individuals into the often unmarked higher status group and is thus more difficult to see in practice.
-Indigenous ancestry (American Indian, Pacific Islander, and Alaska Native) has historically followed a logic of hyperdescent in which a mixing of indigenous ancestry does not disqualify individuals from claiming to be White (==CITATIONS==).
-==NEED TO MIX IN SOME OF THE COMMENTS IN ANN AND ALIYAS RECENT PAPER==
-
-Norms of hyperdescent and hypodescent may account for cases where the biracial group has outcomes similar to one of the monoracial constituent groups.
-However, two issues may diminish the relevance of these norms for understanding the placement of contemporary biracial populations.
-First, for individuals with Asian and Latino ancestry, no historical norm governs the placement of individuals with mixed race.
-Second, the historical norms for those of part-Black and part-Indigenous ancestry may also be breaking down (CITATION).
-MORE TO DO HERE In both of those cases, it is not entirely clear if such norms exist outside of cases in which the other component in the mixing is Whiteness.
-Thus, while these norms may give us some historical expectations for certain groups, they are of limited predictive value more generally.
-
-Patterns of hyperdescent and hypodescent expects biracial individuals to have outcomes more similar to one monoracial group.
-Another possibility is that biracial individuals will have outcomes that are somewhere between their two constituent monoracial groups.
-Why might such an in-between status exist?
-There are several possibilities.
-
-Racial ambiguity in physical appearance may help to produce an in-between status.
-This racial ambiguity may protect biracial respondents from the worst stigmatization and discrimination in social settings.
-However, by the same coin, it may allow them to benefit less from membership in advantaged groups.
-The mechanisms for this process may be two-fold.
-Racial ambiguity may "soften" any particalar experience of discrimination because the person in question does not conform to an ideal type and therefore may be less likely to trigger a sense of racial threat or antipathy.
-Additionally, racial ambiguity may contribute to more probabilistic and uncertain assessments of racial identification by various social actors that contribute to a lower sum total of disadvantage.
-
-Notably, both of these mechanisms could co-exist with norms of hypo/hyperdescent.
-We know, for example, that lighter skin tone among monoracial Black individuals is associated with better outcomes (==CITATION==).
-So, while the one-drop rule provides a consistent classification system, it does not create uniform experiences for all Blacks and skin tone may help to "soften" the experience of discrimination.
-(==CITE MONK HERE==)
-
-Aside from ambiguity in physical appearance, a variety of non-material symbolic resources (e.g.
-social and cultural capital) may also play a role in producing an in-between status.
-We know that racial disparities persist across a variety of outcomes even after controlling for a wide variety of socioeconomic variables (==CITATIONS==).
-The residual effect of race here is a "black box" that we often think of as compromising some combination of (a) discrimination and (b) other forms of social and cultural capital not well captured in our data that are differently distributed by race.
-Biracial individuals may inherit this black box of unobserved resources in weights related to their parent's races, thus helping to produce results that are in-between the two constituent groups.
-(==I SHOULD GIVE SOME EXAMPLES AND CITATIONS OF WHAT I AM TALKING ABOUT HERE==)
-
-The experience of hybridity itself may also positively and/or negative affect the outcomes for biracial individuals.
-On the positive side, hybridity may enlarge the cultural "toolkit" (==CITATION==) of biracial individuals.
-This enlarged toolkit may allow such individuals to more easily navigate a variety of social situations (e.g.
-code switch).
-On the negative side, Stonequist's (==CITATION==) "marginal man" hypothesis suggests that the dissonance of an identity that crosses strong and salient racial boundaries will be experienced by biracial individuals as not fitting in anywhere.
-This liminality may result in negative mental health experiences for biracial individuals that can then affect other types of outcomes as well.
-
-==NO FORMAL HYPOTHESIS==
-
-## Selection into Intermarriage
-
-The placement of biracial individuals may also be affected by mechanisms that drive selection into interracial unions in the previous generation.
-Such selection may lead to differences on a variety of parental background characteristics between biracial individuals and their monoracial peers.
+However, the actual resource position of biracial individuals is likely far more complicated because individuals are not randomly selected into interracial unions.
 How this selection will affect a biracial child's outcomes for a given type of pairing are complex because it depends on the strength and direction of that selection on each partner, its consistency across different types of unions, and how strongly the selected characteristic predicts outcomes in the next generation.
 
-Most research on such selection has focused on socioeconomic characteristics.
 Status exchange theory predicts that given an unequal racial hierarchy, individuals from a lower status racial group will be positively selected on socioeconomic characteristics into intermarriage with a higher status group, while individuals from a higher status racial group will be similarly negatively selected into the same marriage (CITATION).
 Using education as the socioeconomic parameter, prior research has shown strong evidence of this pattern in the case of Black/White marriages in the US (CITATION).
-Evidence of this pattern for other forms of intermarriage is less clear (CITATION).
+Evidence of this pattern for other forms of intermarriage is less clear (CITATION), as is research on other forms of interracial unions such as cohabitations (CITATIONS).
+In any case, because status exchange positively selects one parent and negatively selects the other partner, the two effects may somewhat cancel each other out in terms of the overall distribution of resources.
 
-Alternatively,
+Regardless of the underlying complexity of this selection process, delineating its role as opposed to other mechanisms is key to understanding the overall process that sorts biracial individuals within the racial hierarchy.
+Although this approach is sometimes characterized as pitting race versus class and minimizing the importance of race, that is not case here.
+Instead, what this delineation allows us to do is understand how race affects the outcomes of biracial individuals in a multigenerational process to better pinpoint the generational locus of specific mechanisms.
 
-1.  its complicated
-2.
-Rather than sort out a priori predictions, its important to control for most likely selected characteristics.
-In the case I explore here socioeconomic status and traits measuring assimilation such as foreign-born status and english language proficiency.
-3.  This is not a "race vs. class" argument that diminishes the importance of race. Rather it delineates where generationally and by what mechanisms race might matter for biracial individuals.
+## Hybridity
 
-------------------------------------------------------------------------
+The experience of hybridity itself may also positively and/or negatively affect outcomes for biracial individuals.
+The "marginal man" theory argues that biracial individuals will find it difficult to fit in because of the dissonance of an identity that crosses strong and salient racial boundaries (CITE PARK AND STONEQUIST).
+This liminal identity conflict will then result in negative mental health experiences that can feed into other negative outcomes as well (CITE CHENG AND LIVELY?).
 
-The aforementioned mechanisms all relate to
+Marginal man theory, developed in the first half of the Twentieth century, saw cultures as monolithic and static and thus to be trapped in-between could be a difficult experience.
+Contemporary notions of culture emphasis its "toolkit" nature, asserting a more flexible and fluid notion of culture (CITATION).
+In this case, biracial individuals may benefit from an enlarged cultural toolkit that allows them to navigate a variety of diverse social contexts (i.e.
+code switch).
 
-In addition to the aforementioned racial mechanisms that may drive the placement of biracial individuals, researchers must also pay careful attention to how socioeconomic background may affect the placement of biracial individuals within a racial hierarchy.
-Accounting for socioeconomic difference often reduces the magnitude of observed racial inequality on a given outcome, suggesting that some of the racial inequality we observe operates through class differences (CITATION).
-This issue is particularly important for biracial individuals because of potential class selection into interracial unions.
-The likelihood of being an interracial union varies by class background for at least some groups (CITATION), although it varies in complex ways and we know more about interracial marriages than other types of unions that may produce children.
-For example, black partners tend to be positively selected with regard to education into Black/White marriages while white partners are negatively selected (CITATION).
-According to status exchange theory (CITATIONS), this pattern may be common to all types of interracial marriages within an unequal status hierarchy, although the evidence of status exchange for other forms of marriages is weaker (CITATION).
-How this affects the relative resources of biracial individuals will depend then in some part on the relative strength of these phenomena for a given pairing, the gender balance in that pairing, and how the gender of parents affects outcomes.
-
-Regardless of the underlying complexity, delineating the role of this selection process in the prior generation as opposed to other mechanisms is key to understanding the overall process that sorts biracial individuals into a racialized system.
-Although this approach is sometimes characterized as pitting race vs class and minimizing the importance of race, that is not case here.
-Instead, what this delineation allows us to do is understanding how race affects the outcomes of biracial individuals in a multigenerational process and allow us to better pinpoint the generational locus of certain mechanisms.
+The complexity of the underlying mechanisms described above prevents simple an straightforward hypothesis regarding the position of biracial individuals.
+In part, this is difficult because similar mechanisms can produce the same outcome.
+The weight of different mechanisms may also vary by groups as these groups may governed by different regimes of classification (CITATION).
+However, comparing outcomes across biracial groups may help us better understand the likely mechanisms that drive biracial classification and to disregard others as implausible.
+Accounting for the effect of parental resources on outcomes to the best of our ability is also key to reaching any conclusions about the placement of biracials within the racial hierarchy.
 

### 2022-08-31 analysis/analysis.Rmd
Use counterfactual model to get at parent resource differences; add more detailed resource tables

@@
                   "% Either parent, BA degree","% parents married")) %>%
+  kable_paper() %>%
+  row_spec(index_biracial, background = "#D3D3D3")
+```
+Lets break down parent resource stuff better. First lets do SES.
+
+```{r desc-stats-ses}
+tbl_summary <- acs %>%
+  group_by(race) %>%
+  summarize(fam_inc=round(mean(family_income), 0),
+            own_home=round(mean(own_home), 3)*100,
+            par_mar=round(mean(parents_married), 3)*100,
+            f_lhs=round(mean(degree_father=="LHS"), 3)*100,
+            f_hs=round(mean(degree_father=="HS"), 3)*100,
+            f_aa=round(mean(degree_father=="AA"), 3)*100,
+            f_ba=round(mean(degree_father=="BA" | degree_father=="G"), 3)*100,
+            m_lhs=round(mean(degree_mother=="LHS"), 3)*100,
+            m_hs=round(mean(degree_mother=="HS"), 3)*100,
+            m_aa=round(mean(degree_mother=="AA"), 3)*100,
+            m_ba=round(mean(degree_mother=="BA" | degree_mother=="G"), 3)*100)
+
+index_biracial <- which(str_detect(tbl_summary$race, "/"))
+
+tbl_summary %>%
+  mutate_if(is.numeric, format, big.mark=",") %>%
+  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.",
+      align = c("l",rep("r", ncol(tbl_summary)-1)),
+      col.names=c("race","mean family income","% own home",
+                  "% parents married",
+                  "% father, no HS diploma","% father, HS diploma",
+                  "% father, AA degree",  "% father, BA degree",
+                  "% mother, no HS diploma","% mother, HS diploma", 
+                  "%  mother, AA degree", "% mother, BA degree")) %>%
+  kable_paper() %>%
+  row_spec(index_biracial, background = "#D3D3D3")
+```
+
+```{r desc-stats-cult}
+tbl_summary <- acs %>%
+  group_by(race) %>%
+  summarize(f_born=round(mean(foreign_born), 3)*100,
+            father_fb=round(mean(foreign_born_father), 3)*100,
+            mother_fb=round(mean(foreign_born_mother), 3)*100,
+            spk_eng_no=round(mean(spk_eng=="No"), 3)*100,
+            spk_eng_some=round(mean(spk_eng=="Somewhat"), 3)*100,
+            spk_eng_yes=round(mean(spk_eng=="Yes"), 3)*100,
+            f_spk_eng_no=round(mean(spk_eng_father=="No"), 3)*100,
+            f_spk_eng_some=round(mean(spk_eng_father=="Somewhat"), 3)*100,
+            f_spk_eng_yes=round(mean(spk_eng_father=="Yes"), 3)*100,
+            m_spk_eng_no=round(mean(spk_eng_mother=="No"), 3)*100,
+            m_spk_eng_some=round(mean(spk_eng_mother=="Somewhat"), 3)*100,
+            m_spk_eng_yes=round(mean(spk_eng_mother=="Yes"), 3)*100)
+
+index_biracial <- which(str_detect(tbl_summary$race, "/"))
+
+tbl_summary %>%
+  mutate_if(is.numeric, format, big.mark=",") %>%
+  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.",
+      align = c("l",rep("r", ncol(tbl_summary)-1)),
+      col.names=c("race",
+                  "% foreign-born","% father foreign-born",
+                  "% mother foreign-born",
+                  "% speak no English", "% speak some English", 
+                  "% speak English",
+                  "% father speak no English", "% father speak some English", 
+                  "% father speak English",
+                  "% mother speak no English", "% mother speak some English",
+                  "% mother speak English")) %>%
   kable_paper() %>%
@@
 
-I want a parsimonious way to visualize differences in cultural and family resources.
-I can sort of do it by looking at the big descriptive statistics table, but it can be a little hard to summarize.
-Doing each possible variable separately would be extremely tedious here, so I am going to create scores across variables for these two measures and then model them as outcomes.
-I can then plot the results just like I do for my actual outcome measure.
+A parsimonious way of doing this would be to calculate average predicted probability for each monoracial constituent group and a given biracial group, while holding constant all non-resource variables *including race*.
+This gives you an effective counterfactual of what the average probability would be for each group if they only differed on the resource variables.
+Where to set the other variables constant to might have some effect because of the transformation from logits to probs, but should not make a big difference.
+For convenience, I might just set it to white, but I could also try setting it to the biracial group of interest.
 
-I will start by just looking at the $\alpha$ score for the correlation matrix of each case.
-I then create scores scaled to have a mean of zero and a standard deviation of one and fit these with a linear model using the same structure as for the models above.
+```{r counterfactual-ses, warning=FALSE, fig.height=12, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if there was no difference by race and the only compositional differences between groups were resources (household income, parental education, home ownership, parent and child foreign-born status, parent and child English proficiency, and whether parents are married).
+83.4% confidence bands shown around each estimate.
+Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
+# set up the initial counterfactual dataset
+counter_df <- acs %>%
+  mutate(year=2014.5, # rough midpoint
+         current_grade="6th", # rough midpoint from cumulative probability (55%)
+         state="Minnesota", # median state effect in model 4 (Minnesota)
+         metro="Metro non-central",
+         sex="Male",
+         race_real=race,
+         race="Black")
 
-## Cultural Resources
+coefs <- lapply(unique(acs$race), function(x) {
+  means_margin <- marg(model4, "race", data=subset(counter_df, race_real==x))[[1]]
+  return(tibble(term=x, estimate=means_margin$Margin, 
+                se=means_margin$Standard.Error))
+}) %>%
+  bind_rows() %>%
+  mutate(multiracial=str_detect(term, "/"))
 
-```{r cult-resource-alpha}
-cult_resources <- cbind(as.numeric(!acs$foreign_born), 
-                        as.numeric(!acs$foreign_born_father), 
-                        as.numeric(!acs$foreign_born_mother),
-                        as.numeric(factor(acs$spk_eng, 
-                                          levels=c("No","Somewhat","Yes"))),
-                        as.numeric(factor(acs$spk_eng_father, 
-                                          levels=c("No","Somewhat","Yes"))),
-                        as.numeric(factor(acs$spk_eng_mother, 
-                                          levels=c("No","Somewhat","Yes"))))
+coef_table <- map(as.list(str_subset(coefs$term, "/")), function(x) {
+    coefs %>%
+      filter(term %in% c(x, str_split(x, "/")[[1]])) %>%
+      mutate(mrace=factor(x,
+                          levels=c("Black/White",
+                                   "White/Indigenous",
+                                   "Latino/Asian",
+                                   "Black/Latino",
+                                   "White/Latino",
+                                   "Indigenous/Latino",
+                                   "Black/Asian",
+                                   "White/Asian",
+                                   "Indigenous/Asian",
+                                   "Black/Indigenous")),
+             estimate_mid=ifelse(multiracial, 
+                                 mean(estimate[!multiracial]),
+                                 estimate),
+             se_mid=ifelse(multiracial, 
+                                  sqrt(sum(se[!multiracial]^2)/4),
+                                  se))
+  }) %>%
+    bind_rows()
 
-alpha(cor(cult_resources))
-```
+  #get terms ordered
+  race_names <- levels(reorder(factor(coef_table$term), coef_table$estimate))
+  #pull out single race and multiple race as separate vectors
+  race_single <- str_subset(race_names, "/", negate=TRUE)
+  race_multi <- str_subset(race_names, "/", negate=FALSE)
+  #start with single races
+  race_full <- race_single
+  #now loop through multiple race names and append based on the higher group
+  for(race in race_multi) {
+    temp <- str_split(race, "/")[[1]]
+    idx <- max(which(race_full==temp[1]), which(race_full==temp[2]))
+    race_full <- append(race_full, race, idx-1)
+  }
 
-That all holds together well.
-However, I run into a problem when I think about making a score out of this because the foreign-born status of parents actually helps to reduce the likelihood of grade retention while all of the other variables increase it.
-For now, I am just reverse coding it, but this leads to results that are far from intuitive.
-I think the simple score approach might not work well in this case.
+  #now use this as the levels of term as a factor
+  coef_table <- coef_table %>%
+    mutate(term=factor(term, levels=race_full))
 
-```{r cult-resource-score}
-#scale them all, add them up, and then scale again
-#one catch though, I want to reverse score foreign born mother and father
-#because these reduce the likelihood of grade retention, other things
-#controlled
-acs$cult_resources <- scale(scale(as.numeric(!acs$foreign_born))+
-                              scale(as.numeric(acs$foreign_born_father))+
-                              scale(as.numeric(acs$foreign_born_mother))+
-                              scale(as.numeric(factor(acs$spk_eng, 
-                                          levels=c("No","Somewhat","Yes"))))+
-                              scale(as.numeric(factor(acs$spk_eng_father, 
-                                          levels=c("No","Somewhat","Yes"))))+
-                              scale(as.numeric(factor(acs$spk_eng_mother, 
-                                          levels=c("No","Somewhat","Yes")))))
-```
-
-```{r cult-resource-model, warning=FALSE, fig.cap="Average cultural resources for biracial respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
-Cultural resources are determined by a summated score of standardized native-born status, parents' native-born status, English proficieny, and parents' English proficiency.
-Parents' native-born status is reverse coded based on model results which show it increases grade retention risk."}
-#If I use the glm framework then I should be able to feed this back into
-#the marg command
-model_cultr <- glm(cult_resources~race+I(year-2010)*current_grade+state+metro+sex+
-                    +degree_mother+degree_father+sqrt(family_income/10000)+
-                     own_home+parents_married, 
-                  data=acs, family="gaussian")
-coef_table_cultr <- calculate_marg_means(model_cultr)
-
-ggplot(coef_table_cultr, aes(x=term, y=estimate, color=multiracial,
+ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
                        ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
@@
             color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_cultr, multiracial))+
+  geom_linerange(data=subset(coef_table, multiracial))+
   geom_point()+
@@
   scale_linetype_manual(values=c(3,1))+
+  scale_y_continuous(labels = scales::percent)+
   labs(x=NULL,
-       y="average adjusted cultural resource score (standard deviation units)")
+       y="average adjusted predicted probability of being behind expected grade")
+
 ```
 
-Because of the issue with foreign-born parents, these results are not particularly intuitive.
-It tends to show that part-Asian and part-Latino respondents have high cultural resources, but generally not higher than the highest of their constituent groups.
-Other groups have results half-way in between.
+To summarize the results here:
 
-## Family Resources
+|                   | Lower than both | Same as lower | Between | Same as higher | Higher than both |
+|-------------------|:---------------:|:-------------:|:-------:|:--------------:|:----------------:|
+| White/Asian       |       X         |               |         |                |                  |
+| Black/Asian       |                 |  X            |         |                |                  |
+| Black/Latino      |                 |  X            |         |                |                  |
+| White/Latino      |                 |  X            |         |                |                  |
+| Indigenous/Latino |                 |  X            |         |                |                  |
+| Latino/Asian      |                 |  X            |         |                |                  |
+| White/Indigenous  |                 |               |    X    |                |                  |
+| Indigenous/Asian  |                 |               |    X    |                |                  |
+| Black/Indigenous  |                 |               |    X    |                |                  |
+| Black/White       |                 |               |         |         X      |                  |
 
-```{r fam-resource-score, warning=FALSE}
-fam_resources <- cbind(acs$family_income, as.numeric(acs$degree_father), 
-                       as.numeric(acs$degree_mother), as.numeric(acs$own_home))
+In half the cases, the biracial group would have outcomes similar to the group with better resources suggesting that the parents in these matches are being positively selected.
+The White/Asian case is even more dramatic with a resource advantage over both single race groups.
+You can see this particularly for the White/Asian case in terms of family income where they have the highest of *any* group by a considerable amount.
+White/Asians are on average a very privileged group.
 
-alpha(cor(fam_resources))
+Three groups that have the "expected" in-between status based on resources all share one thing in common - they are part-Indigenous.
+That is not true of the Indigenous/Latino case, but is of all others.
+It suggests less selectivity in crossing this boundary.
 
-#scale them all, add them up, and then scale again
-acs$fam_resources <- scale(scale(acs$family_income)+scale(acs$own_home)+
-  scale(as.numeric(acs$degree_father)+scale(as.numeric(acs$degree_mother))))
-```
+Finally, the Black/White case stands out in that Black/White kids have the same risk of being held back based on their resources as do Black kids.
+Thus, they are not getting a resource boost nor does it seem like they are particularly positively selected.
 
-```{r fam-resource-model, warning=FALSE, fig.cap="Average family resources for biracial respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
-Family resources are determined by a summated score of standardized family income, parental education, and home ownership."}
-#If I use the glm framework then I should be able to feed this back into
-#the marg command
-model_famr <- glm(fam_resources~race+I(year-2010)*current_grade+state+metro+sex+
-                    foreign_born+foreign_born_mother+foreign_born_father+
-                    spk_eng+spk_eng_mother+spk_eng_father, 
-                  data=acs, family="gaussian")
-coef_table_famr <- calculate_marg_means(model_famr)
-
-ggplot(coef_table_famr, aes(x=term, y=estimate, color=multiracial,
-                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_famr, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_wrap(~mrace, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  labs(x=NULL,
-       y="average adjusted family resource score (standard deviation units)")
-```
-
-In most cases, family resources are roughly midway between those of the two monoracial groups. However, there are a couple of notable exceptions:
-
-* The average family resources of Black/White students are only marginally better than those of Black students which are far below those of the average white students.
-* The White/Asian group has family resources higher then either of the two constituent groups which already have the highest resources of any monoracial group.
-The White/Asian group has the highest family resources of any group here.
-* The Indigenous/Latino and Black/Latino groups have family resources slightly lower than either of hte constituent groups, but in both cases there are hardly any differences between the monoracial groups themselves.
-So you would not expect controlling for these variables to change much for these groups.
+So the effects of controlling for these resources will be complex and different across groups.
+In general for those that are the same as lower or lower than both, we see that it pulls their relative probability up somewhat.
+For the Black/White case, it pulls their probability down on the other hand.
+For the "between" groups its more complex and doesn't seem to do much in total.
 

### 2022-09-01 analysis/check_packages.R
Add fastDummies package

@@ -7,7 +7,8 @@
 #add new packages to the chain here
 packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr","broom",
              "ggrepel","purrr", "kableExtra", "tibble","scales","janitor",
-             "survey","modmarg","sandwich","psych","rmdformats","scales")
+             "survey","modmarg","sandwich","psych","rmdformats","scales",
+             "fastDummies")
 
 package.check <- lapply(packages, FUN = function(x) {
   if (!require(x, character.only = TRUE)) {

### 2022-09-02 analysis/analysis.Rmd
Add code chunks to create dummied dataset, run models, and test margins

@@
 
+```{r create-dummied-data}
+acs_dummied <- acs %>%
+  # first we need to unorder current grade for dummy_cols to work
+  mutate(current_grade=factor(current_grade, levels=levels(current_grade), 
+                              ordered = FALSE)) %>%
+
+  dummy_cols(remove_first_dummy = TRUE) %>%
+  select(cluster, strata, perwt, below_exp_grade, year, family_income, 
+         parents_married, own_home, degree_mother, degree_father,
+         foreign_born, foreign_born_mother, foreign_born_father, 
+         spk_eng, spk_eng_father, spk_eng_mother,
+         starts_with(c("state_","current_grade_","race","sex_","metro_"))) %>%
+  mutate(year=year-2010) %>%
+  clean_names()
+```
+
+```{r model-dummied}
+model_full <- glm(below_exp_grade~current_grade_1st*year+current_grade_2nd*year+
+                    current_grade_3rd*year+current_grade_4th*year+
+                    current_grade_5th+current_grade_6th*year+
+                    current_grade_7th*year+current_grade_8th*year+
+                    current_grade_9th*year+current_grade_10th*year+
+                    current_grade_11th*year+current_grade_12th*year+
+                    race_black+race_indigenous+race_asian+race_latino+
+                    race_black_white+race_black_indigenous+race_black_latino+
+                    race_black_asian+race_white_indigenous+race_white_latino+
+                    race_white_asian+race_indigenous_latino+
+                    race_indigenous_asian+race_latino_asian+
+                    sex_female+
+                    metro_non_metro+metro_central_city+metro_metro_non_central+
+                    metro_metro_indeterminable+state_arizona+state_arkansas+
+                    state_california+state_colorado+state_connecticut+state_dc+
+                    state_delaware+state_florida+state_georgia+state_hawaii+
+                    state_idaho+state_illinois+state_indiana+state_iowa+
+                    state_kansas+state_kentucky+state_louisiana+state_maine+
+                    state_maryland+state_massachusetts+state_michigan+
+                    state_minnesota+state_mississippi+state_missouri+
+                    state_montana+state_nebraska+state_nevada+
+                    state_new_hampshire+state_new_jersey+state_new_mexico+
+                    state_new_york+state_north_carolina+state_north_dakota+
+                    state_ohio+state_oklahoma+state_oregon+state_pennsylvania+
+                    state_rhode_island+state_south_carolina+state_south_dakota+
+                    state_tennessee+state_texas+state_utah+state_vermont+
+                    state_virginia+state_washington+state_west_virginia+
+                    state_wisconsin+state_wyoming+
+                    +sqrt(family_income/10000)+parents_married+own_home+
+                    degree_mother+degree_father+
+                    spk_eng+spk_eng_mother+spk_eng_father+
+                    foreign_born_mother+foreign_born_father+foreign_born, 
+                  data=acs_dummied, family=binomial)
+deviance(model_full)
+# FIXME - its a little higher so I am missing something
+```
+
+```{r}
+# what needs to get set at mean?
+# - current_grade
+# - sex
+# - race
+# - state
+# - metro
+# I will not do this for year because I want to use that to grab the modal
+# year of 2014
+means <- acs_dummied %>%
+  select(year, starts_with(c("state_","current_grade_","race_","sex_","metro_")))
+means <- lapply(means, function(x) {
+  rep(mean(x), length(x))
+}) %>%
+  bind_rows()
+
+cf_df <- acs_dummied %>%
+  select(below_exp_grade, race,family_income, parents_married, own_home, 
+         degree_mother, degree_father, foreign_born, foreign_born_mother, 
+         foreign_born_father, spk_eng, spk_eng_father, spk_eng_mother) %>%
+  bind_cols(means)
+```
+
+```{r}
+# I think I can use year in marg to get the results for the modal year
+temp <- subset(cf_df, race=="Black/White")
+#temp$year <- 4
+lor <- predict(model_full, newdata=temp)
+mean(exp(lor)/(1+exp(lor)))
+marg(model_full, "year", data=subset(cf_df, race=="Black/White"))
+marg(model_full, "year", data=subset(cf_df, race=="White"))
+marg(model_full, "year", data=subset(cf_df, race=="Black"))
+```
+
 ```{r counterfactual-ses, warning=FALSE, fig.height=12, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if there was no difference by race and the only compositional differences between groups were resources (household income, parental education, home ownership, parent and child foreign-born status, parent and child English proficiency, and whether parents are married).

### 2022-09-02 analysis/check_packages.R

@@ -8,7 +8,7 @@
 packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr","broom",
              "ggrepel","purrr", "kableExtra", "tibble","scales","janitor",
              "survey","modmarg","sandwich","psych","rmdformats","scales",
-             "fastDummies")
+             "fastDummies","janitor")
 
 package.check <- lapply(packages, FUN = function(x) {
   if (!require(x, character.only = TRUE)) {

### 2022-09-02 analysis/analysis.Rmd
Add missing variable (state_alaska) to dummied model

--- 2022-08-31 analysis/analysis.Rmd
@@
 
-```{r counterfactual-ses, warning=FALSE, fig.height=12, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if there was no difference by race and the only compositional differences between groups were resources (household income, parental education, home ownership, parent and child foreign-born status, parent and child English proficiency, and whether parents are married).
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
-# set up the initial counterfactual dataset
-counter_df <- acs %>%
-  mutate(year=2014.5, # rough midpoint
-         current_grade="6th", # rough midpoint from cumulative probability (55%)
-         state="Minnesota", # median state effect in model 4 (Minnesota)
-         metro="Metro non-central",
-         sex="Male",
-         race_real=race,
-         race="Black")
+```{r create-dummied-data}
+acs_dummied <- acs %>%
+  # first we need to unorder current grade for dummy_cols to work
+  mutate(current_grade=factor(current_grade, levels=levels(current_grade), 
+                              ordered = FALSE)) %>%
 
+  dummy_cols(remove_first_dummy = TRUE) %>%
+  select(cluster, strata, perwt, below_exp_grade, year, family_income, 
+         parents_married, own_home, degree_mother, degree_father,
+         foreign_born, foreign_born_mother, foreign_born_father, 
+         spk_eng, spk_eng_father, spk_eng_mother,
+         starts_with(c("state_","current_grade_","race","sex_","metro_"))) %>%
+  mutate(year=year-2010) %>%
+  clean_names()
+```
+
+```{r model-dummied}
+model_full <- glm(below_exp_grade~
+                    year*(current_grade_1st+current_grade_2nd+current_grade_3rd+
+                            current_grade_4th+current_grade_5th+
+                            current_grade_6th+current_grade_7th+
+                            current_grade_8th+current_grade_9th+
+                            current_grade_10th+current_grade_11th+
+                            current_grade_12th)+
+                    race_black+race_indigenous+race_asian+race_latino+
+                    race_black_white+race_black_indigenous+race_black_latino+
+                    race_black_asian+race_white_indigenous+race_white_latino+
+                    race_white_asian+race_indigenous_latino+
+                    race_indigenous_asian+race_latino_asian+
+                    sex_female+
+                    metro_non_metro+metro_central_city+metro_metro_non_central+
+                    metro_metro_indeterminable+
+                    state_alaska+state_arizona+state_arkansas+state_california+
+                    state_colorado+state_connecticut+state_dc+state_delaware+
+                    state_florida+state_georgia+state_hawaii+state_idaho+
+                    state_illinois+state_indiana+state_iowa+state_kansas+
+                    state_kentucky+state_louisiana+state_maine+state_maryland+
+                    state_massachusetts+state_michigan+state_minnesota+
+                    state_mississippi+state_missouri+state_montana+
+                    state_nebraska+state_nevada+state_new_hampshire+
+                    state_new_jersey+state_new_mexico+state_new_york+
+                    state_north_carolina+state_north_dakota+state_ohio+
+                    state_oklahoma+state_oregon+state_pennsylvania+
+                    state_rhode_island+state_south_carolina+state_south_dakota+
+                    state_tennessee+state_texas+state_utah+state_vermont+
+                    state_virginia+state_washington+state_west_virginia+
+                    state_wisconsin+state_wyoming+
+                    +sqrt(family_income/10000)+parents_married+own_home+
+                    degree_mother+degree_father+
+                    spk_eng+spk_eng_mother+spk_eng_father+
+                    foreign_born_mother+foreign_born_father+foreign_born, 
+                  data=acs_dummied, family=binomial)
+deviance(model_full)
+# FIXME - its a little higher so I am missing something
+```
+
+```{r create-cf-df}
+# Create the counterfactual dataset where non-resource variables are held
+# at their mean. What needs to be set at tis mean?
+# - current_grade
+# - sex
+# - race
+# - state
+# - metro
+# - year
+means <- acs_dummied %>%
+  select(year, starts_with(c("state_","current_grade_","race_","sex_",
+                             "metro_"))) %>%
+  lapply(function(x) {rep(mean(x), length(x))}) %>%
+  bind_rows()
+
+cf_df <- acs_dummied %>%
+  select(cluster, strata, perwt, below_exp_grade, race,family_income, 
+         parents_married, own_home, degree_mother, degree_father, foreign_born, 
+         foreign_born_mother, foreign_born_father, spk_eng, spk_eng_father, 
+         spk_eng_mother) %>%
+  bind_cols(means)
+```
+
+```{r test-marg, eval=FALSE}
+# This code chunk just tests to make sure I get the same value by hand as I get 
+# when I do it through the marg command.
+temp <- subset(cf_df, race=="Black/White")
+lor <- predict(model_full, newdata=temp)
+mean(exp(lor)/(1+exp(lor)))
+marg(model_full, "year", data=temp)
+# It works!
+```
+
+```{r create-cf-marg}
 coefs <- lapply(unique(acs$race), function(x) {
-  means_margin <- marg(model4, "race", data=subset(counter_df, race_real==x))[[1]]
+  temp <- subset(cf_df, race==x)
+  means_margin <- marg(model_full, "year", data=temp)[[1]]
   return(tibble(term=x, estimate=means_margin$Margin, 
@@
     mutate(term=factor(term, levels=race_full))
+```
 
+```{r counterfactual-ses, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were resources (household income, parental education, home ownership, parent and child foreign-born status, parent and child English proficiency, and whether parents are married).
+83.4% confidence bands shown around each estimate.
+Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
   ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
@@
 | White/Asian       |       X         |               |         |                |                  |
-| Black/Asian       |                 |  X            |         |                |                  |
-| Black/Latino      |                 |  X            |         |                |                  |
-| White/Latino      |                 |  X            |         |                |                  |
-| Indigenous/Latino |                 |  X            |         |                |                  |
-| Latino/Asian      |                 |  X            |         |                |                  |
+| Black/Asian       |                 |  X  (Asian)   |         |                |                  |
+| Black/Latino      |                 |  X  (Black)   |         |                |                  |
+| White/Latino      |                 |  X  (White)   |         |                |                  |
+| Indigenous/Latino |                 |X  (Indigenous)|         |                |                  |
+| Latino/Asian      |                 |  X (Asian)    |         |                |                  |
 | White/Indigenous  |                 |               |    X    |                |                  |
@@
 | Black/Indigenous  |                 |               |    X    |                |                  |
-| Black/White       |                 |               |         |         X      |                  |
+| Black/White       |                 |               |         |   X (Black)    |                  |
 

### 2022-09-02 analysis/analysis.Rmd
Refactor calculate_marg_means into a separate subfuncion of create_coef_table
This will allow me to use all of that code even when I have a different method for getting the marginal means as for the counterfactual case

@@
 }) %>%
-  bind_rows() %>%
-  mutate(multiracial=str_detect(term, "/"))
-
-coef_table <- map(as.list(str_subset(coefs$term, "/")), function(x) {
-    coefs %>%
-      filter(term %in% c(x, str_split(x, "/")[[1]])) %>%
-      mutate(mrace=factor(x,
-                          levels=c("Black/White",
-                                   "White/Indigenous",
-                                   "Latino/Asian",
-                                   "Black/Latino",
-                                   "White/Latino",
-                                   "Indigenous/Latino",
-                                   "Black/Asian",
-                                   "White/Asian",
-                                   "Indigenous/Asian",
-                                   "Black/Indigenous")),
-             estimate_mid=ifelse(multiracial, 
-                                 mean(estimate[!multiracial]),
-                                 estimate),
-             se_mid=ifelse(multiracial, 
-                                  sqrt(sum(se[!multiracial]^2)/4),
-                                  se))
-  }) %>%
   bind_rows()
 
-  #get terms ordered
-  race_names <- levels(reorder(factor(coef_table$term), coef_table$estimate))
-  #pull out single race and multiple race as separate vectors
-  race_single <- str_subset(race_names, "/", negate=TRUE)
-  race_multi <- str_subset(race_names, "/", negate=FALSE)
-  #start with single races
-  race_full <- race_single
-  #now loop through multiple race names and append based on the higher group
-  for(race in race_multi) {
-    temp <- str_split(race, "/")[[1]]
-    idx <- max(which(race_full==temp[1]), which(race_full==temp[2]))
-    race_full <- append(race_full, race, idx-1)
-  }
-
-  #now use this as the levels of term as a factor
-  coef_table <- coef_table %>%
-    mutate(term=factor(term, levels=race_full))
+coef_table_cf <- create_coef_table(coefs)
 ```
@@
 Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
-  ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
+  ggplot(coef_table_cf, aes(x=term, y=estimate, color=multiracial,
                          ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
@@
               color=NA, alpha=0.2)+
-    geom_linerange(data=subset(coef_table, multiracial))+
+    geom_linerange(data=subset(coef_table_cf, multiracial))+
     geom_point()+

### 2022-09-02 analysis/analysis.Rmd
Move full resource tables to end; add text discussion resource section

@@
                   "% Either parent, BA degree","% parents married")) %>%
-  kable_paper() %>%
-  row_spec(index_biracial, background = "#D3D3D3")
-```
-Lets break down parent resource stuff better. First lets do SES.
-
-```{r desc-stats-ses}
-tbl_summary <- acs %>%
-  group_by(race) %>%
-  summarize(fam_inc=round(mean(family_income), 0),
-            own_home=round(mean(own_home), 3)*100,
-            par_mar=round(mean(parents_married), 3)*100,
-            f_lhs=round(mean(degree_father=="LHS"), 3)*100,
-            f_hs=round(mean(degree_father=="HS"), 3)*100,
-            f_aa=round(mean(degree_father=="AA"), 3)*100,
-            f_ba=round(mean(degree_father=="BA" | degree_father=="G"), 3)*100,
-            m_lhs=round(mean(degree_mother=="LHS"), 3)*100,
-            m_hs=round(mean(degree_mother=="HS"), 3)*100,
-            m_aa=round(mean(degree_mother=="AA"), 3)*100,
-            m_ba=round(mean(degree_mother=="BA" | degree_mother=="G"), 3)*100)
-
-index_biracial <- which(str_detect(tbl_summary$race, "/"))
-
-tbl_summary %>%
-  mutate_if(is.numeric, format, big.mark=",") %>%
-  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.",
-      align = c("l",rep("r", ncol(tbl_summary)-1)),
-      col.names=c("race","mean family income","% own home",
-                  "% parents married",
-                  "% father, no HS diploma","% father, HS diploma",
-                  "% father, AA degree",  "% father, BA degree",
-                  "% mother, no HS diploma","% mother, HS diploma", 
-                  "%  mother, AA degree", "% mother, BA degree")) %>%
-  kable_paper() %>%
-  row_spec(index_biracial, background = "#D3D3D3")
-```
-
-```{r desc-stats-cult}
-tbl_summary <- acs %>%
-  group_by(race) %>%
-  summarize(f_born=round(mean(foreign_born), 3)*100,
-            father_fb=round(mean(foreign_born_father), 3)*100,
-            mother_fb=round(mean(foreign_born_mother), 3)*100,
-            spk_eng_no=round(mean(spk_eng=="No"), 3)*100,
-            spk_eng_some=round(mean(spk_eng=="Somewhat"), 3)*100,
-            spk_eng_yes=round(mean(spk_eng=="Yes"), 3)*100,
-            f_spk_eng_no=round(mean(spk_eng_father=="No"), 3)*100,
-            f_spk_eng_some=round(mean(spk_eng_father=="Somewhat"), 3)*100,
-            f_spk_eng_yes=round(mean(spk_eng_father=="Yes"), 3)*100,
-            m_spk_eng_no=round(mean(spk_eng_mother=="No"), 3)*100,
-            m_spk_eng_some=round(mean(spk_eng_mother=="Somewhat"), 3)*100,
-            m_spk_eng_yes=round(mean(spk_eng_mother=="Yes"), 3)*100)
-
-index_biracial <- which(str_detect(tbl_summary$race, "/"))
-
-tbl_summary %>%
-  mutate_if(is.numeric, format, big.mark=",") %>%
-  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.",
-      align = c("l",rep("r", ncol(tbl_summary)-1)),
-      col.names=c("race",
-                  "% foreign-born","% father foreign-born",
-                  "% mother foreign-born",
-                  "% speak no English", "% speak some English", 
-                  "% speak English",
-                  "% father speak no English", "% father speak some English", 
-                  "% father speak English",
-                  "% mother speak no English", "% mother speak some English",
-                  "% mother speak English")) %>%
   kable_paper() %>%
@@
 
-A parsimonious way of doing this would be to calculate average predicted probability for each monoracial constituent group and a given biracial group, while holding constant all non-resource variables *including race*.
-This gives you an effective counterfactual of what the average probability would be for each group if they only differed on the resource variables.
-Where to set the other variables constant to might have some effect because of the transformation from logits to probs, but should not make a big difference.
-For convenience, I might just set it to white, but I could also try setting it to the biracial group of interest.
+First, lets compare means of SES and cultural resources across all racial groups.
+
+```{r desc-stats-ses}
+tbl_summary <- acs %>%
+  group_by(race) %>%
+  summarize(fam_inc=round(mean(family_income), 0),
+            own_home=round(mean(own_home), 3)*100,
+            par_mar=round(mean(parents_married), 3)*100,
+            f_lhs=round(mean(degree_father=="LHS"), 3)*100,
+            f_hs=round(mean(degree_father=="HS"), 3)*100,
+            f_aa=round(mean(degree_father=="AA"), 3)*100,
+            f_ba=round(mean(degree_father=="BA" | degree_father=="G"), 3)*100,
+            m_lhs=round(mean(degree_mother=="LHS"), 3)*100,
+            m_hs=round(mean(degree_mother=="HS"), 3)*100,
+            m_aa=round(mean(degree_mother=="AA"), 3)*100,
+            m_ba=round(mean(degree_mother=="BA" | degree_mother=="G"), 3)*100)
+
+index_biracial <- which(str_detect(tbl_summary$race, "/"))
+
+tbl_summary %>%
+  mutate_if(is.numeric, format, big.mark=",") %>%
+  kbl(caption="SES resources by race. Shading indicates biracial group.",
+      align = c("l",rep("r", ncol(tbl_summary)-1)),
+      col.names=c("race","mean family income","% own home",
+                  "% parents married",
+                  "% father, no HS diploma","% father, HS diploma",
+                  "% father, AA degree",  "% father, BA+ degree",
+                  "% mother, no HS diploma","% mother, HS diploma", 
+                  "%  mother, AA degree", "% mother, BA+ degree")) %>%
+  kable_paper() %>%
+  row_spec(index_biracial, background = "#D3D3D3")
+```
+
+There is a lot going on here, but I can summarize a few things:
+
+- Most multiracial groups are more like the more privileged constituent group than the less privileged constituent group across all of these characteristics.
+For example, White/Latino respondents have family incomes of \$130,981 compared to \$138,489 for Whites and \$64,920 for Latinos.
+Similarly, 9.6% and 11.6% of Latino fathers and mothers, respectively, have a BA degree, compared to 43.8% and 47.6% for Whites, and 38.6% and 41.5% for White/Latinos.
+Also White/Latino parents own their home in 86.7% of cases which is similar to the white value of 85.5% and much higher than the 54.3% for Latinos.
+- Black/Whites in contrast, are much more like Blacks in terms of their resources and both groups have fewer resources than Whites.
+- White/Asians are the most privileged group of all, even more so than their two constituent groups.
+Their family income of \$195,028 is off the charts!
+but they also have parents with very high educational attainment.
+- Most part-Indigenous groups also buck the modal trend and have outcomes that are about halfway between their two constituent groups.
+The one part-Indigenous group that does not follow that pattern is the Indigenous/Latino group, which tend to have resources slightly better than both of the monoracial groups (although still very low compared to everyone else).
+
+Now lets look at the cultural characteristics.
+One thing to keep in mind here is that having foreign-born parents actually *lowers* your risk of grade retention, whereas being foreign-born yourself increases it.
+
+```{r desc-stats-cult}
+tbl_summary <- acs %>%
+  group_by(race) %>%
+  summarize(f_born=round(mean(foreign_born), 3)*100,
+            father_fb=round(mean(foreign_born_father), 3)*100,
+            mother_fb=round(mean(foreign_born_mother), 3)*100,
+            spk_eng_no=round(mean(spk_eng=="No"), 3)*100,
+            spk_eng_some=round(mean(spk_eng=="Somewhat"), 3)*100,
+            spk_eng_yes=round(mean(spk_eng=="Yes"), 3)*100,
+            f_spk_eng_no=round(mean(spk_eng_father=="No"), 3)*100,
+            f_spk_eng_some=round(mean(spk_eng_father=="Somewhat"), 3)*100,
+            f_spk_eng_yes=round(mean(spk_eng_father=="Yes"), 3)*100,
+            m_spk_eng_no=round(mean(spk_eng_mother=="No"), 3)*100,
+            m_spk_eng_some=round(mean(spk_eng_mother=="Somewhat"), 3)*100,
+            m_spk_eng_yes=round(mean(spk_eng_mother=="Yes"), 3)*100)
+
+index_biracial <- which(str_detect(tbl_summary$race, "/"))
+
+tbl_summary %>%
+  mutate_if(is.numeric, format, big.mark=",") %>%
+  kbl(caption="Cultural resources by race. Shading indicates biracial group.",
+      align = c("l",rep("r", ncol(tbl_summary)-1)),
+      col.names=c("race",
+                  "% foreign-born","% father foreign-born",
+                  "% mother foreign-born",
+                  "% speak no English", "% speak some English", 
+                  "% speak English",
+                  "% father speak no English", "% father speak some English", 
+                  "% father speak English",
+                  "% mother speak no English", "% mother speak some English",
+                  "% mother speak English")) %>%
+  kable_paper() %>%
+  row_spec(index_biracial, background = "#D3D3D3")
+```
+* In general, the most notable thing here is that in terms of English proficiency (both respondent and parents), all of the biracial groups involving Latino and Asian have almost universal proficiency (>99%) and more like the non-Latino and non-Asian groups.
+This is most notable for Latinos where only about two-thirds of fathers and mothers of monoracial Latinos have full English proficiency but all of the part Latino multiracial groups have parents that are 99% or more fluent.
+So, there is heavy selection on English proficiency into interracial unions, which is not surpising.
+* We also see that part-Latino and part-Asian respondents are much more likely to be native born and to have native-born parents than Latino and Asian respondents.
+The effects are large although not quite as huge as for language.
+However, because foreign-born parents actually *reduce* the effect of grade retention, the ultimate effect of this selection is more complex.
+
+One way to boil down the complexity of these tables into something easier to understand would be to calculate average predicted probability for each racial group, while holding constant all non-resource variables *including race*.
+These probabilities provide a counterfactual estimate of what the average probability would be for each group if they only differed on the resource variables, allowing me to isolate how different distributions of resources across racial groups affect the risk of grade retention.
+
+To do this, the following variables need to be set at some constant value to be held constant:
+
+- current grade
+- state
+- year
+- metro status
+- gender
+- race
+
+One could just plug in some static values (e.g.
+make everyone White, from Minnesota, in 6th grade, etc.), but this feels pretty ad-hoc.
+I think the best approach is to set all of these values at their mean.
+Since most of these variables are categorical variables, this means treating them as 0/1 dummies and then calculating the proportion as the mean.
+Unfortunately, this approach won't work with the models built above because they made use of factor variables.
+So, to run these models, I first need to create actual 0/1 dummies for the categorical variables listed above (current grade, state, metro status, gender, race).
+That is what the code below does.
 
@@
 ```
+
+Now that I have new "dummied up" dataset, I can rereun the full model using the dummy variables which means a very, very long formula.
 
@@
                   data=acs_dummied, family=binomial)
-deviance(model_full)
-# FIXME - its a little higher so I am missing something
-```
+```
+
+I now have the model in the correct form.
+The next step is to create the counterfactual dataset by setting all of the non-resource variables to their mean.
+Thus, in this dataset, the only differences across racial groups are coming from their resource differences.
 
@@
 
+I can then feed subsets (by race) of this counterfactual dataset into the `marg` command to get the predicted probability of grade retention for each racial group as I do below.
+
 ```{r create-cf-marg}
@@
 ```
+
+And now I can plot up the results.
 
@@
 
-To summarize the results here:
+I summarize the results here:
 
@@
 
+Lets see if I can put all of this into a simpler table that summarize the key issues.
+
 ```{r time-end}

### 2022-09-02 analysis/analysis.Rmd
Add condensed summary table of resource differences by racial groups

@@
 ```{r create-cf-marg}
-coefs <- lapply(unique(acs$race), function(x) {
+coefs_cf <- lapply(unique(acs$race), function(x) {
   temp <- subset(cf_df, race==x)
@@
 
-coef_table_cf <- create_coef_table(coefs)
+coef_table_cf <- create_coef_table(coefs_cf)
 ```
@@
 
+```{r tab-resources}
+temp <- coefs_cf %>%
+  select(term, estimate) %>%
+  mutate(estimate=round(estimate,3)*100,
+         term=as.character(term)) %>%
+  rename(race=term, held_back_cf=estimate)
+
+tbl_resources <- acs %>%
+  group_by(race) %>%
+  summarize(fam_inc=round(mean(family_income), 0),
+            own_home=round(mean(own_home), 3)*100,
+            m_ba=round(mean(degree_mother=="BA" | degree_mother=="G"), 3)*100,
+            f_ba=round(mean(degree_father=="BA" | degree_father=="G"), 3)*100,
+            f_born=round(mean(foreign_born), 3)*100,
+            par_fb=round(mean(foreign_born_father | 
+                                       foreign_born_mother), 3)*100,
+            spk_eng_well=round(mean(spk_eng=="Yes"), 3)*100,
+            spk_eng_par=round(mean(spk_eng_father=="Yes" & 
+                                     spk_eng_mother=="Yes"), 3)*100) %>%
+  mutate(race=as.character(race)) %>%
+  left_join(temp) %>%
+  arrange(held_back_cf)
+
+index_biracial <- which(str_detect(tbl_resources$race, "/"))
+
+tbl_resources %>%
+  mutate_if(is.numeric, format, big.mark=",") %>%
+  kbl(caption="Resource differences by race. Shading indicates biracial group.",
+      align = c("l",rep("r", ncol(tbl_summary)-1)),
+      col.names=c("race","mean family income","% own home",
+                  "% mother, BA degree","% father, BA degree",
+                  "% foreign-born","% either parent foreign-born",
+                  "% speak English well", "% Both parents speak english well",
+                  "% held back, counterfactual")) %>%
+  kable_paper() %>%
+  row_spec(index_biracial, background = "#D3D3D3")
+```
+
+The table sorts groups by their counterfactual probability of being held back. I think it does a nice job of summarizing the issues.
+
 ```{r time-end}

### 2022-09-02 analysis/analysis.Rmd
Add subheaders for resource differentials section

@@
 
-# Differences in Cultural and Family Resources
+# Differences in Family Resources
+
+## Full tables of SES and cultural resources
 
@@
 However, because foreign-born parents actually *reduce* the effect of grade retention, the ultimate effect of this selection is more complex.
+
+## Counterfactual models by resource differentials
 
@@
 
+### Senstivity tests separating SES and cultural differences
+
+Lets do the same procedure but I want to isolate the role of SES vs cultural (foreign-born and language).
+
+For ease of use, I am just going to plug in values here for the other resources rather than dummy everything up.
+
+```{r cf-ses-isolate, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were SES resources (household income, parental education, home ownership, and whether parents are married).
+83.4% confidence bands shown around each estimate.
+Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
+cf_df_ses <- cf_df
+cf_df_ses$foreign_born <- FALSE
+cf_df_ses$foreign_born_father <- FALSE
+cf_df_ses$foreign_born_mother <- FALSE
+cf_df_ses$spk_eng <- "Yes"
+cf_df_ses$spk_eng_father <- "Yes"
+cf_df_ses$spk_eng_mother <- "Yes"
+
+coefs_cf_ses <- lapply(unique(acs$race), function(x) {
+  temp <- subset(cf_df_ses, race==x)
+  means_margin <- marg(model_full, "year", data=temp)[[1]]
+  return(tibble(term=x, estimate=means_margin$Margin, 
+                se=means_margin$Standard.Error))
+}) %>%
+  bind_rows()
+
+coef_table_cf_ses <- create_coef_table(coefs_cf_ses)
+
+ggplot(coef_table_cf_ses, aes(x=term, y=estimate, color=multiracial,
+                          ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
+                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
+            color=NA, alpha=0.2)+
+  geom_linerange(data=subset(coef_table_cf_ses, multiracial))+
+  geom_point()+
+  coord_flip()+
+  facet_wrap(~mrace, scales="free_y")+
+  theme_bw()+
+  theme(legend.position = "none", panel.grid = element_blank())+
+  scale_color_manual(values=c("grey40","red"))+
+  scale_fill_manual(values=c("grey40","red"))+
+  scale_linetype_manual(values=c(3,1))+
+  scale_y_continuous(labels = scales::percent)+
+  labs(x=NULL,
+       y="average adjusted predicted probability of being behind expected grade")
+```
+
+```{r cf-culture-isolate, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were cultural resources (foreign-born status and English proficiency).
+83.4% confidence bands shown around each estimate.
+Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
+cf_df_culture <- cf_df
+cf_df_culture$family_income <- mean(cf_df_culture$family_income)
+cf_df_culture$own_home <- TRUE
+cf_df_culture$degree_father <- "BA"
+cf_df_culture$degree_mother <- "BA"
+cf_df_culture$parents_married <- TRUE
+
+coefs_cf_culture <- lapply(unique(acs$race), function(x) {
+  temp <- subset(cf_df_culture, race==x)
+  means_margin <- marg(model_full, "year", data=temp)[[1]]
+  return(tibble(term=x, estimate=means_margin$Margin, 
+                se=means_margin$Standard.Error))
+}) %>%
+  bind_rows()
+
+coef_table_cf_culture <- create_coef_table(coefs_cf_culture)
+
+ggplot(coef_table_cf_culture, aes(x=term, y=estimate, color=multiracial,
+                          ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
+                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
+            color=NA, alpha=0.2)+
+  geom_linerange(data=subset(coef_table_cf_culture, multiracial))+
+  geom_point()+
+  coord_flip()+
+  facet_wrap(~mrace, scales="free_y")+
+  theme_bw()+
+  theme(legend.position = "none", panel.grid = element_blank())+
+  scale_color_manual(values=c("grey40","red"))+
+  scale_fill_manual(values=c("grey40","red"))+
+  scale_linetype_manual(values=c(3,1))+
+  scale_y_continuous(labels = scales::percent)+
+  labs(x=NULL,
+       y="average adjusted predicted probability of being behind expected grade")
+```
+
+The results are a lot messier for cultural differences, partly because of the opposite effect of foreign-born parents.
+In most cases, differences point to an in-between status when there is any daylight between the constituent groups.
+So, its the SES effects that are really pulling this toward the lower group.
+
+## Summary table of resource differences
+
 Lets see if I can put all of this into a simpler table that summarize the key issues.

### 2022-09-02 analysis/analysis.Rmd
Change figure height for counterfactual figures

@@
 
-```{r counterfactual-ses, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were resources (household income, parental education, home ownership, parent and child foreign-born status, parent and child English proficiency, and whether parents are married).
+```{r counterfactual-ses, warning=FALSE, fig.height=12, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were resources (household income, parental education, home ownership, parent and child foreign-born status, parent and child English proficiency, and whether parents are married).
 83.4% confidence bands shown around each estimate.
@@
 
-```{r cf-ses-isolate, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were SES resources (household income, parental education, home ownership, and whether parents are married).
+```{r cf-ses-isolate, warning=FALSE, fig.height=12, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were SES resources (household income, parental education, home ownership, and whether parents are married).
 83.4% confidence bands shown around each estimate.
@@
 
-```{r cf-culture-isolate, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were cultural resources (foreign-born status and English proficiency).
+```{r cf-culture-isolate, warning=FALSE, fig.height=12, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were cultural resources (foreign-born status and English proficiency).
 83.4% confidence bands shown around each estimate.

### 2022-09-02 analysis/analysis.Rmd
Better labeling on state*year interaction model

@@
             "continuous, interaction"=model_inter_cont,
-            "add state*year continuous interaction"=model_inter_state),
+            "continous, interaction + state*year continuous interaction"=model_inter_state),
        BIC) %>%

### 2022-09-02 analysis/analysis.Rmd
Save the tbl_resources object

@@
   row_spec(index_biracial, background = "#D3D3D3")
+
+save(tbl_resources, file=here("analysis","output","tbl_resources.RData"))
 ```

### 2022-09-02 products/paper/main.qmd
Write rough draft of prior work section

@@
 
--   Review previous studies on outcomes
+Prior work on the outcomes for biracial individuals has been hampered by several methodological issues.
+With good reason, researchers have been reluctant to use data in which multiracial or biracial status is determined solely by the self-identification of respondents.
+We know that far fewer people identify as biracial than could feasibly do so based on their parents' races (CITATION) and that various selection effects operate on the decision to identify as biracial among hose who could (CITE DAVENPORT, ROTH, OTHERS).
+Furthermore, data on changes in racial reporting over time suggests that other status markers like education and income may influence how people identify themselves (CITE SAPERSTEIN AND CO).
+For example, a person with one Black and one White parent who is routinely treated by others as Black and discriminated against as Black may be more likely to identify themselves as exclusively Black than a similar person of Black and White ancestry who experiences less consistent discrimination.
+Thus, any analysis of outcomes for biracial individuals based on self-identification must deal with a serious potential for reverse causation in the observed differences between biracial individuals and their constituent groups (CITE BRATTER).
 
-    -   Campbell
-    -   Herman
-    -   Holloway/Ellis stuff
-    -   Udry? - also related stuff by Cheng? How does Cheng and Powell fit in here?
-    -   Do some more bibliographic work, particularly forward searching to see if I am missing something big here.
-    -   Limitations of this work
-        -   Small sample sizes lead to inconsistent estimates in most of the survey data.
-        -   In many cases, impossible to understand outcomes for part-Latino biracials.
+Instead of self-identification, researchers have generally relied upon survey data in which the racial identification of a respondent's biological parents are provided in some form or another.
+CITE CAMPBELL used data from the National Longitudinal Survey of Adolescent to Adult Health (Add Health) to examine high school GPA, advanced placement in math, and four-year college enrollment among biracial and monoracial individuals identified by co-resident biological parents' races.
+CITE HERMAN used data on high school students in California and Wisconsin in which the students reported their parents' races to examine differences in GPA.
+Both of these studies find some support for an "in-between" status for many biracial groups, but are both hampered by a lack of statistical power due to small sample sizes.
+CITE KAO uses data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial Blacks and Asians, and finds that biracial Blacks have outcomes similar to monoracial Blacks, while biracial Asians have outcomes more similar to monoracial Whites.
+This study, however, is hampered by the fact that biracial students are identified by a discrepancy between their reported race and one parent's reported race, rather than by both biological parent's races.
+
+The primary difficulty with the survey data used in the preceding studies is that the sample size for most biracial groups will be small and thus results tend to be imprecise and inconclusive.
+An alternative approach is to use large scale Census or American Community Survey (ACS) data to identify biracial children by the race of the biological parents in their household.
+This approach can generate much larger samples of biracial individuals, although it does exclude examination of children living in non-nuclear households.
+Furthermore, researchers then have a limited range of outcomes to explore because biracial individuals identified by this method are by definition children and the data sources includes relatively little detailed information.
+This approach has been used previously to examine the likelihood of living in poverty (CITATIONS) and residential segregation (CITATIONS).
+SUMMARIZE FINDINGS.
+While these findings provide valuable information of the context of biracial childrens' lived experience, but do not directly measure their own outcomes, but rather the experiences of their parents.
+
+However, one important educational outcome is identifiable for children in Census and ACS data - grade retention.
+Grade retention is partially identifiable in these data sources and provides one of the earliest measurable forms of stratification that individuals face in their lives.
 

### 2022-09-03 analysis/analysis.Rmd
Change all models to use survey design effects

@@
 
+```{r create-design}
+# strata make the models take forever, but do not seem to change anything
+# so I will leave them out
+acs_design <- svydesign(ids=~cluster, weights=~perwt, data=acs)
+```
+
 ```{r build-models, results='asis'}
-model0 <- glm(below_exp_grade~race, data=acs, family=binomial)
+model0 <- svyglm(below_exp_grade~race, design=acs_design, family=binomial)
 model1 <- update(model0, 
@@
                             marg(x, "race", type="effects", 
-                                 vcov_mat=vcovCL(x, cluster=acs$cluster),
                                  at_var_interest=c("White","Black","Indigenous",
@@
 #run separate models by grade groups
-model4_elem <- update(model4, data=subset(acs, current_grade<="5th"))
-model4_middle <- update(model4, data=subset(acs, current_grade>"5th" & 
+model4_elem <- update(model4, design=subset(acs_design, current_grade<="5th"))
+model4_middle <- update(model4, design=subset(acs_design, current_grade>"5th" & 
                                         current_grade<"9th"))
-model4_hs <- update(model4, data=subset(acs, current_grade>="9th"))
-
-coef_table_elem <- calculate_marg_means(model4_elem, 
-                                        cluster_var=subset(acs, current_grade<="5th")$cluster)
-coef_table_middle <- calculate_marg_means(model4_middle,
-                                          cluster_var=subset(acs, current_grade>"5th" & 
-                                        current_grade<"9th")$cluster)
-coef_table_hs <- calculate_marg_means(model4_hs,
-                                      cluster_var=subset(acs, current_grade>="9th")$cluster)
+model4_hs <- update(model4, design=subset(acs_design, current_grade>="9th"))
+
+coef_table_elem <- calculate_marg_means(model4_elem)
+coef_table_middle <- calculate_marg_means(model4_middle)
+coef_table_hs <- calculate_marg_means(model4_hs)
 
@@
 #run separate models by grade groups
-model4_best_link <- update(model4, data=subset(acs, momrule==11 & poprule==11))
-
-coef_table_best_link <- calculate_marg_means(model4_best_link, 
-                                        cluster_var=subset(acs, momrule==11 &
-                                                             poprule==11)$cluster)
+model4_best_link <- update(model4, design=subset(acs_design, momrule==11 & poprule==11))
+
+coef_table_best_link <- calculate_marg_means(model4_best_link)
 
@@
 ```{r model-dummied}
-model_full <- glm(below_exp_grade~
+acs_dummied_design <- svydesign(ids=~cluster, weights=~perwt, data=acs_dummied)
+
+model_full <- svyglm(below_exp_grade~
                        year*(current_grade_1st+current_grade_2nd+current_grade_3rd+
@@
                        foreign_born_mother+foreign_born_father+foreign_born, 
-                  data=acs_dummied, family=binomial)
+                     design=acs_dummied_design, family=binomial)
 ```

### 2022-09-03 analysis/analysis.Rmd
Specify weights manually in marg for counterfactual models
This is necessary because otherwise it tries to use the full weights which don't work for the new dataset

@@
 lor <- predict(model_full, newdata=temp)
-mean(exp(lor)/(1+exp(lor)))
-marg(model_full, "year", data=temp)
-# It works!
+weighted.mean(exp(lor)/(1+exp(lor)), w=temp$perwt)
+marg(model_full, "year", data=temp, weights=temp$perwt)
+# It works
 ```
@@
   temp <- subset(cf_df, race==x)
-  means_margin <- marg(model_full, "year", data=temp)[[1]]
+  means_margin <- marg(model_full, "year", data=temp, weights=temp$perwt)[[1]]
   return(tibble(term=x, estimate=means_margin$Margin, 
@@
 
-```{r counterfactual-ses, warning=FALSE, fig.height=12, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were resources (household income, parental education, home ownership, parent and child foreign-born status, parent and child English proficiency, and whether parents are married).
+```{r counterfactual-ses, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were resources (household income, parental education, home ownership, parent and child foreign-born status, parent and child English proficiency, and whether parents are married).
 83.4% confidence bands shown around each estimate.
@@
 
-```{r cf-ses-isolate, warning=FALSE, fig.height=12, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were SES resources (household income, parental education, home ownership, and whether parents are married).
+```{r cf-ses-isolate, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were SES resources (household income, parental education, home ownership, and whether parents are married).
 83.4% confidence bands shown around each estimate.
@@
 
-```{r cf-culture-isolate, warning=FALSE, fig.height=12, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were cultural resources (foreign-born status and English proficiency).
+```{r cf-culture-isolate, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were cultural resources (foreign-born status and English proficiency).
 83.4% confidence bands shown around each estimate.

### 2022-09-03 analysis/analysis.Rmd
Use survey weighted values in all tables

@@
 load(here("analysis","output","acs.RData"))
+```
+
+```{r create-design}
+# strata make the models take forever, but do not seem to change anything
+# so I will leave them out
+acs_design <- acs %>%
+  as_survey_design(ids = cluster, weights = perwt)
 ```
@@
 
-Lets looks at sample characteristics by race.
+Lets looks at sample size and grade retention by race.
 
 ```{r desc-stats}
-tbl_summary <- acs %>%
+tbl_summary <- acs_design %>%
   group_by(race) %>%
   summarize(n=n(), 
-            held_back=round(mean(below_exp_grade), 3)*100,
-            mean_age=round(mean(age), 1),
-            f_born=round(mean(foreign_born), 3)*100,
-            spk_eng_well=round(mean(spk_eng=="Yes"), 3)*100,
-            spk_eng_par=round(mean(spk_eng_father=="Yes" & 
-                                     spk_eng_mother=="Yes"), 3)*100,
-            par_fb=round(mean(foreign_born_father | 
-                                       foreign_born_mother), 3)*100,
-            fam_inc=round(mean(family_income), 0),
-            own_home=round(mean(own_home), 3)*100,
-            par_hs=round(mean(degree_father!="LHS" | 
-                                degree_mother!="LHS"), 3)*100,
-            par_ba=round(mean(degree_father=="BA" | 
-                                degree_father=="G" | 
-                                degree_mother=="BA" | 
-                                degree_mother=="G"), 3)*100,
-            par_mar=round(mean(parents_married), 3)*100) %>%
+            held_back=round(survey_mean(below_exp_grade, vartype=NULL), 3)*100,
+            mean_age=round(survey_mean(age, vartype=NULL), 1)) %>%
   arrange(desc(n))
@@
   mutate_if(is.numeric, format, big.mark=",") %>%
-  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.",
+  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size. Survey weights applied.",
       align = c("l",rep("r", ncol(tbl_summary)-1)),
-      col.names=c("race","n","% below expected grade", "mean age",
-                  "% foreign-born","% speak English well",
-                  "% Both parents speak english well","% either parent foreign-born",
-                  "mean family income","% own home","% either parent, HS diploma",
-                  "% Either parent, BA degree","% parents married")) %>%
+      col.names=c("race","n","% below expected grade", "mean age")) %>%
   kable_paper() %>%
@@
 
-All models are estimated as logit models. Warning, the code chunks below take a bit of time to run!
-
-```{r create-design}
-# strata make the models take forever, but do not seem to change anything
-# so I will leave them out
-acs_design <- svydesign(ids=~cluster, weights=~perwt, data=acs)
-```
+All models are estimated as logit models.
+I use the `survey` library to adjust for variable sample weights and clustering of respondents in the same household.
+Warning, the code chunks below take a bit of time to run!
 
@@
 Standard errors for these average adjusted predictions are calculated by the [delta method](https://cran.r-project.org/web/packages/modmarg/vignettes/delta-method.html).
-I also adjust all SEs for clustering of respondents within households.
+I also adjust all SEs for clustering of respondents within households and variance in person weights.
 
@@
 ```{r desc-stats-ses}
-tbl_summary <- acs %>%
+tbl_summary <- acs_design %>%
   group_by(race) %>%
-  summarize(fam_inc=round(mean(family_income), 0),
-            own_home=round(mean(own_home), 3)*100,
-            par_mar=round(mean(parents_married), 3)*100,
-            f_lhs=round(mean(degree_father=="LHS"), 3)*100,
-            f_hs=round(mean(degree_father=="HS"), 3)*100,
-            f_aa=round(mean(degree_father=="AA"), 3)*100,
-            f_ba=round(mean(degree_father=="BA" | degree_father=="G"), 3)*100,
-            m_lhs=round(mean(degree_mother=="LHS"), 3)*100,
-            m_hs=round(mean(degree_mother=="HS"), 3)*100,
-            m_aa=round(mean(degree_mother=="AA"), 3)*100,
-            m_ba=round(mean(degree_mother=="BA" | degree_mother=="G"), 3)*100)
+  summarize(fam_inc=survey_mean(family_income, vartype = NULL))
+
+tbl_summary <- acs_design %>%
+  group_by(race) %>%
+  summarize(fam_inc=round(survey_mean(family_income, vartype = NULL), 0),
+            own_home=round(survey_mean(own_home, vartype = NULL), 3)*100,
+            par_mar=round(survey_mean(parents_married, vartype = NULL), 3)*100,
+            f_lhs=round(survey_mean(degree_father=="LHS", vartype = NULL), 
+                        3)*100,
+            f_hs=round(survey_mean(degree_father=="HS", vartype = NULL), 3)*100,
+            f_aa=round(survey_mean(degree_father=="AA", vartype = NULL), 3)*100,
+            f_ba=round(survey_mean(degree_father=="BA" | degree_father=="G", 
+                                   vartype = NULL), 3)*100,
+            m_lhs=round(survey_mean(degree_mother=="LHS", vartype = NULL), 3)*100,
+            m_hs=round(survey_mean(degree_mother=="HS", vartype = NULL), 3)*100,
+            m_aa=round(survey_mean(degree_mother=="AA", vartype = NULL), 3)*100,
+            m_ba=round(survey_mean(degree_mother=="BA" | degree_mother=="G", 
+                                   vartype = NULL), 3)*100)
 
@@
 ```{r desc-stats-cult}
-tbl_summary <- acs %>%
+tbl_summary <- acs_design %>%
   group_by(race) %>%
-  summarize(f_born=round(mean(foreign_born), 3)*100,
-            father_fb=round(mean(foreign_born_father), 3)*100,
-            mother_fb=round(mean(foreign_born_mother), 3)*100,
-            spk_eng_no=round(mean(spk_eng=="No"), 3)*100,
-            spk_eng_some=round(mean(spk_eng=="Somewhat"), 3)*100,
-            spk_eng_yes=round(mean(spk_eng=="Yes"), 3)*100,
-            f_spk_eng_no=round(mean(spk_eng_father=="No"), 3)*100,
-            f_spk_eng_some=round(mean(spk_eng_father=="Somewhat"), 3)*100,
-            f_spk_eng_yes=round(mean(spk_eng_father=="Yes"), 3)*100,
-            m_spk_eng_no=round(mean(spk_eng_mother=="No"), 3)*100,
-            m_spk_eng_some=round(mean(spk_eng_mother=="Somewhat"), 3)*100,
-            m_spk_eng_yes=round(mean(spk_eng_mother=="Yes"), 3)*100)
+  summarize(f_born=round(survey_mean(foreign_born, vartype = NULL), 3)*100,
+            father_fb=round(survey_mean(foreign_born_father, 
+                                        vartype = NULL), 3)*100,
+            mother_fb=round(survey_mean(foreign_born_mother, 
+                                        vartype = NULL), 3)*100,
+            spk_eng_no=round(survey_mean(spk_eng=="No", vartype = NULL), 3)*100,
+            spk_eng_some=round(survey_mean(spk_eng=="Somewhat", 
+                                           vartype = NULL), 3)*100,
+            spk_eng_yes=round(survey_mean(spk_eng=="Yes", vartype = NULL), 3)*100,
+            f_spk_eng_no=round(survey_mean(spk_eng_father=="No", 
+                                           vartype = NULL), 3)*100,
+            f_spk_eng_some=round(survey_mean(spk_eng_father=="Somewhat", 
+                                             vartype = NULL), 3)*100,
+            f_spk_eng_yes=round(survey_mean(spk_eng_father=="Yes", 
+                                            vartype = NULL), 3)*100,
+            m_spk_eng_no=round(survey_mean(spk_eng_mother=="No", 
+                                           vartype = NULL), 3)*100,
+            m_spk_eng_some=round(survey_mean(spk_eng_mother=="Somewhat", 
+                                             vartype = NULL), 3)*100,
+            m_spk_eng_yes=round(survey_mean(spk_eng_mother=="Yes", 
+                                            vartype = NULL), 3)*100)
 
@@
 
-tbl_resources <- acs %>%
+tbl_resources <- acs_design %>%
   group_by(race) %>%
-  summarize(fam_inc=round(mean(family_income), 0),
-            own_home=round(mean(own_home), 3)*100,
-            m_ba=round(mean(degree_mother=="BA" | degree_mother=="G"), 3)*100,
-            f_ba=round(mean(degree_father=="BA" | degree_father=="G"), 3)*100,
-            f_born=round(mean(foreign_born), 3)*100,
-            par_fb=round(mean(foreign_born_father | 
-                                       foreign_born_mother), 3)*100,
-            spk_eng_well=round(mean(spk_eng=="Yes"), 3)*100,
-            spk_eng_par=round(mean(spk_eng_father=="Yes" & 
-                                     spk_eng_mother=="Yes"), 3)*100) %>%
+  summarize(fam_inc=round(survey_mean(family_income, vartype=NULL), 0),
+            own_home=round(survey_mean(own_home, vartype=NULL), 3)*100,
+            m_ba=round(survey_mean(degree_mother=="BA" | degree_mother=="G", 
+                                   vartype=NULL), 3)*100,
+            f_ba=round(survey_mean(degree_father=="BA" | degree_father=="G", 
+                                   vartype=NULL), 3)*100,
+            f_born=round(survey_mean(foreign_born, vartype=NULL), 3)*100,
+            par_fb=round(survey_mean(foreign_born_father | foreign_born_mother, 
+                                     vartype=NULL), 3)*100,
+            spk_eng_well=round(survey_mean(spk_eng=="Yes",
+                                           vartype=NULL), 3)*100,
+            spk_eng_par=round(survey_mean(spk_eng_father=="Yes" & 
+                                     spk_eng_mother=="Yes", vartype=NULL), 
+                              3)*100) %>%
   mutate(race=as.character(race)) %>%
@@
   mutate_if(is.numeric, format, big.mark=",") %>%
-  kbl(caption="Resource differences by race. Shading indicates biracial group.",
-      align = c("l",rep("r", ncol(tbl_summary)-1)),
+  kbl(caption="Resource differences by race. Shading indicates biracial group. All results are survey weighted.",
+      align = c("l",rep("r", ncol(tbl_resources)-1)),
       col.names=c("race","mean family income","% own home",

### 2022-09-03 analysis/check_packages.R

@@ -8,7 +8,7 @@
 packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr","broom",
              "ggrepel","purrr", "kableExtra", "tibble","scales","janitor",
              "survey","modmarg","sandwich","psych","rmdformats","scales",
-             "fastDummies","janitor")
+             "fastDummies","janitor","srvyr")
 
 package.check <- lapply(packages, FUN = function(x) {
   if (!require(x, character.only = TRUE)) {

### 2022-09-03 analysis/analysis.Rmd
Forgot manual weights on two marg commands

@@
   temp <- subset(cf_df_ses, race==x)
-  means_margin <- marg(model_full, "year", data=temp)[[1]]
+  means_margin <- marg(model_full, "year", data=temp, weights=temp$perwt)[[1]]
   return(tibble(term=x, estimate=means_margin$Margin, 
@@
   temp <- subset(cf_df_culture, race==x)
-  means_margin <- marg(model_full, "year", data=temp)[[1]]
+  means_margin <- marg(model_full, "year", data=temp, weights=temp$perwt)[[1]]
   return(tibble(term=x, estimate=means_margin$Margin, 

### 2022-09-04 analysis/analysis_main.Rmd
Refactor analysis.Rmd into three separate Rmd files

--- 2022-09-03 analysis/analysis.Rmd
@@
 ---
-title: "Analysis for Biracial Grade Retention Project"
+title: "Main Analysis for Biracial Grade Retention Project"
 output: 
@@
 
-# Grade retention summary
-
-OK, first lets just confirm that the grade retention variable is working as expected on our sample. I start by looking at the number of kids by age and current grade:
-
-```{r tab-grade-by-age}
-acs %>% 
-  tabyl(age, current_grade) %>%
-  mutate_if(is.numeric, format, big.mark=",") %>%
-  kbl(caption="Cross-tabulation of age and current grade of student",
-      align=c("l",rep("r",13))) %>%
-  kable_paper() %>%
-  column_spec(2, background=c(rep("#D3D3D3",2), rep("white", 14))) %>%
-  column_spec(3, background=c("white", 
-                              rep("#D3D3D3",2), 
-                              rep("white", 13))) %>%
-  column_spec(4, background=c(rep("white", 2), 
-                              rep("#D3D3D3",2), 
-                              rep("white", 12))) %>%
-  column_spec(5, background=c(rep("white", 3), 
-                              rep("#D3D3D3",2), 
-                              rep("white", 11))) %>%
-  column_spec(6, background=c(rep("white", 4), 
-                              rep("#D3D3D3",2), 
-                              rep("white", 10))) %>%
-  column_spec(7, background=c(rep("white", 5), 
-                              rep("#D3D3D3",2), 
-                              rep("white", 9))) %>%
-  column_spec(8, background=c(rep("white", 6), 
-                              rep("#D3D3D3",2), 
-                              rep("white", 8))) %>%
-  column_spec(9, background=c(rep("white", 7), 
-                              rep("#D3D3D3",2), 
-                              rep("white", 7))) %>%
-  column_spec(10, background=c(rep("white", 8), 
-                               rep("#D3D3D3",2), 
-                               rep("white", 6))) %>%
-  column_spec(11, background=c(rep("white", 9), 
-                               rep("#D3D3D3",2), 
-                               rep("white", 5))) %>%
-  column_spec(12, background=c(rep("white", 10), 
-                               rep("#D3D3D3",2), 
-                               rep("white", 4))) %>%
-  column_spec(13, background=c(rep("white", 11), 
-                               rep("#D3D3D3",2), 
-                               rep("white", 3))) %>%
-  column_spec(14, background=c(rep("white", 12), 
-                               rep("#D3D3D3",2), 
-                               rep("white", 2)))
-```
-
-Ok, this looks more or less like what I would expect.
-We observe two modal ages for each grade.
-Some of the more extreme age values for older grades seem a little strange given that these are only supposed to be currently enrolled students.
-20-year old sixth graders?
-I mean there are relatively few of these cases but still seems a bit odd.
-Its also interesting how this really starts expanding past 5th - did the Census Bureau "correct" values below 6th or something?
-
-Now lets calculate the percentage of cases TRUE on the dependent variable across this same table, just to confirm its all working correctly.
-
-```{r check-retention-code, fig.cap="The proportion of students measured at below grade level by age and current grade.
-This graph is used to ensure that the dependent variable is calculated correctly."}
-acs %>%
-  group_by(age, current_grade) %>%
-  summarize(below=mean(below_exp_grade)) %>%
-  ggplot(aes(x=current_grade, y=age, fill=below))+
-  geom_tile(color="grey20")+
-  geom_text(aes(label=below))+
-  theme_bw()+
-  scale_fill_viridis_c()+
-  scale_y_continuous(breaks=5:20, minor_breaks = NULL)+
-  labs(x="current grade", y="age", fill="proportion below grade level")
-```
-
-Ok, that looks good.
-
-What do the grade retention probabilities look like by grade:
-
-```{r retention-by-grade, fig.cap="Percent of students clearly behind grade level by current grade. Blue lines fit a smoothed curve to the data."}
-acs %>%
-  group_by(current_grade) %>%
-  summarize(below_grade=mean(below_exp_grade),
-            n=n(),
-            se=sqrt((below_grade*(1-below_grade))/n)) %>%
-  ggplot(aes(x=current_grade, y=below_grade, group=1,
-             ymin=below_grade-1.96*se, ymax=below_grade+1.96*se))+
-  geom_smooth(se=FALSE)+
-  geom_line(linetype=2, color="grey30")+
-  geom_point()+
-  geom_linerange()+
-  scale_y_continuous(labels=scales::percent)+
-  labs(x="current grade", y="percent clearly behind grade level for age")+
-  theme_bw()
-```
-
-I guess those numbers look pretty reasonable.
-Its hard to benchmark it off anything.
-Warren et.
-al.
-estimate annual retention rates by grade but that is different than what I am measuring here because this is cumulative (but can also reflect students being caught up).
-In general, these must be underestimates because they do not capture kids who were held back but are still within the modal age.
-
-What about by year?
-
-```{r retention-by-year, fig.cap="Trends over time in the percent of students clearly behind grade level. Blue lines fit a smoothed curve to the data."}
-acs %>%
-  group_by(year) %>%
-  summarize(below_grade=mean(below_exp_grade)) %>%
-  ggplot(aes(x=year, y=below_grade))+
-  geom_line()+
-  geom_point()+
-  geom_smooth(se=FALSE)+
-  scale_x_continuous(breaks=c(2010,2014,2018))+
-  scale_y_continuous(labels=scales::percent)+
-  labs(x="current grade", y="percent clearly behind grade level for age")+
-  theme_bw()
-```
-
-Overall, there has been some decline over time in the prevalence of grade retention.
-
-Lets look at this by grade and year
-
-```{r retention-by-year-grade, fig.cap="Trends over time and grade in the percent of students clearly behind grade level. Lines are calculated by smoothing."}
-acs %>%
-  group_by(current_grade, year) %>%
-  summarize(below_grade=mean(below_exp_grade)) %>%
-  ggplot(aes(x=year, y=below_grade, group=current_grade, color=current_grade))+
-  #geom_line()+
-  geom_smooth(se=FALSE)+
-  geom_point()+
-  scale_x_continuous(breaks=c(2010,2014,2018))+
-  scale_y_continuous(labels=scales::percent)+
-  labs(x="current grade", y="percent clearly behind grade level for age")+
-  theme_bw()
-```
-
-The evidence here suggests that the decline is driven by the higher grades where grade retention was higher overall.
-
 # Descriptive statistics
@@
 # Model building
-
-## Functional form of year-grade interactions
-
-Before running the full models, I want to test out how I can parsimoniously address differences in the time trends by grade.
-
-```{r models-year-grade}
-model_add_cat <- glm(below_exp_grade~as.factor(year)+current_grade+state, 
-                     data=acs, family=binomial)
-model_add_cont <- glm(below_exp_grade~I(year-2010)+current_grade+state, 
-                     data=acs, family=binomial)
-model_inter_cont <- glm(below_exp_grade~I(year-2010)*current_grade+state, 
-                     data=acs, family=binomial)
-model_inter_cat <- glm(below_exp_grade~as.factor(year)*current_grade+state, 
-                     data=acs, family=binomial)
-model_inter_state <- glm(below_exp_grade~I(year-2010)*current_grade+
-                            I(year-2010)*state, 
-                     data=acs, family=binomial)
-
-sapply(list("categorical, additive"=model_add_cat,
-            "categorical, interaction"=model_inter_cat, 
-            "continuous, additive"=model_add_cont, 
-            "continuous, interaction"=model_inter_cont,
-            "continous, interaction + state*year continuous interaction"=model_inter_state),
-       BIC) %>%
-    enframe("model","BIC") %>%
-  kbl() %>%
-  kable_styling(full_width=FALSE)
-
-# free some memory
-rm(model_add_cat, model_add_cont, model_inter_cat, model_inter_cont,
-   model_inter_state)
-```
-
-The preferred model by BIC is clearly the model that fits year as a continuous measure and interacts it with current grade, so that is what I will do.
 
@@
 The results here are extraordinarily stable across both specifications.
-
-# Differences in Family Resources
-
-## Full tables of SES and cultural resources
-
-First, lets compare means of SES and cultural resources across all racial groups.
-
-```{r desc-stats-ses}
-tbl_summary <- acs_design %>%
-  group_by(race) %>%
-  summarize(fam_inc=survey_mean(family_income, vartype = NULL))
-
-tbl_summary <- acs_design %>%
-  group_by(race) %>%
-  summarize(fam_inc=round(survey_mean(family_income, vartype = NULL), 0),
-            own_home=round(survey_mean(own_home, vartype = NULL), 3)*100,
-            par_mar=round(survey_mean(parents_married, vartype = NULL), 3)*100,
-            f_lhs=round(survey_mean(degree_father=="LHS", vartype = NULL), 
-                        3)*100,
-            f_hs=round(survey_mean(degree_father=="HS", vartype = NULL), 3)*100,
-            f_aa=round(survey_mean(degree_father=="AA", vartype = NULL), 3)*100,
-            f_ba=round(survey_mean(degree_father=="BA" | degree_father=="G", 
-                                   vartype = NULL), 3)*100,
-            m_lhs=round(survey_mean(degree_mother=="LHS", vartype = NULL), 3)*100,
-            m_hs=round(survey_mean(degree_mother=="HS", vartype = NULL), 3)*100,
-            m_aa=round(survey_mean(degree_mother=="AA", vartype = NULL), 3)*100,
-            m_ba=round(survey_mean(degree_mother=="BA" | degree_mother=="G", 
-                                   vartype = NULL), 3)*100)
-
-index_biracial <- which(str_detect(tbl_summary$race, "/"))
-
-tbl_summary %>%
-  mutate_if(is.numeric, format, big.mark=",") %>%
-  kbl(caption="SES resources by race. Shading indicates biracial group.",
-      align = c("l",rep("r", ncol(tbl_summary)-1)),
-      col.names=c("race","mean family income","% own home",
-                  "% parents married",
-                  "% father, no HS diploma","% father, HS diploma",
-                  "% father, AA degree",  "% father, BA+ degree",
-                  "% mother, no HS diploma","% mother, HS diploma", 
-                  "%  mother, AA degree", "% mother, BA+ degree")) %>%
-  kable_paper() %>%
-  row_spec(index_biracial, background = "#D3D3D3")
-```
-
-There is a lot going on here, but I can summarize a few things:
-
-- Most multiracial groups are more like the more privileged constituent group than the less privileged constituent group across all of these characteristics.
-For example, White/Latino respondents have family incomes of \$130,981 compared to \$138,489 for Whites and \$64,920 for Latinos.
-Similarly, 9.6% and 11.6% of Latino fathers and mothers, respectively, have a BA degree, compared to 43.8% and 47.6% for Whites, and 38.6% and 41.5% for White/Latinos.
-Also White/Latino parents own their home in 86.7% of cases which is similar to the white value of 85.5% and much higher than the 54.3% for Latinos.
-- Black/Whites in contrast, are much more like Blacks in terms of their resources and both groups have fewer resources than Whites.
-- White/Asians are the most privileged group of all, even more so than their two constituent groups.
-Their family income of \$195,028 is off the charts!
-but they also have parents with very high educational attainment.
-- Most part-Indigenous groups also buck the modal trend and have outcomes that are about halfway between their two constituent groups.
-The one part-Indigenous group that does not follow that pattern is the Indigenous/Latino group, which tend to have resources slightly better than both of the monoracial groups (although still very low compared to everyone else).
-
-Now lets look at the cultural characteristics.
-One thing to keep in mind here is that having foreign-born parents actually *lowers* your risk of grade retention, whereas being foreign-born yourself increases it.
-
-```{r desc-stats-cult}
-tbl_summary <- acs_design %>%
-  group_by(race) %>%
-  summarize(f_born=round(survey_mean(foreign_born, vartype = NULL), 3)*100,
-            father_fb=round(survey_mean(foreign_born_father, 
-                                        vartype = NULL), 3)*100,
-            mother_fb=round(survey_mean(foreign_born_mother, 
-                                        vartype = NULL), 3)*100,
-            spk_eng_no=round(survey_mean(spk_eng=="No", vartype = NULL), 3)*100,
-            spk_eng_some=round(survey_mean(spk_eng=="Somewhat", 
-                                           vartype = NULL), 3)*100,
-            spk_eng_yes=round(survey_mean(spk_eng=="Yes", vartype = NULL), 3)*100,
-            f_spk_eng_no=round(survey_mean(spk_eng_father=="No", 
-                                           vartype = NULL), 3)*100,
-            f_spk_eng_some=round(survey_mean(spk_eng_father=="Somewhat", 
-                                             vartype = NULL), 3)*100,
-            f_spk_eng_yes=round(survey_mean(spk_eng_father=="Yes", 
-                                            vartype = NULL), 3)*100,
-            m_spk_eng_no=round(survey_mean(spk_eng_mother=="No", 
-                                           vartype = NULL), 3)*100,
-            m_spk_eng_some=round(survey_mean(spk_eng_mother=="Somewhat", 
-                                             vartype = NULL), 3)*100,
-            m_spk_eng_yes=round(survey_mean(spk_eng_mother=="Yes", 
-                                            vartype = NULL), 3)*100)
-
-index_biracial <- which(str_detect(tbl_summary$race, "/"))
-
-tbl_summary %>%
-  mutate_if(is.numeric, format, big.mark=",") %>%
-  kbl(caption="Cultural resources by race. Shading indicates biracial group.",
-      align = c("l",rep("r", ncol(tbl_summary)-1)),
-      col.names=c("race",
-                  "% foreign-born","% father foreign-born",
-                  "% mother foreign-born",
-                  "% speak no English", "% speak some English", 
-                  "% speak English",
-                  "% father speak no English", "% father speak some English", 
-                  "% father speak English",
-                  "% mother speak no English", "% mother speak some English",
-                  "% mother speak English")) %>%
-  kable_paper() %>%
-  row_spec(index_biracial, background = "#D3D3D3")
-```
-* In general, the most notable thing here is that in terms of English proficiency (both respondent and parents), all of the biracial groups involving Latino and Asian have almost universal proficiency (>99%) and more like the non-Latino and non-Asian groups.
-This is most notable for Latinos where only about two-thirds of fathers and mothers of monoracial Latinos have full English proficiency but all of the part Latino multiracial groups have parents that are 99% or more fluent.
-So, there is heavy selection on English proficiency into interracial unions, which is not surpising.
-* We also see that part-Latino and part-Asian respondents are much more likely to be native born and to have native-born parents than Latino and Asian respondents.
-The effects are large although not quite as huge as for language.
-However, because foreign-born parents actually *reduce* the effect of grade retention, the ultimate effect of this selection is more complex.
-
-## Counterfactual models by resource differentials
-
-One way to boil down the complexity of these tables into something easier to understand would be to calculate average predicted probability for each racial group, while holding constant all non-resource variables *including race*.
-These probabilities provide a counterfactual estimate of what the average probability would be for each group if they only differed on the resource variables, allowing me to isolate how different distributions of resources across racial groups affect the risk of grade retention.
-
-To do this, the following variables need to be set at some constant value to be held constant:
-
-- current grade
-- state
-- year
-- metro status
-- gender
-- race
-
-One could just plug in some static values (e.g.
-make everyone White, from Minnesota, in 6th grade, etc.), but this feels pretty ad-hoc.
-I think the best approach is to set all of these values at their mean.
-Since most of these variables are categorical variables, this means treating them as 0/1 dummies and then calculating the proportion as the mean.
-Unfortunately, this approach won't work with the models built above because they made use of factor variables.
-So, to run these models, I first need to create actual 0/1 dummies for the categorical variables listed above (current grade, state, metro status, gender, race).
-That is what the code below does.
-
-```{r create-dummied-data}
-acs_dummied <- acs %>%
-  # first we need to unorder current grade for dummy_cols to work
-  mutate(current_grade=factor(current_grade, levels=levels(current_grade), 
-                              ordered = FALSE)) %>%
-
-  dummy_cols(remove_first_dummy = TRUE) %>%
-  select(cluster, strata, perwt, below_exp_grade, year, family_income, 
-         parents_married, own_home, degree_mother, degree_father,
-         foreign_born, foreign_born_mother, foreign_born_father, 
-         spk_eng, spk_eng_father, spk_eng_mother,
-         starts_with(c("state_","current_grade_","race","sex_","metro_"))) %>%
-  mutate(year=year-2010) %>%
-  clean_names()
-```
-
-Now that I have new "dummied up" dataset, I can rereun the full model using the dummy variables which means a very, very long formula.
-
-```{r model-dummied}
-acs_dummied_design <- svydesign(ids=~cluster, weights=~perwt, data=acs_dummied)
-
-model_full <- svyglm(below_exp_grade~
-                       year*(current_grade_1st+current_grade_2nd+current_grade_3rd+
-                               current_grade_4th+current_grade_5th+
-                               current_grade_6th+current_grade_7th+
-                               current_grade_8th+current_grade_9th+
-                               current_grade_10th+current_grade_11th+
-                               current_grade_12th)+
-                       race_black+race_indigenous+race_asian+race_latino+
-                       race_black_white+race_black_indigenous+race_black_latino+
-                       race_black_asian+race_white_indigenous+race_white_latino+
-                       race_white_asian+race_indigenous_latino+
-                       race_indigenous_asian+race_latino_asian+
-                       sex_female+
-                       metro_non_metro+metro_central_city+metro_metro_non_central+
-                       metro_metro_indeterminable+
-                       state_alaska+state_arizona+state_arkansas+state_california+
-                       state_colorado+state_connecticut+state_dc+state_delaware+
-                       state_florida+state_georgia+state_hawaii+state_idaho+
-                       state_illinois+state_indiana+state_iowa+state_kansas+
-                       state_kentucky+state_louisiana+state_maine+state_maryland+
-                       state_massachusetts+state_michigan+state_minnesota+
-                       state_mississippi+state_missouri+state_montana+
-                       state_nebraska+state_nevada+state_new_hampshire+
-                       state_new_jersey+state_new_mexico+state_new_york+
-                       state_north_carolina+state_north_dakota+state_ohio+
-                       state_oklahoma+state_oregon+state_pennsylvania+
-                       state_rhode_island+state_south_carolina+state_south_dakota+
-                       state_tennessee+state_texas+state_utah+state_vermont+
-                       state_virginia+state_washington+state_west_virginia+
-                       state_wisconsin+state_wyoming+
-                       +sqrt(family_income/10000)+parents_married+own_home+
-                       degree_mother+degree_father+
-                       spk_eng+spk_eng_mother+spk_eng_father+
-                       foreign_born_mother+foreign_born_father+foreign_born, 
-                     design=acs_dummied_design, family=binomial)
-```
-
-I now have the model in the correct form.
-The next step is to create the counterfactual dataset by setting all of the non-resource variables to their mean.
-Thus, in this dataset, the only differences across racial groups are coming from their resource differences.
-
-```{r create-cf-df}
-# Create the counterfactual dataset where non-resource variables are held
-# at their mean. What needs to be set at tis mean?
-# - current_grade
-# - sex
-# - race
-# - state
-# - metro
-# - year
-means <- acs_dummied %>%
-  select(year, starts_with(c("state_","current_grade_","race_","sex_",
-                             "metro_"))) %>%
-  lapply(function(x) {rep(mean(x), length(x))}) %>%
-  bind_rows()
-
-cf_df <- acs_dummied %>%
-  select(cluster, strata, perwt, below_exp_grade, race,family_income, 
-         parents_married, own_home, degree_mother, degree_father, foreign_born, 
-         foreign_born_mother, foreign_born_father, spk_eng, spk_eng_father, 
-         spk_eng_mother) %>%
-  bind_cols(means)
-```
-
-```{r test-marg, eval=FALSE}
-# This code chunk just tests to make sure I get the same value by hand as I get 
-# when I do it through the marg command.
-temp <- subset(cf_df, race=="Black/White")
-lor <- predict(model_full, newdata=temp)
-weighted.mean(exp(lor)/(1+exp(lor)), w=temp$perwt)
-marg(model_full, "year", data=temp, weights=temp$perwt)
-# It works
-```
-
-I can then feed subsets (by race) of this counterfactual dataset into the `marg` command to get the predicted probability of grade retention for each racial group as I do below.
-
-```{r create-cf-marg}
-coefs_cf <- lapply(unique(acs$race), function(x) {
-  temp <- subset(cf_df, race==x)
-  means_margin <- marg(model_full, "year", data=temp, weights=temp$perwt)[[1]]
-  return(tibble(term=x, estimate=means_margin$Margin, 
-                se=means_margin$Standard.Error))
-}) %>%
-  bind_rows()
-
-coef_table_cf <- create_coef_table(coefs_cf)
-```
-
-And now I can plot up the results.
-
-```{r counterfactual-ses, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were resources (household income, parental education, home ownership, parent and child foreign-born status, parent and child English proficiency, and whether parents are married).
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
-ggplot(coef_table_cf, aes(x=term, y=estimate, color=multiracial,
-                          ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_cf, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_wrap(~mrace, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
-
-```
-
-I summarize the results here:
-
-|                   | Lower than both | Same as lower | Between | Same as higher | Higher than both |
-|-------------------|:---------------:|:-------------:|:-------:|:--------------:|:----------------:|
-| White/Asian       |       X         |               |         |                |                  |
-| Black/Asian       |                 |  X  (Asian)   |         |                |                  |
-| Black/Latino      |                 |  X  (Black)   |         |                |                  |
-| White/Latino      |                 |  X  (White)   |         |                |                  |
-| Indigenous/Latino |                 |X  (Indigenous)|         |                |                  |
-| Latino/Asian      |                 |  X (Asian)    |         |                |                  |
-| White/Indigenous  |                 |               |    X    |                |                  |
-| Indigenous/Asian  |                 |               |    X    |                |                  |
-| Black/Indigenous  |                 |               |    X    |                |                  |
-| Black/White       |                 |               |         |   X (Black)    |                  |
-
-In half the cases, the biracial group would have outcomes similar to the group with better resources suggesting that the parents in these matches are being positively selected.
-The White/Asian case is even more dramatic with a resource advantage over both single race groups.
-You can see this particularly for the White/Asian case in terms of family income where they have the highest of *any* group by a considerable amount.
-White/Asians are on average a very privileged group.
-
-Three groups that have the "expected" in-between status based on resources all share one thing in common - they are part-Indigenous.
-That is not true of the Indigenous/Latino case, but is of all others.
-It suggests less selectivity in crossing this boundary.
-
-Finally, the Black/White case stands out in that Black/White kids have the same risk of being held back based on their resources as do Black kids.
-Thus, they are not getting a resource boost nor does it seem like they are particularly positively selected.
-
-So the effects of controlling for these resources will be complex and different across groups.
-In general for those that are the same as lower or lower than both, we see that it pulls their relative probability up somewhat.
-For the Black/White case, it pulls their probability down on the other hand.
-For the "between" groups its more complex and doesn't seem to do much in total.
-
-### Senstivity tests separating SES and cultural differences
-
-Lets do the same procedure but I want to isolate the role of SES vs cultural (foreign-born and language).
-
-For ease of use, I am just going to plug in values here for the other resources rather than dummy everything up.
-
-```{r cf-ses-isolate, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were SES resources (household income, parental education, home ownership, and whether parents are married).
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
-cf_df_ses <- cf_df
-cf_df_ses$foreign_born <- FALSE
-cf_df_ses$foreign_born_father <- FALSE
-cf_df_ses$foreign_born_mother <- FALSE
-cf_df_ses$spk_eng <- "Yes"
-cf_df_ses$spk_eng_father <- "Yes"
-cf_df_ses$spk_eng_mother <- "Yes"
-
-coefs_cf_ses <- lapply(unique(acs$race), function(x) {
-  temp <- subset(cf_df_ses, race==x)
-  means_margin <- marg(model_full, "year", data=temp)[[1]]
-  return(tibble(term=x, estimate=means_margin$Margin, 
-                se=means_margin$Standard.Error))
-}) %>%
-  bind_rows()
-
-coef_table_cf_ses <- create_coef_table(coefs_cf_ses)
-
-ggplot(coef_table_cf_ses, aes(x=term, y=estimate, color=multiracial,
-                          ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_cf_ses, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_wrap(~mrace, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
-```
-
-```{r cf-culture-isolate, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were cultural resources (foreign-born status and English proficiency).
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
-cf_df_culture <- cf_df
-cf_df_culture$family_income <- mean(cf_df_culture$family_income)
-cf_df_culture$own_home <- TRUE
-cf_df_culture$degree_father <- "BA"
-cf_df_culture$degree_mother <- "BA"
-cf_df_culture$parents_married <- TRUE
-
-coefs_cf_culture <- lapply(unique(acs$race), function(x) {
-  temp <- subset(cf_df_culture, race==x)
-  means_margin <- marg(model_full, "year", data=temp)[[1]]
-  return(tibble(term=x, estimate=means_margin$Margin, 
-                se=means_margin$Standard.Error))
-}) %>%
-  bind_rows()
-
-coef_table_cf_culture <- create_coef_table(coefs_cf_culture)
-
-ggplot(coef_table_cf_culture, aes(x=term, y=estimate, color=multiracial,
-                          ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_cf_culture, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_wrap(~mrace, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
-```
-
-The results are a lot messier for cultural differences, partly because of the opposite effect of foreign-born parents.
-In most cases, differences point to an in-between status when there is any daylight between the constituent groups.
-So, its the SES effects that are really pulling this toward the lower group.
-
-## Summary table of resource differences
-
-Lets see if I can put all of this into a simpler table that summarize the key issues.
-
-```{r tab-resources}
-temp <- coefs_cf %>%
-  select(term, estimate) %>%
-  mutate(estimate=round(estimate,3)*100,
-         term=as.character(term)) %>%
-  rename(race=term, held_back_cf=estimate)
-
-tbl_resources <- acs_design %>%
-  group_by(race) %>%
-  summarize(fam_inc=round(survey_mean(family_income, vartype=NULL), 0),
-            own_home=round(survey_mean(own_home, vartype=NULL), 3)*100,
-            m_ba=round(survey_mean(degree_mother=="BA" | degree_mother=="G", 
-                                   vartype=NULL), 3)*100,
-            f_ba=round(survey_mean(degree_father=="BA" | degree_father=="G", 
-                                   vartype=NULL), 3)*100,
-            f_born=round(survey_mean(foreign_born, vartype=NULL), 3)*100,
-            par_fb=round(survey_mean(foreign_born_father | foreign_born_mother, 
-                                     vartype=NULL), 3)*100,
-            spk_eng_well=round(survey_mean(spk_eng=="Yes",
-                                           vartype=NULL), 3)*100,
-            spk_eng_par=round(survey_mean(spk_eng_father=="Yes" & 
-                                     spk_eng_mother=="Yes", vartype=NULL), 
-                              3)*100) %>%
-  mutate(race=as.character(race)) %>%
-  left_join(temp) %>%
-  arrange(held_back_cf)
-
-index_biracial <- which(str_detect(tbl_resources$race, "/"))
-
-tbl_resources %>%
-  mutate_if(is.numeric, format, big.mark=",") %>%
-  kbl(caption="Resource differences by race. Shading indicates biracial group. All results are survey weighted.",
-      align = c("l",rep("r", ncol(tbl_resources)-1)),
-      col.names=c("race","mean family income","% own home",
-                  "% mother, BA degree","% father, BA degree",
-                  "% foreign-born","% either parent foreign-born",
-                  "% speak English well", "% Both parents speak english well",
-                  "% held back, counterfactual")) %>%
-  kable_paper() %>%
-  row_spec(index_biracial, background = "#D3D3D3")
-
-save(tbl_resources, file=here("analysis","output","tbl_resources.RData"))
-```
-
-The table sorts groups by their counterfactual probability of being held back. I think it does a nice job of summarizing the issues.
-
-```{r time-end}
-timestamp()
-```

### 2022-09-04 analysis/analysis_resources.Rmd
Refactor analysis.Rmd into three separate Rmd files

--- 2022-09-03 analysis/analysis.Rmd
@@
 ---
-title: "Analysis for Biracial Grade Retention Project"
+title: "Examining Resource Differences by Racial Group"
 output: 
@@
 ```
-
-# Grade retention summary
-
-OK, first lets just confirm that the grade retention variable is working as expected on our sample. I start by looking at the number of kids by age and current grade:
-
-```{r tab-grade-by-age}
-acs %>% 
-  tabyl(age, current_grade) %>%
-  mutate_if(is.numeric, format, big.mark=",") %>%
-  kbl(caption="Cross-tabulation of age and current grade of student",
-      align=c("l",rep("r",13))) %>%
-  kable_paper() %>%
-  column_spec(2, background=c(rep("#D3D3D3",2), rep("white", 14))) %>%
-  column_spec(3, background=c("white", 
-                              rep("#D3D3D3",2), 
-                              rep("white", 13))) %>%
-  column_spec(4, background=c(rep("white", 2), 
-                              rep("#D3D3D3",2), 
-                              rep("white", 12))) %>%
-  column_spec(5, background=c(rep("white", 3), 
-                              rep("#D3D3D3",2), 
-                              rep("white", 11))) %>%
-  column_spec(6, background=c(rep("white", 4), 
-                              rep("#D3D3D3",2), 
-                              rep("white", 10))) %>%
-  column_spec(7, background=c(rep("white", 5), 
-                              rep("#D3D3D3",2), 
-                              rep("white", 9))) %>%
-  column_spec(8, background=c(rep("white", 6), 
-                              rep("#D3D3D3",2), 
-                              rep("white", 8))) %>%
-  column_spec(9, background=c(rep("white", 7), 
-                              rep("#D3D3D3",2), 
-                              rep("white", 7))) %>%
-  column_spec(10, background=c(rep("white", 8), 
-                               rep("#D3D3D3",2), 
-                               rep("white", 6))) %>%
-  column_spec(11, background=c(rep("white", 9), 
-                               rep("#D3D3D3",2), 
-                               rep("white", 5))) %>%
-  column_spec(12, background=c(rep("white", 10), 
-                               rep("#D3D3D3",2), 
-                               rep("white", 4))) %>%
-  column_spec(13, background=c(rep("white", 11), 
-                               rep("#D3D3D3",2), 
-                               rep("white", 3))) %>%
-  column_spec(14, background=c(rep("white", 12), 
-                               rep("#D3D3D3",2), 
-                               rep("white", 2)))
-```
-
-Ok, this looks more or less like what I would expect.
-We observe two modal ages for each grade.
-Some of the more extreme age values for older grades seem a little strange given that these are only supposed to be currently enrolled students.
-20-year old sixth graders?
-I mean there are relatively few of these cases but still seems a bit odd.
-Its also interesting how this really starts expanding past 5th - did the Census Bureau "correct" values below 6th or something?
-
-Now lets calculate the percentage of cases TRUE on the dependent variable across this same table, just to confirm its all working correctly.
-
-```{r check-retention-code, fig.cap="The proportion of students measured at below grade level by age and current grade.
-This graph is used to ensure that the dependent variable is calculated correctly."}
-acs %>%
-  group_by(age, current_grade) %>%
-  summarize(below=mean(below_exp_grade)) %>%
-  ggplot(aes(x=current_grade, y=age, fill=below))+
-  geom_tile(color="grey20")+
-  geom_text(aes(label=below))+
-  theme_bw()+
-  scale_fill_viridis_c()+
-  scale_y_continuous(breaks=5:20, minor_breaks = NULL)+
-  labs(x="current grade", y="age", fill="proportion below grade level")
-```
-
-Ok, that looks good.
-
-What do the grade retention probabilities look like by grade:
-
-```{r retention-by-grade, fig.cap="Percent of students clearly behind grade level by current grade. Blue lines fit a smoothed curve to the data."}
-acs %>%
-  group_by(current_grade) %>%
-  summarize(below_grade=mean(below_exp_grade),
-            n=n(),
-            se=sqrt((below_grade*(1-below_grade))/n)) %>%
-  ggplot(aes(x=current_grade, y=below_grade, group=1,
-             ymin=below_grade-1.96*se, ymax=below_grade+1.96*se))+
-  geom_smooth(se=FALSE)+
-  geom_line(linetype=2, color="grey30")+
-  geom_point()+
-  geom_linerange()+
-  scale_y_continuous(labels=scales::percent)+
-  labs(x="current grade", y="percent clearly behind grade level for age")+
-  theme_bw()
-```
-
-I guess those numbers look pretty reasonable.
-Its hard to benchmark it off anything.
-Warren et.
-al.
-estimate annual retention rates by grade but that is different than what I am measuring here because this is cumulative (but can also reflect students being caught up).
-In general, these must be underestimates because they do not capture kids who were held back but are still within the modal age.
-
-What about by year?
-
-```{r retention-by-year, fig.cap="Trends over time in the percent of students clearly behind grade level. Blue lines fit a smoothed curve to the data."}
-acs %>%
-  group_by(year) %>%
-  summarize(below_grade=mean(below_exp_grade)) %>%
-  ggplot(aes(x=year, y=below_grade))+
-  geom_line()+
-  geom_point()+
-  geom_smooth(se=FALSE)+
-  scale_x_continuous(breaks=c(2010,2014,2018))+
-  scale_y_continuous(labels=scales::percent)+
-  labs(x="current grade", y="percent clearly behind grade level for age")+
-  theme_bw()
-```
-
-Overall, there has been some decline over time in the prevalence of grade retention.
-
-Lets look at this by grade and year
-
-```{r retention-by-year-grade, fig.cap="Trends over time and grade in the percent of students clearly behind grade level. Lines are calculated by smoothing."}
-acs %>%
-  group_by(current_grade, year) %>%
-  summarize(below_grade=mean(below_exp_grade)) %>%
-  ggplot(aes(x=year, y=below_grade, group=current_grade, color=current_grade))+
-  #geom_line()+
-  geom_smooth(se=FALSE)+
-  geom_point()+
-  scale_x_continuous(breaks=c(2010,2014,2018))+
-  scale_y_continuous(labels=scales::percent)+
-  labs(x="current grade", y="percent clearly behind grade level for age")+
-  theme_bw()
-```
-
-The evidence here suggests that the decline is driven by the higher grades where grade retention was higher overall.
-
-# Descriptive statistics
-
-Lets looks at sample size and grade retention by race.
-
-```{r desc-stats}
-tbl_summary <- acs_design %>%
-  group_by(race) %>%
-  summarize(n=n(), 
-            held_back=round(survey_mean(below_exp_grade, vartype=NULL), 3)*100,
-            mean_age=round(survey_mean(age, vartype=NULL), 1)) %>%
-  arrange(desc(n))
-
-index_biracial <- which(str_detect(tbl_summary$race, "/"))
-
-tbl_summary %>%
-  mutate_if(is.numeric, format, big.mark=",") %>%
-  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size. Survey weights applied.",
-      align = c("l",rep("r", ncol(tbl_summary)-1)),
-      col.names=c("race","n","% below expected grade", "mean age")) %>%
-  kable_paper() %>%
-  row_spec(index_biracial, background = "#D3D3D3")
-```
-
-# Model building
-
-## Functional form of year-grade interactions
-
-Before running the full models, I want to test out how I can parsimoniously address differences in the time trends by grade.
-
-```{r models-year-grade}
-model_add_cat <- glm(below_exp_grade~as.factor(year)+current_grade+state, 
-                     data=acs, family=binomial)
-model_add_cont <- glm(below_exp_grade~I(year-2010)+current_grade+state, 
-                     data=acs, family=binomial)
-model_inter_cont <- glm(below_exp_grade~I(year-2010)*current_grade+state, 
-                     data=acs, family=binomial)
-model_inter_cat <- glm(below_exp_grade~as.factor(year)*current_grade+state, 
-                     data=acs, family=binomial)
-model_inter_state <- glm(below_exp_grade~I(year-2010)*current_grade+
-                            I(year-2010)*state, 
-                     data=acs, family=binomial)
-
-sapply(list("categorical, additive"=model_add_cat,
-            "categorical, interaction"=model_inter_cat, 
-            "continuous, additive"=model_add_cont, 
-            "continuous, interaction"=model_inter_cont,
-            "continous, interaction + state*year continuous interaction"=model_inter_state),
-       BIC) %>%
-    enframe("model","BIC") %>%
-  kbl() %>%
-  kable_styling(full_width=FALSE)
-
-# free some memory
-rm(model_add_cat, model_add_cont, model_inter_cat, model_inter_cont,
-   model_inter_state)
-```
-
-The preferred model by BIC is clearly the model that fits year as a continuous measure and interacts it with current grade, so that is what I will do.
-
-## Running the models
-
-Here are the model structure that I will use:
-
-0.  Raw race differences
-1.  Add fixed effects for state, current grade, year, metro, and sex
-2.  Add kids and parents foreign-born status.
-3.  Add mother and father highest degree
-4.  Add family income (square rooted), own your own home, and whether parents are married.
-
-All models are estimated as logit models.
-I use the `survey` library to adjust for variable sample weights and clustering of respondents in the same household.
-Warning, the code chunks below take a bit of time to run!
-
-```{r build-models, results='asis'}
-model0 <- svyglm(below_exp_grade~race, design=acs_design, family=binomial)
-model1 <- update(model0, 
-                 .~.+I(year-2010)*current_grade+state+metro+sex)
-model2 <- update(model1, 
-                 .~.+foreign_born+foreign_born_mother+foreign_born_father+
-                   spk_eng+spk_eng_mother+spk_eng_father)
-model3 <- update(model2, .~.+degree_mother+degree_father)
-model4 <- update(model3, .~.+sqrt(family_income/10000)+own_home+parents_married)
-
-knitreg(list(model0, model1, model2, model3, model4))
-```
-
-## Differences for monoracial groups
-
-The model results here are not very interesting directly because everything is relative to a single reference group.
-However, before proceeding with a technique for understanding the outcomes for biracial respondents, its worthwhile to look more closely at the differences between monoracial groups.
-The table below shows these differences using average marginal effects (AME) with whites as the reference.
-
-```{r monoracial-results, warning=FALSE, results='asis'}
-results_monoracial <- map(list(model0, model1, model2, model3, model4),
-                          function(x) {
-                            marg(x, "race", type="effects", 
-                                 at_var_interest=c("White","Black","Indigenous",
-                                                   "Asian","Latino"))[[1]][-1,]
-                          })
-
-knitreg(map(results_monoracial, convert_marg),
-        digits=3, caption="Average marginal differences in probability of being behind grade relative to whites",
-        custom.gof.rows=list("State fixed effects" =  c("No","Yes","Yes","Yes","Yes"),
-                             "Year fixed effects" =   c("No","Yes","Yes","Yes","Yes"),
-                             "Grade fixed effects" =  c("No","Yes","Yes","Yes","Yes"),
-                             "Metro fixed effects" =  c("No","Yes","Yes","Yes","Yes"),
-                             "Nativity and language" = c("No","No","Yes","Yes","Yes"),
-                             "Parent's education" =    c("No","No","No","Yes","Yes"),
-                             "Family resources" =      c("No","No","No","No","Yes")))
-
-save(results_monoracial, file=here("analysis","output","monoracial_results.RData"))
-```
-
-The results indicate that controlling for everything in Model 5 reduces the racial disparities substantially.
-Comparing Model 5 to Model 2, the controls cut the effect in half for the Black and Indigenous groups.
-It has less effect and slightly increases the gap for Asians because Asians already had lower rates than Whites.
-The biggest change is for Latinos.
-In model 2, Latinos had the second highest probabilities and were 2% higher than whites.
-After controlling for everything, that actually reverse direction, such that Latinos have slightly lower expected probabilities than whites (.3% less).
-
-The gap is cut in half for Latinos by the nativity and language variables.
-These variables also increase the advantage of Asians as expected.
-they have no effect on the Black and Indigenous effects as expected.
-
-So, I have to consider the shrinking (and in the case of Latinos, flipping direction) of these effects with controls when evaluating the placement of biracials.
-
-## Average adjusted predictions for biracials
-
-I want to compare each multiracial group to:
-
-1.  Its two constituent monoracial groups (e.g. Compare Black/White to White and Black)
-2.  The assumption that the biracial group will be halfway between the two constituent monoracial groups.
-
-I can do this by calculating the *average adjusted prediction* of the dependent variable across all of the possible race groups.
-The average adjusted prediction (this is the stata-ese language - it is also called variously a predictive margin among other things).
-The adjusted predition in general is just the predicted value of the dependent variable for some set of values on the covariates.
-In this case, I want to predict on the level of the response to get predicted probabilities of being held back.
-The average adjusted prediction essentially gets the adjusted prediction for every observation in the data based on their covariates and then takes the average.
-Essentially, this is equivalent to treating every respondent as if they belonged to group A when calculating the average adjusted prediction for group A.
-The average adjusted prediction is directly related to the *average marginal effect* (AME) in that the AME for a group difference is simply the difference in the average adjusted prediction across groups.
-
-Its easy to get a simple adjusted prediction at specific values using the `predict` command, but more complicated for the average adjusted prediction.
-In order to calculate these average adjusted predictions, I use the `marg` command in the `modmargs` library.
-This function has a nice `type="levels"` argument that directly gives you the average adjusted prediction.
-Standard errors for these average adjusted predictions are calculated by the [delta method](https://cran.r-project.org/web/packages/modmarg/vignettes/delta-method.html).
-I also adjust all SEs for clustering of respondents within households and variance in person weights.
-
-To compare the results for each biracial group to the halfway assumption, I need to calculate the value midway between the two constituent groups and the standard error of that value.
-Calculating the midway value is just a straightforward mean calculation.
-To calculate the standard error, I take:
-
-$$\frac{\sqrt{s_1^2+s_2^2}}{2}$$ where $s_1$ is the standard error for monoracial group 1 and $s_2$ is the standard error for monoracial group 2.
-This formula follows from the standard formulas for the variance of the sum of random variables.
-
-In order to calculate all the values that I need, I have created a function that will put this into a nice table for later plotting.
-
-```{r calc-cond-means, warning=FALSE}
-coef_table_mod0 <- calculate_marg_means(model0)
-```
-
-Now I can use this nice table to plot out the results. I illustrate what this basic plot looks like below for the case of black/white individuals.
-
-```{r example-plot, fig.cap="Probability of being behind grade for biracial black/white respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level"}
-g <- coef_table_mod0 %>%
-  filter(mrace=="Black/White") %>%
-  ggplot(aes(x=term, y=estimate, color=multiracial,
-             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_mod0, multiracial & mrace=="Black/White"))+
-  geom_point()+
-  coord_flip()+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="predicted probability of being behind expected grade")
-
-g+annotate("text", 1, 0.035, hjust="left",
-             label="Grey bands give 83.4%\nconfidence interval around\nmonoracial estimates")+
-  annotate("text", 2, 0.045, hjust="left",
-             label="Red band gives 83.4%\nconfidence interval around\nmidpoint estimate")+
-  annotate("text", 2.5, 0.036, hjust="left",
-             label="Actual biracial estimate\nshown with 83.4% confidence\ninterval")
-```
-
-This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions and their binomial standard errors (although the SEs are slightly different because this model accounts for clustering).
-The grey bands for the monoracial groups bookmark the range of results we might expect and the red band gives what would be the midpoint value between the groups, adjusting for statistical uncertainty.
-I can then see where the actual biracial estimate falls within this range.
-The results here indicate that biracial black/white students have probabilities of grade retention much closer to the lower values of white than black students.
-However, their probability is still a bit higher than that for whites.
-
-Why I am using 83% (technically 83.4%) confidence intervals?
-This is a nice visual trick.
-People often incorrectly believe that overlapping 95% confidence intervals on two means indicates that the difference in the means is not statistically significant at $p<.05$.
-This is not always true and leads in practice to a much lower $\alpha$ for rejection.
-A much better approximation is given when one uses1.386 rather than 1.96 in estimating the confidence interval, which produces an 83.4% confidence interval.
-The intuition behind this technique is given nicely [here](https://chris-said.io/2014/12/01/independent-t-tests-and-the-83-confidence-interval-a-useful-trick-for-eyeballing-your-data/).
-Because I use this interval, we can know that if intervals (or bands) do not overlap, then the difference in means between the two groups is statistically significant at $p<.05$.
-In this case, the biracial estimate is statistically significantly different from both monoracial estimates as well as the "halfway" estimate.
-In short, biracial black/white students are held back at rates much closer but somewhat higher than whites, rather than blacks.
-
-Its important to note that this technique is just an approximation because it depends on the assumption that the standard errors for the two means are the same, which is not going to be true here.
-However, it should give results that are quite close.
-Correcting for unequal SEs is far more complicated because the interval used will vary for any pairwise comparison which I am not doing here.
-
-The results here are the same as the raw differences, but I can also fit this graph to any of the models, using the average adjusted predictions.
-The plot below does this for model 4, by panelizing each multiracial group.
-
-```{r cond-means-model4, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
-Estimates are based on models that adjust for state, year, current grade, metropolitan status, gender, foreign born status, parent's foreign born status, English proficieny, parent's English proficiency, parent's education, family income, home ownership, and family structure."}
-coef_table_mod4 <- calculate_marg_means(model4)
-
-ggplot(coef_table_mod4, aes(x=term, y=estimate, color=multiracial,
-                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_mod4, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_wrap(~mrace, ncol=3, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
-```
-
-As the table here shows, after controlling for other stuff, most multiracial groups are quite similar to the monoracial constituent group with the lower grade retention rate.
-The only case where the biracial group is more like the higher monoracial group is for Latino/Asians.
-The results for indigenous groups are also quite noisy due to small size, but are generally consistent with the "below the halfway point."
-
-To compare these results across models, I can try a very large figure that panelizes by model on the column and group on the row.
-
-```{r cond-means-model-compare, fig.height=12, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
-Each panel is based on a different model that accounts for certain variables.
-Subsequent panels cumulatively include all prior terms."}
-#calculate remaining marginal means
-coef_table_mod1 <- calculate_marg_means(model1)
-coef_table_mod2 <- calculate_marg_means(model2)
-
-coef_table <- coef_table_mod0 %>%
-  mutate(model="crude")
-coef_table <- coef_table_mod1 %>%
-  mutate(model="fixed effects") %>%
-  bind_rows(coef_table)
-coef_table <- coef_table_mod2 %>%
-  mutate(model="cultural resources") %>%
-  bind_rows(coef_table)
-coef_table <- coef_table_mod4 %>%
-  mutate(model="parent resources") %>%
-  bind_rows(coef_table) %>%
-  mutate(model=factor(model, levels=c("crude","fixed effects",
-                                      "cultural resources",
-                                      "parent resources")),
-         mrace=factor(mrace, levels=c("Black/White",
-                                      "Black/Latino",
-                                      "Black/Asian",
-                                      "Black/Indigenous",
-                                      "White/Indigenous",
-                                      "White/Latino",
-                                      "White/Asian",
-                                      "Latino/Asian",
-                                      "Indigenous/Latino",
-                                      "Indigenous/Asian")))
-
-ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
-                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_grid(mrace~model, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
-
-save(coef_table, file=here("analysis","output","coef_table.RData"))
-```
-
-## Summarizing the model results
-
-Ok, let me first try to summarize these results, by group.
-
-### Black/White results
-
-For the first three models, the results are pretty consistent.
-Black/White biracials are much closer to the lower grade retention rates of whites, although they have rates that are clearly still higher than whites.
-Controlling for parental resources changes things considerably.
-First, the gap between Blacks and Whites shrinks considerably (visually it looks like its cut about in half).
-Second, Black/White biracials actually have lower rates than either group.
-They aren't that different than whites in effect size but the difference is statistically significant.
-
-Why does the big change occur in the parental resources model?
-The descriptive stat table shows that the family resources of Black/White biracials are only slightly better than those of Black students and substantially worse than those of White students.
-So, Black/White biracials are at a disadvantage relative to whites, which is controlled for in the last model.
-
-### Black/Latino results
-
-In the crude model, its clear that there isn't much difference in the grade retention rates of Blacks and Latinos, with Blacks having a slightly higher rate than Latinos.
-The Black/Latino group has rates well below either of these groups.
-In fact, their grade retention rates in the crude model are closer to Whites.
-
-Interestingly, the fixed effects model reverses the ordering of Blacks and Latinos, but this is then reversed again in the cultural resources model.
-In addition the gap between Latinos and Blacks grows a little in the cultural resources model, which is expected.
-In all of these models, the Black/Latino group remains below either of those two groups, although the gap shrinks considerably across models.
-
-Things change quite a bit in the parent resources model.
-The gap between blacks and Latinos grows and the Black/Latino group now looks almost identical to the Latino group.
-So, some of their lower probability was due to better family resources, particularly relative to Latinos, as can be seen on the table of descriptive stats.
-Once that is controlled for, their outcomes are indistinguishable from Latino students.
-
-### Black/Asian results
-
-The crude and fixed effects models indicate that Black/Asian students have grade retention rates similar to the lower grade retention rates of Asian students.
-Controlling for cultural resources changes this somewhat with a point estimate that is higher than that of Asian students.
-This makes sense because Black/Asian students are much more likely than Asian students to be native born and speak English well with native-born parents who also speak English well.
-So this gave them and advantage which is removed in that model.
-The family resources model does not change the relative position of Black/Asian students, but does shrink the overall racial gap between Black and Asian students.
-
-In the last two models, the confidence intervals overlap a bit, leading to greater uncertainty in the relatve placement of Black/Asian biracials.
-The point estimate suggests that their grade retention probability is between that of Asians and the half-way point, but the edge of the confidence intervals touch both those bands, so I cannot statistically distinguish their placement relative to those two possibilities.
-In any case, the results strongly rule out them being close to the outcomes of Black students, contra the hypodescent expectation.
-
-### Black/Indigenous results
-
-These results are quite noisy because of the smaller sample sizes involved.
-These are the two monoracial groups with the highest dropout rates, with or without controls.
-However, the difference between them is also quite large due to the very high grade retention probability for Indigenous students.
-This gap shrinks significantly in the model that adjusts for parental resources.
-
-Because of the noise its quite difficult to determine the relative placement of Black/Indigenous students at all, but we can rule out across all models that they have similar grade retentions rates to the higher grade retention of Indigenous students.
-The point estimate suggests a placement somewhere between the lower Black rate and the halfway point, but the wide CIs mean we cannot rule out either of those possibilities.
-
-The point estimate also gets much closer to the lower black rate once we control for family resources.
-
-### White/Indigenous results
-
-The results here consistently show that White/Indigenous students have grade retention rates about midway between the rates of whites and the halfway point.
-So they are clearly in the "lower half" between the groups and closer to White than Indigenous students.
-The overall gap between White and Indigenous students gets smaller with controls, but the relative placement of White/Indigenous students does not change.
-
-###  White/Latino results
-
-Importantly, the ordering of the monoracial groups shifts as controls are added to the model.
-The crude and fixed effets models show that White students are much less likely to be held back relative to Latino students.
-Controlling for cultural resources reduces this gap substantially and the inclusion of family resources flips its so that Latino students have slightly lower grade retention rates than Whites.
-So in the final model, there is not much room to find a midway point for White/Latino biracials.
-The final model shows outcomes that are similar to the lower rates of Latinos, although I need to check this more carefully because the CIs are not visible here relative to the dot (TODO).
-
-In the models prior to the inclusion of family resources, White/Latino biracials had grade retention rates similar to those of Whites, who had the lower rate.
-So despite the switch in ordering between Whites and Latinos, White/Latino biracials tend to look like the group with lower rates.
-
-### White/Asian results
-
-In all of the models prior to the inclusion of family resources, White/Asian students had grade retention rates slightly lower than the rates of Asians - the monoracial group with the lower rates.
-However, once controls are added for family resources, Asian/White students look just like Asian students (the lower group).
-This is because White/Asian students have the greatest family resources of any group in the dataset, even higher than those of White and Asian students.
-
-### Latino/Asian results
-
-This is the group with the biggest outlier in terms of the effect.
-The initial results in the crude model suggested that Latino/Asian students had rates similar to the lower rates of Asian students.
-Controlling for cultural resources shifted this to rates that were exactly midway between the two monoracial groups.
-Controlling for family resources produced a point estimate that was *higher* than either monoracial group although not statistically distinguishable from the higher rates of Latinos.
-Notably, the gap between Asians and Latinos is also quite small in the final model.
-
-The change across models is a result of the high cultural and family resources of Latino/Asian students.
-However, their family resources do not seem particularly advantaged relative to Asian students, so something a bit complex is happening in the final model there.
-They look relatively worse compared to Latinos, but the relative difference doesn't change as much with respect to Asians.
-Essentially, once I control for family resources, they are "leapfrogged" by the Latino group.
-
-### Indigenous/Latino results
-
-Controlling for cultural resources makes a big difference here.
-Initially Indigenous/Latino students look the same as the lower Latino students.
-Controlling for cultural resources makes their outcomes indistinguishably from the midway point, althoug the point estimate is still in the lower half.
-Controlling for family resources has little effect.
-
-Notably, because of sample size the error bars are substantially here.
-
-### Indigenous/Asian results
-
-This is a very small group ($n=521$) so the results are pretty noisy with wide CIs.
-However, the results are consistent across estimates.
-The grade retention rates of Indigenous/Asian students are similar to those of the lower Asian students.
-In the final model, the point estimate is almost identical to those for Asian students.
-
-Overall, I can put these results into a table where I specify the potential placement of each biracial group relative to their monoracial peers. There are seven possible outcomes:
-
-* Lower than both groups
-* Same as lower group
-* Lower half
-* Halfway between
-* Upper half
-* Same as higher
-* Higher than both groups
-
-For the two ends of the distribution and the lower half/upper half, I only consider these cases if it is statistically distinguishable from being in the other categories, respectively.
-For all other cases, I place an X in a cell if the confidence interval indicates this as a possibility.
-Asterisks indicate the placement of the point estimate for cases with uncertainty.
-
-|                   | Lower than both | Same as lower | Lower half | Halfway | Upper half | Same as higher | Higher than both |
-|-------------------|:---------------:|:-------------:|:----------:|:-------:|:----------:|:--------------:|:----------------:|
-| Black/White       |        X        |               |            |         |            |                |                  |
-| Black/Latino      |                 |       X       |            |         |            |                |                  |
-| White/Asian       |                 |       X       |            |         |            |                |                  |
-| White/Latino      |                 |       X       |            |         |            |                |                  |
-| Indigenous/Asian  |                 |       X       |            |         |            |                |                  |
-| Black Asian       |                 |       X       |     X*     |    X    |            |                |                  |
-| Black/Indigenous  |                 |       X       |     X*     |    X    |            |                |                  |
-| Indigenous/White  |                 |               |      X     |         |            |                |                  |
-| Indigenous/Latino |                 |               |            |    X    |            |                |                  |
-| Latino/Asian      |                 |               |            |         |            |        X       |                  |
-
-The results I show here are based on model 4.
-They clearly show that most biracial groups are closer in outcomes to the monoracial group with the lower rates.
-In fact, half the groups have rates that are either indistinguishable from the lower group or lower than *both* groups.
-
-Being a part-black group in no way makes one more likely to suffer from hypodescent.
-In fact, the results for Black/White biracials where we would expect hypodescent to play the largest role are actually the farthest from hypodescent as possible.
-If any group seems to suffer from a pattern of hypodescent to some degree it is part-Indigenous groups and the Latino/Asian group.
-So something else is going on here.
-
-In general, the results show that biracial groups are able to "slip through" the racialized structures that affect their more disadvantaged monoracial groups and this is not just a function of observable differences in class and nativity.
-
-## Sensitivity tests
-
-### Models by grade
-
-I want to separate the results by age more or less, but I cannot take age directly because then some kids cannot logically be behind grade at certain grade levels.
-For example, if I restrict to 11 or younger, then none of the 5th graders will be behind grade, by design.
-So, I will restrict by current grade level which indirectly gets at age.
-I am going to split it by elementary, middle school, and high school.
-
-```{r models-by-grade, warning=FALSE, fig.height=12, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
-Panels compare results when data are restricted to certain grade ranges."}
-#run separate models by grade groups
-model4_elem <- update(model4, design=subset(acs_design, current_grade<="5th"))
-model4_middle <- update(model4, design=subset(acs_design, current_grade>"5th" & 
-                                        current_grade<"9th"))
-model4_hs <- update(model4, design=subset(acs_design, current_grade>="9th"))
-
-coef_table_elem <- calculate_marg_means(model4_elem)
-coef_table_middle <- calculate_marg_means(model4_middle)
-coef_table_hs <- calculate_marg_means(model4_hs)
-
-coef_table <- coef_table_elem %>%
-  mutate(model="elementary")
-coef_table <- coef_table_middle %>%
-  mutate(model="middle school") %>%
-  bind_rows(coef_table)
-coef_table <- coef_table_hs %>%
-  mutate(model="high school") %>%
-  bind_rows(coef_table) %>%
-  mutate(model=factor(model, levels=c("elementary","middle school",
-                                      "high school")),
-         mrace=factor(mrace, levels=c("Black/White",
-                                      "Black/Latino",
-                                      "Black/Asian",
-                                      "Black/Indigenous",
-                                      "White/Indigenous",
-                                      "White/Latino",
-                                      "White/Asian",
-                                      "Latino/Asian",
-                                      "Indigenous/Latino",
-                                      "Indigenous/Asian")))
-
-ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
-                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_grid(mrace~model, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
-```
-
-The results get quite a bit noisier here because we are cutting the sample size substantially in each panel.
-The general pattern is that the results tend to look pretty similar across grade levels.
-What variations we do see could easily be the result of statistical noise as they tend to move around the most in cases with high error bands.
-The two notable cases where something might be going on:
-
-* For White/Black students, grade retention is clearly in the lower-half between Whites and the midway point rather than below whites for high school students.
-Notably, this is also where the white/black grows the largest.
-So the advantage that these students have relative to everyone else only holds for elementary and middle school.
-Nonetheless, their outcomes are still much closer to White students than Black students.
-* For White/Indigenous students, the placement varies across panels.
-In elementary school, these students are indistinguishable from White students.
-They are perfectly in the middle in middle school, and then in the lower half in high school.
-So it seems that their lowest outcomes happen earlier and on average you end up with a "lower half" across all grades.
-
-In no case do the results here suggest a biasing effect of non-biological kids.
-We would expect that to be greater in higher grades, but the results are generally consistent across grades.
-In cases where they are not consistent, you would expect the results to move toward the larger population group with age, but I do not see that.
-
-### Models with stricter parent linking
-
-```{r models-parent-link, warning=FALSE, fig.height=12, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
-Panels compare results depending on whether results are restricted only to children linked to a head of household or not."}
-#run separate models by grade groups
-model4_best_link <- update(model4, design=subset(acs_design, momrule==11 & poprule==11))
-
-coef_table_best_link <- calculate_marg_means(model4_best_link)
-
-coef_table <- coef_table_best_link %>%
-  mutate(model="only direct links")
-coef_table <- coef_table_mod4 %>%
-  mutate(model="all links") %>%
-  bind_rows(coef_table) %>%
-  mutate(model=factor(model, levels=c("all links","only direct links")),
-         mrace=factor(mrace, levels=c("Black/White",
-                                      "Black/Latino",
-                                      "Black/Asian",
-                                      "Black/Indigenous",
-                                      "White/Indigenous",
-                                      "White/Latino",
-                                      "White/Asian",
-                                      "Latino/Asian",
-                                      "Indigenous/Latino",
-                                      "Indigenous/Asian")))
-
-ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
-                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_grid(mrace~model, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
-```
-
-The results here are extraordinarily stable across both specifications.
 

### 2022-09-04 analysis/analysis_spec_tests.Rmd
Refactor analysis.Rmd into three separate Rmd files

--- 2022-09-03 analysis/analysis.Rmd
@@
 ---
-title: "Analysis for Biracial Grade Retention Project"
+title: "Testing specification of baseline models"
 output: 
@@
 ```
+
+I need to determine the best specification to address how current grade, state, and year affect grade retention probabilities so that I can determine the best and most parsimionious mix of variables for the baseline model.
+I am doing this in a separate script because of the length of time it takes to consider some of these models.
 
@@
 
-# Descriptive statistics
-
-Lets looks at sample characteristics by race.
-
-```{r desc-stats}
-tbl_summary <- acs %>%
-  group_by(race) %>%
-  summarize(n=n(), 
-            held_back=round(mean(below_exp_grade), 3)*100,
-            mean_age=round(mean(age), 1),
-            f_born=round(mean(foreign_born), 3)*100,
-            spk_eng_well=round(mean(spk_eng=="Yes"), 3)*100,
-            spk_eng_par=round(mean(spk_eng_father=="Yes" & 
-                                     spk_eng_mother=="Yes"), 3)*100,
-            par_fb=round(mean(foreign_born_father | 
-                                       foreign_born_mother), 3)*100,
-            fam_inc=round(mean(family_income), 0),
-            own_home=round(mean(own_home), 3)*100,
-            par_hs=round(mean(degree_father!="LHS" | 
-                                degree_mother!="LHS"), 3)*100,
-            par_ba=round(mean(degree_father=="BA" | 
-                                degree_father=="G" | 
-                                degree_mother=="BA" | 
-                                degree_mother=="G"), 3)*100,
-            par_mar=round(mean(parents_married), 3)*100) %>%
-  arrange(desc(n))
-
-index_biracial <- which(str_detect(tbl_summary$race, "/"))
-
-tbl_summary %>%
-  mutate_if(is.numeric, format, big.mark=",") %>%
-  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.",
-      align = c("l",rep("r", ncol(tbl_summary)-1)),
-      col.names=c("race","n","% below expected grade", "mean age",
-                  "% foreign-born","% speak English well",
-                  "% Both parents speak english well","% either parent foreign-born",
-                  "mean family income","% own home","% either parent, HS diploma",
-                  "% Either parent, BA degree","% parents married")) %>%
-  kable_paper() %>%
-  row_spec(index_biracial, background = "#D3D3D3")
-```
-
-# Model building
-
 ## Functional form of year-grade interactions
 
-Before running the full models, I want to test out how I can parsimoniously address differences in the time trends by grade.
+So I want to consider the following baseline models.
+
+- Model 1: additive terms for current grade, state, and year, where year is modeled using annual dummy variables.
+- Model 2: additive terms for current grade, state, and year, where year is modeled as a continuous variable.
+- Model 3: same as model 1, but add in an interaction between current grade and year.
+- Model 4: same as model 2, but add in an interaction between current grade and year.
+- Model 5: same as model 4, but add in an interaction between state and year.
+
+I will judge model fit by the BIC statistic. I wil not consider survey design effects here for speed of computation.
 
@@
                      data=acs, family=binomial)
+model_inter_cat <- glm(below_exp_grade~as.factor(year)*current_grade+state, 
+                     data=acs, family=binomial)
 model_inter_cont <- glm(below_exp_grade~I(year-2010)*current_grade+state, 
-                     data=acs, family=binomial)
-model_inter_cat <- glm(below_exp_grade~as.factor(year)*current_grade+state, 
                      data=acs, family=binomial)
@@
 
-sapply(list("categorical, additive"=model_add_cat,
+bic_scores <- sapply(list("categorical, additive"=model_add_cat,
                           "categorical, interaction"=model_inter_cat, 
@@
                      BIC) %>%
-    enframe("model","BIC") %>%
+    enframe("model","BIC")
+
+bic_scores %>%
   kbl() %>%
@@
 
-# free some memory
-rm(model_add_cat, model_add_cont, model_inter_cat, model_inter_cont,
-   model_inter_state)
+save(bic_scores, save=here("analysis","output","bic_spec_tests.RData"))
 ```
 
-The preferred model by BIC is clearly the model that fits year as a continuous measure and interacts it with current grade, so that is what I will do.
-
-## Running the models
-
-Here are the model structure that I will use:
-
-0.  Raw race differences
-1.  Add fixed effects for state, current grade, year, metro, and sex
-2.  Add kids and parents foreign-born status.
-3.  Add mother and father highest degree
-4.  Add family income (square rooted), own your own home, and whether parents are married.
-
-All models are estimated as logit models. Warning, the code chunks below take a bit of time to run!
-
-```{r create-design}
-# strata make the models take forever, but do not seem to change anything
-# so I will leave them out
-acs_design <- svydesign(ids=~cluster, weights=~perwt, data=acs)
-```
-
-```{r build-models, results='asis'}
-model0 <- svyglm(below_exp_grade~race, design=acs_design, family=binomial)
-model1 <- update(model0, 
-                 .~.+I(year-2010)*current_grade+state+metro+sex)
-model2 <- update(model1, 
-                 .~.+foreign_born+foreign_born_mother+foreign_born_father+
-                   spk_eng+spk_eng_mother+spk_eng_father)
-model3 <- update(model2, .~.+degree_mother+degree_father)
-model4 <- update(model3, .~.+sqrt(family_income/10000)+own_home+parents_married)
-
-knitreg(list(model0, model1, model2, model3, model4))
-```
-
-## Differences for monoracial groups
-
-The model results here are not very interesting directly because everything is relative to a single reference group.
-However, before proceeding with a technique for understanding the outcomes for biracial respondents, its worthwhile to look more closely at the differences between monoracial groups.
-The table below shows these differences using average marginal effects (AME) with whites as the reference.
-
-```{r monoracial-results, warning=FALSE, results='asis'}
-results_monoracial <- map(list(model0, model1, model2, model3, model4),
-                          function(x) {
-                            marg(x, "race", type="effects", 
-                                 at_var_interest=c("White","Black","Indigenous",
-                                                   "Asian","Latino"))[[1]][-1,]
-                          })
-
-knitreg(map(results_monoracial, convert_marg),
-        digits=3, caption="Average marginal differences in probability of being behind grade relative to whites",
-        custom.gof.rows=list("State fixed effects" =  c("No","Yes","Yes","Yes","Yes"),
-                             "Year fixed effects" =   c("No","Yes","Yes","Yes","Yes"),
-                             "Grade fixed effects" =  c("No","Yes","Yes","Yes","Yes"),
-                             "Metro fixed effects" =  c("No","Yes","Yes","Yes","Yes"),
-                             "Nativity and language" = c("No","No","Yes","Yes","Yes"),
-                             "Parent's education" =    c("No","No","No","Yes","Yes"),
-                             "Family resources" =      c("No","No","No","No","Yes")))
-
-save(results_monoracial, file=here("analysis","output","monoracial_results.RData"))
-```
-
-The results indicate that controlling for everything in Model 5 reduces the racial disparities substantially.
-Comparing Model 5 to Model 2, the controls cut the effect in half for the Black and Indigenous groups.
-It has less effect and slightly increases the gap for Asians because Asians already had lower rates than Whites.
-The biggest change is for Latinos.
-In model 2, Latinos had the second highest probabilities and were 2% higher than whites.
-After controlling for everything, that actually reverse direction, such that Latinos have slightly lower expected probabilities than whites (.3% less).
-
-The gap is cut in half for Latinos by the nativity and language variables.
-These variables also increase the advantage of Asians as expected.
-they have no effect on the Black and Indigenous effects as expected.
-
-So, I have to consider the shrinking (and in the case of Latinos, flipping direction) of these effects with controls when evaluating the placement of biracials.
-
-## Average adjusted predictions for biracials
-
-I want to compare each multiracial group to:
-
-1.  Its two constituent monoracial groups (e.g. Compare Black/White to White and Black)
-2.  The assumption that the biracial group will be halfway between the two constituent monoracial groups.
-
-I can do this by calculating the *average adjusted prediction* of the dependent variable across all of the possible race groups.
-The average adjusted prediction (this is the stata-ese language - it is also called variously a predictive margin among other things).
-The adjusted predition in general is just the predicted value of the dependent variable for some set of values on the covariates.
-In this case, I want to predict on the level of the response to get predicted probabilities of being held back.
-The average adjusted prediction essentially gets the adjusted prediction for every observation in the data based on their covariates and then takes the average.
-Essentially, this is equivalent to treating every respondent as if they belonged to group A when calculating the average adjusted prediction for group A.
-The average adjusted prediction is directly related to the *average marginal effect* (AME) in that the AME for a group difference is simply the difference in the average adjusted prediction across groups.
-
-Its easy to get a simple adjusted prediction at specific values using the `predict` command, but more complicated for the average adjusted prediction.
-In order to calculate these average adjusted predictions, I use the `marg` command in the `modmargs` library.
-This function has a nice `type="levels"` argument that directly gives you the average adjusted prediction.
-Standard errors for these average adjusted predictions are calculated by the [delta method](https://cran.r-project.org/web/packages/modmarg/vignettes/delta-method.html).
-I also adjust all SEs for clustering of respondents within households.
-
-To compare the results for each biracial group to the halfway assumption, I need to calculate the value midway between the two constituent groups and the standard error of that value.
-Calculating the midway value is just a straightforward mean calculation.
-To calculate the standard error, I take:
-
-$$\frac{\sqrt{s_1^2+s_2^2}}{2}$$ where $s_1$ is the standard error for monoracial group 1 and $s_2$ is the standard error for monoracial group 2.
-This formula follows from the standard formulas for the variance of the sum of random variables.
-
-In order to calculate all the values that I need, I have created a function that will put this into a nice table for later plotting.
-
-```{r calc-cond-means, warning=FALSE}
-coef_table_mod0 <- calculate_marg_means(model0)
-```
-
-Now I can use this nice table to plot out the results. I illustrate what this basic plot looks like below for the case of black/white individuals.
-
-```{r example-plot, fig.cap="Probability of being behind grade for biracial black/white respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level"}
-g <- coef_table_mod0 %>%
-  filter(mrace=="Black/White") %>%
-  ggplot(aes(x=term, y=estimate, color=multiracial,
-             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_mod0, multiracial & mrace=="Black/White"))+
-  geom_point()+
-  coord_flip()+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="predicted probability of being behind expected grade")
-
-g+annotate("text", 1, 0.035, hjust="left",
-             label="Grey bands give 83.4%\nconfidence interval around\nmonoracial estimates")+
-  annotate("text", 2, 0.045, hjust="left",
-             label="Red band gives 83.4%\nconfidence interval around\nmidpoint estimate")+
-  annotate("text", 2.5, 0.036, hjust="left",
-             label="Actual biracial estimate\nshown with 83.4% confidence\ninterval")
-```
-
-This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions and their binomial standard errors (although the SEs are slightly different because this model accounts for clustering).
-The grey bands for the monoracial groups bookmark the range of results we might expect and the red band gives what would be the midpoint value between the groups, adjusting for statistical uncertainty.
-I can then see where the actual biracial estimate falls within this range.
-The results here indicate that biracial black/white students have probabilities of grade retention much closer to the lower values of white than black students.
-However, their probability is still a bit higher than that for whites.
-
-Why I am using 83% (technically 83.4%) confidence intervals?
-This is a nice visual trick.
-People often incorrectly believe that overlapping 95% confidence intervals on two means indicates that the difference in the means is not statistically significant at $p<.05$.
-This is not always true and leads in practice to a much lower $\alpha$ for rejection.
-A much better approximation is given when one uses1.386 rather than 1.96 in estimating the confidence interval, which produces an 83.4% confidence interval.
-The intuition behind this technique is given nicely [here](https://chris-said.io/2014/12/01/independent-t-tests-and-the-83-confidence-interval-a-useful-trick-for-eyeballing-your-data/).
-Because I use this interval, we can know that if intervals (or bands) do not overlap, then the difference in means between the two groups is statistically significant at $p<.05$.
-In this case, the biracial estimate is statistically significantly different from both monoracial estimates as well as the "halfway" estimate.
-In short, biracial black/white students are held back at rates much closer but somewhat higher than whites, rather than blacks.
-
-Its important to note that this technique is just an approximation because it depends on the assumption that the standard errors for the two means are the same, which is not going to be true here.
-However, it should give results that are quite close.
-Correcting for unequal SEs is far more complicated because the interval used will vary for any pairwise comparison which I am not doing here.
-
-The results here are the same as the raw differences, but I can also fit this graph to any of the models, using the average adjusted predictions.
-The plot below does this for model 4, by panelizing each multiracial group.
-
-```{r cond-means-model4, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
-Estimates are based on models that adjust for state, year, current grade, metropolitan status, gender, foreign born status, parent's foreign born status, English proficieny, parent's English proficiency, parent's education, family income, home ownership, and family structure."}
-coef_table_mod4 <- calculate_marg_means(model4)
-
-ggplot(coef_table_mod4, aes(x=term, y=estimate, color=multiracial,
-                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_mod4, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_wrap(~mrace, ncol=3, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
-```
-
-As the table here shows, after controlling for other stuff, most multiracial groups are quite similar to the monoracial constituent group with the lower grade retention rate.
-The only case where the biracial group is more like the higher monoracial group is for Latino/Asians.
-The results for indigenous groups are also quite noisy due to small size, but are generally consistent with the "below the halfway point."
-
-To compare these results across models, I can try a very large figure that panelizes by model on the column and group on the row.
-
-```{r cond-means-model-compare, fig.height=12, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
-Each panel is based on a different model that accounts for certain variables.
-Subsequent panels cumulatively include all prior terms."}
-#calculate remaining marginal means
-coef_table_mod1 <- calculate_marg_means(model1)
-coef_table_mod2 <- calculate_marg_means(model2)
-
-coef_table <- coef_table_mod0 %>%
-  mutate(model="crude")
-coef_table <- coef_table_mod1 %>%
-  mutate(model="fixed effects") %>%
-  bind_rows(coef_table)
-coef_table <- coef_table_mod2 %>%
-  mutate(model="cultural resources") %>%
-  bind_rows(coef_table)
-coef_table <- coef_table_mod4 %>%
-  mutate(model="parent resources") %>%
-  bind_rows(coef_table) %>%
-  mutate(model=factor(model, levels=c("crude","fixed effects",
-                                      "cultural resources",
-                                      "parent resources")),
-         mrace=factor(mrace, levels=c("Black/White",
-                                      "Black/Latino",
-                                      "Black/Asian",
-                                      "Black/Indigenous",
-                                      "White/Indigenous",
-                                      "White/Latino",
-                                      "White/Asian",
-                                      "Latino/Asian",
-                                      "Indigenous/Latino",
-                                      "Indigenous/Asian")))
-
-ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
-                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_grid(mrace~model, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
-
-save(coef_table, file=here("analysis","output","coef_table.RData"))
-```
-
-## Summarizing the model results
-
-Ok, let me first try to summarize these results, by group.
-
-### Black/White results
-
-For the first three models, the results are pretty consistent.
-Black/White biracials are much closer to the lower grade retention rates of whites, although they have rates that are clearly still higher than whites.
-Controlling for parental resources changes things considerably.
-First, the gap between Blacks and Whites shrinks considerably (visually it looks like its cut about in half).
-Second, Black/White biracials actually have lower rates than either group.
-They aren't that different than whites in effect size but the difference is statistically significant.
-
-Why does the big change occur in the parental resources model?
-The descriptive stat table shows that the family resources of Black/White biracials are only slightly better than those of Black students and substantially worse than those of White students.
-So, Black/White biracials are at a disadvantage relative to whites, which is controlled for in the last model.
-
-### Black/Latino results
-
-In the crude model, its clear that there isn't much difference in the grade retention rates of Blacks and Latinos, with Blacks having a slightly higher rate than Latinos.
-The Black/Latino group has rates well below either of these groups.
-In fact, their grade retention rates in the crude model are closer to Whites.
-
-Interestingly, the fixed effects model reverses the ordering of Blacks and Latinos, but this is then reversed again in the cultural resources model.
-In addition the gap between Latinos and Blacks grows a little in the cultural resources model, which is expected.
-In all of these models, the Black/Latino group remains below either of those two groups, although the gap shrinks considerably across models.
-
-Things change quite a bit in the parent resources model.
-The gap between blacks and Latinos grows and the Black/Latino group now looks almost identical to the Latino group.
-So, some of their lower probability was due to better family resources, particularly relative to Latinos, as can be seen on the table of descriptive stats.
-Once that is controlled for, their outcomes are indistinguishable from Latino students.
-
-### Black/Asian results
-
-The crude and fixed effects models indicate that Black/Asian students have grade retention rates similar to the lower grade retention rates of Asian students.
-Controlling for cultural resources changes this somewhat with a point estimate that is higher than that of Asian students.
-This makes sense because Black/Asian students are much more likely than Asian students to be native born and speak English well with native-born parents who also speak English well.
-So this gave them and advantage which is removed in that model.
-The family resources model does not change the relative position of Black/Asian students, but does shrink the overall racial gap between Black and Asian students.
-
-In the last two models, the confidence intervals overlap a bit, leading to greater uncertainty in the relatve placement of Black/Asian biracials.
-The point estimate suggests that their grade retention probability is between that of Asians and the half-way point, but the edge of the confidence intervals touch both those bands, so I cannot statistically distinguish their placement relative to those two possibilities.
-In any case, the results strongly rule out them being close to the outcomes of Black students, contra the hypodescent expectation.
-
-### Black/Indigenous results
-
-These results are quite noisy because of the smaller sample sizes involved.
-These are the two monoracial groups with the highest dropout rates, with or without controls.
-However, the difference between them is also quite large due to the very high grade retention probability for Indigenous students.
-This gap shrinks significantly in the model that adjusts for parental resources.
-
-Because of the noise its quite difficult to determine the relative placement of Black/Indigenous students at all, but we can rule out across all models that they have similar grade retentions rates to the higher grade retention of Indigenous students.
-The point estimate suggests a placement somewhere between the lower Black rate and the halfway point, but the wide CIs mean we cannot rule out either of those possibilities.
-
-The point estimate also gets much closer to the lower black rate once we control for family resources.
-
-### White/Indigenous results
-
-The results here consistently show that White/Indigenous students have grade retention rates about midway between the rates of whites and the halfway point.
-So they are clearly in the "lower half" between the groups and closer to White than Indigenous students.
-The overall gap between White and Indigenous students gets smaller with controls, but the relative placement of White/Indigenous students does not change.
-
-###  White/Latino results
-
-Importantly, the ordering of the monoracial groups shifts as controls are added to the model.
-The crude and fixed effets models show that White students are much less likely to be held back relative to Latino students.
-Controlling for cultural resources reduces this gap substantially and the inclusion of family resources flips its so that Latino students have slightly lower grade retention rates than Whites.
-So in the final model, there is not much room to find a midway point for White/Latino biracials.
-The final model shows outcomes that are similar to the lower rates of Latinos, although I need to check this more carefully because the CIs are not visible here relative to the dot (TODO).
-
-In the models prior to the inclusion of family resources, White/Latino biracials had grade retention rates similar to those of Whites, who had the lower rate.
-So despite the switch in ordering between Whites and Latinos, White/Latino biracials tend to look like the group with lower rates.
-
-### White/Asian results
-
-In all of the models prior to the inclusion of family resources, White/Asian students had grade retention rates slightly lower than the rates of Asians - the monoracial group with the lower rates.
-However, once controls are added for family resources, Asian/White students look just like Asian students (the lower group).
-This is because White/Asian students have the greatest family resources of any group in the dataset, even higher than those of White and Asian students.
-
-### Latino/Asian results
-
-This is the group with the biggest outlier in terms of the effect.
-The initial results in the crude model suggested that Latino/Asian students had rates similar to the lower rates of Asian students.
-Controlling for cultural resources shifted this to rates that were exactly midway between the two monoracial groups.
-Controlling for family resources produced a point estimate that was *higher* than either monoracial group although not statistically distinguishable from the higher rates of Latinos.
-Notably, the gap between Asians and Latinos is also quite small in the final model.
-
-The change across models is a result of the high cultural and family resources of Latino/Asian students.
-However, their family resources do not seem particularly advantaged relative to Asian students, so something a bit complex is happening in the final model there.
-They look relatively worse compared to Latinos, but the relative difference doesn't change as much with respect to Asians.
-Essentially, once I control for family resources, they are "leapfrogged" by the Latino group.
-
-### Indigenous/Latino results
-
-Controlling for cultural resources makes a big difference here.
-Initially Indigenous/Latino students look the same as the lower Latino students.
-Controlling for cultural resources makes their outcomes indistinguishably from the midway point, althoug the point estimate is still in the lower half.
-Controlling for family resources has little effect.
-
-Notably, because of sample size the error bars are substantially here.
-
-### Indigenous/Asian results
-
-This is a very small group ($n=521$) so the results are pretty noisy with wide CIs.
-However, the results are consistent across estimates.
-The grade retention rates of Indigenous/Asian students are similar to those of the lower Asian students.
-In the final model, the point estimate is almost identical to those for Asian students.
-
-Overall, I can put these results into a table where I specify the potential placement of each biracial group relative to their monoracial peers. There are seven possible outcomes:
-
-* Lower than both groups
-* Same as lower group
-* Lower half
-* Halfway between
-* Upper half
-* Same as higher
-* Higher than both groups
-
-For the two ends of the distribution and the lower half/upper half, I only consider these cases if it is statistically distinguishable from being in the other categories, respectively.
-For all other cases, I place an X in a cell if the confidence interval indicates this as a possibility.
-Asterisks indicate the placement of the point estimate for cases with uncertainty.
-
-|                   | Lower than both | Same as lower | Lower half | Halfway | Upper half | Same as higher | Higher than both |
-|-------------------|:---------------:|:-------------:|:----------:|:-------:|:----------:|:--------------:|:----------------:|
-| Black/White       |        X        |               |            |         |            |                |                  |
-| Black/Latino      |                 |       X       |            |         |            |                |                  |
-| White/Asian       |                 |       X       |            |         |            |                |                  |
-| White/Latino      |                 |       X       |            |         |            |                |                  |
-| Indigenous/Asian  |                 |       X       |            |         |            |                |                  |
-| Black Asian       |                 |       X       |     X*     |    X    |            |                |                  |
-| Black/Indigenous  |                 |       X       |     X*     |    X    |            |                |                  |
-| Indigenous/White  |                 |               |      X     |         |            |                |                  |
-| Indigenous/Latino |                 |               |            |    X    |            |                |                  |
-| Latino/Asian      |                 |               |            |         |            |        X       |                  |
-
-The results I show here are based on model 4.
-They clearly show that most biracial groups are closer in outcomes to the monoracial group with the lower rates.
-In fact, half the groups have rates that are either indistinguishable from the lower group or lower than *both* groups.
-
-Being a part-black group in no way makes one more likely to suffer from hypodescent.
-In fact, the results for Black/White biracials where we would expect hypodescent to play the largest role are actually the farthest from hypodescent as possible.
-If any group seems to suffer from a pattern of hypodescent to some degree it is part-Indigenous groups and the Latino/Asian group.
-So something else is going on here.
-
-In general, the results show that biracial groups are able to "slip through" the racialized structures that affect their more disadvantaged monoracial groups and this is not just a function of observable differences in class and nativity.
-
-## Sensitivity tests
-
-### Models by grade
-
-I want to separate the results by age more or less, but I cannot take age directly because then some kids cannot logically be behind grade at certain grade levels.
-For example, if I restrict to 11 or younger, then none of the 5th graders will be behind grade, by design.
-So, I will restrict by current grade level which indirectly gets at age.
-I am going to split it by elementary, middle school, and high school.
-
-```{r models-by-grade, warning=FALSE, fig.height=12, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
-Panels compare results when data are restricted to certain grade ranges."}
-#run separate models by grade groups
-model4_elem <- update(model4, design=subset(acs_design, current_grade<="5th"))
-model4_middle <- update(model4, design=subset(acs_design, current_grade>"5th" & 
-                                        current_grade<"9th"))
-model4_hs <- update(model4, design=subset(acs_design, current_grade>="9th"))
-
-coef_table_elem <- calculate_marg_means(model4_elem)
-coef_table_middle <- calculate_marg_means(model4_middle)
-coef_table_hs <- calculate_marg_means(model4_hs)
-
-coef_table <- coef_table_elem %>%
-  mutate(model="elementary")
-coef_table <- coef_table_middle %>%
-  mutate(model="middle school") %>%
-  bind_rows(coef_table)
-coef_table <- coef_table_hs %>%
-  mutate(model="high school") %>%
-  bind_rows(coef_table) %>%
-  mutate(model=factor(model, levels=c("elementary","middle school",
-                                      "high school")),
-         mrace=factor(mrace, levels=c("Black/White",
-                                      "Black/Latino",
-                                      "Black/Asian",
-                                      "Black/Indigenous",
-                                      "White/Indigenous",
-                                      "White/Latino",
-                                      "White/Asian",
-                                      "Latino/Asian",
-                                      "Indigenous/Latino",
-                                      "Indigenous/Asian")))
-
-ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
-                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_grid(mrace~model, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
-```
-
-The results get quite a bit noisier here because we are cutting the sample size substantially in each panel.
-The general pattern is that the results tend to look pretty similar across grade levels.
-What variations we do see could easily be the result of statistical noise as they tend to move around the most in cases with high error bands.
-The two notable cases where something might be going on:
-
-* For White/Black students, grade retention is clearly in the lower-half between Whites and the midway point rather than below whites for high school students.
-Notably, this is also where the white/black grows the largest.
-So the advantage that these students have relative to everyone else only holds for elementary and middle school.
-Nonetheless, their outcomes are still much closer to White students than Black students.
-* For White/Indigenous students, the placement varies across panels.
-In elementary school, these students are indistinguishable from White students.
-They are perfectly in the middle in middle school, and then in the lower half in high school.
-So it seems that their lowest outcomes happen earlier and on average you end up with a "lower half" across all grades.
-
-In no case do the results here suggest a biasing effect of non-biological kids.
-We would expect that to be greater in higher grades, but the results are generally consistent across grades.
-In cases where they are not consistent, you would expect the results to move toward the larger population group with age, but I do not see that.
-
-### Models with stricter parent linking
-
-```{r models-parent-link, warning=FALSE, fig.height=12, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
-Panels compare results depending on whether results are restricted only to children linked to a head of household or not."}
-#run separate models by grade groups
-model4_best_link <- update(model4, design=subset(acs_design, momrule==11 & poprule==11))
-
-coef_table_best_link <- calculate_marg_means(model4_best_link)
-
-coef_table <- coef_table_best_link %>%
-  mutate(model="only direct links")
-coef_table <- coef_table_mod4 %>%
-  mutate(model="all links") %>%
-  bind_rows(coef_table) %>%
-  mutate(model=factor(model, levels=c("all links","only direct links")),
-         mrace=factor(mrace, levels=c("Black/White",
-                                      "Black/Latino",
-                                      "Black/Asian",
-                                      "Black/Indigenous",
-                                      "White/Indigenous",
-                                      "White/Latino",
-                                      "White/Asian",
-                                      "Latino/Asian",
-                                      "Indigenous/Latino",
-                                      "Indigenous/Asian")))
-
-ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
-                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_grid(mrace~model, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
-```
-
-The results here are extraordinarily stable across both specifications.
-
-# Differences in Family Resources
-
-## Full tables of SES and cultural resources
-
-First, lets compare means of SES and cultural resources across all racial groups.
-
-```{r desc-stats-ses}
-tbl_summary <- acs %>%
-  group_by(race) %>%
-  summarize(fam_inc=round(mean(family_income), 0),
-            own_home=round(mean(own_home), 3)*100,
-            par_mar=round(mean(parents_married), 3)*100,
-            f_lhs=round(mean(degree_father=="LHS"), 3)*100,
-            f_hs=round(mean(degree_father=="HS"), 3)*100,
-            f_aa=round(mean(degree_father=="AA"), 3)*100,
-            f_ba=round(mean(degree_father=="BA" | degree_father=="G"), 3)*100,
-            m_lhs=round(mean(degree_mother=="LHS"), 3)*100,
-            m_hs=round(mean(degree_mother=="HS"), 3)*100,
-            m_aa=round(mean(degree_mother=="AA"), 3)*100,
-            m_ba=round(mean(degree_mother=="BA" | degree_mother=="G"), 3)*100)
-
-index_biracial <- which(str_detect(tbl_summary$race, "/"))
-
-tbl_summary %>%
-  mutate_if(is.numeric, format, big.mark=",") %>%
-  kbl(caption="SES resources by race. Shading indicates biracial group.",
-      align = c("l",rep("r", ncol(tbl_summary)-1)),
-      col.names=c("race","mean family income","% own home",
-                  "% parents married",
-                  "% father, no HS diploma","% father, HS diploma",
-                  "% father, AA degree",  "% father, BA+ degree",
-                  "% mother, no HS diploma","% mother, HS diploma", 
-                  "%  mother, AA degree", "% mother, BA+ degree")) %>%
-  kable_paper() %>%
-  row_spec(index_biracial, background = "#D3D3D3")
-```
-
-There is a lot going on here, but I can summarize a few things:
-
-- Most multiracial groups are more like the more privileged constituent group than the less privileged constituent group across all of these characteristics.
-For example, White/Latino respondents have family incomes of \$130,981 compared to \$138,489 for Whites and \$64,920 for Latinos.
-Similarly, 9.6% and 11.6% of Latino fathers and mothers, respectively, have a BA degree, compared to 43.8% and 47.6% for Whites, and 38.6% and 41.5% for White/Latinos.
-Also White/Latino parents own their home in 86.7% of cases which is similar to the white value of 85.5% and much higher than the 54.3% for Latinos.
-- Black/Whites in contrast, are much more like Blacks in terms of their resources and both groups have fewer resources than Whites.
-- White/Asians are the most privileged group of all, even more so than their two constituent groups.
-Their family income of \$195,028 is off the charts!
-but they also have parents with very high educational attainment.
-- Most part-Indigenous groups also buck the modal trend and have outcomes that are about halfway between their two constituent groups.
-The one part-Indigenous group that does not follow that pattern is the Indigenous/Latino group, which tend to have resources slightly better than both of the monoracial groups (although still very low compared to everyone else).
-
-Now lets look at the cultural characteristics.
-One thing to keep in mind here is that having foreign-born parents actually *lowers* your risk of grade retention, whereas being foreign-born yourself increases it.
-
-```{r desc-stats-cult}
-tbl_summary <- acs %>%
-  group_by(race) %>%
-  summarize(f_born=round(mean(foreign_born), 3)*100,
-            father_fb=round(mean(foreign_born_father), 3)*100,
-            mother_fb=round(mean(foreign_born_mother), 3)*100,
-            spk_eng_no=round(mean(spk_eng=="No"), 3)*100,
-            spk_eng_some=round(mean(spk_eng=="Somewhat"), 3)*100,
-            spk_eng_yes=round(mean(spk_eng=="Yes"), 3)*100,
-            f_spk_eng_no=round(mean(spk_eng_father=="No"), 3)*100,
-            f_spk_eng_some=round(mean(spk_eng_father=="Somewhat"), 3)*100,
-            f_spk_eng_yes=round(mean(spk_eng_father=="Yes"), 3)*100,
-            m_spk_eng_no=round(mean(spk_eng_mother=="No"), 3)*100,
-            m_spk_eng_some=round(mean(spk_eng_mother=="Somewhat"), 3)*100,
-            m_spk_eng_yes=round(mean(spk_eng_mother=="Yes"), 3)*100)
-
-index_biracial <- which(str_detect(tbl_summary$race, "/"))
-
-tbl_summary %>%
-  mutate_if(is.numeric, format, big.mark=",") %>%
-  kbl(caption="Cultural resources by race. Shading indicates biracial group.",
-      align = c("l",rep("r", ncol(tbl_summary)-1)),
-      col.names=c("race",
-                  "% foreign-born","% father foreign-born",
-                  "% mother foreign-born",
-                  "% speak no English", "% speak some English", 
-                  "% speak English",
-                  "% father speak no English", "% father speak some English", 
-                  "% father speak English",
-                  "% mother speak no English", "% mother speak some English",
-                  "% mother speak English")) %>%
-  kable_paper() %>%
-  row_spec(index_biracial, background = "#D3D3D3")
-```
-* In general, the most notable thing here is that in terms of English proficiency (both respondent and parents), all of the biracial groups involving Latino and Asian have almost universal proficiency (>99%) and more like the non-Latino and non-Asian groups.
-This is most notable for Latinos where only about two-thirds of fathers and mothers of monoracial Latinos have full English proficiency but all of the part Latino multiracial groups have parents that are 99% or more fluent.
-So, there is heavy selection on English proficiency into interracial unions, which is not surpising.
-* We also see that part-Latino and part-Asian respondents are much more likely to be native born and to have native-born parents than Latino and Asian respondents.
-The effects are large although not quite as huge as for language.
-However, because foreign-born parents actually *reduce* the effect of grade retention, the ultimate effect of this selection is more complex.
-
-## Counterfactual models by resource differentials
-
-One way to boil down the complexity of these tables into something easier to understand would be to calculate average predicted probability for each racial group, while holding constant all non-resource variables *including race*.
-These probabilities provide a counterfactual estimate of what the average probability would be for each group if they only differed on the resource variables, allowing me to isolate how different distributions of resources across racial groups affect the risk of grade retention.
-
-To do this, the following variables need to be set at some constant value to be held constant:
-
-- current grade
-- state
-- year
-- metro status
-- gender
-- race
-
-One could just plug in some static values (e.g.
-make everyone White, from Minnesota, in 6th grade, etc.), but this feels pretty ad-hoc.
-I think the best approach is to set all of these values at their mean.
-Since most of these variables are categorical variables, this means treating them as 0/1 dummies and then calculating the proportion as the mean.
-Unfortunately, this approach won't work with the models built above because they made use of factor variables.
-So, to run these models, I first need to create actual 0/1 dummies for the categorical variables listed above (current grade, state, metro status, gender, race).
-That is what the code below does.
-
-```{r create-dummied-data}
-acs_dummied <- acs %>%
-  # first we need to unorder current grade for dummy_cols to work
-  mutate(current_grade=factor(current_grade, levels=levels(current_grade), 
-                              ordered = FALSE)) %>%
-
-  dummy_cols(remove_first_dummy = TRUE) %>%
-  select(cluster, strata, perwt, below_exp_grade, year, family_income, 
-         parents_married, own_home, degree_mother, degree_father,
-         foreign_born, foreign_born_mother, foreign_born_father, 
-         spk_eng, spk_eng_father, spk_eng_mother,
-         starts_with(c("state_","current_grade_","race","sex_","metro_"))) %>%
-  mutate(year=year-2010) %>%
-  clean_names()
-```
-
-Now that I have new "dummied up" dataset, I can rereun the full model using the dummy variables which means a very, very long formula.
-
-```{r model-dummied}
-acs_dummied_design <- svydesign(ids=~cluster, weights=~perwt, data=acs_dummied)
-
-model_full <- svyglm(below_exp_grade~
-                       year*(current_grade_1st+current_grade_2nd+current_grade_3rd+
-                               current_grade_4th+current_grade_5th+
-                               current_grade_6th+current_grade_7th+
-                               current_grade_8th+current_grade_9th+
-                               current_grade_10th+current_grade_11th+
-                               current_grade_12th)+
-                       race_black+race_indigenous+race_asian+race_latino+
-                       race_black_white+race_black_indigenous+race_black_latino+
-                       race_black_asian+race_white_indigenous+race_white_latino+
-                       race_white_asian+race_indigenous_latino+
-                       race_indigenous_asian+race_latino_asian+
-                       sex_female+
-                       metro_non_metro+metro_central_city+metro_metro_non_central+
-                       metro_metro_indeterminable+
-                       state_alaska+state_arizona+state_arkansas+state_california+
-                       state_colorado+state_connecticut+state_dc+state_delaware+
-                       state_florida+state_georgia+state_hawaii+state_idaho+
-                       state_illinois+state_indiana+state_iowa+state_kansas+
-                       state_kentucky+state_louisiana+state_maine+state_maryland+
-                       state_massachusetts+state_michigan+state_minnesota+
-                       state_mississippi+state_missouri+state_montana+
-                       state_nebraska+state_nevada+state_new_hampshire+
-                       state_new_jersey+state_new_mexico+state_new_york+
-                       state_north_carolina+state_north_dakota+state_ohio+
-                       state_oklahoma+state_oregon+state_pennsylvania+
-                       state_rhode_island+state_south_carolina+state_south_dakota+
-                       state_tennessee+state_texas+state_utah+state_vermont+
-                       state_virginia+state_washington+state_west_virginia+
-                       state_wisconsin+state_wyoming+
-                       +sqrt(family_income/10000)+parents_married+own_home+
-                       degree_mother+degree_father+
-                       spk_eng+spk_eng_mother+spk_eng_father+
-                       foreign_born_mother+foreign_born_father+foreign_born, 
-                     design=acs_dummied_design, family=binomial)
-```
-
-I now have the model in the correct form.
-The next step is to create the counterfactual dataset by setting all of the non-resource variables to their mean.
-Thus, in this dataset, the only differences across racial groups are coming from their resource differences.
-
-```{r create-cf-df}
-# Create the counterfactual dataset where non-resource variables are held
-# at their mean. What needs to be set at tis mean?
-# - current_grade
-# - sex
-# - race
-# - state
-# - metro
-# - year
-means <- acs_dummied %>%
-  select(year, starts_with(c("state_","current_grade_","race_","sex_",
-                             "metro_"))) %>%
-  lapply(function(x) {rep(mean(x), length(x))}) %>%
-  bind_rows()
-
-cf_df <- acs_dummied %>%
-  select(cluster, strata, perwt, below_exp_grade, race,family_income, 
-         parents_married, own_home, degree_mother, degree_father, foreign_born, 
-         foreign_born_mother, foreign_born_father, spk_eng, spk_eng_father, 
-         spk_eng_mother) %>%
-  bind_cols(means)
-```
-
-```{r test-marg, eval=FALSE}
-# This code chunk just tests to make sure I get the same value by hand as I get 
-# when I do it through the marg command.
-temp <- subset(cf_df, race=="Black/White")
-lor <- predict(model_full, newdata=temp)
-weighted.mean(exp(lor)/(1+exp(lor)), w=temp$perwt)
-marg(model_full, "year", data=temp, weights=temp$perwt)
-# It works
-```
-
-I can then feed subsets (by race) of this counterfactual dataset into the `marg` command to get the predicted probability of grade retention for each racial group as I do below.
-
-```{r create-cf-marg}
-coefs_cf <- lapply(unique(acs$race), function(x) {
-  temp <- subset(cf_df, race==x)
-  means_margin <- marg(model_full, "year", data=temp, weights=temp$perwt)[[1]]
-  return(tibble(term=x, estimate=means_margin$Margin, 
-                se=means_margin$Standard.Error))
-}) %>%
-  bind_rows()
-
-coef_table_cf <- create_coef_table(coefs_cf)
-```
-
-And now I can plot up the results.
-
-```{r counterfactual-ses, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were resources (household income, parental education, home ownership, parent and child foreign-born status, parent and child English proficiency, and whether parents are married).
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
-ggplot(coef_table_cf, aes(x=term, y=estimate, color=multiracial,
-                          ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_cf, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_wrap(~mrace, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
-
-```
-
-I summarize the results here:
-
-|                   | Lower than both | Same as lower | Between | Same as higher | Higher than both |
-|-------------------|:---------------:|:-------------:|:-------:|:--------------:|:----------------:|
-| White/Asian       |       X         |               |         |                |                  |
-| Black/Asian       |                 |  X  (Asian)   |         |                |                  |
-| Black/Latino      |                 |  X  (Black)   |         |                |                  |
-| White/Latino      |                 |  X  (White)   |         |                |                  |
-| Indigenous/Latino |                 |X  (Indigenous)|         |                |                  |
-| Latino/Asian      |                 |  X (Asian)    |         |                |                  |
-| White/Indigenous  |                 |               |    X    |                |                  |
-| Indigenous/Asian  |                 |               |    X    |                |                  |
-| Black/Indigenous  |                 |               |    X    |                |                  |
-| Black/White       |                 |               |         |   X (Black)    |                  |
-
-In half the cases, the biracial group would have outcomes similar to the group with better resources suggesting that the parents in these matches are being positively selected.
-The White/Asian case is even more dramatic with a resource advantage over both single race groups.
-You can see this particularly for the White/Asian case in terms of family income where they have the highest of *any* group by a considerable amount.
-White/Asians are on average a very privileged group.
-
-Three groups that have the "expected" in-between status based on resources all share one thing in common - they are part-Indigenous.
-That is not true of the Indigenous/Latino case, but is of all others.
-It suggests less selectivity in crossing this boundary.
-
-Finally, the Black/White case stands out in that Black/White kids have the same risk of being held back based on their resources as do Black kids.
-Thus, they are not getting a resource boost nor does it seem like they are particularly positively selected.
-
-So the effects of controlling for these resources will be complex and different across groups.
-In general for those that are the same as lower or lower than both, we see that it pulls their relative probability up somewhat.
-For the Black/White case, it pulls their probability down on the other hand.
-For the "between" groups its more complex and doesn't seem to do much in total.
-
-### Senstivity tests separating SES and cultural differences
-
-Lets do the same procedure but I want to isolate the role of SES vs cultural (foreign-born and language).
-
-For ease of use, I am just going to plug in values here for the other resources rather than dummy everything up.
-
-```{r cf-ses-isolate, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were SES resources (household income, parental education, home ownership, and whether parents are married).
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
-cf_df_ses <- cf_df
-cf_df_ses$foreign_born <- FALSE
-cf_df_ses$foreign_born_father <- FALSE
-cf_df_ses$foreign_born_mother <- FALSE
-cf_df_ses$spk_eng <- "Yes"
-cf_df_ses$spk_eng_father <- "Yes"
-cf_df_ses$spk_eng_mother <- "Yes"
-
-coefs_cf_ses <- lapply(unique(acs$race), function(x) {
-  temp <- subset(cf_df_ses, race==x)
-  means_margin <- marg(model_full, "year", data=temp)[[1]]
-  return(tibble(term=x, estimate=means_margin$Margin, 
-                se=means_margin$Standard.Error))
-}) %>%
-  bind_rows()
-
-coef_table_cf_ses <- create_coef_table(coefs_cf_ses)
-
-ggplot(coef_table_cf_ses, aes(x=term, y=estimate, color=multiracial,
-                          ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_cf_ses, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_wrap(~mrace, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
-```
-
-```{r cf-culture-isolate, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were cultural resources (foreign-born status and English proficiency).
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
-cf_df_culture <- cf_df
-cf_df_culture$family_income <- mean(cf_df_culture$family_income)
-cf_df_culture$own_home <- TRUE
-cf_df_culture$degree_father <- "BA"
-cf_df_culture$degree_mother <- "BA"
-cf_df_culture$parents_married <- TRUE
-
-coefs_cf_culture <- lapply(unique(acs$race), function(x) {
-  temp <- subset(cf_df_culture, race==x)
-  means_margin <- marg(model_full, "year", data=temp)[[1]]
-  return(tibble(term=x, estimate=means_margin$Margin, 
-                se=means_margin$Standard.Error))
-}) %>%
-  bind_rows()
-
-coef_table_cf_culture <- create_coef_table(coefs_cf_culture)
-
-ggplot(coef_table_cf_culture, aes(x=term, y=estimate, color=multiracial,
-                          ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_cf_culture, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_wrap(~mrace, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
-```
-
-The results are a lot messier for cultural differences, partly because of the opposite effect of foreign-born parents.
-In most cases, differences point to an in-between status when there is any daylight between the constituent groups.
-So, its the SES effects that are really pulling this toward the lower group.
-
-## Summary table of resource differences
-
-Lets see if I can put all of this into a simpler table that summarize the key issues.
-
-```{r tab-resources}
-temp <- coefs_cf %>%
-  select(term, estimate) %>%
-  mutate(estimate=round(estimate,3)*100,
-         term=as.character(term)) %>%
-  rename(race=term, held_back_cf=estimate)
-
-tbl_resources <- acs %>%
-  group_by(race) %>%
-  summarize(fam_inc=round(mean(family_income), 0),
-            own_home=round(mean(own_home), 3)*100,
-            m_ba=round(mean(degree_mother=="BA" | degree_mother=="G"), 3)*100,
-            f_ba=round(mean(degree_father=="BA" | degree_father=="G"), 3)*100,
-            f_born=round(mean(foreign_born), 3)*100,
-            par_fb=round(mean(foreign_born_father | 
-                                       foreign_born_mother), 3)*100,
-            spk_eng_well=round(mean(spk_eng=="Yes"), 3)*100,
-            spk_eng_par=round(mean(spk_eng_father=="Yes" & 
-                                     spk_eng_mother=="Yes"), 3)*100) %>%
-  mutate(race=as.character(race)) %>%
-  left_join(temp) %>%
-  arrange(held_back_cf)
-
-index_biracial <- which(str_detect(tbl_resources$race, "/"))
-
-tbl_resources %>%
-  mutate_if(is.numeric, format, big.mark=",") %>%
-  kbl(caption="Resource differences by race. Shading indicates biracial group.",
-      align = c("l",rep("r", ncol(tbl_summary)-1)),
-      col.names=c("race","mean family income","% own home",
-                  "% mother, BA degree","% father, BA degree",
-                  "% foreign-born","% either parent foreign-born",
-                  "% speak English well", "% Both parents speak english well",
-                  "% held back, counterfactual")) %>%
-  kable_paper() %>%
-  row_spec(index_biracial, background = "#D3D3D3")
-
-save(tbl_resources, file=here("analysis","output","tbl_resources.RData"))
-```
-
-The table sorts groups by their counterfactual probability of being held back. I think it does a nice job of summarizing the issues.
-
-```{r time-end}
-timestamp()
-```
+The clearly preferred model is a continuous year term interacted with current grade.

### 2022-09-04 analysis/run_entire_project.sh

@@ -21,5 +21,7 @@ Rscript check_packages.R
 # send output to log directory as HTML
 Rscript -e "rmarkdown::render('organize_data.R',output_file='organize_data.html',output_dir='logs')"
 
-# Run the main analysis
-Rscript -e "rmarkdown::render('analysis.Rmd')"
+# Run the analysis
+Rscript -e "rmarkdown::render('analysis_spec_tests.Rmd')"
+Rscript -e "rmarkdown::render('analysis_main.Rmd')"
+Rscript -e "rmarkdown::render('analysis_resources.Rmd')"

### 2022-09-05 analysis/analysis_main.Rmd
Add ending time stamp to main analysis
Remove warnings in main analysis; save model summaries; description of desc stats

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, eval=TRUE)
+knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
 library(here)
@@
 
+The results clearly show substantial differences in the grade retention of single race groups.
+Indigenous students have the highest grade retention at 7.9%, while Asian students have the lowest at 2.6%.
+Whites are also relatively low at 3.5%, while Latinos and Asians have similar and higher grade retention probabilities of 5.1% and 5.5%, respectively.
+Biracial groups also vary substantially, but I will develop better ways of understanding their placement graphically below.
+These numbers also make no adjustment for important covariates.
+
+The biracial groups are generally a little bit younger than the monoracial groups, although the monoracial groups also vary in age a bit.
+Its possible this may bias the grade retention percentages down a bit, but the differences seem quite minor.
+
 # Model building
@@
 knitreg(list(model0, model1, model2, model3, model4))
+```
+
+```{r save-models}
+model0_summary <- summary(model0)
+model1_summary <- summary(model1)
+model2_summary <- summary(model2)
+model3_summary <- summary(model3)
+model4_summary <- summary(model4)
+save(model0_summary, model1_summary, model2_summary, model3_summary,
+     model4_summary, file=here("analysis","output","model_summaries.RData"))
 ```
@@
 The results here are extraordinarily stable across both specifications.
+
+```{r time-end}
+timestamp()
+```

### 2022-09-05 analysis/analysis_main.Rmd
Update description of results for new survey weighted results

@@
 Comparing Model 5 to Model 2, the controls cut the effect in half for the Black and Indigenous groups.
-It has less effect and slightly increases the gap for Asians because Asians already had lower rates than Whites.
+Controlling for resourcess doesn't change the very small gap between Whites and Asians much, and Asians continue to have slightly lower rates.
 The biggest change is for Latinos.
-In model 2, Latinos had the second highest probabilities and were 2% higher than whites.
-After controlling for everything, that actually reverse direction, such that Latinos have slightly lower expected probabilities than whites (.3% less).
+In Model 2, Latinos had the second highest probabilities and were 2.2% higher than whites.
+After controlling for everything, that actually reverse direction, such that Latinos have slightly lower expected probabilities than whites (0.2% less).
 
@@
 
-g+annotate("text", 1, 0.035, hjust="left",
+g+annotate("text", 1, 0.036, hjust="left",
              label="Grey bands give 83.4%\nconfidence interval around\nmonoracial estimates")+
-  annotate("text", 2, 0.045, hjust="left",
+  annotate("text", 2, 0.0445, hjust="left",
              label="Red band gives 83.4%\nconfidence interval around\nmidpoint estimate")+
-  annotate("text", 2.5, 0.036, hjust="left",
-             label="Actual biracial estimate\nshown with 83.4% confidence\ninterval")
+  annotate("text", 2.3, 0.0365, hjust="left",
+             label="Biracial estimate shown\nin red with 83.4% confidence\ninterval")
 ```
 
-This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions and their binomial standard errors (although the SEs are slightly different because this model accounts for clustering).
+This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions (although the SEs are slightly higher due to survey design effects of clustering and variable weights).
 The grey bands for the monoracial groups bookmark the range of results we might expect and the red band gives what would be the midpoint value between the groups, adjusting for statistical uncertainty.
@@
 
-As the table here shows, after controlling for other stuff, most multiracial groups are quite similar to the monoracial constituent group with the lower grade retention rate.
+As the table here shows, after controlling for other stuff, the majority of multiracial groups are quite similar to the monoracial constituent group with the lower grade retention rate.
 The only case where the biracial group is more like the higher monoracial group is for Latino/Asians.
-The results for indigenous groups are also quite noisy due to small size, but are generally consistent with the "below the halfway point."
+The Indigenous/Latino group is the only group statistically disginguishable as midway between the two constituent groups.
+The White/Indigenous group is also statistically distinguishable from both constitutent groups but much closer to Whites in their probability.
+The Black/Asian and Black/Indigenous groups also give point estimates at the midpoint, but the results are very noisy due to large SEs and so conclusions here would have to be speculative.
 
@@
 Why does the big change occur in the parental resources model?
-The descriptive stat table shows that the family resources of Black/White biracials are only slightly better than those of Black students and substantially worse than those of White students.
+As the analysis of family resources shows (separate report), the family resources of Black/White biracials are only slightly better than those of Black students and substantially worse than those of White students.
 So, Black/White biracials are at a disadvantage relative to whites, which is controlled for in the last model.
@@
 Interestingly, the fixed effects model reverses the ordering of Blacks and Latinos, but this is then reversed again in the cultural resources model.
-In addition the gap between Latinos and Blacks grows a little in the cultural resources model, which is expected.
 In all of these models, the Black/Latino group remains below either of those two groups, although the gap shrinks considerably across models.
@@
 The gap between blacks and Latinos grows and the Black/Latino group now looks almost identical to the Latino group.
-So, some of their lower probability was due to better family resources, particularly relative to Latinos, as can be seen on the table of descriptive stats.
+So, some of their lower probability, relative to Latinos, was due to better family resources, as can be seen on in the supplementary analysis of resources.
 Once that is controlled for, their outcomes are indistinguishable from Latino students.
@@
 
-The crude and fixed effects models indicate that Black/Asian students have grade retention rates similar to the lower grade retention rates of Asian students.
-Controlling for cultural resources changes this somewhat with a point estimate that is higher than that of Asian students.
+The crude and fixed effects models indicate that Black/Asian students have grade retention rates similar to the lower grade retention rates of Asian students, although the high SE makes the precision of this effect quite weak.
+Controlling for cultural esources moves this group closer to the midpoint value.
 This makes sense because Black/Asian students are much more likely than Asian students to be native born and speak English well with native-born parents who also speak English well.
 So this gave them and advantage which is removed in that model.
-The family resources model does not change the relative position of Black/Asian students, but does shrink the overall racial gap between Black and Asian students.
+The family resources model moves the Black/Asian group slightly more toward the middle but does not change much from the previous model.
 
-In the last two models, the confidence intervals overlap a bit, leading to greater uncertainty in the relatve placement of Black/Asian biracials.
-The point estimate suggests that their grade retention probability is between that of Asians and the half-way point, but the edge of the confidence intervals touch both those bands, so I cannot statistically distinguish their placement relative to those two possibilities.
-In any case, the results strongly rule out them being close to the outcomes of Black students, contra the hypodescent expectation.
+Due to the shrinking of the differences between the two constituent groups in the later models, the confidence interval here ends up spanning all possibilities.
+So the results here are somewhat indeterminate.
+The point estimate suggests that their grade retention probability is half-way between, but the edge of the confidence intervals touch both those bands, so I cannot statistically distinguish their placement relative to those two possibilities.
 
@@
 These results are quite noisy because of the smaller sample sizes involved.
-These are the two monoracial groups with the highest dropout rates, with or without controls.
+These are the two monoracial groups with the highest grade retention rates, with or without controls.
 However, the difference between them is also quite large due to the very high grade retention probability for Indigenous students.
@@
 
-Because of the noise its quite difficult to determine the relative placement of Black/Indigenous students at all, but we can rule out across all models that they have similar grade retentions rates to the higher grade retention of Indigenous students.
-The point estimate suggests a placement somewhere between the lower Black rate and the halfway point, but the wide CIs mean we cannot rule out either of those possibilities.
-
-The point estimate also gets much closer to the lower black rate once we control for family resources.
+Because of the noise its quite difficult to determine the relative placement of Black/Indigenous students at all.
+The point estimate in the crude model starts closer to Black students but higher than them.
+Controlling for resources consistently moves the point estimate more toward the halfway point.
+The analysis of resources shows that this group has resources midway between the two groups, so this is not surprising.
 
@@
 
-The results here consistently show that White/Indigenous students have grade retention rates about midway between the rates of whites and the halfway point.
-So they are clearly in the "lower half" between the groups and closer to White than Indigenous students.
-The overall gap between White and Indigenous students gets smaller with controls, but the relative placement of White/Indigenous students does not change.
+This group starts fairly close to the midway point, but controlling for resources clearly moves them into the "lower half" space.
+Their results never look as good as whites, but look substantially better than Indigenous students, who continue to have the highest rate of grade retention.
+Its interesting that the resource models move them closer to white given that they still have resources that are quite a bit worse than Whites.
+Presumably this is because those resources are still substantially better than Indigenous students.
 
@@
 The crude and fixed effets models show that White students are much less likely to be held back relative to Latino students.
-Controlling for cultural resources reduces this gap substantially and the inclusion of family resources flips its so that Latino students have slightly lower grade retention rates than Whites.
+Controlling for cultural resources reduces this gap substantially and the inclusion of family resources flips it so that Latino students have slightly lower grade retention rates than Whites.
 So in the final model, there is not much room to find a midway point for White/Latino biracials.
-The final model shows outcomes that are similar to the lower rates of Latinos, although I need to check this more carefully because the CIs are not visible here relative to the dot (TODO).
-
-In the models prior to the inclusion of family resources, White/Latino biracials had grade retention rates similar to those of Whites, who had the lower rate.
-So despite the switch in ordering between Whites and Latinos, White/Latino biracials tend to look like the group with lower rates.
+Nonetheless, the results do give a point estimate for the White/Latino biracial group that is below the confidence band for Latino students and distinguishable from the higher rate of whites and the halfway point.
+So, White/Latino biracials have outcomes similar to the lower grade retention rates of Latinos.
 
@@
 
-In all of the models prior to the inclusion of family resources, White/Asian students had grade retention rates slightly lower than the rates of Asians - the monoracial group with the lower rates.
-However, once controls are added for family resources, Asian/White students look just like Asian students (the lower group).
-This is because White/Asian students have the greatest family resources of any group in the dataset, even higher than those of White and Asian students.
+In all of the models, White/Asian students have grade retention rates lower than the rates of Asians - the monoracial group with the lower rates.
+Controls for family resources reduce this gap a bit because Asian/White students are the most privileged students overall and the confidence interval with Asian students overlaps slightly.
+So, we can determine that Asian/White students have grade retention rates at least as low as the lowest consistuent group (Asians) and possibly even lower.
 
@@
 The initial results in the crude model suggested that Latino/Asian students had rates similar to the lower rates of Asian students.
-Controlling for cultural resources shifted this to rates that were exactly midway between the two monoracial groups.
-Controlling for family resources produced a point estimate that was *higher* than either monoracial group although not statistically distinguishable from the higher rates of Latinos.
-Notably, the gap between Asians and Latinos is also quite small in the final model.
+Controlling for cultural resources shifted this to rates that were clost to the halfway point between the two monoracial groups.
+Controlling for family resources produced a point estimate that was *higher* than either monoracial group although also not statistically distinguishable from either monoracial group, partly because the differences between those two groups shrank so much.
 
-The change across models is a result of the high cultural and family resources of Latino/Asian students.
-However, their family resources do not seem particularly advantaged relative to Asian students, so something a bit complex is happening in the final model there.
-They look relatively worse compared to Latinos, but the relative difference doesn't change as much with respect to Asians.
-Essentially, once I control for family resources, they are "leapfrogged" by the Latino group.
+The change across models is a result of the high cultural and family resources of Latino/Asian students which compensates for slightly worse outcomes.
 
@@
 Initially Indigenous/Latino students look the same as the lower Latino students.
-Controlling for cultural resources makes their outcomes indistinguishably from the midway point, althoug the point estimate is still in the lower half.
-Controlling for family resources has little effect.
+Controlling for cultural resources makes their outcomes indistinguishably from the midway point.
+Controlling for family resources doesn't substantially change this conclusion.
 
-Notably, because of sample size the error bars are substantially here.
+Because of the sample size, the error bars are substantially here.
+Nonetheless, the placement is pretty solidly between both grouips.
+This helped by the fact that large differences remain between the groups in the final model.
+This is the only case where I observe a clearly distinguishable "halfway" result.
+There are two other cases at this level in terms of point estimates, but indeterminate because of high SEs.
 
@@
 
-This is a very small group ($n=521$) so the results are pretty noisy with wide CIs.
+This is a very small group ($n=521$) so the CIs are quite wide.
 However, the results are consistent across estimates.
 The grade retention rates of Indigenous/Asian students are similar to those of the lower Asian students.
-In the final model, the point estimate is almost identical to those for Asian students.
+This placement doesn't shift much across models.
 
@@
 
-For the two ends of the distribution and the lower half/upper half, I only consider these cases if it is statistically distinguishable from being in the other categories, respectively.
-For all other cases, I place an X in a cell if the confidence interval indicates this as a possibility.
-Asterisks indicate the placement of the point estimate for cases with uncertainty.
+I put an X in each cell if the confidence intervals indicate this placement as a possibility with an asterisks indicating the point estimate for each case.
+Results are sorted by placement of the point estimate.
 
@@
 |-------------------|:---------------:|:-------------:|:----------:|:-------:|:----------:|:--------------:|:----------------:|
-| Black/White       |        X        |               |            |         |            |                |                  |
-| Black/Latino      |                 |       X       |            |         |            |                |                  |
-| White/Asian       |                 |       X       |            |         |            |                |                  |
-| White/Latino      |                 |       X       |            |         |            |                |                  |
-| Indigenous/Asian  |                 |       X       |            |         |            |                |                  |
-| Black Asian       |                 |       X       |     X*     |    X    |            |                |                  |
-| Black/Indigenous  |                 |       X       |     X*     |    X    |            |                |                  |
-| Indigenous/White  |                 |               |      X     |         |            |                |                  |
-| Indigenous/Latino |                 |               |            |    X    |            |                |                  |
-| Latino/Asian      |                 |               |            |         |            |        X       |                  |
+| Black/White       |        X*       |               |            |         |            |                |                  |
+| White/Asian       |        X*       |       X       |            |         |            |                |                  |
+| White/Latino      |        X        |       X*      |            |         |            |                |                  |
+| Black/Latino      |        X        |       X*      |      X     |    X    |            |                |                  |
+| Indigenous/Asian  |        X        |       X*      |      X     |         |            |                |                  |
+| Indigenous/White  |                 |               |      X*    |         |            |                |                  |
+| Indigenous/Latino |                 |               |      X     |    X*   |      X     |                |                  |
+| Black Asian       |                 |       X       |      X     |    X*   |      X     |       X        |                  |
+| Black/Indigenous  |                 |       X       |      X     |    X*   |      X     |       X        |                  |
+| Latino/Asian      |                 |       X       |      X     |    X    |      X     |       X        |        x*        |
 
@@
 They clearly show that most biracial groups are closer in outcomes to the monoracial group with the lower rates.
-In fact, half the groups have rates that are either indistinguishable from the lower group or lower than *both* groups.
+6 of the 10 groups have grade retention probabilities lower, indistinguishable from, or closer to the monoracial group with lower rates.
+Of the remaining four groups, three produce results that are statistically indeterminate.
+Only one case (Indigenous/Latino) produces a clear and distinguishable "halfway" placement.
 
@@
 In fact, the results for Black/White biracials where we would expect hypodescent to play the largest role are actually the farthest from hypodescent as possible.
-If any group seems to suffer from a pattern of hypodescent to some degree it is part-Indigenous groups and the Latino/Asian group.
-So something else is going on here.
+In general there is very little evidence of a penalty for being biracial that would be consistent with hypodescent or the marginal man hypothesis.
 

### 2022-09-05 products/paper/main.qmd
Simplify descriptive stat table; Add resource difference table

@@
 load(here("analysis","output","monoracial_results.RData"))
+load(here("analysis","output","tbl_resources.RData"))
 # change names of models for display
@@
                                "Full model")))
+# create design object
+acs_design <- acs %>%
+  as_survey_design(ids = cluster, weights = perwt)
 ```
@@
 #| label: tbl-descriptive
-#| tbl-cap: Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.
-tbl_summary <- acs %>%
+#| tbl-cap: Descriptive survey-weighted statistics by race. Shading indicates biracial group. Groups ordered by sample size.
+tbl_summary <- acs_design %>%
   group_by(race) %>%
   summarize(n=n(), 
-            held_back=percent(mean(below_exp_grade), accuracy=0.01),
-            mean_age=round(mean(age), 1),
-            f_born=percent(mean(foreign_born),  accuracy=0.1),
-            par_fb=percent(mean(foreign_born_father | 
-                                  foreign_born_mother), accuracy=0.1),
-            fam_inc=dollar(round(mean(family_income), 0)),
-            own_home=percent(mean(own_home), accuracy=0.1),
-            par_ba=percent(mean(degree_father=="BA" | 
-                                  degree_father=="G" | 
-                                  degree_mother=="BA" | 
-                                  degree_mother=="G"), accuracy=0.1)
-  ) %>%
+            held_back=percent(as.numeric(survey_mean(below_exp_grade, 
+                                                     vartype=NULL)), 
+                              accuracy=0.01),
+            mean_age=round(survey_mean(age, vartype=NULL), 1)) %>%
   arrange(desc(n)) %>%
@@
                   "clearly below expected grade", 
-                  "mean age",
-                  "foreign-born",
-                  "either parent foreign-born",
-                  "mean family income",
-                  "own home",
-                  "either parent, college degree"
-      )) %>%
-  kable_styling(font_size=8) %>%
-  column_spec(3:10, width = "1.2cm") %>%
+                  "mean age" )) %>%
+  kable_styling() %>%
   row_spec(index_biracial, background = "#D3D3D3")
-#%>% landscape()
 ```
@@
 
-g+annotate("text", 1, 0.035, hjust="left",
-             label="Grey bands give 83.4%\nconfidence interval around\nmonoracial estimates")+
-  annotate("text", 2, 0.045, hjust="left",
-             label="Red band gives 83.4%\nconfidence interval around\nmidpoint estimate")+
-  annotate("text", 2.5, 0.036, hjust="left",
-             label="Actual biracial estimate\nshown with 83.4% confidence\ninterval")
+g+annotate("text", 1, 0.036, hjust="left",
+             label="Grey bands indicate 83.4%\nconfidence interval around\nmonoracial estimates")+
+  annotate("text", 2, 0.0445, hjust="left",
+             label="Red band indicate 83.4%\nconfidence interval around\nmidpoint estimate")+
+  annotate("text", 2.5, 0.0365, hjust="left",
+             label="Actual biracial estimate\nshown with 83.4%\nconfidence interval")
 ```
@@
 
+## Monoracial Differences
+
 I begin by showing the relative risk of being below expected grade among monoracial respondents across models.
@@
 #| tbl-cap: Average marginal differences in the probability of being behind grade by monoracial group, relative to a monoracial white student
-texreg(map(results_monoracial, convert_marg),
+texreg(map(results_monoracial[-1], convert_marg),
        digits=3,
@@
 ```
+
+## Resource Differences by Race
+
+```{r}
+#| label: tbl-resources
+#| tbl-cap: Mean resources by racial group. All results are survey weighted. Results sorted by counterfactual probability of grade retention. Shading indicates biracial group.
+index_biracial <- which(str_detect(tbl_resources$race, "/"))
+
+tbl_resources %>%
+  select(race, held_back_cf, fam_inc, own_home, m_ba, f_ba, f_born, par_fb,
+         spk_eng_well, spk_eng_par) %>%
+  mutate(held_back_cf=percent(held_back_cf/100, accuracy=0.1),
+         own_home=percent(own_home/100, accuracy=0.1),
+         m_ba=percent(m_ba/100, accuracy=0.1),
+         f_ba=percent(f_ba/100, accuracy=0.1),
+         f_born=percent(f_born/100, accuracy=0.1),
+         par_fb=percent(par_fb/100, accuracy=0.1),
+         spk_eng_well=percent(spk_eng_well/100, accuracy=0.1),
+         spk_eng_par=percent(spk_eng_par/100, accuracy=0.1),
+         fam_inc=dollar(fam_inc)) %>%
+  kbl(booktabs = TRUE,
+      linesep = NULL,
+      align = c("l",rep("r", ncol(tbl_resources)-1)),
+      col.names=c("race","held back, counterfactual","family income",
+                  "own home","mother, 4-year college degree",
+                  "father, 4-year college degree",
+                  "foreign-born","either parent foreign-born",
+                  "speak English well", 
+                  "Both parents speak english well")) %>%
+  kable_styling() %>%
+  row_spec(index_biracial, background = "#D3D3D3") %>%
+  column_spec(2:10, width = "1.5cm") %>%
+  landscape()
+```
+
+## Biracial Placement
 

### 2022-09-05 products/paper/main.qmd
Add counterfactual probability placement table

@@
 
+|                   | Lower than both | Same as lower | Between | Same as higher | Higher than both |
+|-------------------|:---------------:|:-------------:|:-------:|:--------------:|:----------------:|
+| White/Asian       |       X         |               |         |                |                  |
+| Black/Asian       |                 |  X  (Asian)   |         |                |                  |
+| Black/Latino      |                 |  X  (Black)   |         |                |                  |
+| White/Latino      |                 |  X  (White)   |         |                |                  |
+| Indigenous/Latino |                 |  X  (Indig.)  |         |                |                  |
+| Latino/Asian      |                 |  X  (Asian)   |         |                |                  |
+| White/Indigenous  |                 |               |    X    |                |                  |
+| Indigenous/Asian  |                 |               |    X    |                |                  |
+| Black/Indigenous  |                 |               |    X    |                |                  |
+| Black/White       |                 |               |         |   X (Black)    |                  |
+
+: Relative placement for biracial groups in comparison to monoracial constituent groups of counterfactual predicted probability of being held back based due to resources.
+
 ## Biracial Placement

### 2022-09-05 analysis/analysis_resources.Rmd

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, eval=TRUE)
+knitr::opts_chunk$set(echo = TRUE, eval=TRUE, warning=FALSE)
 library(here)

### 2022-09-05 products/paper/main.qmd

--- 2022-09-05 products/paper/main.qmd
@@
 The experiences of mixed race individuals will play an important role in the future of race in the US.
-A rise in interracial marriage and coupling since the 1960s has resulted in a "biracial baby boom" of individuals whose parents cross racial lines (==CITE==).
-This demographic phenomenon may have a potentially transformative effect on racial boundaries.
-Although race mixing has always been a feature of US history (==CITE==), this biracial boom is notable for its demographic size and for occurring in a post-Civil Rights era in which historical norms governing how to handle racial mixture may be shifting.
+A rise in interracial marriage and coupling since the 1960s has resulted in a "biracial baby boom" of individuals whose parents cross racial lines [@root_racially_1992].
+This demographic phenomenon may have a potentially transformative effect on racial boundaries [@hochschild_creating_2012; @alba_great_2020].
+Although race mixing has always been a feature of US history [@morning_who_2000; @gullickson_choosing_2011], this biracial boom is notable for its demographic size and for occurring in a post-Civil Rights era in which historical norms governing how to handle racial mixture may be shifting.
 Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of how these biracial individuals identify their own race and how they are sorted into a prevailing racial hierarchy that reinforces a system of racial inequality.
 
-Extensive scholarship documents the identification and classification decisions of multiracial individuals (==CITE==).
-==SUMMARIZE BRIEFLY==.
-On the other hand, only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers.
-(==CITATION==) The lack of work on this topic, despite its acknowledged relevance, results from two important limitations in most available data sources.
+Extensive scholarship documents the identification and classification decisions of multiracial individuals [@xie_racial_1997; @roth_end_2005a; @brunsma_interracial_2005; @herman_forced_2004; @holloway_place_2009] <!-- TODO MORE HERE -->.
+This research has generally identified a surprising degree of fluidity and choice in classification decisions, in the context of historical norms of racial classification <!-- TODO CHECK -->.
+On the other hand, only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers [@kao_racial_1999; @campbell_multiracial_2009; @herman_blackwhiteother_2009].
+The lack of work on this topic, despite its acknowledged relevance, results from two important limitations in most available data sources.
 
 First, using direct identification of biracial respondents on data sources such as the Census or large scale surveys is problematic.
-Research on the identification and classification of biracial individuals indicates that individuals whose parents belong to different races will not always or consistently identify with both races [@liebler_america_2017] (==OTHER CITE?==).
-Furthermore, social status and social mobility has been shown to predict racial identification (==CITE==) and this process of reverse causation is likely to be strongest for more ambiguous groups such as biracials.
-Thus, differences in outcome may in some part produce differences in identification rather than the other way around when examining biracial individuals identified by self-identification.
+Research on the identification and classification of biracial individuals indicates that individuals whose parents belong to different races will not always or consistently identify with both races [@doyle_are_2007; @liebler_america_2017; @gullickson_racial_2019].
+Furthermore, social status and social mobility has been shown to predict racial identification generally [@saperstein_racial_2012; @saperstein_mulatto_2013] and social status has been shown to be related to the identification choices of biracial individuals in particular [@roth_end_2005a; @davenport_role_2016].
+Thus, differences in outcome may in some part produce differences in identification rather than the other way around when examining self-identified biracial individuals.
 
@@
 When children reside with both biological parents, biracial individuals can be identified in large scale data sources such as the Census.
-Researchers have long utilized the ability to identify such biracial children to examine the classification choices made on behalf of these children (==CITATIONS==).
+Researchers have long utilized the ability to identify such biracial children to examine the classification choices made on behalf of these children [@xie_racial_1997; @roth_end_2005a].
 Because of their age, it is more difficult to analyze the outcomes of such children due to the simple fact that they haven't had many measurable life outcomes yet.
@@
 However, one important outcome, grade retention, is measurable and important even for young children.
-Being held back in school is a relatively common occurrence that has consequences for later educational transitions and is disproportionately applied by race (==CITATIONS==).
+Being held back in school is a relatively common occurrence that has consequences for later educational transitions and is disproportionately applied by race <!-- TODO CITATIONS -->.
 Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience and within it we can clearly see the imprint of the racial hierarchy.
@@
 
-The results provide a window into the placement of biracial children within an existing system of racial inequality in our educational systems. ==PITHY SUMMARY==
+The results provide a window into the placement of biracial children within an existing system of racial inequality in our educational systems. <!-- TODO  PITHY SUMMARY -->
 
@@
 For example, in the case of an individual with one Black and one White parent, we would compare the average outcomes for such individuals to the average outcomes for individuals with two White and two Black parents.
-The outcome for the Black/White group may be similar to one of the monoracial groups, it may me somewhere in between the outcomes for the two monoracial groups, or it may be higher or lower than either of the monoracial groups.
+The outcome for the Black/White group may be similar to one of the monoracial groups, it may be somewhere in between the outcomes for the two monoracial groups, or it may be higher or lower than either of the monoracial groups.
 
@@
 Discrimination and bias by race play an important role in overall levels of racial inequality.
-Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unkwnowningly rely upon and reinforce in social interactions.
+Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unknowningly rely upon and reinforce in social interactions <!-- TODO citations, may be Pager, Brubaker?
+-->.
 Such discrimination will be most consequential when performed by important gatekeepers of opportunity such as teachers, health care workers, and employers.
-
-Discrimination need not just occur on an individual basis, but can also be systemic in nature when biases are built into institutional systems.
-Such systemic racism has two effects.
-First, it provides norms and ideologies that create, support, and reinforce the cognitive schemas that individuals use in everyday practice.
-Second, norms, laws, and organizational practices can have built in biases that channel rewards to certain racial groups.
-In our current "colorblind" regime, such systems may be formally race neutral but produce disparate impacts by race.
+Discrimination need not just occur on an individual basis, but can also be systemic in nature when biases are built into organizations and institutional systems <!-- TODO citations, Tilly?
+Bonilla-Silva, Pager again?
+-->.
+In our current "colorblind" regime, such systems may be formally race neutral but produce disparate impacts by race <!-- TODO CITE Bonilla-Silva?
+-->.
 
@@
 
-The strongest historical precedent for understanding such classification in the US is the "one drop rule" norm concering Black ancestry.
-According to the one-drop rule, individuals of known Black ancestry are considered exclusively Black, regardless of other ancestries.
-Davis (==CITE==) generalizes this practice to the idea of *hypodescent* in which individuals with mixed racial ancestry will be classified as members of the lower status group.
-A symmetric, but opposite, pattern of *hyperdescent* has governed the experiences of individuals with American Indian ancestry (==CITE==).
+The strongest historical precedent for understanding such classification in the US is the "one drop rule" norm concerning Black ancestry.
+According to the one-drop rule, individuals of known Black ancestry are considered exclusively Black, regardless of other ancestries[@davis_who_1991].
+@davis_who_1991 generalizes this practice into the concept of *hypodescent* in which individuals with mixed racial ancestry will be classified as members of the lower status group.
+By comparison a pattern of *hyperdescent* implies that individuals of mixed racial ancestry will be classified as members of the higher status group.
+This pattern of hyperdescent has historically governed the experiences of individuals with American Indian ancestry <!-- TODO CITE -->.
 
-Although the hypodescent/hyperdescent paradigm is often used to theoretically frame the possible contemporary experiences of multiracial individuals, there are reasons to be cautious about the usefulness of this paradigm to both the current and historical situations.
-Although often couched in generalizable term, hypodescent and hyperdescent are actually only observed in two specific historical cases involving Black ancestry and American Indian ancestry.
-As CITE note, these two cases could just as parsimoniously be explained by a "dominance" model which ranks ancestries by their tendencies to be supercessive or recessive in determining identification and classification.
-In this framework, among the historical "racial triangle" of American colonialism, Black ancestry is supercessive, Indigenous ancestry is recessive, and White ancestry falls somewhere in between.
+Although the hypodescent/hyperdescent paradigm is often used to theoretically frame the possible contemporary experiences of multiracial individuals <!-- TODO citations -->, there are reasons to be cautious about the generalizability of this paradigm to both the contemporary and historical situation.
+Hypodescent and hyperdescent are observed in two specific historical cases involving Black ancestry and American Indian ancestry, respectively.
+As @iverson_regimes_2022 note, these two cases could just as parsimoniously be explained by a "dominance" model which ranks ancestries by their tendencies to be supercessive or recessive in determining identification and classification.
+In this framework, within the historical "racial triangle" of American colonialism, Black ancestry is supercessive, Indigenous ancestry is recessive, and White ancestry falls somewhere in between.
 
-The development of these particular norms may also reflect the historical regime in which they were developed (CITE).
-In this case, neither the hypodescent/hyperdescent or dominance paradigm may help us completely to understand the experiences of racialized groups such as Latinos and Asians.
-Second, the historical norms for all groups may also be breaking up or transforming in the post-Civil Rights era (CITE).
+The development of these particular norms may also reflect the historical regime in which they were developed [@gullickson_choosing_2011; @iverson_regimes_2022].
+In this case, neither the hypodescent/hyperdescent or dominance paradigm may help us to fully understand the experiences of those of Latino and Asian ancestry due to more recent processes of racialization.
+Second, the historical norms for all groups may also be breaking up or transforming in the post-Civil Rights era <!-- TODO CITE -->.
 
@@
 First, racial ambiguity in appearance may lead to more inconsistent classification into single race categories across observers.
-Even if the norms consistently indicate a certain type of classification, correct identification by observers is still a probabilistic phenomenon in all social interactions and likely to be more heterogenous for biracial individuals.
-As a result, biracial individuals will experience a lower sum total of discrimination and bias affecting the lower status group but will also not reap the same rewards as a member of the higher status group.
+Even if norms consistently indicate a certain type of classification, correct identification by observers is still a probabilistic phenomenon in all social interactions and likely to be more heterogeneous for biracial individuals.
+As a result, biracial individuals will experience a lower sum total of discrimination and bias affecting the lower status group in the case of hypodescent and will not reap the same rewards as a member of the higher status group in the case of hyperdescent.
 
 Second, even when "correctly" classified, greater physical ambiguity may allow biracial individuals to avoid the worst instances of bias and discrimination because they do not conform to the ideal type for the group facing the most discrimination.
-In other words, the experience of discrimination may be "softened" because their ambiguous physical appearance provokes less racial antipathy.
-The US in fact already has a long history of this process.
-Because of the one-drop rule, Black individuals vary substantially in skin tone and those individuals with a lighter skin tone have better outcomes.
-Thus light skin can moderate (but not eliminate) the stigma of Blackness even among monoracial individuals.
+In other words, biracial people's experience of discrimination may be "softened" because their ambiguous physical appearance provokes less racial antipathy.
+The US in fact already has a long history of this kind of "softening." Because of the one-drop rule, Black individuals vary substantially in skin tone and those individuals with a lighter skin tone have better outcomes [@keith_skin_1991; @hughes_significance_1990; @monk_skin_2014] and are viewed more favorably by both Blacks and Whites [@maddox_cognitive_2002a].
+Thus, light skin moderates the stigma of Blackness even among monoracial individuals.
 
-Third, it is also possible that biracial individuals may be classified into a distinct "middleman" minority position.
-This kind of middleman status is common in Latin America (CITATIONS) and Bonilla-Silva (CITE) has argued that this may be a possible future for the US as well.
+Third, it is also possible that biracial individuals may be classified into an ambiguous "middleman" minority position.
+This kind of middle tier status for individuals of mixed race is common in Latin America [@telles_race_2009] and @bonilla-silva_biracial_2004 has argued that such a system may be emerging in the contemporary US as well.
 In this case, many biracial groups may occupy a nebulous and ill-defined middleman or buffer class position in-between whites and darker-skinned non-whites, helping to cement a new kind of racial hierarchy.
@@
 
-Racial inequality may also be at least partly explained by different distributions of material, cultural, and social resources across racial groups.
-On average, non-whites come from a lower socioeconomic background (e.g.
-parental income, wealth, and education) than whites and such socioeconomic background is consequential for outcomes.
+Racial inequality is also at least partly explained by different distributions of material, cultural, and social resources across racial groups <!-- TODO CITATIONS HERE?
+-->.
+On average, socioeconomic background (e.g.
+parental income, wealth, and education) differs by race and such socioeconomic background is consequential for outcomes.
 Other less materially defined resources also vary across groups such as cultural styles, social networks, and neighborhood quality.
-When such characteristics are held constant or "controlled", the remaining differences between racial groups tend to be smaller, although such differences rarely disappear.
-These differences in background resources reflect the accumulation of historical processes of discrimination.
+When such characteristics are held constant or "controlled", the remaining differences between racial groups tend to be smaller, although such differences rarely disappear <!-- TODO WHAT CAN I CITE HERE -->.
+These differences in background resources reflect the accumulation of historical processes of discrimination <!-- CITE PAGER?
+-->.
 Thus, observed racial inequality reflects both contemporary processes of discrimination and bias and the intergenerational accumulation of disadvantage.
 
-How do these differential distributions of resources affect biracials?
-A naive argument would be that because biracial have parents from two different racial groups, they will on average inherit parental resources roughly halfway in-between their two constituent groups (CITE CHEW ET.
-AL.).
-Prior research looking at biracial children in the Census data has found some evidence to support this argument (CITE CHEW ET.
-AL.).
-If such were the case, then the resources of biracial individuals would predict outcomes in-between the two constituent groups, barring other mechanisms.
+How do these differential distributions of resources affect biracial individuals?
+A naive argument would be that because biracial individuals have parents from two different racial groups, they will on average inherit parental resources roughly halfway in-between their two constituent groups [@chew_american_1989].
+Prior research looking at biracial children in the Census data has found some evidence to support this argument [@chew_american_1989].
+In such a case, the resources of biracial individuals would predict outcomes in-between the two constituent groups, barring other mechanisms.
 
@@
 
-Status exchange theory predicts that given an unequal racial hierarchy, individuals from a lower status racial group will be positively selected on socioeconomic characteristics into intermarriage with a higher status group, while individuals from a higher status racial group will be similarly negatively selected into the same marriage (CITATION).
-Using education as the socioeconomic parameter, prior research has shown strong evidence of this pattern in the case of Black/White marriages in the US (CITATION).
-Evidence of this pattern for other forms of intermarriage is less clear (CITATION), as is research on other forms of interracial unions such as cohabitations (CITATIONS).
+Status exchange theory predicts that given an unequal racial hierarchy, individuals from a lower status racial group will be positively selected on socioeconomic characteristics into intermarriage with a higher status group, while individuals from a higher status racial group will be similarly negatively selected into the same marriage [@davis_intermarriage_1941; @merton_intermarriage_1941; @fu_racial_2001].
+Using education as the socioeconomic parameter, prior research has shown strong evidence of this pattern in the case of Black male/White female marriages in the US, but weaker evidence for other types of unions [@gullickson_education_2006; @fu_racial_2001; @kalmijn_educational_2010; @hou_interracial_2011].
+Less research has explored these patterns outside of marriage <!-- TODO cites?
+-->.
 In any case, because status exchange positively selects one parent and negatively selects the other partner, the two effects may somewhat cancel each other out in terms of the overall distribution of resources.
@@
 
+<!-- TODO how can I integrate Cheng and Powell? -->
+
 ## Hybridity
@@
 The experience of hybridity itself may also positively and/or negatively affect outcomes for biracial individuals.
-The "marginal man" theory argues that biracial individuals will find it difficult to fit in because of the dissonance of an identity that crosses strong and salient racial boundaries (CITE PARK AND STONEQUIST).
-This liminal identity conflict will then result in negative mental health experiences that can feed into other negative outcomes as well (CITE CHENG AND LIVELY?).
+The "marginal man" theory argues that biracial individuals will find it difficult to fit in because of the dissonance of an identity that crosses strong and salient racial boundaries [@park_human_1928; @stonequist_problem_1935].
+This liminal identity conflict will then result in negative mental health experiences that could feed into other negative outcomes as well.
+Evidence for such negative mental health experiences, however, has been mixed [@udry_health_2003; @campbell_what_2006; @cheng_multiracial_2009; @bratter_does_2011].
 
-Marginal man theory, developed in the first half of the Twentieth century, saw cultures as monolithic and static and thus to be trapped in-between could be a difficult experience.
-Contemporary notions of culture emphasis its "toolkit" nature, asserting a more flexible and fluid notion of culture (CITATION).
+Marginal man theory, developed in the first half of the Twentieth century, viewed culture in monolithic terms and thus to be trapped in between two static and monolithic cultures was seen as a difficult experience.
+Contemporary notions of culture emphasize its "toolkit" nature, arguing for a more flexible and fluid notion of culture <!-- TODO who to cite here?
+Swidler obviously, but also Dimaggio.
+Look at Dimaggio and Markus article -->.
 In this case, biracial individuals may benefit from an enlarged cultural toolkit that allows them to navigate a variety of diverse social contexts (i.e.
-code switch).
+code switch) <!-- TODO There are citations for this -->.
 
-The complexity of the underlying mechanisms described above prevents simple an straightforward hypothesis regarding the position of biracial individuals.
-In part, this is difficult because similar mechanisms can produce the same outcome.
-The weight of different mechanisms may also vary by groups as these groups may governed by different regimes of classification (CITATION).
+The complexity of the underlying mechanisms described above prevents simple and straightforward hypotheses regarding the position of biracial individuals.
+Different mechanisms may produce the same observed outcome and the weight of different mechanisms may also vary by groups as these groups may governed by different regimes of classification.
 However, comparing outcomes across biracial groups may help us better understand the likely mechanisms that drive biracial classification and to disregard others as implausible.
-Accounting for the effect of parental resources on outcomes to the best of our ability is also key to reaching any conclusions about the placement of biracials within the racial hierarchy.
+Accounting for the effect of parental resources on outcomes to the best of our ability is also key to reaching any conclusions about the placement of biracial individuals within the racial hierarchy.
 
@@
 With good reason, researchers have been reluctant to use data in which multiracial or biracial status is determined solely by the self-identification of respondents.
-We know that far fewer people identify as biracial than could feasibly do so based on their parents' races (CITATION) and that various selection effects operate on the decision to identify as biracial among hose who could (CITE DAVENPORT, ROTH, OTHERS).
-Furthermore, data on changes in racial reporting over time suggests that other status markers like education and income may influence how people identify themselves (CITE SAPERSTEIN AND CO).
+We know that far fewer people identify as biracial than could feasibly do so based on their parents' races <!-- TODO cite morning?
+morning and Goldstein?
+who?
+--> and that various selection effects operate on the decision to identify as biracial among those who could [@roth_end_2005a; @davenport_role_2016].
+Furthermore, data on changes in racial reporting over time suggests that other status markers like education and income may influence how people identify themselves [@saperstein_racial_2012; @saperstein_mulatto_2013].
 For example, a person with one Black and one White parent who is routinely treated by others as Black and discriminated against as Black may be more likely to identify themselves as exclusively Black than a similar person of Black and White ancestry who experiences less consistent discrimination.
-Thus, any analysis of outcomes for biracial individuals based on self-identification must deal with a serious potential for reverse causation in the observed differences between biracial individuals and their constituent groups (CITE BRATTER).
+Thus, any analysis of outcomes for biracial individuals based on self-identification must deal with a serious potential for reverse causation in the observed differences between biracial individuals and their constituent groups [@bratter_multiracial_2018].
 
 Instead of self-identification, researchers have generally relied upon survey data in which the racial identification of a respondent's biological parents are provided in some form or another.
-CITE CAMPBELL used data from the National Longitudinal Survey of Adolescent to Adult Health (Add Health) to examine high school GPA, advanced placement in math, and four-year college enrollment among biracial and monoracial individuals identified by co-resident biological parents' races.
-CITE HERMAN used data on high school students in California and Wisconsin in which the students reported their parents' races to examine differences in GPA.
+@campbell_multiracial_2009 used data from the National Longitudinal Survey of Adolescent to Adult Health (Add Health) to examine high school GPA, advanced placement in math, and four-year college enrollment among biracial and monoracial individuals identified by co-resident biological parents' races.
+@herman_blackwhiteother_2009 used data on high school students in California and Wisconsin in which the students reported their parents' races to examine differences in GPA.
 Both of these studies find some support for an "in-between" status for many biracial groups, but are both hampered by a lack of statistical power due to small sample sizes.
-CITE KAO uses data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial Blacks and Asians, and finds that biracial Blacks have outcomes similar to monoracial Blacks, while biracial Asians have outcomes more similar to monoracial Whites.
+@kao_racial_1999 used data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial Blacks and Asians, and finds that biracial Blacks have outcomes similar to monoracial Blacks, while biracial Asians have outcomes more similar to monoracial Whites.
 This study, however, is hampered by the fact that biracial students are identified by a discrepancy between their reported race and one parent's reported race, rather than by both biological parent's races.
@@
 This approach can generate much larger samples of biracial individuals, although it does exclude examination of children living in non-nuclear households.
-Furthermore, researchers then have a limited range of outcomes to explore because biracial individuals identified by this method are by definition children and the data sources includes relatively little detailed information.
-This approach has been used previously to examine the likelihood of living in poverty (CITATIONS) and residential segregation (CITATIONS).
-SUMMARIZE FINDINGS.
-While these findings provide valuable information of the context of biracial childrens' lived experience, but do not directly measure their own outcomes, but rather the experiences of their parents.
+Furthermore, researchers then have a limited range of outcomes to explore because biracial individuals identified by this method are by definition children and the data sources include relatively little detailed information.
+This approach has been used previously to examine the likelihood of living in poverty [@bratter_multiracial_2013; @bratter_poverty_2013] and residential segregation [@ellis_agents_2012].
+<!-- TODO briefly summarize findings --> While these findings provide valuable information of the context of biracial childrens' lived experience, but do not directly measure their own outcomes, but rather the experiences of their parents.
 
@@
 
+|                   | Lower than both | Same as lower | Between | Same as higher | Higher than both |
+|------------|:----------:|:----------:|:----------:|:----------:|:----------:|
+| White/Asian       |        X        |               |         |                |                  |
+| Black/Asian       |                 |   X (Asian)   |         |                |                  |
+| Black/Latino      |                 |   X (Black)   |         |                |                  |
+| White/Latino      |                 |   X (White)   |         |                |                  |
+| Indigenous/Latino |                 |  X (Indig.)   |         |                |                  |
+| Latino/Asian      |                 |   X (Asian)   |         |                |                  |
+| White/Indigenous  |                 |               |    X    |                |                  |
+| Indigenous/Asian  |                 |               |    X    |                |                  |
+| Black/Indigenous  |                 |               |    X    |                |                  |
+| Black/White       |                 |               |         |   X (Black)    |                  |
+
+: Relative placement for biracial groups in comparison to monoracial constituent groups of counterfactual predicted probability of being held back based due to resources.
+
 ## Biracial Placement

### 2022-09-06 products/paper/main.qmd
Outline of grade retention section

@@
 
--   Why grade retention as an outcome?
-
-    -   Grade retention is a relatively common practice. Discuss overall trends in grade retention and connection to educational reforms.
-    -   Grade retention has negative effect on later outcomes, even controlling for cognitive differences (CITE WORK HERE).
-    -   Clear inequalities by race in practices of grade retention as well as by SES (CITE WORK).
-    -   In effect, grade retention may be for many individuals the first encounter with the wider stratification system that sorts individuals and helps to perpetuate inequality.
-
--   What I do and how it improves our understanding.
+- Define what grade retention is and the era of "social promotion." Recent work suggests a decline in grade retention. Can cite Warren and Saliba on the trends.
+- Grade retention is consequential for later educational outcomes.
+CITE the evidence on this although, Warren and Saliba are also very iff-y on this.
+The Andrew article on "scarring" effects of grade retention is also a good one here as is the Hauser et.
+al.
+2001 piece.
+Furthermore, little evidence on the positive effects of grade retention (again cite Hauser 2001).
+Grade retention also varies by race and socioeconomic background and is thus can be seen as a mechanism for the reproduction of class and racial inequality.
+In fact, because grade retention can happen quite early in the K-12 system, for many children, it may be the first direct encounter with a "sorting" mechanism for inequality.
+- There are many challenges to measuring grade retention in existing data sources.
+One of the most common techniques is to flag students who are behind the expected grade for their age.
+This technique is imperfect because
+  - it captures a cumulative prevalence measure of grade retention rather than incidence.
+  - it cannot distinguish cases in which children were delayed in beginning school ("redshirted") from grade retention. 
+  - Students can reasonable be in two modal grades based on a given calendar year of age.
+Without detailed information on birthdates and survey timing, overall grade retention will be underestimated.
+For example, a child who is reported as 9 years old and in third grade could either be in the expected grade or could have been held back, depending on their birthday.
+A child who is 10 years old, however, is clearly behind expected grade.
+- Unfortunately, the only measure of grade retention available in the American Community Survey is based on students being clearly behind expected grade.
+The CPS has some advantages for better measuring grade retention, but lacks the sample size to precisely estimate parameters for biracial populations.
+- While being clearly behind expected grade is a problematic measure of grade retention for researchers seeking to better understand the process of grade retention itself, that is not my goal here.
+Rather, I am using this measure as an early and accessible measure of racial stratification for young children and adolescents.
+I can thus use it as a window to understand biracial students' placement within a racial hierarchy.
+This approach is similar to that employed by Rosenfeld (CITE) to examine the differential experiences of children from straight and LGBT families.
+From this perspective, the major potential problem is that the overall measure underestimates grade retention.
+However, because this underestimation is largely a function of birthdate, it is unlikely to bias the results presented here.
+- There is a question of how much racial inequality is present in grade retention rates.
+Hauser et.
+al.
+2001 suggest that while Black and Latino students are more likely to have been held back, those differences are largely explained by differences in STUFF.
+However, other more recent research finds at least moderate racial differences, particularly between Black an White students, even after controlling for a variety of other variables (cite Warren and Sariba, Rosenfeld, others?).
+Therefore, understanding the effect of socioeconomic and other resources on the differences between monoracial groups is also key to understanding the larger research question, just as it is for biracial students.
 

### 2022-09-07 products/paper/main.qmd
Edit of data and methods section

@@
 
-Data for this analysis come from the American Community Survey (ACS), an annual 1-in-100 survey of the US population, produced by the Census Bureau.
+Data for this analysis come from the American Community Survey (ACS), an annual 1-in-100 survey of the US population, conducted by the Census Bureau.
 To increase sample size for smaller populations of biracial respondents, I pool ACS data for a full decade from 2010-2019.
@@
 Ideally, I would restrict the sample only to children living in a household with two biological parents.
-However, the ACS only records the direct family relationship between the head of household andother members of the family.
+However, the ACS only records the direct family relationship between the head of household and other members of the family.
 Since 2008, the ACS also distinguishes between biological children, adopted children, and stepchildren.
@@
 Prior work has used a variety of additional restrictions to limit the analysis to children who are more likely to be the biological children of both parents [@saenz_persistence_1995; @xie_racial_1997].
-Some of those procedures rely upon an assumption that most parents would be married that was more tenable in historical data than contemporary data.
-Therefore, I do not explicitly restrict by marital status, I do restrict cases in two other ways that are consistent with prior work.
+Some of those procedures rely upon an assumption that most parents would be married that was more tenable in historical than contemporary data.
+I do not explicitly restrict by marital status, but I do restrict cases in two other ways that are consistent with prior work.
 First, I restrict the sample to cases where both parents were of a reasonable age at the birth of the child (aged 15-44 for mothers and aged 15-60 for fathers).
-Second, I restrict the sample to only those cases where the reported race and Hispanicity of the child is consistent with at least one of the parents.
+Second, I restrict the sample to only those cases where the reported race and Hispanicity of the child is inclusive of at least one of the parents' races.
 
 The final analytical sample consists of children aged 5-20 who are currently enrolled in the K-12 school system and live in a household with two parents who are both likely to be biological parents.
-This necessary sample restriction will bias the family structure of the analytical sample relative to the total population of K-12 schoolchildren, limiting it exclusively to households with two parents present and over-representing the proportion of cases in which those parents are married.
+This necessary sample restriction will bias the family structure of the analytical sample relative to the total population of K-12 schoolchildren, limiting it exclusively to households with two parents present.
 The result of this bias is well-known.
-Children living in two-parent nuclear households have greater educational success than other children as a result of material, social, and cultural resources (CITATIONS).
-Therefore, this sample restriction will upwardly bias estimates of grade retention.
+Children living in two-parent nuclear households have greater educational success than other children as a result of material, social, and cultural resources [@biblarz_family_1999].
+Therefore, this sample restriction will bias overall estimates of grade retention downward.
 However, this bias is less problematic for the goal of this study, which is to estimate differences between groups.
@@
 The cross-tabulation of parent's races leads to ten distinct biracial categories.
-The total sample size for each of these biracial categories is shown in @tbl-descriptive, along with several descriptive statistics.
-The sample sizes are quite large for many of the groups, suggesting substantial statistical power in determining the relative outcomes for these groups.
-Even among the smallest group of Indigenous/Asian, the sample size is substantially greater than one would see in typical survey data.
+The total sample size for each of these biracial categories is shown in @tbl-descriptive, along with the sample size of the monoracial comparison groups.
+The sample sizes are quite large for most groups and far in excess of comparable categories in other available survey data.
 
@@
 
-To measure grade retention, I compare the age of respondents to their current grade in school.
-Because the ACS lacks data on the exact birthday of respondents, it is not possible to exactly identify students who are below the expected grade level.
-A student of any given age in years can always be expected in two different grades.
+Measuring grade retention exactly is not possible in the ACS data.
+Without data on date of birth and survey timing, a student of any given age who has not been held back could reasonably be in two different grades.
 For example, a first grader is expected to be either six or seven years of age, while a second grader is expected to be either seven or eight years of age.
-Therefore, with information only on current grade and age in years, it is impossible to determine whether a seven year old in first grade is behind the expected grade level.
+Therefore, with information only on current grade and age in years, it is impossible to determine whether a seven year old in first grade is behind the expected grade level or in the expected grade.
 
-Instead of trying to identify exact grade retention, I instead identify cases in which students are clearly below expected grade.
+Instead of trying to identify exact grade retention, I instead identify cases in which students are clearly behind expected grade.
 A student is clearly behind expected grade when the age of the student is greater than either of the possible ages for a student of that grade (e.g.
 a first grader who is eight years old).
-This measure will underestimate the total proportion of students that have been held back.
+This indirect measure of grade retention has been frequently used in previous studies of grade retention outcomes [@bianchi_children_1984; @frederick_have_2008; @rosenfeld_nontraditional_2010].
+This measure will underestimate the total proportion of students that have been held back and may also suffer from other biases [@warren_patterns_2014].
 However, the goal of this study is to examine differences across groups rather than total levels.
-The only factor determining this bias is whether the birthday of the student takes place earlier or later in the year, and is unlikely to vary substantially across groups.[^1]
+The primary factor determining bias is whether the birthday of the student takes place earlier or later in the year, which is unlikely to vary substantially across groups.[^1]
 
@@
 
-==ACKNOWLEDGE POTENTIAL ISSUE WITH DROPOUT STUDENTS==
-
-@tbl-descriptive shows the percent of each monoracial and biracial group that is clearly below the expected grade.
+@tbl-descriptive shows the percent of each monoracial and biracial group that is clearly behind expected grade.
 My goal is to understand the degree of grade retention for each biracial group relative to its constituent monoracial groups (e.g.
@@
 The point estimate and confidence interval for the Black/White group are shown in red, while the two monoracial groups are shown via grey dots and confidence bands.
-For all three cases, I use 83.4% confidence intervals, because overlap on such intervals indicates that the observed difference between the two measures in the sample is not statistically distinguishable from zero in the population at $p<.05$ (CITE).[^2] In this case, the lack of overlap between any of the confidence bands indicates that the Black/White probability of being clearly below expected grade is statistically distinguishable from both monoracial probabilities.
+For all three cases, I use 83.4% confidence intervals, because overlap on such intervals indicates that the observed difference between the two measures in the sample is not statistically distinguishable from zero in the population at $p<.05$ [@knol_mis_2011].[^2] In this case, the lack of overlap between any of the confidence bands indicates that the Black/White probability of being clearly behind expected grade is statistically distinguishable from both monoracial probabilities.
 
-[^2]: The correspondence to $p<.05$ is only approximate due to differences in the two standard errors between estimates. However, this approximation will be extremely close in almost all cases.
+[^2]: The correspondence to $p<.05$ is only approximate due to differences in the standard errors between groups.
+Calculating more precise intervals requires that those intervals vary across each pairwise comparison which does not work for the visualizations here.
+The approximation will be very close in all cases used here.
 
@@
 
-[^3]: The standard error of this halfway estimate used to construct the confidence interval is given by $\sqrt{(s_1^2+s_2^2)/2}$ where $s_1$ and $s_2$ are the standard errors of the estimates for the two monoracial groups.
+[^3]: The halfway point is given by taking the mean between the two monoracial group.
+Its standard error is given by $\sqrt{(s_1^2+s_2^2)/2}$ where $s_1$ and $s_2$ are the standard errors of the estimates for the two monoracial groups.
 
@@
 I then construct figures similar to @fig-example by calculating from the model the average predicted probabilities (APP) for each racial group.
-APPs estimate the average probability of the outcome by some covariate across all cases, while holding constant all other variables.
-APPs are akin to average marginal effects (AME) and estimated in the same way via the delta method.
+An APP estimates the average probability of the outcome by a given covariate across all cases, while holding constant all other variables.
+APPs are akin to average marginal effects (AME) and estimated in the same way.
 Whereas AMEs estimate differences or slopes, APPs estimate the level for a given category.[^4]
@@
 First, I include fixed effects for state of residence because states dictate educational policy and thus can differ substantially in the likelihood of grade retention.
-I cannot identify specific school districts, but I also include dummy variables indicating whether the student lived in a central city, suburban, or rural area.[^5]
+I also include dummy variables indicating whether the student lived in a central city, suburban, or rural area.[^5]
 
@@
 
-The probability of being clearly below expected grade also increases with the student's current grade, so I include fixed effects for the current grade of the student.
+The probability of being clearly behind expected grade also increases with the student's current grade, so I include fixed effects for the current grade of the student.
 Additionally, @fig-year-grade shows that the percent of students clearly behind grade level has declined over time, but this decline has been much more substantial at higher grade levels.
@@
 [^6]: Sensitivity analysis showed that this functional form was preferred by the Bayesian information criterion to models with no interaction and a model with interaction terms that treated year as a categorical variable.
-I also considered similar interactions between state and year, but none of the models with these interaction terms were preferred.
+I also considered a similar interaction between state and year, but this model was not preferred.
 
@@
 Second, I include a categorical measure of highest degree earned for each parent.
-Finally, I include measures of family income (square rooted), home ownership, and whether the parents are married that capture other family resources.
-All standard errors reported here also account for the clustering of multiple respondents within the same household.
+Finally, I include measures of family income (square rooted), home ownership, and whether the parents are married.
+
+All statistics reported here adjust for sampling weights among respondents.
+All models incorporate design effects for variance in sample weights abd the clustering of multiple respondents within the same household.
 
@@
 #| label: tbl-monoracial
-#| tbl-cap: Average marginal differences in the probability of being behind grade by monoracial group, relative to a monoracial white student
-texreg(map(results_monoracial[-1], convert_marg),
+#| tbl-cap: Average marginal differences in the probability of being behind grade by monoracial group, relative to a monoracial white student.
+screenreg(map(results_monoracial[-1], convert_marg),
        digits=3,
@@
               "Parent's education" = c("No","No","Yes","Yes"),
-              "Family resources" = c("No","No","No","Yes")))
+              "Other family resources" = c("No","No","No","Yes")))
 ```

### 2022-09-07 products/paper/main.qmd
Add section on resource differentials; edit section on monoracial differences

@@
 
-I begin by showing the relative risk of being below expected grade among monoracial respondents across models.
-@tbl-monoracial reports average marginal effects on the probability of grade retention for each monoracial minority group in comparison to White students.
+I begin by showing the relative risk of being behind expected grade among monoracial respondents across models.
+@tbl-monoracial reports average marginal effects on the probability of being behind expected grade (hereafter BBEG) for each monoracial minority group in comparison to White students.
 Understanding these differences and how they change across models helps me to understand the potential for how biracial respondents might be positioned.
-
-Model 1 only controls for structural factors such as state of residence, student location, grade, and year.
-These results provide a baseline estimate of the differences across monoracial groups without controlling for differences in cultural and material resources across groups.
-The results show that Black, Indigenous, and Latino students all have substantially higher probability of grade retention than White students.
-Indigenous students have substantially higher risk than all other students, with a probability of grade retention that is 4.1% percentage points higher than White students.
-Asian students have the lowest probability of grade retention, and their risk is slightly lower than Whites.
-
-The subsequent models control for a variety of additional variables.
-Model 2 includes controls for whether the student and each of their parents are native-born and speak English well.
-Model 3 controls for the highest degree received for each parent, and Model 4 controls for family income (square rooted), home ownership, and whether the student's parents are married.
-
-In general Model 2 and Model 3 have substantial effects on the observed inequality across racial groups, while the additional variables from Model 4 have less impact.
-Controlling for the characteristics related to immigration in Model 2 cuts the gap between White and Latino students in half and increases the gap between Asian and White students.
-It has no impact on the gap between White and Black or White and Indigenous students because most of these students are native-born with native-born parents.
-
-Controlling for education in Model 3 reduces the gap substantially for Black and Indigenous students and removes the gap entirely for Latino students.
-In fact, the results of both Models 3 and 4 suggest that, when holding constant cultural and family resources, Latino students are slightly less likely than White students to be held back.
-
-SUMMARIZE THIS?
 
@@
 #| tbl-cap: Average marginal differences in the probability of being behind grade by monoracial group, relative to a monoracial white student.
-screenreg(map(results_monoracial[-1], convert_marg),
+texreg(map(results_monoracial[-1], convert_marg),
        digits=3,
@@
 
-## Resource Differences by Race
+Model 1 only controls for structural factors such as state of residence, student location, grade, and year.
+These results provide a baseline estimate of the differences across monoracial groups without controlling for differences in cultural and material resources across groups.
+The results show that Black, Indigenous, and Latino students all have substantially higher probability of BBEG than White students.
+Indigenous students have substantially higher risk than all other students, with a probability of BBEG that is 4.1% percentage points higher than White students.
+Asian students have the lowest probability of BBEG, and their risk is slightly lower than Whites.
+
+The subsequent models control for a variety of additional variables.
+Model 2 includes controls for whether the student and each of their parents are native-born and speak English well.
+Model 3 controls for the highest degree received for each parent, and Model 4 controls for family income (square rooted), home ownership, and whether the student's parents are married.
+
+Model 2 and Model 3 have substantial effects on the observed inequality across racial groups, while the additional variables from Model 4 have less impact.
+Controlling for the characteristics related to immigration and acculturation in Model 2 cuts the gap between White and Latino students in half and increases the gap between Asian and White students.
+It has no impact on the gap between White and Black or White and Indigenous students because most of these students are native-born with native-born parents.
+
+Controlling for education in Model 3 reduces the gap substantially for Black and Indigenous students and removes the gap entirely for Latino students.
+In fact, the results of both Models 3 and 4 suggest that, when holding constant cultural and family resources, Latino students are slightly less likely than White students to be held back.
+
+In total, resource differentials account for a substantial part of the overall racial differences in the risk of BBEG, but not all of these differences.
+Black and Indigenous students remain at higher risk than other students.
+Notably, Asian students rather than White students have the lowest risk of BBEG in all models.
+The most notable shift across models is for Latino students.
+Their higher risk of BBEG relative to White students is completely accounted for by resource differentials.
+When comparing hite and Latino students with the same level or resources in Model 4, Latino students actually have slightly lower risk of BBEG.
+
+Although, the racial differences in risk of BBEG are substantially reduced by controlling for resource differentials, the remaining gaps across all pairwise combinations are still large enough in most cases to sustain the question of where biracial individuals fit within these gaps.
+Before turning to this question, however, I want to better understand how the resources of biracial individuals compare to their monoracial constituent groups.
+
+## The Distribution of Resources for Biracial Individuals
+
+@tbl-resources shows the mean values of the resource variables across monoracial and biracial groups.[^7] Because of the number of variables and groups involved, this table can be overwhelming to read.
+Therefore, I also provide a summation of these resource differentials by racial group by calculating a counterfactual probability of BBEG for each group based on their observed distribution of resource characteristics.
+This counterfactual probability is calculated by estimating the average predicted probability for each group from the fullest model (Model 4 of \@tbl-monoracial) when all other variables, including race, are held at their mean.[^8]
+
+[^7]: For compactness, I combine the parent foreign-born and English proficiency questions into whether either parent is foreign-born or speak English.
+In the models, I use separate variables for mothers and fathers.
+Tables in the supplementary materials provide the full breakdown of these variables by racial group.
+
+[^8]: It might seem odd to hold categorical variables at their mean, but because all categorical variables are technically entered into the models as 0/1 numeric dummy variables, their mean value can be calculated as the proportion of cases with a value of 1.
+This technique is frequently used in multilevel models to deal with grand and group mean centering of variables.
+<!-- TODO CITE raudenbush and bryk?
+-->
 
@@
       align = c("l",rep("r", ncol(tbl_resources)-1)),
-      col.names=c("race","held back, counterfactual","family income",
+      col.names=c("race","behind expected grade, counterfactual","family income",
                   "own home","mother, 4-year college degree",
@@
 ```
+
+The results indicate that we would expect very different risk of BBEG across biracial groups because of differences in resources.
+For example, White/Asian students have the lowest counterfactual risk of BBEG of all groups at 2.4%, owing to their exceptionally high family income and the high college attainment of their parents.
+At the other end of the spectrum, Indigenous/Latino students have a counterfactual risk of BBEG nearly 2% higher, given their much lower family resources.
+
+My particular interest, however, is in how each biracial group compares to its constituent monoracial groups.
+For example, while Indigenous/Latino students have the highest counterfactual risk of BBEG among biracial students, their risk is still lower than their two constituent groups of Indigenous (4.5%) and Latino (5.3%) students, owing in large part to a somewhat higher family income, greater parental educational attainment, and greater acculturation.
+These results suggest positive selection into Indigenous/Latino interracial unions.
 
@@
 
-: Relative placement for biracial groups in comparison to monoracial constituent groups of counterfactual predicted probability of being held back based due to resources.
+: Relative placement for biracial groups in comparison to monoracial constituent groups of counterfactual probability of being behind grade from resources alone.
+
+Table 4 summarizes the patterns from @tbl-resources across all ten biracial groups.
+In six of these cases, the biracial group would have outcomes similar or better than the monoracial group with better resources.
+These results imply that for the majority of these groups, parents in these matches are being positively selected from their constituent groups.
+
+Only three groups exhibit an in-between status based on resources.
+All three of these groups are part-Indigenous.
+This finding does not hold for Indigenous/Latino students, but nonetheless suggests less selectivity in crossing this boundary.
+
+The Black/White case stands out as an outlier.
+Black/White students have similar resources as Black students and both groups have fewer resources than White students.
+This is observable in @tbl-resources across family income, home ownership, and parental educational attainment.
+Unlike most of these other groups, there does not seem to be substantial positive class selectivity into Black/White interracial unions.
+The effect of controlling for these resources will be complex and different across biracial groups.
+For those biracial groups with resources similar to their more advantaged constituent group, we should see that these controlling for these resources should reduce their advantage relative to the less advantaged monoracial group.
+For the Black/White case, on the other hand, we would expect that controlling for resource differentials should eliminate some of their disadvantage relative to Whites.
+For the remaining “in-between” groups, it is more difficult to make a priori predictions.
 
@@
 Full models include control variables for nativity, English proficiency, income, education, home ownership, and marital status of parents.
-#| fig-height: 7.5
+#| fig-height: 7.3
 coef_table %>%

### 2022-09-07 products/paper/main.qmd
Edit biracial results section

@@
 I focus first on biracial groups with one Black parent because of a stronger historical expectation that these students will be identified as Black by the "one-drop rule" and as a result will have similar outcomes to Black students.
-In three cases, this hypothesis can be soundly rejected.
-For Black/White, Black/Latino, and Black/Asian students, The probability of being clearly below expected grade is substantially lower than that for monoracial black students and the difference is statistically distinguishable from zero.
-Only in the case of Black/Indigenous students, do I observe outcomes that are indistinguishable from those of Black students.
-However, in this case,monoracial Black students have lower risk of grade retention than monoracial Indigenous students.
-Thus, In all four cases, part-Black biracial students have a risk of grade retention that is closer to the lower risk monoracial group.
+@fig-black-models provides little evidence for such a pattern, and soundly rejects it in the case of Black/White and Black/Latino students.
+Black/White and Black/Latino students have a risk of BBEG substantially lower than Black students and closer, and perhaps even lower, than their non-Black constituent group.
+Solid conclusions are difficult for the remaining two groups because of smaller sample size and correspondingly wider confidence intervals on estimates.
+Nonetheless the point estimates in each of these cases suggest probabilities of BBEG roughly midway between the two constituent groups.
 
-The results also suggest that each of these biracial groups has a risk of grade retention lower than the halfway point, although this hypothesis cannot be rejected completely for the Black/Asian and Black/Indigenous groups, due to a higher degree of uncertainty for these smaller groups.
-
-Controlling for resource variables has some effect on the placement of Black/White and Black/Latino biracial respondents.
-In the case of Black/White biracial respondents, Black/White students initially have a risk of grade retention slightly higher than and statistically distinguishable from White students.
-After controlling for cultural and family resources, however, Black/White respondents have a risk of grade retention that is slightly lower and statistically distinguishable from White students.
+Controlling for resource variables has some effect on the placement of part-Black biracial respondents.
+In the case of Black/White biracial respondents, Black/White students initially have a risk of BBEG slightly higher than and statistically distinguishable from White students.
+After controlling for cultural and family resources, however, Black/White respondents have a risk of BBEG that is slightly lower and statistically distinguishable from White students.
 This change reflects the relatively low level of parental education and income among Black/White students, which is much closer to Black students than White students, as shown on @tbl-descriptive.
@@
 Black/Latino respondents initially have a risk of grade retention substantially lower than both Black and Latino students, who have similar risks.
-However, after controlling for cultural and family resources, Black/Latino students have a risk of grade retention that is indistinguishable from the grade retention risk of Latino students and substantially lower than Black students.
-The change largely reflects the fact that Black/Latino students have greater cultural resources than Latino students and once this advantage is accounted for, they have a similar risk.
+However, after controlling for cultural and family resources, Black/Latino students have a risk of BBEG that is indistinguishable from the risk of Latino students and substantially lower than Black students.
+The change largely reflects the fact that Black/Latino students have greater resources than Latino (but not Black) students and once this advantage is held constant, they have a similar risk.
 
-Altogether, the results here strongly suggest that the risk of grade retention faced by these biracial students with one black parent is generally similar to that of their constituent monoracial group with the lower risk of grade retention.
-This finding is inconsistent both with the expectation of the one-drop rule and the "in-between" hypothesis.
+The role of parental resources is harder to determine for Black/Asian and Black/Indigenous students due to the greater uncertainty in these estimates.
+Controlling for resources moves the point estimate for Black/Asian students from a point estimate close to Asian students to more of a halfway position between the two constituent groups.
+Because Black/Indigenous students had resources halfway between their constituent groups, controlling for these resources hardly affects the relative placement of Black/Indigenous students at all.
 
 @fig-other-models shows the results for the remaining biracial groups.
-In general, the results here are consistent with those for part-Black biracial respondents, with some important caveats discussed below.
-The point estimates indicate that the risk of each biracial group is either quite close to the monoracial group with the lower risk or in-between this group's risk and the halfway point.
+In general, the results here show that biracial students are more likely to have a risk of BBEG more like the monoracial constituent group with the lower risk than the higher risk, with some important caveats discussed below.
 
@@
 
-For all groups, the full models generally indicate that each biracial group has a risk of grade retention similar to the risk of the lower monoracial constituent group or has risk higher than this group but less than the halfway point.
+For all groups, the full models generally indicate that each biracial group has a risk of BBEG similar to the risk of the lower monoracial constituent group or has risk higher than this group but less than the halfway point.
 There are two exceptions to this general pattern.
-First, for the Indigenous/Latino group, I cannot rule out the possibility that this group has risk roughly halfway between the two constituent groups, although the point estimate favors a result closer to the lower Latino group.
-Second, Latino/Asian students are the only case where the risk for the biracial group is higher than either monoracial group.
-However, the confidence bands suggest that the risk for the Latino/Asian group cannot be differentiated from the rates for Latinos, the higher of the two monoracial constituent groups.
-This result marks the one substantial deviation from the general pattern.
-It only emerges in the models that control for cultural and family resources because Latino/Asian students come from higher income and more educated households than Latino students, on average.
+First, the point estimate of the probability of BBEG for the Indigenous/Latino group is roughly halfway between the two constituent groups and statistically distinguishable from both.
+Second, Latino/Asian students are the only case where the point estimate of the risk for the biracial group is higher than either monoracial group.
+However, the confidence bands suggest that the risk for this group cannot be statistically differentiated from either of the two monoracial constituent groups.
+This result only emerges in the model that controls for resources, Latino/Asian students come from more advantaged households than both Latino and Asian students, on average.
 
-Aside from the effect on Latino/Asian students, controlling for cultural and family resources has moderate effects on the placement of the remaining biracial groups.
-The only case with a substantial shift is that for White/Asian students who initially appear to have lower risk than either White or Asian students.
-After controlling for these resources, the risks of White/Asian students are indistinguishable from the lower risk of Asian students.
-This is due to the fact that White/Asian students come from the most advantaged households of any group in the dataset.
+Aside from the effect on Latino/Asian students, controlling for resources has moderate effects on the placement of the remaining biracial groups.
+White/Asian students are the only case with a substantial shift, owing to their highly advantaged households.
+After controlling for resources, the risks of BBEG for White/Asian students are indistinguishable from the risk of Asian students, but still substantially and statistically distinguishable from the higher risk to White students.
 
-Any results for part-Latino students are complicated by the substantial change in the monoracial Latino risk across models.
-For example, while Latinos have substantially higher risk of grade retention than Whites in the baseline model, there is only a very small difference favoring Latinos in the full model.
+Results for part-Latino students are complicated by the substantial change in the monoracial Latino risk across models.
+For example, while Latinos have substantially higher risk of BBEG than Whites in the baseline model, there is only a very small difference favoring Latinos in the full model.
 Nonetheless, even in this case, we see that White/Latinos have risk closer to those of the lower Latino group, but distinguishable from both groups.

### 2022-09-07 products/paper/main.qmd
Make improvements to tables

@@
       align = c("l",rep("r", ncol(tbl_summary)-1)),
-      col.names=c("race",
-                  "n",
-                  "clearly below expected grade", 
-                  "mean age" )) %>%
+      col.names=c("Race",
+                  "Sample size",
+                  "Clearly behind expected grade", 
+                  "Mean age" )) %>%
   kable_styling() %>%
@@
 
-[^8]: It might seem odd to hold categorical variables at their mean, but because all categorical variables are technically entered into the models as 0/1 numeric dummy variables, their mean value can be calculated as the proportion of cases with a value of 1.
+[^8]: It might seem odd to hold categorical variables at their mean, but because all categorical variables are technically entered into the models as 0/1 numeric indicator variables, their mean value can be calculated as the proportion of cases with the indicated value.
 This technique is frequently used in multilevel models to deal with grand and group mean centering of variables.
@@
   kbl(booktabs = TRUE,
-      linesep = NULL,
+      linesep = NULL, position = "p",
       align = c("l",rep("r", ncol(tbl_resources)-1)),
-      col.names=c("race","behind expected grade, counterfactual","family income",
-                  "own home","mother, 4-year college degree",
-                  "father, 4-year college degree",
-                  "foreign-born","either parent foreign-born",
-                  "speak English well", 
+      col.names=c("Race","Behind expected grade, counterfactual","Family income",
+                  "Own home","Mother, 4-year college degree",
+                  "Father, 4-year college degree",
+                  "Foreign-born","Either parent foreign-born",
+                  "Speak English well", 
                   "Both parents speak english well")) %>%
-  kable_styling() %>%
+  kable_styling(latex_options = c("scale_down")) %>%
   row_spec(index_biracial, background = "#D3D3D3") %>%
-  column_spec(2:10, width = "1.5cm") %>%
-  landscape()
+  column_spec(2:10, width = "1.5cm")
+#%>%
+#  landscape()
 ```
@@
 
-|                   | Lower than both | Same as lower | Between | Same as higher | Higher than both |
-|------------|:----------:|:----------:|:----------:|:----------:|:----------:|
-| White/Asian       |        X        |               |         |                |                  |
-| Black/Asian       |                 |   X (Asian)   |         |                |                  |
-| Black/Latino      |                 |   X (Black)   |         |                |                  |
-| White/Latino      |                 |   X (White)   |         |                |                  |
-| Indigenous/Latino |                 |  X (Indig.)   |         |                |                  |
-| Latino/Asian      |                 |   X (Asian)   |         |                |                  |
-| White/Indigenous  |                 |               |    X    |                |                  |
-| Indigenous/Asian  |                 |               |    X    |                |                  |
-| Black/Indigenous  |                 |               |    X    |                |                  |
-| Black/White       |                 |               |         |   X (Black)    |                  |
+```{r}
+#| label: tbl-resource-summary
+#| tbl-cap: Relative placement for biracial groups in comparison to monoracial constituent groups of counterfactual probability of being behind grade from resources alone.
+race <- c("White/Asian","Black/Asian","Latino/Asian","White/Latino",
+          "Black/Latino","Indigenous/Latino","White/Indigenous",
+          "Black/Indigenous","Indigenous/Latino","Black/White")
+lower_both <- c("X",rep("",9))
+same_lower <- c("","X (Asian)","X (Asian)","X (White)","X (Black)","X (Indig.)",
+                rep("",4))
+between <- c(rep("",6),rep("X",3),"")
+same_higher <- c(rep("",9),"X (Black)")
+higher_both <- rep("",10)
+tibble(race, lower_both, same_lower, between, same_higher, higher_both) %>%
+ kbl(booktabs = TRUE,
+      linesep = NULL,
+      align = c("l",rep("r", 5)),
+      col.names=c("Biacial","Lower than both","Same as lower","Between",
+                  "Same as higher","Higher than both")) %>%
+  kable_styling() %>%
+  column_spec(1:6, width = "2cm")
 
-: Relative placement for biracial groups in comparison to monoracial constituent groups of counterfactual probability of being behind grade from resources alone.
+```
 

### 2022-09-07 analysis/analysis_main.Rmd
Add rough code for new approach to figures

@@
 
+```{r}
+marg(model4, "race", type="effects",
+     at_var_interest=c("White","Black/White","Black"))
+marg(model4, "race", type="effects",
+     at_var_interest=c("Black","Black/Latino","Latino"))
+marg(model4, "race", type="effects",
+     at_var_interest=c("Asian","White/Asian"))
+marg(model4, "race", type="effects",
+     at_var_interest=c("Indigenous","White/Indigenous"))
+marg(model4, "race", type="effects",
+     at_var_interest=c("White","White/Indigenous"))
+marg(model4, "race", type="effects",
+     at_var_interest=c("White","White/Latino"))
+marg(model4, "race", type="effects",
+     at_var_interest=c("Latino","White/Latino"))
+```
+
 I want to compare each multiracial group to:
@@
 Now I can use this nice table to plot out the results. I illustrate what this basic plot looks like below for the case of black/white individuals.
+
+```{r new-plot-attempt}
+means_marginal <- marg(model0, "race", type="levels",
+                       at_var_interest=c("White","Black/White","Black"))[[1]]
+
+coefs <- tibble(term=str_remove(means_marginal$Label, "race = "), 
+                estimate=means_marginal$Margin, 
+                se=means_marginal$Standard.Error) %>%
+  mutate(multiracial=str_detect(term, "/"))
+
+coefs$estimate_mid <- ifelse(coefs$multiracial, 
+                             mean(coefs$estimate[!coefs$multiracial]),
+                             coefs$estimate)
+coefs$se_mid <- ifelse(coefs$multiracial, 
+                       sqrt(sum(coefs$se[!coefs$multiracial]^2)/4),
+                       coefs$se)
+
+coefs$ratio <- coefs$se/coefs$se[2]
+coefs$ratio[2] <- coefs$se[2]/coefs$se_mid[2]
+
+coefs$z <- 1.96*sqrt(1+coefs$ratio^2)/(1+coefs$ratio)
+coefs$upper_self <- ifelse(coefs$multiracial, coefs$estimate_mid+coefs$z*coefs$se_mid,
+                           coefs$estimate+coefs$z*coefs$se)
+coefs$lower_self <- ifelse(coefs$multiracial, coefs$estimate_mid-coefs$z*coefs$se_mid,
+                           coefs$estimate-coefs$z*coefs$se)
+coefs$term <- factor(coefs$term, levels=c("White","Black/White","Black"))
+coefs$colors <- ifelse(coefs$multiracial, "Multiracial", as.character(coefs$term))
+
+se <- coefs %>%
+  filter(!multiracial) %>%
+  select(term, estimate, se) %>%
+  rename(comparison=term, estimate_comparison=estimate, se_comparison=se)
+
+temp <- bind_rows(coefs[2,1:3], coefs[2,1:3]) %>%
+  mutate(comparison=c("Black","White")) %>%
+  left_join(se) %>%
+  mutate(ratio=se/se_comparison,
+         z=1.96*sqrt(1+ratio^2)/(1+ratio),
+         upper=ifelse(estimate<estimate_comparison, estimate+z*se, estimate),
+         lower=ifelse(estimate>estimate_comparison, estimate-z*se, estimate),
+         colors=comparison)
+
+ggplot(coefs, aes(x=term, y=estimate, color=colors))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower_self,
+                ymax=upper_self, fill=colors), color=NA, alpha=0.2)+
+  geom_linerange(data=temp, aes(ymin=lower, ymax=upper, color=colors),
+                  position = position_dodge2(width = 0.1))+
+  geom_point()+
+  coord_flip()+
+  scale_y_continuous(labels = scales::percent)+
+  scale_color_manual(values=c("red","black","blue"))+
+  scale_fill_manual(values=c("red","black","blue"))+
+  labs(x=NULL,
+       y="predicted probability of being behind expected grade")+
+  theme_bw()+
+  theme(legend.position = "none")
+
+```
 

### 2022-09-08 analysis/analysis_main.Rmd
Add example of new graphical approach for model 0, Black/White case

--- 2022-09-05 analysis/analysis_main.Rmd
@@
 load(here("analysis","output","acs.RData"))
+color_choices <- c("#003f5c","#58508d","#bc5090","#ff6361","#000000","#ffa600")
 ```
@@
 ```{r calc-cond-means, warning=FALSE}
-coef_table_mod0 <- calculate_marg_means(model0)
+results_mod0 <- calculate_marg_means(model0)
 ```
@@
 ```{r example-plot, fig.cap="Probability of being behind grade for biracial black/white respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level"}
-g <- coef_table_mod0 %>%
+Non-overlap in color-coded confidence bands roughly indicates statistically significant difference at the 5% level"}
+results_mod0$coef_table %>%
   filter(mrace=="Black/White") %>%
-  ggplot(aes(x=term, y=estimate, color=multiracial,
-             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_mod0, multiracial & mrace=="Black/White"))+
-  geom_point()+
+  ggplot(aes(x=term, y=estimate, color=colors))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower,
+                ymax=upper, fill=colors), color=NA, alpha=0.2)+
+  geom_linerange(data=filter(results_mod0$ci_df, mrace=="Black/White"), 
+                 position=position_dodge2(width = 0.05), size=1.2,
+                 aes(ymin=lower, ymax=upper, color=colors))+
+  geom_point(size=2)+
   coord_flip()+
+  scale_y_continuous(labels = scales::percent)+
+  scale_color_manual(values=color_choices[c(2,5:6)])+
+  scale_fill_manual(values=color_choices[c(2,5:6)])+
+  labs(x=NULL,
+       y="predicted probability of being behind expected grade")+
   theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="predicted probability of being behind expected grade")
-
-g+annotate("text", 1, 0.036, hjust="left",
-             label="Grey bands give 83.4%\nconfidence interval around\nmonoracial estimates")+
-  annotate("text", 2, 0.0445, hjust="left",
-             label="Red band gives 83.4%\nconfidence interval around\nmidpoint estimate")+
-  annotate("text", 2.3, 0.0365, hjust="left",
-             label="Biracial estimate shown\nin red with 83.4% confidence\ninterval")
+  theme(legend.position = "none", panel.grid = element_blank())
 ```

### 2022-09-08 analysis/analysis_main.Rmd
Update explanation of example figure

@@
 
-```{r}
-marg(model4, "race", type="effects",
-     at_var_interest=c("White","Black/White","Black"))
-marg(model4, "race", type="effects",
-     at_var_interest=c("Black","Black/Latino","Latino"))
-marg(model4, "race", type="effects",
-     at_var_interest=c("Asian","White/Asian"))
-marg(model4, "race", type="effects",
-     at_var_interest=c("Indigenous","White/Indigenous"))
-marg(model4, "race", type="effects",
-     at_var_interest=c("White","White/Indigenous"))
-marg(model4, "race", type="effects",
-     at_var_interest=c("White","White/Latino"))
-marg(model4, "race", type="effects",
-     at_var_interest=c("Latino","White/Latino"))
-```
-
 I want to compare each multiracial group to:
@@
 The average adjusted prediction (this is the stata-ese language - it is also called variously a predictive margin among other things).
-The adjusted predition in general is just the predicted value of the dependent variable for some set of values on the covariates.
+The adjusted prediction in general is just the predicted value of the dependent variable for some set of values on the covariates.
 In this case, I want to predict on the level of the response to get predicted probabilities of being held back.
 The average adjusted prediction essentially gets the adjusted prediction for every observation in the data based on their covariates and then takes the average.
-Essentially, this is equivalent to treating every respondent as if they belonged to group A when calculating the average adjusted prediction for group A.
+This is equivalent to treating every respondent as if they belonged to group A when calculating the average adjusted prediction for group A.
 The average adjusted prediction is directly related to the *average marginal effect* (AME) in that the AME for a group difference is simply the difference in the average adjusted prediction across groups.
@@
 
-In order to calculate all the values that I need, I have created a function that will put this into a nice table for later plotting.
+I have created a function that computes these average adjusted predictions, calculates midpoint values and SEs and does some other computaations described below.
+It spits out a list with a `coef_table` showing the average adjusted predictions and a `ci_df` which provides additional information needed for the graphs.
+Basically, this output will allow me to create the plots I need for each biracial comparison.
 
@@
 
-Now I can use this nice table to plot out the results. I illustrate what this basic plot looks like below for the case of black/white individuals.
+Now I can use this output to plot out the results. I illustrate what this basic plot looks like below for the case of black/white individuals.
 
@@
 
-This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions (although the SEs are slightly higher due to survey design effects of clustering and variable weights).
-The grey bands for the monoracial groups bookmark the range of results we might expect and the red band gives what would be the midpoint value between the groups, adjusting for statistical uncertainty.
-I can then see where the actual biracial estimate falls within this range.
-The results here indicate that biracial black/white students have probabilities of grade retention much closer to the lower values of white than black students.
-However, their probability is still a bit higher than that for whites.
+This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions (although the standard errors are higher due to survey design effects of clustering and variable weights).
+The dots give the point estimates for all three groups.
+You can see from the graph that the point estimate of grade retention risk for Black/White students is much closer to the lower for White students than the higher risk for Black students.
 
-Why I am using 83% (technically 83.4%) confidence intervals?
-This is a nice visual trick.
-People often incorrectly believe that overlapping 95% confidence intervals on two means indicates that the difference in the means is not statistically significant at $p<.05$.
-This is not always true and leads in practice to a much lower $\alpha$ for rejection.
-A much better approximation is given when one uses1.386 rather than 1.96 in estimating the confidence interval, which produces an 83.4% confidence interval.
-The intuition behind this technique is given nicely [here](https://chris-said.io/2014/12/01/independent-t-tests-and-the-83-confidence-interval-a-useful-trick-for-eyeballing-your-data/).
-Because I use this interval, we can know that if intervals (or bands) do not overlap, then the difference in means between the two groups is statistically significant at $p<.05$.
-In this case, the biracial estimate is statistically significantly different from both monoracial estimates as well as the "halfway" estimate.
-In short, biracial black/white students are held back at rates much closer but somewhat higher than whites, rather than blacks.
+The figure also provides confidence bands that help to visualize whether the difference between groups are statistically significant.
+People often believe that an overlap in the 95% confidence interval means the difference between the parameters is not statistically significant at the 5% level.
+This is actually a mistake and leads to far more conservative assessments.
+The actual p-value of rejection corresponding to an overlap of 95% CI is about 0.0055 when the SEs of the two estimates are the same as shown in [this paper](https://link.springer.com/article/10.1007/s10654-011-9563-8).
 
-Its important to note that this technique is just an approximation because it depends on the assumption that the standard errors for the two means are the same, which is not going to be true here.
-However, it should give results that are quite close.
-Correcting for unequal SEs is far more complicated because the interval used will vary for any pairwise comparison which I am not doing here.
+If the SEs of the two estimates are the same (and they are independent), then there is a simple correction to this problem: use 83.4% confidence interval (z=1.386).
+In that case, lack of overlap in CIs indicates that the differences between estimates are statistically significant at the 5% level.
+Overlap indicates that they are not statistically significant.
+
+While this provides a nice and easy solution, it unfortunately only works in the case where the SEs are the same.
+Its a pretty good approximation when those SEs are close, but in my case, I get very different SEs due to massive differences in sample size across racial groups.
+To calculate the z-score for the confidence interval needed in the case of unequal SEs (and uncorrelated), first one needs to calculate $\rho=se_1/se_2$ which gives the ratio of the two SEs.
+It doesn't matter which serves as the numerator and denominator as the results are symmetric.
+The required z-score is then given by:
+
+$$z=1.96*\frac{\sqrt{1+\rho^2}}{1+\rho}$$
+
+The problem is that this z-score will be different for each pairwise comparison, and I want to make at least two comparisons in the figures - I want to compare the biracial group to each of its constituent monoracial groups.
+So, the way I handle this in the figure above is to calculate two confidence intervals for each biracial group.
+Each monoracial group gets a confidence band corresponding to that comparison and the biracial group gets two half-confidence intervals.All of these are color coded for the correct comparison and in the biracial case, they are only shown in the relevant direction.
+So in the case, above, the purple line doesn't come close to overlapping with the purple band for Black students and so this difference is clearly statistically distinguishable.
+The yellow line also does not overlap with the yellow band for whites, so this difference is also statistically distinguishable, although as you can see from the graph, that is just barely the case.
+
+To check the graph, I can calculate AMEs for the Black/White group from the same model where I treat White and Black students, respectively, as the reference.
+
+```{r test-marg}
+marg(model0, "race", type="effects",
+     at_var_interest=c("White","Black/White"))
+marg(model0, "race", type="effects",
+     at_var_interest=c("Black","Black/White"))
+```
+
+As you can see, the results are consistent here.
+
+The graph above also shows a grey band for the halfway point.
+To avoid getting cluttery, I do not put a separate confidence line on the Black/White estimate to differentiate this case.
+In general that line will be about halway between the other two, so I think its ok to just eyeball that one.
+If it annoys reviewer, it can be added.
 
@@
 ```{r cond-means-model4, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
+Non-overlap in confidence bands of corresponding color indicates statistically significant difference at the 5% level.
 Estimates are based on models that adjust for state, year, current grade, metropolitan status, gender, foreign born status, parent's foreign born status, English proficieny, parent's English proficiency, parent's education, family income, home ownership, and family structure."}
-coef_table_mod4 <- calculate_marg_means(model4)
+results_mod4 <- calculate_marg_means(model4)
 
-ggplot(coef_table_mod4, aes(x=term, y=estimate, color=multiracial,
-                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_mod4, multiracial))+
-  geom_point()+
+results_mod0$coef_table %>%
+  ggplot(aes(x=term, y=estimate, color=colors))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower,
+                ymax=upper, fill=colors), color=NA, alpha=0.2)+
+  geom_linerange(data=results_mod0$ci_df, 
+                 position=position_dodge2(width = 0.05), size=1.2,
+                 aes(ymin=lower, ymax=upper, color=colors))+
+  geom_point(size=2)+
   coord_flip()+
   facet_wrap(~mrace, ncol=3, scales="free_y")+
+  scale_y_continuous(labels = scales::percent)+
+  scale_color_manual(values=color_choices)+
+  scale_fill_manual(values=color_choices)+
+  labs(x=NULL,
+       y="average adjusted predicted probability of being behind expected grade")+
   theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
+  theme(legend.position = "none", panel.grid = element_blank())
 ```
@@
 ```{r cond-means-model-compare, fig.height=12, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
+Non-overlap in confidence bands of corresponding color indicates statistically significant difference at the 5% level.
 Each panel is based on a different model that accounts for certain variables.
@@
 
-* Lower than both groups
-* Same as lower group
-* Lower half
-* Halfway between
-* Upper half
-* Same as higher
-* Higher than both groups
+-   Lower than both groups
+-   Same as lower group
+-   Lower half
+-   Halfway between
+-   Upper half
+-   Same as higher
+-   Higher than both groups
 
@@
 |                   | Lower than both | Same as lower | Lower half | Halfway | Upper half | Same as higher | Higher than both |
-|-------------------|:---------------:|:-------------:|:----------:|:-------:|:----------:|:--------------:|:----------------:|
-| Black/White       |        X*       |               |            |         |            |                |                  |
-| White/Asian       |        X*       |       X       |            |         |            |                |                  |
-| White/Latino      |        X        |       X*      |            |         |            |                |                  |
-| Black/Latino      |        X        |       X*      |      X     |    X    |            |                |                  |
-| Indigenous/Asian  |        X        |       X*      |      X     |         |            |                |                  |
-| Indigenous/White  |                 |               |      X*    |         |            |                |                  |
-| Indigenous/Latino |                 |               |      X     |    X*   |      X     |                |                  |
-| Black Asian       |                 |       X       |      X     |    X*   |      X     |       X        |                  |
-| Black/Indigenous  |                 |       X       |      X     |    X*   |      X     |       X        |                  |
-| Latino/Asian      |                 |       X       |      X     |    X    |      X     |       X        |        x*        |
+|---------|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|
+| Black/White       |       X\*       |               |            |         |            |                |                  |
+| White/Asian       |       X\*       |       X       |            |         |            |                |                  |
+| White/Latino      |        X        |      X\*      |            |         |            |                |                  |
+| Black/Latino      |        X        |      X\*      |     X      |    X    |            |                |                  |
+| Indigenous/Asian  |        X        |      X\*      |     X      |         |            |                |                  |
+| Indigenous/White  |                 |               |    X\*     |         |            |                |                  |
+| Indigenous/Latino |                 |               |     X      |   X\*   |     X      |                |                  |
+| Black Asian       |                 |       X       |     X      |   X\*   |     X      |       X        |                  |
+| Black/Indigenous  |                 |       X       |     X      |   X\*   |     X      |       X        |                  |
+| Latino/Asian      |                 |       X       |     X      |    X    |     X      |       X        |       x\*        |
 
@@
 
-* For White/Black students, grade retention is clearly in the lower-half between Whites and the midway point rather than below whites for high school students.
+-   For White/Black students, grade retention is clearly in the lower-half between Whites and the midway point rather than below whites for high school students.
 Notably, this is also where the white/black grows the largest.
@@
 Nonetheless, their outcomes are still much closer to White students than Black students.
-* For White/Indigenous students, the placement varies across panels.
+-   For White/Indigenous students, the placement varies across panels.
 In elementary school, these students are indistinguishable from White students.

### 2022-09-08 analysis/analysis_main.Rmd
Correct all models from main analysis, but not sensitivity tests yet

@@
   geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower,
-                ymax=upper, fill=colors), color=NA, alpha=0.2)+
+                ymax=upper, fill=colors), color=NA, alpha=0.5)+
   geom_linerange(data=filter(results_mod0$ci_df, mrace=="Black/White"), 
-                 position=position_dodge2(width = 0.05), size=1.2,
+                 position=position_dodge2(width = 0.05), size=1.2, alpha=0.5,
                  aes(ymin=lower, ymax=upper, color=colors))+
@@
 
-results_mod0$coef_table %>%
+results_mod4$coef_table %>%
   ggplot(aes(x=term, y=estimate, color=colors))+
   geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower,
-                ymax=upper, fill=colors), color=NA, alpha=0.2)+
-  geom_linerange(data=results_mod0$ci_df, 
-                 position=position_dodge2(width = 0.05), size=1.2,
+                ymax=upper, fill=colors), color=NA, alpha=0.5)+
+  geom_linerange(data=results_mod4$ci_df, 
+                 position=position_dodge2(width = 0.25), size=1, alpha=0.5,
                  aes(ymin=lower, ymax=upper, color=colors))+
@@
 As the table here shows, after controlling for other stuff, the majority of multiracial groups are quite similar to the monoracial constituent group with the lower grade retention rate.
-The only case where the biracial group is more like the higher monoracial group is for Latino/Asians.
-The Indigenous/Latino group is the only group statistically disginguishable as midway between the two constituent groups.
-The White/Indigenous group is also statistically distinguishable from both constitutent groups but much closer to Whites in their probability.
-The Black/Asian and Black/Indigenous groups also give point estimates at the midpoint, but the results are very noisy due to large SEs and so conclusions here would have to be speculative.
+The only case where the biracial group is more like the higher monoracial group is for Latino/Asians, but the point estimate here is not statistically distinguishable from either monoracial group and both monoracial groups have similar risks after controlling for resources.
+The Indigenous/Latino, Black/Indigenous, and Black/Asian groups all have point estimates roughly at the halfway pont, but the results are also somewhat inconclusive because of the high SE in all three cases.
+In the case of the Black/Asian and Black/Indigeious gruops, we cannot rule out that they are higher or lower than either monoracial group.
+In the Indigenous/Latin group, we cannot rule out that they are like the lower group (Latinos) and the higher one is very, very close.
+So, basically, in all of the cases where we can make a solid determination, the determination is like the lower group.
 
@@
 #calculate remaining marginal means
-coef_table_mod1 <- calculate_marg_means(model1)
-coef_table_mod2 <- calculate_marg_means(model2)
+results_mod1 <- calculate_marg_means(model1)
+results_mod2 <- calculate_marg_means(model2)
 
-coef_table <- coef_table_mod0 %>%
+coef_table <- results_mod0$coef_table %>%
   mutate(model="crude")
-coef_table <- coef_table_mod1 %>%
+coef_table <- results_mod1$coef_table %>%
   mutate(model="fixed effects") %>%
   bind_rows(coef_table)
-coef_table <- coef_table_mod2 %>%
+coef_table <- results_mod2$coef_table %>%
   mutate(model="cultural resources") %>%
   bind_rows(coef_table)
-coef_table <- coef_table_mod4 %>%
+coef_table <- results_mod4$coef_table %>%
   mutate(model="parent resources") %>%
@@
 
-ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
-                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial))+
+ci_df <- results_mod0$ci_df %>%
+  mutate(model="crude")
+ci_df <- results_mod1$ci_df %>%
+  mutate(model="fixed effects") %>%
+  bind_rows(ci_df)
+ci_df <- results_mod2$ci_df %>%
+  mutate(model="cultural resources") %>%
+  bind_rows(ci_df)
+ci_df <- results_mod4$ci_df %>%
+  mutate(model="parent resources") %>%
+  bind_rows(ci_df) %>%
+  mutate(model=factor(model, levels=c("crude","fixed effects",
+                                      "cultural resources",
+                                      "parent resources")),
+         mrace=factor(mrace, levels=c("Black/White",
+                                      "Black/Latino",
+                                      "Black/Asian",
+                                      "Black/Indigenous",
+                                      "White/Indigenous",
+                                      "White/Latino",
+                                      "White/Asian",
+                                      "Latino/Asian",
+                                      "Indigenous/Latino",
+                                      "Indigenous/Asian")))
+
+ggplot(coef_table, aes(x=term, y=estimate, color=colors))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower, ymax=upper, fill=colors), 
+            color=NA, alpha=0.5)+
+  geom_linerange(data=ci_df, 
+                 position=position_dodge2(width = 0.1), size=1, alpha=0.5,
+                 aes(ymin=lower, ymax=upper, color=colors))+
   geom_point()+
@@
   theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
+  scale_color_manual(values=color_choices)+
+  scale_fill_manual(values=color_choices)+
   scale_y_continuous(labels = scales::percent)+
@@
 First, the gap between Blacks and Whites shrinks considerably (visually it looks like its cut about in half).
-Second, Black/White biracials actually have lower rates than either group.
-They aren't that different than whites in effect size but the difference is statistically significant.
+Second, Black/White biracials actually have lower rates than either group, although the difference from White students is not statistically significant.
 
@@
 Interestingly, the fixed effects model reverses the ordering of Blacks and Latinos, but this is then reversed again in the cultural resources model.
-In all of these models, the Black/Latino group remains below either of those two groups, although the gap shrinks considerably across models.
+In all of these models, the Black/Latino group remains below either of those two groups, although the gap shrinks considerably across models and is not statistically significantly different from Latinos in the cultural resources model.
 
@@
 The crude and fixed effects models indicate that Black/Asian students have grade retention rates similar to the lower grade retention rates of Asian students, although the high SE makes the precision of this effect quite weak.
-Controlling for cultural esources moves this group closer to the midpoint value.
+Controlling for cultural resources moves this group closer to the midpoint value.
 This makes sense because Black/Asian students are much more likely than Asian students to be native born and speak English well with native-born parents who also speak English well.
@@
 This group starts fairly close to the midway point, but controlling for resources clearly moves them into the "lower half" space.
-Their results never look as good as whites, but look substantially better than Indigenous students, who continue to have the highest rate of grade retention.
+Their results never look as good as whites (although the bars are so close to overlapping in the full model), but look substantially better than Indigenous students, who continue to have the highest rate of grade retention.
 Its interesting that the resource models move them closer to white given that they still have resources that are quite a bit worse than Whites.
@@
 In all of the models, White/Asian students have grade retention rates lower than the rates of Asians - the monoracial group with the lower rates.
-Controls for family resources reduce this gap a bit because Asian/White students are the most privileged students overall and the confidence interval with Asian students overlaps slightly.
+Controls for family resources reduce this gap to the point where the difference is not statistically distinguishable from Asian students.
 So, we can determine that Asian/White students have grade retention rates at least as low as the lowest consistuent group (Asians) and possibly even lower.
@@
 |                   | Lower than both | Same as lower | Lower half | Halfway | Upper half | Same as higher | Higher than both |
-|---------|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|
-| Black/White       |       X\*       |               |            |         |            |                |                  |
+|-------------------|:---------------:|:-------------:|:----------:|:-------:|:----------:|:--------------:|:----------------:|
+| Black/White       |       X\*       |       X       |            |         |            |                |                  |
 | White/Asian       |       X\*       |       X       |            |         |            |                |                  |
@@
 | Black/Latino      |        X        |      X\*      |     X      |    X    |            |                |                  |
-| Indigenous/Asian  |        X        |      X\*      |     X      |         |            |                |                  |
-| Indigenous/White  |                 |               |    X\*     |         |            |                |                  |
-| Indigenous/Latino |                 |               |     X      |   X\*   |     X      |                |                  |
+| Indigenous/Asian  |        X\*      |       X       |     X      |    X    |            |                |                  |
+| Indigenous/White  |                 |       X       |    X\*     |         |            |                |                  |
+| Indigenous/Latino |                 |       X       |     X      |   X\*   |     X      |                |                  |
 | Black Asian       |                 |       X       |     X      |   X\*   |     X      |       X        |                  |

### 2022-09-09 products/paper/main.qmd
Update figure code for paper

@@
 source(here("analysis","useful_functions.R"))
+color_choices <- c("#003f5c","#58508d","#bc5090","#ff6361","#000000","#ffa600")
 ```
@@
 coef_table <- coef_table %>%
+  mutate(model=factor(model,
+                      levels=c("crude","fixed effects","cultural resources",
+                               "parent resources"),
+                      labels=c("Raw","Baseline model","Cultural model",
+                               "Full model")))
+ci_df <- ci_df %>%
   mutate(model=factor(model,
@@
 Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
-g <-  coef_table %>%
+
+coef_table %>%
   filter(mrace=="Black/White" & model=="Raw") %>%
-  ggplot(aes(x=term, y=estimate, color=multiracial,
-             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial & 
-                               mrace=="Black/White" &
-                               model=="Raw"))+
-  geom_point()+
+  ggplot(aes(x=term, y=estimate, color=colors))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower,
+                ymax=upper, fill=colors), color=NA, alpha=0.5)+
+  geom_linerange(data=filter(ci_df, mrace=="Black/White" & model=="Raw"), 
+                 position=position_dodge2(width = 0.05), size=1.2, alpha=0.5,
+                 aes(ymin=lower, ymax=upper, color=colors))+
+  geom_point(size=2)+
   coord_flip()+
+  scale_y_continuous(labels = scales::percent)+
+  scale_color_manual(values=color_choices[c(2,5:6)])+
+  scale_fill_manual(values=color_choices[c(2,5:6)])+
+  labs(x=NULL,
+       y="predicted probability of being behind expected grade")+
   theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="predicted probability of being behind expected grade")
-
-g+annotate("text", 1, 0.036, hjust="left",
-             label="Grey bands indicate 83.4%\nconfidence interval around\nmonoracial estimates")+
-  annotate("text", 2, 0.0445, hjust="left",
-             label="Red band indicate 83.4%\nconfidence interval around\nmidpoint estimate")+
-  annotate("text", 2.5, 0.0365, hjust="left",
-             label="Actual biracial estimate\nshown with 83.4%\nconfidence interval")
+  theme(legend.position = "none", panel.grid = element_blank())
 ```
@@
   filter(str_detect(mrace, "Black")) %>%
-  ggplot(aes(x=term, y=estimate, color=multiracial,
-             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial 
-                             & (model=="Baseline model" | 
-                                  model=="Full model") & 
-                               str_detect(mrace, "Black")))+
+  ggplot(aes(x=term, y=estimate, color=colors))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower, ymax=upper, fill=colors), 
+            color=NA, alpha=0.5)+
+  geom_linerange(data=subset(ci_df, 
+                             (model=="Baseline model" |  model=="Full model") &
+                               str_detect(mrace, "Black")), 
+                 position=position_dodge2(width = 0.1), size=1, alpha=0.5,
+                 aes(ymin=lower, ymax=upper, color=colors))+
   geom_point()+
@@
   filter(!str_detect(mrace, "Black")) %>%
-  ggplot(aes(x=term, y=estimate, color=multiracial,
-             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial 
-                             & (model=="Baseline model" | 
-                                  model=="Full model") & 
-                               !str_detect(mrace, "Black")))+
+  ggplot(aes(x=term, y=estimate, color=colors))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower, ymax=upper, fill=colors), 
+            color=NA, alpha=0.5)+
+  geom_linerange(data=subset(ci_df, 
+                             (model=="Baseline model" | model=="Full model") &
+                               !str_detect(mrace, "Black")), 
+                 position=position_dodge2(width = 0.1), size=1, alpha=0.5,
+                 aes(ymin=lower, ymax=upper, color=colors))+
   geom_point()+
@@
   theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
+  scale_color_manual(values=color_choices[-1])+
+  scale_fill_manual(values=color_choices[-1])+
   scale_y_continuous(labels = scales::percent)+

### 2022-09-09 products/paper/main.qmd
Edit explanation of example figure for new design

@@
 @tbl-descriptive shows the percent of each monoracial and biracial group that is clearly behind expected grade.
-My goal is to understand the degree of grade retention for each biracial group relative to its constituent monoracial groups (e.g.
+My goal is to understand the risk of being behind expected grade for each biracial group relative to its constituent monoracial groups (e.g.
 a Black/White student compared to White and Black students).
@@
 
-As an example, I display the percent of students clearly behind grade for White, Black, and Black/White students in @fig-example.
-The point estimate and confidence interval for the Black/White group are shown in red, while the two monoracial groups are shown via grey dots and confidence bands.
-For all three cases, I use 83.4% confidence intervals, because overlap on such intervals indicates that the observed difference between the two measures in the sample is not statistically distinguishable from zero in the population at $p<.05$ [@knol_mis_2011].[^2] In this case, the lack of overlap between any of the confidence bands indicates that the Black/White probability of being clearly behind expected grade is statistically distinguishable from both monoracial probabilities.
+As an example, I display the percent of students clearly behind expected grade for White, Black, and Black/White students in @fig-example.
+I display confidence bands/bars around all three point estimates, but these confidence ranges require some explanation as they are not standard 95% confidence intervals.
+My goal is to determine whether the biracial group's point estimate is statistically distinguishable from a given monoracial constituent group's point estimate by non-overlap of confidence intervals.
+Contrary to common belief, overlap in 95% confidence intervals is not equivalent to failing a hypothesis test of difference at $p<0.05$ and leads to far more stringent tests [@knol_mis_2011].
+When standard errors between groups are equal, overlap in 83.4% confidence intervals will achieve this goal.
+However, in cases where standard errors vary substantially across estimates, the calculation of the appropriate confidence interval is more complex and varies by each pairwise comparison.[^2]
 
-[^2]: The correspondence to $p<.05$ is only approximate due to differences in the standard errors between groups.
-Calculating more precise intervals requires that those intervals vary across each pairwise comparison which does not work for the visualizations here.
-The approximation will be very close in all cases used here.
+[^2]: The formula for calculating the z-score for the correct interval is given by $$1.96 * \frac{\sqrt{1+\rho^2}}{1+\rho}$$ where $\rho$ is the ratio of the standard errors for the two statistics.
+See @knol_mis_2011 for detailed derivation of this formula.
 
@@
 #| label: fig-example
-#| fig-cap: Probability of being clearly below expected grade for biracial black/white respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
+#| fig-cap: Probability of being clearly behind expected grade for biracial black/white respondents in comparison to their monoracial comparison groups.
+Non-overlap in color corresponding confidence bands indicates statistically significant difference at the 5% level.
 
@@
 
-Additionally, I include a measure of the "halfway" point between the two racial groups.[^3] This halfway point allows me to determine where the biracial group falls relative to the expectation of being halfway between the two constituent monoracial groups.
+I calculate two confidence intervals for each biracial group.
+Each confidence interval is in comparison to one of the constituent monoracial groups.
+These confidence intervals are color-coded in @fig-example to indicate the reference group.
+I also only draw half-intervals in the direction of the monoracial group's point estimate.
+Overlap in these color corresponding bars and bands indicates that the difference between the biracial group and the monoracial group is not statistically significant at $p<.05$.
+For example, the yellow bar shown in @fig_example for biracial Black/White students does not quite overlap with the yellow band for White students, indicating that these point estimates for these two groups are statistically distinguishable at $p<.05$.
+<!-- TODO provide actual p-value for comparison -->
+
+Additionally, I include a measure of the "halfway" point between the two monoracial groups in dark grey.[^3] This halfway point allows me to determine where the biracial group falls relative to the expectation of being halfway between the two constituent monoracial groups.
 In this case, Black/White students are much closer to White students in their risk of grade retention, and have probabilities much lower than both Black students and the halfway expectation.
@@
 Its standard error is given by $\sqrt{(s_1^2+s_2^2)/2}$ where $s_1$ and $s_2$ are the standard errors of the estimates for the two monoracial groups.
+To aboid clutter, I do not draw a third half-confidence interval on the biracial estimate itself, but this estimate will always be roughly about halfway between the length of the other two intervals.
 
@@
 For the Black/White case, on the other hand, we would expect that controlling for resource differentials should eliminate some of their disadvantage relative to Whites.
-For the remaining “in-between” groups, it is more difficult to make a priori predictions.
+For the remaining "in-between" groups, it is more difficult to make a priori predictions.
 

### 2022-09-09 products/paper/main.qmd
Update captions in paper

@@
 #| fig-cap: Probability of being behind grade for biracial respondents with one black parent, in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level, Baseline model includes year, grade, location, and state fixed effects.
+.
+Non-overlap in color corresponding confidence bands indicates statistically significant difference at the 5% level, Baseline model includes year, grade, location, and state fixed effects.
 Full models include control variables for nativity, English proficiency, income, education, home ownership, and marital status of parents.
@@
 #| fig-cap: Probability of being behind grade for non-black biracial respondents, in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level, Baseline model includes year, grade, location, and state fixed effects.
+Non-overlap in color corresponding confidence bands indicates statistically significant difference at the 5% level, Baseline model includes year, grade, location, and state fixed effects.
 Full models include control variables for nativity, English proficiency, income, education, home ownership, and marital status of parents.

### 2022-09-09 products/paper/main.qmd
Update text on results for new figures

@@
 In the case of Black/White biracial respondents, Black/White students initially have a risk of BBEG slightly higher than and statistically distinguishable from White students.
-After controlling for cultural and family resources, however, Black/White respondents have a risk of BBEG that is slightly lower and statistically distinguishable from White students.
+After controlling for cultural and family resources, however, Black/White respondents have a risk of BBEG that is slightly lower but not statistically distinguishable from White students.
 This change reflects the relatively low level of parental education and income among Black/White students, which is much closer to Black students than White students, as shown on @tbl-descriptive.
 
-An opposite pattern is observed for Black/Latino respondents.
-Black/Latino respondents initially have a risk of grade retention substantially lower than both Black and Latino students, who have similar risks.
-However, after controlling for cultural and family resources, Black/Latino students have a risk of BBEG that is indistinguishable from the risk of Latino students and substantially lower than Black students.
+Black/Latino students start with a very different position, having a risk of BBEG substantially lower than both Black and Latino students, who have sonewgat similar risks.
+However, after controlling for cultural and family resources, Black/Latino students have a risk of BBEG that is statistically indistinguishable from the risk of Latino students and substantially lower than and statistically distinguishable from Black students.
 The change largely reflects the fact that Black/Latino students have greater resources than Latino (but not Black) students and once this advantage is held constant, they have a similar risk.
+After controlling for family resources, both Black/White and Black/Latino students have a risk of BBEG similar to their monoracial constituent group with lower risk.
 
@@
 Controlling for resources moves the point estimate for Black/Asian students from a point estimate close to Asian students to more of a halfway position between the two constituent groups.
-Because Black/Indigenous students had resources halfway between their constituent groups, controlling for these resources hardly affects the relative placement of Black/Indigenous students at all.
+Black/Indigenous students have BBEG risk roughly halfway between their two constituent groups in both models.
+However, this finding of a halfway position for both groups is highly tentative owing to the wide confidence intervals around both groups, due to small sample size.
 
 @fig-other-models shows the results for the remaining biracial groups.
-In general, the results here show that biracial students are more likely to have a risk of BBEG more like the monoracial constituent group with the lower risk than the higher risk, with some important caveats discussed below.
+In general, the results here show that biracial students are more likely to have a risk of BBEG closer to the monoracial constituent group with the lower risk than the higher risk, with some important caveats discussed below.
 
@@
 There are two exceptions to this general pattern.
-First, the point estimate of the probability of BBEG for the Indigenous/Latino group is roughly halfway between the two constituent groups and statistically distinguishable from both.
+First, the point estimate of the probability of BBEG for the Indigenous/Latino group is roughly halfway between the two constituent groups, although it is not statistically distinguishable from the lower group.
 Second, Latino/Asian students are the only case where the point estimate of the risk for the biracial group is higher than either monoracial group.
-However, the confidence bands suggest that the risk for this group cannot be statistically differentiated from either of the two monoracial constituent groups.
-This result only emerges in the model that controls for resources, Latino/Asian students come from more advantaged households than both Latino and Asian students, on average.
+However, the confidence bands suggest that the risk for this group cannot be statistically differentiated from either of the two monoracial constituent groups who have very similar risks.
+This result only emerges in the model that controls for resources, because Latino/Asian students come from more advantaged households than both Latino and Asian students, on average.
 
@@
 White/Asian students are the only case with a substantial shift, owing to their highly advantaged households.
-After controlling for resources, the risks of BBEG for White/Asian students are indistinguishable from the risk of Asian students, but still substantially and statistically distinguishable from the higher risk to White students.
+Initially white/Asian students have risks of BBEG substantially below both White and Asian students.
+After controlling for resources, the risks of BBEG for White/Asian students are indistinguishable from the those for Asian students, but still substantially and statistically distinguishable from the higher risk to White students.
 

### 2022-09-09 analysis/analysis_main.Rmd
Add ci_df to coef_table save output

@@
 
-save(coef_table, file=here("analysis","output","coef_table.RData"))
+save(coef_table, ci_df, file=here("analysis","output","coef_table.RData"))
 ```
@@
 
-coef_table_elem <- calculate_marg_means(model4_elem)
-coef_table_middle <- calculate_marg_means(model4_middle)
-coef_table_hs <- calculate_marg_means(model4_hs)
+results_elem <- calculate_marg_means(model4_elem)
+results_middle <- calculate_marg_means(model4_middle)
+results_hs <- calculate_marg_means(model4_hs)
 
-coef_table <- coef_table_elem %>%
+coef_table <- results_elem$coef_table %>%
   mutate(model="elementary")
-coef_table <- coef_table_middle %>%
+coef_table <- results_middle$coef_table %>%
   mutate(model="middle school") %>%
   bind_rows(coef_table)
-coef_table <- coef_table_hs %>%
+coef_table <- results_hs$coef_table %>%
   mutate(model="high school") %>%
@@
 
-ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
-                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial))+
+ci_df <- results_elem$ci_df %>%
+  mutate(model="elementary")
+ci_df <- results_middle$ci_df %>%
+  mutate(model="middle school") %>%
+  bind_rows(ci_df)
+ci_df <- results_hs$ci_df %>%
+  mutate(model="high school") %>%
+  bind_rows(ci_df) %>%
+  mutate(model=factor(model, levels=c("elementary","middle school",
+                                      "high school")),
+         mrace=factor(mrace, levels=c("Black/White",
+                                      "Black/Latino",
+                                      "Black/Asian",
+                                      "Black/Indigenous",
+                                      "White/Indigenous",
+                                      "White/Latino",
+                                      "White/Asian",
+                                      "Latino/Asian",
+                                      "Indigenous/Latino",
+                                      "Indigenous/Asian")))
+
+ggplot(coef_table, aes(x=term, y=estimate, color=colors))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower, ymax=upper, fill=colors), 
+            color=NA, alpha=0.5)+
+  geom_linerange(data=ci_df, 
+                 position=position_dodge2(width = 0.1), size=1, alpha=0.5,
+                 aes(ymin=lower, ymax=upper, color=colors))+
   geom_point()+
@@
 
-### Models with stricter parent linking
-
-```{r models-parent-link, warning=FALSE, fig.height=12, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups.
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.
-Panels compare results depending on whether results are restricted only to children linked to a head of household or not."}
-#run separate models by grade groups
-model4_best_link <- update(model4, design=subset(acs_design, momrule==11 & poprule==11))
-
-coef_table_best_link <- calculate_marg_means(model4_best_link)
-
-coef_table <- coef_table_best_link %>%
-  mutate(model="only direct links")
-coef_table <- coef_table_mod4 %>%
-  mutate(model="all links") %>%
-  bind_rows(coef_table) %>%
-  mutate(model=factor(model, levels=c("all links","only direct links")),
-         mrace=factor(mrace, levels=c("Black/White",
-                                      "Black/Latino",
-                                      "Black/Asian",
-                                      "Black/Indigenous",
-                                      "White/Indigenous",
-                                      "White/Latino",
-                                      "White/Asian",
-                                      "Latino/Asian",
-                                      "Indigenous/Latino",
-                                      "Indigenous/Asian")))
-
-ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
-                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table, multiracial))+
-  geom_point()+
-  coord_flip()+
-  facet_grid(mrace~model, scales="free_y")+
-  theme_bw()+
-  theme(legend.position = "none", panel.grid = element_blank())+
-  scale_color_manual(values=c("grey40","red"))+
-  scale_fill_manual(values=c("grey40","red"))+
-  scale_linetype_manual(values=c(3,1))+
-  scale_y_continuous(labels = scales::percent)+
-  labs(x=NULL,
-       y="average adjusted predicted probability of being behind expected grade")
-```
-
-The results here are extraordinarily stable across both specifications.
-
 ```{r time-end}

### 2022-09-09 analysis/analysis_resources.Rmd

@@
 }) %>%
-  bind_rows()
+  bind_rows() %>%
+    mutate(multiracial=str_detect(term, "/"))
 
-coef_table_cf <- create_coef_table(coefs_cf)
+results_cf <- create_coef_table(coefs_cf)
 ```
@@
 
-```{r counterfactual-ses, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were resources (household income, parental education, home ownership, parent and child foreign-born status, parent and child English proficiency, and whether parents are married).
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
-ggplot(coef_table_cf, aes(x=term, y=estimate, color=multiracial,
-                          ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_cf, multiracial))+
+```{r counterfactual-ses, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were resources (household income, parental education, home ownership, parent and child foreign-born status, parent and child English proficiency, and whether parents are married)"}
+ggplot(results_cf$coef_table, aes(x=term, y=estimate, color=multiracial))+
+  geom_hline(aes(yintercept=estimate), color="grey")+
   geom_point()+
@@
 
-```{r cf-ses-isolate, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were SES resources (household income, parental education, home ownership, and whether parents are married).
-83.4% confidence bands shown around each estimate.
-Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
+```{r cf-ses-isolate, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were SES resources (household income, parental education, home ownership, and whether parents are married)."}
 cf_df_ses <- cf_df
@@
 }) %>%
-  bind_rows()
+  bind_rows() %>%
+    mutate(multiracial=str_detect(term, "/"))
 
-coef_table_cf_ses <- create_coef_table(coefs_cf_ses)
+results_ses <- create_coef_table(coefs_cf_ses)
 
-ggplot(coef_table_cf_ses, aes(x=term, y=estimate, color=multiracial,
-                          ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_cf_ses, multiracial))+
+ggplot(results_ses$coef_table, aes(x=term, y=estimate, color=multiracial))+
+  geom_hline(aes(yintercept=estimate), color="grey")+
   geom_point()+
@@
 }) %>%
-  bind_rows()
+  bind_rows() %>%
+    mutate(multiracial=str_detect(term, "/"))
 
-coef_table_cf_culture <- create_coef_table(coefs_cf_culture)
+results_culture <- create_coef_table(coefs_cf_culture)
 
-ggplot(coef_table_cf_culture, aes(x=term, y=estimate, color=multiracial,
-                          ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
-  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
-                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
-            color=NA, alpha=0.2)+
-  geom_linerange(data=subset(coef_table_cf_culture, multiracial))+
+ggplot(results_culture$coef_table, aes(x=term, y=estimate, color=multiracial))+
+  geom_hline(aes(yintercept=estimate), color="grey")+
   geom_point()+

### 2022-11-02 products/paper/main.qmd
Add rough draft of section on grade retention

@@
 
--   Define what grade retention is and the era of "social promotion." Recent work suggests a decline in grade retention. Can cite Warren and Saliba on the trends.
--   Grade retention is consequential for later educational outcomes.
-CITE the evidence on this although, Warren and Saliba are also very iff-y on this.
-The Andrew article on "scarring" effects of grade retention is also a good one here as is the Hauser et.
-al.
-2001 piece.
-Furthermore, little evidence on the positive effects of grade retention (again cite Hauser 2001).
-Grade retention also varies by race and socioeconomic background and is thus can be seen as a mechanism for the reproduction of class and racial inequality.
-In fact, because grade retention can happen quite early in the K-12 system, for many children, it may be the first direct encounter with a "sorting" mechanism for inequality.
--   There are many challenges to measuring grade retention in existing data sources.
-One of the most common techniques is to flag students who are behind the expected grade for their age.
-This technique is imperfect because
-    -   it captures a cumulative prevalence measure of grade retention rather than incidence.
-    -   it cannot distinguish cases in which children were delayed in beginning school ("redshirted") from grade retention.
-    -   Students can reasonable be in two modal grades based on a given calendar year of age.
-Without detailed information on birthdates and survey timing, overall grade retention will be underestimated.
-For example, a child who is reported as 9 years old and in third grade could either be in the expected grade or could have been held back, depending on their birthday.
-A child who is 10 years old, however, is clearly behind expected grade.
--   Unfortunately, the only measure of grade retention available in the American Community Survey is based on students being clearly behind expected grade.
-The CPS has some advantages for better measuring grade retention, but lacks the sample size to precisely estimate parameters for biracial populations.
--   While being clearly behind expected grade is a problematic measure of grade retention for researchers seeking to better understand the process of grade retention itself, that is not my goal here.
-Rather, I am using this measure as an early and accessible measure of racial stratification for young children and adolescents.
-I can thus use it as a window to understand biracial students' placement within a racial hierarchy.
-This approach is similar to that employed by Rosenfeld (CITE) to examine the differential experiences of children from straight and LGBT families.
-From this perspective, the major potential problem is that the overall measure underestimates grade retention.
-However, because this underestimation is largely a function of birthdate, it is unlikely to bias the results presented here.
--   There is a question of how much racial inequality is present in grade retention rates.
-Hauser et.
-al.
-2001 suggest that while Black and Latino students are more likely to have been held back, those differences are largely explained by differences in STUFF.
-However, other more recent research finds at least moderate racial differences, particularly between Black an White students, even after controlling for a variety of other variables (cite Warren and Sariba, Rosenfeld, others?).
-Therefore, understanding the effect of socioeconomic and other resources on the differences between monoracial groups is also key to understanding the larger research question, just as it is for biracial students.
+Grade retention is the practice of having students repeat a grade due to poor academic performance.
+The frequency of grade retention has fluctuated historically in response to changes in educational policy.
+Rates of grade retention increased throughout the late 20th century [@hauser_raceethnicity_2000] but have declined substantially since 2004 [@warren_patterns_2014].
+
+Prior work on grade retention has failed to show a positive effect of grade retention, and has instead found evidence of "scarring" effects of grade retention that lead to later negative educational outcomes, such as a higher dropout risk [@jimerson_metaanalysis_2001; @stearns_staying_2007; @andrew_scarring_2014; @hughes_effect_2017].
+The risk of grade retention also varies substantially by race and class background, with Black and Latino students at particularly high risk of grade retention relative to other groups [@warren_patterns_2014].
+
+Because grade retention often happens very early in the K-12 system, it is one of the earliest mechanisms that individuals encounter for reproducing social inequality and stratification.
+For these same reasons, it can also provide an important window into understanding the placement of biracial children within a racially stratified educational environment.
+
+Despite its long-standing use in the educational system, grade retention is famously difficult to measure accurately.
+In most data sources, researchers do not have access to specific reports of grade retention, but must rather infer grade retention from a discrepancy between a student's reported age and grade.
+Prior studies have used the concept of a student being behind modal grade [@bianchi_children_1984; @frederick_have_2008].
+This measure captures the cumulative prevalence of grade retention, but not its incidence.
+It also cannot distinguish cases in which children were delayed in starting kindergartern from grade retention.
+The most notable limitation of this approach is the ambiguity created by the fact that at any given age, a student may reasonably be in two modal grades.
+Without detailed information on birthdates and survey timing, the correct modal grade for most students cannot be identified.
+This issue has been somewhat alleviated in prior work by the use of the October supplement to the Current Population Survey which is close to the beginning of the school year.
+
+In the American Community Survey (ACS) data that I use here, I do not have the luxury of being able to identify or limit survey timing, so I instead use a related measure of whether a given student is clearly behind expected grade (CBEG).
+A student is considered CBEG if their age is higher than either of the expected ages for a student of that grade.
+This measure will underestimate overall grade retention because it will miss students who have been retained but not yet had a birthday in their current grade that would place them CBEG.
+However, the goal of this study is not to estimate grade retention accurately, but rather to understand racial differences in grade retention.
+Because this bias is largely a function of when students have birthdays, it should be more or less random with regard to sociodemographic characteristics.
+This approach is equivalent to the method used by @rosenfeld_nontraditional_2010 on 2000 Census data, but more detailed information on current grade in recent ACS data allows for more precise estimation.
 

### 2022-11-02 products/paper/main.qmd
Add changes from first round of copy editing

@@
 This demographic phenomenon may have a potentially transformative effect on racial boundaries [@hochschild_creating_2012; @alba_great_2020].
-Although race mixing has always been a feature of US history [@morning_who_2000; @gullickson_choosing_2011], this biracial boom is notable for its demographic size and for occurring in a post-Civil Rights era in which historical norms governing how to handle racial mixture may be shifting.
+While racial mixing is nothing new to US history [@morning_who_2000; @gullickson_choosing_2011], this biracial boom is notable for its demographic size and for a post-Civil Rights environment in which historical norms of racial classification may be eroding.
 Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of how these biracial individuals identify their own race and how they are sorted into a prevailing racial hierarchy that reinforces a system of racial inequality.
 
-Extensive scholarship documents the identification and classification decisions of multiracial individuals [@xie_racial_1997; @roth_end_2005a; @brunsma_interracial_2005; @herman_forced_2004; @holloway_place_2009] <!-- TODO MORE HERE -->.
-This research has generally identified a surprising degree of fluidity and choice in classification decisions, in the context of historical norms of racial classification <!-- TODO CHECK -->.
+Extensive scholarship documents the identification and classification decisions made by or on behalf of multiracial individuals [@xie_racial_1997; @roth_end_2005a; @brunsma_interracial_2005; @herman_forced_2004; @holloway_place_2009] <!-- TODO MORE HERE -->.
+This research has generally found a surprising degree of fluidity and choice in classification decisions <!-- TODO CHECK -->.
 On the other hand, only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers [@kao_racial_1999; @campbell_multiracial_2009; @herman_blackwhiteother_2009].
-The lack of work on this topic, despite its acknowledged relevance, results from two important limitations in most available data sources.
+Two important data limitations have contributed to the lack of work on this important topic.
 
-First, using direct identification of biracial respondents on data sources such as the Census or large scale surveys is problematic.
+First, using self identification of biracial respondents to analyze outcomes is problematic.
 Research on the identification and classification of biracial individuals indicates that individuals whose parents belong to different races will not always or consistently identify with both races [@doyle_are_2007; @liebler_america_2017; @gullickson_racial_2019].
-Furthermore, social status and social mobility has been shown to predict racial identification generally [@saperstein_racial_2012; @saperstein_mulatto_2013] and social status has been shown to be related to the identification choices of biracial individuals in particular [@roth_end_2005a; @davenport_role_2016].
-Thus, differences in outcome may in some part produce differences in identification rather than the other way around when examining self-identified biracial individuals.
+Furthermore, social status and social mobility can predict racial identification generally [@saperstein_racial_2012; @saperstein_mulatto_2013] and social status is related to the identification choices of biracial individuals specifically [@roth_end_2005a; @davenport_role_2016].
+Thus, differences in outcome among biracial individuals may in part produce differences in identification, rather than the other way around.
 
@@
 When children reside with both biological parents, biracial individuals can be identified in large scale data sources such as the Census.
-Researchers have long utilized the ability to identify such biracial children to examine the classification choices made on behalf of these children [@xie_racial_1997; @roth_end_2005a].
-Because of their age, it is more difficult to analyze the outcomes of such children due to the simple fact that they haven't had many measurable life outcomes yet.
+Researchers have previously utilized this feature to examine the classification choices made on behalf of these children [@xie_racial_1997; @roth_end_2005a].
+Analyzing the outcomes for such children is more difficult, for the simple reason that they do not have many measurable outcomes yet.
 
-However, one important outcome, grade retention, is measurable and important even for young children.
+One important outcome, grade retention, is measurable and important even for young children, however.
 Being held back in school is a relatively common occurrence that has consequences for later educational transitions and is disproportionately applied by race <!-- TODO CITATIONS -->.
@@
 In every case, the relevant comparison is how a biracial person fares relative to the two constituent groups with which their parents identify.
-For example, in the case of an individual with one Black and one White parent, we would compare the average outcomes for such individuals to the average outcomes for individuals with two White and two Black parents.
+For example, in the case of an individual with one Black and one White parent, we would compare the average outcomes for such individuals to the average outcomes for individuals with two White and two Black parents, respectively.
 The outcome for the Black/White group may be similar to one of the monoracial groups, it may be somewhere in between the outcomes for the two monoracial groups, or it may be higher or lower than either of the monoracial groups.
 
-To better understand where biracial individuals might fit, it is necessary to understand what is producing a contemporary system of racial inequality overall.
-Here, I distinguish three mechanisms that may help us understand the biracial position within this racial hierarchy.
+I distinguish three mechanisms that help us understand where biracials might be positioned within overall system of racial inequality.
 First, I consider arguments for the racial hierarchy that focus on the effects of racial discrimination at both an individual and systemic level.
-Second, I consider arguments regarding different distributions of resources across racial groups that produce different average outcomes.
+Second, I consider arguments regarding different distributions of resources across racial groups.
 Third, I consider arguments regarding how the nature of hybridity itself may affect outcomes.
@@
 -->.
-Thus, observed racial inequality reflects both contemporary processes of discrimination and bias and the intergenerational accumulation of disadvantage.
+Thus, observed racial inequality reflects both contemporary processes and the intergenerational accumulation of disadvantage.
 
@@
 
-However, the actual resource position of biracial individuals is likely far more complicated because individuals are not randomly selected into interracial unions.
-How this selection will affect a biracial child's outcomes for a given type of pairing are complex because it depends on the strength and direction of that selection on each partner, its consistency across different types of unions, and how strongly the selected characteristic predicts outcomes in the next generation.
+However, the actual resource position of biracial individuals is likely far more complicated because people are not randomly selected into interracial unions.
+The relationship between this selection process and a biracial child's outcomes are complex because it depends on the strength and direction of that selection on each partner, its consistency across different types of unions, and how strongly the selected characteristic predicts outcomes in the next generation.
 
@@
 Although this approach is sometimes characterized as pitting race versus class and minimizing the importance of race, that is not case here.
-Instead, what this delineation allows us to do is understand how race affects the outcomes of biracial individuals in a multigenerational process to better pinpoint the generational locus of specific mechanisms.
+Instead, what this delineation allows us to do is understand how race affects the outcomes of biracial individuals in a multigenerational process.
 
@@
 The experience of hybridity itself may also positively and/or negatively affect outcomes for biracial individuals.
-The "marginal man" theory argues that biracial individuals will find it difficult to fit in because of the dissonance of an identity that crosses strong and salient racial boundaries [@park_human_1928; @stonequist_problem_1935].
+The "marginal man" theory argues that biracial individuals will find it difficult to fit in due to the dissonance of an identity that crosses strong and salient racial boundaries [@park_human_1928; @stonequist_problem_1935].
 This liminal identity conflict will then result in negative mental health experiences that could feed into other negative outcomes as well.
@@
 
-Marginal man theory, developed in the first half of the Twentieth century, viewed culture in monolithic terms and thus to be trapped in between two static and monolithic cultures was seen as a difficult experience.
-Contemporary notions of culture emphasize its "toolkit" nature, arguing for a more flexible and fluid notion of culture <!-- TODO who to cite here?
+Marginal man theory viewed culture in monolithic terms and thus to be trapped in between two static and monolithic cultures was seen as a difficult experience.
+Contemporary notions of culture often use a "toolkit" approach that emphasizes the often flexible and fluid use of cultural elements <!-- TODO who to cite here?
 Swidler obviously, but also Dimaggio.
 Look at Dimaggio and Markus article -->.
-In this case, biracial individuals may benefit from an enlarged cultural toolkit that allows them to navigate a variety of diverse social contexts (i.e.
+From this perspective, biracial individuals may benefit from an enlarged cultural toolkit that allows them to navigate a variety of diverse social contexts (i.e.
 code switch) <!-- TODO There are citations for this -->.
@@
 --> and that various selection effects operate on the decision to identify as biracial among those who could [@roth_end_2005a; @davenport_role_2016].
-Furthermore, data on changes in racial reporting over time suggests that other status markers like education and income may influence how people identify themselves [@saperstein_racial_2012; @saperstein_mulatto_2013].
+Furthermore, research on changes in racial reporting over time suggests that other status markers like education and income may influence how people identify themselves [@saperstein_racial_2012; @saperstein_mulatto_2013].
 For example, a person with one Black and one White parent who is routinely treated by others as Black and discriminated against as Black may be more likely to identify themselves as exclusively Black than a similar person of Black and White ancestry who experiences less consistent discrimination.
@@
 @campbell_multiracial_2009 used data from the National Longitudinal Survey of Adolescent to Adult Health (Add Health) to examine high school GPA, advanced placement in math, and four-year college enrollment among biracial and monoracial individuals identified by co-resident biological parents' races.
-@herman_blackwhiteother_2009 used data on high school students in California and Wisconsin in which the students reported their parents' races to examine differences in GPA.
+@herman_blackwhiteother_2009 examines differences in GPA using data on high school students in California and Wisconsin in which the students reported their parents' races.
 Both of these studies find some support for an "in-between" status for many biracial groups, but are both hampered by a lack of statistical power due to small sample sizes.
-@kao_racial_1999 used data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial Blacks and Asians, and finds that biracial Blacks have outcomes similar to monoracial Blacks, while biracial Asians have outcomes more similar to monoracial Whites.
+@kao_racial_1999 used data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial Blacks and Asians<!-- With what?
+-->, and finds that biracial Blacks have outcomes similar to monoracial Blacks, while biracial Asians have outcomes more similar to monoracial Whites.
 This study, however, is hampered by the fact that biracial students are identified by a discrepancy between their reported race and one parent's reported race, rather than by both biological parent's races.
 
-The primary difficulty with the survey data used in the preceding studies is that the sample size for most biracial groups will be small and thus results tend to be imprecise and inconclusive.
+Results in prior work tend to be imprecise and inconclusive, due to the sample size for most biracial groups.
 An alternative approach is to use large scale Census or American Community Survey (ACS) data to identify biracial children by the race of the biological parents in their household.
@@
 This approach has been used previously to examine the likelihood of living in poverty [@bratter_multiracial_2013; @bratter_poverty_2013] and residential segregation [@ellis_agents_2012].
-<!-- TODO briefly summarize findings --> While these findings provide valuable information of the context of biracial childrens' lived experience, but do not directly measure their own outcomes, but rather the experiences of their parents.
+<!-- TODO briefly summarize findings --> These studies provide valuable information on biracial childrens' lived experience, but do not directly measure their own outcomes.
 
@@
 The frequency of grade retention has fluctuated historically in response to changes in educational policy.
-Rates of grade retention increased throughout the late 20th century [@hauser_raceethnicity_2000] but have declined substantially since 2004 [@warren_patterns_2014].
+Rates of grade retention increased throughout the late Twentieth century [@hauser_raceethnicity_2000] <!-- Check this because their more nuanced results do not necessarily show this --> but have declined substantially since 2004 [@warren_patterns_2014].
 
-Prior work on grade retention has failed to show a positive effect of grade retention, and has instead found evidence of "scarring" effects of grade retention that lead to later negative educational outcomes, such as a higher dropout risk [@jimerson_metaanalysis_2001; @stearns_staying_2007; @andrew_scarring_2014; @hughes_effect_2017].
+Prior work on grade retention has failed to show a positive effect of grade retention on later educational outcomes, and has instead found evidence of "scarring" effectsthat lead to later negative educational outcomes, such as a higher dropout risk [@jimerson_metaanalysis_2001; @stearns_staying_2007; @andrew_scarring_2014; @hughes_effect_2017].
 The risk of grade retention also varies substantially by race and class background, with Black and Latino students at particularly high risk of grade retention relative to other groups [@warren_patterns_2014].
 
-Because grade retention often happens very early in the K-12 system, it is one of the earliest mechanisms that individuals encounter for reproducing social inequality and stratification.
-For these same reasons, it can also provide an important window into understanding the placement of biracial children within a racially stratified educational environment.
-
-Despite its long-standing use in the educational system, grade retention is famously difficult to measure accurately.
-In most data sources, researchers do not have access to specific reports of grade retention, but must rather infer grade retention from a discrepancy between a student's reported age and grade.
-Prior studies have used the concept of a student being behind modal grade [@bianchi_children_1984; @frederick_have_2008].
-This measure captures the cumulative prevalence of grade retention, but not its incidence.
-It also cannot distinguish cases in which children were delayed in starting kindergartern from grade retention.
-The most notable limitation of this approach is the ambiguity created by the fact that at any given age, a student may reasonably be in two modal grades.
-Without detailed information on birthdates and survey timing, the correct modal grade for most students cannot be identified.
-This issue has been somewhat alleviated in prior work by the use of the October supplement to the Current Population Survey which is close to the beginning of the school year.
-
-In the American Community Survey (ACS) data that I use here, I do not have the luxury of being able to identify or limit survey timing, so I instead use a related measure of whether a given student is clearly behind expected grade (CBEG).
-A student is considered CBEG if their age is higher than either of the expected ages for a student of that grade.
-This measure will underestimate overall grade retention because it will miss students who have been retained but not yet had a birthday in their current grade that would place them CBEG.
-However, the goal of this study is not to estimate grade retention accurately, but rather to understand racial differences in grade retention.
-Because this bias is largely a function of when students have birthdays, it should be more or less random with regard to sociodemographic characteristics.
-This approach is equivalent to the method used by @rosenfeld_nontraditional_2010 on 2000 Census data, but more detailed information on current grade in recent ACS data allows for more precise estimation.
+Because grade retention often happens very early in the K-12 system, it is one of the earliest stratification mechanisms that individuals encounter.
+For these same reasons, it provids an important window into understanding the placement of biracial children within a racially stratified educational environment.
+In the remainder of this article, I examine this placement using data from the American Community Survey.
 
@@
 Prior work has used a variety of additional restrictions to limit the analysis to children who are more likely to be the biological children of both parents [@saenz_persistence_1995; @xie_racial_1997].
-Some of those procedures rely upon an assumption that most parents would be married that was more tenable in historical than contemporary data.
-I do not explicitly restrict by marital status, but I do restrict cases in two other ways that are consistent with prior work.
+Consistent with this prior work, I restrict cases in two ways.
 First, I restrict the sample to cases where both parents were of a reasonable age at the birth of the child (aged 15-44 for mothers and aged 15-60 for fathers).
@@
 The final analytical sample consists of children aged 5-20 who are currently enrolled in the K-12 school system and live in a household with two parents who are both likely to be biological parents.
-This necessary sample restriction will bias the family structure of the analytical sample relative to the total population of K-12 schoolchildren, limiting it exclusively to households with two parents present.
+This necessary sample restriction will bias the family structure of the analytical sample relative to the total population of K-12 schoolchildren.
 The result of this bias is well-known.
-Children living in two-parent nuclear households have greater educational success than other children as a result of material, social, and cultural resources [@biblarz_family_1999].
+Children living in two-parent nuclear households have greater educational success than other children as a result of material and social resources [@biblarz_family_1999].
 Therefore, this sample restriction will bias overall estimates of grade retention downward.
@@
 
-Measuring grade retention exactly is not possible in the ACS data.
-Without data on date of birth and survey timing, a student of any given age who has not been held back could reasonably be in two different grades.
-For example, a first grader is expected to be either six or seven years of age, while a second grader is expected to be either seven or eight years of age.
-Therefore, with information only on current grade and age in years, it is impossible to determine whether a seven year old in first grade is behind the expected grade level or in the expected grade.
+Grade retention is famously difficult to measure accurately.
+In most data sources, researchers do not have access to specific reports of grade retention, but rather infer grade retention from a discrepancy between a student's reported age and grade.
+Prior studies have used the concept of a student being behind modal grade [@bianchi_children_1984; @frederick_have_2008].
+This measure captures the cumulative prevalence of grade retention, but not its incidence.
+It also cannot distinguish cases in which children were delayed in starting kindergarten from grade retention.
+The most notable limitation of this approach is the ambiguity created by the fact that at any given age, a student may reasonably be in two modal grades.
+Without detailed information on birthdates and survey timing, the correct modal grade for most students cannot be identified.
+This issue has been somewhat alleviated in prior work by the use of the October supplement to the Current Population Survey which is close to the beginning of the school year.
 
-Instead of trying to identify exact grade retention, I instead identify cases in which students are clearly behind expected grade.
-A student is clearly behind expected grade when the age of the student is greater than either of the possible ages for a student of that grade (e.g.
-a first grader who is eight years old).
-This indirect measure of grade retention has been frequently used in previous studies of grade retention outcomes [@bianchi_children_1984; @frederick_have_2008; @rosenfeld_nontraditional_2010].
-This measure will underestimate the total proportion of students that have been held back and may also suffer from other biases [@warren_patterns_2014].
-However, the goal of this study is to examine differences across groups rather than total levels.
-The primary factor determining bias is whether the birthday of the student takes place earlier or later in the year, which is unlikely to vary substantially across groups.[^1]
+The ACS data do not provide the ability to identify or limit survey timing, so I instead use a related measure of whether a given student is clearly behind expected grade (CBEG).
+A student is considered CBEG if their age is higher than either of the expected ages for a student of that grade.
+This measure will underestimate overall grade retention because it will miss students who have been retained but not yet had a birthday in their current grade that would place them CBEG.
+However, the goal of this study is not to estimate grade retention accurately, but rather to understand racial differences in grade retention.
+Because this bias is largely a function of when students have birthdays, it should be more or less random with regard to sociodemographic characteristics.[^1] This approach is equivalent to the method used by @rosenfeld_nontraditional_2010 on 2000 Census data, but more detailed information on current grade in recent ACS data allows for more precise estimation.
+@tbl-descriptive shows the percent of each monoracial and biracial group that is CBEG.
 
@@
 
-@tbl-descriptive shows the percent of each monoracial and biracial group that is clearly behind expected grade.
-My goal is to understand the risk of being behind expected grade for each biracial group relative to its constituent monoracial groups (e.g.
+My goal is to understand the risk of being CBEG for each biracial group relative to its constituent monoracial groups (e.g.
 a Black/White student compared to White and Black students).
 Throughout this article, I use a visual approach to illustrate this placement.
-
 As an example, I display the percent of students clearly behind expected grade for White, Black, and Black/White students in @fig-example.
@@
 My goal is to determine whether the biracial group's point estimate is statistically distinguishable from a given monoracial constituent group's point estimate by non-overlap of confidence intervals.
-Contrary to common belief, overlap in 95% confidence intervals is not equivalent to failing a hypothesis test of difference at $p<0.05$ and leads to far more stringent tests [@knol_mis_2011].
+Overlap in 95% confidence intervals is not equivalent to failing a hypothesis test of difference at $p<0.05$ and when used in this way, such assessments lead to far more stringent tests [@knol_mis_2011].
 When standard errors between groups are equal, overlap in 83.4% confidence intervals will achieve this goal.
@@
 
-```{r example-plot}
+```{r exampl_plot}
 #| label: fig-example
@@
 Overlap in these color corresponding bars and bands indicates that the difference between the biracial group and the monoracial group is not statistically significant at $p<.05$.
-For example, the yellow bar shown in @fig_example for biracial Black/White students does not quite overlap with the yellow band for White students, indicating that these point estimates for these two groups are statistically distinguishable at $p<.05$.
+For example, the yellow bar shown in @fig-example for biracial Black/White students does not quite overlap with the yellow band for White students, indicating that the point estimates for these two groups are statistically distinguishable at $p<.05$.
 <!-- TODO provide actual p-value for comparison -->
@@
 However, in practice, I want to estimate differences across groups while holding constant a variety of potentially confounding variables.
-To do this, I estimate a set of logit models that predict the likelihood of a student being clearly behind expected grade by race and a variety of other variables.
+To do this, I estimate a set of logit models that predict the likelihood of a student being CBEG by race and a variety of other variables.
 I then construct figures similar to @fig-example by calculating from the model the average predicted probabilities (APP) for each racial group.
@@
 
-The probability of being clearly behind expected grade also increases with the student's current grade, so I include fixed effects for the current grade of the student.
-Additionally, @fig-year-grade shows that the percent of students clearly behind grade level has declined over time, but this decline has been much more substantial at higher grade levels.
+The probability of being CBEG also increases with the student's current grade, so I include fixed effects for the current grade of the student.
+Additionally, @fig-year-grade shows that the percent of students who are CBEG has declined over time, but this decline has been much more substantial at higher grade levels.
 For the elementary grades, there is no evidence of a decline at all.
@@
 All statistics reported here adjust for sampling weights among respondents.
-All models incorporate design effects for variance in sample weights abd the clustering of multiple respondents within the same household.
+All models incorporate design effects for variance in sample weights and the clustering of multiple respondents within the same household.
 
@@
 
-I begin by showing the relative risk of being behind expected grade among monoracial respondents across models.
-@tbl-monoracial reports average marginal effects on the probability of being behind expected grade (hereafter BBEG) for each monoracial minority group in comparison to White students.
-Understanding these differences and how they change across models helps me to understand the potential for how biracial respondents might be positioned.
+I begin by showing the relative risk of being CBEG among monoracial respondents across models.
+@tbl-monoracial reports average marginal effects on the probability of being CBEG for each monoracial minority group in comparison to White students.
+Understanding these differences helps to clarify the potential for how biracial respondents might be positioned between monoracial groups.
 
@@
 These results provide a baseline estimate of the differences across monoracial groups without controlling for differences in cultural and material resources across groups.
-The results show that Black, Indigenous, and Latino students all have substantially higher probability of BBEG than White students.
-Indigenous students have substantially higher risk than all other students, with a probability of BBEG that is 4.1% percentage points higher than White students.
-Asian students have the lowest probability of BBEG, and their risk is slightly lower than Whites.
+The results show that Black, Indigenous, and Latino students all have substantially higher probability of being CBEG than White students.
+Indigenous students have substantially higher risk than all other students, with a probability of being CBEG that is 4.1% percentage points higher than White students.
+Asian students have the lowest probability of being CBEG, and their risk is slightly lower than Whites.
 
@@
 Model 2 includes controls for whether the student and each of their parents are native-born and speak English well.
-Model 3 controls for the highest degree received for each parent, and Model 4 controls for family income (square rooted), home ownership, and whether the student's parents are married.
+Model 3 controls for the highest degree received by each parent, and Model 4 controls for family income (square rooted), home ownership, and whether the student's parents are married.
 
@@
 Controlling for education in Model 3 reduces the gap substantially for Black and Indigenous students and removes the gap entirely for Latino students.
-In fact, the results of both Models 3 and 4 suggest that, when holding constant cultural and family resources, Latino students are slightly less likely than White students to be held back.
+In fact, the results of both Models 3 and 4 suggest that, when holding constant cultural and family resources, Latino students are slightly less likely than White students to be CBEG.
 
-In total, resource differentials account for a substantial part of the overall racial differences in the risk of BBEG, but not all of these differences.
+In total, resource differentials account for a substantial part of the overall racial differences in the risk of being CBEG, but not all of these differences.
 Black and Indigenous students remain at higher risk than other students.
-Notably, Asian students rather than White students have the lowest risk of BBEG in all models.
+Notably, Asian students rather than White students have the lowest risk of being CBEG in all models.
 The most notable shift across models is for Latino students.
-Their higher risk of BBEG relative to White students is completely accounted for by resource differentials.
-When comparing hite and Latino students with the same level or resources in Model 4, Latino students actually have slightly lower risk of BBEG.
+Their higher risk of being CBEG relative to White students is completely accounted for by resource differentials.
+When comparing hite and Latino students with the same level or resources in Model 4, Latino students actually have slightly lower risk of being CBEG.
 
@@
 Therefore, I also provide a summation of these resource differentials by racial group by calculating a counterfactual probability of BBEG for each group based on their observed distribution of resource characteristics.
-This counterfactual probability is calculated by estimating the average predicted probability for each group from the fullest model (Model 4 of \@tbl-monoracial) when all other variables, including race, are held at their mean.[^8]
+This counterfactual probability is calculated by estimating the average predicted probability for each group from the fullest model (Model 4 of \@tbl-monoracial) when all non-resource variables, including race, are held at their mean.[^8]
 
@@
 #| label: tbl-resources
-#| tbl-cap: Mean resources by racial group. All results are survey weighted. Results sorted by counterfactual probability of grade retention. Shading indicates biracial group.
+#| tbl-cap: Mean resources by racial group. All results are survey weighted. Results sorted by counterfactual probability of being clearly behind expected grade. Shading indicates biracial group.
 index_biracial <- which(str_detect(tbl_resources$race, "/"))
@@
       align = c("l",rep("r", ncol(tbl_resources)-1)),
-      col.names=c("Race","Behind expected grade, counterfactual","Family income",
+      col.names=c("Race","CBEG, counterfactual","Family income",
                   "Own home","Mother, 4-year college degree",
@@
 
-The results indicate that we would expect very different risk of BBEG across biracial groups because of differences in resources.
-For example, White/Asian students have the lowest counterfactual risk of BBEG of all groups at 2.4%, owing to their exceptionally high family income and the high college attainment of their parents.
+The results show different risks of being CBEG across biracial groups because of differences in resources.
+For example, White/Asian students have the lowest counterfactual risk of being CBEG of all groups at 2.4%, owing to their exceptionally high family income and the high college attainment of their parents.
 At the other end of the spectrum, Indigenous/Latino students have a counterfactual risk of BBEG nearly 2% higher, given their much lower family resources.
@@
 My particular interest, however, is in how each biracial group compares to its constituent monoracial groups.
-For example, while Indigenous/Latino students have the highest counterfactual risk of BBEG among biracial students, their risk is still lower than their two constituent groups of Indigenous (4.5%) and Latino (5.3%) students, owing in large part to a somewhat higher family income, greater parental educational attainment, and greater acculturation.
+For example, while Indigenous/Latino students have the highest counterfactual risk of being CBEG among biracial students, their risk is still lower than their two constituent groups of Indigenous (4.5%) and Latino (5.3%) students, owing in large part to somewhat higher family income, greater parental educational attainment, and greater acculturation.
 These results suggest positive selection into Indigenous/Latino interracial unions.
@@
 #| label: tbl-resource-summary
-#| tbl-cap: Relative placement for biracial groups in comparison to monoracial constituent groups of counterfactual probability of being behind grade from resources alone.
+#| tbl-cap: Relative placement for biracial groups in comparison to monoracial constituent groups based on counterfactual probability of being clearly behind expected grade due to resources alone.
 race <- c("White/Asian","Black/Asian","Latino/Asian","White/Latino",
@@
 The effect of controlling for these resources will be complex and different across biracial groups.
-For those biracial groups with resources similar to their more advantaged constituent group, we should see that these controlling for these resources should reduce their advantage relative to the less advantaged monoracial group.
+For those biracial groups with resources similar to their more advantaged constituent group, we should see that controlling for these resources should reduce their advantage relative to the less advantaged monoracial group.
 For the Black/White case, on the other hand, we would expect that controlling for resource differentials should eliminate some of their disadvantage relative to Whites.
@@
 
-I now turn to the results for biracial respondents.
+I now turn to the risk of CBEG for biracial respondents.
 Due to the number of comparisons being made, I split these results into two separate figures.
@@
 #| fig-pos: p
-#| fig-cap: Probability of being behind grade for biracial respondents with one black parent, in comparison to their monoracial comparison groups.
+#| fig-cap: Probability of being clearly behind expected grade for biracial respondents with one black parent, in comparison to their monoracial comparison groups.
 .
@@
 @fig-black-models provides little evidence for such a pattern, and soundly rejects it in the case of Black/White and Black/Latino students.
-Black/White and Black/Latino students have a risk of BBEG substantially lower than Black students and closer, and perhaps even lower, than their non-Black constituent group.
+Black/White and Black/Latino students have a risk of being CBEG substantially lower than Black students and closer to their non-Black constituent group.
 Solid conclusions are difficult for the remaining two groups because of smaller sample size and correspondingly wider confidence intervals on estimates.
-Nonetheless the point estimates in each of these cases suggest probabilities of BBEG roughly midway between the two constituent groups.
+Nonetheless the point estimates in each of these cases suggest probabilities of being CBEG roughly midway between the two constituent groups.
 
-Controlling for resource variables has some effect on the placement of part-Black biracial respondents.
-In the case of Black/White biracial respondents, Black/White students initially have a risk of BBEG slightly higher than and statistically distinguishable from White students.
-After controlling for cultural and family resources, however, Black/White respondents have a risk of BBEG that is slightly lower but not statistically distinguishable from White students.
+Controlling for resource variables has some effect on the placement of part-Black biracial respondents.Black/White biracial students initially have a risk of being CBEG slightly higher than and statistically distinguishable from White students.
+After controlling for cultural and family resources, however, Black/White studentss have a risk of being CBEG that is slightly lower but not statistically distinguishable from White students.
 This change reflects the relatively low level of parental education and income among Black/White students, which is much closer to Black students than White students, as shown on @tbl-descriptive.
 
-Black/Latino students start with a very different position, having a risk of BBEG substantially lower than both Black and Latino students, who have sonewgat similar risks.
-However, after controlling for cultural and family resources, Black/Latino students have a risk of BBEG that is statistically indistinguishable from the risk of Latino students and substantially lower than and statistically distinguishable from Black students.
-The change largely reflects the fact that Black/Latino students have greater resources than Latino (but not Black) students and once this advantage is held constant, they have a similar risk.
-After controlling for family resources, both Black/White and Black/Latino students have a risk of BBEG similar to their monoracial constituent group with lower risk.
+Black/Latino students start from a very different position, having a risk of being CBEG substantially lower than both Black and Latino students, who have somewhat similar risks.
+However, after controlling for cultural and family resources, Black/Latino students have a risk of being CBEG that is statistically indistinguishable from the risk of Latino students and substantially lower than and statistically distinguishable from Black students.
+This change largely reflects the fact that Black/Latino students have greater resources than Latino (but not Black) students and once this advantage is held constant, they have a similar risk.
+After controlling for family resources, both Black/White and Black/Latino students have a risk of being CBEG similar to their monoracial constituent group with lower risk.
 
@@
 Controlling for resources moves the point estimate for Black/Asian students from a point estimate close to Asian students to more of a halfway position between the two constituent groups.
-Black/Indigenous students have BBEG risk roughly halfway between their two constituent groups in both models.
-However, this finding of a halfway position for both groups is highly tentative owing to the wide confidence intervals around both groups, due to small sample size.
-
-@fig-other-models shows the results for the remaining biracial groups.
-In general, the results here show that biracial students are more likely to have a risk of BBEG closer to the monoracial constituent group with the lower risk than the higher risk, with some important caveats discussed below.
+Black/Indigenous students have risk of being CBEG roughly halfway between their two constituent groups in both models.
+However, this finding of a halfway position for both groups is highly tentative owing to the wide confidence intervals around both groups.
 
@@
 #| fig-pos: p
-#| fig-cap: Probability of being behind grade for non-black biracial respondents, in comparison to their monoracial comparison groups.
+#| fig-cap: Probability of being clearly behind expected grade for non-black biracial respondents, in comparison to their monoracial comparison groups.
 Non-overlap in color corresponding confidence bands indicates statistically significant difference at the 5% level, Baseline model includes year, grade, location, and state fixed effects.
@@
 
-For all groups, the full models generally indicate that each biracial group has a risk of BBEG similar to the risk of the lower monoracial constituent group or has risk higher than this group but less than the halfway point.
+@fig-other-models shows the results for the remaining biracial groups.
+For most of these groups, the full models indicate that each biracial group has a risk of being CBEG similar to the risk of the lower monoracial constituent group or has risk higher than this group but less than the halfway point.
 There are two exceptions to this general pattern.
-First, the point estimate of the probability of BBEG for the Indigenous/Latino group is roughly halfway between the two constituent groups, although it is not statistically distinguishable from the lower group.
+First, the point estimate of the probability of being CBEG for the Indigenous/Latino group is roughly halfway between the two constituent groups, although it is not statistically distinguishable from the lower group.
 Second, Latino/Asian students are the only case where the point estimate of the risk for the biracial group is higher than either monoracial group.
@@
 White/Asian students are the only case with a substantial shift, owing to their highly advantaged households.
-Initially white/Asian students have risks of BBEG substantially below both White and Asian students.
-After controlling for resources, the risks of BBEG for White/Asian students are indistinguishable from the those for Asian students, but still substantially and statistically distinguishable from the higher risk to White students.
+Initially White/Asian students have risk of being CBEG substantially below both White and Asian students.
+After controlling for resources, the risk of being CBEG for White/Asian students is indistinguishable from those for Asian students, but still substantially and statistically distinguishable from the higher risk to White students.
 
 Results for part-Latino students are complicated by the substantial change in the monoracial Latino risk across models.
-For example, while Latinos have substantially higher risk of BBEG than Whites in the baseline model, there is only a very small difference favoring Latinos in the full model.
+For example, while Latinos have substantially higher risk of being CBEG than Whites in the baseline model, there is only a very small difference favoring Latinos in the full model.
 Nonetheless, even in this case, we see that White/Latinos have risk closer to those of the lower Latino group, but distinguishable from both groups.

### 2022-11-11 products/paper/main.qmd
Clean out most TODO notes from paper draft

@@
 This demographic phenomenon may have a potentially transformative effect on racial boundaries [@hochschild_creating_2012; @alba_great_2020].
-While racial mixing is nothing new to US history [@morning_who_2000; @gullickson_choosing_2011], this biracial boom is notable for its demographic size and for a post-Civil Rights environment in which historical norms of racial classification may be eroding.
+While racial mixing is nothing new to US history [@morning_who_2000; @gullickson_choosing_2011], this biracial boom is notable for both its demographic size and for a post-Civil Rights environment in which historical norms of racial classification may be eroding.
 Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of how these biracial individuals identify their own race and how they are sorted into a prevailing racial hierarchy that reinforces a system of racial inequality.
 
-Extensive scholarship documents the identification and classification decisions made by or on behalf of multiracial individuals [@xie_racial_1997; @roth_end_2005a; @brunsma_interracial_2005; @herman_forced_2004; @holloway_place_2009] <!-- TODO MORE HERE -->.
-This research has generally found a surprising degree of fluidity and choice in classification decisions <!-- TODO CHECK -->.
+Extensive scholarship documents the identification and classification decisions made by or on behalf of multiracial individuals [@xie_racial_1997; @qian_options_2004; @roth_end_2005a; @brunsma_interracial_2005; @herman_forced_2004; @holloway_place_2009; @bratter_mother_2009; @davenport_role_2016; @liebler_boundaries_2016].
+This research has generally found a surprising degree of fluidity and choice in classification decisions that is nonetheless somewhat predictable by local context, socioeconomic status and other characteristics.
 On the other hand, only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers [@kao_racial_1999; @campbell_multiracial_2009; @herman_blackwhiteother_2009].
@@
 One important outcome, grade retention, is measurable and important even for young children, however.
-Being held back in school is a relatively common occurrence that has consequences for later educational transitions and is disproportionately applied by race <!-- TODO CITATIONS -->.
-Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience and within it we can clearly see the imprint of the racial hierarchy.
+Being held back in school is a relatively common occurrence that has consequences for later educational transitions [@jimerson_metaanalysis_2001] and is disproportionately applied by race [@warren_patterns_2014].
+Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience and reveals the imprint of the racial hierarchy.
 In this article, I use data from the American Community Survey to examine the likelihood of being held back in school for biracial children in comparison to their monoracial peers.
@@
 
-How might we expect the outcomes for biracial individuals do differ from their monoracial peers?
+How might we expect the outcomes for biracial individuals to differ from their monoracial peers?
 In every case, the relevant comparison is how a biracial person fares relative to the two constituent groups with which their parents identify.
@@
 
-Discrimination and bias by race play an important role in overall levels of racial inequality.
-Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unknowningly rely upon and reinforce in social interactions <!-- TODO citations, may be Pager, Brubaker?
--->.
+Discrimination and bias by race play an important role in overall levels of racial inequality [@quillian_new_2006; @pager_sociology_2008].
+Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unknowningly rely upon and reinforce in social interactions @quillian_new_2006].
 Such discrimination will be most consequential when performed by important gatekeepers of opportunity such as teachers, health care workers, and employers.
-Discrimination need not just occur on an individual basis, but can also be systemic in nature when biases are built into organizations and institutional systems <!-- TODO citations, Tilly?
-Bonilla-Silva, Pager again?
--->.
-In our current "colorblind" regime, such systems may be formally race neutral but produce disparate impacts by race <!-- TODO CITE Bonilla-Silva?
--->.
+Discrimination need not just occur on an individual basis, but can also be systemic in nature when biases are built into and reinforced by organizations and institutional systems [@reskin_race_2012].
+In our current "colorblind" regime [@bonilla-silva_racism_2006], such systems may be formally race neutral but produce disparate impacts by race [@pager_sociology_2008].
 
@@
 By comparison a pattern of *hyperdescent* implies that individuals of mixed racial ancestry will be classified as members of the higher status group.
-This pattern of hyperdescent has historically governed the experiences of individuals with American Indian ancestry <!-- TODO CITE -->.
+This pattern of hyperdescent has historically governed the experiences of individuals with American Indian ancestry [@snipp_american_1989; @wolfe_land_2001; @gullickson_choosing_2011].
 
-Although the hypodescent/hyperdescent paradigm is often used to theoretically frame the possible contemporary experiences of multiracial individuals <!-- TODO citations -->, there are reasons to be cautious about the generalizability of this paradigm to both the contemporary and historical situation.
+Although the hypodescent/hyperdescent paradigm is often used to theoretically frame the possible contemporary experiences of multiracial individuals, there are reasons to be cautious about the generalizability of this paradigm to both the contemporary and historical situation.
 Hypodescent and hyperdescent are observed in two specific historical cases involving Black ancestry and American Indian ancestry, respectively.
@@
 In this case, neither the hypodescent/hyperdescent or dominance paradigm may help us to fully understand the experiences of those of Latino and Asian ancestry due to more recent processes of racialization.
-Second, the historical norms for all groups may also be breaking up or transforming in the post-Civil Rights era <!-- TODO CITE -->.
+Second, the historical norms for all groups may also be breaking up or transforming in the post-Civil Rights era.
 
@@
 
-Racial inequality is also at least partly explained by different distributions of material, cultural, and social resources across racial groups <!-- TODO CITATIONS HERE?
--->.
+Racial inequality is also at least partly explained by different distributions of material, cultural, and social resources across racial groups [@conley_being_1999].
 On average, socioeconomic background (e.g.
@@
 Other less materially defined resources also vary across groups such as cultural styles, social networks, and neighborhood quality.
-When such characteristics are held constant or "controlled", the remaining differences between racial groups tend to be smaller, although such differences rarely disappear <!-- TODO WHAT CAN I CITE HERE -->.
-These differences in background resources reflect the accumulation of historical processes of discrimination <!-- CITE PAGER?
--->.
-Thus, observed racial inequality reflects both contemporary processes and the intergenerational accumulation of disadvantage.
+These differences in background resources reflect the accumulation of historical processes of discrimination and, when combined with contemporary processes of discrimination and bias, magnify the overall level of racial inequality we observe.
 
@@
 Using education as the socioeconomic parameter, prior research has shown strong evidence of this pattern in the case of Black male/White female marriages in the US, but weaker evidence for other types of unions [@gullickson_education_2006; @fu_racial_2001; @kalmijn_educational_2010; @hou_interracial_2011].
-Less research has explored these patterns outside of marriage <!-- TODO cites?
--->.
-In any case, because status exchange positively selects one parent and negatively selects the other partner, the two effects may somewhat cancel each other out in terms of the overall distribution of resources.
+
+Regardless of its overall prevalence, the overall effects of status exchange on resources are complex because it positively selects one parent and negatively selects the other partner.
+Therefore the two effects may somewhat cancel each other out in terms of the overall distribution of resources across parents.
+Complicating this issue further, research has shown that the parents of biracial children may differ from parents of monoracial children in the magnitude and types of resources that they allocate to children [@cheng_constraints_2007], even among parents of similar resources.
 
@@
 
-<!-- TODO how can I integrate Cheng and Powell? -->
-
 ## Hybridity
 
-The experience of hybridity itself may also positively and/or negatively affect outcomes for biracial individuals.
+The experience of hybridity itself may also positively or negatively affect outcomes for biracial individuals.
 The "marginal man" theory argues that biracial individuals will find it difficult to fit in due to the dissonance of an identity that crosses strong and salient racial boundaries [@park_human_1928; @stonequist_problem_1935].
@@
 
-Marginal man theory viewed culture in monolithic terms and thus to be trapped in between two static and monolithic cultures was seen as a difficult experience.
-Contemporary notions of culture often use a "toolkit" approach that emphasizes the often flexible and fluid use of cultural elements <!-- TODO who to cite here?
-Swidler obviously, but also Dimaggio.
-Look at Dimaggio and Markus article -->.
-From this perspective, biracial individuals may benefit from an enlarged cultural toolkit that allows them to navigate a variety of diverse social contexts (i.e.
-code switch) <!-- TODO There are citations for this -->.
+Marginal man theory viewed culture in monolithic and static terms and thus to be trapped in between two cultures was seen as a difficult experience.
+Contemporary views of culture instead emphasize the often flexible, fragmented, and fluid use of cultural bits that comprise a "toolkit" [@swidler_culture_1986; @dimaggio_culture_1997].
+From this perspective, biracial individuals may benefit from an enlarged cultural toolkit that allows them to navigate a variety of diverse social contexts [@shih_multiple_2019].
 
@@
 Both of these studies find some support for an "in-between" status for many biracial groups, but are both hampered by a lack of statistical power due to small sample sizes.
-@kao_racial_1999 used data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial Blacks and Asians<!-- With what?
--->, and finds that biracial Blacks have outcomes similar to monoracial Blacks, while biracial Asians have outcomes more similar to monoracial Whites.
+@kao_racial_1999 used data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial students with Black and Asian parentage and finds that biracial Blacks have outcomes similar to monoracial Blacks, while biracial Asians have outcomes more similar to monoracial Whites.
 This study, however, is hampered by the fact that biracial students are identified by a discrepancy between their reported race and one parent's reported race, rather than by both biological parent's races.
@@
 This approach has been used previously to examine the likelihood of living in poverty [@bratter_multiracial_2013; @bratter_poverty_2013] and residential segregation [@ellis_agents_2012].
-<!-- TODO briefly summarize findings --> These studies provide valuable information on biracial childrens' lived experience, but do not directly measure their own outcomes.
+These studies provide valuable information on biracial childrens' lived experience, but do not directly measure their own outcomes.
 
@@
 Grade retention is the practice of having students repeat a grade due to poor academic performance.
-The frequency of grade retention has fluctuated historically in response to changes in educational policy.
-Rates of grade retention increased throughout the late Twentieth century [@hauser_raceethnicity_2000] <!-- Check this because their more nuanced results do not necessarily show this --> but have declined substantially since 2004 [@warren_patterns_2014].
+Rates of grade retention likely increased throughout the late Twentieth century [@frederick_have_2008] but have declined substantially since 2004 [@warren_patterns_2014].
 
-Prior work on grade retention has failed to show a positive effect of grade retention on later educational outcomes, and has instead found evidence of "scarring" effectsthat lead to later negative educational outcomes, such as a higher dropout risk [@jimerson_metaanalysis_2001; @stearns_staying_2007; @andrew_scarring_2014; @hughes_effect_2017].
+Prior work on grade retention has failed to show a positive effect of grade retention on later educational outcomes, and has instead found evidence of "scarring" effects that lead to later negative educational outcomes, such as a higher dropout risk [@jimerson_metaanalysis_2001; @stearns_staying_2007; @andrew_scarring_2014; @hughes_effect_2017].
 The risk of grade retention also varies substantially by race and class background, with Black and Latino students at particularly high risk of grade retention relative to other groups [@warren_patterns_2014].
@@
 Because grade retention often happens very early in the K-12 system, it is one of the earliest stratification mechanisms that individuals encounter.
-For these same reasons, it provids an important window into understanding the placement of biracial children within a racially stratified educational environment.
+For these same reasons, it provides an important window into understanding the placement of biracial children within a racially stratified educational environment.
 In the remainder of this article, I examine this placement using data from the American Community Survey.
@@
 
-[^1]: Biracial respondents are slightly younger than monoracial respondents in the sample as a whole because interracial unions are becoming more common over time HOW STRONG COULD THE BIAS BE HERE AND WHAT CAN I DO ABOUT IT?
+[^1]: Biracial respondents are slightly younger than monoracial respondents in the sample as a whole because interracial unions are becoming more common over time which may contribute to a slight downward bias in their risk of grade retention.
 
@@
 For example, the yellow bar shown in @fig-example for biracial Black/White students does not quite overlap with the yellow band for White students, indicating that the point estimates for these two groups are statistically distinguishable at $p<.05$.
-<!-- TODO provide actual p-value for comparison -->
 
@@
 
-[^8]: It might seem odd to hold categorical variables at their mean, but because all categorical variables are technically entered into the models as 0/1 numeric indicator variables, their mean value can be calculated as the proportion of cases with the indicated value.
-This technique is frequently used in multilevel models to deal with grand and group mean centering of variables.
-<!-- TODO CITE raudenbush and bryk?
--->
+[^8]: It might seem odd to hold categorical variables at their mean, but because all categorical variables are entered into the models as 0/1 numeric indicator variables, their mean value can be calculated as the proportion of cases with the indicated value.
 

### 2022-11-11 products/paper/main.qmd
Update bibliography for previous changes

@@
 Discrimination and bias by race play an important role in overall levels of racial inequality [@quillian_new_2006; @pager_sociology_2008].
-Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unknowningly rely upon and reinforce in social interactions @quillian_new_2006].
+Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unknowningly rely upon and reinforce in social interactions [@quillian_new_2006].
 Such discrimination will be most consequential when performed by important gatekeepers of opportunity such as teachers, health care workers, and employers.

### 2022-11-16 products/paper/main.qmd
Make copy edit corrections

@@
 The experiences of mixed race individuals will play an important role in the future of race in the US.
-A rise in interracial marriage and coupling since the 1960s has resulted in a "biracial baby boom" of individuals whose parents cross racial lines [@root_racially_1992].
-This demographic phenomenon may have a potentially transformative effect on racial boundaries [@hochschild_creating_2012; @alba_great_2020].
-While racial mixing is nothing new to US history [@morning_who_2000; @gullickson_choosing_2011], this biracial boom is notable for both its demographic size and for a post-Civil Rights environment in which historical norms of racial classification may be eroding.
-Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of how these biracial individuals identify their own race and how they are sorted into a prevailing racial hierarchy that reinforces a system of racial inequality.
+A rise in interracial marriage and coupling since the 1960s has resulted in a "biracial baby boom" of individuals whose parents cross racial lines [@root_racially_1992] and this demographic surge may have a transformative effect on racial boundaries [@hochschild_creating_2012; @alba_great_2020].
+While racial mixing is nothing new to US history [@morning_who_2000; @gullickson_choosing_2011], this biracial boom is notable for its demographic size and has occurred in a post-Civil Rights environment in which historical norms of racial classification may be eroding.
 
-Extensive scholarship documents the identification and classification decisions made by or on behalf of multiracial individuals [@xie_racial_1997; @qian_options_2004; @roth_end_2005a; @brunsma_interracial_2005; @herman_forced_2004; @holloway_place_2009; @bratter_mother_2009; @davenport_role_2016; @liebler_boundaries_2016].
-This research has generally found a surprising degree of fluidity and choice in classification decisions that is nonetheless somewhat predictable by local context, socioeconomic status and other characteristics.
-On the other hand, only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers [@kao_racial_1999; @campbell_multiracial_2009; @herman_blackwhiteother_2009].
-Two important data limitations have contributed to the lack of work on this important topic.
+Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of both how these biracial individuals identify their own race and how they are sorted into a prevailing system of racial inequality.
+While extensive scholarship documents the identification and classification decisions made by or on behalf of multiracial individuals [@xie_racial_1997; @qian_options_2004; @roth_end_2005a; @brunsma_interracial_2005; @herman_forced_2004; @holloway_place_2009; @bratter_mother_2009; @davenport_role_2016; @liebler_boundaries_2016], only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers [@kao_racial_1999; @campbell_multiracial_2009; @herman_blackwhiteother_2009].
+In this article, I expand on this work by examining the risk of grade retention for K-12 students whose parents identify with different races.
 
+By examining this outcome, I am able to overcome two important limitations that have hampered prior work on the life chances of biracial individuals.
 First, using self identification of biracial respondents to analyze outcomes is problematic.
-Research on the identification and classification of biracial individuals indicates that individuals whose parents belong to different races will not always or consistently identify with both races [@doyle_are_2007; @liebler_america_2017; @gullickson_racial_2019].
+Research on the identification and classification of biracial individuals indicates that individuals whose parents belong to different races do not always consistently identify with both races [@doyle_are_2007; @liebler_america_2017; @gullickson_racial_2019].
 Furthermore, social status and social mobility can predict racial identification generally [@saperstein_racial_2012; @saperstein_mulatto_2013] and social status is related to the identification choices of biracial individuals specifically [@roth_end_2005a; @davenport_role_2016].
-Thus, differences in outcome among biracial individuals may in part produce differences in identification, rather than the other way around.
+Thus, differences in outcomes among biracial individuals may in part produce differences in identification, rather than the other way around.
 
 This problem of reverse causation can be largely addressed when researchers have access to the racial identification of a respondent's biological parents.
-While such information is available on some large scale longitudinal surveys, researchers then run into a second problem of sample size.
-The samples of biracial respondents in such cases are generally far too small to make reliable statistical inferences.
+While such information is available on some large scale longitudinal surveys, researchers then run into a second limitation of sample size.
+The samples of biracial respondents in such data sources are generally far too small to make reliable statistical inferences.
 This limitation is particularly severe for biracial individuals with two non-white parents.
@@
 
-One important outcome, grade retention, is measurable and important even for young children, however.
+However, one important outcome, grade retention, is measurable and important even for young children.
 Being held back in school is a relatively common occurrence that has consequences for later educational transitions [@jimerson_metaanalysis_2001] and is disproportionately applied by race [@warren_patterns_2014].
-Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience and reveals the imprint of the racial hierarchy.
+Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience.
 In this article, I use data from the American Community Survey to examine the likelihood of being held back in school for biracial children in comparison to their monoracial peers.
@@
 
-I distinguish three mechanisms that help us understand where biracials might be positioned within overall system of racial inequality.
+I distinguish three mechanisms that help us understand where biracials might be positioned within a system of racial inequality.
 First, I consider arguments for the racial hierarchy that focus on the effects of racial discrimination at both an individual and systemic level.
@@
 The strongest historical precedent for understanding such classification in the US is the "one drop rule" norm concerning Black ancestry.
-According to the one-drop rule, individuals of known Black ancestry are considered exclusively Black, regardless of other ancestries[@davis_who_1991].
+According to the one-drop rule, individuals of known Black ancestry are considered exclusively Black, regardless of other ancestries [@davis_who_1991].
 @davis_who_1991 generalizes this practice into the concept of *hypodescent* in which individuals with mixed racial ancestry will be classified as members of the lower status group.
@@
 
-Although the hypodescent/hyperdescent paradigm is often used to theoretically frame the possible contemporary experiences of multiracial individuals, there are reasons to be cautious about the generalizability of this paradigm to both the contemporary and historical situation.
+Although the hypodescent/hyperdescent paradigm is often used to theoretically frame the possible contemporary experiences of multiracial individuals, there are reasons to be cautious about the generalizability of this paradigm.
 Hypodescent and hyperdescent are observed in two specific historical cases involving Black ancestry and American Indian ancestry, respectively.
@@
 Status exchange theory predicts that given an unequal racial hierarchy, individuals from a lower status racial group will be positively selected on socioeconomic characteristics into intermarriage with a higher status group, while individuals from a higher status racial group will be similarly negatively selected into the same marriage [@davis_intermarriage_1941; @merton_intermarriage_1941; @fu_racial_2001].
-Using education as the socioeconomic parameter, prior research has shown strong evidence of this pattern in the case of Black male/White female marriages in the US, but weaker evidence for other types of unions [@gullickson_education_2006; @fu_racial_2001; @kalmijn_educational_2010; @hou_interracial_2011].
+Prior research has shown substantial evidence of this pattern in the case of Black male/White female marriages in the US, but weaker evidence for other types of unions [@gullickson_education_2006; @fu_racial_2001; @kalmijn_educational_2010; @hou_interracial_2011].
 
-Regardless of its overall prevalence, the overall effects of status exchange on resources are complex because it positively selects one parent and negatively selects the other partner.
+Regardless of its overall prevalence, the net effect of status exchange on resources is complex because status exchange positively selects one parent and negatively selects the other partner.
 Therefore the two effects may somewhat cancel each other out in terms of the overall distribution of resources across parents.
@@
 
-Marginal man theory viewed culture in monolithic and static terms and thus to be trapped in between two cultures was seen as a difficult experience.
+Marginal man theorists viewed culture in monolithic and static terms and thus to be trapped in between two cultures was seen as a difficult experience.
 Contemporary views of culture instead emphasize the often flexible, fragmented, and fluid use of cultural bits that comprise a "toolkit" [@swidler_culture_1986; @dimaggio_culture_1997].
@@
 The complexity of the underlying mechanisms described above prevents simple and straightforward hypotheses regarding the position of biracial individuals.
-Different mechanisms may produce the same observed outcome and the weight of different mechanisms may also vary by groups as these groups may governed by different regimes of classification.
+Different mechanisms may produce the same observed outcome and the weight of different mechanisms can also vary by groups as these groups may be governed by different regimes of classification.
 However, comparing outcomes across biracial groups may help us better understand the likely mechanisms that drive biracial classification and to disregard others as implausible.
@@
 This approach can generate much larger samples of biracial individuals, although it does exclude examination of children living in non-nuclear households.
-Furthermore, researchers then have a limited range of outcomes to explore because biracial individuals identified by this method are by definition children and the data sources include relatively little detailed information.
+Furthermore, researchers then have a limited range of outcomes to explore because biracial individuals identified by this method are by definition children.
 This approach has been used previously to examine the likelihood of living in poverty [@bratter_multiracial_2013; @bratter_poverty_2013] and residential segregation [@ellis_agents_2012].
@@
 
-However, one important educational outcome is identifiable for children in Census and ACS data - grade retention.
-Grade retention is partially identifiable in these data sources and provides one of the earliest measurable forms of stratification that individuals face in their lives.
+However, one important educational outcome is partially identifiable for children in Census and ACS data - grade retention. I now turn to a closer examination of this particular outcome.
 
@@
 However, this bias is less problematic for the goal of this study, which is to estimate differences between groups.
-The sample restriction effectively eliminates differences in family structure across racial groups which may account for some observed differences in grade retention.
+The sample restriction effectively eliminates differences in family structure across racial groups that may account for some observed differences in grade retention.
 However, the remaining differences observed here are likely to hold for the full population in the absence of strong racial differences in the effect of family structure on grade retention.
@@
 I also considered a similar interaction between state and year, but this model was not preferred.
+Model fit statistics for all senstivity analyses are provided in the supplementary materials.
 
@@
 #| label: fig-year-grade
-#| fig-cap: Trends over time and grade in the percent of students clearly behind grade level. Lines are fit to each set of points by grade via lowess smoothing.
+#| fig-cap: Trends over time and grade in the percent of students clearly behind expected grade. Lines are fit to each set of points by grade via lowess smoothing.
 
@@
   scale_y_continuous(labels=scales::percent)+
-  labs(x="year", y="percent clearly behind grade level for age",
+  labs(x="year", y="percent clearly behind expected grade",
        color="current grade")+
@@
 I begin by showing the relative risk of being CBEG among monoracial respondents across models.
-@tbl-monoracial reports average marginal effects on the probability of being CBEG for each monoracial minority group in comparison to White students.
+Table 2 reports average marginal effects on the probability of being CBEG for each monoracial minority group in comparison to White students.
 Understanding these differences helps to clarify the potential for how biracial respondents might be positioned between monoracial groups.
@@
 Model 1 only controls for structural factors such as state of residence, student location, grade, and year.
-These results provide a baseline estimate of the differences across monoracial groups without controlling for differences in cultural and material resources across groups.
+These results provide a baseline estimate of the differences across monoracial groups without controlling for differences in cultural and material resources.
 The results show that Black, Indigenous, and Latino students all have substantially higher probability of being CBEG than White students.
 Indigenous students have substantially higher risk than all other students, with a probability of being CBEG that is 4.1% percentage points higher than White students.
-Asian students have the lowest probability of being CBEG, and their risk is slightly lower than Whites.
+Asian students have the lowest probability of being CBEG, and their risk is slightly lower than White students.
 
@@
 Their higher risk of being CBEG relative to White students is completely accounted for by resource differentials.
-When comparing hite and Latino students with the same level or resources in Model 4, Latino students actually have slightly lower risk of being CBEG.
+When comparing White and Latino students with the same level or resources in Model 4, Latino students actually have slightly lower risk of being CBEG.
 
-Although, the racial differences in risk of BBEG are substantially reduced by controlling for resource differentials, the remaining gaps across all pairwise combinations are still large enough in most cases to sustain the question of where biracial individuals fit within these gaps.
+Although, the racial differences in risk of CBEG are substantially reduced by controlling for resource differentials, the remaining gaps across all pairwise combinations are still large enough in most cases to sustain the question of where biracial individuals fit within these gaps.
 Before turning to this question, however, I want to better understand how the resources of biracial individuals compare to their monoracial constituent groups.
@@
 @tbl-resources shows the mean values of the resource variables across monoracial and biracial groups.[^7] Because of the number of variables and groups involved, this table can be overwhelming to read.
-Therefore, I also provide a summation of these resource differentials by racial group by calculating a counterfactual probability of BBEG for each group based on their observed distribution of resource characteristics.
-This counterfactual probability is calculated by estimating the average predicted probability for each group from the fullest model (Model 4 of \@tbl-monoracial) when all non-resource variables, including race, are held at their mean.[^8]
+Therefore, I also provide a summation of these resource differentials by racial group by calculating a counterfactual probability of CBEG for each group based on their observed distribution of resource characteristics.
+This counterfactual probability is calculated by estimating the average predicted probability for each group from the fullest model (Model 4 of Table 2) when all non-resource variables, including race, are held at their mean.[^8]
 
@@
 For example, White/Asian students have the lowest counterfactual risk of being CBEG of all groups at 2.4%, owing to their exceptionally high family income and the high college attainment of their parents.
-At the other end of the spectrum, Indigenous/Latino students have a counterfactual risk of BBEG nearly 2% higher, given their much lower family resources.
+At the other end of the spectrum, Indigenous/Latino students have a counterfactual risk of CBEG nearly 2% higher, given their much lower family resources.
 
@@
 Table 4 summarizes the patterns from @tbl-resources across all ten biracial groups.
-In six of these cases, the biracial group would have outcomes similar or better than the monoracial group with better resources.
+In six of these cases, the biracial group would have outcomes similar to or better than the monoracial group with better resources.
 These results imply that for the majority of these groups, parents in these matches are being positively selected from their constituent groups.
@@
 In @fig-black-models, I show the placement of each biracial group involving one black parent in comparison to their two monoracial constituent groups.
-To allow comparison across models, I compare the results from the baseline model (Model 1 in @tbl-monoracial) and the full model that accounts for all cultural and family resources (Model 4 in @tbl-monoracial).
+To allow comparison across models, I compare the results from the baseline model (Model 1 in Table 2) and the full model that accounts for all cultural and family resources (Model 4 in Table 2).
 
@@
 
-Controlling for resource variables has some effect on the placement of part-Black biracial respondents.Black/White biracial students initially have a risk of being CBEG slightly higher than and statistically distinguishable from White students.
+Controlling for resource variables has some effect on the placement of part-Black biracial respondents.
+Black/White biracial students initially have a risk of being CBEG slightly higher than and statistically distinguishable from White students.
 After controlling for cultural and family resources, however, Black/White studentss have a risk of being CBEG that is slightly lower but not statistically distinguishable from White students.
-This change reflects the relatively low level of parental education and income among Black/White students, which is much closer to Black students than White students, as shown on @tbl-descriptive.
+This change reflects the relatively low level of parental education and income among Black/White students, which is much closer to Black students than White students, as shown on @tbl-resources.
 
@@
 The role of parental resources is harder to determine for Black/Asian and Black/Indigenous students due to the greater uncertainty in these estimates.
-Controlling for resources moves the point estimate for Black/Asian students from a point estimate close to Asian students to more of a halfway position between the two constituent groups.
+Controlling for resources moves the point estimate for Black/Asian students from a risk similar to Asian students to more of a halfway position between the two constituent groups.
 Black/Indigenous students have risk of being CBEG roughly halfway between their two constituent groups in both models.
-However, this finding of a halfway position for both groups is highly tentative owing to the wide confidence intervals around both groups.
+However, this finding of a halfway position for both groups is highly tentative owing to wide confidence intervals.
 
@@
 Second, Latino/Asian students are the only case where the point estimate of the risk for the biracial group is higher than either monoracial group.
-However, the confidence bands suggest that the risk for this group cannot be statistically differentiated from either of the two monoracial constituent groups who have very similar risks.
+However, the confidence bands suggest that the risk for this group cannot be statistically differentiated from either of the two monoracial constituent groups.
 This result only emerges in the model that controls for resources, because Latino/Asian students come from more advantaged households than both Latino and Asian students, on average.
@@
 Results for part-Latino students are complicated by the substantial change in the monoracial Latino risk across models.
-For example, while Latinos have substantially higher risk of being CBEG than Whites in the baseline model, there is only a very small difference favoring Latinos in the full model.
-Nonetheless, even in this case, we see that White/Latinos have risk closer to those of the lower Latino group, but distinguishable from both groups.
+For example, while Latino students have substantially higher risk of being CBEG than White students in the baseline model, there is only a very small difference favoring Latino students in the full model.
+Nonetheless, even in this case, we see that White/Latino students have risk closer to those of the lower Latino group, but distinguishable from both groups.
 

### 2022-11-16 products/paper/main.qmd
Add bullet points for conclusion

@@
 
--   Big claim: The results are consistent with a new "regime of hyperdescent" operating across all racial groups.
-This contributes to a muddying of racial distinctiveness via hybridity, while at the same time isolating monoracial members of the most stigmatized racial groups, consistent with arguments like Bonilla-Silva's Latin Americanization.
+-   Unlike prior work on this topic, most of the results here are robust due to large sample size and where the results are robust, they show a relatively consistent story across all groups: biracial students tend to have a grade retention risk similar to their monoracial constituent group with the lower risk.
+-   What does this pattern mean given the hypotheses of difference. We can strongly rule out a marginal man problem or a strong one-drop rule. Instead what emerges are two possible stories.
+    -   The first is a new regime of hyperdescent that seems to apply generally across a wide variety of cases, in contrast to the historical work on this topic.
+
+    -   The second is the possibility of a positive effect of hybridity on biracial students ability to negotiate racialized institutions.
+This may help to bring outcomes down from what might otherwise by a more in-between position based on other dynamics.
+-   What role do family resources and intergenerational selection play?
+The results vary somewhat by group, but generally show positive selection.
+The big exception is Black/White biracials who are more like Blacks in their outcomes.
+White/Asians stand out as particularly privileged relative to every other monoracial and biracial group in the sample.
+Nonetheless, while these resource differentials exist, they do not generally change the underlying story substantially - biracial outcomes are like those of the lower monoracial group.
+The one exception here is the Latino/Asian case.
+-   Better outcomes for biracial individuals is often incorrectly treated as a success story in the American context.
+While it may indicate that biracial individuals do better than some other groups within an American system of racial inequality, it does little to reduce that overall inequality.
+Instead it may heighten the relationship between outcomes BLAH BLAH
+-   What are the limitations of this study?
+    -   Although sample sizes are big generally, we still suffer from large sample errors in some cases of non-white biracials, particularly those of part Indigenous ancestry.
+There is some indication that indigeneity may work differently here, but that finding must be tempered by the wide margins of error for these groups.
+
+    -   It is possible that some of the lower risk for biracials reflects their younger average age.
+
+    -   Grade retention is an important early life outcome, but far from the only one.
+It is possible that the context in which it happens, often early in the educational system, does not continue throughout the life course and that we might see very different patterns emerge for other kinds of outcomes.
+More work should be done!
 

### 2022-11-17 products/paper/main.qmd
Add first full rough draft of conclusion

--- 2022-11-16 products/paper/main.qmd
@@
 
--   Big claim: The results are consistent with a new "regime of hyperdescent" operating across all racial groups.
-This contributes to a muddying of racial distinctiveness via hybridity, while at the same time isolating monoracial members of the most stigmatized racial groups, consistent with arguments like Bonilla-Silva's Latin Americanization.
+In this article, I have used the risk of grade retention among K-12 students to better understand how biracial students' life chances compare to their monoracial peers.
+Unlike prior work on this topic, due to the large samples involved, the estimates for most biracial groups are relatively precise.
+For groups with a high level of precision, the results tell a consistent story - biracial students' risk of grade retention is similar to their monoracial constituent group with the lower risk.
+
+These results contradict expectations of both a "halfway" position and the dominance of the one drop rule.
+For no group do I observe clear evidence of a pattern of hypodescent in which the risk was similar to the higher risk constituent monoracial group.
+These results also strongly contradict the expectations of the marginal man hypothesis.
+
+Two possibilities emerge that may help us understand these results better.
+First, we may be witnessing a shift toward a new regime of hyperdescent that applies broadly across a wide variety of racial groups.
+Second, the relatively low risk for biracial students may be a product of the strength of hybridity in allowing biracial students to better negotiate racialized systems than their monoracial minority peers.
+
+Differences in family resources were expected to play some role in the relative position of biracial students' risk.
+The results on parent selection vary somewhat by group, but the majority of biracial groups had resources more similar to their monoracial constituent group with higher resources, indicating a strong positive selection into interracial unions in the prior generation.
+The most important exception to this trend is the case of Black/White biracial students who have far fewer resources than white students and resources closer to black students.
+
+Regardless of variation in biracial students' parental resources, accounting for these resources differentials does not substantially change the overall result that biracial students' risk is more similar to the constituent monoracial group with lower risk.
+The one exception to this finding is for the Latino/Asian case where the much lower risk of these students compared to Latino and Asian students was entirely driven by resource differentials.
+
+The relative risk for biracial individuals is sometimes incorrectly treated as an indication of a positive future direction for the US in terms of ameliorating racial inequality.
+The improved prospects for biracial individuals do nothing to ameliorate the often strong and persistent inequalities between monoracial groups.
+The growth and relative success of biracial populations may isolate remaining members of monoracial minority groups as much as it assimilates mixed race individuals.
+The real question at stake is how existing divides and identities may be restructured in the context of a growing biracial population.
+This research suggests the potential for a growing divide between these mixed race populations and the monoracial minority populations from which they derive at least part of their ancestry.
+
+Because these findings rely upon large samples, they present an important step forward in our understanding of how biracial individuals will fit into America's system of racial inequality.
+However, this study also has important limitations that need to be noted.
+
+Although sample sizes are big generally, they remain quite small for some groups of non-white biracial respondents, limiting our understanding of the outcomes for those groups.
+In particular, results for non-white part-Indigenous populations are quite imprecise.
+The results interestingly and counterintuitively suggest that part-Indigenous populations are the least likely to be in the position expected by hyperdescent, but due to low sample sizes, that finding is highly tentative.
+
+Wile grade retention is an important early life outcome, it is far from the only one that these individuals will face in their life.
+Although, we often observe similar racial inequalities across a variety of outcomes, knowing the result for one outcome does not perfectly predict other outcomes.
+As biracial individuals age and face later outcomes in the educational system, the labor market, and elsewhere, their experiences may differ from what I observe here.
+Thus, collecting more detailed information on these outcomes in future studies remains important.
 

### 2022-11-17 products/paper/main.qmd
Add concluding paragraph to introduction

@@
 
-The results provide a window into the placement of biracial children within an existing system of racial inequality in our educational systems. <!-- TODO  PITHY SUMMARY -->
+The results provide a window into the placement of biracial children within an existing system of racial inequality in our educational systems.
+Given the large sample size I am able to draw, I am able to estimate relatively precise estimates for most biracial groups.
+In contrast to prevailing hypothesis predicting either a pattern of hypodescent or an in-between position for biracial individuals, I find a consistent pattern in which the risk of grade retention is similar to that of the lower risk constituent monoracial group.
+This result holds even after adjusting for differences in resources across respondents, despite the fact that most biracial groups are positively selected by parental resources relative to their monoracial group.
+I discuss the implications for these findings in the conclusions.
 

### 2022-12-06 products/paper/main.qmd
Final copy edit

@@
 First, using self identification of biracial respondents to analyze outcomes is problematic.
-Research on the identification and classification of biracial individuals indicates that individuals whose parents belong to different races do not always consistently identify with both races [@doyle_are_2007; @liebler_america_2017; @gullickson_racial_2019].
+Prior research indicates that individuals whose parents belong to different races do not always consistently identify with both races [@doyle_are_2007; @liebler_america_2017; @gullickson_racial_2019].
 Furthermore, social status and social mobility can predict racial identification generally [@saperstein_racial_2012; @saperstein_mulatto_2013] and social status is related to the identification choices of biracial individuals specifically [@roth_end_2005a; @davenport_role_2016].
@@
 Researchers have previously utilized this feature to examine the classification choices made on behalf of these children [@xie_racial_1997; @roth_end_2005a].
-Analyzing the outcomes for such children is more difficult, for the simple reason that they do not have many measurable outcomes yet.
-
+Analyzing outcomes for such children is more difficult, for the simple reason that they do not have many measurable outcomes yet.
 However, one important outcome, grade retention, is measurable and important even for young children.
@@
 Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience.
+
 In this article, I use data from the American Community Survey to examine the likelihood of being held back in school for biracial children in comparison to their monoracial peers.
 Pooling data from 2010 to 2019, I am able to draw a large enough sample to make reliable statistical inferences for most biracial groups, including those with two non-white parents.
-
-The results provide a window into the placement of biracial children within an existing system of racial inequality in our educational systems.
+The results provide a window into the placement of biracial children within an existing system of racial inequality.
 Given the large sample size I am able to draw, I am able to estimate relatively precise estimates for most biracial groups.
-In contrast to prevailing hypothesis predicting either a pattern of hypodescent or an in-between position for biracial individuals, I find a consistent pattern in which the risk of grade retention is similar to that of the lower risk constituent monoracial group.
+In contrast to prevailing hypotheses that predict either a pattern of hypodescent or an in-between position for biracial individuals, I find a consistent pattern in which the risk of grade retention is similar to that of the lower risk constituent monoracial group.
 This result holds even after adjusting for differences in resources across respondents, despite the fact that most biracial groups are positively selected by parental resources relative to their monoracial group.
@@
 
-I distinguish three mechanisms that help us understand where biracials might be positioned within a system of racial inequality.
-First, I consider arguments for the racial hierarchy that focus on the effects of racial discrimination at both an individual and systemic level.
+I distinguish three mechanisms that help us understand where biracial individuals might be positioned among these possibilities.
+First, I consider arguments that focus on the effects of racial discrimination at both an individual and systemic level.
 Second, I consider arguments regarding different distributions of resources across racial groups.
@@
 Discrimination and bias by race play an important role in overall levels of racial inequality [@quillian_new_2006; @pager_sociology_2008].
-Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unknowningly rely upon and reinforce in social interactions [@quillian_new_2006].
+Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unknowingly rely upon and reinforce in social interactions [@quillian_new_2006].
 Such discrimination will be most consequential when performed by important gatekeepers of opportunity such as teachers, health care workers, and employers.
@@
 
-The strongest historical precedent for understanding such classification in the US is the "one drop rule" norm concerning Black ancestry.
+The strongest historical precedent for understanding such classification in the US is the "one drop rule" norm governing classification as Black.
 According to the one-drop rule, individuals of known Black ancestry are considered exclusively Black, regardless of other ancestries [@davis_who_1991].
@@
 In this case, neither the hypodescent/hyperdescent or dominance paradigm may help us to fully understand the experiences of those of Latino and Asian ancestry due to more recent processes of racialization.
-Second, the historical norms for all groups may also be breaking up or transforming in the post-Civil Rights era.
+Furthermore, the historical norms for all groups may be weakening or transforming in the post-Civil Rights era.
 
@@
 How might classification and discrimination produce such an in-between status?
-There are several possibilities here.
+Several scenarios might produce such an in-between status.
 
@@
 In other words, biracial people's experience of discrimination may be "softened" because their ambiguous physical appearance provokes less racial antipathy.
-The US in fact already has a long history of this kind of "softening." Because of the one-drop rule, Black individuals vary substantially in skin tone and those individuals with a lighter skin tone have better outcomes [@keith_skin_1991; @hughes_significance_1990; @monk_skin_2014] and are viewed more favorably by both Blacks and Whites [@maddox_cognitive_2002a].
+The US already has a long history of this kind of softening.
+Because of the one-drop rule, Black individuals vary substantially in skin tone and those individuals with a lighter skin tone have better outcomes [@keith_skin_1991; @hughes_significance_1990; @monk_skin_2014] and are viewed more favorably by both Blacks and Whites [@maddox_cognitive_2002a].
 Thus, light skin moderates the stigma of Blackness even among monoracial individuals.
@@
 How do these differential distributions of resources affect biracial individuals?
-A naive argument would be that because biracial individuals have parents from two different racial groups, they will on average inherit parental resources roughly halfway in-between their two constituent groups [@chew_american_1989].
+A naive argument would be that because biracial individuals have parents from two different racial groups, they will on average inherit parental resources roughly halfway between their two constituent groups [@chew_american_1989].
 Prior research looking at biracial children in the Census data has found some evidence to support this argument [@chew_american_1989].
@@
 
-Regardless of its overall prevalence, the net effect of status exchange on resources is complex because status exchange positively selects one parent and negatively selects the other partner.
+Regardless of its overall prevalence, the net effect of status exchange on resources is complex because status exchange positively selects one partner and negatively selects the other partner.
 Therefore the two effects may somewhat cancel each other out in terms of the overall distribution of resources across parents.
@@
 Regardless of the underlying complexity of this selection process, delineating its role as opposed to other mechanisms is key to understanding the overall process that sorts biracial individuals within the racial hierarchy.
-Although this approach is sometimes characterized as pitting race versus class and minimizing the importance of race, that is not case here.
-Instead, what this delineation allows us to do is understand how race affects the outcomes of biracial individuals in a multigenerational process.
+Although this approach is sometimes characterized as pitting race versus class and minimizing the importance of race, that is not the case here.
+Instead, this delineation allows us to understand how race affects the outcomes of biracial individuals in a multigenerational process.
 
@@
 Both of these studies find some support for an "in-between" status for many biracial groups, but are both hampered by a lack of statistical power due to small sample sizes.
-@kao_racial_1999 used data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial students with Black and Asian parentage and finds that biracial Blacks have outcomes similar to monoracial Blacks, while biracial Asians have outcomes more similar to monoracial Whites.
-This study, however, is hampered by the fact that biracial students are identified by a discrepancy between their reported race and one parent's reported race, rather than by both biological parent's races.
+@kao_racial_1999 used data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial students with Black and Asian parentage and finds that biracial Black students have outcomes similar to monoracial Black students, while biracial Asian students have outcomes more similar to monoracial White students.
+This study, however, is hampered by the fact that biracial students are identified by a discrepancy between their reported race and one parent's reported race, rather than by both biological parents' races.
 
@@
 This approach can generate much larger samples of biracial individuals, although it does exclude examination of children living in non-nuclear households.
-Furthermore, researchers then have a limited range of outcomes to explore because biracial individuals identified by this method are by definition children.
+However, researchers then have a limited range of outcomes to explore because biracial individuals identified by this method are by definition children.
 This approach has been used previously to examine the likelihood of living in poverty [@bratter_multiracial_2013; @bratter_poverty_2013] and residential segregation [@ellis_agents_2012].
 These studies provide valuable information on biracial childrens' lived experience, but do not directly measure their own outcomes.
-
-However, one important educational outcome is partially identifiable for children in Census and ACS data - grade retention. I now turn to a closer examination of this particular outcome.
+However, one important educational outcome is partially identifiable for children in Census and ACS data - grade retention.
+I now turn to a closer examination of this particular outcome.
 
@@
 To increase sample size for smaller populations of biracial respondents, I pool ACS data for a full decade from 2010-2019.
-While pooling data helps to increase sample size, it may generate biases if the likelihood of grade retention is changing over time in complex ways.
-I address this issue in the modeling techniques discussed below.
 All data were extracted from the Minnesota Population Center's Integrated Public Use Microdata Series [@ruggles_ipums_2020].
@@
 
-The race of children is calculated by a cross-tabulation of parent's races.
+The race of children is calculated by a cross-tabulation of parents' races.
 A full cross-tabulation of the race/Hispanicity of each parent would produce far too many combined categories to be analytically useful.
@@
                                                      vartype=NULL)), 
-                              accuracy=0.01),
-            mean_age=round(survey_mean(age, vartype=NULL), 1)) %>%
+                              accuracy=0.01)) %>%
   arrange(desc(n)) %>%
@@
                   "Sample size",
-                  "Clearly behind expected grade", 
-                  "Mean age" )) %>%
+                  "Clearly behind expected grade")) %>%
   kable_styling() %>%
@@
 However, the goal of this study is not to estimate grade retention accurately, but rather to understand racial differences in grade retention.
-Because this bias is largely a function of when students have birthdays, it should be more or less random with regard to sociodemographic characteristics.[^1] This approach is equivalent to the method used by @rosenfeld_nontraditional_2010 on 2000 Census data, but more detailed information on current grade in recent ACS data allows for more precise estimation.
+Because this bias is largely a function of when students have birthdays, it should be more or less random with regard to sociodemographic characteristics.
+This approach is equivalent to the method used by @rosenfeld_nontraditional_2010 on 2000 Census data, but more detailed information on current grade in recent ACS data allows for more precise estimation.
 @tbl-descriptive shows the percent of each monoracial and biracial group that is CBEG.
-
-[^1]: Biracial respondents are slightly younger than monoracial respondents in the sample as a whole because interracial unions are becoming more common over time which may contribute to a slight downward bias in their risk of grade retention.
 
@@
 Throughout this article, I use a visual approach to illustrate this placement.
-As an example, I display the percent of students clearly behind expected grade for White, Black, and Black/White students in @fig-example.
+As an example, I display the percent of White, Black, and Black/White students clearly behind expected grade in @fig-example.
 I display confidence bands/bars around all three point estimates, but these confidence ranges require some explanation as they are not standard 95% confidence intervals.
@@
 
-I begin by showing the relative risk of being CBEG among monoracial respondents across models.
+I begin by showing the relative risk of being CBEG among monoracial respondents across models.Understanding these differences helps to clarify the potential for how biracial respondents might be positioned between monoracial groups.
 Table 2 reports average marginal effects on the probability of being CBEG for each monoracial minority group in comparison to White students.
-Understanding these differences helps to clarify the potential for how biracial respondents might be positioned between monoracial groups.
 
@@
 
-@tbl-resources shows the mean values of the resource variables across monoracial and biracial groups.[^7] Because of the number of variables and groups involved, this table can be overwhelming to read.
-Therefore, I also provide a summation of these resource differentials by racial group by calculating a counterfactual probability of CBEG for each group based on their observed distribution of resource characteristics.
+@tbl-resources shows the mean values of the resource variables across monoracial and biracial groups.[^7] I also calculate a counterfactual probability of CBEG for each racial group based solely on their observed distribution of resources.
+Differences in this counterfactual probability provides a summary measure of the resource differentials between groups.
 This counterfactual probability is calculated by estimating the average predicted probability for each group from the fullest model (Model 4 of Table 2) when all non-resource variables, including race, are held at their mean.[^8]
 
-[^7]: For compactness, I combine the parent foreign-born and English proficiency questions into whether either parent is foreign-born or speak English.
+[^7]: For compactness, I combine the parent foreign-born and English proficiency questions into whether either parent is foreign-born or speaks English.
 In the models, I use separate variables for mothers and fathers.
@@
 
-My particular interest, however, is in how each biracial group compares to its constituent monoracial groups.
+I am particularly interested in how each biracial group compares to its constituent monoracial groups.
 For example, while Indigenous/Latino students have the highest counterfactual risk of being CBEG among biracial students, their risk is still lower than their two constituent groups of Indigenous (4.5%) and Latino (5.3%) students, owing in large part to somewhat higher family income, greater parental educational attainment, and greater acculturation.
@@
 Unlike most of these other groups, there does not seem to be substantial positive class selectivity into Black/White interracial unions.
+
 The effect of controlling for these resources will be complex and different across biracial groups.
-For those biracial groups with resources similar to their more advantaged constituent group, we should see that controlling for these resources should reduce their advantage relative to the less advantaged monoracial group.
-For the Black/White case, on the other hand, we would expect that controlling for resource differentials should eliminate some of their disadvantage relative to Whites.
+For those biracial groups with resources similar to their more advantaged constituent group, controlling for these resources should reduce their advantage relative to the less advantaged monoracial group.
+For the Black/White case, on the other hand, controlling for resource differentials should eliminate some of their disadvantage relative to Whites.
 For the remaining "in-between" groups, it is more difficult to make a priori predictions.
@@
 #| fig-cap: Probability of being clearly behind expected grade for biracial respondents with one black parent, in comparison to their monoracial comparison groups.
-.
 Non-overlap in color corresponding confidence bands indicates statistically significant difference at the 5% level, Baseline model includes year, grade, location, and state fixed effects.
@@
 
-Regardless of variation in biracial students' parental resources, accounting for these resources differentials does not substantially change the overall result that biracial students' risk is more similar to the constituent monoracial group with lower risk.
+Regardless of variation in biracial students' parental resources, accounting for these resource differentials does not substantially change the overall result that biracial students' risk is more similar to the constituent monoracial group with lower risk.
 The one exception to this finding is for the Latino/Asian case where the much lower risk of these students compared to Latino and Asian students was entirely driven by resource differentials.
 
-The relative risk for biracial individuals is sometimes incorrectly treated as an indication of a positive future direction for the US in terms of ameliorating racial inequality.
-The improved prospects for biracial individuals do nothing to ameliorate the often strong and persistent inequalities between monoracial groups.
+Low relative risk for biracial individuals is sometimes treated as an indication of a positive future direction for the US in terms of ameliorating racial inequality.
+However, the improved prospects for biracial individuals do nothing to ameliorate the often strong and persistent inequalities between monoracial groups.
 The growth and relative success of biracial populations may isolate remaining members of monoracial minority groups as much as it assimilates mixed race individuals.
 The real question at stake is how existing divides and identities may be restructured in the context of a growing biracial population.
-This research suggests the potential for a growing divide between these mixed race populations and the monoracial minority populations from which they derive at least part of their ancestry.
+This research suggests the potential for a growing divide between these mixed race populations and the most disadvantaged monoracial minority populations from which they derive at least part of their ancestry.
 
@@
 
-Although sample sizes are big generally, they remain quite small for some groups of non-white biracial respondents, limiting our understanding of the outcomes for those groups.
+Sample sizes for some non-white biracial respondents remain small, limiting our understanding of the outcomes for those groups.
 In particular, results for non-white part-Indigenous populations are quite imprecise.
@@
 As biracial individuals age and face later outcomes in the educational system, the labor market, and elsewhere, their experiences may differ from what I observe here.
-Thus, collecting more detailed information on these outcomes in future studies remains important.
+Thus, collecting more detailed information on these outcomes in future studies remains critically important.
 

### 2022-12-06 products/paper/main.qmd
Reduce word count of manuscript
If submitting to SRE, it needs to be under 10K words total, including abstract, tables, figures, footnotes, and references. it should now be about 100 words below that.

@@
 ---
-title: Grade Retention and Mixed Race
+title: Differences in the Risk of Grade Retention for Biracial and Monoracial Students in the US, 2010-2019
 author: 
   - Aaron Gullickson, *University of Oregon*
-abstract: This is a test abstract
-thanks: The corresponding author can be reached at aarong@uoregon.edu. All code and data for this project are available at ...
+abstract: test abstract
+thanks: The corresponding author can be reached at aarong@uoregon.edu.
+Supplementary materials provided with this article include full model results upon which figures are based as well as any senstivity analysis described herein.
+All code and data for this project are available at ...
 format:
@@
   error: false
+prefer-html: true
 editor: source
+csl: american-sociological-association.csl
 ---
@@
 The experiences of mixed race individuals will play an important role in the future of race in the US.
-A rise in interracial marriage and coupling since the 1960s has resulted in a "biracial baby boom" of individuals whose parents cross racial lines [@root_racially_1992] and this demographic surge may have a transformative effect on racial boundaries [@hochschild_creating_2012; @alba_great_2020].
-While racial mixing is nothing new to US history [@morning_who_2000; @gullickson_choosing_2011], this biracial boom is notable for its demographic size and has occurred in a post-Civil Rights environment in which historical norms of racial classification may be eroding.
+A rise in interracial unions since the 1960s has resulted in a "biracial baby boom" of individuals whose parents cross racial lines [@root_racially_1992] and this demographic surge may have transformative effects on racial boundaries [@hochschild_creating_2012; @alba_great_2020].
+While racial mixing is nothing new to US history [@morning_who_2000; @gullickson_choosing_2011], this biracial boom is notable for its size and has occurred in an environment in which historical norms of racial classification may be eroding.
 
-Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of both how these biracial individuals identify their own race and how they are sorted into a prevailing system of racial inequality.
+The transformative potential of this biracial baby boom is contingent on both how these biracial individuals identify their own race and how they are sorted into a prevailing system of racial inequality.
 While extensive scholarship documents the identification and classification decisions made by or on behalf of multiracial individuals [@xie_racial_1997; @qian_options_2004; @roth_end_2005a; @brunsma_interracial_2005; @herman_forced_2004; @holloway_place_2009; @bratter_mother_2009; @davenport_role_2016; @liebler_boundaries_2016], only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers [@kao_racial_1999; @campbell_multiracial_2009; @herman_blackwhiteother_2009].
@@
 First, using self identification of biracial respondents to analyze outcomes is problematic.
-Prior research indicates that individuals whose parents belong to different races do not always consistently identify with both races [@doyle_are_2007; @liebler_america_2017; @gullickson_racial_2019].
+Prior research indicates that individuals whose parents belong to different races do not consistently identify with both races [@doyle_are_2007; @liebler_america_2017; @gullickson_racial_2019].
 Furthermore, social status and social mobility can predict racial identification generally [@saperstein_racial_2012; @saperstein_mulatto_2013] and social status is related to the identification choices of biracial individuals specifically [@roth_end_2005a; @davenport_role_2016].
@@
 
-This problem of reverse causation can be largely addressed when researchers have access to the racial identification of a respondent's biological parents.
-While such information is available on some large scale longitudinal surveys, researchers then run into a second limitation of sample size.
-The samples of biracial respondents in such data sources are generally far too small to make reliable statistical inferences.
+This problem of reverse causation can be eliminated when researchers have access to the racial identification of a respondent's biological parents.
+Even when such information is available, however, researchers run into a second limitation of sample size.
+The samples of biracial respondents in many data sources are too small to make reliable statistical inferences.
 This limitation is particularly severe for biracial individuals with two non-white parents.
 
-When children reside with both biological parents, biracial individuals can be identified in large scale data sources such as the Census.
+Biracial children who reside with both biological parents can be identified in large scale data sources such as the Census.
 Researchers have previously utilized this feature to examine the classification choices made on behalf of these children [@xie_racial_1997; @roth_end_2005a].
 Analyzing outcomes for such children is more difficult, for the simple reason that they do not have many measurable outcomes yet.
-However, one important outcome, grade retention, is measurable and important even for young children.
+However, grade retention is measurable and important even for young children.
 Being held back in school is a relatively common occurrence that has consequences for later educational transitions [@jimerson_metaanalysis_2001] and is disproportionately applied by race [@warren_patterns_2014].
-Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience.
+Grade retention is one of the earliest stratifiers of life chances that an individual will experience.
 
@@
 The results provide a window into the placement of biracial children within an existing system of racial inequality.
-Given the large sample size I am able to draw, I am able to estimate relatively precise estimates for most biracial groups.
 In contrast to prevailing hypotheses that predict either a pattern of hypodescent or an in-between position for biracial individuals, I find a consistent pattern in which the risk of grade retention is similar to that of the lower risk constituent monoracial group.
 This result holds even after adjusting for differences in resources across respondents, despite the fact that most biracial groups are positively selected by parental resources relative to their monoracial group.
-I discuss the implications for these findings in the conclusions.
 
@@
 I distinguish three mechanisms that help us understand where biracial individuals might be positioned among these possibilities.
-First, I consider arguments that focus on the effects of racial discrimination at both an individual and systemic level.
+First, I consider arguments regarding racial discrimination at both an individual and systemic level.
 Second, I consider arguments regarding different distributions of resources across racial groups.
@@
 Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unknowingly rely upon and reinforce in social interactions [@quillian_new_2006].
-Such discrimination will be most consequential when performed by important gatekeepers of opportunity such as teachers, health care workers, and employers.
-Discrimination need not just occur on an individual basis, but can also be systemic in nature when biases are built into and reinforced by organizations and institutional systems [@reskin_race_2012].
+Discrimination can also be systemic in nature when biases are built into and reinforced by organizations and institutional systems [@reskin_race_2012].
 In our current "colorblind" regime [@bonilla-silva_racism_2006], such systems may be formally race neutral but produce disparate impacts by race [@pager_sociology_2008].
@@
 
-The strongest historical precedent for understanding such classification in the US is the "one drop rule" norm governing classification as Black.
+The strongest historical precedent for understanding such classification in the US is the "one-drop rule" norm governing classification as Black.
 According to the one-drop rule, individuals of known Black ancestry are considered exclusively Black, regardless of other ancestries [@davis_who_1991].
 @davis_who_1991 generalizes this practice into the concept of *hypodescent* in which individuals with mixed racial ancestry will be classified as members of the lower status group.
-By comparison a pattern of *hyperdescent* implies that individuals of mixed racial ancestry will be classified as members of the higher status group.
-This pattern of hyperdescent has historically governed the experiences of individuals with American Indian ancestry [@snipp_american_1989; @wolfe_land_2001; @gullickson_choosing_2011].
+A parallel pattern of *hyperdescent* implies that individuals of mixed racial ancestry will be classified as members of the higher status group.
+This pattern of hyperdescent has historically governed the experiences of individuals with American Indian ancestry [@snipp_american_1989; @wolfe_land_2001].
 
-Although the hypodescent/hyperdescent paradigm is often used to theoretically frame the possible contemporary experiences of multiracial individuals, there are reasons to be cautious about the generalizability of this paradigm.
+Although the hypodescent/hyperdescent paradigm is often used to frame the experiences of multiracial individuals, its generalizability and contemporary relevance is questionable.
 Hypodescent and hyperdescent are observed in two specific historical cases involving Black ancestry and American Indian ancestry, respectively.
 As @iverson_regimes_2022 note, these two cases could just as parsimoniously be explained by a "dominance" model which ranks ancestries by their tendencies to be supercessive or recessive in determining identification and classification.
-In this framework, within the historical "racial triangle" of American colonialism, Black ancestry is supercessive, Indigenous ancestry is recessive, and White ancestry falls somewhere in between.
 
-The development of these particular norms may also reflect the historical regime in which they were developed [@gullickson_choosing_2011; @iverson_regimes_2022].
+The development of particular norms may also reflect the historical regime in which they were developed [@gullickson_choosing_2011; @iverson_regimes_2022].
 In this case, neither the hypodescent/hyperdescent or dominance paradigm may help us to fully understand the experiences of those of Latino and Asian ancestry due to more recent processes of racialization.
-Furthermore, the historical norms for all groups may be weakening or transforming in the post-Civil Rights era.
+Furthermore, the historical norms for all groups may be transforming in the post-Civil Rights era.
 
@@
 However, what if those outcomes fall in-between the two groups?
-How might classification and discrimination produce such an in-between status?
-Several scenarios might produce such an in-between status.
+Processes of classification and discrimination could still produce an in-between status for several reasons.
 
 First, racial ambiguity in appearance may lead to more inconsistent classification into single race categories across observers.
-Even if norms consistently indicate a certain type of classification, correct identification by observers is still a probabilistic phenomenon in all social interactions and likely to be more heterogeneous for biracial individuals.
+Even if norms consistently indicate a certain type of classification, correct identification by observers is a probabilistic phenomenon and likely to be more heterogeneous for biracial individuals.
 As a result, biracial individuals will experience a lower sum total of discrimination and bias affecting the lower status group in the case of hypodescent and will not reap the same rewards as a member of the higher status group in the case of hyperdescent.
 
-Second, even when "correctly" classified, greater physical ambiguity may allow biracial individuals to avoid the worst instances of bias and discrimination because they do not conform to the ideal type for the group facing the most discrimination.
-In other words, biracial people's experience of discrimination may be "softened" because their ambiguous physical appearance provokes less racial antipathy.
+Second, biracial people's experience of discrimination may be "softened" because a more ambiguous physical appearance provokes less racial antipathy.
 The US already has a long history of this kind of softening.
@@
 
-Third, it is also possible that biracial individuals may be classified into an ambiguous "middleman" minority position.
+Third, biracial individuals may be classified into an ambiguous "middleman" minority position.
 This kind of middle tier status for individuals of mixed race is common in Latin America [@telles_race_2009] and @bonilla-silva_biracial_2004 has argued that such a system may be emerging in the contemporary US as well.
@@
 
-Racial inequality is also at least partly explained by different distributions of material, cultural, and social resources across racial groups [@conley_being_1999].
+Racial inequality is at least partly explained by different distributions of material, cultural, and social resources across racial groups [@conley_being_1999].
 On average, socioeconomic background (e.g.
@@
 Other less materially defined resources also vary across groups such as cultural styles, social networks, and neighborhood quality.
-These differences in background resources reflect the accumulation of historical processes of discrimination and, when combined with contemporary processes of discrimination and bias, magnify the overall level of racial inequality we observe.
+These differences in background resources reflect the accumulation of historical processes of discrimination and, when combined with contemporary processes of discrimination, magnify the overall level of racial inequality we observe.
 
-How do these differential distributions of resources affect biracial individuals?
-A naive argument would be that because biracial individuals have parents from two different racial groups, they will on average inherit parental resources roughly halfway between their two constituent groups [@chew_american_1989].
+How do differences in the distribution of resources affect biracial individuals?
+Naively, we might expect biracial individuals to inherit parental resources roughly halfway between their two constituent groups, on average [@chew_american_1989].
 Prior research looking at biracial children in the Census data has found some evidence to support this argument [@chew_american_1989].
@@
 However, the actual resource position of biracial individuals is likely far more complicated because people are not randomly selected into interracial unions.
-The relationship between this selection process and a biracial child's outcomes are complex because it depends on the strength and direction of that selection on each partner, its consistency across different types of unions, and how strongly the selected characteristic predicts outcomes in the next generation.
+The relationship between this selection process and a biracial child's outcomes depends on the strength and direction of that selection on each partner, its consistency across different types of unions, and how strongly the selected characteristic predicts outcomes in the next generation.
 
@@
 Prior research has shown substantial evidence of this pattern in the case of Black male/White female marriages in the US, but weaker evidence for other types of unions [@gullickson_education_2006; @fu_racial_2001; @kalmijn_educational_2010; @hou_interracial_2011].
-
-Regardless of its overall prevalence, the net effect of status exchange on resources is complex because status exchange positively selects one partner and negatively selects the other partner.
+Even in cases where status exchange is present, its net effect on resources is complex because status exchange positively selects one partner and negatively selects the other partner.
 Therefore the two effects may somewhat cancel each other out in terms of the overall distribution of resources across parents.
-Complicating this issue further, research has shown that the parents of biracial children may differ from parents of monoracial children in the magnitude and types of resources that they allocate to children [@cheng_constraints_2007], even among parents of similar resources.
 
 Regardless of the underlying complexity of this selection process, delineating its role as opposed to other mechanisms is key to understanding the overall process that sorts biracial individuals within the racial hierarchy.
-Although this approach is sometimes characterized as pitting race versus class and minimizing the importance of race, that is not the case here.
+Although this approach is sometimes characterized as pitting race versus class, that is not the case here.
 Instead, this delineation allows us to understand how race affects the outcomes of biracial individuals in a multigenerational process.
@@
 The experience of hybridity itself may also positively or negatively affect outcomes for biracial individuals.
-The "marginal man" theory argues that biracial individuals will find it difficult to fit in due to the dissonance of an identity that crosses strong and salient racial boundaries [@park_human_1928; @stonequist_problem_1935].
-This liminal identity conflict will then result in negative mental health experiences that could feed into other negative outcomes as well.
+According to "marginal man" theory, biracial individuals will find it difficult to fit in due to the dissonance of an identity that crosses strong and salient racial boundaries [@park_human_1928; @stonequist_problem_1935].
+This liminal identity conflict will result in negative mental health experiences that could feed into other negative outcomes as well.
 Evidence for such negative mental health experiences, however, has been mixed [@udry_health_2003; @campbell_what_2006; @cheng_multiracial_2009; @bratter_does_2011].
@@
 Marginal man theorists viewed culture in monolithic and static terms and thus to be trapped in between two cultures was seen as a difficult experience.
-Contemporary views of culture instead emphasize the often flexible, fragmented, and fluid use of cultural bits that comprise a "toolkit" [@swidler_culture_1986; @dimaggio_culture_1997].
-From this perspective, biracial individuals may benefit from an enlarged cultural toolkit that allows them to navigate a variety of diverse social contexts [@shih_multiple_2019].
+Contemporary views of culture instead emphasize the flexible, fragmented, and fluid use of cultural bits that comprise a "toolkit" [@swidler_culture_1986; @dimaggio_culture_1997].
+From this perspective, biracial individuals may actually benefit from an enlarged cultural toolkit that allows them to navigate a variety of diverse social contexts [@shih_multiple_2019].
 
@@
 Different mechanisms may produce the same observed outcome and the weight of different mechanisms can also vary by groups as these groups may be governed by different regimes of classification.
-However, comparing outcomes across biracial groups may help us better understand the likely mechanisms that drive biracial classification and to disregard others as implausible.
+However, comparing outcomes across biracial groups may help us better understand the likely mechanisms involved and to disregard others as implausible.
 Accounting for the effect of parental resources on outcomes to the best of our ability is also key to reaching any conclusions about the placement of biracial individuals within the racial hierarchy.
@@
 With good reason, researchers have been reluctant to use data in which multiracial or biracial status is determined solely by the self-identification of respondents.
-We know that far fewer people identify as biracial than could feasibly do so based on their parents' races <!-- TODO cite morning?
-morning and Goldstein?
-who?
---> and that various selection effects operate on the decision to identify as biracial among those who could [@roth_end_2005a; @davenport_role_2016].
+We know that far fewer people identify as biracial than could feasibly do so based on their parents' races [@morning_generational_2018] and that various selection effects operate on the decision to identify as biracial among those who could [@roth_end_2005a; @davenport_role_2016].
 Furthermore, research on changes in racial reporting over time suggests that other status markers like education and income may influence how people identify themselves [@saperstein_racial_2012; @saperstein_mulatto_2013].
 For example, a person with one Black and one White parent who is routinely treated by others as Black and discriminated against as Black may be more likely to identify themselves as exclusively Black than a similar person of Black and White ancestry who experiences less consistent discrimination.
-Thus, any analysis of outcomes for biracial individuals based on self-identification must deal with a serious potential for reverse causation in the observed differences between biracial individuals and their constituent groups [@bratter_multiracial_2018].
+Thus, any analysis of outcomes for biracial individuals based on self-identification must deal with the potential for reverse causation in the observed differences between biracial individuals and their constituent groups [@bratter_multiracial_2018].
 
-Instead of self-identification, researchers have generally relied upon survey data in which the racial identification of a respondent's biological parents are provided in some form or another.
+Instead of self-identification, researchers have generally relied upon survey data in which the racial identification of a respondent's biological parents are provided.
 @campbell_multiracial_2009 used data from the National Longitudinal Survey of Adolescent to Adult Health (Add Health) to examine high school GPA, advanced placement in math, and four-year college enrollment among biracial and monoracial individuals identified by co-resident biological parents' races.
-@herman_blackwhiteother_2009 examines differences in GPA using data on high school students in California and Wisconsin in which the students reported their parents' races.
-Both of these studies find some support for an "in-between" status for many biracial groups, but are both hampered by a lack of statistical power due to small sample sizes.
+@herman_blackwhiteother_2009 examined differences in GPA using data on high school students in California and Wisconsin in which the students reported their parents' races.
+Both of these studies found some support for an "in-between" status for many biracial groups, but are hampered by a lack of statistical power due to small sample sizes.
 @kao_racial_1999 used data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial students with Black and Asian parentage and finds that biracial Black students have outcomes similar to monoracial Black students, while biracial Asian students have outcomes more similar to monoracial White students.
-This study, however, is hampered by the fact that biracial students are identified by a discrepancy between their reported race and one parent's reported race, rather than by both biological parents' races.
+This study, however, is limited by the fact that biracial students are identified by a discrepancy between their reported race and one parent's reported race, rather than by both biological parents' races.
 
-Results in prior work tend to be imprecise and inconclusive, due to the sample size for most biracial groups.
+Results in prior work tend to be inconclusive, due to the sample size for most biracial groups.
 An alternative approach is to use large scale Census or American Community Survey (ACS) data to identify biracial children by the race of the biological parents in their household.
-This approach can generate much larger samples of biracial individuals, although it does exclude examination of children living in non-nuclear households.
+This approach can generate much larger samples of biracial individuals.
 However, researchers then have a limited range of outcomes to explore because biracial individuals identified by this method are by definition children.
@@
 However, one important educational outcome is partially identifiable for children in Census and ACS data - grade retention.
-I now turn to a closer examination of this particular outcome.
-
-# Grade Retention as an Outcome
 
 Grade retention is the practice of having students repeat a grade due to poor academic performance.
-Rates of grade retention likely increased throughout the late Twentieth century [@frederick_have_2008] but have declined substantially since 2004 [@warren_patterns_2014].
-
 Prior work on grade retention has failed to show a positive effect of grade retention on later educational outcomes, and has instead found evidence of "scarring" effects that lead to later negative educational outcomes, such as a higher dropout risk [@jimerson_metaanalysis_2001; @stearns_staying_2007; @andrew_scarring_2014; @hughes_effect_2017].
@@
 
-Ideally, I would restrict the sample only to children living in a household with two biological parents.
-However, the ACS only records the direct family relationship between the head of household and other members of the family.
-Since 2008, the ACS also distinguishes between biological children, adopted children, and stepchildren.
+Ideally, I would restrict the sample to children living in a household with two biological parents.
+However, while the ACS distinguishes between biological, adopted, and stepchildren, it only records the direct family relationship between the head of household and other members of the household.
 Respondents can therefore be identified as biological children of another member of the household only if that member is the head of household or the biological child of the head of household.
@@
 First, I restrict the sample to cases where both parents were of a reasonable age at the birth of the child (aged 15-44 for mothers and aged 15-60 for fathers).
-Second, I restrict the sample to only those cases where the reported race and Hispanicity of the child is inclusive of at least one of the parents' races.
+Second, I restrict the sample to those cases where the reported race and Hispanicity of the child is inclusive of at least one of the parents' races.
 
@@
 This necessary sample restriction will bias the family structure of the analytical sample relative to the total population of K-12 schoolchildren.
-The result of this bias is well-known.
 Children living in two-parent nuclear households have greater educational success than other children as a result of material and social resources [@biblarz_family_1999].
@@
 The race of children is calculated by a cross-tabulation of parents' races.
-A full cross-tabulation of the race/Hispanicity of each parent would produce far too many combined categories to be analytically useful.
-Therefore, I collapse each parent's race into the categories of White, Black, Asian, Latino, and Indigenous.
+To produce reasonably parsimonious categories, I collapse each parent's race into the categories of White, Black, Asian, Latino, and Indigenous.
 These categories form the "ethnoracial pentagon" that is commonly used in popular practice and government tabulation to identify race in the US [@hollinger_postethnic_1995].
@@
 Because the goal of my analysis is to examine specifically the outcomes of "first-generation" biracial children, I exclude cases where at least one parent identified with multiple races.
-
-The cross-tabulation of parent's races leads to ten distinct biracial categories.
+The cross-tabulation of parents' races leads to ten distinct biracial categories.
 The total sample size for each of these biracial categories is shown in @tbl-descriptive, along with the sample size of the monoracial comparison groups.
-The sample sizes are quite large for most groups and far in excess of comparable categories in other available survey data.
 
@@
 
-Grade retention is famously difficult to measure accurately.
-In most data sources, researchers do not have access to specific reports of grade retention, but rather infer grade retention from a discrepancy between a student's reported age and grade.
+Grade retention is difficult to measure accurately.
+In most data sources, researchers lack specific reports of grade retention and instead infer grade retention from a discrepancy between a student's reported age and grade.
 Prior studies have used the concept of a student being behind modal grade [@bianchi_children_1984; @frederick_have_2008].
-This measure captures the cumulative prevalence of grade retention, but not its incidence.
-It also cannot distinguish cases in which children were delayed in starting kindergarten from grade retention.
 The most notable limitation of this approach is the ambiguity created by the fact that at any given age, a student may reasonably be in two modal grades.
@@
 Throughout this article, I use a visual approach to illustrate this placement.
-As an example, I display the percent of White, Black, and Black/White students clearly behind expected grade in @fig-example.
+As an example, I display the percent of White, Black, and Black/White students who are CBEG in @fig-example.
 I display confidence bands/bars around all three point estimates, but these confidence ranges require some explanation as they are not standard 95% confidence intervals.
@@
 [^2]: The formula for calculating the z-score for the correct interval is given by $$1.96 * \frac{\sqrt{1+\rho^2}}{1+\rho}$$ where $\rho$ is the ratio of the standard errors for the two statistics.
-See @knol_mis_2011 for detailed derivation of this formula.
+See @knol_mis_2011 for a detailed derivation.
 
@@
 These confidence intervals are color-coded in @fig-example to indicate the reference group.
-I also only draw half-intervals in the direction of the monoracial group's point estimate.
+I only draw half-intervals in the direction of the monoracial group's point estimate.
 Overlap in these color corresponding bars and bands indicates that the difference between the biracial group and the monoracial group is not statistically significant at $p<.05$.
@@
 Its standard error is given by $\sqrt{(s_1^2+s_2^2)/2}$ where $s_1$ and $s_2$ are the standard errors of the estimates for the two monoracial groups.
-To aboid clutter, I do not draw a third half-confidence interval on the biracial estimate itself, but this estimate will always be roughly about halfway between the length of the other two intervals.
+To avoid clutter, I do not draw a third half-confidence interval for this point.
+Its length will be roughly halfway between the length of the other two intervals.
 
 @fig-example shows the raw differences between the three racial groups of interest.
-However, in practice, I want to estimate differences across groups while holding constant a variety of potentially confounding variables.
+However, in practice, I want to estimate differences across groups while holding constant a variety of variables.
 To do this, I estimate a set of logit models that predict the likelihood of a student being CBEG by race and a variety of other variables.
@@
 APPs are akin to average marginal effects (AME) and estimated in the same way.
-Whereas AMEs estimate differences or slopes, APPs estimate the level for a given category.[^4]
-
-[^4]: Output from the full models used to construct these APPs is available in the supplementary materials provided with this article.
+Whereas AMEs estimate differences or slopes, APPs estimate the level for a given category.
 
@@
 The probability of being CBEG also increases with the student's current grade, so I include fixed effects for the current grade of the student.
-Additionally, @fig-year-grade shows that the percent of students who are CBEG has declined over time, but this decline has been much more substantial at higher grade levels.
+Additionally, as @fig-year-grade shows, the percent of students who are CBEG has declined over time, but this decline has been much more substantial at higher grade levels.
 For the elementary grades, there is no evidence of a decline at all.
-To account for this grade-specific decline in the models I include an interaction between a linear year term and current grade.[^6]
-
-[^6]: Sensitivity analysis showed that this functional form was preferred by the Bayesian information criterion to models with no interaction and a model with interaction terms that treated year as a categorical variable.
+To account for this grade-specific decline in the models I include an interaction between a linear year term and current grade.
+Sensitivity analysis showed that this functional form was preferred by the Bayesian information criterion to models with no interaction and a model with interaction terms that treated year as a categorical variable.
 I also considered a similar interaction between state and year, but this model was not preferred.
-Model fit statistics for all senstivity analyses are provided in the supplementary materials.
 
@@
 
-All statistics reported here adjust for sampling weights among respondents.
+All statistical analysis adjusts for sampling weights among respondents.
 All models incorporate design effects for variance in sample weights and the clustering of multiple respondents within the same household.
@@
 
-I begin by showing the relative risk of being CBEG among monoracial respondents across models.Understanding these differences helps to clarify the potential for how biracial respondents might be positioned between monoracial groups.
+I begin by showing the relative risk of being CBEG among monoracial respondents across models.
+Understanding these differences helps to clarify the potential for how biracial respondents might be positioned between monoracial groups.
 Table 2 reports average marginal effects on the probability of being CBEG for each monoracial minority group in comparison to White students.
@@
 #| tbl-cap: Average marginal differences in the probability of being behind grade by monoracial group, relative to a monoracial white student.
-texreg(map(results_monoracial[-1], convert_marg),
+wordreg(map(results_monoracial[-1], convert_marg),
        digits=3,
@@
        float.pos="t",
+       file="regtable.doc",
        custom.gof.rows=
@@
 
-In total, resource differentials account for a substantial part of the overall racial differences in the risk of being CBEG, but not all of these differences.
+In total, resource differentials account for a substantial part of the overall racial differences in the risk of being CBEG, but not its entireity.
 Black and Indigenous students remain at higher risk than other students.
-Notably, Asian students rather than White students have the lowest risk of being CBEG in all models.
+Asian students remain the group with lowest risk across all models.
 The most notable shift across models is for Latino students.
@@
 @tbl-resources shows the mean values of the resource variables across monoracial and biracial groups.[^7] I also calculate a counterfactual probability of CBEG for each racial group based solely on their observed distribution of resources.
-Differences in this counterfactual probability provides a summary measure of the resource differentials between groups.
-This counterfactual probability is calculated by estimating the average predicted probability for each group from the fullest model (Model 4 of Table 2) when all non-resource variables, including race, are held at their mean.[^8]
+Differences in this counterfactual probability provide a summary measure of the resource differentials between groups.
+This counterfactual probability is calculated by estimating the average predicted probability for each group from Model 4 of Table 2 when all non-resource variables, including race, are held at their mean.[^8]
 
@@
 The Black/White case stands out as an outlier.
-Black/White students have similar resources as Black students and both groups have fewer resources than White students.
-This is observable in @tbl-resources across family income, home ownership, and parental educational attainment.
-Unlike most of these other groups, there does not seem to be substantial positive class selectivity into Black/White interracial unions.
+Black/White students have similar resources as Black students and both groups have fewer resources than White students, as can be seen in @tbl-resources for family income, home ownership, and parental educational attainment.
+Unlike most of these other groups, I observe no positive class selectivity into Black/White interracial unions.
 
@@
 Solid conclusions are difficult for the remaining two groups because of smaller sample size and correspondingly wider confidence intervals on estimates.
-Nonetheless the point estimates in each of these cases suggest probabilities of being CBEG roughly midway between the two constituent groups.
+Nonetheless, the point estimates in each of these cases suggest probabilities of being CBEG roughly halfway between the two constituent groups.
 
@@
 Black/White biracial students initially have a risk of being CBEG slightly higher than and statistically distinguishable from White students.
-After controlling for cultural and family resources, however, Black/White studentss have a risk of being CBEG that is slightly lower but not statistically distinguishable from White students.
+After controlling for cultural and family resources, however, Black/White students have a risk of being CBEG that is slightly lower but not statistically distinguishable from White students.
 This change reflects the relatively low level of parental education and income among Black/White students, which is much closer to Black students than White students, as shown on @tbl-resources.
@@
 
-Aside from the effect on Latino/Asian students, controlling for resources has moderate effects on the placement of the remaining biracial groups.
+Aside from the case of Latino/Asian students, controlling for resources has moderate effects on the placement of the remaining biracial groups.
 White/Asian students are the only case with a substantial shift, owing to their highly advantaged households.
@@
 In this article, I have used the risk of grade retention among K-12 students to better understand how biracial students' life chances compare to their monoracial peers.
-Unlike prior work on this topic, due to the large samples involved, the estimates for most biracial groups are relatively precise.
+Unlike prior work on this topic, the estimates used here for most biracial groups are relatively precise due to large sample sizes.
 For groups with a high level of precision, the results tell a consistent story - biracial students' risk of grade retention is similar to their monoracial constituent group with the lower risk.
@@
 
-# Acknowledgments {.appendix}
-
-I am grateful to...
-
 # References {.appendix}

### 2022-12-07 products/paper/main.qmd
Add abstract

@@
   - Aaron Gullickson, *University of Oregon*
-abstract: test abstract
+abstract: Understanding how outcomes for biracial individuals compare to their monoracial peers is critical for understanding how patterns of racial inequality in the contemporary United States might be shifting.
+Yet, we know very little about the life chances of biracial individuals due to limitations in most available data sources.
+In this article, I utilize American Community Survey data from 2010-2019 to examine the risk of being clearly behind expected grade among biracial and monoracial K-12 students, helping to fill a gap in our understanding.
+With large sample sizes for most biracial groups, I am able to estimate grade retention risk for biracial students with enough precision to differentiate even modest differences in risk relative to monoracial groups.
+The results indicate that for most biracial groups, biracial students have risk similar to their monoracial constituent group with lower risk.
+Although biracial students tend to have favorable family resource characteristics, controlling for these characteristics does little to change the overall placement of their outcomes.
 thanks: The corresponding author can be reached at aarong@uoregon.edu.

### 2022-12-07 products/paper/main.qmd
Include a few changes based on final manuscript preparation

@@
 thanks: The corresponding author can be reached at aarong@uoregon.edu.
-Supplementary materials provided with this article include full model results upon which figures are based as well as any senstivity analysis described herein.
-All code and data for this project are available at ...
+Supplementary materials provided with this article include full model results upon which figures are based as well as any sensitivity analysis described herein.
+All code and data for this project are available at [https://osf.io/4fevt/?view_only=4abc6d86595c4313a8d4792471e9bc0d](https://osf.io/4fevt/?view_only=4abc6d86595c4313a8d4792471e9bc0d).
 format:
@@
 #| label: tbl-monoracial
-#| tbl-cap: Average marginal differences in the probability of being behind grade by monoracial group, relative to a monoracial white student.
-wordreg(map(results_monoracial[-1], convert_marg),
+#| tbl-cap: Average marginal differences in the probability of being clearly behind expected grade by monoracial group, relative to a monoracial white student
+texreg(map(results_monoracial[-1], convert_marg),
+#wordreg(map(results_monoracial[-1], convert_marg),
        digits=3,
-       caption = "Average marginal differences in the probability of being clearly behind grade by monoracial group, relative to a monoracial white student",
+       caption = "Average marginal differences in the probability of being clearly behind expected grade by monoracial group, relative to a monoracial white student",
        caption.above=TRUE,
@@
        float.pos="t",
-       file="regtable.doc",
+#       file="regtable.doc",
        custom.gof.rows=
@@
       align = c("l",rep("r", 5)),
-      col.names=c("Biacial","Lower than both","Same as lower","Between",
-                  "Same as higher","Higher than both")) %>%
+      col.names=c("Biacial","Lower probability than both","Similar probability as lower","Between",
+                  "Similar probability as higher","Higher probability than both")) %>%
   kable_styling() %>%

### 2022-12-07 products/paper/main.qmd
Use correct quarto code chunk headers in paper

@@
 
-```{r setup}
+```{r}
+#: label: setup
 #| include: false
@@
 ```{r load-data}
+#| label: load-data
+#| include: false
 load(here("analysis","output","acs.RData"))

### 2022-12-07 products/paper/supp_materials.qmd
Add supplementary materials quarto doc

---
title: "Supplementary Materials for Differences in the Risk of Grade Retention for Biracial and Monoracial Students in the US, 2010-2019"
format:
  pdf:
    documentclass: scrartcl
    include-in-header: 
      text: |
        \usepackage{longtable}
    papersize: letter
    fig-height: 4
    fig-pos: h
    keep-tex: false
  docx:
    toc: false
bibliography: ../project.bib
execute: 
  eval: true
  echo: false
  warning: false
  error: false
prefer-html: true
editor: source
csl: american-sociological-association.csl
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("analysis","check_packages.R"))
source(here("analysis","useful_functions.R"))
load(here("analysis","output","bic_spec_tests.RData"))
load(here("analysis","output","model_summaries.RData"))
load(here("analysis","output","acs.RData"))
```

```{r}
#| label: func-convert-model
#| include: false

#  because I only saved model summaries I need a function to convert this to 
# something that knitreg likes
convert_model <- function(model) {
  tr <- createTexreg(
    coef.names = rownames(model$coefficients),
    coef = model$coefficients[,"Estimate"], 
    se = model$coefficients[,"Std. Error"], 
    pvalues = model$coefficients[,"Pr(>|t|)"],
    gof.names = c("N","Null deviance","Model deviance", "Pseudo-R2"), 
    gof = c(model$df.null+1, model$null.deviance, model$deviance, 
            (model$null.deviance-model$deviance)/model$null.deviance), 
    gof.decimal = c(F,F,F,T)
  )
}
```

```{r}
#| label: create-design
acs_design <- acs %>%
  as_survey_design(ids = cluster, weights = perwt)
```

```{r}
#| results: asis
#| label: tbl-full-models
#| tbl-cap: Log-odds ratios from full models predicting a student being clearly behind expected grade. All models used sample weights and adjust for design effects from sample weight variance and clustering of students within the same household.
texreg(lapply(list(model0_summary, model1_summary, model2_summary, 
                   model3_summary, model4_summary), convert_model),
       digits = 3,
       longtable = TRUE,
       use.packages = FALSE,
       caption.above = TRUE,
       custom.coef.names = 
         c("Intercept","Black","Indigenous","Asian","Latino","Black/White",
           "Black/Indigenous","Black/Latino", "Black/Asian",
           "White/Indigenous","White/Latino","White/Asian",
           "Indigenous/Latino","Indigenous/Asian","Latino/Asian",
           "Year (2010 origin)", "1st grade","2nd grade","3rd grade",
           "4th grade","5th grade","6th grade","7th grade","8th grade",
           "9th grade","10th grade","11th grade","12th grade","Alaska",
           "Arizona","Arkansas","California","Colorado","Connecticut",
           "Washington DC","Delaware","Florida","Georgia","Hawaii","Idaho",
           "Illinois","Indiana","Iowa","Kansas","Kentucky","Louisiana",
           "Maine","Maryland","Massachusetts","Michigan","Minnesota",
           "Mississippi","Missouri","Montana","Nebraska","Nevada",
           "New Hampshire","New Jersey","New Mexico","New York",
           "North Carolina","North Dakota","Ohio","Oklahoma","Oregon",
           "Pennsylvania","Rhode Island","South Carolina","South Dakota",
           "Tennessee","Texas","Utah","Vermont","Virginia","Washington",
           "West Virginia","Wisconsin","Wyoming","Non-metro","Central city",
           "Metro non-central","Metro indeterminable","Female",
           "Year x 1st grade","Year x 2nd grade","Year x 3rd grade",
           "Year x 4th grade","Year x 5th grade","Year x 6th grade",
           "Year x 7th grade","Year x 8th grade","Year x 9th grade",
           "Year x 10th grade","Year x 11th grade","Year x 12th grade",
           "Foreign-born","Mother foreign-born","Father foreign-born",
           "Speaks English somewhat", "Speaks English well", 
           "Mother speaks somewhat", "Mother speaks well",
           "Father speaks somewhat", "Father speaks well",
           "Mother High school diploma", "Mother AA degree", 
           "Mother BA degree","Mother Grad degree",
           "Father High school diploma", "Father AA degree", 
           "Father BA degree","Father Grad degree",
           "Family income (10K USD, sq. root)", "Owns home",
           "parents married"),
       groups = list("Race (ref. White)"=2:15, 
                     "Grade (ref. Kindergarten)"=17:28,
                     "State (ref. Alabama)"=29:78,
                     "Location (ref. Indeterminable)"=79:82,
                     "English proficiency (ref. none)"=99:100,
                     "Mother's English proficiency (ref. none)"=101:102,
                     "Father's English proficiency (ref. none)"=103:104,
                     "Mother's highest degree (ref. none)"=105:108,
                     "Father's highest degree (ref. none)"=109:112))
```

```{r tbl-bic}
#| label: tbl-bic
#| tbl-cap: Bayesian Information Criterion (BIC) statistics for models accounting for change over time using different specifications. All models also include categorical dummy variables for race, grade, state, and location.
bic_scores$model[1] <- "Categorical year dummies, no interaction"
bic_scores$model[2] <- "Categorical year dummies, year * grade interaction"
bic_scores$model[3] <- "Continous year variable, no interaction"
bic_scores$model[4] <- "Continous year variable, year * grade interaction"
bic_scores$model[5] <- "Continous year variable, year * grade interaction + year * state interaction"

bic_scores <- bic_scores[c(1,3,2,4,5),]

bic_scores %>%
  kbl(booktabs = TRUE,
      linesep = NULL,
      align = c("l","r"),
      col.names=c("Model", "BIC Score")) %>%
  kable_styling()
```

```{r desc-stats-ses}
#| label: tbl-desc-stats-ses
#| tbl-cap: Socioeconomic resources by race. Shading indicates biracial group. All results apply survey weights.

tbl_summary <- acs_design %>%
  group_by(race) %>%
  summarize(fam_inc=survey_mean(family_income, vartype = NULL))

tbl_summary <- acs_design %>%
  group_by(race) %>%
  summarize(fam_inc=round(survey_mean(family_income, vartype = NULL), 0),
            own_home=round(survey_mean(own_home, vartype = NULL), 3)*100,
            par_mar=round(survey_mean(parents_married, vartype = NULL), 3)*100,
            f_lhs=round(survey_mean(degree_father=="LHS", vartype = NULL), 
                        3)*100,
            f_hs=round(survey_mean(degree_father=="HS", vartype = NULL), 3)*100,
            f_aa=round(survey_mean(degree_father=="AA", vartype = NULL), 3)*100,
            f_ba=round(survey_mean(degree_father=="BA" | degree_father=="G", 
                                   vartype = NULL), 3)*100,
            m_lhs=round(survey_mean(degree_mother=="LHS", vartype = NULL), 3)*100,
            m_hs=round(survey_mean(degree_mother=="HS", vartype = NULL), 3)*100,
            m_aa=round(survey_mean(degree_mother=="AA", vartype = NULL), 3)*100,
            m_ba=round(survey_mean(degree_mother=="BA" | degree_mother=="G", 
                                   vartype = NULL), 3)*100)

index_biracial <- which(str_detect(tbl_summary$race, "/"))

tbl_summary %>%
  mutate_if(is.numeric, format, big.mark=",") %>%
  kbl(booktabs=TRUE, linesep=NULL,
      align = c("l",rep("r", ncol(tbl_summary)-1)),
      col.names=c("race","mean family income","% own home",
                  "% parents married",
                  "% father, no HS diploma","% father, HS diploma",
                  "% father, AA degree",  "% father, BA+ degree",
                  "% mother, no HS diploma","% mother, HS diploma", 
                  "%  mother, AA degree", "% mother, BA+ degree")) %>%
  kable_styling(font_size=8) %>%
  column_spec(2:12, width = "0.75cm") %>%
  row_spec(index_biracial, background = "#D3D3D3")
```

```{r}
#| label: tbl-desc-stats-cult
#| tbl-cap: Cuiltural resources by race. Shading indicates biracial group. All results apply survey weights.
tbl_summary <- acs_design %>%
  group_by(race) %>%
  summarize(f_born=round(survey_mean(foreign_born, vartype = NULL), 3)*100,
            father_fb=round(survey_mean(foreign_born_father, 
                                        vartype = NULL), 3)*100,
            mother_fb=round(survey_mean(foreign_born_mother, 
                                        vartype = NULL), 3)*100,
            spk_eng_no=round(survey_mean(spk_eng=="No", vartype = NULL), 3)*100,
            spk_eng_some=round(survey_mean(spk_eng=="Somewhat", 
                                           vartype = NULL), 3)*100,
            spk_eng_yes=round(survey_mean(spk_eng=="Yes", vartype = NULL), 3)*100,
            f_spk_eng_no=round(survey_mean(spk_eng_father=="No", 
                                           vartype = NULL), 3)*100,
            f_spk_eng_some=round(survey_mean(spk_eng_father=="Somewhat", 
                                             vartype = NULL), 3)*100,
            f_spk_eng_yes=round(survey_mean(spk_eng_father=="Yes", 
                                            vartype = NULL), 3)*100,
            m_spk_eng_no=round(survey_mean(spk_eng_mother=="No", 
                                           vartype = NULL), 3)*100,
            m_spk_eng_some=round(survey_mean(spk_eng_mother=="Somewhat", 
                                             vartype = NULL), 3)*100,
            m_spk_eng_yes=round(survey_mean(spk_eng_mother=="Yes", 
                                            vartype = NULL), 3)*100)

index_biracial <- which(str_detect(tbl_summary$race, "/"))

tbl_summary %>%
  mutate_if(is.numeric, format, big.mark=",") %>%
  kbl(booktabs=TRUE, linesep=NULL,
      align = c("l",rep("r", ncol(tbl_summary)-1)),
      col.names=c("race",
                  "% foreign-born","% father foreign-born",
                  "% mother foreign-born",
                  "% speak no English", "% speak some English", 
                  "% speak English",
                  "% father speak no English", "% father speak some English", 
                  "% father speak English",
                  "% mother speak no English", "% mother speak some English",
                  "% mother speak English")) %>%
  kable_styling(font_size=8) %>%
  column_spec(2:13, width = "0.75cm") %>%
  row_spec(index_biracial, background = "#D3D3D3")
```

### 2022-12-07 products/paper/supp_materials.qmd
Correct typo

@@
 All models used sample weights and adjust for design effects from sample weight variance and clustering of students within the same household.
-texreg(lapply(list(model0_summary, model1_summary, model2_summary, 
+wordreg(lapply(list(model0_summary, model1_summary, model2_summary, 
                    model3_summary, model4_summary), convert_model),
@@
        caption.above = TRUE,
+       file="full_reg.doc",
        custom.coef.names = 
@@
 #| label: tbl-desc-stats-cult
-#| tbl-cap: Cuiltural resources by race. Shading indicates biracial group. All results apply survey weights.
+#| tbl-cap: Cultural resources by race. Shading indicates biracial group. All results apply survey weights.
 tbl_summary <- acs_design %>%

### 2022-12-07 products/paper/main.qmd
Testing out html option in quarto doc

@@
   docx:
+    toc: false
+  html:
     toc: false
@@
 #| tbl-cap: Average marginal differences in the probability of being clearly behind expected grade by monoracial group, relative to a monoracial white student
-texreg(map(results_monoracial[-1], convert_marg),
+#texreg(map(results_monoracial[-1], convert_marg),
+htmlreg(map(results_monoracial[-1], convert_marg),
 #wordreg(map(results_monoracial[-1], convert_marg),

### 2023-01-09 products/paper/main.qmd
Small changes for Soc Sci submission

--- 2022-12-07 products/paper/main.qmd
@@
 title: Differences in the Risk of Grade Retention for Biracial and Monoracial Students in the US, 2010-2019
-author: 
-  - Aaron Gullickson, *University of Oregon*
+#author: 
+#  - Aaron Gullickson, *University of Oregon*
 abstract: Understanding how outcomes for biracial individuals compare to their monoracial peers is critical for understanding how patterns of racial inequality in the contemporary United States might be shifting.
@@
 Although biracial students tend to have favorable family resource characteristics, controlling for these characteristics does little to change the overall placement of their outcomes.
-thanks: The corresponding author can be reached at aarong@uoregon.edu.
-Supplementary materials provided with this article include full model results upon which figures are based as well as any sensitivity analysis described herein.
+thanks: Supplementary materials provided with this article include full model results upon which figures are based as well as any sensitivity analysis described herein.
 All code and data for this project are available at [https://osf.io/4fevt/?view_only=4abc6d86595c4313a8d4792471e9bc0d](https://osf.io/4fevt/?view_only=4abc6d86595c4313a8d4792471e9bc0d).
@@
     fig-pos: t
+    linestretch: 1.5
     keep-tex: false
   docx:
+    toc: false
+  html:
     toc: false
@@
 texreg(map(results_monoracial[-1], convert_marg),
+#htmlreg(map(results_monoracial[-1], convert_marg),
 #wordreg(map(results_monoracial[-1], convert_marg),

### 2023-01-09 products/paper/supp_materials.qmd
Small changes for Soc Sci submission

@@
 All models used sample weights and adjust for design effects from sample weight variance and clustering of students within the same household.
-wordreg(lapply(list(model0_summary, model1_summary, model2_summary, 
+texreg(lapply(list(model0_summary, model1_summary, model2_summary, 
                    model3_summary, model4_summary), convert_model),
@@
        caption.above = TRUE,
-       file="full_reg.doc",
+       #file="full_reg.doc",
        custom.coef.names = 

### 2023-03-08 analysis/analysis_spec_tests.Rmd
Fix error in save of bic table
Change code for spec tests
Use svyglm models and report marginal probs for racial groups across model types

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, eval=TRUE)
+knitr::opts_chunk$set(echo = TRUE, eval=TRUE, warning=FALSE, message=FALSE)
 library(here)
@@
 
-I will judge model fit by the BIC statistic. I wil not consider survey design effects here for speed of computation.
+I will judge model fit by the BIC statistic.
 
 ```{r models-year-grade}
-model_add_cat <- glm(below_exp_grade~as.factor(year)+current_grade+state, 
-                     data=acs, family=binomial)
-model_add_cont <- glm(below_exp_grade~I(year-2010)+current_grade+state, 
-                     data=acs, family=binomial)
-model_inter_cat <- glm(below_exp_grade~as.factor(year)*current_grade+state, 
-                     data=acs, family=binomial)
-model_inter_cont <- glm(below_exp_grade~I(year-2010)*current_grade+state, 
-                     data=acs, family=binomial)
-model_inter_state <- glm(below_exp_grade~I(year-2010)*current_grade+
-                            I(year-2010)*state, 
-                     data=acs, family=binomial)
+acs_design <- acs %>%
+  as_survey_design(ids = cluster, weights = perwt)
+model0 <- svyglm(below_exp_grade~race, design=acs_design, family=binomial)
+model_add_cat <- svyglm(below_exp_grade~as.factor(year)+current_grade+state+race, 
+                     design=acs_design, family=binomial)
+model_add_cont <- svyglm(below_exp_grade~I(year-2010)+current_grade+state+race, 
+                     design=acs_design, family=binomial)
+model_inter_cat <- svyglm(below_exp_grade~as.factor(year)*current_grade+state+race, 
+                     design=acs_design, family=binomial)
+model_inter_cont <- svyglm(below_exp_grade~I(year-2010)*current_grade+state+race, 
+                     design=acs_design, family=binomial)
+model_inter_state <- svyglm(below_exp_grade~I(year-2010)*current_grade+
+                            I(year-2010)*state+race, 
+                     design=acs_design, family=binomial)
+```
 
-bic_scores <- sapply(list("categorical, additive"=model_add_cat,
+```{r bic-scores}
+bic_scores <- sapply(list("just race"=model0,
+                          "categorical, additive"=model_add_cat,
                           "categorical, interaction"=model_inter_cat, 
@@
                           "continous, interaction + state*year continuous interaction"=model_inter_state),
-                     BIC) %>%
+                     function(x) {
+                       as.numeric(BIC(x, maximal=x)["BIC"])
+                     }) %>%
     enframe("model","BIC")
@@
 
-save(bic_scores, save=here("analysis","output","bic_spec_tests.RData"))
+save(bic_scores, file=here("analysis","output","bic_spec_tests.RData"))
 ```
@@
 The clearly preferred model is a continuous year term interacted with current grade.
+
+Now lets take a look at how much the marginal probabilities of CBEG by race varies across model specifications.
+
+```{r marg-fit-table}
+marg_prob_alt <- list(model0, model_add_cat, model_add_cont, model_inter_cat, 
+                  model_inter_cont, model_inter_state) |>
+  map_dfc(function(x) {
+    marg(x, "race", type="levels")[[1]]$Margin
+  })
+
+marg_prob_alt <- bind_cols(levels(acs$race), marg_prob_alt) |>
+  rename(race=...1, baseline=...2, year_dummy=...3, year_linear=...4,
+         year_dummy_interact=...5, year_linear_interact=...6, year_grade_state=...7)
+
+marg_prob_alt |>
+  gt() |>
+  fmt_percent(2:7, decimals = 3) |>
+  cols_label(race="Race", baseline="No Covariates", year_dummy="Year treated as dummy", 
+             year_linear="Year treated as linear", year_dummy_interact="Year (dummy) x Grade",
+             year_linear_interact="Year (linear) x Grade", 
+             year_grade_stat="Year (linear) x Grade and State") |>
+  tab_source_note("*Notes*: All models except no covariate case include dummies for current grade and state") |>
+  tab_spanner("Sensitivity tests on calendar year x grade coding", 3:7)
+
+save(marg_prob_alt, file=here("analysis","output","marg_prob_alt.RData"))
+```
+
+The answer is barely any.
+The only difference is the baseline model which makes sense because we are adding controls, but the difference between the model specifications is only visible out to the fifth decimal place on the proportions.
+
+```{r time-end}
+timestamp()
+```

### 2023-03-08 analysis/check_packages.R
Add gt package

@@ -8,7 +8,7 @@
 packages = c("readr","ggplot2","texreg","dplyr","purrr","stringr","broom",
              "ggrepel","purrr", "kableExtra", "tibble","scales","janitor",
              "survey","modmarg","sandwich","psych","rmdformats","scales",
-             "fastDummies","janitor","srvyr")
+             "fastDummies","janitor","srvyr", "gt")
 
 package.check <- lapply(packages, FUN = function(x) {
   if (!require(x, character.only = TRUE)) {

### 2023-03-08 analysis/analysis_main.Rmd
Add race by grade figures and tables.

@@
 Its possible this may bias the grade retention percentages down a bit, but the differences seem quite minor.
+
+Lets look at these probabilities by race and grade.
+
+```{r prob-race-grade, fig.cap="Probability of grade retention by race and grade for monoracial groups"}
+tbl_race_grade <- acs_design %>%
+  group_by(race, current_grade) %>%
+  summarize(held_back=survey_mean(below_exp_grade, vartype=NULL)) %>%
+  ungroup()
+
+tbl_race_grade %>%
+  filter(!str_detect(race, "/")) %>%
+  ggplot(aes(x=current_grade, y=held_back, 
+             group=race, color=race))+
+  geom_point()+
+  geom_smooth(se=FALSE)+
+  scale_y_continuous(labels=scales::percent)+
+  labs(x="current grade", y="probability of being clearly behind expected grade")+
+  theme_bw()
+```
+
+In kindergarten, its clear that white students actually have a higher probability than Blacks and (slightly) Latinos but this quickly changes direction by even first grade.
+The kindergarten effect may be capturing some of the redshirting effect and lead to overall underestimates of racial inequality among monoracial groups.
+
+```{r}
+tbl_race_grade %>%
+  filter(race=="White" | race=="Black" | race=="Black/White") %>%
+  ggplot(aes(x=current_grade, y=held_back, 
+             group=race, color=race))+
+  geom_point()+
+  geom_smooth(se=FALSE)+
+  scale_y_continuous(labels=scales::percent)+
+  labs(x="current grade", y="probability of being clearly behind expected grade")+
+  theme_bw()
+```
+
+This figure shows how Black/White kids risk evolves over grade compared to their monoracial constituent groups. You don't get a full in-between status until higher grades.
+
+```{r tab-race-grade}
+tbl_race_grade %>%
+  pivot_wider(id_cols=race, names_from=current_grade, 
+              values_from=held_back) %>%
+  gt() %>%
+  fmt_percent(2:14) %>%
+  tab_caption("Probability of being clearly behind expected grade by race and current grade")
+```
 
@@
 #run separate models by grade groups
-model4_elem <- update(model4, design=subset(acs_design, current_grade<="5th"))
+model4_elem <- update(model4, design=subset(acs_design, current_grade>"K" & current_grade<="5th"))
 model4_middle <- update(model4, design=subset(acs_design, current_grade>"5th" & 
@@
 results_hs <- calculate_marg_means(model4_hs)
+
+save(results_elem, results_middle, results_hs, file=here("analysis","output","models_grade_robust.RData"))
 

### 2023-03-09 analysis/analysis_spec_tests.Rmd
Correct error in spec test

@@
              year_linear_interact="Year (linear) x Grade", 
-             year_grade_stat="Year (linear) x Grade and State") |>
+             year_grade_state="Year (linear) x Grade and State") |>
   tab_source_note("*Notes*: All models except no covariate case include dummies for current grade and state") |>

### 2023-03-13 products/paper/supp_materials.qmd
Add table in supplementary materials to address issue #33

@@
 load(here("analysis","output","model_summaries.RData"))
+load(here("analysis","output","marg_prob_alt.RData"))
 load(here("analysis","output","acs.RData"))
@@
 #| label: create-design
+#| eval: false
 acs_design <- acs %>%
@@
 
+```{r marg-fit-table}
+#| label: tbl-marg-fit
+#| tbl-cap: Marginal probabilities of grade retention by racial group across models using different specifications to model the the temporal trend in grade retention by current grade of the student.
+marg_prob_alt %>%
+  mutate(baseline=percent(round(baseline, 5)),
+         year_dummy=percent(round(year_dummy, 5)),
+         year_linear=percent(round(year_linear, 5)),
+         year_dummy_interact=percent(round(year_dummy_interact, 5)),
+         year_linear_interact=percent(round(year_linear_interact, 5)),
+         year_grade_state=percent(round(year_grade_state, 5))) %>%
+  kbl(booktabs=TRUE, linesep=NULL,
+      align = c("l",rep("r", ncol(marg_prob_alt)-1)),
+      col.names=c("Race", "No Covariates", "Year as dummy", 
+                  "Year as linear", "Year (dummy) x Grade",
+                  "Year (linear) x Grade", 
+                  "Year (linear) x Grade and State")) %>%
+  add_header_above(c(" " = 2, 
+                     "Sensitivity tests on calendar year x grade coding" = 5)) %>%
+  column_spec(2:7, width = "1.75cm") %>%
+  footnote(general = "All models except no covariate case include dummies for current grade and state")
+```
+
 ```{r desc-stats-ses}
@@
 #| tbl-cap: Socioeconomic resources by race. Shading indicates biracial group. All results apply survey weights.
-
+#| eval: false
 tbl_summary <- acs_design %>%
@@
 #| tbl-cap: Cultural resources by race. Shading indicates biracial group. All results apply survey weights.
+#| eval: false
 tbl_summary <- acs_design %>%

### 2023-03-13 products/paper/supp_materials.qmd
Add figure to supplementary appendix to address issue #34

@@
 load(here("analysis","output","marg_prob_alt.RData"))
+load(here("analysis","output","models_grade_robust.RData"))
 load(here("analysis","output","acs.RData"))
+color_choices <- c("#003f5c","#58508d","#bc5090","#ff6361","#000000","#ffa600")
 ```
@@
 ```
+
+```{r}
+#| label: fig-results-grade-group
+#| fig-cap: Probability of being clearly behind expected grade for biracial respondents, in comparison to their monoracial comparison groups, separately by elementary (not including kindergarten), middle school, and high school.
+Non-overlap in color corresponding confidence bands indicates statistically significant difference at the 5% level, Baseline model includes year, grade, location, and state fixed effects.
+Full models include control variables for nativity, English proficiency, income, education, home ownership, and marital status of parents.
+#| fig-height: 7.1
+coef_table <- results_elem$coef_table %>%
+  mutate(model="elementary")
+coef_table <- results_middle$coef_table %>%
+  mutate(model="middle school") %>%
+  bind_rows(coef_table)
+coef_table <- results_hs$coef_table %>%
+  mutate(model="high school") %>%
+  bind_rows(coef_table) %>%
+  mutate(model=factor(model, levels=c("elementary","middle school",
+                                      "high school")),
+         mrace=factor(mrace, levels=c("Black/White",
+                                      "Black/Latino",
+                                      "Black/Asian",
+                                      "Black/Indigenous",
+                                      "White/Indigenous",
+                                      "White/Latino",
+                                      "White/Asian",
+                                      "Latino/Asian",
+                                      "Indigenous/Latino",
+                                      "Indigenous/Asian")))
+
+ci_df <- results_elem$ci_df %>%
+  mutate(model="elementary")
+ci_df <- results_middle$ci_df %>%
+  mutate(model="middle school") %>%
+  bind_rows(ci_df)
+ci_df <- results_hs$ci_df %>%
+  mutate(model="high school") %>%
+  bind_rows(ci_df) %>%
+  mutate(model=factor(model, levels=c("elementary","middle school",
+                                      "high school")),
+         mrace=factor(mrace, levels=c("Black/White",
+                                      "Black/Latino",
+                                      "Black/Asian",
+                                      "Black/Indigenous",
+                                      "White/Indigenous",
+                                      "White/Latino",
+                                      "White/Asian",
+                                      "Latino/Asian",
+                                      "Indigenous/Latino",
+                                      "Indigenous/Asian")))
+
+ggplot(coef_table, aes(x=term, y=estimate, color=colors))+
+  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower, ymax=upper, fill=colors), 
+            color=NA, alpha=0.5)+
+  geom_linerange(data=ci_df, 
+                 position=position_dodge2(width = 0.1), size=1, alpha=0.5,
+                 aes(ymin=lower, ymax=upper, color=colors))+
+  geom_point()+
+  coord_flip()+
+  facet_grid(mrace~model, scales="free_y")+
+  theme_bw()+
+  theme(legend.position = "none", panel.grid = element_blank(), 
+        strip.text.y = element_text(size = 4))+
+  scale_color_manual(values=color_choices)+
+  scale_fill_manual(values=color_choices)+
+  scale_y_continuous(labels = scales::percent)+
+  labs(x=NULL,
+       y="average adjusted predicted probability of being behind expected grade")
+```

### 2023-03-13 products/paper/main.qmd
Add footnote to main text regarding issue #33

@@
 Sensitivity analysis showed that this functional form was preferred by the Bayesian information criterion to models with no interaction and a model with interaction terms that treated year as a categorical variable.
-I also considered a similar interaction between state and year, but this model was not preferred.
+I also considered a similar interaction between state and year, but this model was not preferred.[^yeargrade]
+
+[^yeargrade]: The supplementary materials provide, for each specification, the BIC scores and marginal probabilities of being clearly behind expected grade for each racial group.
+The marginal probabilities for each racial group are almost identical across specifications, indicating that results are not driven by the modeling choice made here.
 

### 2023-03-13 products/paper/supp_materials.qmd
Add footnote to main text regarding issue #33

@@
 #| label: tbl-marg-fit
-#| tbl-cap: Marginal probabilities of grade retention by racial group across models using different specifications to model the the temporal trend in grade retention by current grade of the student.
+#| tbl-cap: Marginal probabilities of being clearly behind expected grade by racial group across models using different specifications to model the the temporal trend in grade retention by current grade of the student.
 marg_prob_alt %>%

### 2023-03-13 products/paper/main.qmd
Forgot slight edit to the previous footnote

@@
 
-[^yeargrade]: The supplementary materials provide, for each specification, the BIC scores and marginal probabilities of being clearly behind expected grade for each racial group.
+[^yeargrade]: The supplementary materials provide, for each model specification, the BIC scores and marginal probabilities of being clearly behind expected grade for each racial group.
 The marginal probabilities for each racial group are almost identical across specifications, indicating that results are not driven by the modeling choice made here.

### 2023-03-13 products/paper/supp_materials.qmd
Add table of CBEG probability by race and grade to address issue #34

@@
 #| label: create-design
-#| eval: false
 acs_design <- acs %>%
@@
 
+```{r prob-race-grade, fig.cap=""}
+#| label: tbl-prob-race-grade
+#| tbl-cap: Probability of being clearly behind expected grade by race and current grade.
+tbl_race_grade <- acs_design %>%
+  group_by(race, current_grade) %>%
+  summarize(held_back=survey_mean(below_exp_grade, vartype=NULL)) %>%
+  ungroup() %>%
+  pivot_wider(id_cols=race, names_from=current_grade, 
+              values_from=held_back) %>%
+  rename(Race=race) %>%
+  mutate(K=percent(K, accuracy=0.01),
+         `1st`=percent(`1st`, accuracy=0.01),
+         `2nd`=percent(`2nd`, accuracy=0.01),
+         `3rd`=percent(`3rd`, accuracy=0.01),
+         `4th`=percent(`4th`, accuracy=0.01),
+         `5th`=percent(`5th`, accuracy=0.01),
+         `6th`=percent(`6th`, accuracy=0.01),
+         `7th`=percent(`7th`, accuracy=0.01),
+         `8th`=percent(`8th`, accuracy=0.01),
+         `9th`=percent(`9th`, accuracy=0.01),
+         `10th`=percent(`10th`, accuracy=0.01),
+         `11th`=percent(`11th`, accuracy=0.01),
+         `12th`=percent(`12th`, accuracy=0.01))
+
+tbl_race_grade %>%
+  kbl(booktabs = TRUE,
+      linesep = NULL,
+      align = c("l",rep("r", ncol(tbl_race_grade)-1))) %>%
+  kable_styling(font_size=8) %>%
+  column_spec(2:14, width = "0.6cm")
+```
+
 ```{r tbl-bic}
@@
 #| tbl-cap: Socioeconomic resources by race. Shading indicates biracial group. All results apply survey weights.
-#| eval: false
 tbl_summary <- acs_design %>%
@@
 #| tbl-cap: Cultural resources by race. Shading indicates biracial group. All results apply survey weights.
-#| eval: false
 tbl_summary <- acs_design %>%

### 2023-03-13 products/paper/main.qmd
Add paragraph to address redshirt issue #34 and footnoote to address issue #34 and #35

@@
 @tbl-descriptive shows the percent of each monoracial and biracial group that is CBEG.
+
+Using grade-age comparisons to infer grade retention may inadvertently capture cases of "academic redshirting" in which parents intentionally delay their child's kindergarten enrollment by a year.
+@frederick_have_2008 suggest that academic redshirting is in many cases be a form of preemptive retention for children with developmental delays and thus the error induced by these false positive cases may be minimal.
+However, @bassok_academic_2013 have shown that academic redshirting is more common among white and SES privileged parents, suggesting that the demographic covariates of academic redshirting may operate in the opposite direction of grade retention.
+Regardless, academic redshirting should be of minimal concern for the measure of CBEG used here.
+Academic redshirting most frequently occurs for students whose birthdays fall close to the cutoff period for enrollment [@graue_redshirting_2000; @bassok_academic_2013].
+Therefore, academic redshirts spend most of a given school year at the older, but correct, modal age for their grade and most of these students will not be identified as CBEG.[^redshirts]
+
+[^redshirts]: Analysis of the probability of being CBEG by race and grade, shown in the supplementary materials, indicates that Black and Latino children in kindergarten are less likely to be CBEG than White children, which may be a consequence of redshirting.
+However, these differences quickly reverse direction by first grade.
+As a sensitivity analysis, I repeated the main analysis shown here separately for students in elementary (1st-5th), middle school (6th-8th), and high school (9th-12th) grades.
+These results are presented in the supplementary materials.
+Although more statistically noisy due to smaller sample sizes, those results are consistent with the conclusions drawn here.
 

### 2023-03-14 products/paper/main.qmd
Rework conclusions to address issue #36

@@
 The results on parent selection vary somewhat by group, but the majority of biracial groups had resources more similar to their monoracial constituent group with higher resources, indicating a strong positive selection into interracial unions in the prior generation.
-The most important exception to this trend is the case of Black/White biracial students who have far fewer resources than white students and resources closer to black students.
+The most important exception to this trend is the case of Black/White biracial students who have far fewer resources than White students and resources closer to Black students.
 
@@
 
-Because these findings rely upon large samples, they present an important step forward in our understanding of how biracial individuals will fit into America's system of racial inequality.
-However, this study also has important limitations that need to be noted.
+These findings stand in contrast to earlier work which found more evidence of an "in-between" status in other educational outcomes [@herman_blackwhiteother_2009; @campbell_multiracial_2009].
+I raise two possible explanations for this discrepancy that point to strengths and weaknesses of the current research and may help drive future work on the topic.
+First, this discrepancy may be due to the small sample sizes of multiracial respondents in prior work which led to statistically imprecise estimates.
+Because the current findings rely upon large samples, they present an important step forward in our understanding of how biracial individuals will fit into America's system of racial inequality.
+Nonetheless, sample sizes for some non-white biracial respondents remain small, limiting our understanding of the outcomes for those groups.
+In particular, results for non-white part-Indigenous populations are quite imprecise.
+The results counter-intuitively suggest that part-Indigenous populations are the least likely to be in the position expected by hyperdescent, but due to low sample sizes, that finding is highly tentative.
 
-Sample sizes for some non-white biracial respondents remain small, limiting our understanding of the outcomes for those groups.
-In particular, results for non-white part-Indigenous populations are quite imprecise.
-The results interestingly and counterintuitively suggest that part-Indigenous populations are the least likely to be in the position expected by hyperdescent, but due to low sample sizes, that finding is highly tentative.
-
-Wile grade retention is an important early life outcome, it is far from the only one that these individuals will face in their life.
+Second, The current findings may also diverge from the findings of prior work because different outcomes were examined.
+While grade retention is an important early life outcome, it is far from the only one that these individuals will face in their life.
 Although, we often observe similar racial inequalities across a variety of outcomes, knowing the result for one outcome does not perfectly predict other outcomes.

### 2023-03-14 products/submissions/SocSci/randr/response_memo.md
Add submission materials for SocSci R&R

# Response Memo, Sociological Science

Thank you to the reviewer and editor for the helpful comments and to the editor for the opportunity to revise my manuscript. I have made changes to the manuscript as requested and I believe those changes have improved the overall manuscript. Those changes are described below.

## Issue 1

From the editor's letter:

> Please provide descriptives on the distribution of CBEG across racial groups and discuss the (im-)possibility of false positives and how this could (not) have impacted the results. DE: “What is the distribution of this measure across across groups? Some descriptives here would help us. For example, do some groups systematically enroll their kids in school later than others? In other words, is the measure truly capturing what we want it to capture? (For example, in Figure 2, we see that it is possible to have a CBEG status even in kindergarten. What is keeping these students behind? Or are they simply starting later than others?)”

There are a couple of points to address here. First, I believe the concern that the reviewer is raising, although they do not name it directly, is the phenomenon of "academic redshirting" in which parents hold back kindergarten enrollment for a child whose birthday falls close to the cutoff date for enrollment. It is true that I should have addressed this issue more directly in the paper and I have done so in the revised draft.

* The distribution of CBEG across racial groups was already provided in the first draft in table 1. However, I think what the reviewer is interested in is the distribution of CBEG across racial groups and grades with a particular focus on kindergarten. I have added a table to the supplementary appendix (Table 2) that shows the probability of CBEG jointly across race and current grade. There is some evidence at the kindergarten level of the usual racial disparities being reversed. Most notably, Black and Latino kindergarteners are less likely to be CBEG than White kindergartners. This finding is consistent with prior work on the topic of redshirting (Bassok and Reardon 2013). However, these inverted racial differences quickly flip at higher grades.
* Based on the belief that redshirting is more likely to affect results at younger grades where it is a larger share, I ran a sensitivity analysis that repeated my full models but separately for the elementary grades w/o kindergarten (1st-5th), middle school (6th-8th), and high school (9th-12th). The results from this sensitivity analysis are provided in Figure 1 of the supplementary materials. Although somewhat noisier, they are consistent with the patterns found in the main text. This sensitivity analysis is described in footnote 1 of the main text which reads:

> Analysis of the probability of being CBEG by race and grade, shown in the supplementary materials, indicates that Black and Latino children in kindergarten are less likely to be CBEG than White children, which may be a consequence of redshirting. However, these differences quickly reverse direction by first grade. As a sensitivity analysis, I repeated the main analysis shown here separately for students in elementary (1st-5th), middle school (6th-8th), and high school (9th-12th) grades. These results are presented in the supplementary materials. Although more statistically noisy due to smaller sample sizes, those results are consistent with the conclusions drawn here.

* I added a discussion of redshirting to the main body of the text on pg. 13-14 of the main text which reads:

> Using grade-age comparisons to infer grade retention may inadvertently capture cases of “academic redshirting” in which parents intentionally delay their child’s kindergarten enrollment by a year. Frederick and Hauser (2008) suggest that academic redshirting is in many cases be a form of preemptive retention for children with developmental delays and thus the error induced by these false positive cases may be minimal. However, Bassok and Reardon (2013) have shown that academic redshirting is more common among white and SES privileged parents, suggesting that the demographic covariates of academic redshirting may operate in the opposite direction of grade retention. Regardless, academic redshirting should be of minimal concern for the measure of CBEG used here. Academic redshirting most frequently occurs for students whose birthdays fall close to the cutoff period for enrollment (Bassok and Reardon 2013; Graue and DiPerna 2000). Therefore, academic redshirts spend most of a given school year at the older, but correct, modal age for their grade and most of these students will not be identified as CBEG.

## Issue 2

From the editor's letter:

> Please clarify whether results remain unaltered when year dummies are used. DE: “One question about the specification: The author notes that linear year term interacted with current grade was preferred by the Bayesian Information Criterion in model comparison. But did the main results remain similar across different specifications (e.g., year dummies, year dummies x current grade)? It would be good to include more robustness checks to ensure the results are not driven by any of the decisions made along the way.”

This is a good point. I have now added a new table (Table 4) to the supplementary materials which shows the marginal probabilities of CBEG for each racial group across all the different specifications of year, current grade, and state that were considered. These probabilities are virtually identical down to the fifth decimal place across specifications. I discuss this analysis in footnote 5 of the main text.

## Issue 3

From the editor's letter:

> The paper is now framed as providing the first good evidence on the matter at hand, saying previous studies had two forms of data issues. But previous studies did have some significant findings despite smaller samples, and these findings seem to deviate from the ones obtained here, specifically Herman (2009) and Campbell (2009). Consider whether the paper’s impact would increase if you engaged with these prior findings more, perhaps presenting the results as in contrast with earlier findings, and considering how they may be reconciled. The paper could also engage more with findings from earlier studies using identity measures of ethic/racial group membership, and more clearly articulate how the issue of reverse causality solved in the present paper reconciles differences in findings. This may help make clear how the paper moves the field forward. This could all be done in the discussion section, which currently present the results on their own, without relating them to earlier findings.

Thank you for the nudge on this. My final three paragraphs of the conclusions were meant to address but they did so indirectly. I have now edited those paragraphs to discuss the prior work and how that relates to both the strengths and weaknesses of the current work and pathways for future work. It now reads as follows:

> These findings stand in contrast to earlier work which found more evidence of an “in-between” status in other educational outcomes (Campbell 2009; Herman 2009). I raise two possible explanations for this discrepancy that point to strengths and weaknesses of the current research and may help drive future work on the topic. First, this discrepancy may be due to the small sample sizes of multiracial respondents in prior work which led to statistically imprecise estimates. Because the current findings rely upon large samples, they present an important step forward in our understanding of how biracial individuals will fit into America’s system of racial inequality. Nonetheless, sample sizes for some non-white biracial respondents remain small, limiting our understanding of the outcomes for those groups. In particular, results for non-white part-Indigenous populations are quite imprecise. The results counter-intuitively suggest that part-Indigenous populations are the least likely to be in the position expected by hyperdescent, but due to low sample sizes, that finding is highly tentative.

> Second, The current findings may also diverge from the findings of prior work because different outcomes were examined. While grade retention is an important early life outcome, it is far from the only one that these individuals will face in their life. Although, we often observe similar racial inequalities across a variety of outcomes, knowing the result for one outcome does not perfectly predict other outcomes. As biracial individuals age and face later outcomes in the educational system, the labor market, and elsewhere, their experiences may differ from what I observe here. Thus, collecting more detailed information on these outcomes in future studies remains critically important.

### 2023-03-14 products/submissions/SocSci/randr/main.tex
Add submission materials for SocSci R&R

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Differences in the Risk of Grade Retention for Biracial and Monoracial Students in the US, 2010-2019},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Differences in the Risk of Grade Retention for Biracial and
Monoracial Students in the US, 2010-2019\thanks{Supplementary materials
provided with this article include full model results upon which figures
are based as well as any sensitivity analysis described herein. All code
and data for this project are available at
\url{https://osf.io/4fevt/?view_only=4abc6d86595c4313a8d4792471e9bc0d}.}}
\author{}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Understanding how outcomes for biracial individuals compare to their
monoracial peers is critical for understanding how patterns of racial
inequality in the contemporary United States might be shifting. Yet, we
know very little about the life chances of biracial individuals due to
limitations in most available data sources. In this article, I utilize
American Community Survey data from 2010-2019 to examine the risk of
being clearly behind expected grade among biracial and monoracial K-12
students, helping to fill a gap in our understanding. With large sample
sizes for most biracial groups, I am able to estimate grade retention
risk for biracial students with enough precision to differentiate even
modest differences in risk relative to monoracial groups. The results
indicate that for most biracial groups, biracial students have risk
similar to their monoracial constituent group with lower risk. Although
biracial students tend to have favorable family resource
characteristics, controlling for these characteristics does little to
change the overall placement of their outcomes.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable, boxrule=0pt, frame hidden, sharp corners]}{\end{tcolorbox}}\fi

\setstretch{1.5}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The experiences of mixed race individuals will play an important role in
the future of race in the US. A rise in interracial unions since the
1960s has resulted in a ``biracial baby boom'' of individuals whose
parents cross racial lines (Root 1992) and this demographic surge may
have transformative effects on racial boundaries (Alba 2020; Hochschild,
Weaver, and Burch 2012). While racial mixing is nothing new to US
history (Gullickson and Morning 2011; Morning 2000), this biracial boom
is notable for its size and has occurred in an environment in which
historical norms of racial classification may be eroding.

The transformative potential of this biracial baby boom is contingent on
both how these biracial individuals identify their own race and how they
are sorted into a prevailing system of racial inequality. While
extensive scholarship documents the identification and classification
decisions made by or on behalf of multiracial individuals (Bratter and
Heard 2009; Brunsma 2005; Davenport 2016; Herman 2004; Holloway et al.
2009; Liebler 2016; Qian 2004; Roth 2005; Xie and Goyette 1997), only a
smattering of research has documented how the life chances of biracial
individuals compare to their monoracial peers (Campbell 2009; Herman
2009; Kao 1999). In this article, I expand on this work by examining the
risk of grade retention for K-12 students whose parents identify with
different races.

By examining this outcome, I am able to overcome two important
limitations that have hampered prior work on the life chances of
biracial individuals. First, using self identification of biracial
respondents to analyze outcomes is problematic. Prior research indicates
that individuals whose parents belong to different races do not
consistently identify with both races (Doyle and Kao 2007; Gullickson
2019; Liebler et al. 2017). Furthermore, social status and social
mobility can predict racial identification generally (Saperstein and
Gullickson 2013; Saperstein and Penner 2012) and social status is
related to the identification choices of biracial individuals
specifically (Davenport 2016; Roth 2005). Thus, differences in outcomes
among biracial individuals may in part produce differences in
identification, rather than the other way around.

This problem of reverse causation can be eliminated when researchers
have access to the racial identification of a respondent's biological
parents. Even when such information is available, however, researchers
run into a second limitation of sample size. The samples of biracial
respondents in many data sources are too small to make reliable
statistical inferences. This limitation is particularly severe for
biracial individuals with two non-white parents.

Biracial children who reside with both biological parents can be
identified in large scale data sources such as the Census. Researchers
have previously utilized this feature to examine the classification
choices made on behalf of these children (Roth 2005; Xie and Goyette
1997). Analyzing outcomes for such children is more difficult, for the
simple reason that they do not have many measurable outcomes yet.
However, grade retention is measurable and important even for young
children. Being held back in school is a relatively common occurrence
that has consequences for later educational transitions (Jimerson 2001)
and is disproportionately applied by race (Warren, Hoffman, and Andrew
2014). Grade retention is one of the earliest stratifiers of life
chances that an individual will experience.

In this article, I use data from the American Community Survey to
examine the likelihood of being held back in school for biracial
children in comparison to their monoracial peers. Pooling data from 2010
to 2019, I am able to draw a large enough sample to make reliable
statistical inferences for most biracial groups, including those with
two non-white parents. The results provide a window into the placement
of biracial children within an existing system of racial inequality. In
contrast to prevailing hypotheses that predict either a pattern of
hypodescent or an in-between position for biracial individuals, I find a
consistent pattern in which the risk of grade retention is similar to
that of the lower risk constituent monoracial group. This result holds
even after adjusting for differences in resources across respondents,
despite the fact that most biracial groups are positively selected by
parental resources relative to their monoracial group.

\hypertarget{where-do-biracials-fit}{%
\section{Where do Biracials Fit?}\label{where-do-biracials-fit}}

How might we expect the outcomes for biracial individuals to differ from
their monoracial peers? In every case, the relevant comparison is how a
biracial person fares relative to the two constituent groups with which
their parents identify. For example, in the case of an individual with
one Black and one White parent, we would compare the average outcomes
for such individuals to the average outcomes for individuals with two
White and two Black parents, respectively. The outcome for the
Black/White group may be similar to one of the monoracial groups, it may
be somewhere in between the outcomes for the two monoracial groups, or
it may be higher or lower than either of the monoracial groups.

I distinguish three mechanisms that help us understand where biracial
individuals might be positioned among these possibilities. First, I
consider arguments regarding racial discrimination at both an individual
and systemic level. Second, I consider arguments regarding different
distributions of resources across racial groups. Third, I consider
arguments regarding how the nature of hybridity itself may affect
outcomes.

\hypertarget{discrimination-and-classification}{%
\subsection{Discrimination and
Classification}\label{discrimination-and-classification}}

Discrimination and bias by race play an important role in overall levels
of racial inequality (Pager and Shepherd 2008; Quillian 2006).
Discrimination need not be consciously recognized by actors, but may
instead reflect cognitive biases and schema that individuals unknowingly
rely upon and reinforce in social interactions (Quillian 2006).
Discrimination can also be systemic in nature when biases are built into
and reinforced by organizations and institutional systems (Reskin 2012).
In our current ``colorblind'' regime (Bonilla-Silva 2006), such systems
may be formally race neutral but produce disparate impacts by race
(Pager and Shepherd 2008).

The experiences of biracial individuals within such a system of
discrimination depend almost entirely on how they are classified by
others. To the extent that biracial individuals are classified as more
like one of their monoracial constituent groups than the other, they
will receive the same treatment and reap similar rewards or suffer
similar penalties.

The strongest historical precedent for understanding such classification
in the US is the ``one-drop rule'' norm governing classification as
Black. According to the one-drop rule, individuals of known Black
ancestry are considered exclusively Black, regardless of other
ancestries (Davis 1991). Davis (1991) generalizes this practice into the
concept of \emph{hypodescent} in which individuals with mixed racial
ancestry will be classified as members of the lower status group. A
parallel pattern of \emph{hyperdescent} implies that individuals of
mixed racial ancestry will be classified as members of the higher status
group. This pattern of hyperdescent has historically governed the
experiences of individuals with American Indian ancestry (Snipp 1989;
Wolfe 2001).

Although the hypodescent/hyperdescent paradigm is often used to frame
the experiences of multiracial individuals, its generalizability and
contemporary relevance is questionable. Hypodescent and hyperdescent are
observed in two specific historical cases involving Black ancestry and
American Indian ancestry, respectively. As Iverson et al. (2022) note,
these two cases could just as parsimoniously be explained by a
``dominance'' model which ranks ancestries by their tendencies to be
supercessive or recessive in determining identification and
classification.

The development of particular norms may also reflect the historical
regime in which they were developed (Gullickson and Morning 2011;
Iverson et al. 2022). In this case, neither the hypodescent/hyperdescent
or dominance paradigm may help us to fully understand the experiences of
those of Latino and Asian ancestry due to more recent processes of
racialization. Furthermore, the historical norms for all groups may be
transforming in the post-Civil Rights era.

If outcomes for biracial individuals all closer to one of their
constituent groups, then we may have stronger evidence about such norms
of classification. However, what if those outcomes fall in-between the
two groups? Processes of classification and discrimination could still
produce an in-between status for several reasons.

First, racial ambiguity in appearance may lead to more inconsistent
classification into single race categories across observers. Even if
norms consistently indicate a certain type of classification, correct
identification by observers is a probabilistic phenomenon and likely to
be more heterogeneous for biracial individuals. As a result, biracial
individuals will experience a lower sum total of discrimination and bias
affecting the lower status group in the case of hypodescent and will not
reap the same rewards as a member of the higher status group in the case
of hyperdescent.

Second, biracial people's experience of discrimination may be
``softened'' because a more ambiguous physical appearance provokes less
racial antipathy. The US already has a long history of this kind of
softening. Because of the one-drop rule, Black individuals vary
substantially in skin tone and those individuals with a lighter skin
tone have better outcomes (Hughes and Hertel 1990; Keith and Herring
1991; Monk 2014) and are viewed more favorably by both Blacks and Whites
(Maddox and Gray 2002). Thus, light skin moderates the stigma of
Blackness even among monoracial individuals.

Third, biracial individuals may be classified into an ambiguous
``middleman'' minority position. This kind of middle tier status for
individuals of mixed race is common in Latin America (Telles and Sue
2009) and Bonilla-Silva (2004) has argued that such a system may be
emerging in the contemporary US as well. In this case, many biracial
groups may occupy a nebulous and ill-defined middleman or buffer class
position in-between whites and darker-skinned non-whites, helping to
cement a new kind of racial hierarchy.

\hypertarget{differential-resources}{%
\subsection{Differential Resources}\label{differential-resources}}

Racial inequality is at least partly explained by different
distributions of material, cultural, and social resources across racial
groups (Conley 1999). On average, socioeconomic background
(e.g.~parental income, wealth, and education) differs by race and such
socioeconomic background is consequential for outcomes. Other less
materially defined resources also vary across groups such as cultural
styles, social networks, and neighborhood quality. These differences in
background resources reflect the accumulation of historical processes of
discrimination and, when combined with contemporary processes of
discrimination, magnify the overall level of racial inequality we
observe.

How do differences in the distribution of resources affect biracial
individuals? Naively, we might expect biracial individuals to inherit
parental resources roughly halfway between their two constituent groups,
on average (Chew, Eggebeen, and Uhlenberg 1989). Prior research looking
at biracial children in the Census data has found some evidence to
support this argument (Chew et al. 1989). In such a case, the resources
of biracial individuals would predict outcomes in-between the two
constituent groups, barring other mechanisms.

However, the actual resource position of biracial individuals is likely
far more complicated because people are not randomly selected into
interracial unions. The relationship between this selection process and
a biracial child's outcomes depends on the strength and direction of
that selection on each partner, its consistency across different types
of unions, and how strongly the selected characteristic predicts
outcomes in the next generation.

Status exchange theory predicts that given an unequal racial hierarchy,
individuals from a lower status racial group will be positively selected
on socioeconomic characteristics into intermarriage with a higher status
group, while individuals from a higher status racial group will be
similarly negatively selected into the same marriage (Davis 1941; Fu
2001; Merton 1941). Prior research has shown substantial evidence of
this pattern in the case of Black male/White female marriages in the US,
but weaker evidence for other types of unions (Fu 2001; Gullickson 2006;
Hou and Myles 2011; Kalmijn 2010). Even in cases where status exchange
is present, its net effect on resources is complex because status
exchange positively selects one partner and negatively selects the other
partner. Therefore the two effects may somewhat cancel each other out in
terms of the overall distribution of resources across parents.

Regardless of the underlying complexity of this selection process,
delineating its role as opposed to other mechanisms is key to
understanding the overall process that sorts biracial individuals within
the racial hierarchy. Although this approach is sometimes characterized
as pitting race versus class, that is not the case here. Instead, this
delineation allows us to understand how race affects the outcomes of
biracial individuals in a multigenerational process.

\hypertarget{hybridity}{%
\subsection{Hybridity}\label{hybridity}}

The experience of hybridity itself may also positively or negatively
affect outcomes for biracial individuals. According to ``marginal man''
theory, biracial individuals will find it difficult to fit in due to the
dissonance of an identity that crosses strong and salient racial
boundaries (Park 1928; Stonequist 1935). This liminal identity conflict
will result in negative mental health experiences that could feed into
other negative outcomes as well. Evidence for such negative mental
health experiences, however, has been mixed (Bratter and Gorman 2011;
Campbell and Eggerling-Boeck 2006; Cheng and Lively 2009; Udry, Li, and
Hendrickson-Smith 2003).

Marginal man theorists viewed culture in monolithic and static terms and
thus to be trapped in between two cultures was seen as a difficult
experience. Contemporary views of culture instead emphasize the
flexible, fragmented, and fluid use of cultural bits that comprise a
``toolkit'' (DiMaggio 1997; Swidler 1986). From this perspective,
biracial individuals may actually benefit from an enlarged cultural
toolkit that allows them to navigate a variety of diverse social
contexts (Shih et al. 2019).

The complexity of the underlying mechanisms described above prevents
simple and straightforward hypotheses regarding the position of biracial
individuals. Different mechanisms may produce the same observed outcome
and the weight of different mechanisms can also vary by groups as these
groups may be governed by different regimes of classification. However,
comparing outcomes across biracial groups may help us better understand
the likely mechanisms involved and to disregard others as implausible.
Accounting for the effect of parental resources on outcomes to the best
of our ability is also key to reaching any conclusions about the
placement of biracial individuals within the racial hierarchy.

\hypertarget{prior-work}{%
\section{Prior Work}\label{prior-work}}

Prior work on the outcomes for biracial individuals has been hampered by
several methodological issues. With good reason, researchers have been
reluctant to use data in which multiracial or biracial status is
determined solely by the self-identification of respondents. We know
that far fewer people identify as biracial than could feasibly do so
based on their parents' races (Morning and Saperstein 2018) and that
various selection effects operate on the decision to identify as
biracial among those who could (Davenport 2016; Roth 2005). Furthermore,
research on changes in racial reporting over time suggests that other
status markers like education and income may influence how people
identify themselves (Saperstein and Gullickson 2013; Saperstein and
Penner 2012). For example, a person with one Black and one White parent
who is routinely treated by others as Black and discriminated against as
Black may be more likely to identify themselves as exclusively Black
than a similar person of Black and White ancestry who experiences less
consistent discrimination. Thus, any analysis of outcomes for biracial
individuals based on self-identification must deal with the potential
for reverse causation in the observed differences between biracial
individuals and their constituent groups (Bratter 2018).

Instead of self-identification, researchers have generally relied upon
survey data in which the racial identification of a respondent's
biological parents are provided. Campbell (2009) used data from the
National Longitudinal Survey of Adolescent to Adult Health (Add Health)
to examine high school GPA, advanced placement in math, and four-year
college enrollment among biracial and monoracial individuals identified
by co-resident biological parents' races. Herman (2009) examined
differences in GPA using data on high school students in California and
Wisconsin in which the students reported their parents' races. Both of
these studies found some support for an ``in-between'' status for many
biracial groups, but are hampered by a lack of statistical power due to
small sample sizes. Kao (1999) used data from the National Educational
Longitudinal Study of 1988 to examine mathematics achievement scores and
GPA for biracial students with Black and Asian parentage and finds that
biracial Black students have outcomes similar to monoracial Black
students, while biracial Asian students have outcomes more similar to
monoracial White students. This study, however, is limited by the fact
that biracial students are identified by a discrepancy between their
reported race and one parent's reported race, rather than by both
biological parents' races.

Results in prior work tend to be inconclusive, due to the sample size
for most biracial groups. An alternative approach is to use large scale
Census or American Community Survey (ACS) data to identify biracial
children by the race of the biological parents in their household. This
approach can generate much larger samples of biracial individuals.
However, researchers then have a limited range of outcomes to explore
because biracial individuals identified by this method are by definition
children. This approach has been used previously to examine the
likelihood of living in poverty (Bratter and Damaske 2013; Bratter and
Kimbro 2013) and residential segregation (Ellis et al. 2012). These
studies provide valuable information on biracial childrens' lived
experience, but do not directly measure their own outcomes. However, one
important educational outcome is partially identifiable for children in
Census and ACS data - grade retention.

Grade retention is the practice of having students repeat a grade due to
poor academic performance. Prior work on grade retention has failed to
show a positive effect of grade retention on later educational outcomes,
and has instead found evidence of ``scarring'' effects that lead to
later negative educational outcomes, such as a higher dropout risk
(Andrew 2014; Hughes et al. 2017; Jimerson 2001; Stearns et al. 2007).
The risk of grade retention also varies substantially by race and class
background, with Black and Latino students at particularly high risk of
grade retention relative to other groups (Warren et al. 2014).

Because grade retention often happens very early in the K-12 system, it
is one of the earliest stratification mechanisms that individuals
encounter. For these same reasons, it provides an important window into
understanding the placement of biracial children within a racially
stratified educational environment. In the remainder of this article, I
examine this placement using data from the American Community Survey.

\hypertarget{data-and-methods}{%
\section{Data and Methods}\label{data-and-methods}}

Data for this analysis come from the American Community Survey (ACS), an
annual 1-in-100 survey of the US population, conducted by the Census
Bureau. To increase sample size for smaller populations of biracial
respondents, I pool ACS data for a full decade from 2010-2019. All data
were extracted from the Minnesota Population Center's Integrated Public
Use Microdata Series (Ruggles et al. 2020).

Ideally, I would restrict the sample to children living in a household
with two biological parents. However, while the ACS distinguishes
between biological, adopted, and stepchildren, it only records the
direct family relationship between the head of household and other
members of the household. Respondents can therefore be identified as
biological children of another member of the household only if that
member is the head of household or the biological child of the head of
household. Determining whether the partner/spouse of that biological
parent is the other biological parent is more difficult.

Prior work has used a variety of additional restrictions to limit the
analysis to children who are more likely to be the biological children
of both parents (Saenz et al. 1995; Xie and Goyette 1997). Consistent
with this prior work, I restrict cases in two ways. First, I restrict
the sample to cases where both parents were of a reasonable age at the
birth of the child (aged 15-44 for mothers and aged 15-60 for fathers).
Second, I restrict the sample to those cases where the reported race and
Hispanicity of the child is inclusive of at least one of the parents'
races.

The final analytical sample consists of children aged 5-20 who are
currently enrolled in the K-12 school system and live in a household
with two parents who are both likely to be biological parents. This
necessary sample restriction will bias the family structure of the
analytical sample relative to the total population of K-12
schoolchildren. Children living in two-parent nuclear households have
greater educational success than other children as a result of material
and social resources (Biblarz and Raftery 1999). Therefore, this sample
restriction will bias overall estimates of grade retention downward.
However, this bias is less problematic for the goal of this study, which
is to estimate differences between groups. The sample restriction
effectively eliminates differences in family structure across racial
groups that may account for some observed differences in grade
retention. However, the remaining differences observed here are likely
to hold for the full population in the absence of strong racial
differences in the effect of family structure on grade retention.

The race of children is calculated by a cross-tabulation of parents'
races. To produce reasonably parsimonious categories, I collapse each
parent's race into the categories of White, Black, Asian, Latino, and
Indigenous. These categories form the ``ethnoracial pentagon'' that is
commonly used in popular practice and government tabulation to identify
race in the US (Hollinger 1995). Parents are identified as Latino based
on their response to the Hispanicity question, regardless of their
response to the race question. Parents are identified as Indigenous if
they responded as either American Indian/Alaska Native or as Pacific
Islander. Because the goal of my analysis is to examine specifically the
outcomes of ``first-generation'' biracial children, I exclude cases
where at least one parent identified with multiple races. The
cross-tabulation of parents' races leads to ten distinct biracial
categories. The total sample size for each of these biracial categories
is shown in Table~\ref{tbl-descriptive}, along with the sample size of
the monoracial comparison groups.

\hypertarget{tbl-descriptive}{}
\begin{table}
\caption{\label{tbl-descriptive}Descriptive survey-weighted statistics by race. Shading indicates
biracial group. Groups ordered by sample size. }\tabularnewline

\centering
\begin{tabular}[t]{lrr}
\toprule
Race & Sample size & Clearly behind expected grade\\
\midrule
White & 1,703,079 & 3.51\%\\
Latino & 386,718 & 5.11\%\\
Asian & 146,479 & 2.56\%\\
Black & 130,849 & 5.45\%\\
\cellcolor[HTML]{D3D3D3}{White/Latino} & \cellcolor[HTML]{D3D3D3}{120,369} & \cellcolor[HTML]{D3D3D3}{3.13\%}\\
\cellcolor[HTML]{D3D3D3}{White/Asian} & \cellcolor[HTML]{D3D3D3}{38,014} & \cellcolor[HTML]{D3D3D3}{1.82\%}\\
\cellcolor[HTML]{D3D3D3}{Black/White} & \cellcolor[HTML]{D3D3D3}{26,829} & \cellcolor[HTML]{D3D3D3}{3.86\%}\\
Indigenous & 16,224 & 7.86\%\\
\cellcolor[HTML]{D3D3D3}{White/Indigenous} & \cellcolor[HTML]{D3D3D3}{13,689} & \cellcolor[HTML]{D3D3D3}{5.25\%}\\
\cellcolor[HTML]{D3D3D3}{Black/Latino} & \cellcolor[HTML]{D3D3D3}{9,578} & \cellcolor[HTML]{D3D3D3}{3.93\%}\\
\cellcolor[HTML]{D3D3D3}{Latino/Asian} & \cellcolor[HTML]{D3D3D3}{5,570} & \cellcolor[HTML]{D3D3D3}{2.43\%}\\
\cellcolor[HTML]{D3D3D3}{Indigenous/Latino} & \cellcolor[HTML]{D3D3D3}{2,627} & \cellcolor[HTML]{D3D3D3}{5.25\%}\\
\cellcolor[HTML]{D3D3D3}{Black/Asian} & \cellcolor[HTML]{D3D3D3}{2,084} & \cellcolor[HTML]{D3D3D3}{2.98\%}\\
\cellcolor[HTML]{D3D3D3}{Black/Indigenous} & \cellcolor[HTML]{D3D3D3}{701} & \cellcolor[HTML]{D3D3D3}{5.96\%}\\
\cellcolor[HTML]{D3D3D3}{Indigenous/Asian} & \cellcolor[HTML]{D3D3D3}{521} & \cellcolor[HTML]{D3D3D3}{2.60\%}\\
\bottomrule
\end{tabular}
\end{table}

Grade retention is difficult to measure accurately. In most data
sources, researchers lack specific reports of grade retention and
instead infer grade retention from a discrepancy between a student's
reported age and grade. Prior studies have used the concept of a student
being behind modal grade (Bianchi 1984; Frederick and Hauser 2008). The
most notable limitation of this approach is the ambiguity created by the
fact that at any given age, a student may reasonably be in two modal
grades. Without detailed information on birthdates and survey timing,
the correct modal grade for most students cannot be identified. This
issue has been somewhat alleviated in prior work by the use of the
October supplement to the Current Population Survey which is close to
the beginning of the school year.

The ACS data do not provide the ability to identify or limit survey
timing, so I instead use a related measure of whether a given student is
clearly behind expected grade (CBEG). A student is considered CBEG if
their age is higher than either of the expected ages for a student of
that grade. This measure will underestimate overall grade retention
because it will miss students who have been retained but not yet had a
birthday in their current grade that would place them CBEG. However, the
goal of this study is not to estimate grade retention accurately, but
rather to understand racial differences in grade retention. Because this
bias is largely a function of when students have birthdays, it should be
more or less random with regard to sociodemographic characteristics.
This approach is equivalent to the method used by Rosenfeld (2010) on
2000 Census data, but more detailed information on current grade in
recent ACS data allows for more precise estimation.
Table~\ref{tbl-descriptive} shows the percent of each monoracial and
biracial group that is CBEG.

Using grade-age comparisons to infer grade retention may inadvertently
capture cases of ``academic redshirting'' in which parents intentionally
delay their child's kindergarten enrollment by a year. Frederick and
Hauser (2008) suggest that academic redshirting is in many cases be a
form of preemptive retention for children with developmental delays and
thus the error induced by these false positive cases may be minimal.
However, Bassok and Reardon (2013) have shown that academic redshirting
is more common among white and SES privileged parents, suggesting that
the demographic covariates of academic redshirting may operate in the
opposite direction of grade retention. Regardless, academic redshirting
should be of minimal concern for the measure of CBEG used here. Academic
redshirting most frequently occurs for students whose birthdays fall
close to the cutoff period for enrollment (Bassok and Reardon 2013;
Graue and DiPerna 2000). Therefore, academic redshirts spend most of a
given school year at the older, but correct, modal age for their grade
and most of these students will not be identified as CBEG.\footnote{Analysis
  of the probability of being CBEG by race and grade, shown in the
  supplementary materials, indicates that Black and Latino children in
  kindergarten are less likely to be CBEG than White children, which may
  be a consequence of redshirting. However, these differences quickly
  reverse direction by first grade. As a sensitivity analysis, I
  repeated the main analysis shown here separately for students in
  elementary (1st-5th), middle school (6th-8th), and high school
  (9th-12th) grades. These results are presented in the supplementary
  materials. Although more statistically noisy due to smaller sample
  sizes, those results are consistent with the conclusions drawn here.}

My goal is to understand the risk of being CBEG for each biracial group
relative to its constituent monoracial groups (e.g.~a Black/White
student compared to White and Black students). Throughout this article,
I use a visual approach to illustrate this placement. As an example, I
display the percent of White, Black, and Black/White students who are
CBEG in Figure~\ref{fig-example}. I display confidence bands/bars around
all three point estimates, but these confidence ranges require some
explanation as they are not standard 95\% confidence intervals. My goal
is to determine whether the biracial group's point estimate is
statistically distinguishable from a given monoracial constituent
group's point estimate by non-overlap of confidence intervals. Overlap
in 95\% confidence intervals is not equivalent to failing a hypothesis
test of difference at \(p<0.05\) and when used in this way, such
assessments lead to far more stringent tests (Knol, Pestman, and Grobbee
2011). When standard errors between groups are equal, overlap in 83.4\%
confidence intervals will achieve this goal. However, in cases where
standard errors vary substantially across estimates, the calculation of
the appropriate confidence interval is more complex and varies by each
pairwise comparison.\footnote{The formula for calculating the z-score
  for the correct interval is given by
  \[1.96 * \frac{\sqrt{1+\rho^2}}{1+\rho}\] where \(\rho\) is the ratio
  of the standard errors for the two statistics. See Knol et al. (2011)
  for a detailed derivation.}

\begin{figure}[t]

{\centering \includegraphics{main_files/figure-pdf/fig-example-1.pdf}

}

\caption{\label{fig-example}Probability of being clearly behind expected
grade for biracial black/white respondents in comparison to their
monoracial comparison groups. Non-overlap in color corresponding
confidence bands indicates statistically significant difference at the
5\% level.}

\end{figure}

I calculate two confidence intervals for each biracial group. Each
confidence interval is in comparison to one of the constituent
monoracial groups. These confidence intervals are color-coded in
Figure~\ref{fig-example} to indicate the reference group. I only draw
half-intervals in the direction of the monoracial group's point
estimate. Overlap in these color corresponding bars and bands indicates
that the difference between the biracial group and the monoracial group
is not statistically significant at \(p<.05\). For example, the yellow
bar shown in Figure~\ref{fig-example} for biracial Black/White students
does not quite overlap with the yellow band for White students,
indicating that the point estimates for these two groups are
statistically distinguishable at \(p<.05\).

Additionally, I include a measure of the ``halfway'' point between the
two monoracial groups in dark grey.\footnote{The halfway point is given
  by taking the mean between the two monoracial group. Its standard
  error is given by \(\sqrt{(s_1^2+s_2^2)/2}\) where \(s_1\) and \(s_2\)
  are the standard errors of the estimates for the two monoracial
  groups. To avoid clutter, I do not draw a third half-confidence
  interval for this point. Its length will be roughly halfway between
  the length of the other two intervals.} This halfway point allows me
to determine where the biracial group falls relative to the expectation
of being halfway between the two constituent monoracial groups. In this
case, Black/White students are much closer to White students in their
risk of grade retention, and have probabilities much lower than both
Black students and the halfway expectation.

Figure~\ref{fig-example} shows the raw differences between the three
racial groups of interest. However, in practice, I want to estimate
differences across groups while holding constant a variety of variables.
To do this, I estimate a set of logit models that predict the likelihood
of a student being CBEG by race and a variety of other variables. I then
construct figures similar to Figure~\ref{fig-example} by calculating
from the model the average predicted probabilities (APP) for each racial
group. An APP estimates the average probability of the outcome by a
given covariate across all cases, while holding constant all other
variables. APPs are akin to average marginal effects (AME) and estimated
in the same way. Whereas AMEs estimate differences or slopes, APPs
estimate the level for a given category.

I begin with a baseline model that adjusts for a variety of nuisance
characteristics that may vary across groups and thus need to be
controlled in all models. First, I include fixed effects for state of
residence because states dictate educational policy and thus can differ
substantially in the likelihood of grade retention. I also include dummy
variables indicating whether the student lived in a central city,
suburban, or rural area.\footnote{The location cannot be determined for
  all students and I therefore also include a fourth category of unknown
  in all models.}

The probability of being CBEG also increases with the student's current
grade, so I include fixed effects for the current grade of the student.
Additionally, as Figure~\ref{fig-year-grade} shows, the percent of
students who are CBEG has declined over time, but this decline has been
much more substantial at higher grade levels. For the elementary grades,
there is no evidence of a decline at all. To account for this
grade-specific decline in the models I include an interaction between a
linear year term and current grade. Sensitivity analysis showed that
this functional form was preferred by the Bayesian information criterion
to models with no interaction and a model with interaction terms that
treated year as a categorical variable. I also considered a similar
interaction between state and year, but this model was not
preferred.\footnote{The supplementary materials provide, for each model
  specification, the BIC scores and marginal probabilities of being
  clearly behind expected grade for each racial group. The marginal
  probabilities for each racial group are almost identical across
  specifications, indicating that results are not driven by the modeling
  choice made here.}

\begin{figure}[t]

{\centering \includegraphics{main_files/figure-pdf/fig-year-grade-1.pdf}

}

\caption{\label{fig-year-grade}Trends over time and grade in the percent
of students clearly behind expected grade. Lines are fit to each set of
points by grade via lowess smoothing.}

\end{figure}

The baseline model also includes the race of the respondent. I then add
additional terms measuring material and cultural resources that may
account for racial differences in a set of subsequent models. First, I
include measures of nativity and English proficiency for both the
respondent and each of their parents. Second, I include a categorical
measure of highest degree earned for each parent. Finally, I include
measures of family income (square rooted), home ownership, and whether
the parents are married.

All statistical analysis adjusts for sampling weights among respondents.
All models incorporate design effects for variance in sample weights and
the clustering of multiple respondents within the same household.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{monoracial-differences}{%
\subsection{Monoracial Differences}\label{monoracial-differences}}

I begin by showing the relative risk of being CBEG among monoracial
respondents across models. Understanding these differences helps to
clarify the potential for how biracial respondents might be positioned
between monoracial groups. Table 2 reports average marginal effects on
the probability of being CBEG for each monoracial minority group in
comparison to White students.

\hypertarget{tbl-monoracial}{}
\begin{table}[t]
\caption{\label{tbl-monoracial}Average marginal differences in the probability of being clearly behind
expected grade by monoracial group, relative to a monoracial white
student }
\begin{center}
\begin{tabular}{l c c c c}
\hline
 & Model 1 & Model 2 & Model 3 & Model 4 \\
\hline
Black                  & $0.015^{***}$  & $0.016^{***}$  & $0.010^{***}$  & $0.007^{***}$  \\
                       & $(0.001)$      & $(0.001)$      & $(0.001)$      & $(0.001)$      \\
Indigenous             & $0.039^{***}$  & $0.039^{***}$  & $0.026^{***}$  & $0.022^{***}$  \\
                       & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      \\
Asian                  & $-0.003^{***}$ & $-0.006^{***}$ & $-0.005^{***}$ & $-0.005^{***}$ \\
                       & $(0.001)$      & $(0.001)$      & $(0.001)$      & $(0.001)$      \\
Latino                 & $0.022^{***}$  & $0.013^{***}$  & $-0.001$       & $-0.002^{**}$  \\
                       & $(0.001)$      & $(0.001)$      & $(0.001)$      & $(0.001)$      \\
\hline
State fixed effects    & Yes            & Yes            & Yes            & Yes            \\
Location fixed effects & Yes            & Yes            & Yes            & Yes            \\
Grade fixed effects    & Yes            & Yes            & Yes            & Yes            \\
Year linear effects    & Yes            & Yes            & Yes            & Yes            \\
Year x grade effects   & Yes            & Yes            & Yes            & Yes            \\
Nativity and language  & No             & Yes            & Yes            & Yes            \\
Parent's education     & No             & No             & Yes            & Yes            \\
Other family resources & No             & No             & No             & Yes            \\
N                      & $2603331$      & $2603331$      & $2603331$      & $2603331$      \\
\hline
\multicolumn{5}{l}{\scriptsize{$^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$}}
\end{tabular}
\label{tbl-monoracial}
\end{center}
\end{table}

Model 1 only controls for structural factors such as state of residence,
student location, grade, and year. These results provide a baseline
estimate of the differences across monoracial groups without controlling
for differences in cultural and material resources. The results show
that Black, Indigenous, and Latino students all have substantially
higher probability of being CBEG than White students. Indigenous
students have substantially higher risk than all other students, with a
probability of being CBEG that is 4.1\% percentage points higher than
White students. Asian students have the lowest probability of being
CBEG, and their risk is slightly lower than White students.

The subsequent models control for a variety of additional variables.
Model 2 includes controls for whether the student and each of their
parents are native-born and speak English well. Model 3 controls for the
highest degree received by each parent, and Model 4 controls for family
income (square rooted), home ownership, and whether the student's
parents are married.

Model 2 and Model 3 have substantial effects on the observed inequality
across racial groups, while the additional variables from Model 4 have
less impact. Controlling for the characteristics related to immigration
and acculturation in Model 2 cuts the gap between White and Latino
students in half and increases the gap between Asian and White students.
It has no impact on the gap between White and Black or White and
Indigenous students because most of these students are native-born with
native-born parents.

Controlling for education in Model 3 reduces the gap substantially for
Black and Indigenous students and removes the gap entirely for Latino
students. In fact, the results of both Models 3 and 4 suggest that, when
holding constant cultural and family resources, Latino students are
slightly less likely than White students to be CBEG.

In total, resource differentials account for a substantial part of the
overall racial differences in the risk of being CBEG, but not its
entireity. Black and Indigenous students remain at higher risk than
other students. Asian students remain the group with lowest risk across
all models. The most notable shift across models is for Latino students.
Their higher risk of being CBEG relative to White students is completely
accounted for by resource differentials. When comparing White and Latino
students with the same level or resources in Model 4, Latino students
actually have slightly lower risk of being CBEG.

Although, the racial differences in risk of CBEG are substantially
reduced by controlling for resource differentials, the remaining gaps
across all pairwise combinations are still large enough in most cases to
sustain the question of where biracial individuals fit within these
gaps. Before turning to this question, however, I want to better
understand how the resources of biracial individuals compare to their
monoracial constituent groups.

\hypertarget{the-distribution-of-resources-for-biracial-individuals}{%
\subsection{The Distribution of Resources for Biracial
Individuals}\label{the-distribution-of-resources-for-biracial-individuals}}

Table~\ref{tbl-resources} shows the mean values of the resource
variables across monoracial and biracial groups.\footnote{For
  compactness, I combine the parent foreign-born and English proficiency
  questions into whether either parent is foreign-born or speaks
  English. In the models, I use separate variables for mothers and
  fathers. Tables in the supplementary materials provide the full
  breakdown of these variables by racial group.} I also calculate a
counterfactual probability of CBEG for each racial group based solely on
their observed distribution of resources. Differences in this
counterfactual probability provide a summary measure of the resource
differentials between groups. This counterfactual probability is
calculated by estimating the average predicted probability for each
group from Model 4 of Table 2 when all non-resource variables, including
race, are held at their mean.\footnote{It might seem odd to hold
  categorical variables at their mean, but because all categorical
  variables are entered into the models as 0/1 numeric indicator
  variables, their mean value can be calculated as the proportion of
  cases with the indicated value.}

\hypertarget{tbl-resources}{}
\begin{table}
\caption{\label{tbl-resources}Mean resources by racial group. All results are survey weighted. Results
sorted by counterfactual probability of being clearly behind expected
grade. Shading indicates biracial group. }\tabularnewline

\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}}
\toprule
Race & CBEG, counterfactual & Family income & Own home & Mother, 4-year college degree & Father, 4-year college degree & Foreign-born & Either parent foreign-born & Speak English well & Both parents speak english well\\
\midrule
\cellcolor[HTML]{D3D3D3}{White/Asian} & \cellcolor[HTML]{D3D3D3}{2.4\%} & \cellcolor[HTML]{D3D3D3}{\$187,049} & \cellcolor[HTML]{D3D3D3}{85.2\%} & \cellcolor[HTML]{D3D3D3}{63.9\%} & \cellcolor[HTML]{D3D3D3}{65.5\%} & \cellcolor[HTML]{D3D3D3}{6.2\%} & \cellcolor[HTML]{D3D3D3}{73.3\%} & \cellcolor[HTML]{D3D3D3}{99.4\%} & \cellcolor[HTML]{D3D3D3}{98.8\%}\\
\cellcolor[HTML]{D3D3D3}{Black/Asian} & \cellcolor[HTML]{D3D3D3}{2.9\%} & \cellcolor[HTML]{D3D3D3}{\$131,156} & \cellcolor[HTML]{D3D3D3}{64.6\%} & \cellcolor[HTML]{D3D3D3}{44.3\%} & \cellcolor[HTML]{D3D3D3}{39.4\%} & \cellcolor[HTML]{D3D3D3}{7.3\%} & \cellcolor[HTML]{D3D3D3}{81.7\%} & \cellcolor[HTML]{D3D3D3}{99.4\%} & \cellcolor[HTML]{D3D3D3}{98.6\%}\\
\cellcolor[HTML]{D3D3D3}{Latino/Asian} & \cellcolor[HTML]{D3D3D3}{2.9\%} & \cellcolor[HTML]{D3D3D3}{\$130,330} & \cellcolor[HTML]{D3D3D3}{70.3\%} & \cellcolor[HTML]{D3D3D3}{43.4\%} & \cellcolor[HTML]{D3D3D3}{40.5\%} & \cellcolor[HTML]{D3D3D3}{4.5\%} & \cellcolor[HTML]{D3D3D3}{71.1\%} & \cellcolor[HTML]{D3D3D3}{99.5\%} & \cellcolor[HTML]{D3D3D3}{96.8\%}\\
White & 3.1\% & \$136,953 & 83.3\% & 46.7\% & 43.6\% & 1.9\% & 9.4\% & 99.6\% & 98.9\%\\
Asian & 3.1\% & \$141,982 & 72.4\% & 56.7\% & 59.6\% & 21.9\% & 96.2\% & 96.5\% & 76.8\%\\
\cellcolor[HTML]{D3D3D3}{White/Latino} & \cellcolor[HTML]{D3D3D3}{3.3\%} & \cellcolor[HTML]{D3D3D3}{\$125,805} & \cellcolor[HTML]{D3D3D3}{73.0\%} & \cellcolor[HTML]{D3D3D3}{38.8\%} & \cellcolor[HTML]{D3D3D3}{36.1\%} & \cellcolor[HTML]{D3D3D3}{2.0\%} & \cellcolor[HTML]{D3D3D3}{32.9\%} & \cellcolor[HTML]{D3D3D3}{99.3\%} & \cellcolor[HTML]{D3D3D3}{98.5\%}\\
\cellcolor[HTML]{D3D3D3}{White/Indigenous} & \cellcolor[HTML]{D3D3D3}{3.6\%} & \cellcolor[HTML]{D3D3D3}{\$97,664} & \cellcolor[HTML]{D3D3D3}{70.7\%} & \cellcolor[HTML]{D3D3D3}{28.2\%} & \cellcolor[HTML]{D3D3D3}{24.2\%} & \cellcolor[HTML]{D3D3D3}{1.1\%} & \cellcolor[HTML]{D3D3D3}{9.6\%} & \cellcolor[HTML]{D3D3D3}{99.8\%} & \cellcolor[HTML]{D3D3D3}{99.9\%}\\
\cellcolor[HTML]{D3D3D3}{Indigenous/Asian} & \cellcolor[HTML]{D3D3D3}{3.6\%} & \cellcolor[HTML]{D3D3D3}{\$100,783} & \cellcolor[HTML]{D3D3D3}{56.1\%} & \cellcolor[HTML]{D3D3D3}{28.8\%} & \cellcolor[HTML]{D3D3D3}{22.6\%} & \cellcolor[HTML]{D3D3D3}{9.9\%} & \cellcolor[HTML]{D3D3D3}{71.3\%} & \cellcolor[HTML]{D3D3D3}{100.0\%} & \cellcolor[HTML]{D3D3D3}{97.0\%}\\
\cellcolor[HTML]{D3D3D3}{Black/White} & \cellcolor[HTML]{D3D3D3}{3.7\%} & \cellcolor[HTML]{D3D3D3}{\$94,841} & \cellcolor[HTML]{D3D3D3}{57.6\%} & \cellcolor[HTML]{D3D3D3}{31.9\%} & \cellcolor[HTML]{D3D3D3}{27.1\%} & \cellcolor[HTML]{D3D3D3}{2.0\%} & \cellcolor[HTML]{D3D3D3}{17.0\%} & \cellcolor[HTML]{D3D3D3}{99.8\%} & \cellcolor[HTML]{D3D3D3}{99.7\%}\\
Black & 3.8\% & \$87,159 & 53.9\% & 30.9\% & 25.1\% & 6.4\% & 27.7\% & 99.4\% & 96.2\%\\
\cellcolor[HTML]{D3D3D3}{Black/Latino} & \cellcolor[HTML]{D3D3D3}{3.8\%} & \cellcolor[HTML]{D3D3D3}{\$86,825} & \cellcolor[HTML]{D3D3D3}{47.2\%} & \cellcolor[HTML]{D3D3D3}{25.8\%} & \cellcolor[HTML]{D3D3D3}{23.4\%} & \cellcolor[HTML]{D3D3D3}{2.0\%} & \cellcolor[HTML]{D3D3D3}{32.4\%} & \cellcolor[HTML]{D3D3D3}{98.9\%} & \cellcolor[HTML]{D3D3D3}{98.5\%}\\
\cellcolor[HTML]{D3D3D3}{Black/Indigenous} & \cellcolor[HTML]{D3D3D3}{4.1\%} & \cellcolor[HTML]{D3D3D3}{\$76,097} & \cellcolor[HTML]{D3D3D3}{44.7\%} & \cellcolor[HTML]{D3D3D3}{19.3\%} & \cellcolor[HTML]{D3D3D3}{21.5\%} & \cellcolor[HTML]{D3D3D3}{3.9\%} & \cellcolor[HTML]{D3D3D3}{23.6\%} & \cellcolor[HTML]{D3D3D3}{100.0\%} & \cellcolor[HTML]{D3D3D3}{99.9\%}\\
\cellcolor[HTML]{D3D3D3}{Indigenous/Latino} & \cellcolor[HTML]{D3D3D3}{4.4\%} & \cellcolor[HTML]{D3D3D3}{\$72,164} & \cellcolor[HTML]{D3D3D3}{47.8\%} & \cellcolor[HTML]{D3D3D3}{14.4\%} & \cellcolor[HTML]{D3D3D3}{10.0\%} & \cellcolor[HTML]{D3D3D3}{1.1\%} & \cellcolor[HTML]{D3D3D3}{29.9\%} & \cellcolor[HTML]{D3D3D3}{99.3\%} & \cellcolor[HTML]{D3D3D3}{97.5\%}\\
Indigenous & 4.5\% & \$63,115 & 50.7\% & 13.2\% & 8.8\% & 7.1\% & 23.6\% & 99.2\% & 96.1\%\\
Latino & 5.3\% & \$62,226 & 50.2\% & 10.4\% & 8.5\% & 11.6\% & 83.0\% & 96.4\% & 50.8\%\\
\bottomrule
\end{tabular}}
\end{table}

The results show different risks of being CBEG across biracial groups
because of differences in resources. For example, White/Asian students
have the lowest counterfactual risk of being CBEG of all groups at
2.4\%, owing to their exceptionally high family income and the high
college attainment of their parents. At the other end of the spectrum,
Indigenous/Latino students have a counterfactual risk of CBEG nearly 2\%
higher, given their much lower family resources.

I am particularly interested in how each biracial group compares to its
constituent monoracial groups. For example, while Indigenous/Latino
students have the highest counterfactual risk of being CBEG among
biracial students, their risk is still lower than their two constituent
groups of Indigenous (4.5\%) and Latino (5.3\%) students, owing in large
part to somewhat higher family income, greater parental educational
attainment, and greater acculturation. These results suggest positive
selection into Indigenous/Latino interracial unions.

\hypertarget{tbl-resource-summary}{}
\begin{table}
\caption{\label{tbl-resource-summary}Relative placement for biracial groups in comparison to monoracial
constituent groups based on counterfactual probability of being clearly
behind expected grade due to resources alone. }\tabularnewline

\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
\toprule
Biacial & Lower probability than both & Similar probability as lower & Between & Similar probability as higher & Higher probability than both\\
\midrule
White/Asian & X &  &  &  & \\
Black/Asian &  & X (Asian) &  &  & \\
Latino/Asian &  & X (Asian) &  &  & \\
White/Latino &  & X (White) &  &  & \\
Black/Latino &  & X (Black) &  &  & \\
Indigenous/Latino &  & X (Indig.) &  &  & \\
White/Indigenous &  &  & X &  & \\
Black/Indigenous &  &  & X &  & \\
Indigenous/Latino &  &  & X &  & \\
Black/White &  &  &  & X (Black) & \\
\bottomrule
\end{tabular}
\end{table}

Table 4 summarizes the patterns from Table~\ref{tbl-resources} across
all ten biracial groups. In six of these cases, the biracial group would
have outcomes similar to or better than the monoracial group with better
resources. These results imply that for the majority of these groups,
parents in these matches are being positively selected from their
constituent groups.

Only three groups exhibit an in-between status based on resources. All
three of these groups are part-Indigenous. This finding does not hold
for Indigenous/Latino students, but nonetheless suggests less
selectivity in crossing this boundary.

The Black/White case stands out as an outlier. Black/White students have
similar resources as Black students and both groups have fewer resources
than White students, as can be seen in Table~\ref{tbl-resources} for
family income, home ownership, and parental educational attainment.
Unlike most of these other groups, I observe no positive class
selectivity into Black/White interracial unions.

The effect of controlling for these resources will be complex and
different across biracial groups. For those biracial groups with
resources similar to their more advantaged constituent group,
controlling for these resources should reduce their advantage relative
to the less advantaged monoracial group. For the Black/White case, on
the other hand, controlling for resource differentials should eliminate
some of their disadvantage relative to Whites. For the remaining
``in-between'' groups, it is more difficult to make a priori
predictions.

\hypertarget{biracial-placement}{%
\subsection{Biracial Placement}\label{biracial-placement}}

I now turn to the risk of CBEG for biracial respondents. Due to the
number of comparisons being made, I split these results into two
separate figures. In Figure~\ref{fig-black-models}, I show the placement
of each biracial group involving one black parent in comparison to their
two monoracial constituent groups. To allow comparison across models, I
compare the results from the baseline model (Model 1 in Table 2) and the
full model that accounts for all cultural and family resources (Model 4
in Table 2).

\begin{figure}[p]

{\centering \includegraphics{main_files/figure-pdf/fig-black-models-1.pdf}

}

\caption{\label{fig-black-models}Probability of being clearly behind
expected grade for biracial respondents with one black parent, in
comparison to their monoracial comparison groups. Non-overlap in color
corresponding confidence bands indicates statistically significant
difference at the 5\% level, Baseline model includes year, grade,
location, and state fixed effects. Full models include control variables
for nativity, English proficiency, income, education, home ownership,
and marital status of parents.}

\end{figure}

I focus first on biracial groups with one Black parent because of a
stronger historical expectation that these students will be identified
as Black by the ``one-drop rule'' and as a result will have similar
outcomes to Black students. Figure~\ref{fig-black-models} provides
little evidence for such a pattern, and soundly rejects it in the case
of Black/White and Black/Latino students. Black/White and Black/Latino
students have a risk of being CBEG substantially lower than Black
students and closer to their non-Black constituent group. Solid
conclusions are difficult for the remaining two groups because of
smaller sample size and correspondingly wider confidence intervals on
estimates. Nonetheless, the point estimates in each of these cases
suggest probabilities of being CBEG roughly halfway between the two
constituent groups.

Controlling for resource variables has some effect on the placement of
part-Black biracial respondents. Black/White biracial students initially
have a risk of being CBEG slightly higher than and statistically
distinguishable from White students. After controlling for cultural and
family resources, however, Black/White students have a risk of being
CBEG that is slightly lower but not statistically distinguishable from
White students. This change reflects the relatively low level of
parental education and income among Black/White students, which is much
closer to Black students than White students, as shown on
Table~\ref{tbl-resources}.

Black/Latino students start from a very different position, having a
risk of being CBEG substantially lower than both Black and Latino
students, who have somewhat similar risks. However, after controlling
for cultural and family resources, Black/Latino students have a risk of
being CBEG that is statistically indistinguishable from the risk of
Latino students and substantially lower than and statistically
distinguishable from Black students. This change largely reflects the
fact that Black/Latino students have greater resources than Latino (but
not Black) students and once this advantage is held constant, they have
a similar risk. After controlling for family resources, both Black/White
and Black/Latino students have a risk of being CBEG similar to their
monoracial constituent group with lower risk.

The role of parental resources is harder to determine for Black/Asian
and Black/Indigenous students due to the greater uncertainty in these
estimates. Controlling for resources moves the point estimate for
Black/Asian students from a risk similar to Asian students to more of a
halfway position between the two constituent groups. Black/Indigenous
students have risk of being CBEG roughly halfway between their two
constituent groups in both models. However, this finding of a halfway
position for both groups is highly tentative owing to wide confidence
intervals.

\begin{figure}[p]

{\centering \includegraphics{main_files/figure-pdf/fig-other-models-1.pdf}

}

\caption{\label{fig-other-models}Probability of being clearly behind
expected grade for non-black biracial respondents, in comparison to
their monoracial comparison groups. Non-overlap in color corresponding
confidence bands indicates statistically significant difference at the
5\% level, Baseline model includes year, grade, location, and state
fixed effects. Full models include control variables for nativity,
English proficiency, income, education, home ownership, and marital
status of parents.}

\end{figure}

Figure~\ref{fig-other-models} shows the results for the remaining
biracial groups. For most of these groups, the full models indicate that
each biracial group has a risk of being CBEG similar to the risk of the
lower monoracial constituent group or has risk higher than this group
but less than the halfway point. There are two exceptions to this
general pattern. First, the point estimate of the probability of being
CBEG for the Indigenous/Latino group is roughly halfway between the two
constituent groups, although it is not statistically distinguishable
from the lower group. Second, Latino/Asian students are the only case
where the point estimate of the risk for the biracial group is higher
than either monoracial group. However, the confidence bands suggest that
the risk for this group cannot be statistically differentiated from
either of the two monoracial constituent groups. This result only
emerges in the model that controls for resources, because Latino/Asian
students come from more advantaged households than both Latino and Asian
students, on average.

Aside from the case of Latino/Asian students, controlling for resources
has moderate effects on the placement of the remaining biracial groups.
White/Asian students are the only case with a substantial shift, owing
to their highly advantaged households. Initially White/Asian students
have risk of being CBEG substantially below both White and Asian
students. After controlling for resources, the risk of being CBEG for
White/Asian students is indistinguishable from those for Asian students,
but still substantially and statistically distinguishable from the
higher risk to White students.

Results for part-Latino students are complicated by the substantial
change in the monoracial Latino risk across models. For example, while
Latino students have substantially higher risk of being CBEG than White
students in the baseline model, there is only a very small difference
favoring Latino students in the full model. Nonetheless, even in this
case, we see that White/Latino students have risk closer to those of the
lower Latino group, but distinguishable from both groups.

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

In this article, I have used the risk of grade retention among K-12
students to better understand how biracial students' life chances
compare to their monoracial peers. Unlike prior work on this topic, the
estimates used here for most biracial groups are relatively precise due
to large sample sizes. For groups with a high level of precision, the
results tell a consistent story - biracial students' risk of grade
retention is similar to their monoracial constituent group with the
lower risk.

These results contradict expectations of both a ``halfway'' position and
the dominance of the one drop rule. For no group do I observe clear
evidence of a pattern of hypodescent in which the risk was similar to
the higher risk constituent monoracial group. These results also
strongly contradict the expectations of the marginal man hypothesis.

Two possibilities emerge that may help us understand these results
better. First, we may be witnessing a shift toward a new regime of
hyperdescent that applies broadly across a wide variety of racial
groups. Second, the relatively low risk for biracial students may be a
product of the strength of hybridity in allowing biracial students to
better negotiate racialized systems than their monoracial minority
peers.

Differences in family resources were expected to play some role in the
relative position of biracial students' risk. The results on parent
selection vary somewhat by group, but the majority of biracial groups
had resources more similar to their monoracial constituent group with
higher resources, indicating a strong positive selection into
interracial unions in the prior generation. The most important exception
to this trend is the case of Black/White biracial students who have far
fewer resources than White students and resources closer to Black
students.

Regardless of variation in biracial students' parental resources,
accounting for these resource differentials does not substantially
change the overall result that biracial students' risk is more similar
to the constituent monoracial group with lower risk. The one exception
to this finding is for the Latino/Asian case where the much lower risk
of these students compared to Latino and Asian students was entirely
driven by resource differentials.

Low relative risk for biracial individuals is sometimes treated as an
indication of a positive future direction for the US in terms of
ameliorating racial inequality. However, the improved prospects for
biracial individuals do nothing to ameliorate the often strong and
persistent inequalities between monoracial groups. The growth and
relative success of biracial populations may isolate remaining members
of monoracial minority groups as much as it assimilates mixed race
individuals. The real question at stake is how existing divides and
identities may be restructured in the context of a growing biracial
population. This research suggests the potential for a growing divide
between these mixed race populations and the most disadvantaged
monoracial minority populations from which they derive at least part of
their ancestry.

These findings stand in contrast to earlier work which found more
evidence of an ``in-between'' status in other educational outcomes
(Campbell 2009; Herman 2009). I raise two possible explanations for this
discrepancy that point to strengths and weaknesses of the current
research and may help drive future work on the topic. First, this
discrepancy may be due to the small sample sizes of multiracial
respondents in prior work which led to statistically imprecise
estimates. Because the current findings rely upon large samples, they
present an important step forward in our understanding of how biracial
individuals will fit into America's system of racial inequality.
Nonetheless, sample sizes for some non-white biracial respondents remain
small, limiting our understanding of the outcomes for those groups. In
particular, results for non-white part-Indigenous populations are quite
imprecise. The results counter-intuitively suggest that part-Indigenous
populations are the least likely to be in the position expected by
hyperdescent, but due to low sample sizes, that finding is highly
tentative.

Second, The current findings may also diverge from the findings of prior
work because different outcomes were examined. While grade retention is
an important early life outcome, it is far from the only one that these
individuals will face in their life. Although, we often observe similar
racial inequalities across a variety of outcomes, knowing the result for
one outcome does not perfectly predict other outcomes. As biracial
individuals age and face later outcomes in the educational system, the
labor market, and elsewhere, their experiences may differ from what I
observe here. Thus, collecting more detailed information on these
outcomes in future studies remains critically important.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-alba_great_2020}{}}%
Alba, Richard. 2020.
\emph{\href{https://press.princeton.edu/books/hardcover/9780691201634/the-great-demographic-illusion}{The
{Great Demographic Illusion}}}. {Princeton University Press}.

\leavevmode\vadjust pre{\hypertarget{ref-andrew_scarring_2014}{}}%
Andrew, Megan. 2014. {``The {Scarring Effects} of {Primary-Grade
Retention}? {A Study} of {Cumulative Advantage} in the {Educational
Career}.''} \emph{Social Forces} 93(2):653--85. doi:
\href{https://doi.org/10.1093/sf/sou074}{10.1093/sf/sou074}.

\leavevmode\vadjust pre{\hypertarget{ref-bassok_academic_2013}{}}%
Bassok, Daphna, and Sean F. Reardon. 2013. {``{`{Academic Redshirting}'}
in {Kindergarten}: {Prevalence}, {Patterns}, and {Implications}.''}
\emph{Educational Evaluation and Policy Analysis} 35(3):283--97. doi:
\href{https://doi.org/10.3102/0162373713482764}{10.3102/0162373713482764}.

\leavevmode\vadjust pre{\hypertarget{ref-bianchi_children_1984}{}}%
Bianchi, Suzanne M. 1984. {``Children's {Progress Through School}: {A
Research Note}.''} \emph{Sociology of Education} 57(3):184--92. doi:
\href{https://doi.org/10.2307/2112602}{10.2307/2112602}.

\leavevmode\vadjust pre{\hypertarget{ref-biblarz_family_1999}{}}%
Biblarz, Timothy J., and Adrian E. Raftery. 1999. {``Family {Structure},
{Educational Attainment}, and {Socioeconomic Success}: {Rethinking} the
{`{Pathology} of {Matriarchy}'}.''} \emph{American Journal of Sociology}
105(2):321--65.

\leavevmode\vadjust pre{\hypertarget{ref-bonilla-silva_biracial_2004}{}}%
Bonilla-Silva, Eduardo. 2004. {``From Bi-Racial to Tri-Racial: {Towards}
a New System of Racial Stratification in the {USA}.''} \emph{Ethnic and
Racial Studies} 27(6):931--50.

\leavevmode\vadjust pre{\hypertarget{ref-bonilla-silva_racism_2006}{}}%
Bonilla-Silva, Eduardo. 2006.
\emph{\href{https://books.google.com?id=HKJE4rVZG1EC}{Racism Without
{Racists}: {Color-Blind Racism} and the {Persistence} of {Racial
Inequality} in the {United States}}}. {Rowman \& Littlefield
Publishers}.

\leavevmode\vadjust pre{\hypertarget{ref-bratter_multiracial_2018}{}}%
Bratter, Jenifer L. 2018. {``Multiracial {Identification} and {Racial
Gaps}: {A Work} in {Progress}.''} \emph{The ANNALS of the American
Academy of Political and Social Science} 677(1):69--80. doi:
\href{https://doi.org/10.1177/0002716218758622}{10.1177/0002716218758622}.

\leavevmode\vadjust pre{\hypertarget{ref-bratter_poverty_2013}{}}%
Bratter, Jenifer L., and Sarah Damaske. 2013. {``Poverty at a {Racial
Crossroads}: {Poverty Among Multiracial Children} of {Single
Mothers}.''} \emph{Journal of Marriage and Family} 75(2):486--502. doi:
\href{https://doi.org/10.1111/jomf.12012}{10.1111/jomf.12012}.

\leavevmode\vadjust pre{\hypertarget{ref-bratter_does_2011}{}}%
Bratter, Jenifer L., and Bridget K. Gorman. 2011. {``Does {Multiracial
Matter}? {A Study} of {Racial Disparities} in {Self-Rated Health}.''}
\emph{Demography} 48(1):127--52. doi:
\href{https://doi.org/10.1007/s13524-010-0005-0}{10.1007/s13524-010-0005-0}.

\leavevmode\vadjust pre{\hypertarget{ref-bratter_mother_2009}{}}%
Bratter, Jenifer, and Holly E. Heard. 2009.
{``\href{http://www.jstor.org/stable/40542696}{Mother's, {Father}'s, or
{Both}? {Parental Gender} and {Parent-Child Interactions} in the {Racial
Classification} of {Adolescents}}.''} \emph{Sociological Forum}
24(3):658--88.

\leavevmode\vadjust pre{\hypertarget{ref-bratter_multiracial_2013}{}}%
Bratter, Jenifer, and Rachel Tolbert Kimbro. 2013. {``Multiracial
{Children} and {Poverty}: {Evidence From} the {Early Childhood
Longitudinal Study} of~{Kindergartners}.''} \emph{Family Relations}
62(1):175--89. doi:
\href{https://doi.org/10.1111/j.1741-3729.2012.00750.x}{10.1111/j.1741-3729.2012.00750.x}.

\leavevmode\vadjust pre{\hypertarget{ref-brunsma_interracial_2005}{}}%
Brunsma, David L. 2005. {``Interracial {Families} and the {Racial
Identification} of {Mixed-Race Children}: {Evidence} from the {Early
Childhood Longitudinal Study}.''} \emph{Social Forces} 84(2):1131--57.
doi: \href{https://doi.org/10.2307/3598493}{10.2307/3598493}.

\leavevmode\vadjust pre{\hypertarget{ref-campbell_multiracial_2009}{}}%
Campbell, Mary E. 2009. {``Multiracial {Groups} and {Educational
Inequality}: {A Rainbow} or a {Divide}?''} \emph{Social Problems}
56(3):425--46.

\leavevmode\vadjust pre{\hypertarget{ref-campbell_what_2006}{}}%
Campbell, Mary E., and Jennifer Eggerling-Boeck. 2006. {``"{What About}
the {Children}?" {The Psychological} and {Social Well-Being} of
{Multiracial Adolescents}.''} \emph{The Sociological Quarterly}
47(1):147--73.

\leavevmode\vadjust pre{\hypertarget{ref-cheng_multiracial_2009}{}}%
Cheng, Simon, and Kathryn J. Lively. 2009. {``Multiracial
{Self-Identification} and {Adolescent Outcomes}: {A Social Psychological
Approach} to the {Marginal Man Theory}.''} \emph{Social Forces}
88(1):61--98.

\leavevmode\vadjust pre{\hypertarget{ref-chew_american_1989}{}}%
Chew, Kenneth S. Y., David J. Eggebeen, and Peter R. Uhlenberg. 1989.
{``American {Children} in {Multiracial Households}.''}
\emph{Sociological Perspectives} 32(1):65--85. doi:
\href{https://doi.org/10.2307/1389008}{10.2307/1389008}.

\leavevmode\vadjust pre{\hypertarget{ref-conley_being_1999}{}}%
Conley, Dalton. 1999. \emph{Being {Black}. {Living} in the {Red}:
{Race}, {Wealth}, and {Social Policy} in {America}}. {Berkeley}:
{University of California Press}.

\leavevmode\vadjust pre{\hypertarget{ref-davenport_role_2016}{}}%
Davenport, Lauren D. 2016.
{``\href{http://asr.sagepub.com/content/81/1/57.short}{The {Role} of
{Gender}, {Class}, and {Religion} in {Biracial Americans}' {Racial
Labeling Decisions}}.''} \emph{American Sociological Review}
81(1):57--84.

\leavevmode\vadjust pre{\hypertarget{ref-davis_who_1991}{}}%
Davis, F. James. 1991. \emph{Who Is {Black}?: {One Nation}'s
{Definition}}. {University Park, PA}: {Pennsylvania State University
Press}.

\leavevmode\vadjust pre{\hypertarget{ref-davis_intermarriage_1941}{}}%
Davis, Kingsley. 1941. {``Intermarriage in {Caste Societies}.''}
\emph{American Anthropologist} 43:376--95.

\leavevmode\vadjust pre{\hypertarget{ref-dimaggio_culture_1997}{}}%
DiMaggio, Paul. 1997.
{``\href{http://www.jstor.org/stable/2952552}{Culture and
{Cognition}}.''} \emph{Annual Review of Sociology} 23:263--87.

\leavevmode\vadjust pre{\hypertarget{ref-doyle_are_2007}{}}%
Doyle, Jamie Mihoko, and Grace Kao. 2007.
{``\href{http://www.jstor.org/stable/20141804}{Are {Racial Identities}
of {Multiracials Stable}? {Changing Self-Identification} Among {Single}
and {Multiple Race Individuals}}.''} \emph{Social Psychology Quarterly}
70(4):405--23.

\leavevmode\vadjust pre{\hypertarget{ref-ellis_agents_2012}{}}%
Ellis, Mark, Steven R. Holloway, Richard Wright, and Christopher S.
Fowler. 2012. {``Agents of {Change}: {Mixed-Race Households} and the
{Dynamics} of {Neighborhood Segregation} in the {United States}.''}
\emph{Annals of the Association of American Geographers} 102(3):549--70.
doi:
\href{https://doi.org/10.1080/00045608.2011.627057}{10.1080/00045608.2011.627057}.

\leavevmode\vadjust pre{\hypertarget{ref-frederick_have_2008}{}}%
Frederick, Carl B., and Robert M. Hauser. 2008. {``Have We Put an End to
Social Promotion? {Changes} in School Progress Among Children Aged 6 to
17 from 1972 to 2005.''} \emph{Demography} 45(3):719--40. doi:
\href{https://doi.org/10.1353/dem.0.0015}{10.1353/dem.0.0015}.

\leavevmode\vadjust pre{\hypertarget{ref-fu_racial_2001}{}}%
Fu, Vincent Kang. 2001. {``Racial {Intermarriage Pairings}.''}
\emph{Demography} 38(2):147--59.

\leavevmode\vadjust pre{\hypertarget{ref-graue_redshirting_2000}{}}%
Graue, M. Elizabeth, and James DiPerna. 2000. {``Redshirting and {Early
Retention}: {Who Gets} the "{Gift} of {Time}" and {What Are} {Its
Outcomes}?''} \emph{American Educational Research Journal}
37(2):509--34. doi:
\href{https://doi.org/10.3102/00028312037002509}{10.3102/00028312037002509}.

\leavevmode\vadjust pre{\hypertarget{ref-gullickson_education_2006}{}}%
Gullickson, Aaron. 2006. {``Education and Black-White Interracial
Marriage.''} \emph{Demography} 43(4):673--89. doi:
\href{https://doi.org/10.1353/dem.2006.0033}{10.1353/dem.2006.0033}.

\leavevmode\vadjust pre{\hypertarget{ref-gullickson_racial_2019}{}}%
Gullickson, Aaron. 2019. {``The {Racial Identification} of {Young
Adults} in a {Racially Complex Society}.''} \emph{Emerging Adulthood}
7(2):150--61. doi:
\href{https://doi.org/10.1177/2167696818790306}{10.1177/2167696818790306}.

\leavevmode\vadjust pre{\hypertarget{ref-gullickson_choosing_2011}{}}%
Gullickson, Aaron, and Ann Morning. 2011. {``Choosing Race:
{Multiracial} Ancestry and Identification.''} \emph{Social Science
Research} 40(2):498--512. doi:
\href{https://doi.org/10.1016/j.ssresearch.2010.12.010}{10.1016/j.ssresearch.2010.12.010}.

\leavevmode\vadjust pre{\hypertarget{ref-herman_forced_2004}{}}%
Herman, Melissa. 2004. {``Forced to {Choose}: {Some Determinants} of
{Racial Identification} in {Multiracial Adolescents}.''} \emph{Child
Development} 75(3):730--48. doi:
\href{https://doi.org/10.2307/3696590}{10.2307/3696590}.

\leavevmode\vadjust pre{\hypertarget{ref-herman_blackwhiteother_2009}{}}%
Herman, Melissa R. 2009. {``The {Black-White-Other Achievement Gap}:
{Testing Theories} of {Academic Performance Among Multiracial} and
{Monoracial Adolescents}.''} \emph{Sociology of Education}
82(20):20--46.

\leavevmode\vadjust pre{\hypertarget{ref-hochschild_creating_2012}{}}%
Hochschild, Jennifer L., Vesla M. Weaver, and Traci R. Burch. 2012.
\emph{\href{https://books.google.com?id=BeTeF3iWG9sC}{Creating a {New
Racial Order}: {How Immigration}, {Multiracialism}, {Genomics}, and the
{Young Can Remake Race} in {America}}}. {Princeton University Press}.

\leavevmode\vadjust pre{\hypertarget{ref-hollinger_postethnic_1995}{}}%
Hollinger, David A. 1995. \emph{Postethnic {America}: {Beyond
Multiculturalism}}. {New York, NY}: {Basic Books}.

\leavevmode\vadjust pre{\hypertarget{ref-holloway_place_2009}{}}%
Holloway, Steven R., Richard Wright, Mark Ellis, and Margaret East.
2009. {``Place, Scale and the Racial Claims Made for Multiracial
Children in the 1990 {US Census}.''} \emph{Ethnic and Racial Studies}
32(3):522--47.

\leavevmode\vadjust pre{\hypertarget{ref-hou_interracial_2011}{}}%
Hou, Feng, and John Myles. 2011. {``Interracial Marriage and
Status-Caste Exchange in {Canada} and the {United States}.''}
\emph{Ethnic and Racial Studies}.

\leavevmode\vadjust pre{\hypertarget{ref-hughes_effect_2017}{}}%
Hughes, Jan N., Qian Cao, Stephen G. West, Paula Allee Smith, and
Carissa Cerda. 2017. {``Effect of Retention in Elementary Grades on
Dropping Out of School Early.''} \emph{Journal of School Psychology}
65:11--27. doi:
\href{https://doi.org/10.1016/j.jsp.2017.06.003}{10.1016/j.jsp.2017.06.003}.

\leavevmode\vadjust pre{\hypertarget{ref-hughes_significance_1990}{}}%
Hughes, Michael, and Bradley R. Hertel. 1990. {``The {Significance} of
{Color Remains}: {A Study} of {Life Chances}, {Mate Selection}, and
{Ethnic Consciousness Among} \{\vphantom\}{B}\vphantom\{\}lack
\{\vphantom\}{A}\vphantom\{\}mericans.''} \emph{Social Forces}
68(4):1105--20.

\leavevmode\vadjust pre{\hypertarget{ref-iverson_regimes_2022}{}}%
Iverson, Sarah, Ann Morning, Aliya Saperstein, and Janet Xu. 2022.
{``Regimes Beyond the {One-Drop Rule}: {New Models} of {Multiracial
Identity}.''} \emph{Genealogy} 6(2):57. doi:
\href{https://doi.org/10.3390/genealogy6020057}{10.3390/genealogy6020057}.

\leavevmode\vadjust pre{\hypertarget{ref-jimerson_metaanalysis_2001}{}}%
Jimerson, Shane R. 2001. {``Meta-Analysis of {Grade Retention Research}:
{Implications} for {Practice} in the 21st {Century}.''} \emph{School
Psychology Review} 30(3):420--37. doi:
\href{https://doi.org/10.1080/02796015.2001.12086124}{10.1080/02796015.2001.12086124}.

\leavevmode\vadjust pre{\hypertarget{ref-kalmijn_educational_2010}{}}%
Kalmijn, Matthijs. 2010. {``Educational Inequality, Homogamy, and Status
Exchange in Black-White Intermarriage: A Comment on {Rosenfeld}.''}
\emph{American Journal of Sociology} 115(4):1252--63.

\leavevmode\vadjust pre{\hypertarget{ref-kao_racial_1999}{}}%
Kao, Grace. 1999. {``Racial {Identity} and {Academic Performance}: {An
Examination} of {Biracial Asian} and {African American Youth}.''}
\emph{Journal of Asian American Studies} 2(3):223--49. doi:
\href{https://doi.org/10.1353/jaas.1999.0023}{10.1353/jaas.1999.0023}.

\leavevmode\vadjust pre{\hypertarget{ref-keith_skin_1991}{}}%
Keith, Verna M., and Cedric Herring. 1991. {``Skin {Tone} and
{Stratification} in the {Black Community}.''} \emph{American Journal of
Sociology} 97(3):760--78.

\leavevmode\vadjust pre{\hypertarget{ref-knol_mis_2011}{}}%
Knol, Mirjam J., Wiebe R. Pestman, and Diederick E. Grobbee. 2011.
{``The (Mis)use of Overlap of Confidence Intervals to Assess Effect
Modification.''} \emph{European Journal of Epidemiology} 26(4):253--54.
doi:
\href{https://doi.org/10.1007/s10654-011-9563-8}{10.1007/s10654-011-9563-8}.

\leavevmode\vadjust pre{\hypertarget{ref-liebler_boundaries_2016}{}}%
Liebler, Carolyn A. 2016. {``On the {Boundaries} of {Race}:
{Identification} of {Mixed-heritage Children} in the {United States},
1960 to 2010.''} \emph{Sociology of Race and Ethnicity} 2(4):548--68.
doi:
\href{https://doi.org/10.1177/2332649216632546}{10.1177/2332649216632546}.

\leavevmode\vadjust pre{\hypertarget{ref-liebler_america_2017}{}}%
Liebler, Carolyn A., Sonya R. Porter, Leticia E. Fernandez, James M.
Noon, and Sharon R. Ennis. 2017. {``America's {Churning Races}: {Race}
and {Ethnicity Response Changes Between Census} 2000 and the 2010
{Census}.''} \emph{Demography} 54(1):259--84. doi:
\href{https://doi.org/10.1007/s13524-016-0544-0}{10.1007/s13524-016-0544-0}.

\leavevmode\vadjust pre{\hypertarget{ref-maddox_cognitive_2002a}{}}%
Maddox, Keith B., and Stephanie A. Gray. 2002. {``Cognitive
{Representations} of {Black Americans}: {Reexploring} the {Role} of
{Skin Tone}.''} \emph{Personality and Social Psychology Bulletin}
28(2):250--59. doi:
\href{https://doi.org/10.1177/0146167202282010}{10.1177/0146167202282010}.

\leavevmode\vadjust pre{\hypertarget{ref-merton_intermarriage_1941}{}}%
Merton, Robert K. 1941. {``Intermarriage and the {Social Structure}:
{Fact} and {Theory}.''} \emph{Psychiatry} 4(3):361--74.

\leavevmode\vadjust pre{\hypertarget{ref-monk_skin_2014}{}}%
Monk, Ellis P. 2014.
{``\href{https://www.jstor.org/stable/43287571}{Skin {Tone
Stratification} Among {Black Americans}, 2001-2003}.''} \emph{Social
Forces} 92(4):1313--37.

\leavevmode\vadjust pre{\hypertarget{ref-morning_who_2000}{}}%
Morning, Ann. 2000. {``Who Is {Multiracial}? {Definitions} and
{Decisions}.''} \emph{Sociological Imagination} 37(4):209--29.

\leavevmode\vadjust pre{\hypertarget{ref-morning_generational_2018}{}}%
Morning, Ann, and Aliya Saperstein. 2018. {``The {Generational Locus} of
{Multiraciality} and {Its Implications} for {Racial
Self-Identification}.''} \emph{The ANNALS of the American Academy of
Political and Social Science} 677(1):57--68. doi:
\href{https://doi.org/10.1177/0002716218754774}{10.1177/0002716218754774}.

\leavevmode\vadjust pre{\hypertarget{ref-pager_sociology_2008}{}}%
Pager, Devah, and Hana Shepherd. 2008.
{``\href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2915460}{The
{Sociology} of {Discrimination}: {Racial Discrimination} in
{Employment}, {Housing}, {Credit}, and {Consumer Markets}}.''}
\emph{Annual Review of Sociology} 34:181--209.

\leavevmode\vadjust pre{\hypertarget{ref-park_human_1928}{}}%
Park, Robert Ezra. 1928. {``Human {Migration} and the {Marginal Man}.''}
\emph{American Journal of Sociology} 33(8):881--92.

\leavevmode\vadjust pre{\hypertarget{ref-qian_options_2004}{}}%
Qian, Zenchao. 2004. {``Options: {Racial}/{Ethnic Identification} of
{Children} of {Intermarried Couples}.''} \emph{Social Science Quarterly}
85(3):746--66.

\leavevmode\vadjust pre{\hypertarget{ref-quillian_new_2006}{}}%
Quillian, Lincoln. 2006.
{``\href{https://www.jstor.org/stable/29737741}{New {Approaches} to
{Understanding Racial Prejudice} and {Discrimination}}.''} \emph{Annual
Review of Sociology} 32:299--328.

\leavevmode\vadjust pre{\hypertarget{ref-reskin_race_2012}{}}%
Reskin, Barbara. 2012. {``The {Race Discrimination System}.''}
\emph{Annual Review of Sociology} 38(1):17--35. doi:
\href{https://doi.org/10.1146/annurev-soc-071811-145508}{10.1146/annurev-soc-071811-145508}.

\leavevmode\vadjust pre{\hypertarget{ref-root_racially_1992}{}}%
Root, Maria P. P., ed. 1992. \emph{Racially {Mixed People} in
{America}}. 1st edition. {Newbury Park, Calif}: {SAGE Publications,
Inc}.

\leavevmode\vadjust pre{\hypertarget{ref-rosenfeld_nontraditional_2010}{}}%
Rosenfeld, Michael J. 2010.
{``\href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3000058}{Nontraditional
{Families} and {Childhood Progress Through School}}.''}
\emph{Demography} 47(3):755--75.

\leavevmode\vadjust pre{\hypertarget{ref-roth_end_2005a}{}}%
Roth, Wendy D. 2005. {``\href{http://www.jstor.org/stable/4540881}{The
{End} of the {One-Drop Rule}? {Labeling} of {Multiracial Children} in
{Black Intermarriages}}.''} \emph{Sociological Forum} 20(1):35--67.

\leavevmode\vadjust pre{\hypertarget{ref-ruggles_ipums_2020}{}}%
Ruggles, Steven, Sarah Flood, Ronald Goeken, Josiah Grover, Erin Meyer,
Jose Pacas, and Matthew Sobek. 2020.
\emph{\href{https://doi.org/10.18128/D010.V10.0}{{IPUMS USA}: {Version}
10.0 {[}Dataset{]}}}. {Minneapolis, MN}: {IPUMS}.

\leavevmode\vadjust pre{\hypertarget{ref-saenz_persistence_1995}{}}%
Saenz, Rogelio, Sean-Shong Hwang, Benigno E. Aguirre, and Robert N.
Anderson. 1995. {``Persistence and {Change} in {Asian Identity Among
Children} of {Intermarried Couples}.''} \emph{Sociological Perspectives}
38(2):175--94.

\leavevmode\vadjust pre{\hypertarget{ref-saperstein_mulatto_2013}{}}%
Saperstein, Aliya, and Aaron Gullickson. 2013. {``A "{Mulatto Escape
Hatch}" in the {United States}? {Examining Evidence} of {Racial} and
{Social Mobility During} the {Jim Crow Era}.''} \emph{Demography}
50(5):1921--42. doi:
\href{https://doi.org/10.1007/s13524-013-0210-8}{10.1007/s13524-013-0210-8}.

\leavevmode\vadjust pre{\hypertarget{ref-saperstein_racial_2012}{}}%
Saperstein, Aliya, and Andrew M. Penner. 2012. {``Racial {Fluidity} and
{Inequality} in the {United States}.''} \emph{American Journal of
Sociology} 118(3):676--727.

\leavevmode\vadjust pre{\hypertarget{ref-shih_multiple_2019}{}}%
Shih, Margaret, Leigh S. Wilton, Serena Does, Brianna M. Goodale, and
Diana T. Sanchez. 2019. {``Multiple Racial Identities as Sources of
Psychological Resilience.''} \emph{Social and Personality Psychology
Compass} 13(6):e12469. doi:
\href{https://doi.org/10.1111/spc3.12469}{10.1111/spc3.12469}.

\leavevmode\vadjust pre{\hypertarget{ref-snipp_american_1989}{}}%
Snipp, C. Matthew. 1989.
\emph{\href{https://books.google.com?id=E0CsvVoVA90C}{American
{Indians}: {The First} of {This Land}}}. {Russell Sage Foundation}.

\leavevmode\vadjust pre{\hypertarget{ref-stearns_staying_2007}{}}%
Stearns, Elizabeth, Stephanie Moller, Judith Blau, and Stephanie
Potochnick. 2007. {``Staying {Back} and {Dropping Out}: {The
Relationship Between Grade Retention} and {School Dropout}.''}
\emph{Sociology of Education} 80(3):210--40. doi:
\href{https://doi.org/10.1177/003804070708000302}{10.1177/003804070708000302}.

\leavevmode\vadjust pre{\hypertarget{ref-stonequist_problem_1935}{}}%
Stonequist, Everett V. 1935. {``The {Problem} of the {Marginal Man}.''}
\emph{American Journal of Sociology} 41(1):1--12.

\leavevmode\vadjust pre{\hypertarget{ref-swidler_culture_1986}{}}%
Swidler, Ann. 1986. {``Culture in {Action}: {Symbols} and
{Strategies}.''} \emph{American Sociological Review} 51(2):273--86. doi:
\href{https://doi.org/10.2307/2095521}{10.2307/2095521}.

\leavevmode\vadjust pre{\hypertarget{ref-telles_race_2009}{}}%
Telles, Edward E., and Christina A. Sue. 2009. {``Race {Mixture}:
{Boundary Crossing} in {Comparative Perspective}.''} \emph{Annual Review
of Sociology} 35:129--46.

\leavevmode\vadjust pre{\hypertarget{ref-udry_health_2003}{}}%
Udry, J. Richard, Rose Maria Li, and Janet Hendrickson-Smith. 2003.
{``Health and {Behavior Risks} of {Adolescents} with {Mixed-Race
Identity}.''} \emph{American Journal of Public Health} 93(11):1865--70.

\leavevmode\vadjust pre{\hypertarget{ref-warren_patterns_2014}{}}%
Warren, John Robert, Emily Hoffman, and Megan Andrew. 2014. {``Patterns
and {Trends} in {Grade Retention Rates} in the {United States},
1995--2010.''} \emph{Educational Researcher} 43(9):433--43. doi:
\href{https://doi.org/10.3102/0013189X14563599}{10.3102/0013189X14563599}.

\leavevmode\vadjust pre{\hypertarget{ref-wolfe_land_2001}{}}%
Wolfe, Patrick. 2001. {``Land, {Labor}, and {Difference}: {Elementary
Structures} of {Race}.''} \emph{The American Historical Review}
106(3):866--905.

\leavevmode\vadjust pre{\hypertarget{ref-xie_racial_1997}{}}%
Xie, Yu, and Kimberly Goyette. 1997. {``The {Racial Identification} of
{Biracial Children} with {One Asian Parent}: {Evidence} from the 1990
{Census}.''} \emph{Social Forces} 76(2):547--70. doi:
\href{https://doi.org/10.2307/2580724}{10.2307/2580724}.

\end{CSLReferences}

\end{document}

### 2023-03-21 products/submissions/SocSci/randr/response_memo.md
A few additional edits to soc sci response memo

@@
+---
+output:
+  pdf_document: default
+  html_document: default
+---
 # Response Memo, Sociological Science
@@
 
-There are a couple of points to address here.
-First, I believe the concern that the reviewer is raising, although they do not name it directly, is the phenomenon of "academic redshirting" in which parents hold back kindergarten enrollment for a child whose birthday falls close to the cutoff date for enrollment.
+I believe the concern that the reviewer is raising, although they do not name it directly, is the phenomenon of "academic redshirting" in which parents hold back kindergarten enrollment for a child whose birthday falls close to the cutoff date for enrollment.
 It is true that I should have addressed this issue more directly in the paper and I have done so in the revised draft.

### 2023-03-21 products/paper/main.qmd
Prepare final manuscript for Soc Sci

@@
 title: Differences in the Risk of Grade Retention for Biracial and Monoracial Students in the US, 2010-2019
-#author: 
-#  - Aaron Gullickson, *University of Oregon*
+author: 
+  - name: Aaron Gullickson
+    orcid: 0000-0001-7237-8131
+    email: aarong@uoregon.edu
+    url: https://aarongullickson.netlify.app/
+    attributes:
+      corresponding: true
+      equal-contributor: true
+    affiliations:
+      - name: University of Oregon
+        department: Sociology
+        address: 727 PLC
+        city: Eugene
+        region: OR
+        country: USA
+        postal-code: 97405
 abstract: Understanding how outcomes for biracial individuals compare to their monoracial peers is critical for understanding how patterns of racial inequality in the contemporary United States might be shifting.
@@
 format:
+  submittable-pdf:
+    papersize: letter
+    fontsize: 12pt          # font size (typically 10pt, 11pt, or 12pt)
+    geometry: margin=1in    # controls the margin size
+    fig-height: 4
+    fig-pos: t
+    linestretch: 1
+    keep-tex: true
+    titlepage: false
+    cite-method: biblatex
+    #biblatexoptions: 
+    #natbiboptions: authoryear,round
+    biblio-style: apa
   pdf:
@@
     fig-pos: t
-    linestretch: 1.5
+    linestretch: 1
     keep-tex: false

### 2023-03-21 products/submissions/SocSci/final_materials/main.tex
Prepare final manuscript for Soc Sci

--- 2023-03-14 products/submissions/SocSci/randr/main.tex
@@
 \documentclass[
+  12pt,
   letterpaper,
-  DIV=11,
-  numbers=noendperiod]{scrartcl}
+]{article}
 
@@
 \usepackage{xcolor}
+\usepackage[margin=1in]{geometry}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 \makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
+
+% -----------------------
+% CUSTOM PREAMBLE STUFF
+% -----------------------
+
+\usepackage{orcidlink}
+\definecolor{mypink}{RGB}{219, 48, 122}
+\usepackage{endnotes}
+\usepackage[nolists]{endfloat}
+
+% ---------------------------
+% END CUSTOM PREAMBLE STUFF
+% ---------------------------
 \usepackage{booktabs}
@@
 \usepackage{xcolor}
-\KOMAoption{captions}{tableheading}
 \makeatletter
@@
 \fi
+\usepackage[style=apa,]{biblatex}
+\addbibresource{project.bib}
 \IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
@@
   pdftitle={Differences in the Risk of Grade Retention for Biracial and Monoracial Students in the US, 2010-2019},
+  pdfauthor={Aaron Gullickson},
   colorlinks=true,
@@
 \url{https://osf.io/4fevt/?view_only=4abc6d86595c4313a8d4792471e9bc0d}.}}
-\author{}
+\author{
+Aaron Gullickson~\orcidlink{0000-0001-7237-8131}\\University of
+Oregon, Sociology\\\href{mailto:aarong@uoregon.edu}{aarong@uoregon.edu}}
 \date{}
-
 \begin{document}
+
 \maketitle
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable, boxrule=0pt, frame hidden, sharp corners]}{\end{tcolorbox}}\fi
-
-\setstretch{1.5}
+
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
+
+\setstretch{1}
 \hypertarget{introduction}{%
@@
 1960s has resulted in a ``biracial baby boom'' of individuals whose
-parents cross racial lines (Root 1992) and this demographic surge may
-have transformative effects on racial boundaries (Alba 2020; Hochschild,
-Weaver, and Burch 2012). While racial mixing is nothing new to US
-history (Gullickson and Morning 2011; Morning 2000), this biracial boom
+parents cross racial lines \autocite{root_racially_1992} and this
+demographic surge may have transformative effects on racial boundaries
+\autocite{hochschild_creating_2012,alba_great_2020}. While racial mixing
+is nothing new to US history
+\autocite{morning_who_2000,gullickson_choosing_2011}, this biracial boom
 is notable for its size and has occurred in an environment in which
@@
 extensive scholarship documents the identification and classification
-decisions made by or on behalf of multiracial individuals (Bratter and
-Heard 2009; Brunsma 2005; Davenport 2016; Herman 2004; Holloway et al.
-2009; Liebler 2016; Qian 2004; Roth 2005; Xie and Goyette 1997), only a
-smattering of research has documented how the life chances of biracial
-individuals compare to their monoracial peers (Campbell 2009; Herman
-2009; Kao 1999). In this article, I expand on this work by examining the
-risk of grade retention for K-12 students whose parents identify with
-different races.
+decisions made by or on behalf of multiracial individuals
+\autocite{xie_racial_1997,qian_options_2004,roth_end_2005a,brunsma_interracial_2005,herman_forced_2004,holloway_place_2009,bratter_mother_2009,davenport_role_2016,liebler_boundaries_2016},
+only a smattering of research has documented how the life chances of
+biracial individuals compare to their monoracial peers
+\autocite{kao_racial_1999,campbell_multiracial_2009,herman_blackwhiteother_2009}.
+In this article, I expand on this work by examining the risk of grade
+retention for K-12 students whose parents identify with different races.
 
@@
 that individuals whose parents belong to different races do not
-consistently identify with both races (Doyle and Kao 2007; Gullickson
-2019; Liebler et al. 2017). Furthermore, social status and social
-mobility can predict racial identification generally (Saperstein and
-Gullickson 2013; Saperstein and Penner 2012) and social status is
-related to the identification choices of biracial individuals
-specifically (Davenport 2016; Roth 2005). Thus, differences in outcomes
-among biracial individuals may in part produce differences in
-identification, rather than the other way around.
+consistently identify with both races
+\autocite{doyle_are_2007,liebler_america_2017,gullickson_racial_2019}.
+Furthermore, social status and social mobility can predict racial
+identification generally
+\autocite{saperstein_racial_2012,saperstein_mulatto_2013} and social
+status is related to the identification choices of biracial individuals
+specifically \autocite{roth_end_2005a,davenport_role_2016}. Thus,
+differences in outcomes among biracial individuals may in part produce
+differences in identification, rather than the other way around.
 
@@
 have previously utilized this feature to examine the classification
-choices made on behalf of these children (Roth 2005; Xie and Goyette
-1997). Analyzing outcomes for such children is more difficult, for the
-simple reason that they do not have many measurable outcomes yet.
-However, grade retention is measurable and important even for young
-children. Being held back in school is a relatively common occurrence
-that has consequences for later educational transitions (Jimerson 2001)
-and is disproportionately applied by race (Warren, Hoffman, and Andrew
-2014). Grade retention is one of the earliest stratifiers of life
-chances that an individual will experience.
+choices made on behalf of these children
+\autocite{xie_racial_1997,roth_end_2005a}. Analyzing outcomes for such
+children is more difficult, for the simple reason that they do not have
+many measurable outcomes yet. However, grade retention is measurable and
+important even for young children. Being held back in school is a
+relatively common occurrence that has consequences for later educational
+transitions \autocite{jimerson_metaanalysis_2001} and is
+disproportionately applied by race \autocite{warren_patterns_2014}.
+Grade retention is one of the earliest stratifiers of life chances that
+an individual will experience.
 
@@
 Discrimination and bias by race play an important role in overall levels
-of racial inequality (Pager and Shepherd 2008; Quillian 2006).
+of racial inequality \autocite{quillian_new_2006,pager_sociology_2008}.
 Discrimination need not be consciously recognized by actors, but may
 instead reflect cognitive biases and schema that individuals unknowingly
-rely upon and reinforce in social interactions (Quillian 2006).
-Discrimination can also be systemic in nature when biases are built into
-and reinforced by organizations and institutional systems (Reskin 2012).
-In our current ``colorblind'' regime (Bonilla-Silva 2006), such systems
+rely upon and reinforce in social interactions
+\autocite{quillian_new_2006}. Discrimination can also be systemic in
+nature when biases are built into and reinforced by organizations and
+institutional systems \autocite{reskin_race_2012}. In our current
+``colorblind'' regime \autocite{bonilla-silva_racism_2006}, such systems
 may be formally race neutral but produce disparate impacts by race
-(Pager and Shepherd 2008).
+\autocite{pager_sociology_2008}.
 
@@
 ancestry are considered exclusively Black, regardless of other
-ancestries (Davis 1991). Davis (1991) generalizes this practice into the
-concept of \emph{hypodescent} in which individuals with mixed racial
-ancestry will be classified as members of the lower status group. A
-parallel pattern of \emph{hyperdescent} implies that individuals of
-mixed racial ancestry will be classified as members of the higher status
-group. This pattern of hyperdescent has historically governed the
-experiences of individuals with American Indian ancestry (Snipp 1989;
-Wolfe 2001).
+ancestries \autocite{davis_who_1991}. \textcite{davis_who_1991}
+generalizes this practice into the concept of \emph{hypodescent} in
+which individuals with mixed racial ancestry will be classified as
+members of the lower status group. A parallel pattern of
+\emph{hyperdescent} implies that individuals of mixed racial ancestry
+will be classified as members of the higher status group. This pattern
+of hyperdescent has historically governed the experiences of individuals
+with American Indian ancestry
+\autocite{snipp_american_1989,wolfe_land_2001}.
 
@@
 observed in two specific historical cases involving Black ancestry and
-American Indian ancestry, respectively. As Iverson et al. (2022) note,
-these two cases could just as parsimoniously be explained by a
-``dominance'' model which ranks ancestries by their tendencies to be
-supercessive or recessive in determining identification and
-classification.
+American Indian ancestry, respectively. As
+\textcite{iverson_regimes_2022} note, these two cases could just as
+parsimoniously be explained by a ``dominance'' model which ranks
+ancestries by their tendencies to be supercessive or recessive in
+determining identification and classification.
 
 The development of particular norms may also reflect the historical
-regime in which they were developed (Gullickson and Morning 2011;
-Iverson et al. 2022). In this case, neither the hypodescent/hyperdescent
-or dominance paradigm may help us to fully understand the experiences of
-those of Latino and Asian ancestry due to more recent processes of
-racialization. Furthermore, the historical norms for all groups may be
-transforming in the post-Civil Rights era.
+regime in which they were developed
+\autocite{gullickson_choosing_2011,iverson_regimes_2022}. In this case,
+neither the hypodescent/hyperdescent or dominance paradigm may help us
+to fully understand the experiences of those of Latino and Asian
+ancestry due to more recent processes of racialization. Furthermore, the
+historical norms for all groups may be transforming in the post-Civil
+Rights era.
 
@@
 substantially in skin tone and those individuals with a lighter skin
-tone have better outcomes (Hughes and Hertel 1990; Keith and Herring
-1991; Monk 2014) and are viewed more favorably by both Blacks and Whites
-(Maddox and Gray 2002). Thus, light skin moderates the stigma of
-Blackness even among monoracial individuals.
+tone have better outcomes
+\autocite{keith_skin_1991,hughes_significance_1990,monk_skin_2014} and
+are viewed more favorably by both Blacks and Whites
+\autocite{maddox_cognitive_2002a}. Thus, light skin moderates the stigma
+of Blackness even among monoracial individuals.
 
@@
 ``middleman'' minority position. This kind of middle tier status for
-individuals of mixed race is common in Latin America (Telles and Sue
-2009) and Bonilla-Silva (2004) has argued that such a system may be
-emerging in the contemporary US as well. In this case, many biracial
-groups may occupy a nebulous and ill-defined middleman or buffer class
-position in-between whites and darker-skinned non-whites, helping to
-cement a new kind of racial hierarchy.
+individuals of mixed race is common in Latin America
+\autocite{telles_race_2009} and \textcite{bonilla-silva_biracial_2004}
+has argued that such a system may be emerging in the contemporary US as
+well. In this case, many biracial groups may occupy a nebulous and
+ill-defined middleman or buffer class position in-between whites and
+darker-skinned non-whites, helping to cement a new kind of racial
+hierarchy.
 
@@
 distributions of material, cultural, and social resources across racial
-groups (Conley 1999). On average, socioeconomic background
-(e.g.~parental income, wealth, and education) differs by race and such
-socioeconomic background is consequential for outcomes. Other less
-materially defined resources also vary across groups such as cultural
-styles, social networks, and neighborhood quality. These differences in
-background resources reflect the accumulation of historical processes of
-discrimination and, when combined with contemporary processes of
-discrimination, magnify the overall level of racial inequality we
-observe.
+groups \autocite{conley_being_1999}. On average, socioeconomic
+background (e.g.~parental income, wealth, and education) differs by race
+and such socioeconomic background is consequential for outcomes. Other
+less materially defined resources also vary across groups such as
+cultural styles, social networks, and neighborhood quality. These
+differences in background resources reflect the accumulation of
+historical processes of discrimination and, when combined with
+contemporary processes of discrimination, magnify the overall level of
+racial inequality we observe.
 
@@
 parental resources roughly halfway between their two constituent groups,
-on average (Chew, Eggebeen, and Uhlenberg 1989). Prior research looking
-at biracial children in the Census data has found some evidence to
-support this argument (Chew et al. 1989). In such a case, the resources
-of biracial individuals would predict outcomes in-between the two
-constituent groups, barring other mechanisms.
+on average \autocite{chew_american_1989}. Prior research looking at
+biracial children in the Census data has found some evidence to support
+this argument \autocite{chew_american_1989}. In such a case, the
+resources of biracial individuals would predict outcomes in-between the
+two constituent groups, barring other mechanisms.
 
@@
 group, while individuals from a higher status racial group will be
-similarly negatively selected into the same marriage (Davis 1941; Fu
-2001; Merton 1941). Prior research has shown substantial evidence of
-this pattern in the case of Black male/White female marriages in the US,
-but weaker evidence for other types of unions (Fu 2001; Gullickson 2006;
-Hou and Myles 2011; Kalmijn 2010). Even in cases where status exchange
-is present, its net effect on resources is complex because status
-exchange positively selects one partner and negatively selects the other
-partner. Therefore the two effects may somewhat cancel each other out in
-terms of the overall distribution of resources across parents.
+similarly negatively selected into the same marriage
+\autocite{davis_intermarriage_1941,merton_intermarriage_1941,fu_racial_2001}.
+Prior research has shown substantial evidence of this pattern in the
+case of Black male/White female marriages in the US, but weaker evidence
+for other types of unions
+\autocite{gullickson_education_2006,fu_racial_2001,kalmijn_educational_2010,hou_interracial_2011}.
+Even in cases where status exchange is present, its net effect on
+resources is complex because status exchange positively selects one
+partner and negatively selects the other partner. Therefore the two
+effects may somewhat cancel each other out in terms of the overall
+distribution of resources across parents.
 
@@
 dissonance of an identity that crosses strong and salient racial
-boundaries (Park 1928; Stonequist 1935). This liminal identity conflict
-will result in negative mental health experiences that could feed into
-other negative outcomes as well. Evidence for such negative mental
-health experiences, however, has been mixed (Bratter and Gorman 2011;
-Campbell and Eggerling-Boeck 2006; Cheng and Lively 2009; Udry, Li, and
-Hendrickson-Smith 2003).
+boundaries \autocite{park_human_1928,stonequist_problem_1935}. This
+liminal identity conflict will result in negative mental health
+experiences that could feed into other negative outcomes as well.
+Evidence for such negative mental health experiences, however, has been
+mixed
+\autocite{udry_health_2003,campbell_what_2006,cheng_multiracial_2009,bratter_does_2011}.
 
@@
 flexible, fragmented, and fluid use of cultural bits that comprise a
-``toolkit'' (DiMaggio 1997; Swidler 1986). From this perspective,
-biracial individuals may actually benefit from an enlarged cultural
-toolkit that allows them to navigate a variety of diverse social
-contexts (Shih et al. 2019).
+``toolkit'' \autocite{swidler_culture_1986,dimaggio_culture_1997}. From
+this perspective, biracial individuals may actually benefit from an
+enlarged cultural toolkit that allows them to navigate a variety of
+diverse social contexts \autocite{shih_multiple_2019}.
 
@@
 that far fewer people identify as biracial than could feasibly do so
-based on their parents' races (Morning and Saperstein 2018) and that
-various selection effects operate on the decision to identify as
-biracial among those who could (Davenport 2016; Roth 2005). Furthermore,
-research on changes in racial reporting over time suggests that other
-status markers like education and income may influence how people
-identify themselves (Saperstein and Gullickson 2013; Saperstein and
-Penner 2012). For example, a person with one Black and one White parent
-who is routinely treated by others as Black and discriminated against as
-Black may be more likely to identify themselves as exclusively Black
-than a similar person of Black and White ancestry who experiences less
-consistent discrimination. Thus, any analysis of outcomes for biracial
-individuals based on self-identification must deal with the potential
-for reverse causation in the observed differences between biracial
-individuals and their constituent groups (Bratter 2018).
+based on their parents' races \autocite{morning_generational_2018} and
+that various selection effects operate on the decision to identify as
+biracial among those who could
+\autocite{roth_end_2005a,davenport_role_2016}. Furthermore, research on
+changes in racial reporting over time suggests that other status markers
+like education and income may influence how people identify themselves
+\autocite{saperstein_racial_2012,saperstein_mulatto_2013}. For example,
+a person with one Black and one White parent who is routinely treated by
+others as Black and discriminated against as Black may be more likely to
+identify themselves as exclusively Black than a similar person of Black
+and White ancestry who experiences less consistent discrimination. Thus,
+any analysis of outcomes for biracial individuals based on
+self-identification must deal with the potential for reverse causation
+in the observed differences between biracial individuals and their
+constituent groups \autocite{bratter_multiracial_2018}.
 
@@
 survey data in which the racial identification of a respondent's
-biological parents are provided. Campbell (2009) used data from the
-National Longitudinal Survey of Adolescent to Adult Health (Add Health)
-to examine high school GPA, advanced placement in math, and four-year
-college enrollment among biracial and monoracial individuals identified
-by co-resident biological parents' races. Herman (2009) examined
-differences in GPA using data on high school students in California and
-Wisconsin in which the students reported their parents' races. Both of
-these studies found some support for an ``in-between'' status for many
-biracial groups, but are hampered by a lack of statistical power due to
-small sample sizes. Kao (1999) used data from the National Educational
+biological parents are provided. \textcite{campbell_multiracial_2009}
+used data from the National Longitudinal Survey of Adolescent to Adult
+Health (Add Health) to examine high school GPA, advanced placement in
+math, and four-year college enrollment among biracial and monoracial
+individuals identified by co-resident biological parents' races.
+\textcite{herman_blackwhiteother_2009} examined differences in GPA using
+data on high school students in California and Wisconsin in which the
+students reported their parents' races. Both of these studies found some
+support for an ``in-between'' status for many biracial groups, but are
+hampered by a lack of statistical power due to small sample sizes.
+\textcite{kao_racial_1999} used data from the National Educational
 Longitudinal Study of 1988 to examine mathematics achievement scores and
@@
 children. This approach has been used previously to examine the
-likelihood of living in poverty (Bratter and Damaske 2013; Bratter and
-Kimbro 2013) and residential segregation (Ellis et al. 2012). These
-studies provide valuable information on biracial childrens' lived
-experience, but do not directly measure their own outcomes. However, one
-important educational outcome is partially identifiable for children in
-Census and ACS data - grade retention.
+likelihood of living in poverty
+\autocite{bratter_multiracial_2013,bratter_poverty_2013} and residential
+segregation \autocite{ellis_agents_2012}. These studies provide valuable
+information on biracial childrens' lived experience, but do not directly
+measure their own outcomes. However, one important educational outcome
+is partially identifiable for children in Census and ACS data - grade
+retention.
 
@@
 later negative educational outcomes, such as a higher dropout risk
-(Andrew 2014; Hughes et al. 2017; Jimerson 2001; Stearns et al. 2007).
+\autocite{jimerson_metaanalysis_2001,stearns_staying_2007,andrew_scarring_2014,hughes_effect_2017}.
 The risk of grade retention also varies substantially by race and class
 background, with Black and Latino students at particularly high risk of
-grade retention relative to other groups (Warren et al. 2014).
+grade retention relative to other groups
+\autocite{warren_patterns_2014}.
 
@@
 were extracted from the Minnesota Population Center's Integrated Public
-Use Microdata Series (Ruggles et al. 2020).
+Use Microdata Series \autocite{ruggles_ipums_2020}.
 
@@
 analysis to children who are more likely to be the biological children
-of both parents (Saenz et al. 1995; Xie and Goyette 1997). Consistent
-with this prior work, I restrict cases in two ways. First, I restrict
-the sample to cases where both parents were of a reasonable age at the
-birth of the child (aged 15-44 for mothers and aged 15-60 for fathers).
-Second, I restrict the sample to those cases where the reported race and
-Hispanicity of the child is inclusive of at least one of the parents'
-races.
+of both parents \autocite{saenz_persistence_1995,xie_racial_1997}.
+Consistent with this prior work, I restrict cases in two ways. First, I
+restrict the sample to cases where both parents were of a reasonable age
+at the birth of the child (aged 15-44 for mothers and aged 15-60 for
+fathers). Second, I restrict the sample to those cases where the
+reported race and Hispanicity of the child is inclusive of at least one
+of the parents' races.
 
@@
 greater educational success than other children as a result of material
-and social resources (Biblarz and Raftery 1999). Therefore, this sample
-restriction will bias overall estimates of grade retention downward.
-However, this bias is less problematic for the goal of this study, which
-is to estimate differences between groups. The sample restriction
-effectively eliminates differences in family structure across racial
-groups that may account for some observed differences in grade
-retention. However, the remaining differences observed here are likely
-to hold for the full population in the absence of strong racial
+and social resources \autocite{biblarz_family_1999}. Therefore, this
+sample restriction will bias overall estimates of grade retention
+downward. However, this bias is less problematic for the goal of this
+study, which is to estimate differences between groups. The sample
+restriction effectively eliminates differences in family structure
+across racial groups that may account for some observed differences in
+grade retention. However, the remaining differences observed here are
+likely to hold for the full population in the absence of strong racial
 differences in the effect of family structure on grade retention.
@@
 commonly used in popular practice and government tabulation to identify
-race in the US (Hollinger 1995). Parents are identified as Latino based
-on their response to the Hispanicity question, regardless of their
-response to the race question. Parents are identified as Indigenous if
-they responded as either American Indian/Alaska Native or as Pacific
-Islander. Because the goal of my analysis is to examine specifically the
-outcomes of ``first-generation'' biracial children, I exclude cases
-where at least one parent identified with multiple races. The
-cross-tabulation of parents' races leads to ten distinct biracial
-categories. The total sample size for each of these biracial categories
-is shown in Table~\ref{tbl-descriptive}, along with the sample size of
-the monoracial comparison groups.
+race in the US \autocite{hollinger_postethnic_1995}. Parents are
+identified as Latino based on their response to the Hispanicity
+question, regardless of their response to the race question. Parents are
+identified as Indigenous if they responded as either American
+Indian/Alaska Native or as Pacific Islander. Because the goal of my
+analysis is to examine specifically the outcomes of ``first-generation''
+biracial children, I exclude cases where at least one parent identified
+with multiple races. The cross-tabulation of parents' races leads to ten
+distinct biracial categories. The total sample size for each of these
+biracial categories is shown in Table~\ref{tbl-descriptive}, along with
+the sample size of the monoracial comparison groups.
 
@@
 reported age and grade. Prior studies have used the concept of a student
-being behind modal grade (Bianchi 1984; Frederick and Hauser 2008). The
-most notable limitation of this approach is the ambiguity created by the
-fact that at any given age, a student may reasonably be in two modal
-grades. Without detailed information on birthdates and survey timing,
-the correct modal grade for most students cannot be identified. This
-issue has been somewhat alleviated in prior work by the use of the
-October supplement to the Current Population Survey which is close to
-the beginning of the school year.
+being behind modal grade
+\autocite{bianchi_children_1984,frederick_have_2008}. The most notable
+limitation of this approach is the ambiguity created by the fact that at
+any given age, a student may reasonably be in two modal grades. Without
+detailed information on birthdates and survey timing, the correct modal
+grade for most students cannot be identified. This issue has been
+somewhat alleviated in prior work by the use of the October supplement
+to the Current Population Survey which is close to the beginning of the
+school year.
 
@@
 more or less random with regard to sociodemographic characteristics.
-This approach is equivalent to the method used by Rosenfeld (2010) on
-2000 Census data, but more detailed information on current grade in
-recent ACS data allows for more precise estimation.
-Table~\ref{tbl-descriptive} shows the percent of each monoracial and
-biracial group that is CBEG.
+This approach is equivalent to the method used by
+\textcite{rosenfeld_nontraditional_2010} on 2000 Census data, but more
+detailed information on current grade in recent ACS data allows for more
+precise estimation. Table~\ref{tbl-descriptive} shows the percent of
+each monoracial and biracial group that is CBEG.
 
@@
 capture cases of ``academic redshirting'' in which parents intentionally
-delay their child's kindergarten enrollment by a year. Frederick and
-Hauser (2008) suggest that academic redshirting is in many cases be a
-form of preemptive retention for children with developmental delays and
-thus the error induced by these false positive cases may be minimal.
-However, Bassok and Reardon (2013) have shown that academic redshirting
-is more common among white and SES privileged parents, suggesting that
-the demographic covariates of academic redshirting may operate in the
-opposite direction of grade retention. Regardless, academic redshirting
-should be of minimal concern for the measure of CBEG used here. Academic
-redshirting most frequently occurs for students whose birthdays fall
-close to the cutoff period for enrollment (Bassok and Reardon 2013;
-Graue and DiPerna 2000). Therefore, academic redshirts spend most of a
-given school year at the older, but correct, modal age for their grade
-and most of these students will not be identified as CBEG.\footnote{Analysis
-  of the probability of being CBEG by race and grade, shown in the
-  supplementary materials, indicates that Black and Latino children in
-  kindergarten are less likely to be CBEG than White children, which may
-  be a consequence of redshirting. However, these differences quickly
-  reverse direction by first grade. As a sensitivity analysis, I
-  repeated the main analysis shown here separately for students in
-  elementary (1st-5th), middle school (6th-8th), and high school
-  (9th-12th) grades. These results are presented in the supplementary
-  materials. Although more statistically noisy due to smaller sample
-  sizes, those results are consistent with the conclusions drawn here.}
+delay their child's kindergarten enrollment by a year.
+\textcite{frederick_have_2008} suggest that academic redshirting is in
+many cases be a form of preemptive retention for children with
+developmental delays and thus the error induced by these false positive
+cases may be minimal. However, \textcite{bassok_academic_2013} have
+shown that academic redshirting is more common among white and SES
+privileged parents, suggesting that the demographic covariates of
+academic redshirting may operate in the opposite direction of grade
+retention. Regardless, academic redshirting should be of minimal concern
+for the measure of CBEG used here. Academic redshirting most frequently
+occurs for students whose birthdays fall close to the cutoff period for
+enrollment \autocite{graue_redshirting_2000,bassok_academic_2013}.
+Therefore, academic redshirts spend most of a given school year at the
+older, but correct, modal age for their grade and most of these students
+will not be identified as CBEG.\footnote{Analysis of the probability of
+  being CBEG by race and grade, shown in the supplementary materials,
+  indicates that Black and Latino children in kindergarten are less
+  likely to be CBEG than White children, which may be a consequence of
+  redshirting. However, these differences quickly reverse direction by
+  first grade. As a sensitivity analysis, I repeated the main analysis
+  shown here separately for students in elementary (1st-5th), middle
+  school (6th-8th), and high school (9th-12th) grades. These results are
+  presented in the supplementary materials. Although more statistically
+  noisy due to smaller sample sizes, those results are consistent with
+  the conclusions drawn here.}
 
@@
 test of difference at \(p<0.05\) and when used in this way, such
-assessments lead to far more stringent tests (Knol, Pestman, and Grobbee
-2011). When standard errors between groups are equal, overlap in 83.4\%
+assessments lead to far more stringent tests \autocite{knol_mis_2011}.
+When standard errors between groups are equal, overlap in 83.4\%
 confidence intervals will achieve this goal. However, in cases where
@@
   \[1.96 * \frac{\sqrt{1+\rho^2}}{1+\rho}\] where \(\rho\) is the ratio
-  of the standard errors for the two statistics. See Knol et al. (2011)
-  for a detailed derivation.}
+  of the standard errors for the two statistics. See
+  \textcite{knol_mis_2011} for a detailed derivation.}
 
@@
 evidence of an ``in-between'' status in other educational outcomes
-(Campbell 2009; Herman 2009). I raise two possible explanations for this
-discrepancy that point to strengths and weaknesses of the current
-research and may help drive future work on the topic. First, this
-discrepancy may be due to the small sample sizes of multiracial
-respondents in prior work which led to statistically imprecise
-estimates. Because the current findings rely upon large samples, they
-present an important step forward in our understanding of how biracial
-individuals will fit into America's system of racial inequality.
-Nonetheless, sample sizes for some non-white biracial respondents remain
-small, limiting our understanding of the outcomes for those groups. In
-particular, results for non-white part-Indigenous populations are quite
-imprecise. The results counter-intuitively suggest that part-Indigenous
-populations are the least likely to be in the position expected by
-hyperdescent, but due to low sample sizes, that finding is highly
-tentative.
+\autocite{herman_blackwhiteother_2009,campbell_multiracial_2009}. I
+raise two possible explanations for this discrepancy that point to
+strengths and weaknesses of the current research and may help drive
+future work on the topic. First, this discrepancy may be due to the
+small sample sizes of multiracial respondents in prior work which led to
+statistically imprecise estimates. Because the current findings rely
+upon large samples, they present an important step forward in our
+understanding of how biracial individuals will fit into America's system
+of racial inequality. Nonetheless, sample sizes for some non-white
+biracial respondents remain small, limiting our understanding of the
+outcomes for those groups. In particular, results for non-white
+part-Indigenous populations are quite imprecise. The results
+counter-intuitively suggest that part-Indigenous populations are the
+least likely to be in the position expected by hyperdescent, but due to
+low sample sizes, that finding is highly tentative.
 
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-alba_great_2020}{}}%
-Alba, Richard. 2020.
-\emph{\href{https://press.princeton.edu/books/hardcover/9780691201634/the-great-demographic-illusion}{The
-{Great Demographic Illusion}}}. {Princeton University Press}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-andrew_scarring_2014}{}}%
-Andrew, Megan. 2014. {``The {Scarring Effects} of {Primary-Grade
-Retention}? {A Study} of {Cumulative Advantage} in the {Educational
-Career}.''} \emph{Social Forces} 93(2):653--85. doi:
-\href{https://doi.org/10.1093/sf/sou074}{10.1093/sf/sou074}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bassok_academic_2013}{}}%
-Bassok, Daphna, and Sean F. Reardon. 2013. {``{`{Academic Redshirting}'}
-in {Kindergarten}: {Prevalence}, {Patterns}, and {Implications}.''}
-\emph{Educational Evaluation and Policy Analysis} 35(3):283--97. doi:
-\href{https://doi.org/10.3102/0162373713482764}{10.3102/0162373713482764}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bianchi_children_1984}{}}%
-Bianchi, Suzanne M. 1984. {``Children's {Progress Through School}: {A
-Research Note}.''} \emph{Sociology of Education} 57(3):184--92. doi:
-\href{https://doi.org/10.2307/2112602}{10.2307/2112602}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-biblarz_family_1999}{}}%
-Biblarz, Timothy J., and Adrian E. Raftery. 1999. {``Family {Structure},
-{Educational Attainment}, and {Socioeconomic Success}: {Rethinking} the
-{`{Pathology} of {Matriarchy}'}.''} \emph{American Journal of Sociology}
-105(2):321--65.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bonilla-silva_biracial_2004}{}}%
-Bonilla-Silva, Eduardo. 2004. {``From Bi-Racial to Tri-Racial: {Towards}
-a New System of Racial Stratification in the {USA}.''} \emph{Ethnic and
-Racial Studies} 27(6):931--50.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bonilla-silva_racism_2006}{}}%
-Bonilla-Silva, Eduardo. 2006.
-\emph{\href{https://books.google.com?id=HKJE4rVZG1EC}{Racism Without
-{Racists}: {Color-Blind Racism} and the {Persistence} of {Racial
-Inequality} in the {United States}}}. {Rowman \& Littlefield
-Publishers}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bratter_multiracial_2018}{}}%
-Bratter, Jenifer L. 2018. {``Multiracial {Identification} and {Racial
-Gaps}: {A Work} in {Progress}.''} \emph{The ANNALS of the American
-Academy of Political and Social Science} 677(1):69--80. doi:
-\href{https://doi.org/10.1177/0002716218758622}{10.1177/0002716218758622}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bratter_poverty_2013}{}}%
-Bratter, Jenifer L., and Sarah Damaske. 2013. {``Poverty at a {Racial
-Crossroads}: {Poverty Among Multiracial Children} of {Single
-Mothers}.''} \emph{Journal of Marriage and Family} 75(2):486--502. doi:
-\href{https://doi.org/10.1111/jomf.12012}{10.1111/jomf.12012}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bratter_does_2011}{}}%
-Bratter, Jenifer L., and Bridget K. Gorman. 2011. {``Does {Multiracial
-Matter}? {A Study} of {Racial Disparities} in {Self-Rated Health}.''}
-\emph{Demography} 48(1):127--52. doi:
-\href{https://doi.org/10.1007/s13524-010-0005-0}{10.1007/s13524-010-0005-0}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bratter_mother_2009}{}}%
-Bratter, Jenifer, and Holly E. Heard. 2009.
-{``\href{http://www.jstor.org/stable/40542696}{Mother's, {Father}'s, or
-{Both}? {Parental Gender} and {Parent-Child Interactions} in the {Racial
-Classification} of {Adolescents}}.''} \emph{Sociological Forum}
-24(3):658--88.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bratter_multiracial_2013}{}}%
-Bratter, Jenifer, and Rachel Tolbert Kimbro. 2013. {``Multiracial
-{Children} and {Poverty}: {Evidence From} the {Early Childhood
-Longitudinal Study} of~{Kindergartners}.''} \emph{Family Relations}
-62(1):175--89. doi:
-\href{https://doi.org/10.1111/j.1741-3729.2012.00750.x}{10.1111/j.1741-3729.2012.00750.x}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-brunsma_interracial_2005}{}}%
-Brunsma, David L. 2005. {``Interracial {Families} and the {Racial
-Identification} of {Mixed-Race Children}: {Evidence} from the {Early
-Childhood Longitudinal Study}.''} \emph{Social Forces} 84(2):1131--57.
-doi: \href{https://doi.org/10.2307/3598493}{10.2307/3598493}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-campbell_multiracial_2009}{}}%
-Campbell, Mary E. 2009. {``Multiracial {Groups} and {Educational
-Inequality}: {A Rainbow} or a {Divide}?''} \emph{Social Problems}
-56(3):425--46.
-
-\leavevmode\vadjust pre{\hypertarget{ref-campbell_what_2006}{}}%
-Campbell, Mary E., and Jennifer Eggerling-Boeck. 2006. {``"{What About}
-the {Children}?" {The Psychological} and {Social Well-Being} of
-{Multiracial Adolescents}.''} \emph{The Sociological Quarterly}
-47(1):147--73.
-
-\leavevmode\vadjust pre{\hypertarget{ref-cheng_multiracial_2009}{}}%
-Cheng, Simon, and Kathryn J. Lively. 2009. {``Multiracial
-{Self-Identification} and {Adolescent Outcomes}: {A Social Psychological
-Approach} to the {Marginal Man Theory}.''} \emph{Social Forces}
-88(1):61--98.
-
-\leavevmode\vadjust pre{\hypertarget{ref-chew_american_1989}{}}%
-Chew, Kenneth S. Y., David J. Eggebeen, and Peter R. Uhlenberg. 1989.
-{``American {Children} in {Multiracial Households}.''}
-\emph{Sociological Perspectives} 32(1):65--85. doi:
-\href{https://doi.org/10.2307/1389008}{10.2307/1389008}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-conley_being_1999}{}}%
-Conley, Dalton. 1999. \emph{Being {Black}. {Living} in the {Red}:
-{Race}, {Wealth}, and {Social Policy} in {America}}. {Berkeley}:
-{University of California Press}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-davenport_role_2016}{}}%
-Davenport, Lauren D. 2016.
-{``\href{http://asr.sagepub.com/content/81/1/57.short}{The {Role} of
-{Gender}, {Class}, and {Religion} in {Biracial Americans}' {Racial
-Labeling Decisions}}.''} \emph{American Sociological Review}
-81(1):57--84.
-
-\leavevmode\vadjust pre{\hypertarget{ref-davis_who_1991}{}}%
-Davis, F. James. 1991. \emph{Who Is {Black}?: {One Nation}'s
-{Definition}}. {University Park, PA}: {Pennsylvania State University
-Press}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-davis_intermarriage_1941}{}}%
-Davis, Kingsley. 1941. {``Intermarriage in {Caste Societies}.''}
-\emph{American Anthropologist} 43:376--95.
-
-\leavevmode\vadjust pre{\hypertarget{ref-dimaggio_culture_1997}{}}%
-DiMaggio, Paul. 1997.
-{``\href{http://www.jstor.org/stable/2952552}{Culture and
-{Cognition}}.''} \emph{Annual Review of Sociology} 23:263--87.
-
-\leavevmode\vadjust pre{\hypertarget{ref-doyle_are_2007}{}}%
-Doyle, Jamie Mihoko, and Grace Kao. 2007.
-{``\href{http://www.jstor.org/stable/20141804}{Are {Racial Identities}
-of {Multiracials Stable}? {Changing Self-Identification} Among {Single}
-and {Multiple Race Individuals}}.''} \emph{Social Psychology Quarterly}
-70(4):405--23.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ellis_agents_2012}{}}%
-Ellis, Mark, Steven R. Holloway, Richard Wright, and Christopher S.
-Fowler. 2012. {``Agents of {Change}: {Mixed-Race Households} and the
-{Dynamics} of {Neighborhood Segregation} in the {United States}.''}
-\emph{Annals of the Association of American Geographers} 102(3):549--70.
-doi:
-\href{https://doi.org/10.1080/00045608.2011.627057}{10.1080/00045608.2011.627057}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-frederick_have_2008}{}}%
-Frederick, Carl B., and Robert M. Hauser. 2008. {``Have We Put an End to
-Social Promotion? {Changes} in School Progress Among Children Aged 6 to
-17 from 1972 to 2005.''} \emph{Demography} 45(3):719--40. doi:
-\href{https://doi.org/10.1353/dem.0.0015}{10.1353/dem.0.0015}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-fu_racial_2001}{}}%
-Fu, Vincent Kang. 2001. {``Racial {Intermarriage Pairings}.''}
-\emph{Demography} 38(2):147--59.
-
-\leavevmode\vadjust pre{\hypertarget{ref-graue_redshirting_2000}{}}%
-Graue, M. Elizabeth, and James DiPerna. 2000. {``Redshirting and {Early
-Retention}: {Who Gets} the "{Gift} of {Time}" and {What Are} {Its
-Outcomes}?''} \emph{American Educational Research Journal}
-37(2):509--34. doi:
-\href{https://doi.org/10.3102/00028312037002509}{10.3102/00028312037002509}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gullickson_education_2006}{}}%
-Gullickson, Aaron. 2006. {``Education and Black-White Interracial
-Marriage.''} \emph{Demography} 43(4):673--89. doi:
-\href{https://doi.org/10.1353/dem.2006.0033}{10.1353/dem.2006.0033}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gullickson_racial_2019}{}}%
-Gullickson, Aaron. 2019. {``The {Racial Identification} of {Young
-Adults} in a {Racially Complex Society}.''} \emph{Emerging Adulthood}
-7(2):150--61. doi:
-\href{https://doi.org/10.1177/2167696818790306}{10.1177/2167696818790306}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gullickson_choosing_2011}{}}%
-Gullickson, Aaron, and Ann Morning. 2011. {``Choosing Race:
-{Multiracial} Ancestry and Identification.''} \emph{Social Science
-Research} 40(2):498--512. doi:
-\href{https://doi.org/10.1016/j.ssresearch.2010.12.010}{10.1016/j.ssresearch.2010.12.010}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-herman_forced_2004}{}}%
-Herman, Melissa. 2004. {``Forced to {Choose}: {Some Determinants} of
-{Racial Identification} in {Multiracial Adolescents}.''} \emph{Child
-Development} 75(3):730--48. doi:
-\href{https://doi.org/10.2307/3696590}{10.2307/3696590}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-herman_blackwhiteother_2009}{}}%
-Herman, Melissa R. 2009. {``The {Black-White-Other Achievement Gap}:
-{Testing Theories} of {Academic Performance Among Multiracial} and
-{Monoracial Adolescents}.''} \emph{Sociology of Education}
-82(20):20--46.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hochschild_creating_2012}{}}%
-Hochschild, Jennifer L., Vesla M. Weaver, and Traci R. Burch. 2012.
-\emph{\href{https://books.google.com?id=BeTeF3iWG9sC}{Creating a {New
-Racial Order}: {How Immigration}, {Multiracialism}, {Genomics}, and the
-{Young Can Remake Race} in {America}}}. {Princeton University Press}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hollinger_postethnic_1995}{}}%
-Hollinger, David A. 1995. \emph{Postethnic {America}: {Beyond
-Multiculturalism}}. {New York, NY}: {Basic Books}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-holloway_place_2009}{}}%
-Holloway, Steven R., Richard Wright, Mark Ellis, and Margaret East.
-2009. {``Place, Scale and the Racial Claims Made for Multiracial
-Children in the 1990 {US Census}.''} \emph{Ethnic and Racial Studies}
-32(3):522--47.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hou_interracial_2011}{}}%
-Hou, Feng, and John Myles. 2011. {``Interracial Marriage and
-Status-Caste Exchange in {Canada} and the {United States}.''}
-\emph{Ethnic and Racial Studies}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hughes_effect_2017}{}}%
-Hughes, Jan N., Qian Cao, Stephen G. West, Paula Allee Smith, and
-Carissa Cerda. 2017. {``Effect of Retention in Elementary Grades on
-Dropping Out of School Early.''} \emph{Journal of School Psychology}
-65:11--27. doi:
-\href{https://doi.org/10.1016/j.jsp.2017.06.003}{10.1016/j.jsp.2017.06.003}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hughes_significance_1990}{}}%
-Hughes, Michael, and Bradley R. Hertel. 1990. {``The {Significance} of
-{Color Remains}: {A Study} of {Life Chances}, {Mate Selection}, and
-{Ethnic Consciousness Among} \{\vphantom\}{B}\vphantom\{\}lack
-\{\vphantom\}{A}\vphantom\{\}mericans.''} \emph{Social Forces}
-68(4):1105--20.
-
-\leavevmode\vadjust pre{\hypertarget{ref-iverson_regimes_2022}{}}%
-Iverson, Sarah, Ann Morning, Aliya Saperstein, and Janet Xu. 2022.
-{``Regimes Beyond the {One-Drop Rule}: {New Models} of {Multiracial
-Identity}.''} \emph{Genealogy} 6(2):57. doi:
-\href{https://doi.org/10.3390/genealogy6020057}{10.3390/genealogy6020057}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jimerson_metaanalysis_2001}{}}%
-Jimerson, Shane R. 2001. {``Meta-Analysis of {Grade Retention Research}:
-{Implications} for {Practice} in the 21st {Century}.''} \emph{School
-Psychology Review} 30(3):420--37. doi:
-\href{https://doi.org/10.1080/02796015.2001.12086124}{10.1080/02796015.2001.12086124}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kalmijn_educational_2010}{}}%
-Kalmijn, Matthijs. 2010. {``Educational Inequality, Homogamy, and Status
-Exchange in Black-White Intermarriage: A Comment on {Rosenfeld}.''}
-\emph{American Journal of Sociology} 115(4):1252--63.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kao_racial_1999}{}}%
-Kao, Grace. 1999. {``Racial {Identity} and {Academic Performance}: {An
-Examination} of {Biracial Asian} and {African American Youth}.''}
-\emph{Journal of Asian American Studies} 2(3):223--49. doi:
-\href{https://doi.org/10.1353/jaas.1999.0023}{10.1353/jaas.1999.0023}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-keith_skin_1991}{}}%
-Keith, Verna M., and Cedric Herring. 1991. {``Skin {Tone} and
-{Stratification} in the {Black Community}.''} \emph{American Journal of
-Sociology} 97(3):760--78.
-
-\leavevmode\vadjust pre{\hypertarget{ref-knol_mis_2011}{}}%
-Knol, Mirjam J., Wiebe R. Pestman, and Diederick E. Grobbee. 2011.
-{``The (Mis)use of Overlap of Confidence Intervals to Assess Effect
-Modification.''} \emph{European Journal of Epidemiology} 26(4):253--54.
-doi:
-\href{https://doi.org/10.1007/s10654-011-9563-8}{10.1007/s10654-011-9563-8}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-liebler_boundaries_2016}{}}%
-Liebler, Carolyn A. 2016. {``On the {Boundaries} of {Race}:
-{Identification} of {Mixed-heritage Children} in the {United States},
-1960 to 2010.''} \emph{Sociology of Race and Ethnicity} 2(4):548--68.
-doi:
-\href{https://doi.org/10.1177/2332649216632546}{10.1177/2332649216632546}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-liebler_america_2017}{}}%
-Liebler, Carolyn A., Sonya R. Porter, Leticia E. Fernandez, James M.
-Noon, and Sharon R. Ennis. 2017. {``America's {Churning Races}: {Race}
-and {Ethnicity Response Changes Between Census} 2000 and the 2010
-{Census}.''} \emph{Demography} 54(1):259--84. doi:
-\href{https://doi.org/10.1007/s13524-016-0544-0}{10.1007/s13524-016-0544-0}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-maddox_cognitive_2002a}{}}%
-Maddox, Keith B., and Stephanie A. Gray. 2002. {``Cognitive
-{Representations} of {Black Americans}: {Reexploring} the {Role} of
-{Skin Tone}.''} \emph{Personality and Social Psychology Bulletin}
-28(2):250--59. doi:
-\href{https://doi.org/10.1177/0146167202282010}{10.1177/0146167202282010}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-merton_intermarriage_1941}{}}%
-Merton, Robert K. 1941. {``Intermarriage and the {Social Structure}:
-{Fact} and {Theory}.''} \emph{Psychiatry} 4(3):361--74.
-
-\leavevmode\vadjust pre{\hypertarget{ref-monk_skin_2014}{}}%
-Monk, Ellis P. 2014.
-{``\href{https://www.jstor.org/stable/43287571}{Skin {Tone
-Stratification} Among {Black Americans}, 2001-2003}.''} \emph{Social
-Forces} 92(4):1313--37.
-
-\leavevmode\vadjust pre{\hypertarget{ref-morning_who_2000}{}}%
-Morning, Ann. 2000. {``Who Is {Multiracial}? {Definitions} and
-{Decisions}.''} \emph{Sociological Imagination} 37(4):209--29.
-
-\leavevmode\vadjust pre{\hypertarget{ref-morning_generational_2018}{}}%
-Morning, Ann, and Aliya Saperstein. 2018. {``The {Generational Locus} of
-{Multiraciality} and {Its Implications} for {Racial
-Self-Identification}.''} \emph{The ANNALS of the American Academy of
-Political and Social Science} 677(1):57--68. doi:
-\href{https://doi.org/10.1177/0002716218754774}{10.1177/0002716218754774}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pager_sociology_2008}{}}%
-Pager, Devah, and Hana Shepherd. 2008.
-{``\href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2915460}{The
-{Sociology} of {Discrimination}: {Racial Discrimination} in
-{Employment}, {Housing}, {Credit}, and {Consumer Markets}}.''}
-\emph{Annual Review of Sociology} 34:181--209.
-
-\leavevmode\vadjust pre{\hypertarget{ref-park_human_1928}{}}%
-Park, Robert Ezra. 1928. {``Human {Migration} and the {Marginal Man}.''}
-\emph{American Journal of Sociology} 33(8):881--92.
-
-\leavevmode\vadjust pre{\hypertarget{ref-qian_options_2004}{}}%
-Qian, Zenchao. 2004. {``Options: {Racial}/{Ethnic Identification} of
-{Children} of {Intermarried Couples}.''} \emph{Social Science Quarterly}
-85(3):746--66.
-
-\leavevmode\vadjust pre{\hypertarget{ref-quillian_new_2006}{}}%
-Quillian, Lincoln. 2006.
-{``\href{https://www.jstor.org/stable/29737741}{New {Approaches} to
-{Understanding Racial Prejudice} and {Discrimination}}.''} \emph{Annual
-Review of Sociology} 32:299--328.
-
-\leavevmode\vadjust pre{\hypertarget{ref-reskin_race_2012}{}}%
-Reskin, Barbara. 2012. {``The {Race Discrimination System}.''}
-\emph{Annual Review of Sociology} 38(1):17--35. doi:
-\href{https://doi.org/10.1146/annurev-soc-071811-145508}{10.1146/annurev-soc-071811-145508}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-root_racially_1992}{}}%
-Root, Maria P. P., ed. 1992. \emph{Racially {Mixed People} in
-{America}}. 1st edition. {Newbury Park, Calif}: {SAGE Publications,
-Inc}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rosenfeld_nontraditional_2010}{}}%
-Rosenfeld, Michael J. 2010.
-{``\href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3000058}{Nontraditional
-{Families} and {Childhood Progress Through School}}.''}
-\emph{Demography} 47(3):755--75.
-
-\leavevmode\vadjust pre{\hypertarget{ref-roth_end_2005a}{}}%
-Roth, Wendy D. 2005. {``\href{http://www.jstor.org/stable/4540881}{The
-{End} of the {One-Drop Rule}? {Labeling} of {Multiracial Children} in
-{Black Intermarriages}}.''} \emph{Sociological Forum} 20(1):35--67.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ruggles_ipums_2020}{}}%
-Ruggles, Steven, Sarah Flood, Ronald Goeken, Josiah Grover, Erin Meyer,
-Jose Pacas, and Matthew Sobek. 2020.
-\emph{\href{https://doi.org/10.18128/D010.V10.0}{{IPUMS USA}: {Version}
-10.0 {[}Dataset{]}}}. {Minneapolis, MN}: {IPUMS}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-saenz_persistence_1995}{}}%
-Saenz, Rogelio, Sean-Shong Hwang, Benigno E. Aguirre, and Robert N.
-Anderson. 1995. {``Persistence and {Change} in {Asian Identity Among
-Children} of {Intermarried Couples}.''} \emph{Sociological Perspectives}
-38(2):175--94.
-
-\leavevmode\vadjust pre{\hypertarget{ref-saperstein_mulatto_2013}{}}%
-Saperstein, Aliya, and Aaron Gullickson. 2013. {``A "{Mulatto Escape
-Hatch}" in the {United States}? {Examining Evidence} of {Racial} and
-{Social Mobility During} the {Jim Crow Era}.''} \emph{Demography}
-50(5):1921--42. doi:
-\href{https://doi.org/10.1007/s13524-013-0210-8}{10.1007/s13524-013-0210-8}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-saperstein_racial_2012}{}}%
-Saperstein, Aliya, and Andrew M. Penner. 2012. {``Racial {Fluidity} and
-{Inequality} in the {United States}.''} \emph{American Journal of
-Sociology} 118(3):676--727.
-
-\leavevmode\vadjust pre{\hypertarget{ref-shih_multiple_2019}{}}%
-Shih, Margaret, Leigh S. Wilton, Serena Does, Brianna M. Goodale, and
-Diana T. Sanchez. 2019. {``Multiple Racial Identities as Sources of
-Psychological Resilience.''} \emph{Social and Personality Psychology
-Compass} 13(6):e12469. doi:
-\href{https://doi.org/10.1111/spc3.12469}{10.1111/spc3.12469}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-snipp_american_1989}{}}%
-Snipp, C. Matthew. 1989.
-\emph{\href{https://books.google.com?id=E0CsvVoVA90C}{American
-{Indians}: {The First} of {This Land}}}. {Russell Sage Foundation}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-stearns_staying_2007}{}}%
-Stearns, Elizabeth, Stephanie Moller, Judith Blau, and Stephanie
-Potochnick. 2007. {``Staying {Back} and {Dropping Out}: {The
-Relationship Between Grade Retention} and {School Dropout}.''}
-\emph{Sociology of Education} 80(3):210--40. doi:
-\href{https://doi.org/10.1177/003804070708000302}{10.1177/003804070708000302}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-stonequist_problem_1935}{}}%
-Stonequist, Everett V. 1935. {``The {Problem} of the {Marginal Man}.''}
-\emph{American Journal of Sociology} 41(1):1--12.
-
-\leavevmode\vadjust pre{\hypertarget{ref-swidler_culture_1986}{}}%
-Swidler, Ann. 1986. {``Culture in {Action}: {Symbols} and
-{Strategies}.''} \emph{American Sociological Review} 51(2):273--86. doi:
-\href{https://doi.org/10.2307/2095521}{10.2307/2095521}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-telles_race_2009}{}}%
-Telles, Edward E., and Christina A. Sue. 2009. {``Race {Mixture}:
-{Boundary Crossing} in {Comparative Perspective}.''} \emph{Annual Review
-of Sociology} 35:129--46.
-
-\leavevmode\vadjust pre{\hypertarget{ref-udry_health_2003}{}}%
-Udry, J. Richard, Rose Maria Li, and Janet Hendrickson-Smith. 2003.
-{``Health and {Behavior Risks} of {Adolescents} with {Mixed-Race
-Identity}.''} \emph{American Journal of Public Health} 93(11):1865--70.
-
-\leavevmode\vadjust pre{\hypertarget{ref-warren_patterns_2014}{}}%
-Warren, John Robert, Emily Hoffman, and Megan Andrew. 2014. {``Patterns
-and {Trends} in {Grade Retention Rates} in the {United States},
-1995--2010.''} \emph{Educational Researcher} 43(9):433--43. doi:
-\href{https://doi.org/10.3102/0013189X14563599}{10.3102/0013189X14563599}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wolfe_land_2001}{}}%
-Wolfe, Patrick. 2001. {``Land, {Labor}, and {Difference}: {Elementary
-Structures} of {Race}.''} \emph{The American Historical Review}
-106(3):866--905.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xie_racial_1997}{}}%
-Xie, Yu, and Kimberly Goyette. 1997. {``The {Racial Identification} of
-{Biracial Children} with {One Asian Parent}: {Evidence} from the 1990
-{Census}.''} \emph{Social Forces} 76(2):547--70. doi:
-\href{https://doi.org/10.2307/2580724}{10.2307/2580724}.
-
-\end{CSLReferences}
-
-\end{document}
+\printbibliography[title=References]
+
+\processdelayedfloats\end{document}
